### 2018-03-07 latex/sc18/abstract.tex
splitting sections

\begin{abstract}
The abstract goes here.
\end{abstract}

### 2018-03-07 latex/sc18/algorithm.tex
splitting sections

\section{Algorithm} \label{sec:algorithm}

\subsection{Block Coordinate Descent}

\subsection{Dimension Trees}

\subsection{Parallel Algorithm}

\begin{algorithm}
\caption{$\lt[\Mn{H}{1},\dots,\Mn{H}{N}\rt] = \text{Par-NNCP-ALS}(\TA,k)$}
\label{alg:2D}
\begin{algorithmic}[1]
\Require $\TA$ is an $I_1\times \cdots \times I_N$ tensor distributed across a $P_1\times \cdots \times P_N$ grid of $P$ processors, so that $\TA_{\V{p}}$ is $(I_1/P_1)\times \cdots \times (I_N/P_N)$ and is owned by processor $\V{p}=(p_1,\dots,p_N)$, $k$ is rank of approximation
%\Require Local memory: space for $\Mn{H}{i}_{p_i}$ of size $(M_i/P_i) \times k$, for $1\leq i\leq N$
\State \Comment{Initialize data}
\State $a = \text{Norm-Squared}(\TA_{\V{p}})$
\State $\alpha = \text{All-Reduce}(a,\textsc{All-Procs})$
\For{$n=2$ to $N$}
	\State Initialize $\Mn{H}{n}_{\V{p}}$ of dimensions $(I_n/P)\times k$ 
	\State $\M{U} = \text{Local-SYRK}(\Mn{H}{n}_{\V{p}})$
	\State $\Mn{G}{n} = \text{All-Reduce}(\M{U},\textsc{All-Procs})$
	\State $\Mn{H}{n}_{p_n} = \text{All-Gather}(\Mn{H}{n}_{\V{p}},\textsc{Proc-Slice}(n,\VE{p}{n}))$
\EndFor
\State \Comment{Compute NNCP approximation}
\While{not converged}
	\State \Comment{Perform outer iteration of NNCP-ALS}
	\For{$n=1$ to $N$}
	\State \Comment{Compute new factor matrix in $n$th mode}
	\State $\M{V} = \text{Local-MTTKRP}(\TA_{p_1\cdots p_N},\{\Mn{H}{i}_{p_i}\},n)$
		\label{line:locMTTKRP}
	\State $\M{W} = \text{Reduce-Scatter}(\M{V},\textsc{Proc-Slice}(n,\VE{p}{n}))$ 
		\label{line:reduce-scatter}
	\State $\M{S} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$
		\label{line:hadamard}
	\State $\Mn[\hat]{H}{n}_{\V{p}} = \text{Local-NLS-Update}(\M{S},\M{W})$
		\label{line:locNLS}
	\State \Comment{Normalize columns}
	\State $\V{c} = \text{Local-Col-Norms}(\Mn[\hat]{H}{n}_{\V{p}})$
	\State $\V{\lambda} = \text{All-Reduce}(\V{c},\textsc{All-Procs})$
	\State $\Mn{H}{n}_{\V{p}} = \text{Local-Col-Scale}(\Mn[\hat]{H}{n}_{\V{p}},\V{\lambda})$
	\State \Comment{Organize data for later modes}
	\State $\M{U} = \text{Local-SYRK}(\Mn{H}{n}_{\V{p}})$
		\label{line:locSYRK}
	\State $\Mn{G}{n} = \text{All-Reduce}(\M{U},\textsc{All-Procs})$
		\label{line:all-reduce}
	\State $\Mn{H}{n}_{p_n} = \text{All-Gather}(\Mn{H}{n}_{\V{p}},\textsc{Proc-Slice}(n,\VE{p}{n}))$
		\label{line:all-gather}
	\EndFor
	\State \Comment{Compute error}
	\State $b = \text{Inner-Product}(\M{W},\Mn[\hat]{H}{N}_{\V{p}})$
	\State $\beta = \text{All-Reduce}(b,\textsc{All-Procs})$
	\State $\gamma = \V{\lambda}^\Tra (\M{S} \Hada \Mn{G}{N}) \V{\lambda}$
	\State $\epsilon = \sqrt{(\alpha-2\beta+\gamma)/\alpha}$ 
\EndWhile
\Ensure $\lt\|\TA - \Dsquare{\V{\lambda}; \Mn{H}{1},\dots,\Mn{H}{N}}\rt\| /\|\TA\| = \epsilon$
\Ensure Local matrices: $\Mn{H}{n}_{\V{p}}$ is $(I_n/P)\times k$ and owned by processor $\V{p}=(p_1,\dots,p_N)$, for $1\leq n \leq N$, $\V{\lambda}$ stored redundantly on every processor
\end{algorithmic}
\end{algorithm}

### 2018-03-07 latex/sc18/experiments.tex
splitting sections

\section{Performance Results} \label{sec:experiments}

\subsection{Datasets}

\subsubsection{Real World}
\subsubsection{Synthetic}

\subsection{Machine Detatils}

\subsection{Strong Scaling}

\subsection{Weak Scaling}

### 2018-03-07 latex/sc18/introduction.tex
splitting sections

\section{Introduction}

\grey{Contributions:
\begin{itemize}
	\item fastest scalable NNCP (and unsonstrained CP) software for dense tensors
	\item computation efficient using dimension trees, communication efficient using optimal parallel algorithm
	\item available open-source software, with flexibility to incorporate multiple NLS algorithms
\end{itemize}
}

### 2018-03-07 latex/sc18/algorithm.tex

@@
 
+\subsection{Relative Error Computation}
+
+Given a model $\T{M}=\CP$, we compute the relative error $\|\TA - \T{M}\|/\|\TA\|$ efficiently by using the identity $\|\TA-\T{M}\|^2 = \|\TA\|^2 - 2\langle \TA, \T{M} \rangle + \|\T{M}\|^2.$
+The quantity $\|\TA\|$ is fixed, and the other two terms can be computed cheaply given the temporary matrices computed during the course of the BCD algorithm.
+The second term can be computed using the identity $\langle \TA, \T{M} \rangle = \langle \Mn{M}{N}, \Mn{H}{N} \rangle$, where $\Mn{M}{N} = \Mz{A}{N} (\Mn{H}{N-1} \Khat \cdots \Khat \Mn{H}{1})$ is the MTTKRP result in the $N$th mode (the MTTKRP result is computed during the course of the algorithm to update the factor matrix).
+The third term can be computed using the identity $\|\T{M}\|^2 = \V{1}^\Tra(\M{S} \Hada $
+
 \subsection{Parallel Algorithm}
@@
 \begin{algorithm}
-\caption{$\lt[\Mn{H}{1},\dots,\Mn{H}{N}\rt] = \text{Par-NNCP-ALS}(\TA,k)$}
+\caption{$(\CPl,\epsilon) = \text{Par-NNCP}(\TA,k)$}
 \label{alg:2D}
@@
 \While{not converged}
-	\State \Comment{Perform outer iteration of NNCP-ALS}
+	\State \Comment{Perform outer iteration of BCD}
 	\For{$n=1$ to $N$}
@@
 		\label{line:locMTTKRP}
-	\State $\M{W} = \text{Reduce-Scatter}(\M{V},\textsc{Proc-Slice}(n,\VE{p}{n}))$ 
+	\State $\Mn{M}{n}_{\V{p}} = \text{Reduce-Scatter}(\M{V},\textsc{Proc-Slice}(n,\VE{p}{n}))$ 
 		\label{line:reduce-scatter}
@@
 		\label{line:hadamard}
-	\State $\Mn[\hat]{H}{n}_{\V{p}} = \text{Local-NLS-Update}(\M{S},\M{W})$
+	\State $\Mn[\hat]{H}{n}_{\V{p}} = \text{Local-NLS-Update}(\M{S},\Mn{M}{n}_{\V{p}})$
 		\label{line:locNLS}
@@
 	\EndFor
-	\State \Comment{Compute error}
-	\State $b = \text{Inner-Product}(\M{W},\Mn[\hat]{H}{N}_{\V{p}})$
+	\State \Comment{Compute relative error $\epsilon$ from mode-$N$ matrices}
+	\State $b = \text{Inner-Product}(\Mn{M}{N}_{\V{p}},\Mn[\hat]{H}{N}_{\V{p}})$
 	\State $\beta = \text{All-Reduce}(b,\textsc{All-Procs})$
@@
 \EndWhile
-\Ensure $\lt\|\TA - \Dsquare{\V{\lambda}; \Mn{H}{1},\dots,\Mn{H}{N}}\rt\| /\|\TA\| = \epsilon$
+\Ensure $\|\TA - \dsquare{\V{\lambda}; \Mn{H}{1},\dots,\Mn{H}{N}}\| /\|\TA\| = \epsilon$
 \Ensure Local matrices: $\Mn{H}{n}_{\V{p}}$ is $(I_n/P)\times k$ and owned by processor $\V{p}=(p_1,\dots,p_N)$, for $1\leq n \leq N$, $\V{\lambda}$ stored redundantly on every processor

### 2018-03-07 latex/sc18/algorithm.tex
minor updates

@@
-\section{Algorithm} \label{sec:algorithm}
+\section{Algorithm} 
+\label{sec:algorithm}
 
@@
 The quantity $\|\TA\|$ is fixed, and the other two terms can be computed cheaply given the temporary matrices computed during the course of the BCD algorithm.
-The second term can be computed using the identity $\langle \TA, \T{M} \rangle = \langle \Mn{M}{N}, \Mn{H}{N} \rangle$, where $\Mn{M}{N} = \Mz{A}{N} (\Mn{H}{N-1} \Khat \cdots \Khat \Mn{H}{1})$ is the MTTKRP result in the $N$th mode (the MTTKRP result is computed during the course of the algorithm to update the factor matrix).
-The third term can be computed using the identity $\|\T{M}\|^2 = \V{1}^\Tra(\M{S} \Hada $
+The second term can be computed using the identity $\langle \TA, \T{M} \rangle = \langle \Mn{M}{N}, \Mn{H}{N} \rangle$, where $\Mn{M}{N} = \Mz{A}{N} (\Mn{H}{N-1} \Khat \cdots \Khat \Mn{H}{1})$ is the MTTKRP result in the $N$th mode.
+The third term can be computed using the identity $\|\T{M}\|^2 = \V{1}^\Tra(\Mn{S}{N} \Hada \MnTra{H}{N} \Mn{H}{N})\V{1}$ where $\Mn{S}{N}=\MnTra{H}{1} \Mn{H}{1} \Hada \cdots \Hada \MnTra{H}{N-1} \Mn{H}{N-1}$.
+Both matrices $\Mn{M}{N}$ and $\Mn{S}{N}$ are computed during the course of the BCD algorithm for updating the factor matrix $\Mn{H}{N}$.
 
@@
 		\label{line:reduce-scatter}
-	\State $\M{S} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$
+	\State $\Mn{S}{N} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$
 		\label{line:hadamard}
-	\State $\Mn[\hat]{H}{n}_{\V{p}} = \text{Local-NLS-Update}(\M{S},\Mn{M}{n}_{\V{p}})$
+	\State $\Mn[\hat]{H}{n}_{\V{p}} = \text{Local-NLS-Update}(\Mn{S}{N},\Mn{M}{n}_{\V{p}})$
 		\label{line:locNLS}
@@
 	\State $\beta = \text{All-Reduce}(b,\textsc{All-Procs})$
-	\State $\gamma = \V{\lambda}^\Tra (\M{S} \Hada \Mn{G}{N}) \V{\lambda}$
+	\State $\gamma = \V{\lambda}^\Tra (\Mn{S}{N} \Hada \Mn{G}{N}) \V{\lambda}$
 	\State $\epsilon = \sqrt{(\alpha-2\beta+\gamma)/\alpha}$ 

### 2018-03-07 latex/sc18/prelims.tex
added lists after conversation

\section{Preliminaries} 
\label{sec:prelims}

\begin{itemize}
	\item tensor notation, Hadamard, Khatri-Rao, MTTKRP
	\item NNCP definition
	\item BCD/ALS
	\begin{itemize}
		\item MU, HALS, BPP, Nesterov, ADMM
	\end{itemize}
	\item MPI collectives
\end{itemize}

### 2018-03-07 latex/sc18/survey.tex
added lists after conversation

\section{Related Work} 
\label{sec:survey}

\grey{I haven't organized these very well...}
\begin{itemize}
	\item NLS algorithms
	\begin{itemize}
		\item MU, HALS, BPP, etc.
	\end{itemize}
	\item parallel sparse CP
	\begin{itemize}
		\item Karypis/Smith
		\item Kaya/Ucar
		\item Vagelis/Faloutsos?
		\item Li/Vuduc?
	\end{itemize}
	\item dense tensor computations
	\begin{itemize}
		\item Phan/Cichocki
		\item Austin/Ballard/Kolda IPDPS16
		\item Ballard/Knight/Rouse IPDPS18
	\end{itemize}
	\item parallel dense CP
	\begin{itemize}
		\item Liavs et al.
		\item Cichocki?
		\item distributed incremental descent?
	\end{itemize} 
\end{itemize}

### 2018-03-08 latex/sc18/algorithm.tex
added lists after conversation
finished error computation subsection
added algorithm figure, adapted from IPDPS18 paper

@@
+% !TEX root = paper.tex
+
 \section{Algorithm} 
 \label{sec:algorithm}
-
-\subsection{Block Coordinate Descent}
 
@@
 Both matrices $\Mn{M}{N}$ and $\Mn{S}{N}$ are computed during the course of the BCD algorithm for updating the factor matrix $\Mn{H}{N}$.
+The extra computation involved in computing the relative error is negligible.
+These identities have been used in previous work \cite{KB09,TensorBox,SK16}.
 
 \subsection{Parallel Algorithm}
+
+\input{fig/Alg_header.tex}
+\begin{figure*}[t]
+\centering
+  \subfloat[Start $n$th iteration with redundant subset of rows of each input matrix.]{\input{fig/AlgStart.tex}} \quad
+  \subfloat[Compute local MTTKRP for contribution to output matrix $\Mn{M}{2}$.]{\input{fig/AlgMTTKRP.tex}} \quad
+  \subfloat[Reduce-Scatter to compute and distribute rows of $\Mn{M}{2}$.]{\input{fig/AlgRS.tex}} \quad
+  \subfloat[Compute local NLS update to obtain $\Mn{H}{2}_{\V{p}}$ from $\Mn{M}{2}_{\V{p}}$ (and $\Mn{S}{2}$).]{\input{fig/AlgNLS.tex}} \quad
+  \subfloat[All-Gather to collect rows of $\Mn{H}{2}$ needed for later inner iterations.]{\input{fig/AlgAG.tex}}
+  \caption{Illustration of 2nd inner iteration of Par-NNCP algorithm for 3-way tensor on a $3\times3\times3$ processor grid, showing data distribution, communication, and computation across steps.
+Highlighted areas correspond to processor $(1,3,1)$ and its processor slice with which it communicates.
+The column normalization and computation of $\Mn{G}{2}$, which involve communication across all processors, is not shown here.}
+  \label{fig:par_staten} 
+\end{figure*}
 
@@
 \Require $\TA$ is an $I_1\times \cdots \times I_N$ tensor distributed across a $P_1\times \cdots \times P_N$ grid of $P$ processors, so that $\TA_{\V{p}}$ is $(I_1/P_1)\times \cdots \times (I_N/P_N)$ and is owned by processor $\V{p}=(p_1,\dots,p_N)$, $k$ is rank of approximation
-%\Require Local memory: space for $\Mn{H}{i}_{p_i}$ of size $(M_i/P_i) \times k$, for $1\leq i\leq N$
 \State \Comment{Initialize data}
@@
 	\State Initialize $\Mn{H}{n}_{\V{p}}$ of dimensions $(I_n/P)\times k$ 
-	\State $\M{U} = \text{Local-SYRK}(\Mn{H}{n}_{\V{p}})$
-	\State $\Mn{G}{n} = \text{All-Reduce}(\M{U},\textsc{All-Procs})$
+	\State $\M[\overline]{G} = \text{Local-SYRK}(\Mn{H}{n}_{\V{p}})$
+	\State $\Mn{G}{n} = \text{All-Reduce}(\M[\overline]{G},\textsc{All-Procs})$
 	\State $\Mn{H}{n}_{p_n} = \text{All-Gather}(\Mn{H}{n}_{\V{p}},\textsc{Proc-Slice}(n,\VE{p}{n}))$
@@
 	\State \Comment{Compute new factor matrix in $n$th mode}
-	\State $\M{V} = \text{Local-MTTKRP}(\TA_{p_1\cdots p_N},\{\Mn{H}{i}_{p_i}\},n)$
+	\State $\M[\overline]{M} = \text{Local-MTTKRP}(\TA_{p_1\cdots p_N},\{\Mn{H}{i}_{p_i}\},n)$
 		\label{line:locMTTKRP}
-	\State $\Mn{M}{n}_{\V{p}} = \text{Reduce-Scatter}(\M{V},\textsc{Proc-Slice}(n,\VE{p}{n}))$ 
+	\State $\Mn{M}{n}_{\V{p}} = \text{Reduce-Scatter}(\M[\overline]{M},\textsc{Proc-Slice}(n,\VE{p}{n}))$ 
 		\label{line:reduce-scatter}
-	\State $\Mn{S}{N} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$
+	\State $\Mn{S}{n} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$
 		\label{line:hadamard}
-	\State $\Mn[\hat]{H}{n}_{\V{p}} = \text{Local-NLS-Update}(\Mn{S}{N},\Mn{M}{n}_{\V{p}})$
+	\State $\Mn[\hat]{H}{n}_{\V{p}} = \text{Local-NLS-Update}(\Mn{S}{n},\Mn{M}{n}_{\V{p}})$
 		\label{line:locNLS}
 	\State \Comment{Normalize columns}
-	\State $\V{c} = \text{Local-Col-Norms}(\Mn[\hat]{H}{n}_{\V{p}})$
-	\State $\V{\lambda} = \text{All-Reduce}(\V{c},\textsc{All-Procs})$
+	\State $\V[\overline]{\lambda} = \text{Local-Col-Norms}(\Mn[\hat]{H}{n}_{\V{p}})$
+	\State $\V{\lambda} = \text{All-Reduce}(\V[\overline]{\lambda},\textsc{All-Procs})$
 	\State $\Mn{H}{n}_{\V{p}} = \text{Local-Col-Scale}(\Mn[\hat]{H}{n}_{\V{p}},\V{\lambda})$
 	\State \Comment{Organize data for later modes}
-	\State $\M{U} = \text{Local-SYRK}(\Mn{H}{n}_{\V{p}})$
+	\State $\M[\overline]{G} = \text{Local-SYRK}(\Mn{H}{n}_{\V{p}})$
 		\label{line:locSYRK}
-	\State $\Mn{G}{n} = \text{All-Reduce}(\M{U},\textsc{All-Procs})$
+	\State $\Mn{G}{n} = \text{All-Reduce}(\M[\overline]{G},\textsc{All-Procs})$
 		\label{line:all-reduce}
@@
 	\State \Comment{Compute relative error $\epsilon$ from mode-$N$ matrices}
-	\State $b = \text{Inner-Product}(\Mn{M}{N}_{\V{p}},\Mn[\hat]{H}{N}_{\V{p}})$
-	\State $\beta = \text{All-Reduce}(b,\textsc{All-Procs})$
+	\State $\overline{\beta} = \text{Inner-Product}(\Mn{M}{N}_{\V{p}},\Mn[\hat]{H}{N}_{\V{p}})$
+	\State $\beta = \text{All-Reduce}(\overline{\beta},\textsc{All-Procs})$
 	\State $\gamma = \V{\lambda}^\Tra (\Mn{S}{N} \Hada \Mn{G}{N}) \V{\lambda}$

### 2018-03-08 latex/sc18/fig/AlgMTTKRP.tex
added algorithm figure, adapted from IPDPS18 paper

% !TEX root = Alg_header.tex

\begin{tikzpicture}[x={(-0.5cm,-0.4cm)}, y={(1cm,0cm)}, z={(0cm,1cm)},every node/.append style={transform shape},scale=\parscale]

% highlight all-gather/reduce-scatter patterns

% local computation
% front face
\begin{scope}[canvas is yz plane at x=.5]
	% highlight block of 1st factor matrix now owned by processor
	\draw[fill=\highlight,shift={(-1,.5)},xscale=.5] (0,0) rectangle (-1,-1);
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=-90]
	% highlight block of 3rd factor matrix now owned by processor
	\draw[fill=\highlight,yscale=.5,shift={(-.5,-6)}] (0,0) rectangle (1,-1);
\end{scope}

% front face
\begin{scope}[canvas is yz plane at x=.5]
	% highlight block of 2nd factor matrix contribution computed by processor
	\draw[fill=\highlight,shift={(1.5,-3)},yscale=.5] (0,0) rectangle (1,-1);
\end{scope}

\parbasepic

\end{tikzpicture}

### 2018-03-12 latex/sc18/algorithm.tex
added discussion and figure for dimension trees

@@
 \subsection{Dimension Trees}
+
+An important optimization of the CP-ALS algorithm is to re-use temporary values across inner iterations \cite{PTC13a,KU16-TR,LCPSV17,Kaya17}.
+To illustrate the idea, consider a 3-way tensor $\T{X}$ approximated by $\dsquare{\M{U},\M{V},\M{W}}$ and the two MTTKRP computations $\underline{\Mz{X}{1}(\M{W}}\Khat\M{V})$ and $\underline{\Mz{X}{2}(\M{W}}\Khat\M{U})$ used to update factor matrices $\M{U}$ and $\M{V}$, respectively.
+The underlined parts of the expressions correspond to the shared dependence of the outputs on the tensor $\T{X}$ and the third factor matrix $\M{W}$.
+Indeed, a temporary quantity, which we refer to as a \emph{partial MTTKRP}, can be computed and re-used across the two MTTKRP expressions.
+We refer to the computation that combines the temporary quantity with the other factor matrix to complete the MTTKRP computation as a multi-tensor-times-vector or \emph{multi-TTV}, as it consists of multiple operations that multiply a tensor times a set of vectors, each corresponding to a different mode.
+This terminology was used in \cite{HBJT17-TR}.
+
+For a larger number of modes, a more general approach can organize the temporary quantities to be used over a maximal number of MTTKRPs.
+The general approach can yield significant benefit, decreasing the computation by a factor of approximately $N/2$ for dense $N$-way tensors.
+The idea is introduced in \cite{PTC13a}, but we adopt the terminology and notation of \emph{dimension trees} used in \cite{KU16-TR,Kaya17}.
+In this notation, the root node is labeled $\{1,\dots,N\}$ and corresponds to the original tensor, a leaf is labeled $\{n\}$ and corresponds to the $n$th MTTKRP result, and an internal node is labeled by a set of modes $\{i,\dots,j\}$ and corresponds to a temporary tensor whose values contribute to the MTTKRP results of modes $i,\dots,j$.
+
+\Cref{fig:DT} illustrates a dimension tree for the case $N=5$.
+Various shapes of binary trees are possible \cite{PTC13a,Kaya17}.
+For dense tensors, the computational cost is dominated by the root's branches, which correspond to partial MTTKRP computations.
+All other edges in a tree correspond to multi-TTVs and are typically much cheaper.
+
+\begin{figure}
+\input{fig/dim_tree.tex}
+\caption{Dimension tree example for $N=5$. 
+The data associated with the root node is the original tensor, the data associated with the leaf nodes are MTTKRP results, and the data associated with internal nodes are temporary tensors.  
+Edges labeled with PM correspond to partial MTTKRP computations, and edges labeled with mTTV correspond to multi-TTV computations.}
+\label{fig:DT}
+\end{figure}
 

### 2018-03-12 latex/sc18/fig/dim_tree.tex
added discussion and figure for dimension trees

% !TEX root = ../paper.tex

\begin{center}
\begin{tikzpicture}

\node (12345) at (4,2) {$\{1,2,3,4,5\}$};
\node (12) at (2,1) {$\{1,2\}$};
\node (345) at (6,1) {$\{3,4,5\}$};
\node (1) at (1,0) {$\{1\}$};
\node (2) at (3,0) {$\{2\}$};
\node (3) at (5,0) {$\{3\}$};
\node (45) at (7,0) {$\{4,5\}$};
\node (4) at (6,-1) {$\{4\}$};
\node (5) at (8,-1) {$\{5\}$};

\scriptsize
\path[draw] (12345) edge [left,align=right] node {PM} (12);
\path[draw] (12345) edge [right,align=left] node {PM} (345);
\path[draw] (12) edge [left,align=right] node {mTTV} (1);
\path[draw] (12) edge [right,align=left] node {mTTV} (2);
\path[draw] (345) edge [left,align=right] node {mTTV} (3);
\path[draw] (345) edge [right,align=left] node {mTTV} (45);
\path[draw] (45) edge [left,align=right] node {mTTV} (4);
\path[draw] (45) edge [right,align=left] node {mTTV} (5);
\normalsize

\end{tikzpicture}
\end{center}

### 2018-03-20 latex/sc18/abstract.tex
first draft of abstract

@@
+% !TEX root = paper.tex
+
 \begin{abstract}
-The abstract goes here.
+
+The CP tensor decomposition is a low-rank approximation of multidimensional (or tensor) data. 
+We present a distributed-memory parallel algorithm and implementation of an alternating optimization method for computing a CP decomposition of dense tensor data that can enforce nonnegativity of the computed low-rank factors.
+The principal task is to parallelize the matricized-tensor times Khatri-Rao product (MTTKRP) bottleneck subcomputation.
+The algorithm is computation efficient, using dimension trees to avoid redundant computation across MTTKRPs within the alternating method.
+Our approach is also communication efficient, using a data distribution and parallel algorithm across a multidimensional processor grid that can be tuned to minimize communication.
+We benchmark our software on synthetic as well as hyperspectral image data, demonstrating that our algorithm scales well to 100s of nodes (up to 9600 cores) and is faster than the currently available parallel software.
+
 \end{abstract}

### 2018-03-22 latex/sc18/plots.tex
added plots

% macros for plotting
\newcommand{\datafile}{}
\newcommand{\run}{}
\newcommand{\numiterations}{30}
\newcommand{\minvalue}{1}

% toggles whether or not to plot naive results (if not, the rows of the data file also need to be commented out)
\newif\ifnaive
% toggles ksweep vs scaling plot
\newif\ifksweep
% toggles legend
\newif\iflegend
% toggles y axis label
\newif\ifylabel
% toggles wider plot area
\newif\ifwider

\newcommand{\setcolors}{
\pgfplotsset{cycle list={
	red, fill=red \\ 
	blue, fill=blue \\ 
	green, fill=green \\ 
	red, pattern=crosshatch, pattern color=red \\
	blue, pattern=crosshatch, pattern color=blue \\
	green, pattern=crosshatch, pattern color=green \\
}};
}

% set options for grouped bar plot
\newcommand{\plotoptions}{
	ybar stacked,
	reverse legend,
	bar width=6pt,
	width=9cm, height=3.85cm,
	ylabel={Time (seconds)}, 
	y label style={yshift=-.5cm},
	ymin=0,
	\ifnaive 
		\ifksweep
			symbolic x coords={10-0,10-1,10-2,10-3,,20-0,20-1,20-2,20-3,,30-0,30-1,30-2,30-3,,40-0,40-1,40-2,40-3,,50-0,50-1,50-2,50-3},
		\else
			symbolic x coords={16-0,16-1,16-2,16-3,,96-0,96-1,96-2,96-3,,384-0,384-1,384-2,384-3,,864-0,864-1,864-2,864-3,,1536-0,1536-1,1536-2,1536-3},
		\fi
		xticklabels={MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive},
	\else  
		\ifksweep
			symbolic x coords={10-0,10-1,10-2,,,20-0,20-1,20-2,,,30-0,30-1,30-2,,,40-0,40-1,40-2,,,50-0,50-1,50-2,},
		\else
			symbolic x coords={16-0,16-1,16-2,,,96-0,96-1,96-2,,,384-0,384-1,384-2,,,864-0,864-1,864-2,,,1536-0,1536-1,1536-2,}, 
		\fi
		xticklabels={MU,HALS,ABPP,MU,HALS,ABPP,MU,HALS,ABPP,MU,HALS,ABPP,MU,HALS,ABPP},
	\fi
	xtick=data,
	xticklabel style={xshift=.15cm,rotate=75,anchor=east},
	\ifksweep
		xlabel={Low Rank ($k$)}, 
	\else
		xlabel={Number of Processes ($p$)},
	\fi
	xlabel style={yshift=-0.75cm},
	%legend style={at={(0.4,-0.9)},anchor=north},
	legend style={at={(0.5,1.3)},anchor=north},
	legend columns=-1,
	legend style={draw=none, cells={align=left}, nodes={scale=0.8}}
}

% grouped bar plot command
\newcommand{\makeplot}{
\begin{axis}[\plotoptions]
	\setcolors
	\addplot table[x=algo, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{gram}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{allgather}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{reducescatter}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
	\iflegend
		\legend{MM, \NLS, Gram, All-Gather, Reduce-Scatter, All-Reduce};
	\fi
\end{axis}
}

% labels for bar groups (manually positioned)
\newcommand{\labels}{
%\node [align=center,text width=3cm] at (1.25cm, -.95cm)   {\ifksweep 10 \else 16 \fi};
%\node [align=center,text width=3cm] at (3cm, -0.95cm)   {\ifksweep 20 \else 96 \fi};
%\node [align=center,text width=3cm] at (4.6cm, -0.95cm)   {\ifksweep 30 \else 384 \fi};
%\node [align=center,text width=3cm] at (6.3cm, -0.95cm) {\ifksweep 40 \else 864 \fi};
%\node [align=center,text width=3cm] at (8.15cm, -0.95cm) {\ifksweep 50 \else 1536 \fi};
\node [align=center,text width=3cm] at (1cm, -1.15cm)   {\ifksweep 10 \else 16 \fi};
\node [align=center,text width=3cm] at (2.4cm, -1.15cm)   {\ifksweep 20 \else 96 \fi};
\node [align=center,text width=3cm] at (3.7cm, -1.15cm)   {\ifksweep 30 \else 384 \fi};
\node [align=center,text width=3cm] at (5.1cm, -1.15cm) {\ifksweep 40 \else 864 \fi};
\node [align=center,text width=3cm] at (6.4cm, -1.15cm) {\ifksweep 50 \else 1536 \fi};
}

% allows for filtering rows from dat file
\pgfplotsset{
    discard if not/.style 2 args={
        x filter/.append code={
            \edef\tempa{\thisrow{#1}}
            \edef\tempb{#2}
            \ifx\tempa\tempb
            \else
                \def\pgfmathresult{inf}
            \fi
        }
    }
}

% set options for relative error over time plot
\newcommand{\relerrplotoptions}{
	xlabel=Time (s),
	\ifylabel
		ylabel=Relative Error,
		ylabel style={yshift=-.3cm},
	\fi
	\ifwider
		width=1.8in, height=2in,
	\else
		width=1.35in, height=2in,
	\fi
	y tick label style={
        /pgf/number format/.cd,
            precision=4,
	},
	legend style={draw=none, cells={align=left}, nodes={scale=0.7}}
}

% plot relative error over time, filtering rows out appropriately
\newcommand{\relerrplot}{
\begin{axis}[\relerrplotoptions]
	\addplot+ [discard if not={algo}{liavas},discard if not={runs}{\run},green,thick,mark options={solid}] table [x={totaltime}, y={relerr}] {\datafile};
	\addplot+ [discard if not={algo}{nodimtree},discard if not={runs}{\run},red,thick,mark options={solid},mark=square*] table [x={totaltime}, y={relerr}] {\datafile};
	\addplot+ [discard if not={algo}{dimtree},discard if not={runs}{\run},blue,thick,mark options={solid},mark=triangle*] table [x={totaltime}, y={relerr}] {\datafile};
	\iflegend
		\legend{Liavas, Classic, DimTree}
	\fi
\end{axis}
}

\newcommand{\strongscalingplot}{
\begin{axis}[\relerrplotoptions]
	\addplot+ [discard if not={algo}{liavas},discard if not={runs}{\run},green,thick,mark options={solid}] table [x={nodes}, y={totaltime}] {\datafile};
	\addplot+ [discard if not={algo}{nodimtree},discard if not={runs}{\run},red,thick,mark options={solid},mark=square*] table [x={nodes}, y={totaltime}] {\datafile};
	\addplot+ [discard if not={algo}{dimtree},discard if not={runs}{\run},blue,thick,mark options={solid},mark=triangle*] table [x={nodes}, y={totaltime}] {\datafile};
	\iflegend
		\legend{Liavas, Classic, DimTree}
	\fi
\end{axis}
}

### 2018-03-22 latex/sc18/plots.tex
starting to put together plots

@@
 \newcommand{\datafile}{}
-\newcommand{\run}{}
-\newcommand{\numiterations}{30}
-\newcommand{\minvalue}{1}
+\newcommand{\alg}{}
+%\newcommand{\numiterations}{30}
+%\newcommand{\minvalue}{1}
 
@@
 
-% set options for relative error over time plot
-\newcommand{\relerrplotoptions}{
+% set options for strongscaling plot
+\newcommand{\strongscalingplotoptions}{
 	xlabel=Time (s),
@@
 
-% plot relative error over time, filtering rows out appropriately
-\newcommand{\relerrplot}{
-\begin{axis}[\relerrplotoptions]
-	\addplot+ [discard if not={algo}{liavas},discard if not={runs}{\run},green,thick,mark options={solid}] table [x={totaltime}, y={relerr}] {\datafile};
-	\addplot+ [discard if not={algo}{nodimtree},discard if not={runs}{\run},red,thick,mark options={solid},mark=square*] table [x={totaltime}, y={relerr}] {\datafile};
-	\addplot+ [discard if not={algo}{dimtree},discard if not={runs}{\run},blue,thick,mark options={solid},mark=triangle*] table [x={totaltime}, y={relerr}] {\datafile};
+% plot time over p, filtering rows out appropriately
+\newcommand{\strongscalingplot}{
+\begin{axis}[\strongscalingplotoptions]
+	%\addplot+ [discard if not={algo}{liavas},thick,mark options={solid}] table [x={nodes}, y={totaltime}] {\datafile};
+	%\addplot+ [discard if not={algo}{nodimtree},thick,mark options={solid},mark=square*] table [x={nodes}, y={totaltime}] {\datafile};
+	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=triangle*] table [x={p}, y={total_d}] {\datafile};
 	\iflegend
@@
 }
-
-\newcommand{\strongscalingplot}{
-\begin{axis}[\relerrplotoptions]
-	\addplot+ [discard if not={algo}{liavas},discard if not={runs}{\run},green,thick,mark options={solid}] table [x={nodes}, y={totaltime}] {\datafile};
-	\addplot+ [discard if not={algo}{nodimtree},discard if not={runs}{\run},red,thick,mark options={solid},mark=square*] table [x={nodes}, y={totaltime}] {\datafile};
-	\addplot+ [discard if not={algo}{dimtree},discard if not={runs}{\run},blue,thick,mark options={solid},mark=triangle*] table [x={nodes}, y={totaltime}] {\datafile};
-	\iflegend
-		\legend{Liavas, Classic, DimTree}
-	\fi
-\end{axis}
-}

### 2018-03-22 latex/sc18/plots.tex
fixing up the strong scaling plot

@@
+% !TEX root = experiments.tex
+
 % macros for plotting
@@
 \newcommand{\strongscalingplotoptions}{
-	xlabel=Time (s),
-	\ifylabel
-		ylabel=Relative Error,
-		ylabel style={yshift=-.3cm},
-	\fi
-	\ifwider
-		width=1.8in, height=2in,
-	\else
-		width=1.35in, height=2in,
-	\fi
+	log basis y={2},
+	log basis x={2},
+	xlabel=Nodes,
+	ylabel=Time (s),
 	y tick label style={
@@
 \newcommand{\strongscalingplot}{
-\begin{axis}[\strongscalingplotoptions]
+\begin{loglogaxis}[\strongscalingplotoptions]
 	%\addplot+ [discard if not={algo}{liavas},thick,mark options={solid}] table [x={nodes}, y={totaltime}] {\datafile};
@@
 	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=triangle*] table [x={p}, y={total_d}] {\datafile};
-	\iflegend
-		\legend{Liavas, Classic, DimTree}
-	\fi
-\end{axis}
+	\legend{DimTree}
+\end{loglogaxis}
 }

### 2018-03-22 common/distutils.hpp
change to omp_get_wtime and timed dimension trees

@@ -9,13 +9,16 @@
 #include "utils.hpp"
 
 inline void mpitic() {
-    tictoc_stack.push(clock());
+    // tictoc_stack.push(clock());
+    tictoc_stack_omp_clock.push(omp_get_wtime());
 }
 
 inline void mpitic(int rank) {
-    double temp = clock();    
+    // double temp = clock();
+    double temp = omp_get_wtime();
     std::cout << "tic::" << rank << "::" << temp << std::endl;
-    tictoc_stack.push(temp);
+    // tictoc_stack.push(temp);
+    tictoc_stack_omp_clock.push(omp_get_wtime());
 }
 
 inline double mpitoc(int rank) {
@@ -23,10 +26,12 @@ inline double mpitoc(int rank) {
     MPI_Barrier(MPI_COMM_WORLD);
 #endif
     std::cout << "toc::" << rank << "::"
-         << tictoc_stack.top() << std::endl;
-    double rc = (static_cast<double>
-                 (clock() - tictoc_stack.top())) / CLOCKS_PER_SEC;
-    tictoc_stack.pop();
+              << tictoc_stack_omp_clock.top() << std::endl;
+    // double rc = (static_cast<double>
+    //              (clock() - tictoc_stack.top())) / CLOCKS_PER_SEC;
+    //tictoc_stack.pop();
+    double rc = omp_get_wtime() - tictoc_stack_omp_clock.top();
+    tictoc_stack_omp_clock.pop();
     return rc;
 }
 
@@ -34,9 +39,11 @@ inline double mpitoc() {
 #ifdef __WITH__BARRIER__TIMING__
     MPI_Barrier(MPI_COMM_WORLD);
 #endif
-    double rc = (static_cast<double>
-                 (clock() - tictoc_stack.top())) / CLOCKS_PER_SEC;
-    tictoc_stack.pop();
+    // double rc = (static_cast<double>
+    //              (clock() - tictoc_stack.top())) / CLOCKS_PER_SEC;
+    // tictoc_stack.pop();
+    double rc = omp_get_wtime() - tictoc_stack_omp_clock.top();
+    tictoc_stack_omp_clock.pop();
     return rc;
 }
 

### 2018-03-22 common/utils.hpp

@@ -23,6 +23,7 @@ static ULONG powersof10[16] = {
 };
 
 static std::stack<clock_t> tictoc_stack;
+static std::stack<double> tictoc_stack_omp_clock;
 
 inline void tic() {
     tictoc_stack.push(clock());

### 2018-03-22 distntf/distauntf.hpp

@@ -198,9 +198,15 @@ class DistAUNTF {
         }
         MPITIC;  // mttkrp tic
         if (this->m_enable_dim_tree) {
+            double multittv_time = 0;
+            double mttkrp_time = 0;
             kdt->in_order_reuse_MTTKRP(current_mode,
                                        ncp_mttkrp_t[current_mode].memptr(),
-                                       false);
+                                       false, multittv_time, mttkrp_time);
+            this->time_stats.compute_duration(multittv_time);
+            this->time_stats.compute_duration(mttkrp_time);
+            this->time_stats.multittv_duration(multittv_time);
+            this->time_stats.mttkrp_duration(mttkrp_time);
 
         } else {
             m_input_tensor.mttkrp(current_mode, ncp_krp[current_mode],
@@ -320,6 +326,7 @@ class DistAUNTF {
         this->reportTime(this->time_stats.gram_duration(), "total_gram");
         this->reportTime(this->time_stats.krp_duration(), "total_krp");
         this->reportTime(this->time_stats.mttkrp_duration(), "total_mttkrp");
+        this->reportTime(this->time_stats.multittv_duration(), "total_multittv");
         this->reportTime(this->time_stats.nnls_duration(), "total_nnls");
         if (this->m_compute_error) {
             this->reportTime(this->time_stats.err_compute_duration(),
@@ -345,7 +352,7 @@ class DistAUNTF {
         m_local_ncp_factors_t(i_local_dims, i_k, true),
         m_gathered_ncp_factors(i_tensor.dimensions(), i_k, false),
         m_gathered_ncp_factors_t(i_tensor.dimensions(), i_k, true),
-        time_stats(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0) {
+        time_stats(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0) {
         this->m_compute_error = false;
         this->m_enable_dim_tree = false;
         this->m_num_it = 30;
@@ -464,7 +471,7 @@ class DistAUNTF {
         // rel_Error = sqrt(max(init.nr_X^2 + lambda^T * Hadamard of all gram * lambda - 2 * innerprod(X,F_kten),0))/init.nr_X;
         MPITIC;  // err compute
         hadamard_all_grams = global_gram % factor_global_grams[this->m_modes - 1];
-        VEC local_lambda = m_local_ncp_factors.lambda();        
[…]

### 2018-03-22 latex/sc18/plots.tex
added start of breakdown plot

@@
 \newcommand{\alg}{}
-%\newcommand{\numiterations}{30}
-%\newcommand{\minvalue}{1}
+\newcommand{\numiterations}{30}
+\newcommand{\minvalue}{1}
 
@@
 % set options for grouped bar plot
-\newcommand{\plotoptions}{
+\newcommand{\breakdownplotoptions}{
 	ybar stacked,
@@
 	ymin=0,
-	\ifnaive 
-		\ifksweep
-			symbolic x coords={10-0,10-1,10-2,10-3,,20-0,20-1,20-2,20-3,,30-0,30-1,30-2,30-3,,40-0,40-1,40-2,40-3,,50-0,50-1,50-2,50-3},
-		\else
-			symbolic x coords={16-0,16-1,16-2,16-3,,96-0,96-1,96-2,96-3,,384-0,384-1,384-2,384-3,,864-0,864-1,864-2,864-3,,1536-0,1536-1,1536-2,1536-3},
-		\fi
-		xticklabels={MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive},
-	\else  
-		\ifksweep
-			symbolic x coords={10-0,10-1,10-2,,,20-0,20-1,20-2,,,30-0,30-1,30-2,,,40-0,40-1,40-2,,,50-0,50-1,50-2,},
-		\else
-			symbolic x coords={16-0,16-1,16-2,,,96-0,96-1,96-2,,,384-0,384-1,384-2,,,864-0,864-1,864-2,,,1536-0,1536-1,1536-2,}, 
-		\fi
-		xticklabels={MU,HALS,ABPP,MU,HALS,ABPP,MU,HALS,ABPP,MU,HALS,ABPP,MU,HALS,ABPP},
-	\fi
+	symbolic x coords={1,2,4,8,16,32},
 	xtick=data,
-	xticklabel style={xshift=.15cm,rotate=75,anchor=east},
-	\ifksweep
-		xlabel={Low Rank ($k$)}, 
-	\else
-		xlabel={Number of Processes ($p$)},
-	\fi
-	xlabel style={yshift=-0.75cm},
-	%legend style={at={(0.4,-0.9)},anchor=north},
-	legend style={at={(0.5,1.3)},anchor=north},
-	legend columns=-1,
-	legend style={draw=none, cells={align=left}, nodes={scale=0.8}}
 }
 
-% grouped bar plot command
-\newcommand{\makeplot}{
-\begin{axis}[\plotoptions]
+% stacked bar plot command
+\newcommand{\breakdownplot}{
+\begin{axis}[\breakdownplotoptions]
 	\setcolors
-	\addplot table[x=algo, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{gram}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{allgather}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{reducescatter}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
-	\iflegend
-		\legend{MM, \NLS, Gram, All-Gather, Reduce-Scatter, All-Reduce};
-	\fi
+	\addplot table[x=p, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=p, y expr=(\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=p, y expr=(\thisrow{gram}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=p, y expr=(\thisrow{allgather}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=p, y expr=(\thisrow{reducescatter}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=p, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
 \end{axis}
@@
 	%\addplot+ [discard if not={algo}{nodimtree},thick,mark options={solid},mark=square*] table [x={nodes}, y={totaltime}] {\datafile};
-	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=triangle*] table [x={p}, y={total_d}] {\datafile};
+	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=triangle*] table [x={p}, y={total}] {\datafile};
 	\legend{DimTree}

### 2018-03-23 latex/sc18/plots.tex
added some data and weak scaling breakdown plot

@@
 	ymin=0,
-	symbolic x coords={1,2,4,8,16,32},
+	symbolic x coords={1,16,81,256},
 	xtick=data,
+	legend={DimTree}
 }

### 2018-03-25 latex/sc18/plots.tex
working on plot formatting

--- 2018-03-22 latex/sc18/plots.tex
@@
 \newif\ifwider
+% toggles including liavas results
+\newif\ifliavas
 
@@
 	width=9cm, height=3.85cm,
-	ylabel={Time (seconds)}, 
-	y label style={yshift=-.5cm},
+	ylabel={Time (s)}, 
+	%y label style={yshift=-.5cm},
 	ymin=0,
-	symbolic x coords={1,2,4,8,16,32},
+	symbolic x coords={1,16,81,256},
 	xtick=data,
@@
 	\setcolors
-	\addplot table[x=p, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=p, y expr=(\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=p, y expr=(\thisrow{gram}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=p, y expr=(\thisrow{allgather}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=p, y expr=(\thisrow{reducescatter}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=p, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=p, y expr=((\thisrow{mttkrp}+\thisrow{krp})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=p, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=p, y expr=((\thisrow{nnls}+\thisrow{gram}+\thisrow{allreduce})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=p, y expr=((\thisrow{err_compute}+\thisrow{err_communication})/(\minvalue*\numiterations))] {\datafile};
+	\legend{MTTRKP,Factor Comm,NLS,Error};
 \end{axis}
@@
 \begin{loglogaxis}[\strongscalingplotoptions]
-	%\addplot+ [discard if not={algo}{liavas},thick,mark options={solid}] table [x={nodes}, y={totaltime}] {\datafile};
-	%\addplot+ [discard if not={algo}{nodimtree},thick,mark options={solid},mark=square*] table [x={nodes}, y={totaltime}] {\datafile};
+	%\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={p}, y={totaltime}] {\datafile};
 	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=triangle*] table [x={p}, y={total}] {\datafile};
-	\legend{DimTree}
+	\ifliavas
+		\addplot+ [discard if not={alg}{L},thick,mark options={solid}] table [x={p}, y={total}] {\datafile};
+	\fi
+	\legend{DimTree,Liavas};
 \end{loglogaxis}

### 2018-03-25 latex/sc18/experiments.tex
added algorithm figure, adapted from IPDPS18 paper
added plots
starting to put together plots
fixing up the strong scaling plot
added start of breakdown plot
added some data and weak scaling breakdown plot
working on plot formatting
formatting plots

@@
+% !TEX root = paper.tex
+
 \section{Performance Results} \label{sec:experiments}
+
+\input{plots}
 
@@
 
-\subsection{Machine Detatils}
+\subsection{Machine Details}
 
@@
 
+\begin{itemize}
+	\item 3D HSI (3 algs)
+	\item 3D synthetic (3 algs)
+	\item 5D synthetic (2 algs)
+\end{itemize}
+
+\begin{figure}
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/str_3D_syn.dat}
+\renewcommand{\numiterations}{42}
+\liavastrue
+\strongscalingplot
+\end{tikzpicture}
+\caption{Strong scaling of 3D synthetic tensor with dimension $1024\times 1024\times 1024$ on processor grids $2^k\times 2^k\times 2^k$ for $k\in\{0,1,2,3\}$.  The rank is fixed at 32.}
+\end{figure}
+
+\begin{figure}
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/str_5D_syn.dat}
+\renewcommand{\numiterations}{10}
+\liavasfalse
+\strongscalingplot
+\end{tikzpicture}
+\caption{Strong scaling of 5D synthetic tensor with dimension $64\times 64\times 64\times 64\times 64$ on processor grids $1\times1\times1\times1\times1$, $2\times1\times1\times1\times1$, $\dots$, $2\times2\times2\times2\times2$.
+The rank is fixed at 32.}
+\end{figure}
+
 \subsection{Weak Scaling}
+
+\begin{itemize}
+	\item 4D synthetic (2 algs)
+	\item 3D neuroscience, increasing one mode? (2 algs)
+\end{itemize}
+
+\begin{figure}
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/wk_4D_syn.dat}
+\renewcommand{\numiterations}{10}
+\breakdownplot
+%\labels
+\end{tikzpicture}
+\caption{Weak scaling of 4D synthetic tensors with (D) and without (N) the use of dimension trees.
+The tensor and processor grid dimensions are $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$, and the rank is fixed at 32.
+The reported times are per iteration.}
+\end{figure}
+
+\subsection{Varying Approximation Rank}
+
+\begin{itemize}
+	\item 4D HSI (2 algs)
+	\item 3D neuroscience (2 algs)
+\end{itemize}
+
+%\begin{figure}
+
+% set which run to plot
+%\renewcommand{\run}{3}
+%
+%\begin{subfigure}{0.3 \columnwidth}
+%\ylabeltrue
+%\begin{tikzpicture}
+%\renewcommand{\datafile}{data/denserwerr-time.dat}
+%\renewcommand{\run}{1}
+%\relerrplot
+%\renewcommand{\run}{3}
+%\end{tikzpicture}
+%\ylabelfalse
+%\subcaption{Video}
+%\label{fig:denserwerr}
+%\end{subfigure}
+%~
+%\begin{subfigure}{0.3 \columnwidth}
+%\begin{tikzpicture}
+%\renewcommand{\datafile}{data/stkx-5runs-err.dat}
+%\legendtrue
+%\relerrplot
+%\legendfalse
+%\end{tikzpicture}
+%\subcaption{Stack Exchange}
+%\label{fig:stackexchangeerr}
+%\end{subfigure}
+%~
+%\begin{subfigure}{0.3 \columnwidth}
+%\begin{tikzpicture}
+%\renewcommand{\datafile}{data/webbase1M-5runs-err.dat}
+%\relerrplot
+%\end{tikzpicture}
+%\subcaption{Webbase}
+%\label{fig:sparserwerr}
+%\end{subfigure}
+%
+%\caption{Relative error comparison of \MU, \HALS, \BPP\ on real world datasets.}
+%\label{fig:convergence}
+%\end{figure}
+
+% set which run to plot
+%\renewcommand{\run}{3}
+%
+%\begin{subfigure}{0.3 \columnwidth}
+%\ylabeltrue
+%\begin{tikzpicture}
+%\renewcommand{\datafile}{data/denserwerr-time.dat}
+%\renewcommand{\run}{1}
+%\strongscalingplot
+%\renewcommand{\run}{3}
+%\end{tikzpicture}
+%\ylabelfalse
+%\subcaption{Video}
+%\label{fig:denserwerr}
+%\end{subfigure}
+%~
+%\begin{subfigure}{0.3 \columnwidth}
+%\begin{tikzpicture}
+%\renewcommand{\datafile}{data/stkx-5runs-err.dat}
+%\legendtrue
+%\strongscalingplot
+%\legendfalse
+%\end{tikzpicture}
+%\subcaption{Stack Exchange}
+%\label{fig:stackexchangeerr}
+%\end{subfigure}
+%~
+%\begin{subfigure}{0.3 \columnwidth}
+%\begin{tikzpicture}
+%\renewcommand{\datafile}{data/webbase1M-5runs-err.dat}
+%\strongscalingplot
+%\end{tikzpicture}
+%\subcaption{Webbase}
+%\label{fig:sparserwerr}
+%\end{subfigure}
+%
+%\caption{Relative error comparison of \MU, \HALS, \BPP\ on real world datasets.}
+%\label{fig:convergence}
+%\end{figure}

### 2018-03-25 latex/sc18/plots.tex
formatting plots

@@
 	reverse legend,
-	bar width=6pt,
+	bar width=8pt,
 	width=9cm, height=3.85cm,
@@
 	ymin=0,
-	symbolic x coords={1,16,81,256},
+	symbolic x coords={D1,N1,,D16,N16,,D81,N81,,D256},
 	xtick=data,
+	legend style={at={(0.5,1.3)},anchor=north},
+	legend columns=-1,
+	reverse legend
 }
@@
 	\setcolors
-	\addplot table[x=p, y expr=((\thisrow{mttkrp}+\thisrow{krp})/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=p, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=p, y expr=((\thisrow{nnls}+\thisrow{gram}+\thisrow{allreduce})/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=p, y expr=((\thisrow{err_compute}+\thisrow{err_communication})/(\minvalue*\numiterations))] {\datafile};
-	\legend{MTTRKP,Factor Comm,NLS,Error};
+	\addplot table[x=alg-p, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-p, y expr=(\thisrow{krp}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-p, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-p, y expr=((\thisrow{nnls}+\thisrow{gram}+\thisrow{allreduce})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-p, y expr=((\thisrow{err_compute}+\thisrow{err_communication})/(\minvalue*\numiterations))] {\datafile};
+	\legend{MTTRKP,KRP,Factor Comm,NLS,Error};
 \end{axis}
@@
 %\node [align=center,text width=3cm] at (8.15cm, -0.95cm) {\ifksweep 50 \else 1536 \fi};
-\node [align=center,text width=3cm] at (1cm, -1.15cm)   {\ifksweep 10 \else 16 \fi};
-\node [align=center,text width=3cm] at (2.4cm, -1.15cm)   {\ifksweep 20 \else 96 \fi};
-\node [align=center,text width=3cm] at (3.7cm, -1.15cm)   {\ifksweep 30 \else 384 \fi};
-\node [align=center,text width=3cm] at (5.1cm, -1.15cm) {\ifksweep 40 \else 864 \fi};
-\node [align=center,text width=3cm] at (6.4cm, -1.15cm) {\ifksweep 50 \else 1536 \fi};
+\node [align=center,text width=3cm] at (1cm, -1.15cm)   {\ifksweep 10 \else 1 \fi};
+\node [align=center,text width=3cm] at (2.4cm, -1.15cm)   {\ifksweep 20 \else 16 \fi};
+\node [align=center,text width=3cm] at (3.7cm, -1.15cm)   {\ifksweep 30 \else 81 \fi};
+\node [align=center,text width=3cm] at (5.1cm, -1.15cm) {\ifksweep 40 \else 256 \fi};
 }
@@
 	%\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={p}, y={totaltime}] {\datafile};
-	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=triangle*] table [x={p}, y={total}] {\datafile};
+	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=triangle*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
+	\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
 	\ifliavas
-		\addplot+ [discard if not={alg}{L},thick,mark options={solid}] table [x={p}, y={total}] {\datafile};
+		\addplot+ [discard if not={alg}{L},thick,mark options={solid}] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
+		\legend{DimTree,NoDimTree,Liavas}
+	\else
+		\legend{DimTree,NoDimTree};
 	\fi
-	\legend{DimTree,Liavas};
 \end{loglogaxis}

### 2018-03-26 latex/slides/talk.tex
started creating slides

\documentclass[xcolor=dvipsnames]{beamer}

\usepackage{setspace}
\usepackage{beamerthemesplit}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage{color}
\usepackage{colortbl}
%\usepackage{algorithm,algorithmic}
\usepackage{multimedia,xmpmulti}
\usepackage{multirow,multicol}
\usepackage[absolute,overlay]{textpos}
\usepackage{multimedia,xmpmulti}
\usepackage{tabularx}
\usepackage{algorithm,algpseudocode}

\usepackage{ourmacros}

%% Tikz - For making pretty pictures
\usepackage{tikz}
\usetikzlibrary{3d}
\usetikzlibrary{patterns}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\multiplycolor}{blue}
\newcommand{\zero}{}

\graphicspath{{fig/}}

\mode<presentation>

\definecolor{wfugold}{rgb}{0.6196078,0.494117647,0.21960784}
\usetheme{Warsaw}
\usecolortheme[named=wfugold]{structure}
%Madrid
%\usecolortheme{dolphin}
%\useinnertheme{rounded}
%\usefonttheme{serif}
\setbeamertemplate{navigation symbols}{} % gets rid of navigation bars
\setbeamertemplate{footline}
{
  \hbox{
  \begin{beamercolorbox}[wd=.33\paperwidth,ht=2.25ex,dp=1ex,left]{author in head/foot}%
    \usebeamerfont{author in head/foot}
    Ballard
  \end{beamercolorbox}%
  \begin{beamercolorbox}[wd=.34\paperwidth,ht=2.25ex,dp=1ex,center]{title in head/foot}%
    \usebeamerfont{title in head/foot}

  \end{beamercolorbox}%
  \begin{beamercolorbox}[wd=.33\paperwidth,ht=2.25ex,dp=1ex,right]{date in head/foot}%
    \usebeamerfont{date in head/foot}
    \insertframenumber{} \hspace*{2ex} 
  \end{beamercolorbox}}%
}

\title{Parallel Nonnegative CP Decomposition \\ of Dense Tensors}
\author[]{Grey Ballard, Koby Hayashi, Ramakrishnan Kannan}
\institute{}
\date{March 26, 2018}
%\titlegraphic{\includegraphics[scale=.15]{wfulogo}}

\begin{document}

\begin{frame}[plain]
\maketitle
\end{frame}
\addtocounter{framenumber}{-1}

\begin{frame}
\frametitle{Nonnegative CP via Block Coordinate Descent (ALS)}

\footnotesize
\begin{itemize}
	\item Basic algorithm is block coordinate descent, each block is factor matrix
	\item For each block, solve nonnegative linear least squares problem
\end{itemize}
\normalsize

\vfill

%\begin{algorithm}
%\caption{$(\CPl,\epsilon) = \text{NNCP}(\TA,k)$}
%\caption{$\CP = \text{NNCP}(\TA,k)$}
%\label{alg:nncp}
\begin{algorithmic}[1]
\Require $\TA$ is $I_1\times \cdots \times I_N$ tensor, $k$ is approximation rank
%\State \Comment{Compute NNCP approximation}
\While{not converged}
	\State \Comment{Perform outer iteration of BCD}
	\For{$n=1$ to $N$}
	\State \Comment{Compute new factor matrix in $n$th mode}
	\State \red{$\Mn{M}{n} = \text{MTTKRP}(\TA,\{\Mn{H}{i}\},n)$}
	\State $\Mn{S}{n} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$
	\State \blue{$\Mn{H}{n} = \text{NLS-Update}(\Mn{S}{n},\Mn{M}{n})$}
%	\State $\Mn[\hat]{H}{n} = \text{NLS-Update}(\Mn{S}{n},\Mn{M}{n})$
%	\State \Comment{Normalize columns}
%	\State $\V{\lambda} = \text{Col-Norms}(\Mn[\hat]{H}{n})$
%	\State $\Mn{H}{n} = \text{Col-Scale}(\Mn[\hat]{H}{n},\V{\lambda})$
%	\State \Comment{Organize data for later modes}
	\State $\Mn{G}{n} = \MnTra{H}{n}\Mn{H}{n}$
	\EndFor
%	\State \Comment{Compute relative error $\epsilon$ from mode-$N$ matrices}
%	\State $\beta = \langle \Mn{M}{N},\Mn[\hat]{H}{N} \rangle$
%	\State $\gamma = \V{\lambda}^\Tra (\Mn{S}{N} \Hada \Mn{G}{N}) \V{\lambda}$
%	\State $\epsilon = \sqrt{(\alpha-2\beta+\gamma)/\alpha}$ 
\EndWhile
%\Ensure $\|\TA - \dsquare{\V{\lambda}; \Mn{H}{1},\dots,\Mn{H}{N}}\| /\|\TA\| = \epsilon$
\Ensure $\TA \approx \CP$
\end{algorithmic}
%\end{algorithm}

\end{frame}

\begin{frame}

\end{frame}

\end{document}

### 2018-03-26 latex/sc18/plots.tex
added ksweep plot

@@
 	ymin=0,
-	symbolic x coords={D1,N1,,D16,N16,,D81,N81,,D256},
+	symbolic x coords={D1,N1,,D16,N16,,D81,N81,,D256,N256},
 	xtick=data,

### 2018-03-26 latex/sc18/experiments.tex
added ksweep plot
finished talk slides

@@
 \end{itemize}
+
+\begin{figure}
+\renewcommand{\datafile}{data/ksw_4D_HSI.dat}
+\renewcommand{\numiterations}{10}
+\begin{tikzpicture}
+\begin{axis}[	
+	ybar stacked,
+	bar width=8pt,
+	width=\columnwidth,
+	height =.75\columnwidth,
+	%width=9cm, height=3.85cm,
+	ylabel={Time (s)}, 
+	xlabel={Rank $k$},
+	y label style={yshift=-.5cm},
+	ymin=0,
+	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
+	symbolic x coords={D10,D20,D30,D40,D50},
+	%xtick=data,
+	xticklabels={,10,20,30,40,50},
+	legend style={at={(0.5,1.3)},anchor=north},
+	legend columns=-1,
+]
+	\setcolors
+	\addplot table[x=alg-K, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-K, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-K, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-K, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-K, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
+	\legend{PM,mTTV,NLS,Factor Comm,Gram Comm};
+\end{axis}
+%\labels
+\end{tikzpicture}
+\caption{Per-iteration time breakdown of our implementation (using dimension trees) over various ranks for a time-lapse HSI dataset with dimensions $1344\times 1024\times 33 \times 9$ on 64 processors arranged in a $8\times8\times1\times1$ grid.}
+\end{figure}
 

### 2018-03-26 latex/sc18/plots.tex
finished talk slides

--- 2018-03-25 latex/sc18/plots.tex
@@
 	red, pattern=crosshatch, pattern color=red \\
+	green, pattern=crosshatch, pattern color=green \\
 	blue, pattern=crosshatch, pattern color=blue \\
-	green, pattern=crosshatch, pattern color=green \\
 }};
@@
 	ylabel={Time (s)}, 
-	%y label style={yshift=-.5cm},
+	y label style={yshift=-.5cm},
 	ymin=0,
-	symbolic x coords={D1,N1,,D16,N16,,D81,N81,,D256},
+	symbolic x coords={D1,N1,D16,N16,D81,N81,D256,N256},
 	xtick=data,

### 2018-03-26 latex/slides/talk.tex
finished talk slides

@@
 \usepackage{algorithm,algpseudocode}
+\usepackage{subfig}
+
+\usepackage{pgfplots}
+\usepackage{pgfplotstable}
+\usepackage{etoolbox}
+\usetikzlibrary{patterns}
 
@@
 	\item For each block, solve nonnegative linear least squares problem
+	\begin{itemize}
+		\item \scriptsize assuming least squares loss function (Gaussian distribution)
+	\end{itemize}
 \end{itemize}
@@
 
+\only<1>{
 %\begin{algorithm}
@@
 %\end{algorithm}
+}
+\only<2>{
+\begin{itemize}
+	\item Lots of ways to solve nonnegative linear least squares 
+	\begin{enumerate}
+		\item multiplicative updates (one element at a time, Lee-Seung)
+		\item hierarchical alternating least squares (one column at a time, Cichocki et al.)
+		\item block principal pivoting (active set method, Park et al.)
+		\item Nesterov algorithm (used by Sidiropolous et al.)
+		\item alternating directions method of multipliers
+	\end{enumerate}
+	\item All are bottlenecked by MTTKRP and also need Gram 
+	\begin{itemize}	
+		\item MTTKRP bottleneck in dense case, maybe not sparse?
+	\end{itemize}
+\end{itemize}
+}
 
@@
 \begin{frame}
+\frametitle{Parallel Algorithm for NNCP via BCD/ALS}
+
+\footnotesize
+\textbf{Data distribution}
+\begin{itemize}
+	\item block distribution of tensor over $N$-D logical processor grid
+	\item row-wise distribution of each factor matrix 
+\end{itemize}
+
+\vfill
+
+\textbf{Communication steps}
+\begin{itemize}
+	\item All-Gather and Reduce-Scatter across processor slices for MTTKRP
+	\item All-Reduce across all processors for Gram
+	\item All-Reduce across all processors for normalization and error comp.
+\end{itemize}
+
+\vfill
+
+\textbf{Local Computation}
+\begin{itemize}
+	\item local MTTKRP (which we do using dimension trees)
+	\item local NLS solves/updates (which we do using block principal pivoting)
+	\item normalization and error computation (very cheap)
+\end{itemize}
+
+\end{frame}
+
+\input{../sc18/fig/Alg_header.tex}
+
+\begin{frame}
+\frametitle{Inner Iteration of Parallel Algorithm}
+
+\begin{center}
+\scalebox{1.5}{
+\only<1>{\input{../sc18/fig/AlgStart.tex}}
+\only<2>{\input{../sc18/fig/AlgMTTKRP.tex}}
+\only<3>{\input{../sc18/fig/AlgRS.tex}}
+\only<4>{\input{../sc18/fig/AlgNLS.tex}}
+\only<5>{\input{../sc18/fig/AlgAG.tex}}
+}
+
+\vfill
+
+\only<1>{Start $n$th iteration with redundant subset of rows \\ of each input matrix.}
+\only<2>{Compute local MTTKRP for contribution to output matrix $\Mn{M}{2}$.}
+\only<3>{Reduce-Scatter to compute and distribute rows of $\Mn{M}{2}$.}
+\only<4>{Compute local NLS update to obtain $\Mn{H}{2}_{\V{p}}$ from $\Mn{M}{2}_{\V{p}}$ (and $\Mn{S}{2}$).}
+\only<5>{All-Gather to collect rows of $\Mn{H}{2}$ needed \\ for later inner iterations.}
+
+\end{center}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Inner Iteration of Parallel Algorithm}
+
+\begin{figure}
+\centering
+  \subfloat[Start]{\input{../sc18/fig/AlgStart.tex}} \quad
+  \subfloat[Local MTTKRP]{\input{../sc18/fig/AlgMTTKRP.tex}} \quad
+  \subfloat[Reduce-Scatter]{\input{../sc18/fig/AlgRS.tex}} \quad
+  \subfloat[Local NLS]{\input{../sc18/fig/AlgNLS.tex}} \quad
+  \subfloat[All-Gather]{\input{../sc18/fig/AlgAG.tex}}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Dimension Trees}
+
+\begin{center}
+\footnotesize
+Dimension trees systematically avoid recomputation across MTTKRPs 
+\normalsize
+
+\vfill
+
+\input{../sc18/fig/dim_tree.tex}
+\end{center}
+
+\vfill
+
+\footnotesize
+Nearly all the time spent in partial MTTKRP (PM), same number of flops as MTTKRP but easier to implement/parallelize (one DGEMM call)
+\normalsize
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Error Computation}
+
+Given data tensor $\TA$ and CP model $\T{M} = \CP$, we want to compute the squared error
+$$\| \TA - \T{M}\|^2 = \|\TA\|^2 - 2\langle \TA, \T{M} \rangle + \|\T{M}\|^2$$
+which we do efficiently using the identity
+$$\langle \TA, \T{M} \rangle = \langle \Mn{M}{N}, \Mn{H}{N} \rangle$$ 
+where 
+$$\Mn{M}{N} = \Mz{A}{N} (\Mn{H}{N-1} \Khat \cdots \Khat \Mn{H}{1})$$
+
+\vfill
+
+*This technique is noted in a paper by Smith and Karypis and used in code by Phan, Tichavsky, and Cichocki
+
+\end{frame}
+
+\input{../sc18/plots}
+
+\begin{frame}
+\frametitle{Strong Scaling on 3D Tensor}
+
+\begin{figure}
+\begin{tikzpicture}[scale=.9]
+\renewcommand{\datafile}{../sc18/data/str_3D_syn.dat}
+\renewcommand{\numiterations}{42}
+\liavastrue
+\strongscalingplot
+\end{tikzpicture}
+\caption{Strong scaling of 3D synthetic tensor with dimension $1024\times 1024\times 1024$ on processor grids $2^k\times 2^k\times 2^k$ for $k\in\{0,1,2,3\}$.  The rank is fixed at 32.}
+\end{figure}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Strong Scaling on 5D Tensor}
+
+\begin{figure}
+\begin{tikzpicture}[scale=.9]
+\renewcommand{\datafile}{../sc18/data/str_5D_syn.dat}
+\renewcommand{\numiterations}{10}
+\liavasfalse
+\strongscalingplot
+\end{tikzpicture}
+\caption{Strong scaling of 5D synthetic tensor with dimension $64\times 64\times 64\times 64\times 64$ on processor grids $1\times1\times1\times1\times1$, $2\times1\times1\times1\times1$, $\dots$, $2\times2\times2\times2\times2$.
+The rank is fixed at 32.}
+\end{figure}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Weak Scaling Time Breakdown}
+
+\begin{figure}
+\begin{tikzpicture}
+\renewcommand{\datafile}{../sc18/data/wk_4D_syn.dat}
+\renewcommand{\numiterations}{10}
+\breakdownplot
+%\labels
+\end{tikzpicture}
+\caption{Weak scaling of 4D synthetic tensors with (D) and without (N) the use of dimension trees.
+The tensor and processor grid dimensions are $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$, and the rank is fixed at 32.}
+\end{figure}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Varying Rank Time Breakdown}
+
+\begin{figure}
+\renewcommand{\datafile}{../sc18/data/ksw_4D_HSI.dat}
+\renewcommand{\numiterations}{10}
+\begin{tikzpicture}[scale=.8]
+\begin{axis}[	
+	ybar stacked,
+	bar width=12pt,
+	width=\columnwidth,
+	height =.5\columnwidth,
+	%width=9cm, height=3.85cm,
+	ylabel={Time (s)}, 
+	xlabel={Rank $k$},
+	y label style={yshift=-.5cm},
+	ymin=0,
+	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
+	symbolic x coords={D10,D20,D30,D40,D50},
+	%xtick=data,
+	xticklabels={,10,,20,,30,,40,,50},
+	legend style={at={(0.5,1.3)},anchor=north},
+	legend columns=-1,
+]
+	\setcolors
+	\addplot table[x=alg-K, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-K, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-K, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-K, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-K, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
+	\legend{PM,mTTV,NLS,Factor Comm,Gram Comm};
+\end{axis}
+%\labels
+\end{tikzpicture}
+\caption{Per-iteration time breakdown of our implementation (using dimension trees) over various ranks for a time-lapse HSI dataset with dimensions $1344\times 1024\times 33 \times 9$ on 64 processors arranged in a $8\times8\times1\times1$ grid.}
+\end{figure}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{Conclusions (so far)}
+
+\footnotesize
+\begin{enumerate}
+	\item Hard not to be compute bound in dense case
+	\begin{itemize}
+	\scriptsize
+		\item computation is dominated by $O(R (\prod I_k) / P)$
+		\item communication is dominated by $O(R \sum I_k / P_k)$
+		\item computation scales with $P$, communication scales with $P^{1/N}$, but it doesn't matter until $P$ is huge
+	\end{itemize}
+	\vfill
+	\item Dimension trees make huge difference for higher-order tensors
+	\begin{itemize}
+	\scriptsize
+		\item dominant computation term predicts $N/2$ speedup
+		\item avoiding large Khatri-Rao products also makes a big difference
+		\item DGEMM calls have more favorable shapes
+	\end{itemize}
+	\vfill
+	\item Liavas et al.'s code has some shortcomings
+	\begin{itemize}
+	\scriptsize
+		\item limited to 3D tensors
+		\item permutes the tensor entries for each matricization 
+		\item does not use dimension trees or efficient error computation
+		\item this doesn't explain everything, but we don't have a time breakdown for further detail
+	\end{itemize}
+\end{enumerate}
 

### 2018-03-26 latex/sc18/prelims.tex
added algorithm figure, adapted from IPDPS18 paper
completed ramki prelims

@@
+% !TEX root = paper.tex
+
 \section{Preliminaries} 
@@
 \end{itemize}
+
+\begin{algorithm}
+%\caption{$(\CPl,\epsilon) = \text{NNCP}(\TA,k)$}
+\caption{$\CP = \text{NNCP}(\TA,k)$}
+\label{alg:nncp}
+\begin{algorithmic}[1]
+\Require $\TA$ is $I_1\times \cdots \times I_N$ tensor, $k$ is approximation rank
+\State \Comment{Initialize data}
+%\State $\alpha = \|\TA\|^2$
+\For{$n=2$ to $N$}
+	\State Initialize $\Mn{H}{n}$ 
+	\State $\Mn{G}{n} = \MnTra{H}{n}\Mn{H}{n}$
+\EndFor
+\State \Comment{Compute NNCP approximation}
+\While{not converged}
+	\State \Comment{Perform outer iteration of BCD}
+	\For{$n=1$ to $N$}
+	\State \Comment{Compute new factor matrix in $n$th mode}
+	\State $\Mn{M}{n} = \text{MTTKRP}(\TA,\{\Mn{H}{i}\},n)$
+	\State $\Mn{S}{n} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$
+	\State $\Mn{H}{n} = \text{NLS-Update}(\Mn{S}{n},\Mn{M}{n})$
+%	\State $\Mn[\hat]{H}{n} = \text{NLS-Update}(\Mn{S}{n},\Mn{M}{n})$
+%	\State \Comment{Normalize columns}
+%	\State $\V{\lambda} = \text{Col-Norms}(\Mn[\hat]{H}{n})$
+%	\State $\Mn{H}{n} = \text{Col-Scale}(\Mn[\hat]{H}{n},\V{\lambda})$
+%	\State \Comment{Organize data for later modes}
+	\State $\Mn{G}{n} = \MnTra{H}{n}\Mn{H}{n}$
+	\EndFor
+%	\State \Comment{Compute relative error $\epsilon$ from mode-$N$ matrices}
+%	\State $\beta = \langle \Mn{M}{N},\Mn[\hat]{H}{N} \rangle$
+%	\State $\gamma = \V{\lambda}^\Tra (\Mn{S}{N} \Hada \Mn{G}{N}) \V{\lambda}$
+%	\State $\epsilon = \sqrt{(\alpha-2\beta+\gamma)/\alpha}$ 
+\EndWhile
+%\Ensure $\|\TA - \dsquare{\V{\lambda}; \Mn{H}{1},\dots,\Mn{H}{N}}\| /\|\TA\| = \epsilon$
+\Ensure $\TA \approx \CP$
+\end{algorithmic}
+\end{algorithm}
+
+\subsection {Block Coordinate Descent(BCD) for NTF:}
+
+In this section, we will see relevant foundation for using  this
+framework.  The explanation of this section is mainly motivated out of
+Kim, He and Park \cite{KHP2014}. Consider a constrained non-linear optimization problem
+as follows:
+\begin{gather}
+\min f(x)\:\mbox{ subject to }\:x \in\mathcal{X},\label{eq:general_nonlinear}
+\end{gather}
+Here,  $\mathcal{X}$ is a closed convex subset of $\mathbb{R}^{n}$.
+An important assumption to be exploited in the BCD method is that
+the set $\mathcal{X}$ is represented by a Cartesian product:
+\begin{equation}
+\mathcal{X}=\mathcal{X}_{1}\times\cdots\times\mathcal{X}_{m},\label{eq:bcd-cartesian-product}
+\end{equation}
+where $\mathcal{X}_{j}$, $j=1,\cdots,m$, is a closed convex subset
+of $\mathbb{R}^{N_{j}}$, satisfying $n=\sum_{j=1}^{m}N_{j}$.
+Accordingly, the vector $\mathbf{x}$ is partitioned as
+$\mathbf{x}=(\mathbf{x}_{1},\cdots,\mathbf{x}_{m})$ so that
+$\mathbf{x}_{j}\in\mathcal{X}_{j}$ for $j=1,\cdots,m$. The BCD
+method solves for $\mathbf{x}_{j}$ by fixing all other subvectors of
+$\mathbf{x}$ in a cyclic manner. That is, if
+$\mathbf{x}^{(i)}=(\mathbf{x}_{1}^{(i)},\cdots,\mathbf{x}_{m}^{(i)})$
+is given as the current iterate at the $i^{th}$ step, the algorithm
+generates the next iterate
+$\mathbf{x}^{(i+1)}=(\mathbf{x}_{1}^{(i+1)},\cdots,\mathbf{x}_{m}^{(i+1)})$
+block by block, according to the solution of the following
+subproblem:
+\begin{equation}
+\mathbf{x}_{j}^{(k+1)}\leftarrow\underset{\mathbf{\xi}\in\mathcal{X}_{j}}{\text{argmin}}
+f(\mathbf{x}_{1}^{(k+1)},\cdots,\mathbf{x}_{j-1}^{(k+1)},\mathbf{\xi},\mathbf{x}_{j+1}^{(k)},\cdots,\mathbf{x}_{m}^{(k)}).\label{eq:bcd-method}
+\end{equation}
+Also known as a \textit{non-linear Gauss-Seidel}
+method~\cite{Bertsekas1999}, this algorithm updates one block each
+time, always using the most recently updated values of other blocks
+$\mathbf{x}_{\tilde{j}},\tilde{j}\ne j$. This is important since it
+ensures that after each update, the objective function value does
+not increase. For a sequence $\left\lbrace
+\mathbf{x}^{(i)}\right\rbrace $ where each $\mathbf{x}^{(i)}$ is
+generated by the BCD method, the following property holds.
+\begin{theorem}
+\label{thm:bcd}
+Suppose $f$ is continuously differentiable in $\mathcal{X}=\mathcal{X}_{1}\times\dots\times\mathcal{X}_{m}$,
+where $\mathcal{X}_{j}$, $j=1,\cdots,m$, are closed convex sets.
+Furthermore, suppose that for all $j$ and $i$, the minimum of
+\[
+\min_{\mathbf{\mathbf{\xi}}\in\mathcal{X}_{j}}f(\mathbf{x}_{1}^{(k+1)},\cdots,\mathbf{x}_{j-1}^{(k+1)},\mathbf{\xi},\mathbf{x}_{j+1}^{(k)},\cdots,\mathbf{x}_{m}^{(k)})
+\]
+is uniquely attained. Let $\left\lbrace
+\mathbf{x}^{(i)}\right\rbrace $ be the sequence generated by the
+block coordinate descent method as in Eq.~\eqref{eq:bcd-method}.
+Then, every limit point of $\left\lbrace
+\mathbf{x}^{(i)}\right\rbrace $ is a stationary point. The
+uniqueness of the minimum is not required for the case when  $m=2$
+\cite{GS2000}.
+\end{theorem}
+
+The proof of this theorem for an arbitrary number of blocks is shown
+in Bertsekas~\cite{Bertsekas1999}.
+For a non-convex optimization problem, most algorithms only guarantee
+the stationarity of a limit point \cite{Lin2007}.
+
+When applying the BCD method to a constrained non-linear programming
+problem, we have to ensure the following (a) it is critical to wisely choose a partition of
+$\mathcal{X}$, whose Cartesian product constitutes $\mathcal{X}$. 
+(b)  the dependency among the subproblems -- independent blocks can be 
+computed simultaneously  (c) when updating a block, ensure we are 
+using the most recent other blocks and (d) importantly, every block
+must be solved optimally. Every NTF variant will have different the partition
+scheme for blocks with natural partition such as scalar, vector \cite{CP2009} and matrix \cite{KP2011}. 
+Also, the update equation for determining the optimal block given
+other blocks such as projected gradient, block principal pivoting \cite{KP2011}  etc. 
+A detailed discussion of equivalences and comparison of different NMF and NTF algorithm
+is presented at \cite{KHP2014}.
+
+In the rest of the section, we will discuss the update equation for different 
+algorithms, given the local matrix $\Mn{S}{n}$ and $\Mn{M}{n}_{\V{p}}$ as described in the
+Algorithm \ref{alg:2D}. In a similar direction, The paper $MPIFAUN$ \cite{KBP2018}, have detailed some of 
+these algorithms on their paper towards NMF.
+
+\subsubsection{Hierarchical Alternating Least Squares (HALS)}
+\label{sec:HALS}
+
+In the case of HALS \cite{CP2009}, updates are performed on individual columns of $\HH$ with all other entries in the factor matrices fixed.
+The update rules \cite[Algorithm 2]{CP2009} can be written in closed form:
+\SplitN{\label{eqn:halsupdate}} {
+\Mn[\hat]{H}{n}_{\V{p}}(:,i) &\leftarrow \lt[ \Mn{H}{n}_{\V{p}}(:,i) + \Mn{M}{n}_{\V{p}}(:,i) - (\Mn{H}{n}_{\V{p}} \Mn{S}{n})(:,i)  \rt]_+\\
+% \ww^i &\leftarrow \lt[ \ww^i + (\AA\HH^T)^i - \WW (\HH \HH^T)^i \rt]_+
+},
+
+where, $\Mn{H}{n}_{\V{p}}(:,i) \in \Rn{I_n/P}$ is the $i^{th}$ column of the factor matrix $\Mn{H}{n}_{\V{p}}$.
+Note that the columns of $\Mn{H}{n}_{\V{p}}$ are updated in order, so that the most up-to-date values are always used, and these $nk$ updates can be done in an arbitrary order.
+In our case we are updating all the $k$ vectors of a factor $\Mn{H}{n}_{\V{p}}$ one by one.
+
+\subsection{Alternating Nonnegative Least Squares with Block Principal Pivoting}
+\label{sec:BPP}
+
+Kim and Park \cite{KP2011} first proposed the algorithm and the detail of the BPP algorithm for NTF is explained subsequently in detailed at \cite{KHP2014}.
+In this paper, we will leverage it to solve the factors $\Mn{H}{n}$.
+Block Principal Pivoting (BPP) is an active-set-like method for solving the NLS subproblems in Eq. \eqref{eqn:single NLS}.
+\SplitN{\label{eqn:single NLS}}{
+\min_{\xx\geq 0} \|\CC\XX-\BB\|_2.
+}. 
+Kannan, Ballard and Park \cite{KBP2018}, have explained solving the active-set based non-negative least squares 
+using normal equations with $\CC^T\CC$ and $\CC^T\BB$ for determining the non-negative factors of NMF. 
+In our case, $\CC^T\CC$ is $\Mn{S}{n}$ and $\CC^T\BB$ is $\Mn{M}{n}_{\V{p}}$. We use the same implementation
+of BPP from $MPIFAUN$.
+
+Our paper can be extended to other NTF algorithms such as AO-ADMM \cite{HSL2015} and Nestrov based updates
+\cite{LKLHS2017} by changing the $Local-NLS-Update$ function of the proposed Algorithm \ref{alg:2D}.

### 2018-03-27 latex/sc18/prelims.tex

@@
 must be solved optimally. Every NTF variant will have different the partition
-scheme for blocks with natural partition such as scalar, vector \cite{CP2009} and matrix \cite{KP2011}. 
+scheme for blocks with natural partition such as scalar \cite{SL2001}, vector \cite{CP2009} and matrix \cite{KP2011}. 
 Also, the update equation for determining the optimal block given
@@
 
-Our paper can be extended to other NTF algorithms such as AO-ADMM \cite{HSL2015} and Nestrov based updates
+Our paper can be extended to other NTF algorithms such as Alternating Optimization and Alternating Direction Method of Multipliers (AO-ADMM) \cite{HSL2015, SBK2017} and Nestrov based updates
 \cite{LKLHS2017} by changing the $Local-NLS-Update$ function of the proposed Algorithm \ref{alg:2D}.

### 2018-03-27 latex/sc18/prelims.tex
Some minor prelims work

--- 2018-03-26 latex/sc18/prelims.tex
@@
 
+	Tensors will be denoted using Euler script (e.g., $\T{T}$), matrices will be denoted (e.g., $\M{M}$), vectors will be denoted as lower case bold face type (e.g., $\V{v}$), and scalars will not be bolded (e.g., $s$).
+Matlab style notation will be sued to index into tensors, matrices, and vectors.
+For example, $\M{M}(:,c)$ gives the cth column of the matrix M.
+	The Hadamard product is the element wise matrix product and will be denoted using $*$. 
+	The Khatri-Rao product, abbreviated KRP, will be denoted with $\Khat$.
+Given matrices $\M{A}$ and $\M{B}$ that are $I_{A} \times R$ and $I_{B} \times R$, the KRP $\M{K} = \M{A} \Khat \M{B}$ is $I_{A}I_{B} \times R$.
+It can be thought of as a series row-wise Hadamard products s.t.
+\\$\M{K}(0:) = \M{A}(0,:) * \M{B}(0,:)\\ \M{K}(1:) = \M{A}(0,:) * \M{B}(1,:)  \\.\\.\\.
+\\ \M{K}(I_{A}I_{B} - 1,:) = \M{A}(I_{A}-1,:) * \M{B}(I_{B}-1,:)$
+
 \begin{itemize}
-	\item tensor notation, Hadamard, Khatri-Rao, MTTKRP
+	%\item tensor notation, Hadamard, Khatri-Rao, 
 	\item NNCP definition

### 2018-03-27 latex/sc18/survey.tex
added algorithm figure, adapted from IPDPS18 paper

@@
+% !TEX root = paper.tex
+
 \section{Related Work} 
@@
 \end{itemize}
+
+In recent time scaling tensor operations have been of interest to the data mining/machine learning (DM/ML) and the high performance computing (HPC) community.
+A detailed survey of  from Sidiropoulos et.al., \cite{SLFHPF2017} that includes basic tensor factorization models and their relationships and properties, broad coverage of algorithms ranging from alternating optimization to stochastic gradient and different applications ranging is a good starting point for Non-negative Tensor Factorization.
+Similarly, the tensor operations used in this paper is motivated out of Bader and Kolda \cite{KB2009, BK2012}.
+
+Most of the work is for sparse tensors because of the general availability of the 
+internet data. A collection of sparse tensor datasets can be found at FROSTT repository \cite{forsttdataset}. We  
+briefly capture some of the key results on sparse tensors towards tensor factorization.
+
+\subsection{Sparse NTF:}
+Smith and Karypis \cite{SBK2017} has recently presented parallelization strategies and two approaches for accelerating AO-ADMM.
+They also developed a method of exploiting dynamic sparsity in the factors to speed up tensor-matrix kernels and efficient use of cache resources that offered 8x speed up over the state of the art sparse tensor methods.
+Their entire software is available at \cite{splattsoftware}.
+Kaya and Ucar have contributed significantly in the area of sparse tensor decomposition.
+A comprehensive coverage of their entire work is detailed at \cite{Kaya2017}.
+They have investigated hyper-graph based tensor decomposition on distributed memory \cite{KU15},  effective shared memory parallelizations of these algorithms \cite{KU16} and CP decomposition for sparse tensor \cite{KU18}.
+Li et.al., \cite{LCPSV17}  have optimized MTTKRP for sparse tensors on shared memory systems using an adaptive tensor memoization algorithm that achieve 8× faster over SPLATT \cite{SRSK2015}.
+The optimization of the kernel Sparse Tensor-Times-dense matrix Multiply (SpTTM) on shared memory CPU and GPU by parallelizing, avoiding locks, and exploiting data locality have been proposed Li, Ma, Yan and Vuduc \cite{LMYV16}.
+
+Given these recent work on the sparse tensors, we survey the literature around dense tensor -- the interest of this paper.
+Unlike sparse tensor, there are very few activities around dense tensors that are important for analyzing tensors from images and scientific world such as 2D Ronchigram from pytography data.
+With the advent of the deep learning and the interest of the community in processing higher order tensors from hyper spectral image and medical imaging, we have used these real world in conducting experiments of this paper.
+
+Austin, Ballard, and Kolda \cite{ABK16} distributed-memory parallel implementation for the Tucker decomposition to compress 8TB of scientific data.
+Tucker decomposition is unconstrained and a completely different decomposition over constrained Non-negative Tensor Factorization explained in this paper.
+Recently,  Ballard, Knight, Rouse \cite{BKR2018} have proposed Communication Lower Bounds for Matricized Tensor Times Khatri-Rao Product under particular distribution of dense tensors.
+The proposed algorithm \ref{alg:2D} is motivated out of this work.
+
+With this detailed survey, we would like to highlight the contribution of this paper.
+
+\subsection{Contributions:}

### 2018-03-27 latex/sc18/algorithm.tex
just committing in order to merge, minor updates
filling in algorithm section, still need to do analysis

@@
 The extra computation involved in computing the relative error is negligible.
-These identities have been used in previous work \cite{KB09,TensorBox,SK16}.
+These identities have been used in previous work \cite{KB09,TensorBox,SK16,LKLHS2017}.
 
 \subsection{Parallel Algorithm}
+
+\subsubsection{Algorithm Overview}
+
+The basic sequential algorithm is given in \Cref{alg:nncp}, and the parallel version is given in \Cref{alg:Par-NNCP-short}.
+We will refer to both the inner iteration, in which one factor matrix is updated (\cref{line:for} to \cref{line:endfor}), and the outer iteration, in which all factor matrices are updated (\cref{line:while} to \cref{line:endwhile}).
+In the parallel algorithm, the processors are organized into a logical multidimensional grid (tensor) with as many modes as the data tensor.
+The communication patterns used in the algorithm are all MPI collectives, including All-Reduce, Reduce-Scatter, and All-Gather.
+The processor communicators (across which the collectives are performed) include the set of all processors and the sets of processors within the same processor slice.
+Processors within a mode-$n$ slice all have the same $n$th coordinate.
+
+The method of enforcing the nonnegativity constraints of the linear least squares solve (or update) generally affects only local computation because each row of a factor matrix can be updated independently.
+In our algorithm, each processor solves the linear problem or computes the update for its subset of rows (see \cref{line:locNLS}). 
+The most expensive (and most complicated) part of the parallel algorithm is the computation of the MTTKRP, which corresponds to \cref{line:locMTTKRP,line:reduce-scatter,line:all-gather}.
+
+The details that are omitted from this presentation of the algorithm include the normalization of each factor matrix after it is computed and the computation of the residual error at the end of an outer iteration.
+The computations do involve both local computation and communication, but their costs are negligible.
+A more detailed pseudocode is given in \Cref{alg:Par-NNCP-long}.
+
+\subsubsection{Data Distribution}
+
+Given a logical processor grid of processors $P_1\times \cdots \times P_N$, we distribute the tensor $\TA$ in a block or Cartesian partition.
+Each processor owns a local tensor of dimensions $(I_1/P_1)\times \cdots \times (I_N/P_N)$, and only one copy of the tensor is stored.
+Locally, the tensor is stored linearly, with entries ordered in a natural mode-descending way that generalizes column-major layout of matrices.
+Given a processor $\V{p}=(\VE{p}{1},\dots,\VE{p}{N})$, we denote its local tensor by $\TA_{\V{p}}$.
+
+Each factor matrix is distributed across processors in a block row partition, so that each processor owns a subset of the rows.
+We use the notation $\Mn{H}{n}_{\V{p}}$, which has dimensions $I_n/P\times R$ to denote the local part of the $n$th factor matrix stored on processor $\V{p}$.
+However, we also make use a redundant distribution of the factor matrices across processors, because all processors in a mode-$n$ processor slice need access to the same entries of $\Mn{H}{n}$ to perform their computations.
+The notation $\Mn{H}{n}_{\VE{p}{n}}$ denotes the $I_n/P_n\times R$ submatrix of $\Mn{H}{n}$ that is redundantly stored on all processors whose $n$th coordinate is $\VE{p}{n}$ (there are $P/P_n$ such processors).
+
+Other matrices involved in the algorithm include $\Mn{M}{n}_{\V{p}}$, which is the result of the MTTKRP computation and has the same distribution scheme as $\Mn{H}{n}_{\V{p}}$, and $\Mn{G}{n}$, which is the $R\times R$ Gram matrix of the factor matrix $\Mn{H}{n}$ and is stored redundantly on all processors.
+
+\subsubsection{Inner Iteration}
+
+The inner iteration is displayed graphically in \Cref{fig:inner} for a 3-way example and an update of the $2$nd factor matrix.
+The main idea is that at the start of the $n$th inner iteration (\cref{fig:inner_a}), all of the data is in place for each processor to perform a local MTTKRP computation.
+This means that all processors in a slice redundantly own the same rows of the corresponding factor matrix (for all modes except $n$).
+After the local MTTKRP is computed (\cref{fig:inner_b}), each processor has computed a contribution to a subset of the rows of the global MTTKRP $\Mn{M}{n}$, but its contribution must be summed up with the contributions of all other processors in its mode-$n$ slice.
+This summation is performed with a Reduce-Scatter collective across the mode-$n$ processor slice that achieves a row-wise partition of the result (in \cref{fig:inner_c}, the light gray shading corresponds to the rows of $\Mn{M}{2}$ to which processor $(1,3,1)$ contributes and the dark gray shading corresponds to the rows it receives as output).
+The output distribution of the Reduce-Scatter is designed so that afterwards, the update of the factor matrix in that mode can be performed row-wise in parallel.
+Along with $\Mn{S}{n}$, which can be computed locally, each processor updates its own rows of the factor matrix given its rows of the MTTKRP result (\cref{fig:inner_d}).
+The remainder of the inner iteration is preparing and distributing the new factor matrix data for future inner iterations, which includes an All-Gather of the newly computed factor matrix $\Mn{H}{n}$ across mode-$n$ processor slices (\cref{fig:inner_e}) and recomputing $\Mn{G}{n}={\Mn{H}{n}}^\Tra\Mn{H}{n}$.
+
+\subsubsection{Analysis}
+
+\paragraph{Computation}
+
+\paragraph{Communication}
+
+\paragraph{Memory}
 
@@
 \centering
-  \subfloat[Start $n$th iteration with redundant subset of rows of each input matrix.]{\input{fig/AlgStart.tex}} \quad
-  \subfloat[Compute local MTTKRP for contribution to output matrix $\Mn{M}{2}$.]{\input{fig/AlgMTTKRP.tex}} \quad
-  \subfloat[Reduce-Scatter to compute and distribute rows of $\Mn{M}{2}$.]{\input{fig/AlgRS.tex}} \quad
-  \subfloat[Compute local NLS update to obtain $\Mn{H}{2}_{\V{p}}$ from $\Mn{M}{2}_{\V{p}}$ (and $\Mn{S}{2}$).]{\input{fig/AlgNLS.tex}} \quad
-  \subfloat[All-Gather to collect rows of $\Mn{H}{2}$ needed for later inner iterations.]{\input{fig/AlgAG.tex}}
+  \subfloat[Start $n$th iteration with redundant subset of rows of each input matrix. \label{fig:inner_a}]{\input{fig/AlgStart.tex}}  \quad
+  \subfloat[Compute local MTTKRP for contribution to output matrix $\Mn{M}{2}$. \label{fig:inner_b}]{\input{fig/AlgMTTKRP.tex}}  \quad
+  \subfloat[Reduce-Scatter to compute and distribute rows of $\Mn{M}{2}$. \label{fig:inner_c}]{\input{fig/AlgRS.tex}} \quad
+  \subfloat[Compute local NLS update to obtain $\Mn{H}{2}_{\V{p}}$ from $\Mn{M}{2}_{\V{p}}$ (and $\Mn{S}{2}$). \label{fig:inner_d}]{\input{fig/AlgNLS.tex}} \quad
+  \subfloat[All-Gather to collect rows of $\Mn{H}{2}$ needed for later inner iterations. \label{fig:inner_e}]{\input{fig/AlgAG.tex}}
   \caption{Illustration of 2nd inner iteration of Par-NNCP algorithm for 3-way tensor on a $3\times3\times3$ processor grid, showing data distribution, communication, and computation across steps.
@@
 The column normalization and computation of $\Mn{G}{2}$, which involve communication across all processors, is not shown here.}
-  \label{fig:par_staten} 
+  \label{fig:inner} 
 \end{figure*}
@@
 \begin{algorithm}
-\caption{$(\CPl,\epsilon) = \text{Par-NNCP}(\TA,k)$}
-\label{alg:2D}
+\caption{$\CP = \text{Par-NNCP}(\TA,k)$}
+\label{alg:Par-NNCP-short}
 \begin{algorithmic}[1]
 \Require $\TA$ is an $I_1\times \cdots \times I_N$ tensor distributed across a $P_1\times \cdots \times P_N$ grid of $P$ processors, so that $\TA_{\V{p}}$ is $(I_1/P_1)\times \cdots \times (I_N/P_N)$ and is owned by processor $\V{p}=(p_1,\dots,p_N)$, $k$ is rank of approximation
-\State \Comment{Initialize data}
-\State $a = \text{Norm-Squared}(\TA_{\V{p}})$
-\State $\alpha = \text{All-Reduce}(a,\textsc{All-Procs})$
 \For{$n=2$ to $N$}
@@
 \While{not converged}
+	\label{line:while}
 	\State \Comment{Perform outer iteration of BCD}
 	\For{$n=1$ to $N$}
+		\label{line:for}
 		\State \Comment{Compute new factor matrix in $n$th mode}
@@
 			\label{line:hadamard}
-	\State $\Mn[\hat]{H}{n}_{\V{p}} = \text{Local-NLS-Update}(\Mn{S}{n},\Mn{M}{n}_{\V{p}})$
+		\State $\Mn{H}{n}_{\V{p}} = \text{Local-NLS-Update}(\Mn{S}{n},\Mn{M}{n}_{\V{p}})$
 			\label{line:locNLS}
-	\State \Comment{Normalize columns}
-	\State $\V[\overline]{\lambda} = \text{Local-Col-Norms}(\Mn[\hat]{H}{n}_{\V{p}})$
-	\State $\V{\lambda} = \text{All-Reduce}(\V[\overline]{\lambda},\textsc{All-Procs})$
-	\State $\Mn{H}{n}_{\V{p}} = \text{Local-Col-Scale}(\Mn[\hat]{H}{n}_{\V{p}},\V{\lambda})$
 		\State \Comment{Organize data for later modes}
@@
 	\EndFor 
-	\State \Comment{Compute relative error $\epsilon$ from mode-$N$ matrices}
-	\State $\overline{\beta} = \text{Inner-Product}(\Mn{M}{N}_{\V{p}},\Mn[\hat]{H}{N}_{\V{p}})$
-	\State $\beta = \text{All-Reduce}(\overline{\beta},\textsc{All-Procs})$
-	\State $\gamma = \V{\lambda}^\Tra (\Mn{S}{N} \Hada \Mn{G}{N}) \V{\lambda}$
-	\State $\epsilon = \sqrt{(\alpha-2\beta+\gamma)/\alpha}$ 
+		\label{line:endfor}
 \EndWhile
-\Ensure $\|\TA - \dsquare{\V{\lambda}; \Mn{H}{1},\dots,\Mn{H}{N}}\| /\|\TA\| = \epsilon$
+	\label{line:endwhile}
+\Ensure $\TA \approx \CP$
 \Ensure Local matrices: $\Mn{H}{n}_{\V{p}}$ is $(I_n/P)\times k$ and owned by processor $\V{p}=(p_1,\dots,p_N)$, for $1\leq n \leq N$, $\V{\lambda}$ stored redundantly on every processor

### 2018-03-27 latex/sc18/appendix.tex

--- 2018-03-08 latex/sc18/algorithm.tex
@@
 
-\section{Algorithm} 
-\label{sec:algorithm}
-
-\subsection{Dimension Trees}
-
-\subsection{Relative Error Computation}
-
-Given a model $\T{M}=\CP$, we compute the relative error $\|\TA - \T{M}\|/\|\TA\|$ efficiently by using the identity $\|\TA-\T{M}\|^2 = \|\TA\|^2 - 2\langle \TA, \T{M} \rangle + \|\T{M}\|^2.$
-The quantity $\|\TA\|$ is fixed, and the other two terms can be computed cheaply given the temporary matrices computed during the course of the BCD algorithm.
-The second term can be computed using the identity $\langle \TA, \T{M} \rangle = \langle \Mn{M}{N}, \Mn{H}{N} \rangle$, where $\Mn{M}{N} = \Mz{A}{N} (\Mn{H}{N-1} \Khat \cdots \Khat \Mn{H}{1})$ is the MTTKRP result in the $N$th mode.
-The third term can be computed using the identity $\|\T{M}\|^2 = \V{1}^\Tra(\Mn{S}{N} \Hada \MnTra{H}{N} \Mn{H}{N})\V{1}$ where $\Mn{S}{N}=\MnTra{H}{1} \Mn{H}{1} \Hada \cdots \Hada \MnTra{H}{N-1} \Mn{H}{N-1}$.
-Both matrices $\Mn{M}{N}$ and $\Mn{S}{N}$ are computed during the course of the BCD algorithm for updating the factor matrix $\Mn{H}{N}$.
-The extra computation involved in computing the relative error is negligible.
-These identities have been used in previous work \cite{KB09,TensorBox,SK16}.
-
-\subsection{Parallel Algorithm}
-
-\input{fig/Alg_header.tex}
-\begin{figure*}[t]
-\centering
-  \subfloat[Start $n$th iteration with redundant subset of rows of each input matrix.]{\input{fig/AlgStart.tex}} \quad
-  \subfloat[Compute local MTTKRP for contribution to output matrix $\Mn{M}{2}$.]{\input{fig/AlgMTTKRP.tex}} \quad
-  \subfloat[Reduce-Scatter to compute and distribute rows of $\Mn{M}{2}$.]{\input{fig/AlgRS.tex}} \quad
-  \subfloat[Compute local NLS update to obtain $\Mn{H}{2}_{\V{p}}$ from $\Mn{M}{2}_{\V{p}}$ (and $\Mn{S}{2}$).]{\input{fig/AlgNLS.tex}} \quad
-  \subfloat[All-Gather to collect rows of $\Mn{H}{2}$ needed for later inner iterations.]{\input{fig/AlgAG.tex}}
-  \caption{Illustration of 2nd inner iteration of Par-NNCP algorithm for 3-way tensor on a $3\times3\times3$ processor grid, showing data distribution, communication, and computation across steps.
-Highlighted areas correspond to processor $(1,3,1)$ and its processor slice with which it communicates.
-The column normalization and computation of $\Mn{G}{2}$, which involve communication across all processors, is not shown here.}
-  \label{fig:par_staten} 
-\end{figure*}
+\appendix
+\label{sec:appendix}
 
@@
 \caption{$(\CPl,\epsilon) = \text{Par-NNCP}(\TA,k)$}
-\label{alg:2D}
+\label{alg:Par-NNCP-long}
 \begin{algorithmic}[1]
@@
 \State $\alpha = \text{All-Reduce}(a,\textsc{All-Procs})$
+\State $\epsilon = $ \texttt{Inf}
 \For{$n=2$ to $N$}
@@
 \State \Comment{Compute NNCP approximation}
-\While{not converged}
+\While{$\epsilon > $ \texttt{tol}}
 	\State \Comment{Perform outer iteration of BCD}
@@
 	\State $\M[\overline]{M} = \text{Local-MTTKRP}(\TA_{p_1\cdots p_N},\{\Mn{H}{i}_{p_i}\},n)$
-		\label{line:locMTTKRP}
 	\State $\Mn{M}{n}_{\V{p}} = \text{Reduce-Scatter}(\M[\overline]{M},\textsc{Proc-Slice}(n,\VE{p}{n}))$ 
-		\label{line:reduce-scatter}
 	\State $\Mn{S}{n} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$
-		\label{line:hadamard}
 	\State $\Mn[\hat]{H}{n}_{\V{p}} = \text{Local-NLS-Update}(\Mn{S}{n},\Mn{M}{n}_{\V{p}})$
-		\label{line:locNLS}
 	\State \Comment{Normalize columns}
@@
 	\State $\M[\overline]{G} = \text{Local-SYRK}(\Mn{H}{n}_{\V{p}})$
-		\label{line:locSYRK}
 	\State $\Mn{G}{n} = \text{All-Reduce}(\M[\overline]{G},\textsc{All-Procs})$
-		\label{line:all-reduce}
 	\State $\Mn{H}{n}_{p_n} = \text{All-Gather}(\Mn{H}{n}_{\V{p}},\textsc{Proc-Slice}(n,\VE{p}{n}))$
-		\label{line:all-gather}
 	\EndFor

### 2018-03-27 latex/sc18/algorithm.tex
finished draft of algorithm section

@@
 
-\paragraph{Computation}
+We will analyze the cost of a single outer iteration.
+While the number of outer iterations is sensitive to the NLS method used, the outer iteration time is generally the same across NLS methods.
 
-\paragraph{Communication}
+\paragraph{\emph{Computation}}
+The local computation occurs at \cref{line:locMTTKRP,line:hadamard,line:locNLS,line:locSYRK}.
+The cost of \cref{line:hadamard} is $O(NR^2)$, the cost of \cref{line:locNLS} is $O(R^3I_n/P)$, which is a loose upper bound for BPP and other methods \cite{KBP16}, and the cost of \cref{line:locSYRK} is $O(R(I_n/P)^2)$.
+The sum of these three costs across all inner iterations is $O(R^2N^2+(R^3/P+R/P^2)\sum I_n)$, which is dominated by the cost of the MTTKRP.
+When using dimension trees to perform the MTTKRP (\cref{line:locMTTKRP}), we compute the cost amortized over all inner iterations.
+In this case, the cost is dominated by the two partial MTTKRP computations (from the root of the tree), which together are $O((R/P) \prod I_n)=O(IR/P)$ and dominate the costs of the multi-TTVs.
+We note that this cost involves the product of all the tensor dimensions, which is why it dominates, and we note that it scales linearly with $P$.
 
-\paragraph{Memory}
+\paragraph{\emph{Communication}}
+The communication within the inner iteration occurs at \cref{line:reduce-scatter,line:all-reduce,line:all-gather}.
+\Cref{line:all-reduce} involves $O(R^2)$ data and a collective across all processors.
+\Cref{line:reduce-scatter,line:all-gather} involve $O(I_nR/P_n)$ data across a subset of $P/P_n$ processors.
+Thus, the All-Reduce dominates the latency cost and the Reduce-Scatter/All-Gather dominate the bandwidth cost, for a total outer iteration communication cost of $O(R\sum I_n/P_n)$ words and $O(N\log P)$ messages.
+If the optimal processor grid can be chosen to minimize communication (assuming $P$ is sufficiently factorable), then the bandwidth cost can achieve a value of $O(RI^{1/N}/P^{1/N})$ by making the local tensors as cubical as possible.
+We note that this cost scales with $P^{1/N}$, which is far from linear scaling.
+However, it is proportional to the geometric mean of the tensor dimensions (on the order of one tensor dimension), which is much less than the computation cost dependence on the product of all dimensions.
+
+\paragraph{\emph{Memory}}
+The algorithm requires extra local memory to run.
+Aside from the memory required to store the local tensor of $O(I/P)$ words and factor matrices of cumulative size $O((R/P)\sum I_n)$, each processor must be able to store a redundant subset of the rows of the factor matrices it needs to perform MTTKRP computations.
+This corresponds to storing $P/P_n$ redundant copies of every factor matrix, which results in a local memory requirement of $O(R \sum I_n/P_n)$ for a general processor grid.
+The processor grid that minimizes communication also minimizes local memory, and the extra memory requirement can be as low as $O(RI^{1/N}/P^{1/N})$.
+
+The dimension tree algorithm also requires extra temporary memory space, but the space required tends to be much smaller than what is required to store the local tensor.
+If the tensor dimensions can be partitioned into two parts with approximately equal geometric means, the extra memory requirement for running a dimension tree is as small as $O(R\sqrt{I}/P)$, which is typically dominated by $O(I/P)$.
 
@@
 		\State \Comment{Organize data for later modes}
-		\State $\M[\overline]{G} = \text{Local-SYRK}(\Mn{H}{n}_{\V{p}})$
+		\State $\M[\overline]{G} = {\Mn{H}{n}_{\V{p}}}^\Tra\Mn{H}{n}_{\V{p}}$
 			\label{line:locSYRK}

### 2018-03-28 latex/sc18/prelims.tex
Prelims
MPI collectives to prelims

@@
 
-	Tensors will be denoted using Euler script (e.g., $\T{T}$), matrices will be denoted (e.g., $\M{M}$), vectors will be denoted as lower case bold face type (e.g., $\V{v}$), and scalars will not be bolded (e.g., $s$).
+Tensors will be denoted using Euler script (e.g., $\T{T}$), 
+matrices will be denoted (e.g., $\M{M}$), vectors will be denoted as lower case bold face type 
+(e.g., $\V{v}$), and scalars will not be bolded (e.g., $s$). 
 Matlab style notation will be sued to index into tensors, matrices, and vectors. 
@@
 It can be thought of as a series row-wise Hadamard products s.t.
-\\$\M{K}(0:) = \M{A}(0,:) * \M{B}(0,:)\\ \M{K}(1:) = \M{A}(0,:) * \M{B}(1,:)  \\.\\.\\.
-\\ \M{K}(I_{A}I_{B} - 1,:) = \M{A}(I_{A}-1,:) * \M{B}(I_{B}-1,:)$
+ \\$\M{K}(0:) = \M{A}(0,:) * \M{B}(0,:)\\ \M{K}(1:) = \M{A}(0,:) * \M{B}(1,:)  \\.\\.\\. \\ \M{K}(I_{A}I_{B} - 1,:) = \M{A}(I_{A}-1,:) * \M{B}(I_{B}-1,:)$ \\
+ The Matricized Tensor Times Khatri Rao Product or MTTKRP will be central to this work and takes the form $\M{M} = \M{X}_{(n)}( \Mn{H}{N-1} \Khat \cdots \Khat \Mn{H}{n-1} \Khat \Mn{H}{n+1} \Khat  \cdots \Khat \Mn{H}{0})$.
+$\M{X}_{(n)}$ is a matrix formed by taking the nth mode fibers of a tensor $\T{X}$ with dimensions $I_{0} \times I_{1} \times ...
+\times I_{N-1}$ and making them the columns of a matrix.
+
+%MPI collectives for Prelims
+To implement and analyze our algorithms we use the MPI collective communication model of distributed-memory parallel computation.
+Setting $W$ to be the number of words in a message between two processors, $\alpha$ to be the latency, and $\beta$ to be the per word cost, the total cost of a message is $\alpha + \beta W$.
+Reduce operations also include a flop cost but we will ignore it both for simplicity and the fact that cost is often dominated by communication.
+For a more in-depth discussion of the MPI model and its efficient collectives see [NEED CITATION],[NEED CITATION].
+The collectives utilized in this work and their costs are the
+\begin{itemize}
+\item All Reduce $2\alpha log P + 2\beta W (P-1)/P$
+\item All Gather $\alpha log P + \beta W (P-1)/P$
+\item Reduce Scatter $\alpha log P + \beta W (P-1)/P$
+\end{itemize}
 

### 2018-03-28 latex/sc18/prelims.tex
made pass, tightened up preliminaries section

--- 2018-03-27 latex/sc18/prelims.tex
@@
 
-	Tensors will be denoted using Euler script (e.g., $\T{T}$), matrices will be denoted (e.g., $\M{M}$), vectors will be denoted as lower case bold face type (e.g., $\V{v}$), and scalars will not be bolded (e.g., $s$).
-Matlab style notation will be sued to index into tensors, matrices, and vectors.
-For example, $\M{M}(:,c)$ gives the cth column of the matrix M.
-	The Hadamard product is the element wise matrix product and will be denoted using $*$. 
+\subsection{Notation}
+
+Tensors will be denoted using Euler script (e.g., $\T{T}$), 
+matrices will be denoted with uppercase boldface (e.g., $\M{M}$), vectors will be denoted with lowercase boldface  
+(e.g., $\V{v}$), and scalars will not be boldface (e.g., $s$). 
+We use Matlab style notation to index into tensors, matrices, and vectors, and we use 1-indexing. 
+For example, $\M{M}(:,c)$ gives the $c$th column of the matrix M.
+
+We use $\circ$ to denote the outer product of two or more vectors.
+The Hadamard product is the element-wise matrix product and will be denoted using $\Hada$. 
 The Khatri-Rao product, abbreviated KRP, will be denoted with $\Khat$. 
 Given matrices $\M{A}$ and $\M{B}$ that are $I_{A} \times R$ and $I_{B} \times R$, the KRP $\M{K} = \M{A} \Khat \M{B}$ is $I_{A}I_{B} \times R$. 
-It can be thought of as a series row-wise Hadamard products s.t.
-\\$\M{K}(0:) = \M{A}(0,:) * \M{B}(0,:)\\ \M{K}(1:) = \M{A}(0,:) * \M{B}(1,:)  \\.\\.\\.
-\\ \M{K}(I_{A}I_{B} - 1,:) = \M{A}(I_{A}-1,:) * \M{B}(I_{B}-1,:)$
+It can be thought of as a row-wise Hadamard product, where $\M{K}(i+I_{A}(j{-}1),:) = \M{A}(i,:) \Hada \M{B}(j,:)$, or a column-wise Kronecker product, where $\M{K}(,c) =  \M{A}(:,c) \Kron \M{B}(:,c)$.
 
-\begin{itemize}
-	%\item tensor notation, Hadamard, Khatri-Rao, 
-	\item NNCP definition
-	\item BCD/ALS
-	\begin{itemize}
-		\item MU, HALS, BPP, Nesterov, ADMM
-	\end{itemize}
-	\item MPI collectives
-\end{itemize}
+The CP decomposition of a tensor (also referred to as the CANDECOMP/PARAFAC or canonical polyadic decomposition) is a low-rank approximation of a tensor, where the approximation is a sum of rank-one tensors and each rank-one tensor is the outer product of vectors.
+We use the notation
+$$\TA \approx \CP = \sum_{r=1}^R \Mn{H}{1}(:,r) \circ \cdots \circ \Mn{H}{N}(:,r)$$
+to represent a rank-$R$ CP model, where $\Mn{H}{n}$ is called a factor matrix and collects the mode-$n$ vectors of the rank-one tensors as columns.
+The columns of the factor matrices are often normalized, with weights collected into an auxiliary vector $\V{\lambda}$ of length $R$; in this case we use the notation $\CPl$.
+
+A Nonnegative CP decomposition (NNCP) constrains the factor matrices to have nonnegative values.
+In this work, we are interested in NNCP models that are good approximations to $\TA$ in the least squares sense.
+That is, we seek to minimize $\|\TA-\CPl\|$, where the tensor norm is a generalization of the matrix Frobenius or vector 2-norm, the square root of the sum of the squares of the entries.
+
+The $n$th mode matricized tensor denoted by $\M{A}_{(n)}$ is a $I_n\times I/I_n$ matrix formed by organizing the $n$th mode fibers of a tensor $\T{X}$ with dimensions $I_{1} \times ...
+\times I_{N}$ into the columns of a matrix.
+The Matricized Tensor Times Khatri Rao Product or MTTKRP will be central to this work and takes the form $\Mn{M}{n} = \Mz{A}{n}( \Mn{H}{N} \Khat \cdots \Khat \Mn{H}{n+1} \Khat \Mn{H}{n-1} \Khat  \cdots \Khat \Mn{H}{1})$.
+
+\subsection {Block Coordinate Descent for NNCP}
+
+While there are multiple optimization methods to compute NNCP, we will focus on a class of methods that use Block Coordinate Descent (BCD), which is also known as the nonlinear Gauss-Seidel method \cite{Bertsekas1999}.
+In BCD, the variables are partitioned into blocks, and each variable block is cyclically updated with all other blocks fixed.
+For details on the convergence properties and comparisons of BCD methods for nonnegative matrix and tensor decomposition problems, see \cite{KHP2014}.
+We consider BCD methods for NNCP that choose the factor matrices as the blocks, which is also often referred to as Alternating Least Squares.
+In this case, every subproblem is a linear nonnegative least squares problem.
+
+\Cref{alg:nncp} shows the pseudocode for BCD applied to NNCP.
+\Cref{line:MTTKRP,line:GH,line:Gn} compute matrices involved in the gradients of the subproblem objective functions, and \cref{line:NLS} uses those matrices to update the current factor matrix.
+
+The NLS-Update in \cref{line:NLS} can be implemented in different ways.
+In a faithful BCD algorithm, the subproblems are solved exactly; in this case, the subproblem is a nonnegative linear least squares problem, which is convex.
+We use the Block Principal Pivoting (BPP) method \cite{KP2011,KHP2014}, which is an active-set-like method, to solve the subproblem exactly.
+BPP is an active-set-like method for solving the NLS subproblems in Eq. \eqref{eqn:single NLS}.
+
+However, as discussed in \cite{KBP2018} for the matrix case, there are other reasonable alternatives to updating the factor matrix without solving the subproblem exactly.
+For example, we can more efficiently update individual columns of the factor matrix as is done in the Hierarchical Alternating Least Squares (HALS) method \cite{CP2009}.
+In this case, the update rule is 
+$$\Mn{H}{n}(:,r) \leftarrow \lt[ \Mn{H}{n}(:,r) + \Mn{M}{n}(:,r) - (\Mn{H}{n} \Mn{S}{n})(:,r)  \rt]_+$$
+which involves the same matrices $\Mn{M}{n}$ and $\Mn{S}{n}$ as BPP.
+Other possible BCD methods include Alternating Optimization and Alternating Direction Method of Multipliers (AO-ADMM) \cite{HSL2015, SBK2017} and Nestrov-based algorithms \cite{LKLHS2017}.
+The parallel algorithm presented in this paper is generally agnostic to the approach used to solve the nonnegative least squares subproblems, as all these methods are bottlenecked by the subroutine they have in common, the MTTKRP.
+
+%In this section, we will see relevant foundation for using  this
+%framework.  The explanation of this section is mainly motivated out of
+%Kim, He and Park \cite{KHP2014}. Consider a constrained non-linear optimization problem
+%as follows:
+%\begin{gather}
+%\min f(x)\:\mbox{ subject to }\:x \in\mathcal{X},\label{eq:general_nonlinear}
+%\end{gather}
+%Here,  $\mathcal{X}$ is a closed convex subset of $\mathbb{R}^{n}$.
+%An important assumption to be exploited in the BCD method is that
+%the set $\mathcal{X}$ is represented by a Cartesian product:
+%\begin{equation}
+%\mathcal{X}=\mathcal{X}_{1}\times\cdots\times\mathcal{X}_{m},\label{eq:bcd-cartesian-product}
+%\end{equation}
+%where $\mathcal{X}_{j}$, $j=1,\cdots,m$, is a closed convex subset
+%of $\mathbb{R}^{N_{j}}$, satisfying $n=\sum_{j=1}^{m}N_{j}$.
+%Accordingly, the vector $\mathbf{x}$ is partitioned as
+%$\mathbf{x}=(\mathbf{x}_{1},\cdots,\mathbf{x}_{m})$ so that
+%$\mathbf{x}_{j}\in\mathcal{X}_{j}$ for $j=1,\cdots,m$. The BCD
+%method solves for $\mathbf{x}_{j}$ by fixing all other subvectors of
+%$\mathbf{x}$ in a cyclic manner. That is, if
+%$\mathbf{x}^{(i)}=(\mathbf{x}_{1}^{(i)},\cdots,\mathbf{x}_{m}^{(i)})$
+%is given as the current iterate at the $i^{th}$ step, the algorithm
+%generates the next iterate
+%$\mathbf{x}^{(i+1)}=(\mathbf{x}_{1}^{(i+1)},\cdots,\mathbf{x}_{m}^{(i+1)})$
+%block by block, according to the solution of the following
+%subproblem:
+%\begin{equation}
+%\mathbf{x}_{j}^{(k+1)}\leftarrow\underset{\mathbf{\xi}\in\mathcal{X}_{j}}{\text{argmin}}
+%f(\mathbf{x}_{1}^{(k+1)},\cdots,\mathbf{x}_{j-1}^{(k+1)},\mathbf{\xi},\mathbf{x}_{j+1}^{(k)},\cdots,\mathbf{x}_{m}^{(k)}).\label{eq:bcd-method}
+%\end{equation}
+%Also known as a \textit{non-linear Gauss-Seidel}
+%method~\cite{Bertsekas1999}, this algorithm updates one block each
+%time, always using the most recently updated values of other blocks
+%$\mathbf{x}_{\tilde{j}},\tilde{j}\ne j$. This is important since it
+%ensures that after each update, the objective function value does
+%not increase. For a sequence $\left\lbrace
+%\mathbf{x}^{(i)}\right\rbrace $ where each $\mathbf{x}^{(i)}$ is
+%generated by the BCD method, the following property holds.
+%\begin{theorem}
+%\label{thm:bcd}
+%Suppose $f$ is continuously differentiable in $\mathcal{X}=\mathcal{X}_{1}\times\dots\times\mathcal{X}_{m}$,
+%where $\mathcal{X}_{j}$, $j=1,\cdots,m$, are closed convex sets.
+%Furthermore, suppose that for all $j$ and $i$, the minimum of
+%\[
+%\min_{\mathbf{\mathbf{\xi}}\in\mathcal{X}_{j}}f(\mathbf{x}_{1}^{(k+1)},\cdots,\mathbf{x}_{j-1}^{(k+1)},\mathbf{\xi},\mathbf{x}_{j+1}^{(k)},\cdots,\mathbf{x}_{m}^{(k)})
+%\]
+%is uniquely attained. Let $\left\lbrace
+%\mathbf{x}^{(i)}\right\rbrace $ be the sequence generated by the
+%block coordinate descent method as in Eq.~\eqref{eq:bcd-method}.
+%Then, every limit point of $\left\lbrace
+%\mathbf{x}^{(i)}\right\rbrace $ is a stationary point. The
+%uniqueness of the minimum is not required for the case when  $m=2$
+%\cite{GS2000}.
+%\end{theorem}
+%
+%The proof of this theorem for an arbitrary number of blocks is shown
+%in Bertsekas~\cite{Bertsekas1999}.
+%For a non-convex optimization problem, most algorithms only guarantee
+%the stationarity of a limit point \cite{Lin2007}.
+%
+%When applying the BCD method to a constrained non-linear programming
+%problem, we have to ensure the following (a) it is critical to wisely choose a partition of
+%$\mathcal{X}$, whose Cartesian product constitutes $\mathcal{X}$. 
+%(b)  the dependency among the subproblems -- independent blocks can be 
+%computed simultaneously  (c) when updating a block, ensure we are 
+%using the most recent other blocks and (d) importantly, every block
+%must be solved optimally. Every NTF variant will have different the partition
+%scheme for blocks with natural partition such as scalar \cite{SL2001}, vector \cite{CP2009} and matrix \cite{KP2011}. 
+%Also, the update equation for determining the optimal block given
+%other blocks such as projected gradient, block principal pivoting \cite{KP2011}  etc. 
+%A detailed discussion of equivalences and comparison of different NMF and NTF algorithm
+%is presented at \cite{KHP2014}.
 
@@
 	\State $\Mn{M}{n} = \text{MTTKRP}(\TA,\{\Mn{H}{i}\},n)$
+		\label{line:MTTKRP}
 	\State $\Mn{S}{n} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$
+		\label{line:GH}
 	\State $\Mn{H}{n} = \text{NLS-Update}(\Mn{S}{n},\Mn{M}{n})$
+		\label{line:NLS}
 %	\State $\Mn[\hat]{H}{n} = \text{NLS-Update}(\Mn{S}{n},\Mn{M}{n})$
@@
 	\State $\Mn{G}{n} = \MnTra{H}{n}\Mn{H}{n}$
+		\label{line:Gn}
 	\EndFor
@@
 
-\subsection {Block Coordinate Descent(BCD) for NTF:}
+%\subsubsection{Hierarchical Alternating Least Squares (HALS)}
+%\label{sec:HALS}
+%
+%In the case of HALS \cite{CP2009}, updates are performed on individual columns of $\HH$ with all other entries in the factor matrices fixed.
+%The update rules \cite[Algorithm 2]{CP2009} can be written in closed form:
+%\SplitN{\label{eqn:halsupdate}} {
+%\Mn[\hat]{H}{n}_{\V{p}}(:,i) &\leftarrow \lt[ \Mn{H}{n}_{\V{p}}(:,i) + \Mn{M}{n}_{\V{p}}(:,i) - (\Mn{H}{n}_{\V{p}} \Mn{S}{n})(:,i)  \rt]_+\\
+%% \ww^i &\leftarrow \lt[ \ww^i + (\AA\HH^T)^i - \WW (\HH \HH^T)^i \rt]_+
+%}, 
+%
+%where, $\Mn{H}{n}_{\V{p}}(:,i) \in \Rn{I_n/P}$ is the $i^{th}$ column of the factor matrix $\Mn{H}{n}_{\V{p}}$.
+%Note that the columns of $\Mn{H}{n}_{\V{p}}$ are updated in order, so that the most up-to-date values are always used, and these $nk$ updates can be done in an arbitrary order.
+In our case we are updating all the $k$ vectors of a factor $\Mn{H}{n}_{\V{p}}$ one by one.
+%
+%\subsection{Alternating Nonnegative Least Squares with Block Principal Pivoting}
+%\label{sec:BPP}
+%
+%Kim and Park \cite{KP2011} first proposed the algorithm and the detail of the BPP algorithm for NTF is explained subsequently in detailed at \cite{KHP2014}.
+In this paper, we will leverage it to solve the factors $\Mn{H}{n}$.
+%Block Principal Pivoting (BPP) is an active-set-like method for solving the NLS subproblems in Eq. \eqref{eqn:single NLS}.
+%\SplitN{\label{eqn:single NLS}}{
+%\min_{\xx\geq 0} \|\CC\XX-\BB\|_2.
+%}. 
+%Kannan, Ballard and Park \cite{KBP2018}, have explained solving the active-set based non-negative least squares 
+%using normal equations with $\CC^T\CC$ and $\CC^T\BB$ for determining the non-negative factors of NMF. 
+%In our case, $\CC^T\CC$ is $\Mn{S}{n}$ and $\CC^T\BB$ is $\Mn{M}{n}_{\V{p}}$. We use the same implementation
+%of BPP from $MPIFAUN$.
+%
+%Our paper can be extended to other NTF algorithms such as Alternating Optimization and Alternating Direction Method of Multipliers (AO-ADMM) \cite{HSL2015, SBK2017} and Nestrov based updates
+%\cite{LKLHS2017} by changing the $Local-NLS-Update$ function of the proposed Algorithm \ref{alg:2D}.
 
-In this section, we will see relevant foundation for using  this
-framework.  The explanation of this section is mainly motivated out of
-Kim, He and Park \cite{KHP2014}. Consider a constrained non-linear optimization problem
-as follows:
-\begin{gather}
-\min f(x)\:\mbox{ subject to }\:x \in\mathcal{X},\label{eq:general_nonlinear}
-\end{gather}
-Here,  $\mathcal{X}$ is a closed convex subset of $\mathbb{R}^{n}$.
-An important assumption to be exploited in the BCD method is that
-the set $\mathcal{X}$ is represented by a Cartesian product:
-\begin{equation}
-\mathcal{X}=\mathcal{X}_{1}\times\cdots\times\mathcal{X}_{m},\label{eq:bcd-cartesian-product}
-\end{equation}
-where $\mathcal{X}_{j}$, $j=1,\cdots,m$, is a closed convex subset
-of $\mathbb{R}^{N_{j}}$, satisfying $n=\sum_{j=1}^{m}N_{j}$.
-Accordingly, the vector $\mathbf{x}$ is partitioned as
-$\mathbf{x}=(\mathbf{x}_{1},\cdots,\mathbf{x}_{m})$ so that
-$\mathbf{x}_{j}\in\mathcal{X}_{j}$ for $j=1,\cdots,m$. The BCD
-method solves for $\mathbf{x}_{j}$ by fixing all other subvectors of
-$\mathbf{x}$ in a cyclic manner. That is, if
-$\mathbf{x}^{(i)}=(\mathbf{x}_{1}^{(i)},\cdots,\mathbf{x}_{m}^{(i)})$
-is given as the current iterate at the $i^{th}$ step, the algorithm
-generates the next iterate
-$\mathbf{x}^{(i+1)}=(\mathbf{x}_{1}^{(i+1)},\cdots,\mathbf{x}_{m}^{(i+1)})$
-block by block, according to the solution of the following
-subproblem:
-\begin{equation}
-\mathbf{x}_{j}^{(k+1)}\leftarrow\underset{\mathbf{\xi}\in\mathcal{X}_{j}}{\text{argmin}}
-f(\mathbf{x}_{1}^{(k+1)},\cdots,\mathbf{x}_{j-1}^{(k+1)},\mathbf{\xi},\mathbf{x}_{j+1}^{(k)},\cdots,\mathbf{x}_{m}^{(k)}).\label{eq:bcd-method}
-\end{equation}
-Also known as a \textit{non-linear Gauss-Seidel}
-method~\cite{Bertsekas1999}, this algorithm updates one block each
-time, always using the most recently updated values of other blocks
-$\mathbf{x}_{\tilde{j}},\tilde{j}\ne j$. This is important since it
-ensures that after each update, the objective function value does
-not increase. For a sequence $\left\lbrace
-\mathbf{x}^{(i)}\right\rbrace $ where each $\mathbf{x}^{(i)}$ is
-generated by the BCD method, the following property holds.
-\begin{theorem}
-\label{thm:bcd}
-Suppose $f$ is continuously differentiable in $\mathcal{X}=\mathcal{X}_{1}\times\dots\times\mathcal{X}_{m}$,
-where $\mathcal{X}_{j}$, $j=1,\cdots,m$, are closed convex sets.
-Furthermore, suppose that for all $j$ and $i$, the minimum of
-\[
-\min_{\mathbf{\mathbf{\xi}}\in\mathcal{X}_{j}}f(\mathbf{x}_{1}^{(k+1)},\cdots,\mathbf{x}_{j-1}^{(k+1)},\mathbf{\xi},\mathbf{x}_{j+1}^{(k)},\cdots,\mathbf{x}_{m}^{(k)})
-\]
-is uniquely attained. Let $\left\lbrace
-\mathbf{x}^{(i)}\right\rbrace $ be the sequence generated by the
-block coordinate descent method as in Eq.~\eqref{eq:bcd-method}.
-Then, every limit point of $\left\lbrace
-\mathbf{x}^{(i)}\right\rbrace $ is a stationary point. The
-uniqueness of the minimum is not required for the case when  $m=2$
-\cite{GS2000}.
-\end{theorem}
+\subsection{Parallel Computation Model}
 
-The proof of this theorem for an arbitrary number of blocks is shown
-in Bertsekas~\cite{Bertsekas1999}.
-For a non-convex optimization problem, most algorithms only guarantee
-the stationarity of a limit point \cite{Lin2007}.
-
-When applying the BCD method to a constrained non-linear programming
-problem, we have to ensure the following (a) it is critical to wisely choose a partition of
-$\mathcal{X}$, whose Cartesian product constitutes $\mathcal{X}$. 
-(b)  the dependency among the subproblems -- independent blocks can be 
-computed simultaneously  (c) when updating a block, ensure we are 
-using the most recent other blocks and (d) importantly, every block
-must be solved optimally. Every NTF variant will have different the partition
-scheme for blocks with natural partition such as scalar, vector \cite{CP2009} and matrix \cite{KP2011}. 
-Also, the update equation for determining the optimal block given
-other blocks such as projected gradient, block principal pivoting \cite{KP2011}  etc. 
-A detailed discussion of equivalences and comparison of different NMF and NTF algorithm
-is presented at \cite{KHP2014}.
-
-In the rest of the section, we will discuss the update equation for different 
-algorithms, given the local matrix $\Mn{S}{n}$ and $\Mn{M}{n}_{\V{p}}$ as described in the
-Algorithm \ref{alg:2D}. In a similar direction, The paper $MPIFAUN$ \cite{KBP2018}, have detailed some of 
-these algorithms on their paper towards NMF.
-
-\subsubsection{Hierarchical Alternating Least Squares (HALS)}
-\label{sec:HALS}
-
-In the case of HALS \cite{CP2009}, updates are performed on individual columns of $\HH$ with all other entries in the factor matrices fixed.
-The update rules \cite[Algorithm 2]{CP2009} can be written in closed form:
-\SplitN{\label{eqn:halsupdate}} {
-\Mn[\hat]{H}{n}_{\V{p}}(:,i) &\leftarrow \lt[ \Mn{H}{n}_{\V{p}}(:,i) + \Mn{M}{n}_{\V{p}}(:,i) - (\Mn{H}{n}_{\V{p}} \Mn{S}{n})(:,i)  \rt]_+\\
-% \ww^i &\leftarrow \lt[ \ww^i + (\AA\HH^T)^i - \WW (\HH \HH^T)^i \rt]_+
-},
-
-where, $\Mn{H}{n}_{\V{p}}(:,i) \in \Rn{I_n/P}$ is the $i^{th}$ column of the factor matrix $\Mn{H}{n}_{\V{p}}$.
-Note that the columns of $\Mn{H}{n}_{\V{p}}$ are updated in order, so that the most up-to-date values are always used, and these $nk$ updates can be done in an arbitrary order.
-In our case we are updating all the $k$ vectors of a factor $\Mn{H}{n}_{\V{p}}$ one by one.
-
-\subsection{Alternating Nonnegative Least Squares with Block Principal Pivoting}
-\label{sec:BPP}
-
-Kim and Park \cite{KP2011} first proposed the algorithm and the detail of the BPP algorithm for NTF is explained subsequently in detailed at \cite{KHP2014}.
-In this paper, we will leverage it to solve the factors $\Mn{H}{n}$.
-Block Principal Pivoting (BPP) is an active-set-like method for solving the NLS subproblems in Eq. \eqref{eqn:single NLS}.
-\SplitN{\label{eqn:single NLS}}{
-\min_{\xx\geq 0} \|\CC\XX-\BB\|_2.
-}. 
-Kannan, Ballard and Park \cite{KBP2018}, have explained solving the active-set based non-negative least squares 
-using normal equations with $\CC^T\CC$ and $\CC^T\BB$ for determining the non-negative factors of NMF. 
-In our case, $\CC^T\CC$ is $\Mn{S}{n}$ and $\CC^T\BB$ is $\Mn{M}{n}_{\V{p}}$. We use the same implementation
-of BPP from $MPIFAUN$.
-
-Our paper can be extended to other NTF algorithms such as AO-ADMM \cite{HSL2015} and Nestrov based updates
-\cite{LKLHS2017} by changing the $Local-NLS-Update$ function of the proposed Algorithm \ref{alg:2D}.
+%MPI collectives for Prelims
+To analyze our algorithms we use the MPI model of distributed-memory parallel computation, where we assume a fully connected network. 
+Sending a message of $W$ words from one processor to another costs $\alpha + \beta W$, where $\alpha$ is the latency and $\beta$ to be the per word or bandwidth cost. 
+In particular, we will use collective communication over groups of $P$ processors, and we will assume the use of efficient algorithms \cite{TRG05,CH+07}.
+In this case, an All-Reduce, which sums data initially distributed across processors and stores the result of size $W$ redundantly on every processor, costs $2\alpha \log P + 2\beta W (P-1)/P$.
+An All-Gather collects data initially distributed across processors and stores the union of size $W$ redundantly on all processors and costs $\alpha \log P + \beta W (P-1)/P$.
+A Reduce-Scatter sums data initially distributed across processors and partitions the result across processors, which costs $\alpha \log P + \beta W (P-1)/P$, where $W$ is the size of the data that each processor initially stores.
+Reduction operations also include a flop cost but we will omit it because it is usually dominated by communication.

### 2018-03-28 latex/sc18/experiments.tex
synthetic and machine details included

@@
 \subsubsection{Real World}
+
 \subsubsection{Synthetic}
+For the synthetic datasets, our open source code supports (a) low-rank tensor (b) uniform random and (c) positive shifted normal distribution of $\mathfrak{N}(3,1)$ -- that is change the mean such that all the random numbers are positive.
+We considered three different synthetic matrices for different cases.
+For baseline comparison with Liavas \cite{LKLHS18}, we considered a three mode
+uniform of size 1024x1024x1024 on processor grids $2^k \times 2^k \times 2^k$ for $k \in {0,1,2,3}$. We used the uniform five 
+mode synthetic tensor with dimension $64\times 64\times 64\times 64\times 64$ on processor grids 
+$1\times1\times1\times1\times1$, $2\times1\times1\times1\times1$, $\dots$, $2\times2\times2\times2\times2$ for strong scaling 
+experiments.  In the case of weak scaling of four mode synthetic tensors with (D) and without (N) the use of dimension trees.  The 
+tensor and processor grid dimensions are $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$.
+In all the cases the dimensions were considered such that synthetic tensors can be accomodated even on single node with 64GB for scale up plots.
 
 \subsection{Machine Details}
+The entire experimentation was performed on Eos -- a super computer in Oak Ridge Leadership Computing Facility (OLCF). 
+Eos is a 736-node Cray® XC30? cluster with a total of 47.104TB of memory. Its processor is the Intel® Xeon® E5-2670. It features 
+16 I/O service nodes and 2 external login nodes.
+Its compute nodes are organized in blades.
+Each blade contains 4 nodes and every node has 2 sockets with 8 physical cores and 64GB memory.
+Even though the machine support Intel's hyper-threading (HT), we restricted the total number of OpenMP threads to match the number of physical cores as HT offers minimal improvement for
+ BLAS and LAPACK operations. In total, the Eos compute partition contains 11,776 traditional processor cores (23,552 logical cores 
+ with HT Technology enabled) and some of the experiments peaked to 8,192 out of 11,776 physical cores that is approximately 
+ 70\% of the peak capacity.
 

### 2018-03-28 latex/sc18/experiments.tex
strong and weak scaling

@@
 \subsubsection{Real World}
+{\em Hyperspectral Images(HSI):} For real world datasets and comparison with the baseline Liavas \cite{LKLHS18}, we considered the same hyperspectral imaging datasets from \cite{FAN16}.
+Each hyperspectral image has dimensions 1024 x 1344 x 33 and represents a set of 33 greyscale images of size 1344 (H) x 1024 (V) pixels sampled at wavelengths 400, 410, ..., 720 nm, with each pixel value representing spectral radiance in $W m^{-2} sr^{-1} nm^{-1}$.
+In each scene, hyperspectral images were acquired at about 1-hour intervals.
+Since, the baseline can work only for three mode real world data, we considered one image and to demonstrate the proposed algorithm's capability we considered the entire set of 9 images collected over 9 time step as  1024 x 1344 x 33 x 9.
 
@@
 
+The strong scaling on 3D synthetic tensor with baseline is shown in Figure \ref{fig:strongsynthetic3D}. 
+We can observe from the figure that all the three algorithms scales linearly as the problem is more
+compute bound in the case of dense tensors. The computation is dominated by $O(R \prod I_k / P_k)$
+and the communication scales as $P^{1/N}$. As evident from the figure, this communication cost doesn't 
+show up even for $P$ in the order of few hundreds. We are \koby{xx} faster than the baseline 
+{\em Liavas}\cite{LKLHS18}. The advantages of the dimension tree more pronounced only in the higher
+modes because of avoiding unnecessary KRP computation that is more memory bound. Even though, in
+the Figure \ref{fig:strongsynthetic3D} the dimension trees is only as good as the classical MTTKRP, we are witnessing
+2.5x speed up with dimension trees in Figure \ref{fig:strongsynthetic5D}. Also, for tensors with equal dimensions over
+modes, dimension trees has some advantage with favorable shapes for matrix multiplication DGEMM.
+
 \begin{figure}
@@
 \caption{Strong scaling of 3D synthetic tensor with dimension $1024\times 1024\times 1024$ on processor grids $2^k\times 2^k\times 2^k$ for $k\in\{0,1,2,3\}$.  The rank is fixed at 32.}
+\label{fig:strongsynthetic3D}
 \end{figure}
@@
 The rank is fixed at 32.}
+\label{fig:strongsynthetic5D}
 \end{figure}
@@
 \end{itemize}
+
+In the case of weakscaling, we wanted to understand the time it takes to solve bigger problems by increasing the number of nodes with the increase in data size by keeping the problem size (workload) assigned to each processing element constant.
+We conducted this experiment by using synthetic tensor and processor grid of dimensions $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$, and the rank is fixed at 32.
+For eg., in the case of 3 x 3 x 3 x 3 processor grid with 81 nodes, we ran the proposed algorithm with and without dimension trees on a four mode tensor of size 384 x 384 x 384 x 384.
+The results of the breakdown plot is shown in Figure \ref{fig:weaksynthetic4D}.
+Since, the NTF on dense tensors is compute intensive, the total time of the weak scaling is staying fixed for both the case of with and without dimension trees.
+The entire running is completely dominated by the MTTKRP computation.
+As mentioned above, from the Figure   \ref{fig:weaksynthetic4D}, we can observe that the KRP cost of dimension trees is insignificant and it offers 2.5x speed up over classical MTTKRP.
 
@@
 The reported times are per iteration.}
+\label{fig:weaksynthetic4D}
 \end{figure}

### 2018-03-28 latex/sc18/survey.tex

--- 2018-03-07 latex/sc18/survey.tex
@@
 
-\grey{I haven't organized these very well...}
-\begin{itemize}
-	\item NLS algorithms
-	\begin{itemize}
-		\item MU, HALS, BPP, etc.
-	\end{itemize}
-	\item parallel sparse CP
-	\begin{itemize}
-		\item Karypis/Smith
-		\item Kaya/Ucar
-		\item Vagelis/Faloutsos?
-		\item Li/Vuduc?
-	\end{itemize}
-	\item dense tensor computations
-	\begin{itemize}
-		\item Phan/Cichocki
-		\item Austin/Ballard/Kolda IPDPS16
-		\item Ballard/Knight/Rouse IPDPS18
-	\end{itemize}
-	\item parallel dense CP
-	\begin{itemize}
-		\item Liavs et al.
-		\item Cichocki?
-		\item distributed incremental descent?
-	\end{itemize} 
-\end{itemize}
+The formulation of NNCP with least squares error and algorithms for computing it go back to \cite{Paatero97,WW01}, developed in part as a generalization of nonnegative matrix factorization algorithms \cite{LS99} to tensors.
+Sidiropoulos et al.
+\cite{SLFHPF2017} provide a more detailed and complete survey that includes basic tensor factorization models with and without constraints, broad coverage of algorithms, and recent driving applications.
+The tensor operations discussed and the notation used in this paper follow Kolda and Bader's survey \cite{KB2009}.
+
+Recently, there has been growing interest in scaling tensor operations to bigger data and more processors in both the data mining/machine learning and the high performance computing communities. 
+For sparse tensors, there have been parallelization efforts to compute CP decompositions both on shared-memory platforms \cite{SRSK2015,LCPSV17} as well as distributed-memory platforms \cite{KU16,SK16,KU18}, and these approaches can be generalized to constrained problems \cite{SBK2017}.
+The focus of this work is on dense tensors, but many of the ideas for sparse tensors are applicable to the dense case, including parallel data distributions and techniques to avoid recomputation across modes.
+
+In particular, Liavas et al. \cite{LK+17b} extend a parallel algorithm designed for sparse tensors \cite{SK16} to the 3D dense case.
+They use the ``medium-grained'' dense tensor distribution and row-wise factor matrix distribution, which is exactly the same as our distribution strategy (see \cref{sec:datadist}), and they use a Nesterov-based algorithm to enforce the nonnegativity constraints.
+Their code is publicly available, and we compare our performance with theirs in \cref{sec:experiments}.
+A similar data distribution and parallel algorithm for computing a single dense MTTKRP computation is proposed by Ballard, Knight, and Rouse \cite{BKR17-TR}. 
+They prove that the algorithm is communication optimal, but they do not provide an implementation.
+Another approach to parallelizing NNCP decomposition of dense tensors is presented by Phan and Cichocki \cite{PC11}, but they use a dynamic tensor factorization, which performs different, more independent computations across processors.
+
+The idea of using dimension trees (discussed in \cref{sec:dimtrees}) to avoid recomputation within MTTKRPs across modes is introduced in \cite{PTC13a} for computing the CP decomposition of dense tensors.
+It has also been used for sparse CP \cite{LCPSV17,KU18} and other tensor computations \cite{KU16}.
+
+%Given these recent work on the sparse tensors, we survey the literature around dense tensor -- the interest of this paper.
+Unlike sparse tensor, there are very few activities around dense tensors that are important for analyzing tensors from images and scientific world such as 2D Ronchigram from pytography data.
+With the advent of the deep learning and the interest of the community in processing higher order tensors from hyper spectral image and medical imaging, we have used these real world in conducting experiments of this paper.
+%
+%Austin, Ballard, and Kolda \cite{ABK16} distributed-memory parallel implementation for the Tucker decomposition to compress 8TB of scientific data.
+Tucker decomposition is unconstrained and a completely different decomposition over constrained Non-negative Tensor Factorization explained in this paper.
+Recently,  Ballard, Knight, Rouse \cite{BKR17-TR} have proposed Communication Lower Bounds for Matricized Tensor Times Khatri-Rao Product under particular distribution of dense tensors.
+The proposed algorithm \ref{alg:2D} is motivated out of this work.
+%
+%With this detailed survey, we would like to highlight the contribution of this paper.
+
+%\subsection{Contributions:}

### 2018-03-28 latex/sc18/experiments.tex
experiments liavas reference changed to LK+17b

@@
 \subsubsection{Real World}
-{\em Hyperspectral Images(HSI):} For real world datasets and comparison with the baseline Liavas \cite{LKLHS18}, we considered the same hyperspectral imaging datasets from \cite{FAN16}.
+{\em Hyperspectral Images(HSI):} For real world datasets and comparison with the baseline Liavas \cite{LK+17b}, we considered the same hyperspectral imaging datasets from \cite{FAN16}.
 Each hyperspectral image has dimensions 1024 x 1344 x 33 and represents a set of 33 greyscale images of size 1344 (H) x 1024 (V) pixels sampled at wavelengths 400, 410, ..., 720 nm, with each pixel value representing spectral radiance in $W m^{-2} sr^{-1} nm^{-1}$.
@@
 We considered three different synthetic matrices for different cases.
-For baseline comparison with Liavas \cite{LKLHS18}, we considered a three mode
+For baseline comparison with Liavas \cite{LK+17b}, we considered a three mode
 uniform of size 1024x1024x1024 on processor grids $2^k \times 2^k \times 2^k$ for $k \in {0,1,2,3}$. We used the uniform five 
@@
 show up even for $P$ in the order of few hundreds. We are \koby{xx} faster than the baseline 
-{\em Liavas}\cite{LKLHS18}. The advantages of the dimension tree more pronounced only in the higher
+{\em Liavas}\cite{LK+17b}. The advantages of the dimension tree more pronounced only in the higher
 modes because of avoiding unnecessary KRP computation that is more memory bound. Even though, in

### 2018-03-28 latex/sc18/introduction.tex
edits all around apparently

@@
+% !TEX root = paper.tex
+
 \section{Introduction}
 
-\grey{Contributions:
 \begin{itemize}
-	\item fastest scalable NNCP (and unsonstrained CP) software for dense tensors
-	\item computation efficient using dimension trees, communication efficient using optimal parallel algorithm
-	\item available open-source software, with flexibility to incorporate multiple NLS algorithms
+	\item CP and its applications
+	\item NNCP and its specific applications (2D R...g raph \cite{JC+16}, MD protein folding simultation analysis, phenotypes of patients Ghosh/Sun
+	\item importance of preserving multidimensional relationships
+	\item parallelization strategy, bottlenecked by MTTKRP but independence in solve
+	\item challenge is to perform matrix operations with different views of tensor
+	\item summarize our approach, highlight comptuational avoidance, communication avoidance
+	\item state bulleted list of contributions
+	\begin{itemize}
+		\item algorithm and analysis
+		\item implementation and benchmrking, showing near-lienar scaling and improvement over existed code
+		\item generalization to higher modes
+		\item scaled up to 512 GB synthetic data
 	\end{itemize}
-}
+\end{itemize}
+
+The CP decomposition is a low-rank approximation of a multi-dimensional array, or tensor, which generalizes matrix approximations like the truncated singular value decomposition.
+CP is often used for finding hidden patterns, or latent factors, within tensor data, particularly when the goal is to interpret the factors, and it is popular within the signal processing, machine learning, and scientific computing communities.
+To aid in interpretability, domain-specific constraints are often imposed on the computed factors.
+We focus in this paper on constraining solutions to have nonnegative entries, which is useful when the tensor data itself is nonnegative.
+For example, when tensor entries

### 2018-03-28 latex/sc18/plots.tex
edits all around apparently

@@
 	\addplot table[x=alg-p, y expr=((\thisrow{err_compute}+\thisrow{err_communication})/(\minvalue*\numiterations))] {\datafile};
-	\legend{MTTRKP,KRP,Factor Comm,NLS,Error};
+	\legend{MTTKRP,KRP,Factor Comm,NLS,Error};
 \end{axis}
@@
 	%\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={p}, y={totaltime}] {\datafile};
-	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=triangle*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
-	\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
+	%\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=triangle*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
+	%\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
+	\addplot+ [discard if not={alg}{DF},thick,mark options={solid},mark=square*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
+	\addplot+ [discard if not={alg}{NF},thick,mark options={solid},mark=square*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
 	\ifliavas
@@
 	\else
-		\legend{DimTree,NoDimTree};
+		\legend{DimTree,NoDimTree,FlatDimTree,FlatNoDimTree};
 	\fi

### 2018-03-28 latex/sc18/prelims.tex
edits all around apparently

@@
 In this work, we are interested in NNCP models that are good approximations to $\TA$ in the least squares sense.
-That is, we seek to minimize $\|\TA-\CPl\|$, where the tensor norm is a generalization of the matrix Frobenius or vector 2-norm, the square root of the sum of the squares of the entries.
+That is, we seek to minimize $\|\TA-\CPl\|$, where the tensor norm is a generalization of the matrix Frobenius or vector 2-norm, the square root of the sum of squares of the entries.
 
 The $n$th mode matricized tensor denoted by $\M{A}_{(n)}$ is a $I_n\times I/I_n$ matrix formed by organizing the $n$th mode fibers of a tensor $\T{X}$ with dimensions $I_{1} \times ...
-\times I_{N}$ into the columns of a matrix.
+\times I_{N}$ (and $I=\prod I_n$) into the columns of a matrix.
 The Matricized Tensor Times Khatri Rao Product or MTTKRP will be central to this work and takes the form $\Mn{M}{n} = \Mz{A}{n}( \Mn{H}{N} \Khat \cdots \Khat \Mn{H}{n+1} \Khat \Mn{H}{n-1} \Khat  \cdots \Khat \Mn{H}{1})$.
@@
 We use the Block Principal Pivoting (BPP) method \cite{KP2011,KHP2014}, which is an active-set-like method, to solve the subproblem exactly.
-BPP is an active-set-like method for solving the NLS subproblems in Eq. \eqref{eqn:single NLS}.
+%BPP is an active-set-like method for solving the NLS subproblems in Eq. \eqref{eqn:single NLS}.
 

### 2018-03-28 latex/sc18/experiments.tex
k sweep also completed. plots fixed for both the flat and non-flat case

@@
 and the communication scales as $P^{1/N}$. As evident from the figure, this communication cost doesn't 
-show up even for $P$ in the order of few hundreds. We are \koby{xx} faster than the baseline 
+show up even for $P$ in the order of few hundreds. Our proposed algorithm with dimension trees is 1.5x faster than the baseline 
 {\em Liavas}\cite{LK+17b}. The advantages of the dimension tree more pronounced only in the higher
@@
 the Figure \ref{fig:strongsynthetic3D} the dimension trees is only as good as the classical MTTKRP, we are witnessing
-2.5x speed up with dimension trees in Figure \ref{fig:strongsynthetic5D}. Also, for tensors with equal dimensions over
-modes, dimension trees has some advantage with favorable shapes for matrix multiplication DGEMM.
+2.5x speed up with dimension trees over higher modes as in Figure \ref{fig:strongsynthetic5D}. 
+Also, for tensors with equal dimensions over all the
+modes, dimension trees has some advantage with favorable shapes for matrix multiplication with BLAS DGEMM.
 
@@
 \caption{Per-iteration time breakdown of our implementation (using dimension trees) over various ranks for a time-lapse HSI dataset with dimensions $1344\times 1024\times 33 \times 9$ on 64 processors arranged in a $8\times8\times1\times1$ grid.}
+\label{fig:ksweep4DHSI}
 \end{figure}
+
+According to Kim, He and Park \cite{KHP2014}, in the case of dense tensors, as the low rank $k$ increases, the approximation 
+error  $\|\TA - \T{M}\|$ decreases because it is obtaining bigger low rank factors. However, there is a law of diminishing advantages, 
+that an increase in $k$ in lower values, say from 5 to 10, will have significant improvement in relative error over increase in higher 
+values, like 100 to 105. Hence, it is common practice in the community to sweep $k$, to obtain better approximation error within the 
+manageable computation. Towards this end, we wanted to understand the increase in running time over increasing the low rank $k
+$. The results of the experiments on real world four mode HSI tensor is presented in Figure \ref{fig:ksweep4DHSI}. We can observe 
+that, the multi-TTV({\em mTTV}) scales linearly with the increasing $k$,  where as the partial MTTKRP ({\em PM}) is scaling super 
+linearly. This is because, with higher $k$, the BLAS $DGEMM$ gets a bigger chunk of work that it can optimize better.  The local 
+compute $NLS$ is increasing as expected nearly $O(k^3)$ and the $ALL-GATHER$ required for the gram matrices over $k^2$ 
+elements, becomes significant on higher $k$'s.
 

### 2018-03-28 latex/sc18/plots.tex
k sweep also completed. plots fixed for both the flat and non-flat case

--- 2018-03-26 latex/sc18/plots.tex
@@
 	%\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={p}, y={totaltime}] {\datafile};
+	\ifliavas
+		\addplot+ [discard if not={alg}{DF},thick,mark options={solid},mark=square*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
+		\addplot+ [discard if not={alg}{NF},thick,mark options={solid},mark=square*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
+	\else
 		\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=triangle*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
 		\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
+	\fi
 	\ifliavas

### 2018-03-29 latex/sc18/experiments.tex
some changes to experiments section

@@
 The strong scaling on 3D synthetic tensor with baseline is shown in Figure \ref{fig:strongsynthetic3D}. 
-We can observe from the figure that all the three algorithms scales linearly as the problem is more
+We can observe from the figure that all the three algorithms scale linearly as the problem is more
 compute bound in the case of dense tensors. The computation is dominated by $O(R \prod I_k / P_k)$
@@
 show up even for $P$ in the order of few hundreds. Our proposed algorithm with dimension trees is 1.5x faster than the baseline 
-{\em Liavas}\cite{LK+17b}. The advantages of the dimension tree more pronounced only in the higher
-modes because of avoiding unnecessary KRP computation that is more memory bound. Even though, in
+{\em Liavas}\cite{LK+17b}. The advantages of the dimension tree are more pronounced in higher
+mode tensors because of more reuse over partial MTTKRPs and avoiding unnecessary KRP computation that is more memory bound. Even though, in
 the Figure \ref{fig:strongsynthetic3D} the dimension trees is only as good as the classical MTTKRP, we are witnessing

### 2018-03-29 latex/sc18/experiments.tex
strong synthetic scaling fixed arguments

@@
 show up even for $P$ in the order of few hundreds. Our proposed algorithm with dimension trees is 1.5x faster than the baseline 
-{\em Liavas}\cite{LK+17b}. The advantages of the dimension tree are more pronounced in higher
-mode tensors because of more reuse over partial MTTKRPs and avoiding unnecessary KRP computation that is more memory bound. Even though, in
-the Figure \ref{fig:strongsynthetic3D} the dimension trees is only as good as the classical MTTKRP, we are witnessing
+{\em Liavas}\cite{LK+17b}. 
+The dimension trees are faster over classical MTTKRP, in both the three and five mode synthetic tensors. We are witnessing
 2.5x speed up with dimension trees over higher modes as in Figure \ref{fig:strongsynthetic5D}. 
-Also, for tensors with equal dimensions over all the
+The advantages of the dimension tree are more pronounced in higher
+mode tensors because of more reuse over partial MTTKRPs and avoiding unnecessary KRP computation that is more memory bound.  Also, for tensors with equal dimensions over all the
 modes, dimension trees has some advantage with favorable shapes for matrix multiplication with BLAS DGEMM.

### 2018-03-29 latex/sc18/introduction.tex
finished draft of intro

--- 2018-03-07 latex/sc18/introduction.tex
@@
+% !TEX root = paper.tex
+
 \section{Introduction}
 
-\grey{Contributions:
+The CP decomposition is a low-rank approximation of a multi-dimensional array, or tensor, which generalizes matrix approximations like the truncated singular value decomposition.
+CP is often used for finding hidden patterns, or latent factors, within tensor data, particularly when the goal is to interpret the factors, and it is popular within the signal processing, machine learning, and scientific computing communities.
+
+To aid in interpretability, domain-specific constraints are often imposed on the computed factors.
+We focus in this paper on dense tensors (when nearly all of the tensor entries are nonzero) and on constraining solutions to have nonnegative entries, which is useful when the tensor data itself is nonnegative.
+For example, in imaging and microscopy applications, tensor values often correspond to intensities, and NNCP can be used to cluster and analyze the data in a lower-dimensional space \cite{JC+16}.
+In this work, we consider two such applications: a series of time-lapse hyperspectral images \cite{FAN16} and a dynamic functional correlation data set generated from functional magnetic resonance images of human brains \cite{VEU+12}.
+
+One approach to handling multidimensional data is to ``matricize'' it, combining sets of modes to reshape the data into a matrix, so that standard matrix  methods like principal component analysis or nonnegative matrix factorization can be applied.
+While this approach can be effective in certain cases, reshaping the data destroys multidimensional relationships among entries that the matrix methods cannot recover.
+By maintaining the tensor structure of the data, the low-rank representations preserve these relationships, often producing better and more interpretable results.
+
+However, tensor methods are more complicated both mathematically and computationally.
+The kernel computations within standard algorithms for computing NNCP can be formulated as matrix computations, but the complicated layout of tensors in memory prevents the straightforward use of BLAS and LAPACK libraries.
+In particular, the matrix formulation of subcomputations involve different views of the tensor data, so no single layout yields a column- or row-major matrix layout for all subcomputations.
+Likewise, the parallelization approach for tensor methods is not a straightforward application of parallel matrix computation algorithms.
+
+In developing an efficient parallel algorithm for computing a NNCP of a dense tensor, the key is to parallelize the bottleneck computation known as Matricized-Tensor Times Khatri-Rao Product (MTTKRP), which is performed repeatedly for each mode of the tensor.
+The parallelization must load balance the computation, minimize communication across processors, and distribute the results so that the rest of the computation can be performed independently.
+In our algorithm, not only do we load balance the computation, but we also compute and store temporary values that can be used across MTTKRPs of different modes using a technique known as dimension trees, significantly reducing the computational cost compared to standard approaches.
+Our parallelization strategy also avoids communicating tensor entries and minimizes the communication of factor matrix entries, helping the algorithm to remain computation bound and scalable to high core counts.
+
+As we detail in the related work, the general techniques for reducing computation and communication are not new.
+The recomputation avoidance was proposed in a sequential algorithm \cite{PTC13a}, and the parallelization scheme was proposed and analyzed for general tensors \cite{BKR17-TR} and the algorithm was implemented for 3D tensors \cite{LK+17a}.
+The main contribution of this paper is the combination and implementation of these ideas for general $N$D tensors, and we demonstrate superior performance to the existing implementation for 3D tensors \cite{LK+17a}.
+
+We summarize our main contributions as follows:
 \begin{itemize}
-	\item fastest scalable NNCP (and unsonstrained CP) software for dense tensors
-	\item computation efficient using dimension trees, communication efficient using optimal parallel algorithm
-	\item available open-source software, with flexibility to incorporate multiple NLS algorithms
+		\item we present a computation- and communication-efficient algorithm for NNCP with detailed cost analysis,
+		\item our algorithm is designed for general $N$-way tensors,
+		\item we demonstrate a performance improvement of up to $2.2\times$ over the existing state-of-the-art parallel software on 3D tensors,
+		\item we demonstrate efficient parallel scaling of up to \koby{$XX\times$} on \koby{$YY$} cores.
 \end{itemize}
-}

### 2018-03-29 latex/sc18/prelims.tex
finished draft of intro

@@
 \times I_{N}$ (and $I=\prod I_n$) into the columns of a matrix.
-The Matricized Tensor Times Khatri Rao Product or MTTKRP will be central to this work and takes the form $\Mn{M}{n} = \Mz{A}{n}( \Mn{H}{N} \Khat \cdots \Khat \Mn{H}{n+1} \Khat \Mn{H}{n-1} \Khat  \cdots \Khat \Mn{H}{1})$.
+The Matricized-Tensor Times Khatri-Rao Product or MTTKRP will be central to this work and takes the form $\Mn{M}{n} = \Mz{A}{n}( \Mn{H}{N} \Khat \cdots \Khat \Mn{H}{n+1} \Khat \Mn{H}{n-1} \Khat  \cdots \Khat \Mn{H}{1})$.
 

### 2018-03-29 latex/sc18/abstract.tex
edits all around apparently
minor changes

@@
 
-The CP tensor decomposition is a low-rank approximation of multidimensional (or tensor) data. 
+The CP tensor decomposition is a low-rank approximation of a tensor. 
 We present a distributed-memory parallel algorithm and implementation of an alternating optimization method for computing a CP decomposition of dense tensor data that can enforce nonnegativity of the computed low-rank factors.
@@
 Our approach is also communication efficient, using a data distribution and parallel algorithm across a multidimensional processor grid that can be tuned to minimize communication.
-We benchmark our software on synthetic as well as hyperspectral image data, demonstrating that our algorithm scales well to 100s of nodes (up to 9600 cores) and is faster than the currently available parallel software.
+We benchmark our software on synthetic as well as hyperspectral image and neuroscience dynamic functional connectivity data, demonstrating that our algorithm scales well to \grey{100s} of nodes (up to \grey{9600} cores) and is faster and more general than the currently available parallel software.
 

### 2018-03-29 latex/sc18/introduction.tex
minor changes

@@
 As we detail in the related work, the general techniques for reducing computation and communication are not new.
-The recomputation avoidance was proposed in a sequential algorithm \cite{PTC13a}, and the parallelization scheme was proposed and analyzed for general tensors \cite{BKR17-TR} and the algorithm was implemented for 3D tensors \cite{LK+17a}.
-The main contribution of this paper is the combination and implementation of these ideas for general $N$D tensors, and we demonstrate superior performance to the existing implementation for 3D tensors \cite{LK+17a}.
+The recomputation avoidance was proposed in a sequential algorithm \cite{PTC13a}, and the parallelization scheme was proposed and analyzed for general tensors \cite{BKR17-TR} and the algorithm was implemented for 3D tensors \cite{LK+17b}.
+The main contribution of this paper is the combination and implementation of these ideas for general $N$D tensors, and we demonstrate superior performance to the existing implementation for 3D tensors \cite{LK+17b}.
 
@@
 		\item our algorithm is designed for general $N$-way tensors,
-		\item we demonstrate a performance improvement of up to $2.2\times$ over the existing state-of-the-art parallel software on 3D tensors,
-		\item we demonstrate efficient parallel scaling of up to \koby{$XX\times$} on \koby{$YY$} cores.
+		\item we demonstrate a performance improvement of up to \grey{$2.2\times$} over the existing state-of-the-art parallel software on 3D tensors,
+		\item we demonstrate efficient parallel scaling of up to \grey{$527\times$} on \grey{$512$} cores.
 \end{itemize}

### 2018-03-29 latex/sc18/experiments.tex
ksweep neuroscience, strong scaling 3D hsi

@@
 
+The Figure \ref{fig:stronghsi3D} shows the strong scaling on the three modes HSI real world data. We can observe
+that our proposed algorithm with dimension trees is almost 2x faster than baseline Liavas. The proposed algorithm
+with classical MTTKRP was performing better on higher number of processors over the baseline.
+
 \begin{figure}
@@
 \label{fig:strongsynthetic5D}
+\end{figure}
+
+\begin{figure}
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/str_3D_HSI.dat}
+\renewcommand{\numiterations}{10}
+\liavastrue
+\strongscalingplot
+\end{tikzpicture}
+\caption{Strong scaling of 3D HSI real world data with dimension 1024 x1344 x 33 on processor grids of $1 \times k \times k$ for $k \in {1, 2, 4, 8, 16, 32}$. The rank is fixed at 32.}
+\label{fig:stronghsi3D}
 \end{figure}
@@
 \label{fig:ksweep4DHSI}
+\end{figure}
+
+\begin{figure}
+\begin{tikzpicture}
+\begin{axis}[legend style={draw=none, anchor=north,  cells={align=left}, nodes={scale=0.7}}]
+	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
+	\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
+	\addplot+ [discard if not={alg}{L},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
+	\legend{DimTree,NoDimTree,Liavas}
+\end{axis}
+\end{tikzpicture}
+\caption{$k$ sweep plot for Neuroscience dataset}
+\label{fig:ksweepneuro}
 \end{figure}

### 2018-03-29 latex/sc18/experiments.tex
adjusted legend pos for ksweep neuro
minor edits

--- 2018-03-29 latex/sc18/experiments.tex
@@
 
-\subsubsection{Real World}
-{\em Hyperspectral Images(HSI):} For real world datasets and comparison with the baseline Liavas \cite{LK+17b}, we considered the same hyperspectral imaging datasets from \cite{FAN16}.
-Each hyperspectral image has dimensions 1024 x 1344 x 33 and represents a set of 33 greyscale images of size 1344 (H) x 1024 (V) pixels sampled at wavelengths 400, 410, ..., 720 nm, with each pixel value representing spectral radiance in $W m^{-2} sr^{-1} nm^{-1}$.
+\subsubsection{Hyperspectral Images (HSI)}
+For comparison with previous work \cite{LK+17b}, we consider the same 3D hyperspectral imaging dataset called ``Souto\_wood\_pile'' \cite{FAN16}. 
+The hyperspectral datacube has dimensions $1024 \times 1344 \times 33$ and represents a set of 33 grayscale images of size 1344 $\times$ 1024 pixels sampled at wavelengths 400, 410, $\dots$, 720 nm, with each pixel value representing spectral radiance in $W m^{-2} sr^{-1} nm^{-1}$.
+We also consider the Nogueir\'{o} scene dataset, which is a sequence of 9 time-lapse HSI images of the same scene acquired at about 1-hour intervals.
 In each scene, hyperspectral images were acquired at about 1-hour intervals. 
-Since, the baseline can work only for three mode real world data, we considered one image and to demonstrate the proposed algorithm's capability we considered the entire set of 9 images collected over 9 time step as  1024 x 1344 x 33 x 9.
+Each Nogueir\'{o} scene HSI image has the same properties as the Souto\_wood\_pile data set, so the corresponding tensor has dimensions $1024 \times 1344 \times 33 \times 9$.
+
+\subsubsection{Dynamic Functional Connectivity (dFC)}
 
@@
 tensor and processor grid dimensions are $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$.
-In all the cases the dimensions were considered such that synthetic tensors can be accomodated even on single node with 64GB for scale up plots.
+In all the cases the dimensions were considered such that synthetic tensors can be accommodated even on single node with 64GB for scale up plots.
 

### 2018-03-29 latex/sc18/experiments.tex

@@
 \label{fig:ksweep4DHSI}
+\end{figure}
+
+\begin{figure}
+\begin{tikzpicture}
+\begin{axis}[legend style={draw=none, anchor=north,  cells={align=left}, nodes={scale=0.7}}, legend pos=north west]
+	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
+	\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
+	\addplot+ [discard if not={alg}{L},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
+	\legend{DimTree,NoDimTree,Liavas}
+\end{axis}
+\end{tikzpicture}
+\caption{$k$ sweep plot for Neuroscience dataset}
+\label{fig:ksweepneuro}
 \end{figure}

### 2018-03-29 latex/sc18/experiments.tex
strong scaling of HSI complete

--- 2018-03-29 latex/sc18/experiments.tex
@@
 that our proposed algorithm with dimension trees is almost 2x faster than baseline Liavas. The proposed algorithm
-with classical MTTKRP was performing better on higher number of processors over the baseline.
+with classical MTTKRP was performing better on higher number of processors over the baseline. This is because at
+1024 processes on 64 nodes, MTTKRP is no more the dominating cost and hence there is no advantages due to
+dimension trees. Similarly, at this stage, the SVD used for Nestrov Non-negative Least Squares on Liavas dominates the 
+cost over MTTKRP that makes it even slower than the our proposed algorithm without dimension trees.
 
@@
 \begin{tikzpicture}
-\begin{axis}[legend style={draw=none, anchor=north,  cells={align=left}, nodes={scale=0.7}}]
+\begin{axis}[legend style={draw=none, anchor=north,  cells={align=left}, nodes={scale=0.7}}, legend pos=north west]
 	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};

### 2018-03-29 latex/sc18/experiments.tex
committing to sync experiments

--- 2018-03-29 latex/sc18/experiments.tex
@@
 \subsubsection{Dynamic Functional Connectivity (dFC)}
+We also consider dynamic functional connectivity datasets that are generated from fMRI images of human brains.
+Given a 4D fMRI data set of voxel measurements across multiple timesteps, voxels containing brain data are partitioned into a set of ZZ regions of interest (specified using domain-specific knowledge), and a single time-series signal is aggregated for each region of interest.
+For each time point and pair of regions,
 

### 2018-03-29 latex/sc18/plots.tex

@@
 		log basis x={2},
+		\ifliavas
+			xlabel=Num. Procs,
+		\else
 			xlabel=Nodes,
+		\fi
 		ylabel=Time (s),

### 2018-03-29 latex/sc18/experiments.tex
experiments section

@@
 
+In Figure \ref{fig:ksweepneuro} we compare k-sweeps for $R = 10,20,...,50$ for all 3 algorithms.
+Starting at $R=10$ we see the largest speed up of $2\times$ for dimensions trees over Liavas.
+This is due to a combination of the dimension tree performing less flops in the MTTKRPs and KRPs.
+However, as the rank increases this speed up diminishes to $1.6\times$.
+The loss of speed up is a result of the fact that, as we observed in \ref{fig:ksweep4DHSI}, the multi-TTV operations do not scale as well as the partial-MTTKRPs for increasing $R$.
+
 %\begin{figure}

### 2018-03-29 latex/sc18/experiments.tex
itemize to conclusion and types in experiments

--- 2018-03-29 latex/sc18/experiments.tex
@@
 
+In Figure \ref{fig:ksweepneuro} we compare k-sweeps for $R = 10,20,...,50$ for all 3 algorithms.
+Starting at $R=10$ we see the largest speed up of $2\times$ for dimension trees over Liavas.
+This is due to a combination of the dimension tree performing less flops in the MTTKRPs and KRPs.
+However, as the rank increases this speed up diminishes to $1.6\times$.
+The loss of speed up is a result of the fact that, as we observed in Figure \ref{fig:ksweep4DHSI}, the multi-TTV operations do not scale as well as the partial-MTTKRPs for increasing $R$.
+
 %\begin{figure}

### 2018-03-29 latex/sc18/experiments.tex
several updates mostly to experiments section

--- 2018-03-29 latex/sc18/experiments.tex
@@
 For comparison with previous work \cite{LK+17b}, we consider the same 3D hyperspectral imaging dataset called ``Souto\_wood\_pile'' \cite{FAN16}. 
+NNCP is often used on HSI data sets for classification and blind source separation of materials with differing spectral signatures.
 The hyperspectral datacube has dimensions $1024 \times 1344 \times 33$ and represents a set of 33 grayscale images of size 1344 $\times$ 1024 pixels sampled at wavelengths 400, 410, $\dots$, 720 nm, with each pixel value representing spectral radiance in $W m^{-2} sr^{-1} nm^{-1}$.
@@
 We also consider dynamic functional connectivity datasets that are generated from fMRI images of human brains.
-Given a 4D fMRI data set of voxel measurements across multiple timesteps, voxels containing brain data are partitioned into a set of ZZ regions of interest (specified using domain-specific knowledge), and a single time-series signal is aggregated for each region of interest.
-For each time point and pair of regions,
+Given a 4D fMRI data set of voxel measurements across multiple timesteps, voxels containing brain data are partitioned into a set of regions of interest (specified using domain-specific knowledge), and a single time-series signal is aggregated for each region of interest.
+Then, an instantaneous correlation is computed for each time point and pair of regions, and this process is repeated for a number of subjects.
+Computing a CP decomposition of this data helps to discover patterns of brain connectivity among different regions and also differentiate among individuals.
+For our representative dFC data set, we consider 246 brain regions, which yields 30{,}012 unique pairs of regions, 1200 times steps, and 500 subjects, or a tensor of dimension $30{,}012\times 1200 \times 500$ \cite{VEU+12,THBGW17}.
 
 \subsubsection{Synthetic}
-For the synthetic datasets, our open source code supports (a) low-rank tensor (b) uniform random and (c) positive shifted normal distribution of $\mathfrak{N}(3,1)$ -- that is change the mean such that all the random numbers are positive.
+Our synthetic data sets are constructed from a CP model with an exact low rank with no added noise.
+In this case we can confirm that the residual error of our algorithm with a random start converges to zero.
+For the purposes of benchmarking, we run a fixed number of iterations of the BCD algorithm rather than using a convergence check.
+
+%For the synthetic datasets, our open source code supports (a) low-rank tensor (b) uniform random and (c) positive shifted normal distribution of $\mathfrak{N}(3,1)$ -- that is change the mean such that all the random numbers are positive.
 We considered three different synthetic matrices for different cases.
 For baseline comparison with Liavas \cite{LK+17b}, we considered a three mode
-uniform of size 1024x1024x1024 on processor grids $2^k \times 2^k \times 2^k$ for $k \in {0,1,2,3}$. We used the uniform five 
-mode synthetic tensor with dimension $64\times 64\times 64\times 64\times 64$ on processor grids 
-$1\times1\times1\times1\times1$, $2\times1\times1\times1\times1$, $\dots$, $2\times2\times2\times2\times2$ for strong scaling 
-experiments.  In the case of weak scaling of four mode synthetic tensors with (D) and without (N) the use of dimension trees.  The 
-tensor and processor grid dimensions are $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$.
+%uniform of size 1024x1024x1024 on processor grids $2^k \times 2^k \times 2^k$ for $k \in {0,1,2,3}$. We used the uniform five 
+%mode synthetic tensor with dimension $64\times 64\times 64\times 64\times 64$ on processor grids 
+%$1\times1\times1\times1\times1$, $2\times1\times1\times1\times1$, $\dots$, $2\times2\times2\times2\times2$ for strong scaling 
+%experiments.  In the case of weak scaling of four mode synthetic tensors with (D) and without (N) the use of dimension trees.  The 
+%tensor and processor grid dimensions are $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$.
 In all the cases the dimensions were considered such that synthetic tensors can be accommodated even on single node with 64GB for scale up plots.
@@
 \subsection{Machine Details}
-The entire experimentation was performed on Eos -- a super computer in Oak Ridge Leadership Computing Facility (OLCF). 
-Eos is a 736-node Cray® XC30? cluster with a total of 47.104TB of memory. Its processor is the Intel® Xeon® E5-2670. It features 
-16 I/O service nodes and 2 external login nodes.
-Its compute nodes are organized in blades.
-Each blade contains 4 nodes and every node has 2 sockets with 8 physical cores and 64GB memory.
-Even though the machine support Intel's hyper-threading (HT), we restricted the total number of OpenMP threads to match the number of physical cores as HT offers minimal improvement for
- BLAS and LAPACK operations. In total, the Eos compute partition contains 11,776 traditional processor cores (23,552 logical cores 
- with HT Technology enabled) and some of the experiments peaked to 8,192 out of 11,776 physical cores that is approximately 
- 70\% of the peak capacity.
+The entire experimentation was performed on Eos, a supercomputer at the Oak Ridge Leadership Computing Facility. 
+Eos is a 736-node Cray XC30 cluster of Intel Xeon E5-2670 processors with a total of 47.104TB of memory. 
+Its compute nodes are organized in blades where each blade contains 4 nodes, and every node has 2 sockets with 8 physical cores and 64GB memory. 
+The machine support Intel's hyper-threading (HT), but we restricted it because HT offers minimal improvement for BLAS and LAPACK operations. 
+In total, the Eos compute partition contains 11,776 traditional processor cores and our experiments used up to 4,096 cores (35\% of the machine).
+\grey{add compiler and blas details}
+
+\subsection{Comparison Implementation}
+The implementation proposed by Liavas et al.
+\cite{LK+17b} is the only publicly available distributed-memory software (of which we are aware) for computing the CP decomposition of dense tensors, with or without constraints.
+We use the acronym NbAO-NTF for Nesterov-based Alternating Optimization Nonnegative Tensor Factorization to refer to their code.
+
+It is based on the same parallel algorithm as our implementation, though it is limited to 3D tensors.
+The code uses MPI collectives for communication and Eigen as an interface to BLAS and LAPACK.
+We compiled the code linked to OpenBLAS (the same BLAS implementation used by our code) but we were unable to run multithreaded BLAS with their code.
+For fair comparison, we use a flat MPI configuration (one MPI process per core) on all comparisons between the two implementations.
+
+We also point out a difference between the Nesterov-based algorithm and the BPP algorithm for solving the NLS subproblems.
+The Nesterov-based algorithm attempts an acceleration step using a linear combination of the current and proposed future step; however, it re-computes the residual error before deciding whether or not to accept or reject the acceleration step.
+This residual error cannot always be computed cheaply, using the technique described in \cref{sec:error}, and it contributes significantly (approximately 25\%) to the overall run time.
+Because the BPP algorithm does not require this extra computation, and studying convergence behavior of the different NLS algorithms is beyond the scope of this work, we remove the time spent in the acceleration step of NbAO-NTF in all our comparisons.
 
 \subsection{Strong Scaling}
-
-\begin{itemize}
-	\item 3D HSI (3 algs)
-	\item 3D synthetic (3 algs)
-	\item 5D synthetic (2 algs)
-\end{itemize}
 

### 2018-03-29 latex/sc18/conclusion.tex

\section{Conclusion} \label{sec:conclusion}
\begin{itemize}
	\item We present new parallel software for distributed memory NNCP.
	\item We demonstrate consistent speed up against existing state of the art up to $2.2\times$.
	\item We extend our algorithms and software to a general number of modes.
	\item We present analysis of our parallel algorithms showing good strong and weak parallel scaling up to thousands of cores (hundreds of nodes)
\end{itemize}

### 2018-03-29 latex/sc18/prelims.tex
nncp changes in introduction and prelims

@@
 While there are multiple optimization methods to compute NNCP, we will focus on a class of methods that use Block Coordinate Descent (BCD), which is also known as the nonlinear Gauss-Seidel method \cite{Bertsekas1999}.
-In BCD, the variables are partitioned into blocks, and each variable block is cyclically updated with all other blocks fixed.
+In BCD, the variables are partitioned into blocks, and each variable block is cyclically updated to optimality with all other blocks fixed.
 For details on the convergence properties and comparisons of BCD methods for nonnegative matrix and tensor decomposition problems, see \cite{KHP2014}.
-We consider BCD methods for NNCP that choose the factor matrices as the blocks, which is also often referred to as Alternating Least Squares.
-In this case, every subproblem is a linear nonnegative least squares problem.
+We consider BCD methods for NNCP that choose the entire factor matrices as the blocks, which is also often referred to as Alternating Least Squares.
+In this case, every subproblem is a linear nonnegative least squares problem. Formally, the following problem 
+is solved iteratively for $n = 1 \cdots N$.
+
+\SplitN{\label{eqn:bcdnncp}}{
+\HH^{(n)} \leftarrow \argmin_{\HH \geq 0} & \| \Mn{S}{n} \HH^T - \Mn{M}{n}^T \|_F^2
+}
+
+where, $\Mn{S}{n} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$.
 

### 2018-03-29 latex/sc18/experiments.tex

@@
 
-\subsection{Comparison Implementation}
+\subsection{Comparison Implementations}
 The implementation proposed by Liavas et al.
@@
 
+Our proposed algorithm uses dimension trees, but we also benchmark our implementation without that optimization to highlight its importance.
+We use an existing implementation to perform the individual MTTKRPs \cite{HBJT18} with this approach.
+
 \subsection{Strong Scaling}
 
-The strong scaling on 3D synthetic tensor with baseline is shown in Figure \ref{fig:strongsynthetic3D}. 
-We can observe from the figure that all the three algorithms scale linearly as the problem is more
-compute bound in the case of dense tensors. The computation is dominated by $O(R \prod I_k / P_k)$
-and the communication scales as $P^{1/N}$. As evident from the figure, this communication cost doesn't 
-show up even for $P$ in the order of few hundreds. Our proposed algorithm with dimension trees is 1.5x faster than the baseline 
-{\em Liavas}\cite{LK+17b}. 
-The dimension trees are faster over classical MTTKRP, in both the three and five mode synthetic tensors. We are witnessing
-2.5x speed up with dimension trees over higher modes as in Figure \ref{fig:strongsynthetic5D}. 
-The advantages of the dimension tree are more pronounced in higher
-mode tensors because of more reuse over partial MTTKRPs and avoiding unnecessary KRP computation that is more memory bound.  Also, for tensors with equal dimensions over all the
-modes, dimension trees has some advantage with favorable shapes for matrix multiplication with BLAS DGEMM.
+We perform two strong scaling experiments to compare performance with NbAO-NTF.
+The experiments use a cubical synthetic tensor and the HSI image used in \cite{LK+17b}, both of which are 3D.
 
-The Figure \ref{fig:stronghsi3D} shows the strong scaling on the three modes HSI real world data. We can observe
-that our proposed algorithm with dimension trees is almost 2x faster than baseline Liavas. The proposed algorithm
-with classical MTTKRP was performing better on higher number of processors over the baseline. This is because at
-1024 processes on 64 nodes, MTTKRP is no more the dominating cost and hence there is no advantages due to
-dimension trees. Similarly, at this stage, the SVD used for Nestrov Non-negative Least Squares on Liavas dominates the 
-cost over MTTKRP that makes it even slower than the our proposed algorithm without dimension trees.
+The performance on the synthetic tensor is shown in \cref{fig:strongsynthetic3D}. 
+We can observe from the figure that all the three algorithms scale nearly linearly as the problem remains compute bound. 
+Recall from \label{sec:analysis} that the computation scales linearly with $1/P$ while the communication scales with $1/P^{1/N}=1/P^{1/3}$. 
+As is evident from the figure, the communication cost does not degrade performance even for thousands of cores. 
+Our proposed algorithm with dimension trees is 35\% faster than NbAO-NTF at 512 cores (with similar relative difference for other core counts).
+This performance improvement is due in large part to the 50\% reduction in arithmetic operations provided by the dimension tree optimization.
+There is little difference in performance between our implementation without dimension trees and NbAO-NTF.
+
+\Cref{fig:stronghsi3D} shows the strong scaling on the HSI data. 
+In this case, our proposed algorithm with dimension trees is over $2\times$ faster than NbAO-NTF, but part of this speedup is due to differences in the NLS update algorithms.
+For the low core count, the dimension tree provides a 60\% speedup compared to the MTTKRP time in NbAO-NTF.
+At the high core counts for this experiment, the local MTTKRP is no longer the dominating cost.
+%The proposed algorithm with classical MTTKRP was performing better on higher number of processors over the baseline. 
+%Similarly, at this stage, the SVD used for Nestrov Non-negative Least Squares on Liavas dominates the cost over MTTKRP that makes it even slower than the our proposed algorithm without dimension trees.
+
+In \cref{fig:strongsynthetic5D}, we benchmark performance for a 5D tensor, so we can no longer compare against NbAO-NTF.
+We see a $13-16\times$ speed up using a dimension tree over not using one for this problem.
+As predicted, the dimension tree optimization saves relatively more arithmetic for higher-order tensors.
+However, the reduction in leading order flop cost is only $2.5\times$ for $N=5$; the rest of the speedup comes from avoiding  unnecessary KRP computation that is memory bound.  
+That is, although the flop count of KRP computation is lower order, it still contributes to the run time because it is inefficient.
+Also, for tensors with balanced dimensions, the dimension tree approach has some advantage with favorable shapes for BLAS's DGEMM.
 
@@
 \begin{tikzpicture}
+\renewcommand{\datafile}{data/str_3D_HSI.dat}
+\renewcommand{\numiterations}{10}
+\liavastrue
+\strongscalingplot
+\end{tikzpicture}
+\caption{Strong scaling of 3D HSI real world data with dimension 1024 x1344 x 33 on processor grids of $1 \times k \times k$ for $k \in {1, 2, 4, 8, 16, 32}$. The rank is fixed at 32.}
+\label{fig:stronghsi3D}
+\end{figure}
+
+\begin{figure}
+\begin{tikzpicture}
 \renewcommand{\datafile}{data/str_5D_syn.dat}
@@
 \label{fig:strongsynthetic5D}
-\end{figure}
-
-\begin{figure}
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/str_3D_HSI.dat}
-\renewcommand{\numiterations}{10}
-\liavastrue
-\strongscalingplot
-\end{tikzpicture}
-\caption{Strong scaling of 3D HSI real world data with dimension 1024 x1344 x 33 on processor grids of $1 \times k \times k$ for $k \in {1, 2, 4, 8, 16, 32}$. The rank is fixed at 32.}
-\label{fig:stronghsi3D}
 \end{figure}

### 2018-03-29 latex/sc18/prelims.tex
modified prelims

@@
 \times I_{N}$ (and $I=\prod I_n$) into the columns of a matrix.
-The Matricized-Tensor Times Khatri-Rao Product or MTTKRP will be central to this work and takes the form $\Mn{M}{n} = \Mz{A}{n}( \Mn{H}{N} \Khat \cdots \Khat \Mn{H}{n+1} \Khat \Mn{H}{n-1} \Khat  \cdots \Khat \Mn{H}{1})$.
+The Matricized-Tensor Times Khatri-Rao Product or MTTKRP will be central to this work and takes the form $\Mn{M}{n} = \Mz{A}{n} \Mn{S}{n}$, where $\Mn{S}{n}= \Mn{H}{N} \Khat \cdots \Khat \Mn{H}{n+1} \Khat \Mn{H}{n-1} \Khat  \cdots \Khat \Mn{H}{1}$.
 
@@
 
-where, $\Mn{S}{n} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$.
+%where, $\Mn{S}{n} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$.
 

### 2018-03-29 latex/sc18/introduction.tex
included cp diagram
Added some numbers to the intro

@@
 The CP decomposition is a low-rank approximation of a multi-dimensional array, or tensor, which generalizes matrix approximations like the truncated singular value decomposition.
+As in Figure \ref{fig:cpdecomposition}, CP decomposition approximates the given input matrix as sum of $k$ rank-1 tensors.
 CP is often used for finding hidden patterns, or latent factors, within tensor data, particularly when the goal is to interpret the factors, and it is popular within the signal processing, machine learning, and scientific computing communities.
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=3in, height=1.5in]{fig/cpdecomposition.pdf}
+\caption{CP Decomposition}
+\label{fig:cpdecomposition}
+\end{center}
+\end{figure}
 
@@
 We focus in this paper on dense tensors (when nearly all of the tensor entries are nonzero) and on constraining solutions to have nonnegative entries, which is useful when the tensor data itself is nonnegative.
+Formally, NNCP can be defined as
+\SplitN{\label{eqn:nncp}}{
+\min_{\HH^{(i)} \geq 0} & \| \TA - \CP \|_F^2 \\ 
+\min_{\HH^{(i)} \geq 0} & \| \TA - \sum_{r=1}^R \Mn{H}{1}(:,r) \circ \cdots \circ \Mn{H}{N}(:,r) \|_F^2
+}
+where $\|\M{X}\|_F=(\sum_{ij} x_{ij}^2)^{1/2}$ is the Frobenius norm.
+
 For example, in imaging and microscopy applications, tensor values often correspond to intensities, and NNCP can be used to cluster and analyze the data in a lower-dimensional space \cite{JC+16}.
@@
 		\item we demonstrate a performance improvement of up to \grey{$2.2\times$} over the existing state-of-the-art parallel software on 3D tensors,
-		\item we demonstrate efficient parallel scaling of up to \grey{$527\times$} on \grey{$512$} cores.
+		\item we demonstrate efficient parallel scaling of up to \grey{$253\times$} on \grey{$512$} cores.
 \end{itemize}

### 2018-03-29 latex/sc18/experiments.tex
finished pass through weak scaling

@@
 The machine support Intel's hyper-threading (HT), but we restricted it because HT offers minimal improvement for BLAS and LAPACK operations. 
-In total, the Eos compute partition contains 11,776 traditional processor cores and our experiments used up to 4,096 cores (35\% of the machine).
+In total, the Eos compute partition contains 11,776 traditional processor cores and our experiments used up to 8,192 cores (70\% of the machine).
 \grey{add compiler and blas details}
@@
 
-The performance on the synthetic tensor is shown in \cref{fig:strongsynthetic3D}. 
+The performance on the cubical synthetic tensor is shown in \cref{fig:strongsynthetic3D}. 
 We can observe from the figure that all the three algorithms scale nearly linearly as the problem remains compute bound. 
@@
 There is little difference in performance between our implementation without dimension trees and NbAO-NTF.
-
-\Cref{fig:stronghsi3D} shows the strong scaling on the HSI data. 
-In this case, our proposed algorithm with dimension trees is over $2\times$ faster than NbAO-NTF, but part of this speedup is due to differences in the NLS update algorithms.
-For the low core count, the dimension tree provides a 60\% speedup compared to the MTTKRP time in NbAO-NTF.
-At the high core counts for this experiment, the local MTTKRP is no longer the dominating cost.
-%The proposed algorithm with classical MTTKRP was performing better on higher number of processors over the baseline. 
-%Similarly, at this stage, the SVD used for Nestrov Non-negative Least Squares on Liavas dominates the cost over MTTKRP that makes it even slower than the our proposed algorithm without dimension trees.
-
-In \cref{fig:strongsynthetic5D}, we benchmark performance for a 5D tensor, so we can no longer compare against NbAO-NTF.
-We see a $13-16\times$ speed up using a dimension tree over not using one for this problem.
-As predicted, the dimension tree optimization saves relatively more arithmetic for higher-order tensors.
-However, the reduction in leading order flop cost is only $2.5\times$ for $N=5$; the rest of the speedup comes from avoiding  unnecessary KRP computation that is memory bound.  
-That is, although the flop count of KRP computation is lower order, it still contributes to the run time because it is inefficient.
-Also, for tensors with balanced dimensions, the dimension tree approach has some advantage with favorable shapes for BLAS's DGEMM.
 
@@
 
+\Cref{fig:stronghsi3D} shows the strong scaling on the HSI data. 
+In this case, our proposed algorithm with dimension trees is over $2\times$ faster than NbAO-NTF, but part of this speedup is due to differences in the NLS update algorithms.
+For the low core count, the dimension tree provides a 60\% speedup compared to the MTTKRP time in NbAO-NTF.
+At the high core counts for this experiment, the local MTTKRP is no longer the dominating cost.
+%The proposed algorithm with classical MTTKRP was performing better on higher number of processors over the baseline. 
+%Similarly, at this stage, the SVD used for Nestrov Non-negative Least Squares on Liavas dominates the cost over MTTKRP that makes it even slower than the our proposed algorithm without dimension trees.
+
 \begin{figure}
@@
 \end{figure}
+
+In \cref{fig:strongsynthetic5D}, we benchmark performance for a 5D cubical tensor with each dimension set to 64.
+Because the tensor is 5D, we can no longer compare against NbAO-NTF.
+We see a $13-16\times$ speed up using a dimension tree over not using one for this problem.
+As predicted, the dimension tree optimization saves relatively more arithmetic for higher-order tensors.
+However, the reduction in leading order flop cost is only $2.5\times$ for $N=5$; the rest of the speedup comes from avoiding  unnecessary KRP computation that is memory bound.  
+That is, although the flop count of KRP computation is lower order, it still contributes to the run time because it is inefficient.
+Also, for tensors with balanced dimensions, the dimension tree approach has some advantage with favorable shapes for BLAS's DGEMM.
 
@@
 
-\subsection{Weak Scaling}
+\subsection{Weak Scaling Time Breakdown}
 
-\begin{itemize}
-	\item 4D synthetic (2 algs)
-	\item 3D neuroscience, increasing one mode? (2 algs)
-\end{itemize}
-
-In the case of weakscaling, we wanted to understand the time it takes to solve bigger problems by increasing the number of nodes with the increase in data size by keeping the problem size (workload) assigned to each processing element constant.
-We conducted this experiment by using synthetic tensor and processor grid of dimensions $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$, and the rank is fixed at 32.
-For eg., in the case of 3 x 3 x 3 x 3 processor grid with 81 nodes, we ran the proposed algorithm with and without dimension trees on a four mode tensor of size 384 x 384 x 384 x 384.
-The results of the breakdown plot is shown in Figure \ref{fig:weaksynthetic4D}.
-Since, the NTF on dense tensors is compute intensive, the total time of the weak scaling is staying fixed for both the case of with and without dimension trees.
-The entire running is completely dominated by the MTTKRP computation.
-As mentioned above, from the Figure   \ref{fig:weaksynthetic4D}, we can observe that the KRP cost of dimension trees is insignificant and it offers 2.5x speed up over classical MTTKRP.
+We also perform a weak scaling experiment to understand the time it takes to solve bigger problems with more processors.
+In this experiment, we use a synthetic 4D tensor and keep the amount of tensor data assigned to each processor constant, with tensor and processor grid of dimensions $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$ and the rank fixed at 32.
+%For eg., in the case of 3 x 3 x 3 x 3 processor grid with 81 nodes, we ran the proposed algorithm with and without dimension trees on a four mode tensor of size 384 x 384 x 384 x 384. 
+The results of the breakdown plot is shown in \cref{fig:weaksynthetic4D}. 
+In this case, the algorithm is compute bound with and without the use of the dimension tree, so the total time of the weak scaling remains fixed for both cases. 
+Using a dimension tree, the time is completely dominated by the MTTKRP computation.  
+Without using a dimension tree, we observe that the KRP is expensive and yields a $2.5\times$ slower total run time even in the 4D case.
 

### 2018-03-29 latex/sc18/experiments.tex
Note on scaling up to 512 nodes
software environment details

--- 2018-03-29 latex/sc18/experiments.tex
@@
 In total, the Eos compute partition contains 11,776 traditional processor cores and our experiments used up to 4,096 cores (35\% of the machine).
+
+Our objective of the implementation is using open source software as much as possible 
+to promote reproducibility and reuse of our code.
+Unlike Liavas \cite{LK+17b} that uses Eigen matrix library, we use Armadillo \cite{sanderson2010} for matrix representation
+and operations.  
+In Armadillo, the elements of the dense matrix are stored in column major order.
+For dense BLAS and LAPACK operations, we linked Armadillo with the default LAPACK/BLAS wrappers from Cray. 
+For compiler, we use GNU C++ Compiler (g++ (GCC) 6.3.0) and MPI library is from Cray.  We could also 
+compile and run the code in RHEA the commodity cluster from OLCF with entire open source libraries such as OpenBLAS and OpenMPI.
+
 \grey{add compiler and blas details}

### 2018-03-29 latex/sc18/experiments.tex

--- 2018-03-29 latex/sc18/experiments.tex
@@
 The machine support Intel's hyper-threading (HT), but we restricted it because HT offers minimal improvement for BLAS and LAPACK operations. 
-In total, the Eos compute partition contains 11,776 traditional processor cores and our experiments used up to 8,192 cores (70\% of the machine).
-\grey{add compiler and blas details}
+In total, the Eos compute partition contains 11,776 traditional processor cores and our experiments used up to 4,096 cores (35\% of the machine).
+
+Our objective of the implementation is using open source software as much as possible 
+to promote reproducibility and reuse of our code.
+Unlike Liavas \cite{LK+17b} that uses Eigen matrix library, we use Armadillo \cite{sanderson2010} for matrix representation
+and operations.  
+In Armadillo, the elements of the dense matrix are stored in column major order.
+For dense BLAS and LAPACK operations, we linked Armadillo with the default LAPACK/BLAS wrappers from Cray. 
+For compiler, we use GNU C++ Compiler (g++ (GCC) 6.3.0) and MPI library is from Cray.  We could also 
+compile and run the code in RHEA the commodity cluster from OLCF with entire open source libraries such as OpenBLAS and OpenMPI.
 
@@
 %Similarly, at this stage, the SVD used for Nestrov Non-negative Least Squares on Liavas dominates the cost over MTTKRP that makes it even slower than the our proposed algorithm without dimension trees.
+
+Finally, we ran on 512 nodes, utilizing the 16 threads per node via parallel BLAS and OMP, for a $1024\times1024\times1024$ synthetic tensor and processor grid $8\times8\times8$.
+Here we observe a $253\times$ speedup which is the result of each node having less than 20 megabytes of data.
 

### 2018-03-29 latex/sc18/experiments.tex
working in ksweep section

--- 2018-03-29 latex/sc18/experiments.tex
@@
 
-\begin{itemize}
-	\item 4D HSI (2 algs)
-	\item 3D neuroscience (2 algs)
-\end{itemize}
+One of the challenges of the CP (and NNCP) decomposition in practice is the choice of decomposition rank.
+The most common technique is to compute multiple CP decompositions for various ranks.
+As the rank $k$ increases, the approximation error  $\|\TA - \T{M}\|$ decreases with the better approximation power of more parameters. 
+However, the benefit of increasing $k$ eventually diminishes if the data can be well approximated with a CP model.
+%that an increase in $k$ in lower values, say from 5 to 10, will have significant improvement in relative error over increase in higher 
+%values, like 100 to 105. 
+%Hence, it is common practice in the community to sweep $k$, to obtain better approximation error within the 
+%manageable computation. 
+Towards this end, we experiment with various values of $k$ to observe the relative increase in running time for two real-world data sets. 
+The results of the experiments on real world four mode HSI tensor is presented in Figure \ref{fig:ksweep4DHSI}. We can observe 
+that, the multi-TTV({\em mTTV}) scales linearly with the increasing $k$,  where as the partial MTTKRP ({\em PM}) is scaling super 
+linearly. This is because, with higher $k$, the BLAS $DGEMM$ gets a bigger chunk of work that it can optimize better.  The local 
+compute $NLS$ is increasing as expected nearly $O(k^3)$ and the $ALL-GATHER$ required for the gram matrices over $k^2$ 
+elements, becomes significant on higher $k$'s.
+
+In Figure \ref{fig:ksweepneuro} we compare k-sweeps for $R = 10,20,...,50$ for all 3 algorithms.
+Starting at $R=10$ we see the largest speed up of $2\times$ for dimension trees over Liavas.
+This is due to a combination of the dimension tree performing less flops in the MTTKRPs and KRPs.
+However, as the rank increases this speed up diminishes to $1.6\times$.
+The loss of speed up is a result of the fact that, as we observed in Figure \ref{fig:ksweep4DHSI}, the multi-TTV operations do not scale as well as the partial-MTTKRPs for increasing $R$.
 
@@
 \end{figure}
-
-According to Kim, He and Park \cite{KHP2014}, in the case of dense tensors, as the low rank $k$ increases, the approximation 
-error  $\|\TA - \T{M}\|$ decreases because it is obtaining bigger low rank factors. However, there is a law of diminishing advantages, 
-that an increase in $k$ in lower values, say from 5 to 10, will have significant improvement in relative error over increase in higher 
-values, like 100 to 105. Hence, it is common practice in the community to sweep $k$, to obtain better approximation error within the 
-manageable computation. Towards this end, we wanted to understand the increase in running time over increasing the low rank $k
-$. The results of the experiments on real world four mode HSI tensor is presented in Figure \ref{fig:ksweep4DHSI}. We can observe 
-that, the multi-TTV({\em mTTV}) scales linearly with the increasing $k$,  where as the partial MTTKRP ({\em PM}) is scaling super 
-linearly. This is because, with higher $k$, the BLAS $DGEMM$ gets a bigger chunk of work that it can optimize better.  The local 
-compute $NLS$ is increasing as expected nearly $O(k^3)$ and the $ALL-GATHER$ required for the gram matrices over $k^2$ 
-elements, becomes significant on higher $k$'s.
-
-In Figure \ref{fig:ksweepneuro} we compare k-sweeps for $R = 10,20,...,50$ for all 3 algorithms.
-Starting at $R=10$ we see the largest speed up of $2\times$ for dimension trees over Liavas.
-This is due to a combination of the dimension tree performing less flops in the MTTKRPs and KRPs.
-However, as the rank increases this speed up diminishes to $1.6\times$.
-The loss of speed up is a result of the fact that, as we observed in Figure \ref{fig:ksweep4DHSI}, the multi-TTV operations do not scale as well as the partial-MTTKRPs for increasing $R$.
 

### 2018-03-29 latex/sc18/experiments.tex
eigen details from liavas code

--- 2018-03-29 latex/sc18/experiments.tex
@@
 to promote reproducibility and reuse of our code.
-Unlike Liavas \cite{LK+17b} that uses Eigen matrix library, we use Armadillo \cite{sanderson2010} for matrix representation
+%Unlike Liavas \cite{LK+17b} that uses Eigen matrix library \cite{eigenweb}, 
+We use Armadillo \cite{sanderson2010} for matrix representation
 and operations.  
@@
 It is based on the same parallel algorithm as our implementation, though it is limited to 3D tensors.
-The code uses MPI collectives for communication and Eigen as an interface to BLAS and LAPACK.
-We compiled the code linked to OpenBLAS (the same BLAS implementation used by our code) but we were unable to run multithreaded BLAS with their code.
+The code uses MPI collectives for communication and Eigen \cite{eigenweb} as an interface to BLAS and LAPACK.
+We compiled the code linked to BLAS/LAPACK wrappers from Cray  (the same BLAS implementation used by our code) but we were unable to run multithreaded BLAS with their code.
 For fair comparison, we use a flat MPI configuration (one MPI process per core) on all comparisons between the two implementations.

### 2018-03-29 latex/sc18/experiments.tex
minor changes to experiments section

--- 2018-03-29 latex/sc18/experiments.tex
@@
 As predicted, the dimension tree optimization saves relatively more arithmetic for higher-order tensors.
-However, the reduction in leading order flop cost is only $2.5\times$ for $N=5$; the rest of the speedup comes from avoiding  unnecessary KRP computation that is memory bound.  
+However, the reduction in leading order flop cost is only $2.5\times$ for $N=5$; the rest of the speedup comes from more efficient DGEMM performance and avoiding memory-bound KRP computation.  
 That is, although the flop count of KRP computation is lower order, it still contributes to the run time because it is inefficient.
-Also, for tensors with balanced dimensions, the dimension tree approach has some advantage with favorable shapes for BLAS's DGEMM.
+Also, for tensors with balanced dimensions, the dimension tree approach yields more favorable shapes for DGEMM.
 

### 2018-03-29 latex/sc18/experiments.tex
finished pass through performance section

--- 2018-03-29 latex/sc18/experiments.tex
@@
 For compiler, we use GNU C++ Compiler (g++ (GCC) 6.3.0) and MPI library is from Cray.  We could also 
-compile and run the code in RHEA the commodity cluster from OLCF with entire open source libraries such as OpenBLAS and OpenMPI.
+compile and run the code in Rhea the commodity cluster from OLCF with entire open source libraries such as OpenBLAS and OpenMPI.
 
@@
 As predicted, the dimension tree optimization saves relatively more arithmetic for higher-order tensors.
-However, the reduction in leading order flop cost is only $2.5\times$ for $N=5$; the rest of the speedup comes from more efficient DGEMM performance and avoiding memory-bound KRP computation.  
+However, the reduction in leading order flop cost is only $2.5\times$ for $N=5$; the rest of the speedup comes from avoiding  unnecessary KRP computation that is memory bound.  
 That is, although the flop count of KRP computation is lower order, it still contributes to the run time because it is inefficient.
-Also, for tensors with balanced dimensions, the dimension tree approach yields more favorable shapes for DGEMM.
+Also, for tensors with balanced dimensions, the dimension tree approach has some advantage with favorable shapes for BLAS's DGEMM.
 
@@
 Towards this end, we experiment with various values of $k$ to observe the relative increase in running time for two real-world data sets.
-The results of the experiments on real world four mode HSI tensor is presented in Figure \ref{fig:ksweep4DHSI}. We can observe 
-that, the multi-TTV({\em mTTV}) scales linearly with the increasing $k$,  where as the partial MTTKRP ({\em PM}) is scaling super 
-linearly. This is because, with higher $k$, the BLAS $DGEMM$ gets a bigger chunk of work that it can optimize better.  The local 
-compute $NLS$ is increasing as expected nearly $O(k^3)$ and the $ALL-GATHER$ required for the gram matrices over $k^2$ 
-elements, becomes significant on higher $k$'s.
 
-In Figure \ref{fig:ksweepneuro} we compare k-sweeps for $R = 10,20,...,50$ for all 3 algorithms.
-Starting at $R=10$ we see the largest speed up of $2\times$ for dimension trees over Liavas.
-This is due to a combination of the dimension tree performing less flops in the MTTKRPs and KRPs.
+\Cref{fig:ksweep4DHSI} shows the time breakdown of our implementation using a dimension tree on the 4D HSI dataset for $R=\{10,\dots,50\}$. 
+We observe an overall time increase with increased $k$, but each part of the computation scales slightly differently.
+The multi-TTV computation ({\em mTTV}) scales linearly with the increasing $k$,  whereas the partial MTTKRP ({\em PM}) is scaling super-linearly. 
+This is because mTTV is cast as matrix-vector products (DGEMV) and PM is cast as matrix-matrix products (DGEMM).
+As $k$ increases from 10 to 50, DGEMM performance improves but DGEMV performance is constant.  
+The local NLS time is increasing with $O(k^3)$ as expected and the All-Reduce required for the Gram matrices scales with $O(k^2)$, becoming a significant cost for larger $k$.
+
+In \cref{fig:ksweepneuro} we compare performance for various ranks $R$ across all 3 algorithms, again used flat MPI. 
+Starting at $R=10$ we see the largest speed up of $2\times$ for our implementation with a dimension tree over NbAO-NTF. 
+This is due to a combination of the dimension tree performing fewer flops in the MTTKRPs and KRPs. 
 However, as the rank increases this speed up diminishes to $1.6\times$. 
-The loss of speed up is a result of the fact that, as we observed in Figure \ref{fig:ksweep4DHSI}, the multi-TTV operations do not scale as well as the partial-MTTKRPs for increasing $R$.
+The loss of speed up is a result of the fact that, as we observed in \cref{fig:ksweep4DHSI}, the multi-TTV operations do not scale as well as the partial-MTTKRPs for increasing $R$.
+Again, the performance of our implementation without the dimension tree optimization is comparable to NbAO-NTF.
 
@@
 	ylabel={Time (s)}, 
-	xlabel={Rank $k$},
+	xlabel={Rank $R$},
 	y label style={yshift=-.5cm},
@@
 	\addplot+ [discard if not={alg}{L},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
-	\legend{DimTree,NoDimTree,Liavas}
+	\legend{DimTree,NoDimTree,NbAO-NTF}
 \end{axis}
 \end{tikzpicture}
-\caption{$k$ sweep plot for Neuroscience dataset}
+\caption{Overall running time for dFC dataset with dimensions $30{,}012\times 1200 \times 500$ on 1440 cores arranged in a $2\times 6\times 120$ processor grid and varying choices of rank $R$.}
 \label{fig:ksweepneuro}

### 2018-03-29 latex/sc18/algorithm.tex
Changed algorithm1 and algorithm2 to use R instead of k to denoate the rank

@@
 \subsection{Dimension Trees}
+\label{sec:dimtrees}
 
@@
 \subsection{Relative Error Computation}
+\label{sec:error}
 
@@
 \subsubsection{Data Distribution}
+\label{sec:datadist}
 
@@
 \begin{algorithm}
-\caption{$\CP = \text{Par-NNCP}(\TA,k)$}
+\caption{$\CP = \text{Par-NNCP}(\TA,R)$}
 \label{alg:Par-NNCP-short}
 \begin{algorithmic}[1]
-\Require $\TA$ is an $I_1\times \cdots \times I_N$ tensor distributed across a $P_1\times \cdots \times P_N$ grid of $P$ processors, so that $\TA_{\V{p}}$ is $(I_1/P_1)\times \cdots \times (I_N/P_N)$ and is owned by processor $\V{p}=(p_1,\dots,p_N)$, $k$ is rank of approximation
+\Require $\TA$ is an $I_1\times \cdots \times I_N$ tensor distributed across a $P_1\times \cdots \times P_N$ grid of $P$ processors, so that $\TA_{\V{p}}$ is $(I_1/P_1)\times \cdots \times (I_N/P_N)$ and is owned by processor $\V{p}=(p_1,\dots,p_N)$, $R$ is rank of approximation
 \For{$n=2$ to $N$}
-	\State Initialize $\Mn{H}{n}_{\V{p}}$ of dimensions $(I_n/P)\times k$ 
+	\State Initialize $\Mn{H}{n}_{\V{p}}$ of dimensions $(I_n/P)\times R$ 
 	\State $\M[\overline]{G} = \text{Local-SYRK}(\Mn{H}{n}_{\V{p}})$
@@
 \Ensure $\TA \approx \CP$
-\Ensure Local matrices: $\Mn{H}{n}_{\V{p}}$ is $(I_n/P)\times k$ and owned by processor $\V{p}=(p_1,\dots,p_N)$, for $1\leq n \leq N$, $\V{\lambda}$ stored redundantly on every processor
+\Ensure Local matrices: $\Mn{H}{n}_{\V{p}}$ is $(I_n/P)\times R$ and owned by processor $\V{p}=(p_1,\dots,p_N)$, for $1\leq n \leq N$, $\V{\lambda}$ stored redundantly on every processor
 \end{algorithmic}

### 2018-03-29 latex/sc18/appendix.tex
finished draft of algorithm section

@@
 \begin{algorithm}
-\caption{$(\CPl,\epsilon) = \text{Par-NNCP}(\TA,k)$}
+\caption{$(\CPl,\epsilon) = \text{Par-NNCP}(\TA,R)$}
 \label{alg:Par-NNCP-long}
 \begin{algorithmic}[1]
-\Require $\TA$ is an $I_1\times \cdots \times I_N$ tensor distributed across a $P_1\times \cdots \times P_N$ grid of $P$ processors, so that $\TA_{\V{p}}$ is $(I_1/P_1)\times \cdots \times (I_N/P_N)$ and is owned by processor $\V{p}=(p_1,\dots,p_N)$, $k$ is rank of approximation
+\Require $\TA$ is an $I_1\times \cdots \times I_N$ tensor distributed across a $P_1\times \cdots \times P_N$ grid of $P$ processors, so that $\TA_{\V{p}}$ is $(I_1/P_1)\times \cdots \times (I_N/P_N)$ and is owned by processor $\V{p}=(p_1,\dots,p_N)$, $R$ is rank of approximation
 \State \Comment{Initialize data}
@@
 \For{$n=2$ to $N$}
-	\State Initialize $\Mn{H}{n}_{\V{p}}$ of dimensions $(I_n/P)\times k$ 
+	\State Initialize $\Mn{H}{n}_{\V{p}}$ of dimensions $(I_n/P)\times R$ 
 	\State $\M[\overline]{G} = \text{Local-SYRK}(\Mn{H}{n}_{\V{p}})$
@@
 	\State \Comment{Organize data for later modes}
-	\State $\M[\overline]{G} = \text{Local-SYRK}(\Mn{H}{n}_{\V{p}})$
+	\State $\M[\overline]{G} = {\Mn{H}{n}_{\V{p}}}^\Tra\Mn{H}{n}_{\V{p}}$
 	\State $\Mn{G}{n} = \text{All-Reduce}(\M[\overline]{G},\textsc{All-Procs})$
@@
 \Ensure $\|\TA - \dsquare{\V{\lambda}; \Mn{H}{1},\dots,\Mn{H}{N}}\| /\|\TA\| = \epsilon$
-\Ensure Local matrices: $\Mn{H}{n}_{\V{p}}$ is $(I_n/P)\times k$ and owned by processor $\V{p}=(p_1,\dots,p_N)$, for $1\leq n \leq N$, $\V{\lambda}$ stored redundantly on every processor
+\Ensure Local matrices: $\Mn{H}{n}_{\V{p}}$ is $(I_n/P)\times R$ and owned by processor $\V{p}=(p_1,\dots,p_N)$, for $1\leq n \leq N$, $\V{\lambda}$ stored redundantly on every processor
 \end{algorithmic}

### 2018-03-29 latex/sc18/prelims.tex
software environment details
finished pass through performance section
changed algorithm in prelims to use R instead of k
finished conclusion

@@
 In this work, we are interested in NNCP models that are good approximations to $\TA$ in the least squares sense.
-That is, we seek to minimize $\|\TA-\CPl\|$, where the tensor norm is a generalization of the matrix Frobenius or vector 2-norm, the square root of the sum of squares of the entries.
+That is, we seek 
+$$\min_{\HH^{(i)} \geq 0}  \| \TA - \CPl \|,$$
+where the tensor norm is a generalization of the matrix Frobenius or vector 2-norm, the square root of the sum of squares of the entries.
 
@@
 
-\SplitN{\label{eqn:bcdnncp}}{
-\HH^{(n)} \leftarrow \argmin_{\HH \geq 0} & \| \Mn{S}{n} \HH^T - \Mn{M}{n}^T \|_F^2
-}
+%\SplitN{\label{eqn:bcdnncp}}{
+$$\HH^{(n)} \leftarrow \argmin_{\HH \geq 0}  \| \Mn{S}{n} \HH^T - {\Mn{M}{n}}^T \|_F^2$$
+%}
 

### 2018-03-29 latex/sc18/experiments.tex

@@
 This is because mTTV is cast as matrix-vector products (DGEMV) and PM is cast as matrix-matrix products (DGEMM).
-As $k$ increases from 10 to 50, DGEMM performance improves but DGEMV performance is constant.  
-The local NLS time is increasing with $O(k^3)$ as expected and the All-Reduce required for the Gram matrices scales with $O(k^2)$, becoming a significant cost for larger $k$.
+As $R$ increases from 10 to 50, DGEMM performance improves but DGEMV performance is constant.  
+The local NLS time is increasing with $O(R^3)$ as expected and the All-Reduce required for the Gram matrices scales with $O(R^2)$, becoming a significant cost for larger $R$.
 
@@
 \begin{tikzpicture}
-\begin{axis}[legend style={draw=none, anchor=north,  cells={align=left}, nodes={scale=0.7}}, legend pos=north west]
+\begin{axis}[legend style={draw=none, anchor=north,  cells={align=left}, nodes={scale=0.7}}, legend pos=north west, xlabel=Low Rank $R$, ylabel=Time(s), ylabel near ticks]
 	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};

### 2018-03-29 latex/sc18/conclusion.tex

@@
+% !TEX root = paper.tex
+
 \section{Conclusion} \label{sec:conclusion}
-\begin{itemize}
-	\item We present new parallel software for distributed memory NNCP.
-	\item We demonstrate consistent speed up against existing state of the art up to $2.2\times$.
-	\item We extend our algorithms and software to a general number of modes.
-	\item We present analysis of our parallel algorithms showing good strong and weak parallel scaling up to thousands of cores (hundreds of nodes)
-\end{itemize}
+
+In this work, we present a new implementation for distributed-memory NNCP that will be made publicly available.
+The algorithm is general enough to handle any number of modes in the data tensor and can be adapted to use any NLS algorithm within the context of BCD (ALS).
+We use a dimension tree optimization to avoid unnecessary recomputation within the bottleneck local MTTKRP computation, and we use an efficient parallelization that minimizes communication cost.
+Our performance results show the ability to scale well to high processor counts, and we show favorable performance in comparison to state-of-the-art software for 3D tensors.
+
+In particular, the performance results demonstrate that computing NNCP for dense tensors involves heavy computation relative to the sizes of the computed factor matrices.
+By avoiding the communication of tensor entries and communicating only the factor matrices, the parallel algorithm is nearly always compute bound.
+This observation is supported by the theoretical analysis: although the communication does not scale as well with $P$, the total amount of data depends on a sum of tensor dimensions rather the product of the tensor dimensions, which determines the total amount of computation.
+
+We can also conclude from the performance results that the dimension tree optimization is the key to performance improvement over the state-of-the-art approaches.
+For 3D tensors, we observe a benefit larger than the theoretical 50\% reduction in computation, and for larger numbers of modes, the improvement is only magnified.
+Besides the reduction in flops, the dimension tree approach enjoys better DGEMM performance and avoids memory-bound KRP computations.
+
+Given an efficient parallel algorithm for NNCP, it is now possible to compare the convergence properties of various NLS algorithms on large-scale tensor data that is too large to fit on single servers or would require too much computational time to analyze interactively.
+We hope to explore this in future work.

### 2018-03-29 latex/sc18/introduction.tex
finished conclusion
changed formal definition nncp
modified nncp formal definition

@@
 \SplitN{\label{eqn:nncp}}{
-\min_{\HH^{(i)} \geq 0} & \| \TA - \CP \|_F^2 \\ 
-\min_{\HH^{(i)} \geq 0} & \| \TA - \sum_{r=1}^R \Mn{H}{1}(:,r) \circ \cdots \circ \Mn{H}{N}(:,r) \|_F^2
+ \min_{\HH^{(i)} \geq 0} & \| \TA - \sum_{r=1}^R \Mn{H}{1}(:,r) \circ \cdots \circ \Mn{H}{N}(:,r) \|^2
 }
-where $\|\M{X}\|_F=(\sum_{ij} x_{ij}^2)^{1/2}$ is the Frobenius norm.
-
+where $\Mn{H}{1}(:,i) \circ \cdots \circ \Mn{H}{N}(:,i)$ is the outer product of the $i^{th}$ vector from all the $N$ factors that yields a rank one tensor $\T{M}$ and $\sum_{r=1}^R \Mn{H}{1}(:,r) \circ \cdots \circ \Mn{H}{N}(:,r)$ results in a sum of $R$ rank one tensors that will be of the same dimension as the input tensor $\TA$.
 For example, in imaging and microscopy applications, tensor values often correspond to intensities, and NNCP can be used to cluster and analyze the data in a lower-dimensional space \cite{JC+16}.

### 2018-03-29 latex/sc18/abstract.tex
dont know what i did

@@
 Our approach is also communication efficient, using a data distribution and parallel algorithm across a multidimensional processor grid that can be tuned to minimize communication.
-We benchmark our software on synthetic as well as hyperspectral image and neuroscience dynamic functional connectivity data, demonstrating that our algorithm scales well to \grey{100s} of nodes (up to \grey{9600} cores) and is faster and more general than the currently available parallel software.
+We benchmark our software on synthetic as well as hyperspectral image and neuroscience dynamic functional connectivity data, demonstrating that our algorithm scales well to 100s of nodes (up to 8192 cores) and is faster and more general than the currently available parallel software.
 

### 2018-03-29 latex/sc18/introduction.tex
dont know what i did

--- 2018-03-29 latex/sc18/introduction.tex
@@
 We focus in this paper on dense tensors (when nearly all of the tensor entries are nonzero) and on constraining solutions to have nonnegative entries, which is useful when the tensor data itself is nonnegative.
-Formally, NNCP can be defined as
-\SplitN{\label{eqn:nncp}}{
-\min_{\HH^{(i)} \geq 0} & \| \TA - \CP \|_F^2 \\ 
-\min_{\HH^{(i)} \geq 0} & \| \TA - \sum_{r=1}^R \Mn{H}{1}(:,r) \circ \cdots \circ \Mn{H}{N}(:,r) \|_F^2
-}
-where $\|\M{X}\|_F=(\sum_{ij} x_{ij}^2)^{1/2}$ is the Frobenius norm.
 
@@
 		\item our algorithm is designed for general $N$-way tensors,
-		\item we demonstrate a performance improvement of up to \grey{$2.2\times$} over the existing state-of-the-art parallel software on 3D tensors,
-		\item we demonstrate efficient parallel scaling of up to \grey{$253\times$} on \grey{$512$} cores.
+		\item we demonstrate a performance improvement of up to $2.2\times$ over the existing state-of-the-art parallel software on 3D tensors,
+		\item we demonstrate efficient parallel scaling of up to $253\times$ on $512$ cores.
 \end{itemize}

### 2018-04-02 latex/sc18/experiments.tex
dont know what i did

@@
 \renewcommand{\datafile}{data/str_3D_syn.dat}
-\renewcommand{\numiterations}{42}
+\renewcommand{\numiterations}{1}
 \liavastrue
@@
 The most common technique is to compute multiple CP decompositions for various ranks.
-As the rank $k$ increases, the approximation error  $\|\TA - \T{M}\|$ decreases with the better approximation power of more parameters. 
-However, the benefit of increasing $k$ eventually diminishes if the data can be well approximated with a CP model.
-%that an increase in $k$ in lower values, say from 5 to 10, will have significant improvement in relative error over increase in higher 
+As the rank $R$ increases, the approximation error  $\|\TA - \T{M}\|$ decreases with the better approximation power of more parameters. 
+However, the benefit of increasing $R$ eventually diminishes if the data can be well approximated with a CP model.
+%that an increase in $R$ in lower values, say from 5 to 10, will have significant improvement in relative error over increase in higher 
 %values, like 100 to 105. 
@@
 %manageable computation. 
-Towards this end, we experiment with various values of $k$ to observe the relative increase in running time for two real-world data sets.
+Towards this end, we experiment with various values of $R$ to observe the relative increase in running time for two real-world data sets.
 
 \Cref{fig:ksweep4DHSI} shows the time breakdown of our implementation using a dimension tree on the 4D HSI dataset for $R=\{10,\dots,50\}$. 
-We observe an overall time increase with increased $k$, but each part of the computation scales slightly differently.
-The multi-TTV computation ({\em mTTV}) scales linearly with the increasing $k$,  whereas the partial MTTKRP ({\em PM}) is scaling super-linearly. 
+We observe an overall time increase with increased $R$, but each part of the computation scales slightly differently.
+The multi-TTV computation ({\em mTTV}) scales linearly with the increasing $R$,  whereas the partial MTTKRP ({\em PM}) is scaling super-linearly. 
 This is because mTTV is cast as matrix-vector products (DGEMV) and PM is cast as matrix-matrix products (DGEMM).

### 2018-04-02 latex/sc18/plots.tex

@@
 	bar width=8pt,
-	width=9cm, height=3.85cm,
+	%width=9cm, height=3.85cm,
+	%width=\textwidth,
 	ylabel={Time (s)}, 
@@
 		\ifliavas
-			xlabel=Num. Procs,
+			xlabel=Cores,
 		\else
@@
 		\addplot+ [discard if not={alg}{L},thick,mark options={solid}] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
-		\legend{DimTree,NoDimTree,Liavas}
+		\legend{DimTree,NoDimTree,NbAO-NTF \cite{LK+17b}}
 	\else

### 2018-04-02 latex/slides/talk.tex

@@
 
+\begin{frame}
+\frametitle{Comparison against existing code}
+
+From paper: \\
+\footnotesize
+A.
+P.
+Liavas, G.
+Kostoulas, G.
+Lourakis, K.
+Huang, and N.
+D.
+Sidiropoulos, "Nesterov-based Alternating Optimization for Nonnegative Tensor Factorization: Algorithm and Parallel Implementations", \emph{IEEE Transactions on Signal Processing}, vol.
+66, no.
+4, Feb.
+2018.
+
+\vfill
+
+Code available: \url{http://www.telecom.tuc.gr/Greek/Liavas/publications/MPI_EIGEN_Nesterov_NTF.zip}
+
+\vfill
+
+Not a whole lot of detail in paper about implementation, but data distribution and communication pattern the same as ours
+\begin{itemize}
+	\item limited to 3D tensors
+	\item uses Eigen for local dense linear algebra
+	\item no use of dimension trees
+	\item inefficient error computation
+\end{itemize}
+
+\end{frame}
+
 \input{../sc18/plots}
@@
 	\scriptsize
-		\item computation is dominated by $O(R (\prod I_k) / P)$
+		\item computation is dominated by $O(R \prod I_k / P_k)$
 		\item communication is dominated by $O(R \sum I_k / P_k)$
@@
 		\item limited to 3D tensors
-		\item permutes the tensor entries for each matricization 
+		\item stores each matricization separately (each in col-major)
 		\item does not use dimension trees or efficient error computation

### 2018-04-02 common/distutils.h
aoadmm added

@@ -21,6 +21,8 @@ enum iodistributions {ONED_ROW, ONED_COL, ONED_DOUBLE, TWOD};
 #define NUMROWPROCS this->m_mpicomm.pr()
 #define NUMCOLPROCS this->m_mpicomm.pc()
 
+// #define MPITIC tic()
+// #define MPITOC toc()
 #define MPITIC mpitic()
 #define MPITOC mpitoc()
 // #define MPITIC mpitic(MPI_RANK)

### 2018-04-02 distnmf/distio.hpp

@@ -89,6 +89,7 @@ class DistIO {
 #endif
     }
 
+#ifndef BUILD_SPARSE
     /* normalization */
     void normalize(normtype i_normtype) {
         ROWVEC globalnormA = arma::zeros<ROWVEC>(m_A.n_cols);
@@ -96,12 +97,12 @@ class DistIO {
         MATTYPE normmat = arma::zeros <MATTYPE>(m_A.n_rows, m_A.n_cols);
         switch (m_distio) {
         case ONED_ROW:
-            if (i_normtype == L2) {
+            if (i_normtype == L2NORM) {
                 normc = arma::sum(arma::square(m_A));
                 MPI_Allreduce(normc.memptr(), globalnormA.memptr(), m_A.n_cols,
                               MPI_DOUBLE, MPI_SUM, MPI_COMM_WORLD);
 
-            } else if (i_normtype == MAX) {
+            } else if (i_normtype == MAXNORM) {
                 normc = arma::max(m_A);
                 MPI_Allreduce(normc.memptr(), globalnormA.memptr(), m_A.n_cols,
                               MPI_DOUBLE, MPI_MAX, MPI_COMM_WORLD);
@@ -110,18 +111,18 @@ class DistIO {
             break;
         case ONED_COL:
         case ONED_DOUBLE:
-            if (i_normtype == L2) {
+            if (i_normtype == L2NORM) {
                 globalnormA = arma::sum(arma::square(m_Arows));
-            } else if (i_normtype == MAX) {
+            } else if (i_normtype == MAXNORM) {
                 globalnormA = arma::max(arma::square(m_Arows));
             }
             break;
         case TWOD:
-            if (i_normtype == L2) {
+            if (i_normtype == L2NORM) {
                 normc = arma::sum(arma::square(m_A));
                 MPI_Allreduce(normc.memptr(), globalnormA.memptr(), m_A.n_cols,
                               MPI_DOUBLE, MPI_SUM, this->m_mpicomm.commSubs()[1]);
-            } else if (i_normtype == MAX) {
+            } else if (i_normtype == MAXNORM) {
                 normc = arma::max(m_A);
                 MPI_Allreduce(normc.memptr(), globalnormA.memptr(), m_A.n_cols,
                               MPI_DOUBLE, MPI_SUM, this->m_mpicomm.commSubs()[1]);
@@ -134,6 +135,7 @@ class DistIO {
         normmat = arma::repmat(globalnormA, m_A.n_rows, 1);
[…]

### 2018-04-02 distnmf/distnmf.cpp

@@ -265,9 +265,9 @@ class DistNMFDriver {
 #endif  // ifndef USE_PACOSS
     }
     void parseCommandLine() {
-        PLANC::ParseCommandLine pc(this->m_argc, this->m_argv);
+        planc::ParseCommandLine pc(this->m_argc, this->m_argv);
         pc.parseplancopts();
-        this->m_nmfalgo       = pc.nmfalgo();
+        this->m_nmfalgo       = pc.lucalgo();
         this->m_k             = pc.lowrankk();
         this->m_Afile_name    = pc.input_file_name();
         this->m_pr            = pc.pr();

### 2018-04-02 nmf/nmf.cpp

@@ -5,6 +5,7 @@
 #include "utils.hpp"
 #include "hals.hpp"
 #include "mu.hpp"
+#include "aoadmm.hpp"
 #include <string>
 #include <omp.h>
 
@@ -73,10 +74,10 @@ void incrementalGraph(std::string AfileName, std::string WfileName) {
 }
 #endif
 void parseCommandLineandCallNMF(int argc, char *argv[]) {
-    PLANC::ParseCommandLine pc(argc, argv);
+    planc::ParseCommandLine pc(argc, argv);
     pc.parseplancopts();
-    pc.printconfig();
-    switch (pc.nmfalgo()) {
+    pc.printConfig();
+    switch (pc.lucalgo()) {
     case MU:
 #ifdef BUILD_SPARSE
         NMFDriver<MUNMF<SP_MAT > >(pc.lowrankk(), pc.globalm(), pc.globaln(),
@@ -121,6 +122,23 @@ void parseCommandLineandCallNMF(int argc, char *argv[]) {
                                  pc.output_file_name() + "_h",
                                  pc.iterations());
 #endif
+    case AOADMM:
+#ifdef BUILD_SPARSE
+        // NMFDriver<AOADMMNMF<SP_MAT > >(lowRank, m, n, AfileName, WInitfileName,
+        //                                HInitfileName, WfileName, HfileName, numIt);
+        NMFDriver<AOADMMNMF<SP_MAT > >(pc.lowrankk(), pc.globalm(), pc.globaln(),
+                                       pc.input_file_name(),
+                                       pc.output_file_name() + "_w",
+                                       pc.output_file_name() + "_h", 
+                                       pc.iterations());
+
+#else
+        NMFDriver<AOADMMNMF<MAT > >(pc.lowrankk(), pc.globalm(), pc.globaln(),
+                                       pc.input_file_name(),
+                                       pc.output_file_name() + "_w",
+                                       pc.output_file_name() + "_h", 
+                                       pc.iterations());
+#endif        
         break;
     }
 }

### 2018-04-02 ntf/auntf.hpp

@@ -42,7 +42,7 @@ class AUNTF {
     const algotype m_updalgo;
     LUC *m_luc;
     planc::Tensor *lowranktensor;
-    KobyDimensionTree *kdt;
+    // KobyDimensionTree *kdt;
     bool m_enable_dim_tree;
 
   public:

### 2018-04-02 openmp/nmf.cpp

@@ -1,222 +1,0 @@
-/* Copyright 2016 Ramakrishnan Kannan */
-
-#include "bppnmf.hpp"
-#include "utils.hpp"
-#include "hals.hpp"
-#include "mu.hpp"
-#include "aoadmm.hpp"
-#include <string>
-#include <omp.h>
-
-template <class NMFTYPE>
-void NMFDriver(int k, UWORD m, UWORD n, std::string AfileName,
-               std::string WinitFileName, std::string HinitFileName,
-               std::string WfileName, std::string HfileName, int numIt) {
-#ifdef BUILD_SPARSE
-    SP_MAT A;
-    UWORD nnz;
-#else
-    MAT A;
-#endif
-    double t1, t2;
-    if (!AfileName.empty()) {
-#ifdef BUILD_SPARSE
-        A.load(AfileName, arma::coord_ascii);
-        INFO << "Successfully loaded the input matrix" << std::endl;
-#else
-        tic();
-        A.load(AfileName);
-        t2 = toc();
-        INFO << "Successfully loaded dense input matrix. A=" << PRINTMATINFO(A)
-             << " took=" << t2 << std::endl;
-        m = A.n_rows;
-        n = A.n_cols;
-#endif
-    } else {
-#ifdef BUILD_SPARSE
-        A = arma::sprandu<SP_MAT>(m, n, 0.001);
-#else
-        A = arma::randu<MAT>(m, n);
-#endif
-        INFO << "generated random matrix A=" << PRINTMATINFO(A) << std::endl;
-    }
-    MAT W, H;
-    if (!WinitFileName.empty()) {
-        INFO << "Winitfilename = " << WinitFileName << std::endl;
-        W.load(WinitFileName);
-        INFO << "Loaded W." << PRINTMATINFO(W) << std::endl;
-    }
-    if (!HinitFileName.empty()) {
-        INFO << "HInitfilename=" << HinitFileName << std::endl;
-        H.load(HinitFileName);
-        INFO << "Loaded H." << PRINTMATINFO(H) << std::endl;
-    }
-    if (!WinitFileName.empty()) {
-        NMFTYPE nmfAlgorithm(A, W, H);
-        nmfAlgorithm.num_iterations(numIt);
-        INFO << "completed constructor" << std::endl;
-        tic();
-        nmfAlgorithm.computeNMF();
-        t2 = toc();
-        INFO << "time taken:" << t2 << std::endl;
[…]

### 2018-04-02 common/ncpfactors.hpp
sequential ntf was broken after sc18 changes. fixed it now

@@ -289,7 +289,17 @@ class NCPFactors {
             this->ncp_factors[mode].col(i) /= m_lambda(i);
         }
     }
-
+        /*
+    * this is for reinitializing random numbers across different
+    * processors.
+    */
+    void randu(const int i_seed) {
+        arma::arma_rng::set_seed(i_seed);
+        for (int i = 0; i < this->m_modes; i++) {
+            ncp_factors[i].randu();
+        }
+    }
+#ifdef MPI_DISTNTF
     // Distribution normalization of factor matrices
     // To be used for MPI code only
     void distributed_normalize() {
@@ -320,16 +330,8 @@ class NCPFactors {
             m_lambda(j) = global_colnorm;
         }
     }
-    /*
-    * this is for reinitializing random numbers across different
-    * processors.
-    */
-    void randu(const int i_seed) {
-        arma::arma_rng::set_seed(i_seed);
-        for (int i = 0; i < this->m_modes; i++) {
-            ncp_factors[i].randu();
-        }
-    }
+#endif
+
 }; // NCPFactors
 }  // planc
 

### 2018-04-02 common/tensor.hpp

@@ -8,6 +8,7 @@
 #include <type_traits>
 #include <ios>
 #include <fstream>
+#include <cblas.h>
 
 namespace planc {
 /*

### 2018-04-02 ntf/auntf.hpp

@@ -3,6 +3,10 @@
 #ifndef NTF_AUNTF_HPP_
 #define NTF_AUNTF_HPP_
 
+#define MPITIC tic();
+#define MPITOC toc();
+
+
 #include <armadillo>
 #include <cblas.h>
 #include "tensor.hpp"
@@ -42,7 +46,7 @@ class AUNTF {
     const algotype m_updalgo;
     LUC *m_luc;
     planc::Tensor *lowranktensor;
-    // KobyDimensionTree *kdt;
+    KobyDimensionTree *kdt;
     bool m_enable_dim_tree;
 
   public:
@@ -105,7 +109,11 @@ class AUNTF {
                      << ncp_krp[j] << std::endl;
 #endif
                 if (this->m_enable_dim_tree) {
-                    kdt->in_order_reuse_MTTKRP(j, ncp_mttkrp_t[j].memptr(), false);
+                    double multittv_time = 0;
+                    double mttkrp_time = 0;
+                    kdt->in_order_reuse_MTTKRP(j, ncp_mttkrp_t[j].memptr(),
+                                               false, multittv_time,
+                                               mttkrp_time);
                 } else {
                     m_input_tensor.mttkrp(j, ncp_krp[j], &ncp_mttkrp_t[j]);
                 }

### 2018-04-02 sc17/introduction.tex
got all the files checked in into sc17

\section{Introduction}

%\begin{itemize}
%\item Introduce NMF
%\item Sparse NMF
%\item Sparse-dense matrix multiplication as bottleneck
%\item Hypergraph partition
%\item A partition that is not only good for multiplication but at the same time cannot hurt NNLS computation
%\item Introduce Contributions
%\end{itemize}

Non-negative Matrix Factorization (NMF) is the problem of finding two low rank factors $\WW\in \Rnplus{m\times k}$ and $\HH\in \Rnplus{k \times n}$ for a given input matrix  $\AA\in \Rnplus{m\times n}$, such that $\AA \approx \WW \HH$.
Here, $\Rnplus{m\times n}$ denotes the set of $m \times n$ matrices with non-negative real values.
Formally, the NMF problem \cite{seung2001algorithms} can be defined as \SplitN{\label{eqn:original NMF}}{
\min_{\WW \geq 0,\HH \geq 0} & \|\AA-\WW\HH\|_F + \phi (\WW) + \psi (\HH),
}
where $\|\M{X}\|_F=(\sum_{ij} x_{ij}^2)^{1/2}$ is the Frobenius norm. The functions $\phi (\WW)$ and $\psi (\HH)$ are called regularization functions that avoids the model from over fitting on the data and are chosen depending on the characteristics of the data. For this paper, we are considering $\phi(\WW) = \alpha \|\WW\|_F^2$ called as $\ell_2$ regularizer and $\psi(\HH) = \beta \sum_{i=1}^n \| \hh_{i} \|_1^2$ called as $\ell_1$ regularizer for addressing the inherent sparsity in the input matrix. The lack of $\ell_1$ regularization on the matrix $\HH$ can result in numerical instability.

NMF is widely used in data mining and machine learning as a dimension reduction and factor analysis method.
It is a natural fit for many real world problems as the non-negativity is inherent in many representations of real-world data and
the resulting low rank factors are expected to have a natural interpretation. The applications of NMF range from text mining \cite{pauca2004text},  computer vision \cite{hoyer2004non}, and bioinformatics \cite{kim2007sparse} to blind source separation  \cite{cichocki2009nonnegative}, unsupervised clustering \cite{kuang2012symmetric,kuang2013symnmf}  and many other areas.
In most real-world applications $m$ and $n$ can be on the order of millions or more while $k$ being much smaller in the order of tens to thousands.

In the recent years, there has been an explosion in the size of collected data particularly with the widespread use of internet and social media.
Such collected data has big dimensions~(from millions to billions) and is extremely sparse, which is either implied by the nature of the data~(e.g., most web content are not visited by users), or because capturing every possible data point is impractical.
Analyzing such sparse big datasets require enormous computational power as well as efficient algorithms that can leverage the sparsity of the data.
NMF is an effective tool for analyzing such data in many applications, yet there is a lack of high performance algorithms and software that can compute it efficiently.
Our goal is in this paper is to fill in this gap by providing scalable parallel algorithms that employ efficient communication schemes and can scale to thousands of processes.
Thereby, we aim to enable analyzing data in such big data applications.

% How NMF is computed, and in sparse what is important.
We would like to highlight that ``Non-negative'' Matrix Factorization is NOT matrix factorization in collaborative filtering for recommender systems for which many implementations exist.
The collaborative filtering problem is different than NMF -- the focus of our paper -- because it interprets the ``{\em zero}'' entries of the input matrix as missing data, while the NMF problem is defined for a completely known input matrix and does not handle missing values.
This leads to different optimization problems, algorithms, and computational steps.
Specifically, computing NMF involves three main types of operations; multiplication of $\Am$ with a factor matrix $\Wm$ or $\Hm$, computing the Gram matrices $\Wm^T \Wm$ and $\Hm \Hm^T$, and solving a non-linear least squares~(NNLS) problem using these two resulting dense matrices to update factor matrices.
To the best of our knowledge, the only high performance software available for parallelizing this computation is \mpifaun~\cite{KBP16MPIFAUN} which operates both on sparse and dense input matrices.
For parallelism, it employs a fixed 2D uniform partitioning on $\Am$, and partitions factor matrices $\Wm$ and $\Hm$ conformally with this 2D partition.
Each process, works on its local matrix block of $\Am$, and communicates the rows and columns of $\Wm$ and $\Hm$ corresponding to its process row and column in the 2D topology.
This communication is optimal in the dense case; i.e., no process receives a data element not used in its local computations.
In the sparse case, however, the actual communication requirements of processes depend on the sparsity of $\Am$ and its partitioning, and is in general significantly less than the dense case.
Therefore, employing the same all-to-all communication scheme in the sparse case brings about a major redundancy in the communication cost.
Also, \mpifaun similarly uses only fixed 2D uniform partitioning on sparse matrices; this creates significant imbalance both in computation and communication as typically the data is not homogenously distributed in $\Am$.
In overall, the lack of an efficient sparse communication scheme that allows flexibility in partitioning hinders the scalability of this approach.

% Sketch of our software and advantages over MPIFAUN
%In this paper, we provide an efficient parallel NMF algorithm called \distspnmf for sparse matrices, which leverages the sparsity of the matrix, and employs an optimal point-to-point communication scheme that prevents any redundant communication.
%Moreover, our algorithm is flexible in the sense that it can work on any partition of $\Am$ as well as factor matrices $\Wm$ and $\Hm$.
%We use this partitioning potential in two ways.
%First, we employ a 2D cartesion processor topology to eliminate the communication cost in one dimension, and bound the maximum number of messages sent per process.
%Then, we investigate efficient 2D partitioning strategies to embed into this topology iin order to further reduce the communication cost and enhance scalability.

%% List contributions
%
%\todo{\oguz{revise this paragraph}}
%I
%In this paper, we are leveraging the \mpifaun algorithm \cite{KBP16MPIFAUN} which established the high performance framework for different NMF algorithms. NMF is a dimensionality reduction problem and there are various algorithms exists in the literature. The major steps in NMF algorithms are (a)matrix multiplication between the sparse input matrix $\AA$ with the low rank factors $\WW$ or $\HH$  (b) solving Non-negative Least Squares (NLS) subproblems and (c) computing the gram matrix. For the convenience of the readers, we are presenting the general NMF Algorithm \ref{alg:aunmf} based on \mpifaun. The different NMF algorithms such as Multiplicative Update \cite{seung2001algorithms},Hierarchical Alternating Least Square \cite{cichocki2009nonnegative} and Alternative Non-negative Least Squares using Block Principal Pivoting \cite{kim2013nonnegative} differ in the way they use the matrix multiplication and gram of the factor matrix to obtain the NLS factor. 
%
%For sparse NMF, the two popular approaches for scalability are distributing non-zeros \todo{\oguz{add a citation}} ($NNZ-PART$) and the sparse matrix itself\cite{KBP16MPIFAUN, KBP16} ($2D$).  For the latter, 2D distributions are effective for reducing the communication cost and the \mpifaun algorithm  \cite{KBP16MPIFAUN} claims communication minimization for dense matrices under some mild assumptions.  %Even though this can work for some sparse input matrix as well, there are some shortcomings in the existing framework and in Section \ref{sec:sparsenmf}, we explain our key extensions handle really large high dimensional sparse data.  
%Even though this can work for some sparse input matrix as well, there is a huge load imbalance which is typically alleviated by permuting the rows and columns of the matrix. 
%Also, the performance of $NNZ-PART$ can be more useful in practice over $2D$ partition. However, it is very difficult to design an effective non-zero distribution strategy for any general input sparse matrix that balances the load and reduces the communication. 
%
%We will discuss briefly these two issues of computation and communication in the case of sparse NMF. The $NNZ-PART$ of sparse NMF should take into consideration all the three operations - sparse-dense matrix multiplication, NLS and gram of the factor matrices. On very large number of processors when the problem size has become smaller, gram is as important as other operations. The $NNZ-PART$ and $2D$ employs completely different communication MPI techniques. While $2D$ partitions leverage the MPI collective communication which is typically bounded logarithm of the number of processors $p$, the $NNZ-PART$ uses point-to-point($P2P$) communication which is bounded by the number and the size of messages. The major disadvantage of $P2P$ communication is the latency which most of the time in practice dominates the communication cost. 
%
%Towards this end, we are proposing a novel Algorithm \distspnmffull(\distspnmf) that handles both the distribution strategies $NNZ-PART$ and $2D$. We are proposing two $NNZ-PART$ schemes (i) Checkerboard-PaTOH (\cpp)  and (ii) Checkerboard Random (\crp). These schemes consider multiple constraints such as (a) effective for the three operations in NMF algorithm (b) reducing the number of messages between nodes  as the communication within a computational node is insignificant and (c) \todo{\oguz{add additional constraints}}. It is important to notice that our algorithm is independent of the partition schemes employed for $NNZ-PART$ and it is not restricted to only the proposed $\cpp$ and $\crp$.

We summarize our main contributions as follows:
\begin{itemize}
\item We propose an efficient parallel NMF algorithm called \distspnmf for sparse matrices that employs a point-to-point communication scheme to leverage the sparsity of the input matrix, and eliminate any redundant communication existing in \mpifaun.
The algorithm is flexible to work with any partition of the input matrix $\Am$ as well as factor matrices $\Wm$ and $\Hm$.
\item We employ a 2D cartesion process topology to eliminate the communication cost of parallel NMF in one dimension, and bound the maximum number of messages sent per process in our MPI implementation.
\item We introduce effective partitioning strategies to further reduce the communication cost and enhance the parallel scalability of our algorithm.
\item We introduce regularization to NMF computations to prevent potential numerical instabilities using large matrices.
\item We compare our implementation with a state-of-the-art NMF library, and report scalability results up to 32678 processors on two different parallel computing platforms using real-world datasets.
\end{itemize}

### 2018-04-02 sc17/plots.tex
got all the files checked in into sc17

% macros for plotting
\newcommand{\datafile}{}
\newcommand{\numiterations}{30}
\newcommand{\minvalue}{1}
%\newcommand{\xticklabels}{\unp, \urp, \crp, \cpp}

\newif \ifylabel
\newif \iflegend
\newif \ifbgq

\newcommand{\setcolors}{
\pgfplotsset{cycle list={
	red, fill=red \\ 
	blue, fill=blue \\ 
	green, fill=green \\ 
	red, pattern=crosshatch, pattern color=red \\
	blue, pattern=crosshatch, pattern color=blue \\
	green, pattern=crosshatch, pattern color=green \\
}};
}
% set options for grouped bar plot
\newcommand{\plotoptions}{
	ybar stacked,
	reverse legend,
	bar width=8pt,
	width=2in, height=2in,
	\ifylabel
	     ylabel={Time (seconds)}, 
	     y label style={yshift=-.5cm},
	\fi
         ymin=0,
	%xticklabels/.expanded={\unp, \urp, \crp, \cnp}, 
	symbolic x coords={un, ur, cr, cp},
	xtick=data,
	xticklabel style={xshift=.15cm,rotate=45,anchor=east},
	%xlabel={Partitioning Schemes},
	xlabel style={yshift=-0.5cm},
	legend cell align=left,
	legend style={nodes={scale=0.7, transform shape}, draw=none,row sep=-0.1cm},
	%legend style={at={(1,.5)},anchor=west}
	legend pos=north east
}

\newcommand{\plotoptionsnew}{
	ybar stacked,
	reverse legend,
	bar width=8pt,
	width=2in, height=2in,
	\ifylabel
	     ylabel={Time (seconds)}, 
	     y label style={yshift=-.5cm},
	\fi
         ymin=0,
	symbolic x coords={cp-fl,cr-fl,cp-dl,cr-dl},
	xtick=data,
	xticklabel style={xshift=.15cm,rotate=45,anchor=east},	
	xlabel style={yshift=-0.5cm},
	legend cell align=left,
	legend style={nodes={scale=0.7, transform shape}, draw=none,row sep=-0.1cm},
	%legend style={at={(1,.5)},anchor=west}
	legend pos=north west
}

\newcommand{\speedupoptions}{
xlabel=Processors, ylabel=Speed-up,width=3.5in, height=2.0in, 
y label style={yshift=-.5cm}, 
legend pos=north west, legend style={draw=none,row sep=-0.1cm},	
xtick=data, 
\ifbgq
xticklabels={256, 512, 1024, 2048, 4096, 8192, 16384, 32768},
symbolic x coords={256, 512, 1024, 2048, 4096, 8192, 16384, 32768}
\else
xticklabels={16, 48, 96, 192, 384, 768, 1536, 3072},
symbolic x coords={16, 48, 96, 192, 384, 768, 1536, 3072}
\fi
}

% grouped bar plot command
\newcommand{\makeplot}{
\begin{axis}[\plotoptions, xticklabels/.expand once={\cpp, \crp, \unp, \urp}]
	\setcolors
	\addplot table[x=algo, y expr=(\thisrow{mm}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{gram}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
	%\addplot table[x=algo, y expr=(\thisrow{reducescatter}/(\minvalue*\numiterations))] {\datafile};
	\iflegend
	    %\legend{MM, \NLS, Gram, All-Gather, Reduce-Scatter};
	    \legend{MM, \NLS, Gram, Comm};
	\fi
\end{axis}
}

\newcommand{\makeplotnew}{
\begin{axis}[\plotoptionsnew, xticklabels/.expand once={\cpp,\crp,\cpp,\crp},
xlabel/.expand once={Flickr \quad\quad\quad Delicious} ]
	\setcolors
	\addplot table[x=algo, y expr=(\thisrow{mm}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{gram}/(\minvalue*\numiterations))] {\datafile};
	%\addplot table[x=algo, y expr=(\thisrow{err}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{comm}/(\minvalue*\numiterations))] {\datafile};
	\iflegend
	    %\legend{MM, \NLS, Gram, Err, Comm};
	    \legend{MM, \NLS, Gram, Comm};
	\fi
\end{axis}
}

% labels for bar groups (manually positioned)
\newcommand{\labels}{
\node [align=center,text width=3cm] at (1.25cm, -.95cm){\cp};
\node [align=center,text width=3cm] at (3cm, -0.95cm) {\cr};
\node [align=center,text width=3cm] at (4.6cm, -0.95cm){\un};
\node [align=center,text width=3cm] at (6.3cm, -0.95cm){\ur};
}

\newcommand{\speedup}{
\begin{axis}[\speedupoptions]
\addplot [red,very thick, mark=triangle] table [x={p}, y expr=(\minvalue/\thisrow{cr})] {\datafile};
\addplot [green,very thick, mark=x] table [x={p}, y expr=(\minvalue/\thisrow{cp})] {\datafile};
\ifbgq
\else
\addplot [very thick, mark=diamond] table [x={p}, y expr=(\minvalue/\thisrow{ur})] {\datafile};
\addplot [blue,very thick, mark=*] table [x={p}, y expr=(\minvalue/\thisrow{un})] {\datafile};
\fi
\legend{\crp, \cpp, \urp, \unp}
\end{axis}
}

\newcommand{\numkblocksweep}{

\begin{axis}[xlabel=Number of $k$ blocks, 
	ylabel=Sparse-Dense MM (in Secs),
	%xticklabels={1,2,3,4,6},
	y label style={yshift=-.25cm},
	width=3in, height=2.4in]
\addplot[red,very thick] table [x={numkblocks}, y expr=(\thisrow{mm}/(\minvalue*\numiterations))] {\datafile}; \label{plot_one}
\end{axis}
}

### 2018-04-02 latex/sc17/abstract.tex
moved under latex

\begin{abstract}
Non-negative matrix factorization (NMF) is the problem of finding two non-negative low rank factors $\WW$ and $\HH$ for a given input matrix $\AA$ such that $\AA \approx \WW \HH$.
NMF is a useful tool for many internet applications such as topic modeling in text mining and community detection in social networks.
In this paper, we focus on scaling distributed NMF to very large sparse datasets by employing effective algorithms, communications, and partitioning schemes that leverage the sparsity of the input matrix $\AA$.
Our method employs an optimal point-to-point communication scheme, and removes any redundant communication existing in the state of the art.
Unlike the state of the art, our method also permits arbitrary partitions of the sparse input matrix as well as factor matrices, which enables us to employ effective partitioning schemes for better scalability.
We experiment on a cluster using up to 3072 cores, and report significant better scalability with respect to existing methods owing to our efficient communication and partitioning strategies.
We also run our algorithms on an IBM BlueGene/Q supercomputer, and achieve scalability up to 32768 processors with a 32x speedup over the execution using 256 processors.

%We propose a fine-grain communication scheme that can utilize arbitrary partitions of the matrices. For a more scalable parallel NMF algorithm, we employ effective 2D and fine-grain hypergraph partitions. If hypergraph partitioning gets expensive for the purpose of the application, researchers can resort to the proposed uniform and random 2D partitioning NMF algorithms. But in the latter, memory growth of intermediate dense matrices is exorbitant -- a major bottleneck to run in lesser number of nodes. Toward this, we propose memory efficient blocking 2D matrix multiplication for sparse NMF. We compare the performance of 2D and the hypergraph partitions on real-world datasets. 
\end{abstract}

### 2018-04-02 latex/sc17/conclusion.tex
moved under latex

\section{conclusion}

In this paper, we proposed a novel algorithm parallel non-negative matrix factorization of sparse matrices in distributed memory environments.
To the best of our knowledge, this is the first high performance implementation that takes the sparsity of the input matrix into consideration in communication to reduce the communication cost, and employs effective partitionings to further enhance parallel scalability.
We introduce effective partitioning strategies for both the sparse input matrix $\Am$ as well as factor matrices $\Wm$ and $\Hm$ that helps us achieve good computational load balance while minimizing the communication costs.
With all the algorithmic contributions and an efficient parallel implementation, our method outperforms the state of the art by a significant margin, and gracefully scales up to 32768 cores on an IBM BlueGene/Q supercomputer.
Our immediate next steps for extending our work involve adding shared memory parallelism to obtain further speedup.

### 2018-04-02 latex/sc17/experiments.tex
moved under latex

\section{Experiments}
\label{sec:experiment}
\newcommand{\NLS}{LUC }

%In this section, we describe our implementation of \distspnmf\  and evaluate its performance.
%We identify two popular real world data sets to experiment with dimensions that span to tens of millions.  with \distspnmf\. 
%We compare the performance and exploring scaling behavior of \distspnmf\ against \mpifaun. 
%The \mpifaun code for conducting the experiments can be downloaded from \url{https://github.com/ramkikannan/nmflibrary}.

In this section, we compare our algorithm \distspnmf against \mpifaun, and compare its parallel performance on two big sparse matrices formed from real world datasets.
We analyze and compare the computation and communication timings of these algorithms on a smaller cluster, then test the scalability limits of our method on a large supercomputing environment.

\subsection{Experimental Setup}

\subsubsection{Datasets}\label{sec:datasets}

We use two datasets from Flickr.com and Delicious.com that involves images tagged with different labels by users.
The rows of the matrix correspond to different images, whereas the columns of the matrix represent different tags. The value of each nonzero $a_{i,j} \in \Am$ indicates the number of unique users that tagged the image $i$ with the tag $j$.
Flickr and Delicious matrices are of size 28Mx1.6M and 17Mx2.5M, and have 112M and 72M nonzero elements, respectively.
The current implementation of \mpifaun can only operate when $R$ and $C$ can divide $m$ and $n$, hence we trimmed the matrices slightly.

\subsubsection{Implementation Platform}
We conducted our experiments on two different parallel computing platforms.
The first platform is the ``Rhea'' cluster at the Oak Ridge Leadership Computing Facility (OLCF), which is a commodity-type Linux cluster with a total of 512 nodes and a 4X FDR Infiniband interconnect.
Each node contains dual-socket 8-core Intel Sandy Bridge-EP processors operating at 2GHz clock frequency and 128 GB of memory.
Each socket has a shared 20MB L3 cache, and each core has a private 256K L2 cache. 
There, we ran our experiments up to 3072 cores, which is the maximum allowed in the cluster.
The second platform is an IBM BlueGene/Q supercomputer consisting of 6 racks each having 16384 cores.
Each compute node has 16GB of memory and single socket 16-core PowerPC A2 processor at 1.6GHz clock frequency with 16KB of L1 cache per core, and 32MB shared L2 cache.
We ran both algorithms using 16 MPI ranks per node, and set $P_c = 16$ in all partitionings.

Our code for local matrix operations is developed using the matrix library Armadillo \cite{sanderson2010}. 
We use BLAS and LAPACK for dense matrix operations by linking Armadillo with Intel MKL, OpenBLAS~\cite{xianyi2015}, or any other BLAS and LAPACK implementation.
Both codes are compiled using the default GNU C++ Compiler (g++ (GCC) 5.3.0) and MPI library (Open MPI 1.8.4) on RHEA, and Clang compiler~(3.5.0) with IBM MPI library on BlueGene/Q.

\subsubsection{Algorithms}

In our experiments, we considered the following algorithms and partitionings:
\begin{itemize}
	\item \unp: \mpifaun Algorithm \cite{KBP16,KBP16MPIFAUN} with uniform natural  partition~(\unp) where each process holds an input matrix of size $m/P_r \times n/P_c$. 
	\item \urp: The partitioning strategy in \unp\ could result in a significant computational load imbalance in with a skewed nonzero distribution of $\Am$. We alleviate this by randomly permuting the rows and columns of the matrix before executing \mpifaun, and call this scheme $\urp$.
	\item \cpp: \distspnmf (\cref{alg:distnmf}) with checkerboard hypergraph partitioning explained in \cref{sec:cp1d}.
	\item \crp: \distspnmf (\cref{alg:distnmf}) with randomized checkerboard partioning explained in \cref{sec:cp1d}.

\end{itemize}

%The Figure \ref{fig:stackexchangerr} shows the comparison of these algorithms on the stack exchange dataset explained in Section \ref{sec:datasets}. 
\subsection {Strong Scaling}

In \cref{fig:flickr-speedup-16,fig:flickr-speedup-48} we show the speedup results of all four instances on the Rhea cluster using up to 3072 MPI ranks/cores on Flickr data.
The speedup values are with respect to slowest runtime among all four instances using 16 cores~(single node).
We observe in \cref{fig:flickr-speedup-16,fig:flickr-speedup-48} that all algorithms scale up to 1536 cores, yet \mpifaun instances achieve this with significantly lower parallel efficiency.
This mostly is due to higher communication costs involved in the all-to-all communication scheme for both instances.
We also realize that \urp significantly improves the runtime with respect to \unp, meaning that \unp indeed causes load imbalance in partitioning nonzeros of $\Am$.
At 3072 processes, both \urp and \unp lose scalability and slow down, whereas \cpp and \crp scale to 3072 processors.
One point to note is that for $k=16$, \cpp is considerably slower than \crp, whereas for $k=48$, \cpp starts to get faster than \crp.
The former result is due to that \cpp typically causes more imbalance in factor matrix rows and the input matrix nonzeros, whereas in the latter result, the size of each communication unit is tripled; therefore, the reduction in the communication cost using \cpp begins to compensate for the performance loss due to load imbalance.

Similarly, in \cref{fig:delicious-speedup-16,fig:delicious-speedup-48} we provide the same results for the Delicious matrix.
We observe a similar trend in the comparisons of different methods, except that \urp and \unp scale even worse in this case.
Our algorithm also loses scalability after 1536 processes, and similarly to the previous case \cpp starts slower than \crp due to load imbalance, and catches up for $k = 48$.
These two test cases clearly show that employing a point-to-point communication with good partitionings is essential for obtaining high performance inNMF algorithm.

To better test the scalability limit of our algorithms, we ran them on an IBM BlueGene/Q supercomputer up to 32768 processors using the same two matrices.
The results of these two experiments are provided in \cref{fig:flickr-speedup-bgq,fig:delicious-speedup-bgq}.
Our algorithm graciously scales up to 16384 cores in all four instances, and \cpp manages to slightly improve the runtime using 32768 cores on Flickr, while all other slowing down using 32768 ranks.
Again, \cpp is slower than \crp using lower number of processors as the communication cost is negligible in these instances, and \cpp introduces worse load balance than \crp.
However, using 32768 processors \cpp manages to outrun \crp by incurring less communication.

%%%%%%%%%%% Strong scaling flickr %%%%%%%%
\begin{figure}
\begin{subfigure}[t]{0.5\textwidth}
\begin{tikzpicture}
\renewcommand{\minvalue}{291.82}
\renewcommand{\numiterations}{30}
\renewcommand{\datafile}{data/flickr-speedup16.dat}
\speedup
\end{tikzpicture}
\subcaption{Flickr dataset with low rank $k=16$ on Rhea}
\label{fig:flickr-speedup-16}
\end{subfigure}

\begin{subfigure}[t]{0.5\textwidth}
\begin{tikzpicture}
\renewcommand{\minvalue}{1513.73}
\renewcommand{\numiterations}{30}
\renewcommand{\datafile}{data/flickr-speedup48.dat}
\speedup
\end{tikzpicture}
\subcaption{Flickr dataset low rank $k=48$ on Rhea}
\label{fig:flickr-speedup-48}
\end{subfigure}

\begin{subfigure}[t]{0.5\textwidth}
\begin{tikzpicture}
\bgqtrue
\renewcommand{\minvalue}{186.1}
\renewcommand{\numiterations}{10}
\renewcommand{\datafile}{data/flickr-speedup-bgq.dat}
\speedup
\end{tikzpicture}
\subcaption{Flickr dataset low rank $k=48$ on Bluegene/Q}
\label{fig:flickr-speedup-bgq}
\end{subfigure}

%\begin{subfigure}[t]{0.5\textwidth}
%\begin{tikzpicture}
%\bgqtrue
%\renewcommand{\minvalue}{137.1}
%\renewcommand{\numiterations}{10}
%\renewcommand{\datafile}{data/stackoverflow-speedup-bgq.dat}
%\speedup
%\end{tikzpicture}
%\subcaption{Stackoverflow dataset Low Rank $k=48$ on Bluegene/Q}
%\end{subfigure}

\caption{Strong scaling on Flickr dataset}
\label{fig:details}
\end{figure}

%%%%%%%%%%% Strong scaling delicious %%%%%%%%

\begin{figure}
\begin{subfigure}[t]{0.5\textwidth}
\begin{tikzpicture}
\renewcommand{\minvalue}{400.8}
\renewcommand{\datafile}{data/delicious-speedup16.dat}
\speedup
\end{tikzpicture}
\subcaption{Delicious dataset with low rank $k=16$ on Rhea}
\label{fig:delicious-speedup-16}
\end{subfigure}

\begin{subfigure}[t]{0.5\textwidth}
\begin{tikzpicture}
\renewcommand{\minvalue}{1423.05}
\renewcommand{\datafile}{data/delicious-speedup48.dat}
\speedup
\end{tikzpicture}
\subcaption{Delicious dataset low rank $k=48$ on Rhea}
\label{fig:delicious-speedup-48}
\end{subfigure}

%\begin{subfigure}[t]{0.5\textwidth}
%\begin{tikzpicture}
%\bgqtrue
%\renewcommand{\minvalue}{186.1}
%\renewcommand{\numiterations}{10}
%\renewcommand{\datafile}{data/delicious-speedup-bgq.dat}
%\speedup
%\end{tikzpicture}
%\subcaption{Flickr dataset Low Rank $k=48$ on Bluegene/Q}
%\end{subfigure}
%

\begin{subfigure}[t]{0.5\textwidth}
\begin{tikzpicture}
\bgqtrue
\renewcommand{\minvalue}{186.1}
\renewcommand{\numiterations}{10}
\renewcommand{\datafile}{data/delicious-speedup-bgq.dat}
\speedup
\end{tikzpicture}
\subcaption{Delicious dataset low rank $k=48$ on Bluegene/Q}
\label{fig:delicious-speedup-bgq}
\end{subfigure}

\caption{Strong scaling on Delicious dataset}
\label{fig:details}
\end{figure}

\subsection{Time Breakdown Per Iteration}\label{sec:perf-breakdown}

In this section we provide the time spent on each individual operation type and communication within an NMF iteration.
We report the averages over 30 iterations on Rhea, and 10 iterations on BlueGene/Q for each of four runs.
As provided in \cref{alg:distnmf} there are three types computations and two types of communications within an NMF iteration, and we present timings for these steps with the following labels:
\begin{itemize}
	\item \textbf{Gram}: Computing the local contribution to the Gram matrix, and performing an \textsc{All-Reduce} to gather the final result.
	\item \textbf{MM}: Computing the sparse matrix-dense-matrix multiplication using $\Amp$ and one of the factor matrices.
	\item \textbf{\NLS}: Local NNLS computation to compute the final value of the factor matrix.
	\item \textbf{Comm}: Total expand and fold communication time in the case of $\crp$ and $\cpp$,  and the total time spent on \textsc{All-Gather} and \textsc{Reduce-Scatter} steps for $\unp$ and $\urp$.
	%\item \textbf{All-Gather}, to compute the global matrix multiplication;
	%\item \textbf{Reduce-Scatter}, to compute the global matrix multiplication;
	%\item \textbf{All-Reduce}, to compute the global Gram matrix.
\end{itemize}
In our results, we do not distinguish the costs of these tasks for $\WW$ and $\HH$ separately; we instead report their sum.

We report the time breakdown for Flickr and Delicious datasets in \cref{fig:flickrdetails}, \cref{fig:deliciousdetails} for Rhea and \cref{fig:bgqdetailsk48} for BlueGene/Q.
For each cluster and data set, we show the timings for the smallest and the largest number of processors used.
Our objective in this experiment is to better analyze the speedup results and by comparing the computational and communication costs of different communication schemes and partitionings.

\paragraph{Flickr on Rhea:} We observe in \cref{fig:flickrdetails} that in one node configuration with $p=16$, the \unp and \urp\  performs similar to \crp\ and \cpp\ in terms of computation, and the communication time takes a small portion of the execution in all instances.
As the number of processes increases to 3072, the communication time of \crp\ and \cpp\ stays reasonably low, whereas in the case of \unp\  and \urp, we clearly observe that the communication cost dominates the execution time using both low rank values $k=16$ and $k=48$.
Randomization offers load balance to \urp\ which gives it a slight edge over \unp, yet both instances suffer from the high communication cost associated with the all-to-all communication strategy, which explains the drop in the scalability results.

\paragraph{Delicious on Rhea:} In \cref{fig:deliciousdetails}, we see that \crp\ and \cpp\ perform better than \unp\ even in single node configuration. \cref{fig:deliciousdetails} shows that \unp\ takes twice more than \crp\ and \cpp\ in the sparse matrix multiplication step, highlighting the skewed distribution of the matrix nonzeros, which is alleviated to a certain extend by randomly permuting the matrix.
Similar to Flickr data, using 3072 processors, \crp\ and \cpp\ perform significantly better than \unp\  and \urp, whose iteration times are dominated by the communication.

\paragraph{Flickr and Delicious on BlueGene/Q:}
In \cref{fig:bgqdetailsk48} we give the timings for computation and communication steps using our methods with two different partitionings of matrices on BlueGene/Q.
We observe that using 512 processors, communication cost is negligible, and \crp beats \cpp thanks to better load balance.
Using 16384 processors, however, on Flickr matrix \cpp gets faster than \crp due to significant reduction in the communication volume.
On Delicious matrix, \cpp similarly better reduces the communication, yet this is outweighed by the load imbalance in matrix multiplications.

%%%%%%%%%%% Flickr Details %%%%%%%%

\renewcommand{\minvalue}{1}
\begin{figure}
\begin{subfigure}[t]{0.25\textwidth}
\begin{tikzpicture}
\ylabeltrue
\legendfalse
\renewcommand{\numiterations}{30}
\renewcommand{\datafile}{data/flickr-details16-16.dat}
\makeplot
\end{tikzpicture}
\subcaption{Rhea $p=16$, Low Rank $k=16$}
\end{subfigure}
~
\begin{subfigure}[t]{0.25\textwidth}
\begin{tikzpicture}
\ylabelfalse
\legendtrue
\renewcommand{\numiterations}{30}
\renewcommand{\datafile}{data/flickr-details3072-16.dat}
\makeplot
\end{tikzpicture}
\subcaption{Rhea $p=3072$, Low Rank $k=16$}
\end{subfigure}

\begin{subfigure}[t]{0.25\textwidth}
\begin{tikzpicture}
\ylabeltrue
\legendfalse
\renewcommand{\numiterations}{30}
\renewcommand{\datafile}{data/flickr-details16-48.dat}
\makeplot
\end{tikzpicture}
\subcaption{Rhea $p=16$, Low Rank $k=48$}
\end{subfigure}
~
\begin{subfigure}[t]{0.25\textwidth}
\begin{tikzpicture}
\ylabelfalse
\legendtrue
\renewcommand{\numiterations}{30}
\renewcommand{\datafile}{data/flickr-details3072-48.dat}
\makeplot
\end{tikzpicture}
\subcaption{Rhea $p=3072$, Low Rank $k=48$}
\end{subfigure}

\caption{Flickr dataset Time Split up for Low Rank $k=16, 48$ on Rhea}
\label{fig:flickrdetails}
\end{figure}

%%%%%%%%%%% Delicious Details %%%%%%%%

\renewcommand{\minvalue}{1}
\begin{figure}
\begin{subfigure}[t]{0.25\textwidth}
\begin{tikzpicture}
\ylabeltrue
\legendfalse
\renewcommand{\numiterations}{30}
\renewcommand{\datafile}{data/delicious-details16-16.dat}
\makeplot
\end{tikzpicture}
\subcaption{Rhea $p=16$, Low rank $k=16$}
\end{subfigure}
~
\begin{subfigure}[t]{0.25\textwidth}
\begin{tikzpicture}
\ylabelfalse
\legendtrue
\renewcommand{\numiterations}{30}
\renewcommand{\datafile}{data/delicious-details3072-16.dat}
\makeplot
\end{tikzpicture}
\subcaption{Rhea $p=3072$, Low rank $k=16$}
\end{subfigure}

\begin{subfigure}[t]{0.25\textwidth}
\begin{tikzpicture}
\ylabeltrue
\legendfalse
\renewcommand{\numiterations}{30}
\renewcommand{\datafile}{data/delicious-details16-48.dat}
\makeplot
\end{tikzpicture}
\subcaption{Rhea $p=16$, Low Rank $k=48$}
\end{subfigure}
~
\begin{subfigure}[t]{0.25\textwidth}
\begin{tikzpicture}
\ylabelfalse
\legendtrue
\renewcommand{\numiterations}{30}
\renewcommand{\datafile}{data/delicious-details3072-48.dat}
\makeplot
\end{tikzpicture}
\subcaption{Rhea $p=3072$, Low Rank $k=48$}
\end{subfigure}

\caption{Delicious dataset Time Split up for Low Rank $k=16, 48$ on Rhea}
\label{fig:deliciousdetails}
\end{figure}

%%%%%%%%%%% Flickr & Delicious BGQ Details for k=48 %%%%%%%%
\begin{figure}
\begin{subfigure}[t]{0.25\textwidth}
\renewcommand{\numiterations}{10}
\begin{tikzpicture}
\ylabeltrue
\renewcommand{\numiterations}{10}
\renewcommand{\datafile}{data/flickr-details-bgq-512.dat}
\makeplotnew
\end{tikzpicture}
\subcaption{BlueGene/Q $p=512$}
\end{subfigure}
~
\begin{subfigure}[t]{0.25\textwidth}
\begin{tikzpicture}
\ylabelfalse
\renewcommand{\numiterations}{10}
\renewcommand{\datafile}{data/flickr-details-bgq-16384.dat}
\makeplotnew
\end{tikzpicture}
\subcaption{BlueGene/Q $p=16384$}
\end{subfigure}

\caption{Time Split up of Flickr and Delicious for Low Rank $k=48$ on Bluegene/Q. The left two bars are for the Flickr matrix, and the right two bars are for the Delicious matrix}
\label{fig:bgqdetailsk48}
\end{figure}

### 2018-04-02 latex/sc17/notations.tex
moved under latex

\subsection{Notation}
\label{sec:notations}

\Cref{tab:notation} summarizes the notation we use throughout this paper.
We use bold uppercase letters for matrices and lowercase letters for vectors.
For matrix rows and columns, we employ MATLAB notation, i.e., $\AA(i, :)$ and $\AA(:, j)$ refer to the $i$th row and the $j$th column of $\AA$.
We use subscripts to refer to sub-blocks of matrices.
For example, $\AA_ij$ refers to the sub-block $(i, j)$ of $\AA$ in a 2D partition.
We use $m$ and $n$ to denote the numbers of rows and columns of $\AA$, respectively, and assume without loss of generality $m\geq n$ throughout.

\begin{table}%[htdp]
\begin{center}
\begin{tabular}{|l|l|}
\hline
$\AA$ & Input matrix \\
$\WW$ & Left low rank factor \\
$\HH$ & Right low rank factor \\
$m$ & Number of rows of input matrix \\
$n$ & Number of columns of input matrix \\
$k$ & Low rank \\
$P$ & Number of parallel processes \\
$P_r$ & Number of rows in processor grid \\
$P_c$ & Number of columns in processor grid \\
$\Ip$,$\Jp$ & Set of rows/columns of of $\Wm$/$\Hm$ owned by process $p$  \\
$\Fp$,$\Gp$ & Set of unique row and column indices of $\Amp$ \\
$\Amp$ & Submatrix of $\Am$ owned by process p \\
$\Wm(\Ip,:)$ & Owned rows of initial $\Wm$ by process p \\
$\Hm(:,\Jp)$ & Owned columns of initial $\Hm$ by process p\\

\hline
\end{tabular}
\end{center}
\caption{Notation}
\label{tab:notation}
\end{table}%

### 2018-04-02 latex/sc17/survey.tex
moved under latex

\section{survey}

%\begin{itemize}
%\item NMF papers
%\item HPC-NMF papers
%\item Difference between matrix factorization and NMF
%\item Hypergraph partitions paper
%\end{itemize}

In the data mining and machine learning literature, there is an overlap between low rank approximations and matrix factorizations due to the nature of applications.
Despite its name, non-negative matrix ``factorization'' is in fact a low rank approximation.
Recently, there has been a growing interest in collaborative filtering based
recommender systems. One of the popular techniques
for collaborative filtering is matrix factorization, often with nonnegativity constraints,
and its implementation is widely available in many
off-the-shelf distributed machine learning libraries
such as GraphLab \cite{low2012}, MLLib \cite{meng2015mllib},
and many others \cite{satish2014,yun2014}.
However, we would like to emphasize that collaborative filtering using matrix factorization is a different problem than NMF:
In the case of collaborative filtering, nonzeros in the matrix
are considered to be observed ratings and zeros to be missing entries, while in the case of NMF, there is no missing entries and even zero is an observed entry.

There are several recent distributed NMF algorithms in the literature \cite{liao2014cloudnmf,Faloutsos2014,Yin2014,liu2010distributed}.
Liu et al.\ propose running Multiplicative Update (MU) for KL divergence, squared loss, and ``exponential'' loss functions \cite{liu2010distributed}.
Matrix multiplication, element-wise multiplication, and element-wise division are the building blocks of the MU algorithm.
The authors discuss performing these matrix operations effectively in Hadoop for sparse matrices.
Using similar approaches, Liao et al.\ implement an open source Hadoop-based MU algorithm and study its scalability on large-scale biological data sets \cite{liao2014cloudnmf}.
Also, Yin, Gao, and Zhang present a scalable NMF that can perform frequent updates, which aim to use the most recently updated data \cite{Yin2014}.
Similarly Faloutsos et al.\ propose a distributed, scalable method for decomposing matrices, tensors, and coupled data sets through stochastic gradient descent on a variety of objective functions \cite{Faloutsos2014}.
The authors also provide an implementation that can enforce non-negative constraints on the factor matrices.
All of these works use Hadoop framework to implement their algorithms, hence are not very efficient.

%We emphasize that our MPI-based approach has several advantages over Hadoop-based approaches:
%\begin{itemize}
%        \item efficiency -- our approach maintains data in memory, never communicating the data matrix, while Hadoop-based approaches must read/write data to/from disk and involves global shuffles of data matrix entries;
%        \item generality -- our approach is well-designed for both dense and sparse data matrices, whereas Hadoop-based approaches generally require sparse inputs;
%        \item privacy -- our approach allows processors to collaborate on computing an approximation without ever sharing their local input data (important for applications involving sensitive data, such as electronic health records), while Hadoop requires the user to relinquish control of data placement.
%\end{itemize}

Spark \cite{ZCFSS10} is a popular big-data processing infrastructure that is generally more efficient for iterative algorithms such as NMF than Hadoop, as it maintains data in memory and avoids file system I/O.
Even with a Spark implementation of previously proposed Hadoop-based NMF algorithms, the performance still suffers from expensive communication of input matrix entries, and Spark does not have innate mechanisms to overcome this shortcoming. 
Spark has collaborative filtering libraries such as MLlib \cite{meng2015mllib} which use matrix factorization and can impose non-negativity constraints. %none of them implement pure NMF, and so we do not have a direct comparison against NMF running on Spark.
Nevertheless, as mentioned, the problem of collaborative filtering is different from NMF,  hence we do not compare our approach against these.

In parallel with the Hadoop and Spark implementations, there have been growing interest in the HPC community towards efficiently computing these algorithms with tuned high performance implementations. Kannan, Ballard and Park \cite{KBP16, KBP16MPIFAUN}, have proposed \mpifaun framework to implement various NMF algorithms such as multiplicative update (MU), Hierarchical Alternating Least Squares (HALS) and Alternating Non-negative Least Squares using Block Principal Pivoting (ANLS-BPP).
We choose this work as a baseline, as it is the only available high performance implementation of NMF, and it performs significantly faster than Hadoop and Spark-based approaches.
To elaborate this, Gittens et.al., \cite{GDRRGKLMC16} recently benchmarked the implementations of different matrix factorization algorithms, such as NMF and Principal Component Analysis~(PCA), in Spark and in C and MPI, which is tuned for HPC platforms.
They claim native MPI implementations on HPC platforms out perform Spark implementation by a speedup factor of 44x.
Similar observations have been made by Sukumar, Kannan, Matheson and Lim \cite{SKLM16, SMKL16} on super computers at Oak Ridge Leadership Computing Facility. 
Finally, there are implementations of the MU algorithm in a distributed memory setting using X10 \cite{Grove2014} and on a GPU \cite{mejia2015nmf}.

%Given these existing literature, we would like to highlight our contribution in this paper here.
%(a) Our proposed Algorithm \distspnmf that is agnostic to distribution strategies. In our humble opinion, this is the first high performance distributed sparse NMF algorithm in the literature. 
%(b) We compare both the MPI collective communication and $P2P$ in real world datasets on very large number of processors
%(c) Sparse regularized NMF which otherwise will result in numerical in-stability for large high-dimensional sparse matrices 
%(d) The intermediate memory requirement for $MPI-FAUN$ restricts its ability to handle sparse matrices of few hundred millions and more in smaller processor and we have to extend blocking support on the low rank side.

### 2018-04-03 latex/icpp18/algorithm.tex
seeding icpp18

\section{Algorithm}

% Notation for frequently used matrices and indices. Defined in spdistnmfsc17.tex, commented out here for reference.
% \newcommand{\mat}[1]{\mathbf{#1}}  % Denote a matrix
% \newcommand{\matt}[1]{\mathbf{\tilde{#1}}}  % Denote a matrix with hat
% \newcommand{\Am}{\mat{A}} % Input sparse matrix
% \newcommand{\Amp}{\mat{A}_p} % Input sparse matrix
% \newcommand{\Wm}{\mat{W}} % Factor matrix
% \newcommand{\Wmt}{\matt{W}} % AH result
% \newcommand{\Hm}{\mat{H}} % Factor matrix
% \newcommand{\Hmt}{\matt{H}} % WtA result
% \newcommand{\Ip}{\mathcal{I}_p}
% \newcommand{\Jp}{\mathcal{J}_p}
% \newcommand{\Fp}{\mathcal{F}_p}
% \newcommand{\Gp}{\mathcal{G}_p}

Here, we first introduce our parallel NMF algorithm that operates on a partition of the matrices $\Am$, $\Wm$, and $\Hm$.
For a given partition, we describe how parallel computations and communications take place within the algorithm, and illustrate computational and communication costs associated with a partition.
In doing so, we elaborate the advantages of our parallelization scheme compared with the state of the art.
We then discuss efficient partitioning strategies to better establish computational load balance and reduce communication.

\subsection{Distributed Parallel Sparse NMF Algorithm}

\begin{algorithm}
\caption{\distspnmf: \distspnmffull}
\label{alg:distnmf}
  \begin{algorithmic}[1]
    \setcounter{ALC@unique}{0}
    \REQUIRE {$\Amp$: An $m \times n$ sparse matrix \\
      $\Ip$,$\Jp$: Set of rows/columns of $\Wm$ and $\Hm$ owned by process $p$ \\
      $\Fp$,$\Gp$: Footprints of process $p$ on $\Wm$ and $\Hm$ \\
      $\Wm(\Ip,:)$,$\Hm(:,\Jp)$: Owned rows and columns of initial $\Wm$ and $\Hm$ \\
      $k$: The NMF rank}
    \ENSURE {Process $p$ gets the final value of $\Wm(\Ip,:)$ and $\Hm(:,\Jp)$}
      \REPEAT
        \STATE $\Cexpand (\Hm(:,\Gp))$ \label{line:exph}
        \STATE $\Wmt(\Fp, :) \gets \Amp \Hm(:, \Gp)$ \label{line:ah}
        \STATE $\Cfold(\Wmt(\Fp, :))$ \label{line:foldw}
        \STATE $\Hmgr \gets \textsc{All-Reduce}(\Hm(:, \Jp) \Hm(:, \Jp)^T)$ \label{line:hmgr}
        \STATE $\Wm(\Ip, :) \gets \textsc{Nnls}(\Hmgr, \Wmt(\Ip, :))$ \label{line:wmnnls}
        \STATE $\Cexpand(\Wm(\Fp, :))$ \label{line:expw}
        \STATE $\Hmt(:, \Gp) \gets \Wm(\Fp, :)^T \Amp$ \label{line:wta}
        \STATE $\Cfold(\Hmt(:, \Gp))$ \label{line:foldh}
        \STATE $\Wmgr \gets \textsc{All-Reduce}(\Wm(\Ip, :)^T \Wm(\Ip, :))$ \label{line:wmgr}
        \STATE $\Hm(\Jp, :) \gets \textsc{Nnls}(\Wmgr, \Hmt(\Jp, :))$ \label{line:hmnnls}
      \UNTIL{convergence or maximum number of iterations}
  \end{algorithmic}
\end{algorithm}

Parallellizing NMF involves partitioning data and computational tasks to processes, and communicating results wherever necessary, just as any parallel algorithm.
Specifically for NMF, one needs to partition the sparse matrix $\Am$ as well as the factor matrices $\Wm$ and $\Hm$, where the former partitioning distributes the computational load of sparse matrix-dense matrix multiplications $\Am \Hm$ and $\Wm^T \Am$, whereas the latter divides the workload of NNLS computations to processes.
We provide the execution of our parallel algorithm for computing a rank-$k$ NMF of a sparse matrix $\Am \in \R^{m \times n}$ in \cref{alg:distnmf}, which is executed by each process $p$ for $1 \le p \le P$.
The algorithm starts with an arbitrary partition of the input matrix and the factor matrices; process $p$ owns the submatrices $\Wm(\Ip, :)$ and $\Hm(:, \Jp)$ as well as the nonzero elements of the sparse matrix $\Am_p$ where $\Am = \bigcup_{i=1}^{P}\Am_i$, i.e., $\Am_1, \dots, \Am_p$ partitions the nonzeros of $\Am$.
The sets $\Fp$ and $\Gp$ denote the ``footprints'' of the process $p$ on the rows and the columns of matrices $\Wm$ and $\Hm$, respectively, in the course of the algorithm.
That is, we have $i \in \Fp$ or $j \in \Gp$ if only if $i \in \Ip$ or $j \in \Jp$~(row/column is owned), or there is a nonzero element $a_{i,j} \in \Amp$~(row/column is used in local computations).
At each iteration, the process $p$ is responsible for gathering the new value of submatrices $\Wm(\Ip, :)$ and $\Hm(:, \Jp)$, and sending these values to processes in need.

In an iteration of \cref{alg:distnmf}, each process $p$ possesses three types of computational tasks as well as associated pre- and post-communication steps.
The first task involves performing sparse matrix-dense matrix multiplications $\Am_p \Hm(:, \Gp)$ and $\Wm(\Fp, :)^T \Am_p$, whose results are stored in distributed matrices $\Wmt$ and $\Hmt$, which follow the same row-/column-wise data distribution as $\Wm$ and $\Hm$.
Note that carrying out these multiplications must be preceded by a communication step where each process $p$ gets the submatrices $\Hm(:, \Gp \setminus \Jp)$ and $\Wm(\Fp \setminus \Ip, :)$ that are accessed by entries of $\Amp$, and this step is performed at \cref{line:exph,line:expw}.
These multiplications performed by each process $p$ generate partial results for the set $\Fp$ and $\Gp$ of rows of $\Wmt$ and columns of $\Hmt$, respectively, which is highlighted at \cref{line:ah,line:wta}.
Indeed, partial results for the submatrices $\Wmt(\Fp \setminus \Ip, :)$ and $\Hmt(:, \Gp \setminus \Jp)$ correspond to rows and columns owned by other processes; hence, they need to be communicated.
The results for $\Wmt(\Ip, :)$ and $\Hmt(:, \Jp)$, however, should be kept locally, and all partial results for these matrix rows and columns generated by other processes should similarly be received and accumulated in order to obtain the final value for these owned portions.
The second task is to compute the Gram matrices $\Hmgr = \Hm \Hm^T$ and $\Wmgr = \Wm^T \Wm$ of size $k \times k$, and making these matrices available to all processes, which is performed at \cref{line:hmgr,line:wmgr}.
This is done in a row-parallel dense matrix multiplication step, in which the process $p$ computes $\Hm(:, \Jp) \Hm^T(:, \Jp)$ and $\Wm^T(\Ip, :) \Wm(\Ip, :)$, followed by an \textsc{All-Reduce} communication of these partial multiplications.
The third task pertains to updating the factor matrices $\Wm$ and $\Hm$ using matrices $\Wmt$ and $\Hmgr$, or $\Hmt$ and $\Wmgr$, which takes place at \cref{line:wmnnls,line:hmnnls}.
This corresponds to \cref{line:aunmf-w,line:aunmf-h} of \cref{alg:aunmf}, and can be computed locally at each process $p$ by executing \textsc{Nnls} algorithm on dense matrices $\Wmt(\Ip, :)$ and $\Hmgr$, or $\Hmt(:, \Jp)$ and $\Wmgr$, to obtain new $\Wm(\Ip, :)$ or $\Hm(\Ip, :)$, respectively, as described in \cref{sec:sparsenmfmu}.

The first type of communication in \cref{alg:distnmf} pertains to an \textsc{All-Reduce} of a dense matrix of fixed size $k \times k$ at \cref{line:hmgr,line:wmgr}, and the cost of this step is typically negligible in compare to the rest.
The other two communication types involve (i) transferring the partial row results of $\Wmt$ and $\Hmt$ to their owner processes at \cref{line:foldw,line:foldh} to accumulate at the owners, (ii) sending the updated rows of $\Wm$ and $\Hm$ to processes in need at \cref{line:expw,line:exph}.
We respectively call these steps \emph{fold} and \emph{expand} communications, following the convention used by the sparse matrix community.
The way these two communications are carried out plays a vital role in obtaining parallel scalability as they dominate the communication cost of the algorithm.
Here, we employ an efficient point-to-point communication framework to effectuate fold and expand communications.
In this scheme, after determining a partition of the matrix, each process $p$ determines the sets $\Fp$ and $\Gp$ using the row and column indices in its local matrix $\Amp$ and its sets $\Ip$ and $\Jp$ of owned row indices, then identifies the owners of these rows in order to request these rows from the corresponding processes in the expand step.
Next, these requests are exchanged in a communication step so that each process $p$ determines the unique row indices that it needs to send to or receive from each other process in expand and fold steps.
Once this communication structure is formed, it is reused througout all iterations in \cref{alg:distnmf}.
This is possible thanks to the data partition staying fixed during the algorithm, hence data dependencies do not change.

The existing work in the literature \cite{KBP16,KBP16MPIFAUN} amounts to employing an all-to-all communication strategy for both expand and fold steps of \cref{alg:distnmf} for dense as well as sparse $\Am$.
In this strategy, the rows and the columns indices $1, \dots, m$ and $1, \dots, n$ are divided into $P_r$ and $P_c$~($P = P_r P_c$) sets $I_1, \dots, I_{P_r}$ and $J_1, \dots, J_{P_c}$ of equal size and having contiguous indices.
This effectively defines a 2D uniform checkerboard partition of $\Am$ with matrix blocks $\Am_{(r,c)} = \Am(I_r, J_c)$ for all $1 \le r \le P_r$ and $1 \le c \le P_c$, as well as a row partition of $\Wm$ and $\Hm$.
Here, process $p$ owns the matrix subblock $\Am_{(r, c)}$, where $r = \lfloor P/P_c \rfloor + 1$ and $c = (P \mod P_c) + 1$, as well as $m / P$ and $n / P$ rows of $\Wm(I_r, :)$ and $\Hm(J_c, :)$.
The communication is performed within each process row and column using \textsc{All-Gather} and \textsc{Reduce-Scatter} routines, in which the process $p$ receives all matrix rows $\Wm(I_r, :)$ and $\Hm(I_c, :)$ belonging to processes in the same row and column of the process grid.
Despite being favorable due to small number of exchanged messages in all-to-all routines, in this strategy processes receive many rows that they do not need in their local sparse matrix dense matrix multiplication, and this redundancy dramatically increases the communication volume, thus preventing scalability. 
%For convenience of the reader, \cref{tab:costs} discusses the computation, communication and memory complexity of this partitioning scheme that is observed in the \mpifaun algorithm. 
To illustrate, in \cref{fig:partition} we provide a sparse matrix partitioned using a $5 \times 5$ checkerboard scheme where each process owns a subblock of the matrix, and show only the nonzero elements belonging to three columns with indices $j_1$, $j_2$, and $j_3$.
Here, in performing the expand communication of \cref{line:exph}, the all-to-all strategy sends these column vectors to all processes in the same column, whereas point-to-point communication identifies the exact set of proceses in need beforehand, indicated with the same color, and communicates the data only with these processes.
In this example, all-to-all scheme incurs 4, 3, and 2 units of redundant communication for columns $j_3$, $j_2$, and $j_1$, respectively, and this redundancy only worsens with the increasing number of processes.

\begin{figure}
\caption{A 5x5 checkerboard partition of a sparse matrix.}
\label{fig:partition}
\includegraphics[width=0.8\linewidth]{figures/matrix-partition.png}
\end{figure}

\subsection{Partitioning}
\Cref{alg:distnmf} requires a partitioning of the nonzeros of $\Am$ as well as the rows and the columns of $\Wm$ and $\Hm$, and these three partitions completely determine its computational and communication costs.
Here, we compare different partitioning strategies, and argue how they relate to these two performance metrics.

Our algorithm is fine-grained; meaning that it can work any fine-grain, 1D, or 2D/checkerboard partition of the sparse matrix $\Am$.
Here, however, we exclusively focus on $P_r \times P_c$~(where $P = P_r P_c$) checkerboard partitionings, and always set $P_c$ to the number of cores available per compute node in the cluster.
This helps us reduce the communication cost in two ways.
The first one corresponds to that we do not employ shared-memory parallelism in our implementation, hence assign one MPI rank per core.
In this scenario, by assigning each group of $P_c$ processes owning the same row block of $\Am$ to the same node, and similarly assigning the corresponding row blocks of $\Wm$ and $\Wmt$ to these processes, we effectively eliminate the communication in the row dimension of the process grid~(which pertains to communicating matrices $\Wm$ and $\Wmt$) as these row exchanges stay within a node.
This enables us to focus solely on reducing the communication due to columns of $\Hm$ and $\Hmt$.
Second, this topology bounds the maximum number of messages sent per process by $P_r-1$ and $P_c-1$, thereby significantly reduce the communication latency by a factor of $P_c$ in compare with a fine-grain or a 1D partition, which incurs up to $P-1$ messages per process.

\subsubsection{Partitioning sparse input matrix $\Am$} \label{sec:cp1d}
\paragraph{Hypergraph partitioning~(\cpp)} We aim to partition the matrix $\Am$ into $P_r$ row slices first, and $P_c$ column slices next, to obtain an $P_r \times P_c$ checkerboard partition.
To this end, we adopt hypergraph partitioning techniques used in parallel sparse matrix literature to our case~\cite{caay:99}.
A hypergraph consists of vertices with associated weights and hyperedges that connect two or more vertices.
In the literature, a hypergraph is typically formed by adding a vertex for each computational task with the associated execution cost, adding a hyperedge for each data element, and connecting the vertex to a hyperedge whenever the associated task and the data are dependent.
Then, the vertices of the hypergraph is partitioned using a hypergraph partitioner to distribute vertex loads to parts equitably while reducing a metric called \emph{cutsize}, which amounts to minimizing the total number of different parts each hyperedge connects.
This corresponds in the actual computation to minimizing the data dependencies between tasks, hence the communication volume.
Here, we form a \emph{column-net} hypergraph model; for each row $i$ and column $j$ of  $\Am$, we add a vertex $v_i$ and a hyperedge $h_j$ to the hypergraph, and for each nonzero $(i, j) \in \Am$, which implies a dependency to $\Hm(:, j)$ in computing $\Wmt(i, :)$ at \cref{line:ah}, we connect $v_i$ to $h_j$.
We assign the value $(nnz(\Am) / m + nnz(\Am(i, :))$ to each vertex $v_i$.
This weighting is preferred to find a tradeoff between balancing the number of vertices, which corresponds to the cost of dense matrix operations, and balancing the number of nonzeros per part, which corresponds to the cost of sparse matrix-dense matrix multiplication.
We establish these two load balance goals by assigning a fixed cost to each vertex in the first summand, and increasing the cost of the vertices corresponding to denser rows in the second summand.
We then partition this hypergraph into $R$ parts by balancing the vertex weights, and minimizing the cutsize of the hypergraph.
In \cref{fig:partition} we also show this hypergraph construction for a $3 \times 3$ subset of the $A$ where rows corresponding to each vertex are highlighted and connected to related hyperedges.
In this partitioning, minimizing the cutsize of this hypergraph exactly corresponds to minimizing the total communication volume due to columns of $\Hm$ and $\Hmt$ as is clear in \cref{fig:partition}, whereas balancing the vertex weights corresponds to balancing the cost of sparse matrix-dense matrix multiplications as well as dense matrix operations in computing Gram matrices and $\textsc{Nnls}$ in each process row.

Once the matrix is partitioned into $R$ row slices, one can similarly partition columns of $\Am$ into $C$ parts using a multi-constraint formulation~\cite{aycu:08} using a \emph{row-net} hypergraph.
Yet in our case, we skip this common approach for two reasons.
First, existing multi-constraint partitioning tools are known to be not as efficient nor effective as single-constraint partitioners in practice.
Second, the main goal of hypergraph partitioning of columns is to minimize the communication volume due to rows, whereas in our case communication of rows of $\Wm$ and $\Wmt$ stays within a node with a proper mapping of processes, and this communication cost would be negligible.
Therefore, we instead partition columns randomly, which is expected to provide balance the number of nonzeros of $\Am$, number of columns of $\Hm$ and $\Hmt$, the amount of column communication volume due to columns of $\Hm$ and $\Hmt$.

\paragraph{Randomized checkerboard partitioning(\crp)}
This scheme corresponds to partitioning both the rows and the columns of $\Am$ into $R$ segments randomly.
It is expected to provide good load balance both in sparse and dense matrix operations, but it overlooks the communication cost.

\paragraph{Uniform checkerboard partitioning(\unp)}
This partitioning variant forms an $R \times C$ partition of $\Am$ by putting a contiguous set of $m / R$ and $n / C$ rows and columns in each slice.
$\Wm$ and $\Hm$ are partitioned conformally with this topology; each process is assigned a contiguous set of  $m / P$ and $n / P$ rows and columns, respectively, belonging to the corresponding row and column blocks of $\Hm$.
This is the only partitioning scheme employed by $\mpifaun$ \cite{KBP16, KBP16MPIFAUN}, hence we only use it for this implementation.
We also use a randomized variant (\urp) of this scheme in which the rows and columns of $\Am$ are permuted randomly befor executing $\mpifaun$ to better balance its nonzeros.

\subsubsection{Partitioning factors $\Wm$ and $\Hm$}
Once $\Am$ is partitioned with an $R \times C$ checkerboard topology, one has to partition rows and columns of factor matrices conformally with this topology to form the sets $\Ip$ and $\Jp$ in \cref{alg:distnmf}.
In doing so, we are interested in assigning rows and columns to processes in the corresponding process row or column equitably.
For this purpose, we specify imbalance parameters $\alpha_r$ and $\alpha_c$ that correspond to maximum imbalance we allow in this partitioning; i.e., $|\Ip| \le \alpha_r m / P$ and $|\Jp| \le \alpha_c n / P$ for each process $p$.
We normally set these imbalance parameters to $1.1$.
However, when $m >> n$, we increase $\alpha_c$ proportionally in order to give more flexibility in column partitioning which results in smaller communication volume.

Next, for each row and column of $\Wm$ and $\Hm$ we create a list of processes that has a dependency to that row or column, which corresponds to processes owning the matrix blocks of same color in \cref{fig:partition}.
Finally, we randomly assign each row and column to one of the processes satisfying the imbalance constraint in this list.
If all processes in the list are overloaded, we randomly assign it to a process that is not overloaded in the same process row or column.
Note that the latter assignment increases the communication volume due to that row or column by 1; hence, in general larger imbalance parameters yield smaller communication volume by reducing this type of assignment.

%%\newcommand{\smaller}{\scriptsize}
%\begin{table*}[t!]
%\begin{center}
%\begin{tabular}{|c|c|c|c|c|}
%\hline
%\textbf{Algorithm} & \textbf{Flops} & \textbf{Words} & \textbf{Messages} & \textbf{Memory}  \\ \hline
%\smaller \mpifaun ($m/p < n$) & \smaller $O(2*nnz_p*k)+\frac{(m+n)k^2}{p}+F\lt(\frac mp,\frac np,k\rt)$ & \smaller $O\lt( \sqrt{\frac{mnk^2}{p}}\rt)$ & \smaller $O(\log p)^*$ & \smaller $O\lt(\frac{mn}{p}+\sqrt{\frac{mnk^2}{p}}\rt)$ \\ \hline
%\end{tabular}
%\normalsize
%\end{center}
%\caption{Leading order algorithmic costs for \NaiveAlg{} and \ParNMF{} (per iteration).  
%The function $F(\cdot)$ denotes the number of flops required for the particular NMF algorithm's Local Update Computation, aside from the matrix multiplications common across AU-NMF algorithms. \\
%$^*$The stated latency cost assumes no communication is required in \LUC; \HALS\ requires $k\log p$ messages for normalization steps.}
%\label{tab:costs}
%\end{table*}%

### 2018-04-03 latex/icpp18/aunmf.tex
seeding icpp18

\section{Sparse Non-negative Matrix Factorization}
\label{sec:sparsenmf}

%We are extending the $AUNMF$ algorithm in \mpifaun framework \cite{KBP16MPIFAUN, KBP16} for implementing different NMF algorithms to add support for sparse input matrix.
The Alternating-Updating NMF algorithms are those that alternate between updating one of $\WW$ and $\HH$ using the given input matrix $\AA$ and other 'fixed' factor - $\HH$ for updating $\WW$ or $\WW$ for updating $\HH$.
This update is performed using the Gram matrix associated with the fixed factor matrix, and the product of the input matrix $\AA$ with the fixed factor matrix.
We show the structure of the framework in \cref{alg:aunmf}.

%Different algorithms determine $\WW$ and $\HH$ by partitioning these matrices in different ways and can be explained under Block Coordinate Descent (BCD) framework. 
%Generally, these matrices are determined as (a) 2-Blocks of entire matrix $\WW$ and $\HH$; (b) $2k$ blocks such as $[\ww^1, \ww^2, \cdots , \ww^m] \in \Rnplus{k}$ and $[\hh_1^T, \hh_2^T, \cdots , \hh_n^T] \in \Rnplus{k}$ \footnote{For convenience, $\M{x}^i$ is the $i$th column vector of matrix $\M{X}$ and $\M{x}_i$ is the $i$th row vector} and (c) $(m+n)k$ scalar blocks of $w_{ik} \in \WW$ and $h_{kj} \in \HH$.

\begin{algorithm}
\caption{$[\WW,\HH] = \text{AU-NMF}(A,k)$}
\label{alg:aunmf}
\begin{algorithmic}[1]
\REQUIRE $\AA$ is an $m\times n$ matrix, $k$ is rank of approximation
\STATE Initialize $\HH$ with a non-negative matrix in $\Rn{n\times k}_+$.
\WHILE{stopping criteria not satisfied} \label{algo:nmfloop}
  \STATE Update $\WW$ using $\HH \HH^T$ and $\AA \HH^T$ \label{line:aunmf-w}
  %\STATE Update $\WW$ as $\Argmin{\tilde \WW\geq 0}\NormBr{\AA-\tilde\WW\HH}_F$
  %\STATE Update all blocks defined over $\WW$
  \label{line:aunmf:W}
  \STATE Update $\HH$ using $\WW^T\WW$ and $\WW^T \AA$ \label{line:aunmf-h}
  %\STATE Update $\HH$ as $\Argmin{\tilde\HH\geq 0}\NormBr{\AA-\WW\tilde\HH}_F$
  %\STATE Update all blocks defined over $\HH$
  \label{line:aunmf:H}
\ENDWHILE
\end{algorithmic}
\end{algorithm}

After computing the Gram matrix and the multiplication of $\AA$ with the fixed factor matrix, the specifics of the update at \cref{line:aunmf:W,line:aunmf:H} depend on the NMF algorithm, and we refer to the computation associated with these lines as the Local Update Computations (\LUC).
%Because these computations are performed locally, we use a function $F(m,n,k)$ to denote the number of flops required for each algorithm's \LUC\ (and we do not consider communication costs).

We note that AU-NMF is very similar to a two-block, block coordinate descent (BCD) framework as explained by Bertsekas \cite{bertsekas1999nonlinear}.
The BCD framework expresses solving optimization variables in complex non-linear optimization problem as one block at a time, while keeping the others fixed. In NMF,  the two blocks are
the unknown factors $\WW$ and $\HH$, 
and we \emph{solve} the following subproblems,
which have a unique solution for a full rank $\HH$ and $\WW$: 
\SplitN{\label{eqn:two block}} {
\WW &\leftarrow \Argmin{\tilde \WW\geq 0}\NormBr{\AA-\tilde\WW\HH}_F +  \phi (\tilde \WW) + \psi (\HH),\\
\HH &\leftarrow  \Argmin{\tilde\HH\geq 0}\NormBr{\AA-\WW\tilde\HH}_F + \phi (\WW) + \psi (\tilde \HH).
}

Since each subproblem involves non-negative least squares,
this two-block BCD method is also called
the Alternating Non-negative Least Squares (ANLS) method \cite{kim2013nonnegative}.
Block Principal Pivoting (\BPP) is one algorithm that solves these NLS subproblems.
In the context of the AU-NMF algorithm,
 an ANLS method {\em maximally} reduces 
the overall NMF objective function value 
by finding the optimal solution for
 given $\HH$ and $\WW$ in \cref{line:aunmf:W,line:aunmf:H}, respectively.

%There are other popular NMF algorithms 
%that update the factor matrices alternatively
%without maximally reducing the objective function value each time,
%in the same sense as in ANLS. 
From time to time, these updates do not necessarily solve each of the subproblems \eqref{eqn:two block} to optimality but simply improve the overall objective function \eqref{eqn:original NMF}.  
such as Multiplicative Update (\MU) \cite{seung2001algorithms} and Hierarchical Alternating Least Squares (\HALS) \cite{cichocki2009nonnegative}.

The convergence properties of these different NMF algorithms are discussed in detail by Kim, He and Park \cite{kim2013nonnegative}.
While we focus only on the \MU algorithms in this paper, we highlight that our algorithm is not restricted to this, and is seamlessly extensible to other NMF algorithms as well, including \HALS, \BPP, Alternating Direction Method of Multipliers (\textbf{ADMM}) \cite{SF14}, and Nesterov-based methods \cite{GTLY12}.

%Here, we have explained the update equations of the NMF algorithms ANLS/BPP, HALS and MU. In all these three algorithms, the update equations uses $\AA\HH^T$ and $\HH\HH^T$ for line \ref{line:anls:W} and $\WW^T\AA$ and $\WW^T\WW$ for line \ref{line:anls:H}.

%As in the Figure \ref{fig:2blocks}, the ANLS/BPP \cite{kim2011fast} is an example of 2-Blocks BCD that determines $\WW$ and $\HH$ as matrix blocks by solving 
%the following equations using active set method iteratively until a stopping criteria is satisfied:

%\begin{figure*}
%\subfloat[][ANLS/BPP 2-Blocks BCD] {\includegraphics[height=1.5in, width=2.5in]{fig/bcd2blocks}\label{fig:2blocks}}
%\subfloat[][HALS 2k-Blocks BCD] {\includegraphics[height=1.5in, width=2.5in]{fig/bcd2kblocks}\label{fig:2kblocks}}
%\subfloat[][MU (m+n)k Scalar Blocks BCD] {\includegraphics[height=1.5in, width=2.5in]{fig/bcdscalarblocks}\label{fig:bcdscalarblocks}}
%\end{figure*}

\subsection{Multiplicative Update (\MU)}
\label{sec:sparsenmfmu}

In this paper, we are considering $\ell_2$ regularization on the $\WW$ matrix and $\ell_1$ regularization on the $\HH$ matrix to address the sparsity of the input matrix. Hence, the NMF problem becomes
 \SplitN{\label{eqn:original NMF}}{
\min_{\WW \geq 0,\HH \geq 0} & \|\AA-\WW\HH\|_F + \alpha (\| \WW \|)_F^2 + \beta \sum_{i=1}^n \| \hh_{i} \|_1^2.
}

The values $\alpha$ and $\beta$ were fixed for the experiments. 
In the case of \MU\ \cite{seung2001algorithms}, individual entries of $\WW$ and $\HH$ are updated with all other entries fixed.
In this case, the update rules are 
\SplitN{\label{eqn:muupdate}} {
w_{ij} &\leftarrow w_{ij} \frac{(\AA \HH^T)_{ij}}{(\WW (\HH \HH^T+ 2\beta \mathbf{1}_k))_{ij}}, \text{ and }\\
h_{ij} &\leftarrow  h_{ij} \frac{(\WW^T \AA)_{ij}}{((\WW^T \WW+ 2\alpha\mathbf{I}_k) \HH)_{ij}}.
} 
where $\mathbf{1}_k$ is a matrix of $k \times k$ with all one's and $\mathbf{I}_k$ is an identity matrix of size $k \times k$.

After computing the Gram matrices $\HH\HH^T$ and $\WW^T \WW$, adding the appropriate regularizers and the products $\AA\HH^T$ and $\WW^T\AA$, the extra cost of computing $\WW (\HH \HH^T+ 2\beta \mathbf{1}_k)$ and $(\WW^T \WW+ 2\alpha\mathbf{I}_k)$ is $F(m,n,k)=2(m+n)k^2$ flops to perform updates for all entries of $\WW$ and $\HH$, as the other elementwise operations affect only lower-order terms.
The details about using AU-NMF in \cref{alg:aunmf} for other algorithms \HALS\ and \BPP\ are explained in~\cite{KBP16,KBP16MPIFAUN}.

The above update equation \cref{eqn:muupdate} can be easily parallelized using the \cref{alg:distnmf}.

%To begin with let us consider the 1D row distribution of the the matrix $\AA$, where each process hold $\frac{m}{p} \times n$ represented as $\AA^i$ and the entire $\HH \in \Rnplus{k \times n}$. Every  $\AA^i\HH^T \in $

It is important to observe that the update function of $\WW$ is element-wise normalization of the sparse matrix-dense matrix multiplication $\AA \HH^T$ with the denominator. Given that if all the processes owns the $k \times k$ gram of the factor matrix $\HH$, computing the entire denominator $(\WW (\HH \HH^T+ 2\beta \mathbf{1}_k))$ is does not require any communication, hence can be done locally.
%We can see that the \cref{line:hmgr} and \cref{line:foldw} offers the local processes' $\AA \HH^T$.
One can argue that the same holds for updating $\HH$ as well.
Therefore, for row and column index sets $\Ip$ and $\Jp$, one can update these rows of the factor matrices as follows:

\SplitN{\label{eqn:muupdate}} {
\Wm(\Ip, :) &\leftarrow \Wm(\Ip, :) \circledast (\Wmt(\Ip, :)) \oslash (\Wm(\Ip, :) (\Hmgr+ 2\beta \mathbf{1}_k)), \text{ and }\\
\Hm(\Jp, :) &\leftarrow  \Hm(\Jp, :) \circledast (\Hmt(\Jp, :) \oslash (\Wmgr+ 2\alpha\mathbf{I}_k) \Hm(\Jp, :)).
}
where $\circledast$ and $\oslash$ correspond to element-wise multiplication and division of matrices or vectors.
This scheme provides row-wise parallelism in NNLS computation.
\HALS\ and \BPP can similarly be expressed in this row-parallel form.

%\todo{\ramki{ramki can you explain that we can row-wise parallelize this computation? Specificaly, we need to say that $\Wm(I, :)$ can be updated using $\Hm \Hm^T$ and the submatrix $[\Am \Hm] (I, :)$}}

%Apart from the proposed \distspnmf Algorithm \ref{alg:distnmf}, we have to extend the \mpifaun framework for $2D$ partition strategy in the following directions (a) Additional sparse regularization which otherwise will result in numerical in-stability for large high-dimensional sparse matrices (b) The intermediate memory requirement for \mpifaun restricts its ability to handle sparse matrices of few hundred millions and more in smaller processor and we have to extend blocking support on the low rank side.

### 2018-04-03 latex/icpp18/distspnmf-reproducibility-ae.tex
seeding icpp18

% LaTeX template for Artifact Description appendix used for SC17
% V20170220
%% V20170220
% (C)opyright 2017

% Derived with permission by Michael Heroux (Sandia National Laboratories, St. John's University, MN)
% from ae-20160509.tex 
% written by Grigori Fursin (cTuning foundation, France and dividiti, UK) 
% and Bruce Childers (University of Pittsburgh, USA)
% (C)opyright 2014-2016

% sigplanconf.cls is available at http://www.sigplan.org/sites/default/files/sigplanconf.cls
\documentclass[sigconf, review=false]{acmart}

\usepackage{url}

\newcommand{\distspnmffull}{Distributed memory Sparse NMF algorithm}

\begin{document}

\special{papersize=8.5in,11in}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% When adding this appendix to your paper, 
% please remove above part
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix
\section{Artifact Description: \distspnmffull}

Submission and reviewing guidelines and methodology: \\
{\small\em http://sc17.supercomputing.org/submitters/technical-papers/reproducibility-initiatives-for-technical-papers/artifact-description-paper-title/}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Abstract}

This artifact contains all the components and instruction for the paper titled \distspnmffull\ submitted to SC`2017.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Description}

\subsubsection{Check-list (artifact meta information)}

{\small
\begin{itemize}
  \item {\bf Algorithm: } \distspnmffull for Multiplicative Update(MU), Hierarchical Alternating Least Squares(HALS) and ANLS/BPP supporting multi constrained hypergraph partitions 
  \item {\bf Program: } C/C++ code
  \item {\bf Compilation: } C/C++ compiler with C++11 support. We tested with GCC 5.3.0 compiler. 
  \item {\bf Transformations: }
  \item {\bf Binary: } distnmf
  \item {\bf Data set: } Flickr and Delicious dataset. Publicly available. 
  \item {\bf Run-time environment: } MPI on Unix
  \item {\bf Hardware: } Rhea on Oak Ridge Leadership Computing Platform and IBM BlueGene/Q supercomputer
  \item {\bf Run-time state: } 
  \item {\bf Execution: } Rhea execution - mpirun -np 16 ./distnmf \-a [0/1/2 for mu/hals/bpp] \-\-pr=[number of row processes] \-\-pc=[number of col processes] \-k [low rank value] \-i [input file] \-t [number of iterations] \\
BGQ execution -  runjob \-\-ranks-per-node 16 \-\-np 16 : /linkhome/rech/qyy/rqyy003/nmflibrary/mpi/distnmf \-a 0 \-i /linkhome/rech/qyy/rqyy003/job-run/flickr.mtx -t 10 -k 16 \\
  \item {\bf Output: }
total\_allgather::m::17261568::n::2473984::k::48::SIZE::1536::algo::0::root::13.66::min::13.36::avg::13.656::max::13.79 \\
total\_allreduce::m::17261568::n::2473984::k::48::SIZE::1536::algo::0::root::0.1::min::0.03::avg::0.106354::max::0.21 \\
total\_reducescatter::m::17261568::n::2473984::k::48::SIZE::1536::algo::0::root::21.99::min::21.8::avg::22.0041::max::22.18 \\
total\_gram::m::17261568::n::2473984::k::48::SIZE::1536::algo::0::root::1.48::min::1.37::avg::1.47555::max::1.64 \\
total\_mm::m::17261568::n::2473984::k::48::SIZE::1536::algo::0::root::10.04::min::9.96::avg::10.0897::max::10.23 \\
total\_nnls::m::17261568::n::2473984::k::48::SIZE::1536::algo::0::root::7.95::min::7.64::avg::7.93462::max::8.05 \\
  \item {\bf Experiment workflow: }
  \begin{enumerate}
\item   git clone source code
\item download dependencies
\item build source code
\item download datasets
\item prepare data
\item create scheduler scripts
\item queue the job
\item analyze the output.
\end{enumerate}
  \item {\bf Experiment customization: } Yes, the software is general
purpose.
  \item {\bf Publicly available?: } Yes.
\end{itemize}
}

\subsubsection{How software can be obtained (if available):} 
Github/Gitlab repository. The clone URL can reveal the person's identity and can violate the double blind review process.

\subsubsection{Hardware dependencies}. \\

We have tested on two different supercomputers Rhea on Oak Ridge Leadership Computing Facility and IBM Bluegene/Q supercomputer.

\subsubsection{Software dependencies} \label{sec:dependencies}

\begin{itemize}
\item GCC 5.3.0 and above with support for C++11
\item CMake
\item BLAS/LAPACK libraries - Tested w/ MKL 16.0.0, OpenBLAS (https://github.com/xianyi/OpenBLAS), libblas libraries. 
\item Boost 1.58.0  and above (Needed only for ANLS/BPP algorithm)
\item Armadillo - http://arma.sourceforge.net/
\item MPI Library - tested w/ OpenMPI 1.8.4 and Bluegene/Q MPI library
\end{itemize}

\subsubsection{Datasets} \label{sec:datasets}

Delicious - \url{https://www.dropbox.com/s/3602tin4g59n8vl/delicious.mtx.bz2?dl=0}
Flickr - \url{https://www.dropbox.com/s/r9xn19vhq0r0dvi/flickr.tar.bz2?dl=0}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Installation}

Follow the build instruction in README.md

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment workflow}

\begin{enumerate}
\item git clone source code (DistSPNMF, pacoss, tmpi)
\item download dependencies listed in Section \ref{sec:dependencies}
\item build source code following the README.md file
\item download datasets listed in Section \ref{sec:datasets}
\item prepare data - {\em pacoss partition} and {\em pacoss distribute}
\item create scheduler scripts - PBS/Torque scripts
\item queue the job 
\item analyze the output.
\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Evaluation and expected result}

As explained in the paper experiment section, the $P2PHP$ and $P2PRP$ will provide better running time over $FAUN$ in higher number of processors. The output time can be parsed as explained in output above.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Experiment customization}

The software is implemented by extending the $MPI-FAUN$ library and the distribution strategies are available though $pacoss$ software. Users can place with the options available in these software to trying different NMF algorithms and different partition methods.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Notes}

The entire software will be available for users through github.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% When adding this appendix to your paper, 
% please remove below part
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2018-04-03 latex/icpp18/prelim.tex
seeding icpp18

\section{Preliminaries}
\input{notations}

\subsection{Communication model}
\label{sec:comm-model}

To analyze our algorithms, we use the $\alpha$-$\beta$-$\gamma$ model of distributed-memory parallel computation.
In this model, interprocessor communication occurs in the form of messages sent between two processors across a bidirectional link (we assume a fully connected network).
We model the cost of a message of size $n$ words as $\alpha+n\beta$, where $\alpha$ is the per-message latency cost and $\beta$ is the per-word bandwidth cost.
Each processor can compute floating point operations (flops) on data that resides in its local memory; $\gamma$ is the per-flop computation cost.
With this communication model, we can predict the performance of an algorithm in terms of the number of flops it performs as well as the number of words and messages it communicates.
For simplicity, we will ignore the possibilities of overlapping computation with communication in our analysis.
For more details on the $\alpha$-$\beta$-$\gamma$ model, see \cite{TRG05,CH+07}. Since one of the major objective of this paper is comparing the communications strategies between different distribution strategies -- point-to-point communication for $NNZ-PART$ and the collective communications for $2D$. For the collective communication for $2D$, we are observing the same communication model of the \mpifaun algorithm \cite{KBP16, KBP16MPIFAUN}. We present the communication model for P2P here.

\subsection{Point-to-Point Communication}

\todo{\oguz{fill the point to point communication on pacoss}}

%\subsection{MPI collectives}
%\label{sec:collectives}
%
%Point-to-point messages can be organized into collective communication operations that involve more than two processors.
%MPI provides an interface to the most commonly used collectives like broadcast, reduce, and gather, as the algorithms for these collectives can be optimized for particular network topologies and processor characteristics.
%The baseline algorithms $MPIFAUN$ use the all-gather, reduce-scatter, and all-reduce collectives and so we review them here, along with their costs.
%%Our analysis assumes optimal collective algorithms are used (see \cite{TRG05,CH+07}), though our implementation relies on the underlying MPI implementation.
%
%At the start of an all-gather collective, each of $p$ processors owns data of size $n/p$.
%After the all-gather, each processor owns a copy of the entire data of size $n$.
%The cost of an all-gather is $\alpha\cdot \log p + \beta \cdot \frac{p-1}{p}n$.
%%
%At the start of a reduce-scatter collective, each processor owns data of size $n$.
%After the reduce-scatter, each processor owns a subset of the sum over all data, which is of size $n/p$.
%(Note that the reduction can be computed with other associative operators besides addition.)
%The cost of an reduce-scatter is $\alpha\cdot \log p + (\beta+\gamma) \cdot \frac{p-1}{p}n$.
%%
%At the start of an all-reduce collective, each processor owns data of size $n$.
%After the all-reduce, each processor owns a copy of the sum over all data, which is also of size $n$.
%The cost of an all-reduce is $2\alpha\cdot \log p + (2\beta+\gamma) \cdot \frac{p-1}{p}n$.
%%
%Note that the costs of each of the collectives are zero when $p=1$.

\input{aunmf}
%\input{hypergraph}

### 2018-04-03 latex/icpp18/abstract.tex
new draft of abstract and intro

--- 2018-04-02 latex/sc17/abstract.tex
@@
+%!TEX root = icpp18.tex
+
 \begin{abstract}
@@
 
+\grey{ Next draft of abstract...
+Non-negative matrix factorization (NMF), the problem of finding two non-negative low-rank factors whose product approximates an input matrix, is a useful tool for many data mining and scientific applications such as topic modeling in text mining and blind source separation in microscopy.
+In this paper, we focus on scaling algorithms for NMF to very large sparse datasets and massively parallel machines by employing effective algorithms, communication patterns, and partitioning schemes that leverage the sparsity of the input matrix.
+In particular, we consider both a fine-grained, point-to-point communication pattern and a coarse-grain, collective-based communication pattern, and we use both hypergraph-based and randomized partitioning strategies.
+We characterize, both analytically and empirically, the classes of problems for which each combination of communication pattern and partitioning strategy is most advantageous.
+We present experiments on (THESE DATASETS), running on up to (YY) cores of a Cray XK7 supercomputer.
+(MAYBE STATE SIMPLE VERSION OF CONCLUSIONS?)
+}
+
 %We propose a fine-grain communication scheme that can utilize arbitrary partitions of the matrices.

### 2018-04-03 latex/icpp18/aunmf.tex
new draft of abstract and intro

@@
+%!TEX root = icpp18.tex
+
 \section{Sparse Non-negative Matrix Factorization}
 \label{sec:sparsenmf}
+
+\grey{Not sure where the following paragraph should be incorporated:
+We would like to highlight that ``Non-negative'' Matrix Factorization is NOT matrix factorization in collaborative filtering for recommender systems for which many implementations exist.
+The collaborative filtering problem is different than NMF -- the focus of our paper -- because it interprets the ``{\em zero}'' entries of the input matrix as missing data, while the NMF problem is defined for a completely known input matrix and does not handle missing values.
+This leads to different optimization problems, algorithms, and computational steps.
+}
+
+\subsection{Notation}
+\label{sec:notations}
+
+\Cref{tab:notation} summarizes the notation we use throughout this paper.
+We use bold uppercase letters for matrices and lowercase letters for vectors.
+For matrix rows and columns, we employ MATLAB notation, i.e., $\AA(i, :)$ and $\AA(:, j)$ refer to the $i$th row and the $j$th column of $\AA$.
+We use subscripts to refer to sub-blocks of matrices.
+For example, $\AA_ij$ refers to the sub-block $(i, j)$ of $\AA$ in a 2D partition.
+We use $m$ and $n$ to denote the numbers of rows and columns of $\AA$, respectively, and assume without loss of generality $m\geq n$ throughout.
+
+\begin{table}%[htdp]
+\begin{center}
+\begin{tabular}{|l|l|}
+\hline
+$\AA$ & Input matrix \\
+$\WW$ & Left low rank factor \\
+$\HH$ & Right low rank factor \\
+$m$ & Number of rows of input matrix \\
+$n$ & Number of columns of input matrix \\
+$k$ & Low rank \\
+$P$ & Number of parallel processes \\
+$P_r$ & Number of rows in processor grid \\
+$P_c$ & Number of columns in processor grid \\
+$\Ip$,$\Jp$ & Set of rows/columns of of $\Wm$/$\Hm$ owned by process $p$  \\
+$\Fp$,$\Gp$ & Set of unique row and column indices of $\Amp$ \\
+$\Amp$ & Submatrix of $\Am$ owned by process p \\
+$\Wm(\Ip,:)$ & Owned rows of initial $\Wm$ by process p \\
+$\Hm(:,\Jp)$ & Owned columns of initial $\Hm$ by process p\\
+
+\hline
+\end{tabular}
+\end{center}
+\caption{Notation}
+\label{tab:notation}
+\end{table}%
+
+\subsection{Alternating-Updating NMF}
 
@@
  \SplitN{\label{eqn:original NMF}}{
-\min_{\WW \geq 0,\HH \geq 0} & \|\AA-\WW\HH\|_F + \alpha (\| \WW \|)_F^2 + \beta \sum_{i=1}^n \| \hh_{i} \|_1^2.
+\min_{\WW \geq 0,\HH \geq 0} & \|\AA-\WW\HH\|_F + \alpha \| \WW \|_F^2 + \beta \sum_{i=1}^n \| \hh_{i} \|_1^2.
 }

### 2018-04-03 latex/icpp18/introduction.tex
new draft of abstract and intro

--- 2018-04-02 sc17/introduction.tex
@@
+%!TEX root = icpp18.tex
+
 \section{Introduction}
@@
 Here, $\Rnplus{m\times n}$ denotes the set of $m \times n$ matrices with non-negative real values.
-Formally, the NMF problem \cite{seung2001algorithms} can be defined as \SplitN{\label{eqn:original NMF}}{
-\min_{\WW \geq 0,\HH \geq 0} & \|\AA-\WW\HH\|_F + \phi (\WW) + \psi (\HH),
+Formally, the NMF problem can be defined as 
+\SplitN{\label{eqn:origNMF}}{
+\min_{\WW \geq 0,\HH \geq 0} & \|\AA-\WW\HH\|_F,
 }
 where $\|\M{X}\|_F=(\sum_{ij} x_{ij}^2)^{1/2}$ is the Frobenius norm. 
-The functions $\phi (\WW)$ and $\psi (\HH)$ are called regularization functions that avoids the model from over fitting on the data and are chosen depending on the characteristics of the data.
-For this paper, we are considering $\phi(\WW) = \alpha \|\WW\|_F^2$ called as $\ell_2$ regularizer and $\psi(\HH) = \beta \sum_{i=1}^n \| \hh_{i} \|_1^2$ called as $\ell_1$ regularizer for addressing the inherent sparsity in the input matrix.
-The lack of $\ell_1$ regularization on the matrix $\HH$ can result in numerical instability.
+%The functions $\phi (\WW)$ and $\psi (\HH)$ are called regularization functions that avoid the model from over-fitting the data and are chosen depending on the characteristics of the data. 
+%In this work, because we assume $\AA$ to be sparse, we use $\phi(\WW) = \alpha \|\WW\|_F^2$ (an $\ell_2$ regularizer) and $\psi(\HH) = \beta \sum_{i=1}^n \| \hh_{i} \|_1^2$ (an $\ell_1$ regularizer) to address the sparsity in the input matrix.
+%The lack of $\ell_1$ regularization on the matrix $\HH$ can result in numerical instability.
 
@@
 The applications of NMF range from text mining \cite{pauca2004text},  computer vision \cite{hoyer2004non}, and bioinformatics \cite{kim2007sparse} to blind source separation  \cite{cichocki2009nonnegative}, unsupervised clustering \cite{kuang2012symmetric,kuang2013symnmf}  and many other areas.
-In most real-world applications $m$ and $n$ can be on the order of millions or more while $k$ being much smaller in the order of tens to thousands.
-
-In the recent years, there has been an explosion in the size of collected data particularly with the widespread use of internet and social media.
-Such collected data has big dimensions~(from millions to billions) and is extremely sparse, which is either implied by the nature of the data~(e.g., most web content are not visited by users), or because capturing every possible data point is impractical.
-Analyzing such sparse big datasets require enormous computational power as well as efficient algorithms that can leverage the sparsity of the data.
+In most real-world applications $m$ and $n$ are on the order of millions or more while $k$ is much smaller, on the order of tens to thousands.
+Furthermore, data sets from these applications are often quite sparse and have highly irregular nonzero patterns.
 NMF is an effective tool for analyzing such data in many applications, yet there is a lack of high performance algorithms and software that can compute it efficiently.
 Our goal is in this paper is to fill in this gap by providing scalable parallel algorithms that employ efficient communication schemes and can scale to thousands of processes.
-Thereby, we aim to enable analyzing data in such big data applications.
+
+%In the recent years, there has been an explosion in the size of collected data particularly with the widespread use of internet and social media.
+%Such collected data has big dimensions~(from millions to billions) and is extremely sparse, which is either implied by the nature of the data~(e.g., most web content are not visited by users), or because capturing every possible data point is impractical.
+%Analyzing such sparse big datasets require enormous computational power as well as efficient algorithms that can leverage the sparsity of the data.
+
+%Thereby, we aim to enable analyzing data in such big data applications.
 
 % How NMF is computed, and in sparse what is important.
-We would like to highlight that ``Non-negative'' Matrix Factorization is NOT matrix factorization in collaborative filtering for recommender systems for which many implementations exist.
-The collaborative filtering problem is different than NMF -- the focus of our paper -- because it interprets the ``{\em zero}'' entries of the input matrix as missing data, while the NMF problem is defined for a completely known input matrix and does not handle missing values.
-This leads to different optimization problems, algorithms, and computational steps.
-Specifically, computing NMF involves three main types of operations; multiplication of $\Am$ with a factor matrix $\Wm$ or $\Hm$, computing the Gram matrices $\Wm^T \Wm$ and $\Hm \Hm^T$, and solving a non-linear least squares~(NNLS) problem using these two resulting dense matrices to update factor matrices.
-To the best of our knowledge, the only high performance software available for parallelizing this computation is \mpifaun~\cite{KBP16MPIFAUN} which operates both on sparse and dense input matrices.
-For parallelism, it employs a fixed 2D uniform partitioning on $\Am$, and partitions factor matrices $\Wm$ and $\Hm$ conformally with this 2D partition.
-Each process, works on its local matrix block of $\Am$, and communicates the rows and columns of $\Wm$ and $\Hm$ corresponding to its process row and column in the 2D topology.
-This communication is optimal in the dense case; i.e., no process receives a data element not used in its local computations.
-In the sparse case, however, the actual communication requirements of processes depend on the sparsity of $\Am$ and its partitioning, and is in general significantly less than the dense case.
-Therefore, employing the same all-to-all communication scheme in the sparse case brings about a major redundancy in the communication cost.
-Also, \mpifaun similarly uses only fixed 2D uniform partitioning on sparse matrices; this creates significant imbalance both in computation and communication as typically the data is not homogenously distributed in $\Am$.
-In overall, the lack of an efficient sparse communication scheme that allows flexibility in partitioning hinders the scalability of this approach.
+The most common method for solving \cref{eqn:origNMF} is to use an alternating optimization approach, iteratively updating $\WW$ with $\HH$ fixed and then updating $\HH$ with $\WW$ fixed.
+Specifically, updating a factor matrix involves three main operations; computing of $\WW^T\AA$ or $\AA\Hm^T$, computing the Gram matrix $\Wm^T \Wm$ or $\Hm \Hm^T$, and solving a non-negative linear least squares~(NNLS) problem using these two resulting matrices to update the factor matrix.
+The typical bottleneck is the computation that involves the input matrix, $\WW^T\AA$ and $\AA\Hm^T$, which are sparse-dense matrix multiplications (SpMMs) when the input data is sparse.
+
+The focus in this work is the comparison and combination of two previous approaches that can be applied to sparse NMF problems.
+The first approach is based on MPI-FAUN \cite{KBP17}, a software framework for computing the NMF of dense and sparse data.
+The second approach is based on HyperTensor \cite{KU15}, a library for computing a CP decomposition of a sparse tensor.
+In the former case, the algorithm is optimized for dense input matrices, but it can be used for sparse data.
+In the latter case, the algorithm is optimized for tensor data and unconstrained linear least squares problem, but it can be applied to matrices (2D tensors) and combined with NLS updates.
+
+MPI-FAUN uses a regular, Cartesian partitioning of the input matrix $\AA$ across processors that is oblivious to the nonzero pattern.
+The advantages of this approach are that the resulting factor matrix communication (to compute the SpMMs) is also regular and cast as low-latency MPI collective operations and that the rest of the computation (including NLS updates) is load-balanced.
+The disadvantage is that because the partition ignores the sparsity of $\AA$, the approach will often communicate more data than necessary, as some processors will receive data that they do not need for local computation.
+
+HyperTensor solves an input-sparsity-dependent hypergraph partitioning problem to determine the parallel distribution of the entries of $\AA$.
+The advantage of this approach is that by being sensitive to the nonzero pattern, the resulting factor matrix communication required to compute the SpMMs is (approximately) minimized, and the only data that a processor receives is what it needs for its local computation.
+The disadvantages of this approach are an upfront partitioning cost (which can be amortized over the iterations of the NMF algorithm) and possible load imbalance, in terms of the rest of the computation and the number of messages communicated among processors.
+
+The two methods differ both in their partitioning strategy and in their communication pattern.
+We also consider a compromise approach, which uses a randomized partitioning scheme, to achieve load balance without perfectly minimizing communication, and a sparsity-sensitive communication pattern, which avoids communicating unnecessary data.
+We show that this compromise avoids heavy upfront partitioning costs and effectively balances the efficiency constraints of the different NMF computations.
+
+%To the best of our knowledge, the only high performance software available for parallelizing this computation is \mpifaun~\cite{KBP16MPIFAUN} which operates both on sparse and dense input matrices.
+%For parallelism, it employs a fixed 2D uniform partitioning on $\Am$, and partitions factor matrices $\Wm$ and $\Hm$ conformally with this 2D partition.
+%Each process, works on its local matrix block of $\Am$, and communicates the rows and columns of $\Wm$ and $\Hm$ corresponding to its process row and column in the 2D topology.
+%This communication is optimal in the dense case; i.e., no process receives a data element not used in its local computations.
+%In the sparse case, however, the actual communication requirements of processes depend on the sparsity of $\Am$ and its partitioning, and is in general significantly less than the dense case.
+%Therefore, employing the same all-to-all communication scheme in the sparse case brings about a major redundancy in the communication cost.
+%Also, \mpifaun similarly uses only fixed 2D uniform partitioning on sparse matrices; this creates significant imbalance both in computation and communication as typically the data is not homogenously distributed in $\Am$.
+%In overall, the lack of an efficient sparse communication scheme that allows flexibility in partitioning hinders the scalability of this approach.
 
@@
 
+\grey{I have not edited the following contribution list yet, we may want to wait until we have new data...
 We summarize our main contributions as follows:
@@
 \end{itemize}
+}

### 2018-04-04 latex/icpp18/abstract.tex
made updates to abstract and intro based on comments from ramki and oguz

--- 2018-04-02 latex/sc17/abstract.tex
@@
+%!TEX root = icpp18.tex
+
 \begin{abstract}
-Non-negative matrix factorization (NMF) is the problem of finding two non-negative low rank factors $\WW$ and $\HH$ for a given input matrix $\AA$ such that $\AA \approx \WW \HH$.
-NMF is a useful tool for many internet applications such as topic modeling in text mining and community detection in social networks.
-In this paper, we focus on scaling distributed NMF to very large sparse datasets by employing effective algorithms, communications, and partitioning schemes that leverage the sparsity of the input matrix $\AA$.
-Our method employs an optimal point-to-point communication scheme, and removes any redundant communication existing in the state of the art.
-Unlike the state of the art, our method also permits arbitrary partitions of the sparse input matrix as well as factor matrices, which enables us to employ effective partitioning schemes for better scalability.
-We experiment on a cluster using up to 3072 cores, and report significant better scalability with respect to existing methods owing to our efficient communication and partitioning strategies.
-We also run our algorithms on an IBM BlueGene/Q supercomputer, and achieve scalability up to 32768 processors with a 32x speedup over the execution using 256 processors.
+Non-negative matrix factorization (NMF), the problem of finding two non-negative low-rank factors whose product approximates an input matrix, is a useful tool for many data mining and scientific applications such as topic modeling in text mining and blind source separation in microscopy.
+In this paper, we focus on scaling algorithms for NMF to very large sparse datasets and massively parallel machines by employing effective algorithms, communication patterns, and partitioning schemes that leverage the sparsity of the input matrix.
+In particular, we consider two previous works developed for related problems, one that uses a fine-grained partitioning strategy using a point-to-point communication pattern and on that uses a coarse-grained partitioning strategy using a collective-based communication pattern.
+We show that a combination of the previous approaches balances the demands of the various computations within NMF algorithms and achieves high efficiency and scalability.
+\grey{To be completed...
+
+We present experiments on (THESE DATASETS), running on up to (YY) cores of a Cray XK7 supercomputer.
+(MAYBE STATE SIMPLE VERSION OF CONCLUSIONS?)
+}
 

### 2018-04-04 latex/icpp18/introduction.tex
made updates to abstract and intro based on comments from ramki and oguz

@@
 The most common method for solving \cref{eqn:origNMF} is to use an alternating optimization approach, iteratively updating $\WW$ with $\HH$ fixed and then updating $\HH$ with $\WW$ fixed.
-Specifically, updating a factor matrix involves three main operations; computing of $\WW^T\AA$ or $\AA\Hm^T$, computing the Gram matrix $\Wm^T \Wm$ or $\Hm \Hm^T$, and solving a non-negative linear least squares~(NNLS) problem using these two resulting matrices to update the factor matrix.
-The typical bottleneck is the computation that involves the input matrix, $\WW^T\AA$ and $\AA\Hm^T$, which are sparse-dense matrix multiplications (SpMMs) when the input data is sparse.
+Specifically, updating a factor matrix involves three main operations; computing of $\WW^T\AA$ or $\AA\Hm^T$, computing the Gram matrix $\Wm^T \Wm$ or $\Hm \Hm^T$, and solving a non-negative linear least squares (NLS) problem using these two resulting matrices to update the factor matrix.
+When the rank $k$ is small, the typical bottleneck is the computation that involves the input matrix, $\WW^T\AA$ and $\AA\Hm^T$, which are sparse-dense matrix multiplications (SpMMs) when the input data is sparse.
+However, for large $k$, the Gram and NLS computations can become the bottleneck, as they grow more quickly with $k$ than the SpMMs.
+Efficient parallel algorithms for NMF must load balance the computation and avoid communication overheads.
+The distribution of the input $\AA$ across processors affects the load balance and communication of the SpMMs, and the distribution of the factor matrices $\WW$ and $\HH$ affects the load balance of the Gram and NLS computations.
 
@@
 The advantage of this approach is that by being sensitive to the nonzero pattern, the resulting factor matrix communication required to compute the SpMMs is (approximately) minimized, and the only data that a processor receives is what it needs for its local computation.
-The disadvantages of this approach are an upfront partitioning cost (which can be amortized over the iterations of the NMF algorithm) and possible load imbalance, in terms of the rest of the computation and the number of messages communicated among processors.
+The disadvantages of this approach are an upfront partitioning cost (which can be amortized over the iterations of the NMF algorithm) and possible load imbalance, particularly in the NLS computations.
 

### 2018-04-07 distnmf/disthals.hpp
pacoss compiles. some runtime error

@@ -94,6 +94,7 @@ public:
            const int numkblks) :
     DistAUNMF<INPUTMATTYPE>(input, leftlowrankfactor,
                             rightlowrankfactor, communicator, numkblks) {
+      PRINTROOT("DistHALS() constructor successful");
   }
 };
 

### 2018-04-07 distnmf/distmu.hpp

@@ -82,6 +82,7 @@ class DistMU : public DistAUNMF<INPUTMATTYPE>{
     HWtW.zeros(this->globaln() / this->m_mpicomm.size(), this->k);
     localWnorm.zeros(this->k);
     Wnorm.zeros(this->k);
+    PRINTROOT("DistMU() constructor successful");
   }
 };
 #endif  // MPI_DISTMU_HPP_

### 2018-04-07 distnmf/distnmf.cpp

@@ -119,6 +119,8 @@ class DistNMFDriver {
         std::string dim_part_file_name = this->m_Afile_name;
         dim_part_file_name += ".dpart.part" + std::to_string(mpicomm.rank());
         this->m_Afile_name += ".part" + std::to_string(mpicomm.rank());
+        INFO << mpicomm.rank() << ":: part_file_name::" << dim_part_file_name
+             << "::m_Afile_name::" << this->m_Afile_name << std::endl;
         Pacoss_SparseStruct<double> ss;
         ss.load(m_Afile_name.c_str());
         std::vector<std::vector<Pacoss_IntPair> > dim_part;

### 2018-04-07 pacoss/src/cpp-utils.cpp

@@ -41,7 +41,7 @@ void _fclosee(
 std::chrono::time_point<std::chrono::high_resolution_clock> timerTic()
 {
   tic_global = std::chrono::high_resolution_clock::now();
-  return std::chrono::high_resolution_clock::now();
+  return tic_global;
 }
 
 double timerToc(const std::chrono::time_point<std::chrono::high_resolution_clock> &ticPoint)

### 2018-04-07 pacoss/src/cpp-utils.h

@@ -6,6 +6,8 @@
 #include <cstring>
 #include <exception>
 #include <chrono>
+#include <vector>
+#include <iostream>
 
 #define CppUtils_Error(msg) _CppUtils_Error(msg, __LINE__, __FILE__)
 
@@ -161,6 +163,18 @@ const char * STRCAT(T t, Args... args) // recursive variadic function
   strcpy(STRCAT_STR, t);
   strcat(STRCAT_STR, str);
   return STRCAT_STR;
+}
+
+template<class T>
+static void printVector(std::vector<T> vec) {
+  for (auto &i : vec) { std::cout << i << " "; }
+  std::cout << std::endl;
+}
+
+template<class T>
+static void printlnVector(std::vector<T> vec) {
+  for (auto &i : vec) { std::cout << i << " "; }
+  std::cout << std::endl;
 }
 
 // Timer tic-toc routines

### 2018-04-07 pacoss/src/exe/pacoss.cpp

@@ -7,19 +7,22 @@ char *sparseStructFileName;
 char *dimPartitionFileName;
 char *nzPartitionFileName;
 const char *nzPartitionMethod = "fine-random";
-const char *dimPartitionMethod = "balance-comm";
+const char *dimPartitionMethod = "balanced";
 const char *patohSettings = "default";
+std::vector<Pacoss_Int> procGridDims;
 char dimPartitionFileNameDefault[1000]; 
 char nzPartitionFileNameDefault[1000];
 char *command;
 bool oneBasedIdx = true;
+std::vector<std::vector<Pacoss_Int>> partCtorIdx;
+std::vector<double> maxRowImbalance;
 
 void printUsage()
 {
   printfe("Usage: ./pacoss -c command command-args\n");
-  printfe("  command: partition | distribute | check-stats | convert-sparse-struct\n");
+  printfe("  command: partition | partition-dims | distribute | check-stats | convert-sparse-struct | fix-sparse-struct\n");
   printfe("  partition-args: -f sparseStructFileName -m nzPartitionMethod -n dimPartitionMethod -p numParts [-z"
-      " nzPartitionFileName -d dimPartitionFileName -q numColParts -s patohSettings]\n");
+      " nzPartitionFileName -d dimPartitionFileName -q numColParts -s patohSettings -i maxRowImbalance]\n");
   printfe("  distribute-args: -f sparseStructFileName [-z nzPartitionFileName -d dimPartitionFileName]\n");
   printfe("  check-stats-args: -f sparseStructFileName [-z nzPartitionFileName -d dimPartitionFileName]\n");
   printfe("  convert-sparse-struct-args: -f sparseStructFileName [--one-based-idx]\n");
@@ -27,15 +30,15 @@ void printUsage()
   printfe("    fine-random(default): Fine-grain random partitioning of sparse structure nonzeros.\n");
   printfe("    fine-patoh: Fine-grain hypergraph partitioning using PaToH.\n");
   printfe("    checkerboard-random: Random checkerboard partitioning.\n");
+  printfe("    checkerboard-uniform: Uniform checkerboard partitioning.\n");
   printfe("    checkerboard-patoh-1d: Checkerboard partitioning using PaToH for row partitioning, and"
[…]

### 2018-04-07 pacoss/src/keyvalsorter.h

@@ -151,9 +151,9 @@ static size_t partition(
   if (left == right) { return left; }
   size_t pivot = left, i = left + 1, j = right;
   size_t order = key.size();
-  size_t randIdx1 = rand() % (right - left + 1) + left;
-  size_t randIdx2 = rand() % (right - left + 1) + left;
-  size_t randIdx3 = rand() % (right - left + 1) + left;
+  size_t randIdx1 = (size_t) rand() % (right - left + 1) + left;
+  size_t randIdx2 = (size_t) rand() % (right - left + 1) + left;
+  size_t randIdx3 = (size_t) rand() % (right - left + 1) + left;
   randIdx1 = findMedian3(randIdx1, randIdx2, randIdx3, key);
 
   // Swap the random pivot to the very left

### 2018-04-07 pacoss/src/pacoss-communicator.cpp

@@ -408,7 +408,7 @@ void Pacoss_Communicator<DataType>::gatherData(
     Pacoss_Int root)
 {
   Pacoss_Int ownedCount = localOwnedRowCount();
-  Pacoss_Int totalRows;
+  Pacoss_Int totalRows = 0;
   Pacoss_IntVector rowIdx;
   Pacoss_IntVector recvCount; if (_procRank == root) { recvCount.resize(_numProcs); }
   Pacoss_IntVector recvBegin; if (_procRank == root) { recvBegin.resize(_numProcs + 1); }
@@ -452,7 +452,7 @@ void Pacoss_Communicator<DataType>::scatterData(
     Pacoss_Int root)
 {
   Pacoss_Int ownedCount = localOwnedRowCount();
-  Pacoss_Int totalRows;
+  Pacoss_Int totalRows = 0;
   Pacoss_IntVector rowIdx;
   Pacoss_IntVector recvCount; if (_procRank == root) { recvCount.resize(_numProcs); }
   Pacoss_IntVector recvBegin; if (_procRank == root) { recvBegin.resize(_numProcs + 1); }

### 2018-04-07 pacoss/src/pacoss-cpp-libs.h

@@ -2,6 +2,7 @@
 #define PACOSS_CPP_LIBS
 
 // C libraries
+#include <cassert>
 #include <cinttypes>
 #include <cstdio>
 #include <cstdlib>

### 2018-04-07 pacoss/src/pacoss-partitioner.cpp

@@ -20,6 +20,7 @@ void Pacoss_Partitioner<DataType, IntType, IdxType>::loadDimPart(
   for (auto &curDimPart : dimPart) {
     for (auto &i : curDimPart) {
       fscanfe(file, SCNMOD<IntType>(), &i);
+      i--; // Convert to 0-based
     }
   }
   fclosee(file);
@@ -37,7 +38,7 @@ void Pacoss_Partitioner<DataType, IntType, IdxType>::saveDimPart(
   fprintfe(file, "\n");
   for (auto &curDimPart : dimPart) {
     fprintfe(file, "\n");
-    for (auto i : curDimPart) { fprintfe(file, STRCAT(PRIMOD<IntType>(), "\n"), i); }
+    for (auto i : curDimPart) { fprintfe(file, STRCAT(PRIMOD<IntType>(), "\n"), i + 1); }
   }
   fclosee(file);
 }
@@ -51,7 +52,10 @@ void Pacoss_Partitioner<DataType, IntType, IdxType>::loadNzPart(
   IntType nzPartSize;
   fscanf(file, SCNMOD<IntType>(), &nzPartSize); 
   nzPart.resize(nzPartSize);
-  for (IntType i = 0; i < nzPartSize; i++) { fscanf(file, SCNMOD<IntType>(), &nzPart[i]); }
+  for (IntType i = 0; i < nzPartSize; i++) {
+    fscanf(file, SCNMOD<IntType>(), &nzPart[i]);
+    nzPart[i]--; // Convert to 0-based
+  }
   fclosee(file);
 }
 
@@ -62,7 +66,7 @@ void Pacoss_Partitioner<DataType, IntType, IdxType>::saveNzPart(
 {
   FILE *file = fopene(nzPartFileName, "w"); 
   fprintfe(file, "%zu\n\n", nzPart.size());
-  for (auto i : nzPart) { fprintfe(file, STRCAT(PRIMOD<IntType>(), "\n"), i); }
+  for (auto i : nzPart) { fprintfe(file, STRCAT(PRIMOD<IntType>(), "\n"), i + 1); }
   fclosee(file);
 }
 
@@ -85,12 +89,34 @@ void Pacoss_Partitioner<DataType, IntType, IdxType>::partitionDimsRandom(
 
 template <class DataType, class IntType, class IdxType>
 void Pacoss_Partitioner<DataType, IntType, IdxType>::partitionDimsBalanced(
-    const Pacoss_SparseStruct<DataType> &spStruct,
[…]

### 2018-04-07 pacoss/src/pacoss-partitioner.h

@@ -37,11 +37,12 @@ class Pacoss_Partitioner
         std::vector<std::vector<IntType>> &dimPart);
 
     static void partitionDimsBalanced(
-        const Pacoss_SparseStruct<DataType> &spStruct,
+        Pacoss_SparseStruct<DataType> &spStruct,
         const IntType numParts,
         const std::vector<IntType> &nzPart,
         std::vector<std::vector<IntType>> &dimPart,
-        double maxAllowedRowImbalance = 5.0);
+        std::vector<double> maxRowImbalance,
+        std::vector<std::vector<IntType>> procCtorIdx = std::vector<std::vector<IntType>>());
 
     // Partitioning routines for nonzeros
     static void partitionNzFineRandom(
@@ -53,7 +54,15 @@ class Pacoss_Partitioner
         const Pacoss_SparseStruct<DataType> &spStruct,
         const IntType numRowParts,
         const IntType numColParts,
-        std::vector<IntType> &nzPart);
+        std::vector<IntType> &nzPart,
+        std::vector<std::vector<IntType>> &partCtorIdx);
+
+    static void partitionNzCheckerboardUniform(
+        const Pacoss_SparseStruct<DataType> &spStruct,
+        const std::vector<IntType> &numParts,
+        std::vector<IntType> &nzPart,
+        std::vector<std::vector<IntType>> &dimPart,
+        bool randomize = true);
 
     static void partitionHypergraphFast(
         const int numVtx,
@@ -89,7 +98,8 @@ class Pacoss_Partitioner
         const IntType numColParts,
         const char * const patohSettings,
         const char * const colPartitionType,
-        std::vector<IntType> &nzPart);
+        std::vector<IntType> &nzPart,
+        std::vector<std::vector<IntType>> &partCtorIdx);
 #endif
 
     // Distribution routines

### 2018-04-07 pacoss/src/pacoss-sparse-struct.cpp

@@ -1,4 +1,5 @@
 #include "pacoss-sparse-struct.h"
+#include "keyvalsorter.h"
 
 template <class DataType>
 void Pacoss_SparseStruct<DataType>::load(const char * const fileName)
@@ -71,6 +72,40 @@ void Pacoss_SparseStruct<DataType>::form(
   }
   _val = val;
   if (_val.size() == 0) { _val.resize(_nnz, (DataType)1.0); }
+}
+
+template <class DataType>
+void Pacoss_SparseStruct<DataType>::fix()
+{
+  // Remove duplicates
+  KeyValSorter::sort(_idx, _val);   
+  Pacoss_Idx newNnz = 0;
+  for (Pacoss_Idx i = 0, ie = _nnz; i < ie; newNnz++) {
+    DataType newVal = 0.0;
+    while (i < _nnz && compKey(newNnz, i, _idx) == 0) { 
+      newVal += _val[i];
+      i++;
+    }
+    _val[newNnz] = newVal;
+    if (i < _nnz) {
+      for (Pacoss_Int j = 0; j < _order; j++) { _idx[j][newNnz + 1] = _idx[j][i]; }
+    }
+  }
+  _nnz = newNnz;
+  for (Pacoss_Int i = 0, ie = _order; i < ie; i++) { _idx[i].resize(_nnz); }
+  _val.resize(_nnz);
+  // Remove empty slices
+  for (Pacoss_Int dim = 0, dime = _order; dim < dime; dim++) {
+    // Detect empty slices
+    std::vector<Pacoss_Int> newLabel(_dimSize[dim]);
+    auto &idx = _idx[dim];
+    for (Pacoss_Idx i = 0, ie = _nnz; i < ie; i++) { 
+      newLabel[idx[i]] = 1;
+    }
+    // Find new labels and replace with the old ones
+    for (Pacoss_Int i = 1, ie = _dimSize[dim]; i < ie; i++) { newLabel[i] += newLabel[i - 1]; }
+    for (Pacoss_Idx i = 0, ie = _nnz; i < ie; i++) { idx[i] = newLabel[idx[i]] - 1; }
+  }
 }
 
 template class Pacoss_SparseStruct<double>;

### 2018-04-07 pacoss/src/pacoss-sparse-struct.h

@@ -21,6 +21,7 @@ class Pacoss_SparseStruct
         const std::vector<DataType> &val,
         const bool oneBased = true
         );
+    void fix();
 };
 
 #endif

### 2018-04-07 tmpi/src/tmpi.h

@@ -81,39 +81,39 @@ class tmpi_exception : public std::exception
 }
 #endif
 
-// MPI datatypes and templated routines that return the corresponding datatypes
-typedef enum
-{
-  type_int = MPI_INT,
-  type_long = MPI_LONG,
-  type_longlong = MPI_LONG_LONG,
-  type_float = MPI_FLOAT,
-  type_double = MPI_DOUBLE
-} TMPI_DataType;
+//// MPI datatypes and templated routines that return the corresponding datatypes
+//typedef enum
+//{
+//  type_int = MPI_INT,
+//  type_long = MPI_LONG,
+//  type_longlong = MPI_LONG_LONG,
+//  type_float = MPI_FLOAT,
+//  type_double = MPI_DOUBLE
+//} TMPI_Datatype;
 
 template <class T>
-TMPI_DataType getTMPIDataType()
+MPI_Datatype getTMPIDataType()
 { throw tmpi_exception("Data type is not supported by TMPI."); }
 
 template <>
-inline TMPI_DataType getTMPIDataType<int>()
-{ return type_int; }
+inline MPI_Datatype getTMPIDataType<int>()
+{ return MPI_INT; }
 
 template <>
-inline TMPI_DataType getTMPIDataType<long>()
-{ return type_long; }
+inline MPI_Datatype getTMPIDataType<long>()
+{ return MPI_LONG; }
 
 template <>
-inline TMPI_DataType getTMPIDataType<long long>()
-{ return type_longlong; }
+inline MPI_Datatype getTMPIDataType<long long>()
+{ return MPI_LONG_LONG; }
 
 template <>
-inline TMPI_DataType getTMPIDataType<float>()
-{ return type_float; }
+inline MPI_Datatype getTMPIDataType<float>()
+{ return MPI_FLOAT; }
 
 template <>
-inline TMPI_DataType getTMPIDataType<double>()
-{ return type_double; }
+inline MPI_Datatype getTMPIDataType<double>()
+{ return MPI_DOUBLE; }
 
 // TMPI functions
 

### 2018-04-07 distnmf/distnmf.cpp
bugfix: not using lrinitializer when using pacos

@@ -212,6 +212,7 @@ class DistNMFDriver {
         // rand initialization hurts ANLS BPP running time. For a better
         // initializer we run couple of iterations of HALS.
 
+#ifndef USE_PACOS
 #ifdef BUILD_SPARSE
         if (m_nmfalgo == ANLSBPP) {
             DistHALS<SP_MAT> lrinitializer(A,
@@ -226,6 +227,7 @@ class DistNMFDriver {
             H = lrinitializer.getRightLowRankFactor();
         }
 #endif  // ifdef BUILD_SPARSE
+#endif
 
 #ifdef MPI_VERBOSE
         INFO << mpicomm.rank() << "::" << __PRETTY_FUNCTION__ << "::" \

### 2018-04-07 distnmf/distnmf.cpp
bugfix: PACOS should have been PACOSS

@@ -212,7 +212,7 @@ class DistNMFDriver {
         // rand initialization hurts ANLS BPP running time. For a better
         // initializer we run couple of iterations of HALS.
 
-#ifndef USE_PACOS
+#ifndef USE_PACOSS
 #ifdef BUILD_SPARSE
         if (m_nmfalgo == ANLSBPP) {
             DistHALS<SP_MAT> lrinitializer(A,

### 2018-04-09 latex/icpp18/experiments.tex
included stackoverflow

--- 2018-04-02 latex/sc17/experiments.tex
@@
 
+%%%%%%%%%%% NNZ Sweep %%%%%%%%
+
+\begin{figure}
+
+\begin{subfigure}[t]{0.5\textwidth}
+\begin{tikzpicture}
+\begin{axis}[xlabel=Edge Factor, ylabel=Time(in Secs),width=3.5in, height=2.0in, 
+y label style={yshift=-.5cm}, 
+%legend pos=outer, legend style={draw=none,row sep=-0.1cm},
+%legend style={draw=none,at={(0.5,-0.1)},anchor=north},
+%legend columns = -1,
+%legend pos=outer north west, legend style={draw=none,row sep=-0.1cm},
+xtick=data, 
+xticklabels={4, 8, 16, 32, 64}]
+]
+\addplot  table [x={edgefactor}, y={collective}] {data/nnzsweepall_k48_p4096_comp.dat};
+\addplot  table [x={edgefactor}, y={cp1d}] {data/nnzsweepall_k48_p4096_comp.dat};
+\addplot  table [x={edgefactor}, y={cp2d}] {data/nnzsweepall_k48_p4096_comp.dat};
+\addplot  table [x={edgefactor}, y={cr}] {data/nnzsweepall_k48_p4096_comp.dat};
+\addplot  table [x={edgefactor}, y={cu}] {data/nnzsweepall_k48_p4096_comp.dat};
+%\legend{collective, cp1d, cp2d, cr, cu}
+\end{axis}
+\end{tikzpicture}
+\subcaption{Computation cost for low rank $k$=48 on 4096 procs}
+\label{fig:nnzsweepcomp}
+\end{subfigure}
+
+\begin{subfigure}[t]{0.5\textwidth}
+\begin{tikzpicture}
+\begin{axis}[xlabel=Edge Factor, ylabel=Time(in Secs),width=3.5in, height=2.0in, 
+y label style={yshift=-.5cm}, 
+%legend pos=outer north west, legend style={draw=none,row sep=-0.1cm},
+legend style={draw=none,at={(0.5,-0.3)},anchor=north},
+legend columns = -1
+xtick=data, 
+xticklabels={4, 8, 16, 32, 64}]
+]
+\addplot  table [x={edgefactor}, y={collective}] {data/nnzsweepall_k48_p4096_comm.dat};
+\addplot  table [x={edgefactor}, y={cp1d}] {data/nnzsweepall_k48_p4096_comm.dat};
+\addplot  table [x={edgefactor}, y={cp2d}] {data/nnzsweepall_k48_p4096_comm.dat};
+\addplot  table [x={edgefactor}, y={cr}] {data/nnzsweepall_k48_p4096_comm.dat};
+\addplot  table [x={edgefactor}, y={cu}] {data/nnzsweepall_k48_p4096_comm.dat};
+\legend{collective, cp1d, cp2d, cr, cu}
+\end{axis}
+\end{tikzpicture}
+\subcaption{Communication cost for low rank $k$=48 on 4096 procs}
+\label{fig:nnzsweepcomm}
+\end{subfigure}
+
+\end{figure}
+
+%\begin{figure}
+%\begin{subfigure}[t]{0.5\textwidth}
+%\begin{tikzpicture}
+%\renewcommand{\minvalue}{1}
+%\renewcommand{\numiterations}{30}
+%\renewcommand{\datafile}{data/nnzsweepall_k48_p4096_comm.dat}
+%\speedup
+%\end{tikzpicture}
+%\subcaption{Communication cost for low rank $k$=48 on 4096 procs}
+%\label{fig:flickr-speedup-16}
+%\end{subfigure}
+%
+%\begin{subfigure}[t]{0.5\textwidth}
+%\begin{tikzpicture}
+%\renewcommand{\minvalue}{1}
+%\renewcommand{\numiterations}{30}
+%\renewcommand{\datafile}{data/nnzsweepall_k48_p4096_comp.dat}
+%\speedup
+%\end{tikzpicture}
+%\subcaption{Communication cost for low rank $k$=48 on 4096 procs}
+%\label{fig:flickr-speedup-48}
+%\end{subfigure}
+%\end{figure}
+
 %%%%%%%%%%% Strong scaling flickr %%%%%%%%
@@
 \end{subfigure}
-
-%\begin{subfigure}[t]{0.5\textwidth}
-%\begin{tikzpicture}
-%\bgqtrue
-%\renewcommand{\minvalue}{137.1}
-%\renewcommand{\numiterations}{10}
-%\renewcommand{\datafile}{data/stackoverflow-speedup-bgq.dat}
-%\speedup
-%\end{tikzpicture}
-%\subcaption{Stackoverflow dataset Low Rank $k=48$ on Bluegene/Q}
-%\end{subfigure}
 
@@
 
+%%%%%%%%%%% Stackexchange Details %%%%%%%%
+\renewcommand{\minvalue}{1}
+\begin{figure}
+\begin{subfigure}[t]{0.25\textwidth}
+\begin{tikzpicture}
+\ylabelfalse
+\legendtrue
+\renewcommand{\numiterations}{30}
+\renewcommand{\datafile}{data/stkx-details1536-16.dat}
+\makeplot
+\end{tikzpicture}
+\subcaption{Rhea $p=1536$, Low Rank $k=16$}
+\end{subfigure}
+~
+\begin{subfigure}[t]{0.25\textwidth}
+\begin{tikzpicture}
+\ylabelfalse
+\legendtrue
+\renewcommand{\numiterations}{30}
+\renewcommand{\datafile}{data/stkx-details1536-48.dat}
+\makeplot
+\end{tikzpicture}
+\subcaption{Rhea $p=1536$, Low Rank $k=48$}
+\end{subfigure}
+
+\begin{subfigure}[t]{0.5\textwidth}
+\begin{tikzpicture}
+\bgqtrue
+\renewcommand{\minvalue}{137.1}
+\renewcommand{\numiterations}{10}
+\renewcommand{\datafile}{data/stackoverflow-speedup-bgq.dat}
+\speedup
+\end{tikzpicture}
+\subcaption{Stackoverflow dataset Low Rank $k=48$ on Bluegene/Q}
+\end{subfigure}
+
+\caption{Stack exchange dataset Time Split up for Low Rank $k=16, 48$ on Rhea and Strong scaling on Bluegene/Q}
+\label{fig:flickrdetails}
+\end{figure}
+
 %%%%%%%%%%% Delicious Details %%%%%%%%

### 2018-04-09 latex/icpp18/experiments.tex
removed k=16 for all experiments

--- 2018-04-02 latex/sc17/experiments.tex
@@
 
+%%%%%%%%%%% NNZ Sweep %%%%%%%%
+
+\begin{figure}
+\renewcommand{\numiterations}{30}
+\renewcommand{\minvalue}{1}
+\begin{subfigure}[t]{0.5\textwidth}
+\begin{tikzpicture}
+\begin{axis}[xlabel=Edge Factor, ylabel=Time(in Secs),width=3.5in, height=2.0in, 
+y label style={yshift=-.5cm}, 
+%legend pos=outer, legend style={draw=none,row sep=-0.1cm},
+%legend style={draw=none,at={(0.5,-0.1)},anchor=north},
+%legend columns = -1,
+%legend pos=outer north west, legend style={draw=none,row sep=-0.1cm},
+xtick=data, 
+xticklabels={4, 8, 16, 32, 64}
+]
+\addplot  table [x={edgefactor}, y expr=(\thisrow{collective}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comp.dat};
+\addplot  table [x={edgefactor}, y expr=(\thisrow{cp1d}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comp.dat};
+\addplot  table [x={edgefactor}, y expr=(\thisrow{cp2d}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comp.dat};
+\addplot  table [x={edgefactor}, y expr=(\thisrow{cr}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comp.dat};
+\addplot  table [x={edgefactor}, y expr=(\thisrow{cu}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comp.dat};
+%\legend{collective, cp1d, cp2d, cr, cu}
+\end{axis}
+\end{tikzpicture}
+\subcaption{Computation cost for low rank $k$=48 on 4096 procs}
+\label{fig:nnzsweepcomp}
+\end{subfigure}
+
+\begin{subfigure}[t]{0.5\textwidth}
+\begin{tikzpicture}
+\begin{axis}[xlabel=Edge Factor, ylabel=Time(in Secs),width=3.5in, height=2.0in, 
+y label style={yshift=-.5cm}, 
+%legend pos=outer north west, legend style={draw=none,row sep=-0.1cm},
+xtick=data, 
+xticklabels={4, 8, 16, 32, 64},
+legend style={draw=none,at={(0.5,-0.3)},anchor=north},
+legend columns = -1
+]
+
+\addplot  table [x={edgefactor}, y expr=(\thisrow{collective}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comm.dat};
+\addplot  table [x={edgefactor}, y expr=(\thisrow{cp1d}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comm.dat};
+\addplot  table [x={edgefactor}, y expr=(\thisrow{cp2d}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comm.dat};
+\addplot  table [x={edgefactor}, y expr=(\thisrow{cr}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comm.dat};
+\addplot  table [x={edgefactor}, y expr=(\thisrow{cu}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comm.dat};\legend{collective, cp1d, cp2d, cr, cu}
+\end{axis}
+\end{tikzpicture}
+\subcaption{Communication cost for low rank $k$=48 on 4096 procs}
+\label{fig:nnzsweepcomm}
+\end{subfigure}
+
+\end{figure}
+
 %%%%%%%%%%% Strong scaling flickr %%%%%%%%
 \begin{figure}
-\begin{subfigure}[t]{0.5\textwidth}
-\begin{tikzpicture}
-\renewcommand{\minvalue}{291.82}
-\renewcommand{\numiterations}{30}
-\renewcommand{\datafile}{data/flickr-speedup16.dat}
-\speedup
-\end{tikzpicture}
-\subcaption{Flickr dataset with low rank $k=16$ on Rhea}
-\label{fig:flickr-speedup-16}
-\end{subfigure}
+%\begin{subfigure}[t]{0.5\textwidth}
+%\begin{tikzpicture}
+%\renewcommand{\minvalue}{291.82}
+%\renewcommand{\numiterations}{30}
+%\renewcommand{\datafile}{data/flickr-speedup16.dat}
+%\speedup
+%\end{tikzpicture}
+%\subcaption{Flickr dataset with low rank $k=16$ on Rhea}
+%\label{fig:flickr-speedup-16}
+%\end{subfigure}
 
@@
 
-%\begin{subfigure}[t]{0.5\textwidth}
-%\begin{tikzpicture}
-%\bgqtrue
-%\renewcommand{\minvalue}{137.1}
-%\renewcommand{\numiterations}{10}
-%\renewcommand{\datafile}{data/stackoverflow-speedup-bgq.dat}
-%\speedup
-%\end{tikzpicture}
-%\subcaption{Stackoverflow dataset Low Rank $k=48$ on Bluegene/Q}
-%\end{subfigure}
-
 \caption{Strong scaling on Flickr dataset}
@@
 \begin{figure}
-\begin{subfigure}[t]{0.5\textwidth}
-\begin{tikzpicture}
-\renewcommand{\minvalue}{400.8}
-\renewcommand{\datafile}{data/delicious-speedup16.dat}
-\speedup
-\end{tikzpicture}
-\subcaption{Delicious dataset with low rank $k=16$ on Rhea}
-\label{fig:delicious-speedup-16}
-\end{subfigure}
+%\begin{subfigure}[t]{0.5\textwidth}
+%\begin{tikzpicture}
+%\renewcommand{\minvalue}{400.8}
+%\renewcommand{\datafile}{data/delicious-speedup16.dat}
+%\speedup
+%\end{tikzpicture}
+%\subcaption{Delicious dataset with low rank $k=16$ on Rhea}
+%\label{fig:delicious-speedup-16}
+%\end{subfigure}
 
@@
 %
-
+~
 \begin{subfigure}[t]{0.5\textwidth}
@@
 \begin{figure}
-\begin{subfigure}[t]{0.25\textwidth}
-\begin{tikzpicture}
-\ylabeltrue
-\legendfalse
-\renewcommand{\numiterations}{30}
-\renewcommand{\datafile}{data/flickr-details16-16.dat}
-\makeplot
-\end{tikzpicture}
-\subcaption{Rhea $p=16$, Low Rank $k=16$}
-\end{subfigure}
-~
-\begin{subfigure}[t]{0.25\textwidth}
-\begin{tikzpicture}
-\ylabelfalse
-\legendtrue
-\renewcommand{\numiterations}{30}
-\renewcommand{\datafile}{data/flickr-details3072-16.dat}
-\makeplot
-\end{tikzpicture}
-\subcaption{Rhea $p=3072$, Low Rank $k=16$}
-\end{subfigure}
+
+%\begin{subfigure}[t]{0.25\textwidth}
+%\begin{tikzpicture}
+%\ylabeltrue
+%\legendfalse
+%\renewcommand{\numiterations}{30}
+%\renewcommand{\datafile}{data/flickr-details16-16.dat}
+%\makeplot
+%\end{tikzpicture}
+%\subcaption{Rhea $p=16$, Low Rank $k=16$}
+%\end{subfigure}
+%~
+%\begin{subfigure}[t]{0.25\textwidth}
+%\begin{tikzpicture}
+%\ylabelfalse
+%\legendtrue
+%\renewcommand{\numiterations}{30}
+%\renewcommand{\datafile}{data/flickr-details3072-16.dat}
+%\makeplot
+%\end{tikzpicture}
+%\subcaption{Rhea $p=3072$, Low Rank $k=16$}
+%\end{subfigure}
 
@@
 
-%%%%%%%%%%% Delicious Details %%%%%%%%
+%%%%%%%%%%% Stackexchange Details %%%%%%%%
+%\renewcommand{\minvalue}{1}
+\begin{figure}
 
-\renewcommand{\minvalue}{1}
-\begin{figure}
-\begin{subfigure}[t]{0.25\textwidth}
-\begin{tikzpicture}
-\ylabeltrue
-\legendfalse
-\renewcommand{\numiterations}{30}
-\renewcommand{\datafile}{data/delicious-details16-16.dat}
-\makeplot
-\end{tikzpicture}
-\subcaption{Rhea $p=16$, Low rank $k=16$}
-\end{subfigure}
-~
+%\begin{subfigure}[t]{0.25\textwidth}
+%\begin{tikzpicture}
+%\ylabelfalse
+%\legendtrue
+%\renewcommand{\numiterations}{30}
+%\renewcommand{\datafile}{data/stkx-details1536-16.dat}
+%\makeplot
+%\end{tikzpicture}
+%\subcaption{Rhea $p=1536$, Low Rank $k=16$}
+%\end{subfigure}
+
 \begin{subfigure}[t]{0.25\textwidth}
@@
 \renewcommand{\numiterations}{30}
-\renewcommand{\datafile}{data/delicious-details3072-16.dat}
+\renewcommand{\datafile}{data/stkx-details1536-48.dat}
 \makeplot
 \end{tikzpicture}
-\subcaption{Rhea $p=3072$, Low rank $k=16$}
+\subcaption{Rhea $p=1536$, Low Rank $k=48$}
 \end{subfigure}
+
+\begin{subfigure}[t]{0.5\textwidth}
+\begin{tikzpicture}
+\bgqtrue
+\renewcommand{\minvalue}{137.1}
+\renewcommand{\numiterations}{10}
+\renewcommand{\datafile}{data/stackoverflow-speedup-bgq.dat}
+\speedup
+\end{tikzpicture}
+\subcaption{Stackoverflow dataset Low Rank $k=48$ on Bluegene/Q}
+\end{subfigure}
+
+\caption{Stack exchange dataset Time Split up for Low Rank $k=16, 48$ on Rhea and Strong scaling on Bluegene/Q}
+\label{fig:flickrdetails}
+\end{figure}
+
+%%%%%%%%%%% Delicious Details %%%%%%%%
+
+%\renewcommand{\minvalue}{1}
+\begin{figure}
+
+%\begin{subfigure}[t]{0.25\textwidth}
+%\begin{tikzpicture}
+%\ylabeltrue
+%\legendfalse
+%\renewcommand{\numiterations}{30}
+%\renewcommand{\datafile}{data/delicious-details16-16.dat}
+%\makeplot
+%\end{tikzpicture}
+%\subcaption{Rhea $p=16$, Low rank $k=16$}
+%\end{subfigure}
+%~
+%\begin{subfigure}[t]{0.25\textwidth}
+%\begin{tikzpicture}
+%\ylabelfalse
+%\legendtrue
+%\renewcommand{\numiterations}{30}
+%\renewcommand{\datafile}{data/delicious-details3072-16.dat}
+%\makeplot
+%\end{tikzpicture}
+%\subcaption{Rhea $p=3072$, Low rank $k=16$}
+%\end{subfigure}
 

### 2018-04-09 latex/icpp18/introduction.tex
minor

@@
 NMF is an effective tool for analyzing such data in many applications, yet there is a lack of high performance algorithms and software that can compute it efficiently.
-Our goal is in this paper is to fill in this gap by providing scalable parallel algorithms that employ efficient communication schemes and can scale to thousands of processes.
+There are scalable tensor factorization algorithms resembling NMF kernels with sparse irregular computational patterns, yet they require an expensive partitioning step and do not necessarily yield the optimal performance in the context of NMF.
+Our goal is in this paper is to fill in this gap by comparing the existing approaches and providing scalable parallel algorithms that employ efficient communication and partitioning schemes enabling scaling to thousands of processes.
 

### 2018-04-09 latex/icpp18/introduction.tex
pass over intro and title

--- 2018-04-04 latex/icpp18/introduction.tex
@@
 Furthermore, data sets from these applications are often quite sparse and have highly irregular nonzero patterns.
-NMF is an effective tool for analyzing such data in many applications, yet there is a lack of high performance algorithms and software that can compute it efficiently.
-Our goal is in this paper is to fill in this gap by providing scalable parallel algorithms that employ efficient communication schemes and can scale to thousands of processes.
 
@@
 
-The focus in this work is the comparison and combination of two previous approaches that can be applied to sparse NMF problems.
-The first approach is based on MPI-FAUN \cite{KBP17}, a software framework for computing the NMF of dense and sparse data.
-The second approach is based on HyperTensor \cite{KU15}, a library for computing a CP decomposition of a sparse tensor.
-In the former case, the algorithm is optimized for dense input matrices, but it can be used for sparse data.
-In the latter case, the algorithm is optimized for tensor data and unconstrained linear least squares problem, but it can be applied to matrices (2D tensors) and combined with NLS updates.
-
-MPI-FAUN uses a regular, Cartesian partitioning of the input matrix $\AA$ across processors that is oblivious to the nonzero pattern.
+There is a lack of highly scalable parallel algorithms and software for computing sparse NMF.
+MPI-FAUN \cite{KBP17} software framework enables computing the NMF for dense and sparse data, yet the algorithm is optimized for dense input matrices and employs the same communication scheme to carry out sparse NMF computations.
+It uses a regular, Cartesian partitioning of the input matrix $\AA$ across processors that is oblivious to the nonzero pattern.
 The advantages of this approach are that the resulting factor matrix communication (to compute the SpMMs) is also regular and cast as low-latency MPI collective operations and that the rest of the computation (including NLS updates) is load-balanced.
 The disadvantage is that because the partition ignores the sparsity of $\AA$, the approach will often communicate more data than necessary, as some processors will receive data that they do not need for local computation.
-
-HyperTensor solves an input-sparsity-dependent hypergraph partitioning problem to determine the parallel distribution of the entries of $\AA$.
-The advantage of this approach is that by being sensitive to the nonzero pattern, the resulting factor matrix communication required to compute the SpMMs is (approximately) minimized, and the only data that a processor receives is what it needs for its local computation.
-The disadvantages of this approach are an upfront partitioning cost (which can be amortized over the iterations of the NMF algorithm) and possible load imbalance, particularly in the NLS computations.
-
-The two methods differ both in their partitioning strategy and in their communication pattern.
-We also consider a compromise approach, which uses a randomized partitioning scheme, to achieve load balance without perfectly minimizing communication, and a sparsity-sensitive communication pattern, which avoids communicating unnecessary data.
-We show that this compromise avoids heavy upfront partitioning costs and effectively balances the efficiency constraints of the different NMF computations.
+For sparse tensor factorization, there are scalable algorithms and software resembling NMF kernels with sparse irregular computational patterns~\cite{KU15}, yet they require an expensive partitioning step and do not necessarily yield the optimal performance for NMF.
+The main advantage of this approach is that it minimizes the communication cost of the SpMM step through hypergraph partitioning, and that it employs point-to-point communication scheme which communicates elements of $\WW$ and $\HH$ only to processors in need.
+The disadvantages of this approach are an upfront hypergraph partitioning cost, possible load imbalance~(particularly in the NLS computations) and communication imbalance.
+Our goal is in this paper is to fill in this gap between two approaches by comparing and evaluating them in the context of NMF, then proposing a synthesis involving parallel algorithms as well as communication and partitioning schemes enabling scalability to thousands of processes.
 
@@
 
-\grey{I have not edited the following contribution list yet, we may want to wait until we have new data...
-We summarize our main contributions as follows:
-\begin{itemize}
-\item We propose an efficient parallel NMF algorithm called \distspnmf for sparse matrices that employs a point-to-point communication scheme to leverage the sparsity of the input matrix, and eliminate any redundant communication existing in \mpifaun.
-The algorithm is flexible to work with any partition of the input matrix $\Am$ as well as factor matrices $\Wm$ and $\Hm$.
-\item We employ a 2D cartesion process topology to eliminate the communication cost of parallel NMF in one dimension, and bound the maximum number of messages sent per process in our MPI implementation.
-\item We introduce effective partitioning strategies to further reduce the communication cost and enhance the parallel scalability of our algorithm.
-\item We introduce regularization to NMF computations to prevent potential numerical instabilities using large matrices.
-\item We compare our implementation with a state-of-the-art NMF library, and report scalability results up to 32678 processors on two different parallel computing platforms using real-world datasets.
-\end{itemize}
-}
+%\grey{I have not edited the following contribution list yet, we may want to wait until we have new data...
+%We summarize our main contributions as follows:
+%\begin{itemize}
+%\item We propose an efficient parallel NMF algorithm called \distspnmf for sparse matrices that employs a point-to-point communication scheme to leverage the sparsity of the input matrix, and eliminate any redundant communication existing in \mpifaun.
+%The algorithm is flexible to work with any partition of the input matrix $\Am$ as well as factor matrices $\Wm$ and $\Hm$.
+%\item We employ a 2D cartesion process topology to eliminate the communication cost of parallel NMF in one dimension, and bound the maximum number of messages sent per process in our MPI implementation.
+%\item We introduce effective partitioning strategies to further reduce the communication cost and enhance the parallel scalability of our algorithm.
+%\item We introduce regularization to NMF computations to prevent potential numerical instabilities using large matrices.
+%\item We compare our implementation with a state-of-the-art NMF library, and report scalability results up to 32678 processors on two different parallel computing platforms using real-world datasets.
+%\end{itemize}
+%}

### 2018-04-10 latex/icpp18/abstract.tex

@@
 In this paper, we focus on scaling algorithms for NMF to very large sparse datasets and massively parallel machines by employing effective algorithms, communication patterns, and partitioning schemes that leverage the sparsity of the input matrix.
-In particular, we consider two previous works developed for related problems, one that uses a fine-grained partitioning strategy using a point-to-point communication pattern and on that uses a coarse-grained partitioning strategy using a collective-based communication pattern.
+In particular, we consider two previous works developed for related problems, one that uses a fine-grained partitioning strategy using a point-to-point communication pattern and on that uses a checkerboard partitioning strategy using a collective-based communication pattern.
 We show that a combination of the previous approaches balances the demands of the various computations within NMF algorithms and achieves high efficiency and scalability.

### 2018-04-10 latex/icpp18/algorithm.tex

@@
-\section{Algorithm}
+\section{Distributed Sparse NMF}
 
@@
 For a given partition, we describe how parallel computations and communications take place within the algorithm, and illustrate computational and communication costs associated with a partition.
-In doing so, we elaborate the advantages of our parallelization scheme compared with the state of the art.
-We then discuss efficient partitioning strategies to better establish computational load balance and reduce communication.
+We then discuss efficient partitioning strategies to better establish computational load balance and reduce communication in NMF.
+In doing so, we also explain how existing methods compare to this scheme with their advantages and disadvantages in terms of partitioning.
 
-\subsection{Distributed Parallel Sparse NMF Algorithm}
+\subsection{Distributed Sparse NMF Algorithm}
 
@@
       $\Fp$,$\Gp$: Footprints of process $p$ on $\Wm$ and $\Hm$ \\
-      $\Wm(\Ip,:)$,$\Hm(:,\Jp)$: Owned rows and columns of initial $\Wm$ and $\Hm$ \\
+      $\Wm(\Ip,:)$,$\Hm(:,\Jp)$: Owned rows/columns of $\Wm$ and $\Hm$ \\
       $k$: The NMF rank}
-    \ENSURE {Process $p$ gets the final value of $\Wm(\Ip,:)$ and $\Hm(:,\Jp)$}
+    \ENSURE {Process $p$ gets final values of $\Wm(\Ip,:)$ and $\Hm(:,\Jp)$}
       \REPEAT
@@
 
-Parallellizing NMF involves partitioning data and computational tasks to processes, and communicating results wherever necessary, just as any parallel algorithm.
-Specifically for NMF, one needs to partition the sparse matrix $\Am$ as well as the factor matrices $\Wm$ and $\Hm$, where the former partitioning distributes the computational load of sparse matrix-dense matrix multiplications $\Am \Hm$ and $\Wm^T \Am$, whereas the latter divides the workload of NNLS computations to processes.
+Parallelizing sparse NMF involves partition the sparse matrix $\Am$ as well as the factor matrices $\Wm$ and $\Hm$, where the former partitioning distributes the computational load of sparse matrix-dense matrix multiplications $\Am \Hm$ and $\Wm^T \Am$, whereas the latter divides the workload of NNLS computations to processes.
 We provide the execution of our parallel algorithm for computing a rank-$k$ NMF of a sparse matrix $\Am \in \R^{m \times n}$ in \cref{alg:distnmf}, which is executed by each process $p$ for $1 \le p \le P$.
-The algorithm starts with an arbitrary partition of the input matrix and the factor matrices; process $p$ owns the submatrices $\Wm(\Ip, :)$ and $\Hm(:, \Jp)$ as well as the nonzero elements of the sparse matrix $\Am_p$ where $\Am = \bigcup_{i=1}^{P}\Am_i$, i.e., $\Am_1, \dots, \Am_p$ partitions the nonzeros of $\Am$.
-The sets $\Fp$ and $\Gp$ denote the ``footprints'' of the process $p$ on the rows and the columns of matrices $\Wm$ and $\Hm$, respectively, in the course of the algorithm.
-That is, we have $i \in \Fp$ or $j \in \Gp$ if only if $i \in \Ip$ or $j \in \Jp$~(row/column is owned), or there is a nonzero element $a_{i,j} \in \Amp$~(row/column is used in local computations).
-At each iteration, the process $p$ is responsible for gathering the new value of submatrices $\Wm(\Ip, :)$ and $\Hm(:, \Jp)$, and sending these values to processes in need.
+The algorithm starts with an arbitrary partition of the input matrix and the factor matrices; process $p$ owns the submatrices $\Wm(\Ip, :)$ and $\Hm(:, \Jp)$ as well as the nonzero elements of the sparse matrix $\Am_p$ where $\Am = \bigcup_{i=1}^{P}\Am_i$, i.e., $\Am_1, \dots, \Am_P$ partitions the nonzeros of $\Am$.
+The sets $\Fp$ and $\Gp$ denote the ``footprints'' of the process $p$ on the rows and the columns of matrices $\Wm$ and $\Hm$, respectively; hence, these rows need to be stored by this process.
+Specifically, we have $i \in \Fp$ or $j \in \Gp$ if only if $i \in \Ip$ or $j \in \Jp$~(row/column is owned), or there is a nonzero element $a_{i,j} \in \Amp$~(row/column is used in local computations).
+At each iteration, the process $p$ is responsible for gathering the new value of submatrices $\Wm(\Ip, :)$ and $\Hm(:, \Jp)$, and sending them to processes in need.
 
@@
 The way these two communications are carried out plays a vital role in obtaining parallel scalability as they dominate the communication cost of the algorithm.
-Here, we employ an efficient point-to-point communication framework to effectuate fold and expand communications.
-In this scheme, after determining a partition of the matrix, each process $p$ determines the sets $\Fp$ and $\Gp$ using the row and column indices in its local matrix $\Amp$ and its sets $\Ip$ and $\Jp$ of owned row indices, then identifies the owners of these rows in order to request these rows from the corresponding processes in the expand step.
-Next, these requests are exchanged in a communication step so that each process $p$ determines the unique row indices that it needs to send to or receive from each other process in expand and fold steps.
-Once this communication structure is formed, it is reused througout all iterations in \cref{alg:distnmf}.
-This is possible thanks to the data partition staying fixed during the algorithm, hence data dependencies do not change.
 
-The existing work in the literature \cite{KBP16,KBP16MPIFAUN} amounts to employing an all-to-all communication strategy for both expand and fold steps of \cref{alg:distnmf} for dense as well as sparse $\Am$.
+%Here, we employ an efficient point-to-point communication framework to effectuate fold and expand communications.
+%In this scheme, after determining a partition of the matrix, each process $p$ determines the sets $\Fp$ and $\Gp$ using the row and column indices in its local matrix $\Amp$ and its sets $\Ip$ and $\Jp$ of owned row indices, then identifies the owners of these rows in order to request these rows from the corresponding processes in the expand step.
+%Next, these requests are exchanged in a communication step so that each process $p$ determines the unique row indices that it needs to send to or receive from each other process in expand and fold steps.
+%Once this communication structure is formed, it is reused througout all iterations in \cref{alg:distnmf}.
+%This is possible thanks to the data partition staying fixed during the algorithm, hence data dependencies do not change.
+
+%For convenience of the reader, \cref{tab:costs} discusses the computation, communication and memory complexity of this partitioning scheme that is observed in the \mpifaun algorithm. 
+%To illustrate, in \cref{fig:partition} we provide a sparse matrix partitioned using a $5 \times 5$ checkerboard scheme where each process owns a subblock of the matrix, and show only the nonzero elements belonging to three columns with indices $j_1$, $j_2$, and $j_3$.
+%Here, in performing the expand communication of \cref{line:exph}, the all-to-all strategy sends these column vectors to all processes in the same column, whereas point-to-point communication identifies the exact set of proceses in need beforehand, indicated with the same color, and communicates the data only with these processes.
+%In this example, all-to-all scheme incurs 4, 3, and 2 units of redundant communication for columns $j_3$, $j_2$, and $j_1$, respectively, and this redundancy only worsens with the increasing number of processes.
+
+\subsection{Communication Scheme}
+\mpifaun employs collective communication strategies for both expand and fold steps of \cref{alg:distnmf} for dense as well as sparse $\Am$, and partitions $\Am$ using a uniform checkerboard topology.
 In this strategy, the rows and the columns indices $1, \dots, m$ and $1, \dots, n$ are divided into $P_r$ and $P_c$~($P = P_r P_c$) sets $I_1, \dots, I_{P_r}$ and $J_1, \dots, J_{P_c}$ of equal size and having contiguous indices.
@@
 Despite being favorable due to small number of exchanged messages in all-to-all routines, in this strategy processes receive many rows that they do not need in their local sparse matrix dense matrix multiplication, and this redundancy dramatically increases the communication volume, thus preventing scalability.
-%For convenience of the reader, \cref{tab:costs} discusses the computation, communication and memory complexity of this partitioning scheme that is observed in the \mpifaun algorithm. 
-To illustrate, in \cref{fig:partition} we provide a sparse matrix partitioned using a $5 \times 5$ checkerboard scheme where each process owns a subblock of the matrix, and show only the nonzero elements belonging to three columns with indices $j_1$, $j_2$, and $j_3$.
-Here, in performing the expand communication of \cref{line:exph}, the all-to-all strategy sends these column vectors to all processes in the same column, whereas point-to-point communication identifies the exact set of proceses in need beforehand, indicated with the same color, and communicates the data only with these processes.
-In this example, all-to-all scheme incurs 4, 3, and 2 units of redundant communication for columns $j_3$, $j_2$, and $j_1$, respectively, and this redundancy only worsens with the increasing number of processes.
 
-\begin{figure}
-\caption{A 5x5 checkerboard partition of a sparse matrix.}
-\label{fig:partition}
-\includegraphics[width=0.8\linewidth]{figures/matrix-partition.png}
-\end{figure}
+%\begin{figure}
+%\caption{A 5x5 checkerboard partition of a sparse matrix.}
+%\label{fig:partition}
+%\includegraphics[width=0.8\linewidth]{figures/matrix-partition.png}
+%\end{figure}
 

### 2018-04-10 latex/icpp18/introduction.tex

@@
 It uses a regular, Cartesian partitioning of the input matrix $\AA$ across processors that is oblivious to the nonzero pattern.
-The advantages of this approach are that the resulting factor matrix communication (to compute the SpMMs) is also regular and cast as low-latency MPI collective operations and that the rest of the computation (including NLS updates) is load-balanced.
+The advantages of this approach are that the resulting factor matrix communication (to compute the SpMMs) is also regular and cast as low-latency MPI collective operations and that the rest of the computation (including NLS updates) is perfectly load-balanced.
 The disadvantage is that because the partition ignores the sparsity of $\AA$, the approach will often communicate more data than necessary, as some processors will receive data that they do not need for local computation.

### 2018-04-10 latex/icpp18/algorithm.tex
progress

@@
 \subsection{Communication Scheme}
+\paragraph{Collective communication~(\textbf{COL}):} 
 \mpifaun employs collective communication strategies for both expand and fold steps of \cref{alg:distnmf} for dense as well as sparse $\Am$, and partitions $\Am$ using a uniform checkerboard topology.
 In this strategy, the rows and the columns indices $1, \dots, m$ and $1, \dots, n$ are divided into $P_r$ and $P_c$~($P = P_r P_c$) sets $I_1, \dots, I_{P_r}$ and $J_1, \dots, J_{P_c}$ of equal size and having contiguous indices.
-This effectively defines a 2D uniform checkerboard partition of $\Am$ with matrix blocks $\Am_{(r,c)} = \Am(I_r, J_c)$ for all $1 \le r \le P_r$ and $1 \le c \le P_c$, as well as a row partition of $\Wm$ and $\Hm$.
 Here, process $p$ owns the matrix subblock $\Am_{(r, c)}$, where $r = \lfloor P/P_c \rfloor + 1$ and $c = (P \mod P_c) + 1$, as well as $m / P$ and $n / P$ rows of $\Wm(I_r, :)$ and $\Hm(J_c, :)$.  
-The communication is performed within each process row and column using \textsc{All-Gather} and \textsc{Reduce-Scatter} routines, in which the process $p$ receives all matrix rows $\Wm(I_r, :)$ and $\Hm(I_c, :)$ belonging to processes in the same row and column of the process grid.
-Despite being favorable due to small number of exchanged messages in all-to-all routines, in this strategy processes receive many rows that they do not need in their local sparse matrix dense matrix multiplication, and this redundancy dramatically increases the communication volume, thus preventing scalability.
+As $\Am_{(r, c)}$ only touches the rows/columns of processes in the same row/column of the processor grid, the communication of $\WW$ and $\HH$ are performed within each process row and column using \textsc{All-Gather} and \textsc{Reduce-Scatter} routines, in which the process $p$ receives all matrix rows $\Wm(I_r, :)$ and $\Hm(J_c, :)$ belonging to processes in the same row and column of the process grid.
+Despite being favorable due to small number of exchanged messages in collective routines, in this strategy processes might receive rows that they do not need in their local sparse matrix dense matrix multiplication~(particularly if $\AA$ is very sparse), and this redundancy dramatically increases the communication volume, thus preventing scalability.
+
+\paragraph{Point-to-point communication~(\textbf{P2P}):} 
+\hypertensor employs point-to-point communication for fold and expand steps by precomputing set of processes having a row/column in its footprint for each row/column of $\WW$/$\HH$.
+This reduces the communication volume at the cost of increased number of messages with respect to the strategy of \mpifaun.
 
@@
 \subsection{Partitioning}
+
+%This effectively defines a 2D uniform checkerboard partition of $\Am$ with matrix blocks $\Am_{(r,c)} = \Am(I_r, J_c)$ for all $1 \le r \le P_r$ and $1 \le c \le P_c$, as well as a row partition of $\Wm$ and $\Hm$.
+%Here, process $p$ owns the matrix subblock $\Am_{(r, c)}$, where $r = \lfloor P/P_c \rfloor + 1$ and $c = (P \mod P_c) + 1$, as well as $m / P$ and $n / P$ rows of $\Wm(I_r, :)$ and $\Hm(J_c, :)$.
+%
+
 \Cref{alg:distnmf} requires a partitioning of the nonzeros of $\Am$ as well as the rows and the columns of $\Wm$ and $\Hm$, and these three partitions completely determine its computational and communication costs.
-Here, we compare different partitioning strategies, and argue how they relate to these two performance metrics.
+Here, we compare different partitioning strategies, employed by \mpifaun, \hypertensor, and SpMV kernels, and argue how they relate to these two performance metrics.
 
-Our algorithm is fine-grained; meaning that it can work any fine-grain, 1D, or 2D/checkerboard partition of the sparse matrix $\Am$.
-Here, however, we exclusively focus on $P_r \times P_c$~(where $P = P_r P_c$) checkerboard partitionings, and always set $P_c$ to the number of cores available per compute node in the cluster.
-This helps us reduce the communication cost in two ways.
-The first one corresponds to that we do not employ shared-memory parallelism in our implementation, hence assign one MPI rank per core.
-In this scenario, by assigning each group of $P_c$ processes owning the same row block of $\Am$ to the same node, and similarly assigning the corresponding row blocks of $\Wm$ and $\Wmt$ to these processes, we effectively eliminate the communication in the row dimension of the process grid~(which pertains to communicating matrices $\Wm$ and $\Wmt$) as these row exchanges stay within a node.
-This enables us to focus solely on reducing the communication due to columns of $\Hm$ and $\Hmt$.
-Second, this topology bounds the maximum number of messages sent per process by $P_r-1$ and $P_c-1$, thereby significantly reduce the communication latency by a factor of $P_c$ in compare with a fine-grain or a 1D partition, which incurs up to $P-1$ messages per process.
+%Our algorithm is fine-grained; meaning that it can work any fine-grain, 1D, or 2D/checkerboard partition of the sparse matrix $\Am$.
+%Here, however, we exclusively focus on $P_r \times P_c$~(where $P = P_r P_c$) checkerboard partitionings, and always set $P_c$ to the number of cores available per compute node in the cluster.
+%This helps us reduce the communication cost in two ways.
+%The first one corresponds to that we do not employ shared-memory parallelism in our implementation, hence assign one MPI rank per core.
+%In this scenario, by assigning each group of $P_c$ processes owning the same row block of $\Am$ to the same node, and similarly assigning the corresponding row blocks of $\Wm$ and $\Wmt$ to these processes, we effectively eliminate the communication in the row dimension of the process grid~(which pertains to communicating matrices $\Wm$ and $\Wmt$) as these row exchanges stay within a node.
+%This enables us to focus solely on reducing the communication due to columns of $\Hm$ and $\Hmt$.
+%Second, this topology bounds the maximum number of messages sent per process by $P_r-1$ and $P_c-1$, thereby significantly reduce the communication latency by a factor of $P_c$ in compare with a fine-grain or a 1D partition, which incurs up to $P-1$ messages per process.
 
-\subsubsection{Partitioning sparse input matrix $\Am$} \label{sec:cp1d}
-\paragraph{Hypergraph partitioning~(\cpp)} We aim to partition the matrix $\Am$ into $P_r$ row slices first, and $P_c$ column slices next, to obtain an $P_r \times P_c$ checkerboard partition.
+\subsubsection{Partitioning $\Am$} \label{sec:cp1d}
+
+\paragraph{Checkerboard hypergraph partitioning~(\textbf{CH2})} We aim to partition the matrix $\Am$ into $P_r$ row slices first, and $P_c$ column slices next, to obtain an $P_r \times P_c$ checkerboard partition.
 To this end, we adopt hypergraph partitioning techniques used in parallel sparse matrix literature to our case~\cite{caay:99}.
@@
 
-\paragraph{Randomized checkerboard partitioning(\crp)}
+\paragraph{1D-like checkerboard hypergraph partitioning~(\textbf{CH1})}
+
+\paragraph{Randomized checkerboard partitioning(\textbf{CRD})}
 This scheme corresponds to partitioning both the rows and the columns of $\Am$ into $R$ segments randomly.
@@
 
-\paragraph{Uniform checkerboard partitioning(\unp)}
+\paragraph{Uniform checkerboard partitioning(\textbf{CUN})}
 This partitioning variant forms an $R \times C$ partition of $\Am$ by putting a contiguous set of $m / R$ and $n / C$ rows and columns in each slice.
@@
 This is the only partitioning scheme employed by $\mpifaun$ \cite{KBP16, KBP16MPIFAUN}, hence we only use it for this implementation.
-We also use a randomized variant (\urp) of this scheme in which the rows and columns of $\Am$ are permuted randomly befor executing $\mpifaun$ to better balance its nonzeros.
+We also use a randomized variant (\urp) of this scheme in which the rows and columns of $\Am$ are permuted randomly before executing $\mpifaun$ to better balance its nonzeros.
 
-\subsubsection{Partitioning factors $\Wm$ and $\Hm$}
+\subsubsection{Partitioning $\Wm$ and $\Hm$}
 Once $\Am$ is partitioned with an $R \times C$ checkerboard topology, one has to partition rows and columns of factor matrices conformally with this topology to form the sets $\Ip$ and $\Jp$ in \cref{alg:distnmf}.

### 2018-04-10 latex/icpp18/aunmf.tex
progress

@@
 
-\grey{Not sure where the following paragraph should be incorporated:
-We would like to highlight that ``Non-negative'' Matrix Factorization is NOT matrix factorization in collaborative filtering for recommender systems for which many implementations exist.
-The collaborative filtering problem is different than NMF -- the focus of our paper -- because it interprets the ``{\em zero}'' entries of the input matrix as missing data, while the NMF problem is defined for a completely known input matrix and does not handle missing values.
-This leads to different optimization problems, algorithms, and computational steps.
-}
+%\grey{Not sure where the following paragraph should be incorporated:
+%We would like to highlight that ``Non-negative'' Matrix Factorization is NOT matrix factorization in collaborative filtering for recommender systems for which many implementations exist.
+%The collaborative filtering problem is different than NMF -- the focus of our paper -- because it interprets the ``{\em zero}'' entries of the input matrix as missing data, while the NMF problem is defined for a completely known input matrix and does not handle missing values.
+%This leads to different optimization problems, algorithms, and computational steps.
+%}
 

### 2018-04-10 latex/icpp18/introduction.tex
oguz comments

@@
 Furthermore, data sets from these applications are often quite sparse and have highly irregular nonzero patterns.
+We would like to highlight that ``Non-negative'' Matrix Factorization is NOT matrix factorization in collaborative filtering for recommender systems for which many implementations exist.
+The collaborative filtering problem is different than NMF -- the focus of our paper -- because it interprets the ``{\em zero}'' entries of the input matrix as missing data, while the NMF problem is defined for a completely known input matrix and does not handle missing values.
+This leads to different optimization problems, algorithms, and computational steps.
+
+%However, we would like to clarify that collaborative filtering in recommender systems using matrix factorization is a different problem than NMF: in the case of collaborative filtering, non-nonzeros in the matrix are considered to be missing entries, while in the case of NMF, non-nonzeros in the matrix correspond to true zero values.
 

### 2018-04-10 latex/icpp18/algorithm.tex
progress

@@
 
-\paragraph{Checkerboard hypergraph partitioning~(\textbf{CH2})} We aim to partition the matrix $\Am$ into $P_r$ row slices first, and $P_c$ column slices next, to obtain an $P_r \times P_c$ checkerboard partition.
+\paragraph{Checkerboard hypergraph partitioning~(\textbf{CH2})} \textbf{CH2} aims to partition the matrix $\Am$ into $P_r$ row slices first, and $P_c$ column slices next, to obtain an $P_r \times P_c$ checkerboard partition.
+The first partitioning is done using a column-net hypergraph model, in which for each row $\Am(i, :)$, a vertex $v_i$ with weight equaling to the number of nonzeros in $\Am(i, :)$ is created.
+Each column $j$ is represented with a hyperedge $h_j$,  and e
+
 To this end, we adopt hypergraph partitioning techniques used in parallel sparse matrix literature to our case~\cite{caay:99}.

### 2018-04-10 latex/icpp18/experiments.tex
progress up to partitioning

@@
 %The Figure \ref{fig:stackexchangerr} shows the comparison of these algorithms on the stack exchange dataset explained in Section \ref{sec:datasets}.
-\subsection {Strong Scaling}
+
+\subsection {Effect of communication scheme}
+
+\subsection {Effect of partitioning}
+
+\subsection {Strong scaling}
 
@@
 At 3072 processes, both \urp and \unp lose scalability and slow down, whereas \cpp and \crp scale to 3072 processors.
-One point to note is that for $k=16$, \cpp is considerably slower than \crp, whereas for $k=48$, \cpp starts to get faster than \crp.
-The former result is due to that \cpp typically causes more imbalance in factor matrix rows and the input matrix nonzeros, whereas in the latter result, the size of each communication unit is tripled; therefore, the reduction in the communication cost using \cpp begins to compensate for the performance loss due to load imbalance.
+%One point to note is that for $k=16$, \cpp is considerably slower than \crp, whereas for $k=48$, \cpp starts to get faster than \crp.
+%The former result is due to that \cpp typically causes more imbalance in factor matrix rows and the input matrix nonzeros, whereas in the latter result, the size of each communication unit is tripled; therefore, the reduction in the communication cost using \cpp begins to compensate for the performance loss due to load imbalance.
 
@@
 \end{tikzpicture}
-\subcaption{Computation cost for low rank $k$=48 on 4096 procs}
+\subcaption{Computation cost for $k$=48 on 4096 procs}
 \label{fig:nnzsweepcomp}
@@
 \end{tikzpicture}
-\subcaption{Communication cost for low rank $k$=48 on 4096 procs}
+\subcaption{Communication cost for $k$=48 on 4096 procs}
 \label{fig:nnzsweepcomm}
@@
 \end{tikzpicture}
-\subcaption{Flickr dataset low rank $k=48$ on Rhea}
+\subcaption{Flickr dataset $k=48$ on Rhea}
 \label{fig:flickr-speedup-48}
@@
 \end{tikzpicture}
-\subcaption{Flickr dataset low rank $k=48$ on Bluegene/Q}
+\subcaption{Flickr dataset $k=48$ on Bluegene/Q}
 \label{fig:flickr-speedup-bgq}
@@
 \end{tikzpicture}
-\subcaption{Delicious dataset low rank $k=48$ on Rhea}
+\subcaption{Delicious dataset $k=48$ on Rhea}
 \label{fig:delicious-speedup-48}
@@
 \end{tikzpicture}
-\subcaption{Delicious dataset low rank $k=48$ on Bluegene/Q}
+\subcaption{Delicious dataset $k=48$ on Bluegene/Q}
 \label{fig:delicious-speedup-bgq}
@@
 \paragraph{Flickr on Rhea:} We observe in \cref{fig:flickrdetails} that in one node configuration with $p=16$, the \unp and \urp\  performs similar to \crp\ and \cpp\ in terms of computation, and the communication time takes a small portion of the execution in all instances.
-As the number of processes increases to 3072, the communication time of \crp\ and \cpp\ stays reasonably low, whereas in the case of \unp\  and \urp, we clearly observe that the communication cost dominates the execution time using both low rank values $k=16$ and $k=48$.
+As the number of processes increases to 3072, the communication time of \crp\ and \cpp\ stays reasonably low, whereas in the case of \unp\  and \urp, we clearly observe that the communication cost dominates the execution time using both low rank values  $k=48$.
 Randomization offers load balance to \urp\ which gives it a slight edge over \unp, yet both instances suffer from the high communication cost associated with the all-to-all communication strategy, which explains the drop in the scalability results.
@@
 \end{tikzpicture}
-\subcaption{Rhea $p=16$, Low Rank $k=48$}
+\subcaption{Rhea $p=16$ for $k=48$}
 \end{subfigure}
@@
 \end{tikzpicture}
-\subcaption{Rhea $p=3072$, Low Rank $k=48$}
+\subcaption{Rhea $p=3072$ for $k=48$}
 \end{subfigure}
 
-\caption{Flickr dataset Time Split up for Low Rank $k=16, 48$ on Rhea}
+%\caption{Flickr dataset Time Split up for Low Rank $k=16, 48$ on Rhea}
+\caption{Flickr dataset Time Breakdown for $k=48$ on Rhea}
 \label{fig:flickrdetails}
@@
 %\renewcommand{\minvalue}{1}
-\begin{figure}
+%\begin{figure}
 
@@
 
-\begin{subfigure}[t]{0.25\textwidth}
-\begin{tikzpicture}
-\ylabelfalse
-\legendtrue
-\renewcommand{\numiterations}{30}
-\renewcommand{\datafile}{data/stkx-details1536-48.dat}
-\makeplot
-\end{tikzpicture}
-\subcaption{Rhea $p=1536$, Low Rank $k=48$}
-\end{subfigure}
-
-\begin{subfigure}[t]{0.5\textwidth}
-\begin{tikzpicture}
-\bgqtrue
-\renewcommand{\minvalue}{137.1}
-\renewcommand{\numiterations}{10}
-\renewcommand{\datafile}{data/stackoverflow-speedup-bgq.dat}
-\speedup
-\end{tikzpicture}
-\subcaption{Stackoverflow dataset Low Rank $k=48$ on Bluegene/Q}
-\end{subfigure}
-
-\caption{Stack exchange dataset Time Split up for Low Rank $k=16, 48$ on Rhea and Strong scaling on Bluegene/Q}
-\label{fig:flickrdetails}
-\end{figure}
+%\begin{subfigure}[t]{0.25\textwidth}
+%\begin{tikzpicture}
+%\ylabelfalse
+%\legendtrue
+%\renewcommand{\numiterations}{30}
+%\renewcommand{\datafile}{data/stkx-details1536-48.dat}
+%\makeplot
+%\end{tikzpicture}
+%\subcaption{Rhea $p=1536$, Low Rank $k=48$}
+%\end{subfigure}
+%
+%\begin{subfigure}[t]{0.5\textwidth}
+%\begin{tikzpicture}
+%\bgqtrue
+%\renewcommand{\minvalue}{137.1}
+%\renewcommand{\numiterations}{10}
+%\renewcommand{\datafile}{data/stackoverflow-speedup-bgq.dat}
+%\speedup
+%\end{tikzpicture}
+%\subcaption{Stackoverflow dataset Low Rank $k=48$ on Bluegene/Q}
+%\end{subfigure}
+%
+%
+%
+%\caption{Stack exchange dataset Time Split up for Low Rank $k=16, 48$ on Rhea and Strong scaling on Bluegene/Q}
+%\label{fig:flickrdetails}
+%\end{figure}
+%
 
@@
 \end{tikzpicture}
-\subcaption{Rhea $p=3072$, Low Rank $k=48$}
+\subcaption{Rhea $p=3072$ for $k=48$}
 \end{subfigure}
 
-\caption{Delicious dataset Time Split up for Low Rank $k=16, 48$ on Rhea}
+%\caption{Delicious dataset Time Split up for Low Rank $k=16, 48$ on Rhea}
+\caption{Delicious dataset Time Breakdown for $k=48$ on Rhea}
 \label{fig:deliciousdetails}
@@
 
-\caption{Time Split up of Flickr and Delicious for Low Rank $k=48$ on Bluegene/Q. The left two bars are for the Flickr matrix, and the right two bars are for the Delicious matrix}
+\caption{Time Breakdown of Flickr and Delicious for $k=48$ on Bluegene/Q. The left two bars are for the Flickr matrix, and the right two bars are for the Delicious matrix}
 \label{fig:bgqdetailsk48}

### 2018-04-10 latex/icpp18/introduction.tex
spmm dense introduction

@@
 The main advantage of this approach is that it minimizes the communication cost of the SpMM step through hypergraph partitioning, and that it employs point-to-point communication scheme which communicates elements of $\WW$ and $\HH$ only to processors in need.
-The disadvantages of this approach are an upfront hypergraph partitioning cost, possible load imbalance~(particularly in the NLS computations) and communication imbalance.
+The disadvantages of this approach are an upfront hypergraph partitioning cost, possible load imbalance~(particularly in the NLS 
+computations) and communication imbalance. We would like to iterate that in the case of machine learning workflow, the 
+computations after SpMM must deal with dense matrices, as Sparse-Dense matrix multiplication will result in a dense matrix. 
+Hence, the partitioning strategy considering only SpMM will result in a huge imbalance and in this specific case of NMF on NLS computations. 
 Our goal is in this paper is to fill in this gap between two approaches by comparing and evaluating them in the context of NMF, then proposing a synthesis involving parallel algorithms as well as communication and partitioning schemes enabling scalability to thousands of processes.

### 2018-04-10 latex/icpp18/introduction.tex
spmm dense introduction

@@
 computations after SpMM must deal with dense matrices, as Sparse-Dense matrix multiplication will result in a dense matrix. 
-Hence, the partitioning strategy considering only SpMM will result in a huge imbalance and in this specific case of NMF on NLS computations. 
+Hence, the partitioning strategy considering only SpMM will result in a huge imbalance in the overall workflow especially on computations after SpMM and in this specific case of NMF on NLS computations.
 Our goal is in this paper is to fill in this gap between two approaches by comparing and evaluating them in the context of NMF, then proposing a synthesis involving parallel algorithms as well as communication and partitioning schemes enabling scalability to thousands of processes.

### 2018-04-10 latex/icpp18/algorithm.tex
checkerboard progress

--- 2018-04-10 latex/icpp18/algorithm.tex
@@
 
-\paragraph{Checkerboard hypergraph partitioning~(\textbf{CH2})} We aim to partition the matrix $\Am$ into $P_r$ row slices first, and $P_c$ column slices next, to obtain an $P_r \times P_c$ checkerboard partition.
-To this end, we adopt hypergraph partitioning techniques used in parallel sparse matrix literature to our case~\cite{caay:99}.
+\paragraph{Checkerboard hypergraph partitioning~(\textbf{CH2})}
+To this end, we adopt hypergraph partitioning techniques used in parallel sparse matrix literature to our case~\cite{caay:99,aycu:08}.
 A hypergraph consists of vertices with associated weights and hyperedges that connect two or more vertices.
@@
 This corresponds in the actual computation to minimizing the data dependencies between tasks, hence the communication volume.
-Here, we form a \emph{column-net} hypergraph model; for each row $i$ and column $j$ of  $\Am$, we add a vertex $v_i$ and a hyperedge $h_j$ to the hypergraph, and for each nonzero $(i, j) \in \Am$, which implies a dependency to $\Hm(:, j)$ in computing $\Wmt(i, :)$ at \cref{line:ah}, we connect $v_i$ to $h_j$.
-We assign the value $(nnz(\Am) / m + nnz(\Am(i, :))$ to each vertex $v_i$.
-This weighting is preferred to find a tradeoff between balancing the number of vertices, which corresponds to the cost of dense matrix operations, and balancing the number of nonzeros per part, which corresponds to the cost of sparse matrix-dense matrix multiplication.
-We establish these two load balance goals by assigning a fixed cost to each vertex in the first summand, and increasing the cost of the vertices corresponding to denser rows in the second summand.
-We then partition this hypergraph into $R$ parts by balancing the vertex weights, and minimizing the cutsize of the hypergraph.
-In \cref{fig:partition} we also show this hypergraph construction for a $3 \times 3$ subset of the $A$ where rows corresponding to each vertex are highlighted and connected to related hyperedges.
-In this partitioning, minimizing the cutsize of this hypergraph exactly corresponds to minimizing the total communication volume due to columns of $\Hm$ and $\Hmt$ as is clear in \cref{fig:partition}, whereas balancing the vertex weights corresponds to balancing the cost of sparse matrix-dense matrix multiplications as well as dense matrix operations in computing Gram matrices and $\textsc{Nnls}$ in each process row.
 
-Once the matrix is partitioned into $R$ row slices, one can similarly partition columns of $\Am$ into $C$ parts using a multi-constraint formulation~\cite{aycu:08} using a \emph{row-net} hypergraph.
-Yet in our case, we skip this common approach for two reasons.
-First, existing multi-constraint partitioning tools are known to be not as efficient nor effective as single-constraint partitioners in practice.
-Second, the main goal of hypergraph partitioning of columns is to minimize the communication volume due to rows, whereas in our case communication of rows of $\Wm$ and $\Wmt$ stays within a node with a proper mapping of processes, and this communication cost would be negligible.
-Therefore, we instead partition columns randomly, which is expected to provide balance the number of nonzeros of $\Am$, number of columns of $\Hm$ and $\Hmt$, the amount of column communication volume due to columns of $\Hm$ and $\Hmt$.
+Traditional checkerboard hypergraph partitioning aims to partition the matrix $\Am$ into $P_r$ row slices first, and $P_c$ column slices next to obtain an $P_r \times P_c$ checkerboard partition.
+The first partitioning phase is done using a column-net hypergraph model, in which for each row $\Am(i, :)$, a vertex $v_i$ with weight equaling to the number of nonzeros in $\Am(i, :)$ is created.
+Each column $j$ is represented with a hyperedge $h_j$ and for each nonzero $(i, j) \in \Am$, which implies a dependency to $\Hm(:, j)$ in computing $\Wmt(i, :)$ at \cref{line:ah}, we connect $v_i$ to $h_j$.
+This hypergraph is partitioned into $P_r$ parts giving the row partition of the checkerboard topology.
+The second partitioning phase uses a row-net hypergraph model induced by this row partition, where each column is represented with a vertex with $P_r$ weights corresponding to the number of nonzeros in that column in all $P_r$ row segments.
+Partitioning this hypergraph into $P_c$ parts finalizes the $P_r \times P_c$ checkerboard partitionin by balancing the weights~(number of nonzeros of $\Am_{(r, c)}$) of each part while minimizing the communication volume.
+In \textbf{CH2},
+
+%We assign the value $(nnz(\Am) / m + nnz(\Am(i, :))$ to each vertex $v_i$.
+%This weighting is preferred to find a tradeoff between balancing the number of vertices, which corresponds to the cost of dense matrix operations, and balancing the number of nonzeros per part, which corresponds to the cost of sparse matrix-dense matrix multiplication.
+%We establish these two load balance goals by assigning a fixed cost to each vertex in the first summand, and increasing the cost of the vertices corresponding to denser rows in the second summand.
+%We then partition this hypergraph into $R$ parts by balancing the vertex weights, and minimizing the cutsize of the hypergraph.
+%In \cref{fig:partition} we also show this hypergraph construction for a $3 \times 3$ subset of the $A$ where rows corresponding to each vertex are highlighted and connected to related hyperedges.
+%In this partitioning, minimizing the cutsize of this hypergraph exactly corresponds to minimizing the total communication volume due to columns of $\Hm$ and $\Hmt$ as is clear in \cref{fig:partition}, whereas balancing the vertex weights corresponds to balancing the cost of sparse matrix-dense matrix multiplications as well as dense matrix operations in computing Gram matrices and $\textsc{Nnls}$ in each process row.
+
+%Once the matrix is partitioned into $R$ row slices, one can similarly partition columns of $\Am$ into $C$ parts using a multi-constraint formulation~\cite{aycu:08} using a \emph{row-net} hypergraph.
+%Yet in our case, we skip this common approach for two reasons.
+%First, existing multi-constraint partitioning tools are known to be not as efficient nor effective as single-constraint partitioners in practice.
+%Second, the main goal of hypergraph partitioning of columns is to minimize the communication volume due to rows, whereas in our case communication of rows of $\Wm$ and $\Wmt$ stays within a node with a proper mapping of processes, and this communication cost would be negligible.
+%Therefore, we instead partition columns randomly, which is expected to provide balance the number of nonzeros of $\Am$, number of columns of $\Hm$ and $\Hmt$, the amount of column communication volume due to columns of $\Hm$ and $\Hmt$.
 

### 2018-04-10 latex/icpp18/experiments.tex
reintroduced figure1 and remove k=16 references in experiments

@@
 
-In \cref{fig:flickr-speedup-16,fig:flickr-speedup-48} we show the speedup results of all four instances on the Rhea cluster using up to 3072 MPI ranks/cores on Flickr data.
+%In \cref{fig:flickr-speedup-16,fig:flickr-speedup-48} we show the speedup results of all four instances on the Rhea cluster using up 
+In \cref{fig:flickr-speedup-48} we show the speedup results of all four instances on the Rhea cluster using up 
+to 3072 MPI ranks/cores on Flickr data.
 The speedup values are with respect to slowest runtime among all four instances using 16 cores~(single node).
-We observe in \cref{fig:flickr-speedup-16,fig:flickr-speedup-48} that all algorithms scale up to 1536 cores, yet \mpifaun instances achieve this with significantly lower parallel efficiency.
+%We observe in \cref{fig:flickr-speedup-16,fig:flickr-speedup-48} that all algorithms scale up to 1536 cores, yet \mpifaun instances 
+We observe in \cref{fig:flickr-speedup-48} that all algorithms scale up to 1536 cores, yet \mpifaun instances 
+achieve this with significantly lower parallel efficiency.
 This mostly is due to higher communication costs involved in the all-to-all communication scheme for both instances.
@@
 
-Similarly, in \cref{fig:delicious-speedup-16,fig:delicious-speedup-48} we provide the same results for the Delicious matrix.
+%Similarly, in \cref{fig:delicious-speedup-16,fig:delicious-speedup-48} we provide the same results for the Delicious matrix.
+Similarly, in \cref{fig:delicious-speedup-48} we provide the same results for the Delicious matrix.
 We observe a similar trend in the comparisons of different methods, except that \urp and \unp scale even worse in this case.
 Our algorithm also loses scalability after 1536 processes, and similarly to the previous case \cpp starts slower than \crp due to load imbalance, and catches up for $k = 48$.
-These two test cases clearly show that employing a point-to-point communication with good partitionings is essential for obtaining high performance inNMF algorithm.
+These two test cases clearly show that employing a point-to-point communication with good partitionings is essential for obtaining high performance in NMF algorithm.
 
@@
 \addplot  table [x={edgefactor}, y expr=(\thisrow{cr}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comm.dat};
-\addplot  table [x={edgefactor}, y expr=(\thisrow{cu}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comm.dat};\legend{collective, cp1d, cp2d, cr, cu}
+\addplot  table [x={edgefactor}, y expr=(\thisrow{cu}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comm.dat};\legend{COL, CH1, CH2, CRD, CUN}
 \end{axis}

### 2018-04-10 latex/icpp18/algorithm.tex
reintroduced figure1 and remove k=16 references in experiments

@@
 
-%\begin{figure}
-%\caption{A 5x5 checkerboard partition of a sparse matrix.}
-%\label{fig:partition}
-%\includegraphics[width=0.8\linewidth]{figures/matrix-partition.png}
-%\end{figure}
+\begin{figure}
+\caption{A 5x5 checkerboard partition of a sparse matrix.}
+\label{fig:partition}
+\includegraphics[width=0.8\linewidth]{figures/matrix-partition.png}
+\end{figure}
 
@@
 Partitioning this hypergraph into $P_c$ parts finalizes the $P_r \times P_c$ checkerboard partitionin by balancing the weights~(number of nonzeros of $\Am_{(r, c)}$) of each part while minimizing the communication volume.
-In \textbf{CH2},
+In the context of NMF, one potential issue arises when the matrix has a large variance in the numbe r of nonzeros in its rows/columns, which in turns yields unbalanced row/column strides.
+This in turn creates an imbalance in the NLS computations as rows/cols of $\HH$/$\WW$ are distributed to the processes in the same stride.
+To alleviate this issue, we modify this scheme slightly, which we call \textbf{CH2}.
+In both row and column partitioning phases, we add an additional constant weight to vertices.
+Balancing this additional constraint in hypergraph partitioning is expected to prevent such imbalanced strides.
+This partitioning model succesfully grasps the computation~(both SpMM and NLS) and communication requirements using checkerboard topology for sparse NMF, yet it is very costly to compute in practice due to high number of constraints in the row-net hypergraph.
 

### 2018-04-10 latex/icpp18/experiments.tex
synthetic dataset

@@
 \subsection{Experimental Setup}
-
+In this paper, we use both synthetic and realworld datasets. The synthetic datasets are used to evaluate the communication
+ strategies with the increase in number of non-zeros of the sparse matrices. 
 \subsubsection{Datasets}\label{sec:datasets}
 
-We use two datasets from Flickr.com and Delicious.com that involves images tagged with different labels by users.
-The rows of the matrix correspond to different images, whereas the columns of the matrix represent different tags.
-The value of each nonzero $a_{i,j} \in \Am$ indicates the number of unique users that tagged the image $i$ with the tag $j$.
+{\em Synthetic:} For synthetic sparse matrices, we used the popular Kronecker generator from Graph500 benchmark (\url{http://www.graph500.org/}).
+The graph generator is a Kronecker generator similar to the Recursive MATrix (R-MAT) scale-free graph generation algorithm \cite{CZF04}.
+This model recursively sub-divides the adjacency matrix of the graph into four equal-sized partitions and distributes edges within these partitions with unequal probabilities.
+Initially, the adjacency matrix is empty, and edges are added one at a time.
+The parameters to the generator are the number of vertices $N$ and the Edge Factor ($ef$) that defines the ratio of the graph?s edge count to its vertex count (i.e., half the average degree of a vertex in the graph).
+Typically, the number of edges $M$ of the scale free graph will be $ef \times N$.
+In our case, we consider $2^{20}$($\approx$ 1.05 million) vertices and set of edge factor $ef$=\{4,8,16,32,64\}.
+
+{\em Real World:} We use two datasets from Flickr.com and Delicious.com that involves images tagged with different labels by users.
+The rows of the matrix correspond to different images, whereas the columns of the matrix represent different tags. The value of 
+each nonzero $a_{i,j} \in \Am$ indicates the number of unique users that tagged the image $i$ with the tag $j$.
 Flickr and Delicious matrices are of size 28Mx1.6M and 17Mx2.5M, and have 112M and 72M nonzero elements, respectively.

### 2018-04-10 latex/icpp18/experiments.tex
synthetic dataset

@@
 Initially, the adjacency matrix is empty, and edges are added one at a time.
-The parameters to the generator are the number of vertices $N$ and the Edge Factor ($ef$) that defines the ratio of the graph?s edge count to its vertex count (i.e., half the average degree of a vertex in the graph).
+The parameters to the generator are the number of vertices $N$ and the Edge Factor ($ef$) that defines the ratio of the graph's edge count to its vertex count (i.e., half the average degree of a vertex in the graph).
 Typically, the number of edges $M$ of the scale free graph will be $ef \times N$.

### 2018-04-10 latex/icpp18/algorithm.tex

--- 2018-04-10 latex/icpp18/algorithm.tex
@@
 \paragraph{Checkerboard hypergraph partitioning~(\textbf{CH2})}
-To this end, we adopt hypergraph partitioning techniques used in parallel sparse matrix literature to our case~\cite{caay:99,aycu:08}.
 A hypergraph consists of vertices with associated weights and hyperedges that connect two or more vertices.
@@
 
-Traditional checkerboard hypergraph partitioning aims to partition the matrix $\Am$ into $P_r$ row slices first, and $P_c$ column slices next to obtain an $P_r \times P_c$ checkerboard partition.
+Traditional checkerboard hypergraph partitioning aims to partition the matrix $\Am$ into $P_r$ row slices first, and $P_c$ column slices next to obtain an $P_r \times P_c$ checkerboard partition~\cite{caay:99,aycu:08}.
 The first partitioning phase is done using a column-net hypergraph model, in which for each row $\Am(i, :)$, a vertex $v_i$ with weight equaling to the number of nonzeros in $\Am(i, :)$ is created.
@@
 Partitioning this hypergraph into $P_c$ parts finalizes the $P_r \times P_c$ checkerboard partitionin by balancing the weights~(number of nonzeros of $\Am_{(r, c)}$) of each part while minimizing the communication volume.
-In \textbf{CH2},
+In the context of NMF, one issue arises when the matrix has some variance in the number of nonzeros in its rows/columns, which in turns yields unbalanced row/column strides.
+This in turn creates an imbalance in the NNLS computations as rows/cols of $\WW$/$\HH$ are partitioned to the processes in the same stride.
+To alleviate this issue, we modify this scheme slightly as follows.
+In both row and column partitioning phases, we add an additional constant weight to vertices.
+Balancing this additional constraint in hypergraph partitioning is expected to prevent such imbalanced strides.
+This partitioning model~(which we call \textbf{CH2}) succesfully grasps the computation~(both SpMM and NNLS) and communication requirements using checkerboard topology for sparse NMF, yet is costly to compute in practice due to high number of constraints~($P_r + 1$) in the row-net hypergraph.
 
@@
 \paragraph{1D-like checkerboard hypergraph partitioning~(\textbf{CH1})}
+This variant partitions rows same as $\textbf{CH2}$, then partitions columns randomly to avoid multi-constraint partitioning.
+Random column partition provides load and communication balance yet increases the communication volume for the rows of $\WW$.
 
-\paragraph{Randomized checkerboard partitioning(\textbf{CRD})}
-This scheme corresponds to partitioning both the rows and the columns of $\Am$ into $R$ segments randomly.
-It is expected to provide good load balance both in sparse and dense matrix operations, but it overlooks the communication cost.
+\paragraph{Randomized checkerboard partitioning(\textbf{CR})}
+This scheme corresponds to partitioning both the rows and the columns of $\Am$ into $P_r$ and $P_c$ segments randomly.
+It is expected to provide good load and communication balance both in sparse and dense matrix operations, but it overlooks the communication volume.
 
-\paragraph{Uniform checkerboard partitioning(\textbf{CUN})}
-This partitioning variant forms an $R \times C$ partition of $\Am$ by putting a contiguous set of $m / R$ and $n / C$ rows and columns in each slice.
-$\Wm$ and $\Hm$ are partitioned conformally with this topology; each process is assigned a contiguous set of  $m / P$ and $n / P$ rows and columns, respectively, belonging to the corresponding row and column blocks of $\Hm$.
-This is the only partitioning scheme employed by $\mpifaun$ \cite{KBP16, KBP16MPIFAUN}, hence we only use it for this implementation.
-We also use a randomized variant (\urp) of this scheme in which the rows and columns of $\Am$ are permuted randomly before executing $\mpifaun$ to better balance its nonzeros.
+\paragraph{Uniform checkerboard partitioning(\textbf{CN})}
+This partitioning variant forms an $P_R \times P_C$ partition of $\Am$ by putting a contiguous set of $m / P_R$ and $n / P_C$ rows and columns in each slice.
+$\Wm$ and $\Hm$ are partitioned conformally with this topology; each process is assigned a contiguous set of  $m / P_R P_C$ and $n / P_R P_C$ rows and columns of $\Wm$ and $\Hm$.
+This is the partitioning scheme employed by $\mpifaun$~\cite{KBP16, KBP16MPIFAUN}.
+It provides perfect balance in NNLS step yet may incur high communication cost.
+We also use a randomized variant (\urp) of this scheme in which the rows and columns of $\Am$ are permuted randomly to balance its nonzeros among parts.
+
+\paragraph{Fine-grain hypergraph partitioning~(\textbf{FH})}
+This is the partitioning strategy employed by \hypertensor.
 

### 2018-04-10 latex/icpp18/algorithm.tex
partitioning done

@@
 \subsubsection{Partitioning $\Wm$ and $\Hm$}
-Once $\Am$ is partitioned with an $R \times C$ checkerboard topology, one has to partition rows and columns of factor matrices conformally with this topology to form the sets $\Ip$ and $\Jp$ in \cref{alg:distnmf}.
-In doing so, we are interested in assigning rows and columns to processes in the corresponding process row or column equitably.
-For this purpose, we specify imbalance parameters $\alpha_r$ and $\alpha_c$ that correspond to maximum imbalance we allow in this partitioning; i.e., $|\Ip| \le \alpha_r m / P$ and $|\Jp| \le \alpha_c n / P$ for each process $p$.
-We normally set these imbalance parameters to $1.1$.
-However, when $m >> n$, we increase $\alpha_c$ proportionally in order to give more flexibility in column partitioning which results in smaller communication volume.
+Once $\Am$ is partitioned, one has to partition rows and columns of factor matrices to form the sets $\Ip$ and $\Jp$ in \cref{alg:distnmf}.
+In doing so, we are interested in assigning rows and columns to processes equitably.
+For this purpose, we specify imbalance parameters $\alpha$ that correspond to maximum imbalance we allow in this partitioning; i.e., $|\Ip| \le \alpha m / P$ and $|\Jp| \le \alpha n / P$ for each process $p$, and set $\alpha = 1.05$ in the experiments.
 
@@
 Finally, we randomly assign each row and column to one of the processes satisfying the imbalance constraint in this list.
-If all processes in the list are overloaded, we randomly assign it to a process that is not overloaded in the same process row or column.
-Note that the latter assignment increases the communication volume due to that row or column by 1; hence, in general larger imbalance parameters yield smaller communication volume by reducing this type of assignment.
+If all processes in the list are overloaded, we assign it to the process that has the minimum number of rows/columns assigned.
+For a checkerboard partition, the minimum is always chosen from the same processor row/column so that 2D communication topology is not disturbed.
+Note that such an assignment increases the communication volume due to that row or column by 1; hence, in general smaller imbalance parameters yield larger communication volume due to increasing this type of assignment.
+On the other hand, we desire to keep $\alpha$ small as it pertains to the load imbalance in the NNLS step.
 

### 2018-04-10 latex/icpp18/algorithm.tex
finegrain

@@
 
-\paragraph{Randomized checkerboard partitioning(\textbf{CR})}
+\paragraph{Randomized checkerboard partitioning(\textbf{CRD})}
 This scheme corresponds to partitioning both the rows and the columns of $\Am$ into $P_r$ and $P_c$ segments randomly.
@@
 This is the partitioning strategy employed by \hypertensor.
+It forms a fine-grain hypergraph involving a vertex for each nonzero $(\Am(i, j)$ and a hyperedge for each row and column index $i = 1, \dots, m$ and $j = 1, \dots, n$.
+The resulting hypergraph is typically very large and is costly to partition, and unlike checkerboard variants fingerprint of processes are not restricted to a row/column stride.
 

### 2018-04-10 latex/icpp18/experiments.tex
partitioning done

@@
 Flickr and Delicious matrices are of size 28Mx1.6M and 17Mx2.5M, and have 112M and 72M nonzero elements, respectively.
-The current implementation of \mpifaun can only operate when $R$ and $C$ can divide $m$ and $n$, hence we trimmed the matrices slightly.
+The current implementation of \mpifaun can only operate when $P_R$ and $P_C$ can divide $m$ and $n$, hence we trimmed the matrices slightly.
 
@@
 \addplot  table [x={edgefactor}, y expr=(\thisrow{cr}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comm.dat};
-\addplot  table [x={edgefactor}, y expr=(\thisrow{cu}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comm.dat};\legend{COL, CH1, CH2, CRD, CUN}
+\addplot  table [x={edgefactor}, y expr=(\thisrow{cu}/(\minvalue*\numiterations))] {data/nnzsweepall_k48_p4096_comm.dat};\legend{COL, CH1, CH2, CR, CN}
 \end{axis}
@@
 
+\caption{Communication and computation costs of different partitioning strategies on Kronecker graphs}
 \end{figure}

### 2018-04-10 latex/icpp18/experiments.tex
effect of comm scheme

@@
 \newcommand{\NLS}{LUC }
+\newcommand{\krfaun}{\textbf{COL}\xspace}
+\newcommand{\krcun}{\textbf{CN}\xspace}
 
@@
 \subsection {Effect of communication scheme}
+In \cref{fig:nnzsweep} we provide per-iteration computation and communication timings for different partitionings and communication schemes using $4096$ processors and a $64x64$ processor grid.
+\krfaun is the is the implementation of \mpifaun and is the only instance that uses collective communication.
+Comparing it with the \krcun that also uses uniform partitioning but with point-to-point communication, we notice that \mpifaun always incurs significantly higher communication cost.
+As the edge factor increases and matrix becomes denser, the communication cost of \mpifaun does not change as expected since it does not depend on the matrix sparsity.
+The communication cost of \krcun progressively increases with the edge factor, but even with an edge factor of $64$ \mpifaun takes about more time by a factor of two.
+Therefore, we conclude that even though there is a converging trend between the cost of collective and point-to-point communication, it is a rather slow convergence, and the point-to-point scheme should be the method of choice unless the matrix gets very dense.
 
-\subsection {Effect of partitioning}
+\subsection {Effect of partitioning strategies}
 
@@
 \caption{Communication and computation costs of different partitioning strategies on Kronecker graphs}
+\label{fig:nnzsweep}
 \end{figure}

### 2018-04-10 latex/icpp18/abstract.tex
submitted fist draft icpp18

@@
 In this paper, we focus on scaling algorithms for NMF to very large sparse datasets and massively parallel machines by employing effective algorithms, communication patterns, and partitioning schemes that leverage the sparsity of the input matrix.
-In particular, we consider two previous works developed for related problems, one that uses a fine-grained partitioning strategy using a point-to-point communication pattern and on that uses a checkerboard partitioning strategy using a collective-based communication pattern.
+In the case of machine learning workflow, the 
+computations after SpMM must deal with dense matrices, as Sparse-Dense matrix multiplication will result in a dense matrix. 
+Hence, the partitioning strategy considering only SpMM will result in a huge imbalance in the overall workflow especially on computations after SpMM and in this specific case of NMF on non-negative least squares computations.
+Towards this, we consider two previous works developed for related problems, one that uses a fine-grained partitioning strategy using a point-to-point communication pattern and on that uses a checkerboard partitioning strategy using a collective-based communication pattern.
 We show that a combination of the previous approaches balances the demands of the various computations within NMF algorithms and achieves high efficiency and scalability.
-\grey{To be completed...
-
-We present experiments on (THESE DATASETS), running on up to (YY) cores of a Cray XK7 supercomputer.
-(MAYBE STATE SIMPLE VERSION OF CONCLUSIONS?)
-}
+From the experiments, we could see that our proposed algorithm communicates atleast
+4x less than the collective and achieves upto 100x speed over the baseline FAUN on real world datasets.
+Our algorithm was experimented in two different super computing platforms and we could scale up to 32000 processors on Bluegene/Q.
 

### 2018-04-10 latex/icpp18/experiments.tex
submitted fist draft icpp18

--- 2018-04-10 latex/icpp18/experiments.tex
@@
 
-\subsection {Effect of communication scheme}
+%\subsection {Effect of communication scheme}
 
 \subsection {Effect of partitioning}
+To understand the effect of partitioning and its impact on communication and computation, we performed
+experiments on synthetic Kronecker graph and the results are presented in Figure \ref{fig:kronecker}. 
+For $2^{20}$ vertices, we chose the edge factor of the kronecker graph as \{4,8,16,32,64\} and in
+this setting as the edge factor increases,  higher the number of  non-zeros making the sparse matrix
+denser. We ran the baseline FAUN algorithm and the proposed algorithm with different partitioning 
+scheme as explained in Section \ref{sec:partitioning} on 4096 processors in Rhea for low rank $k=48$. 
+From the Figure \ref{fig:nnzsweepcomm},  it is evident that checkerboard partitioning with point-to-point
+communication is approximately 4x faster over the 2D partition with collective communication. In the case of
+computation, from Figure \ref{fig:nnzsweepcomp}, we can observe that in the case of denser matrices
+with higher edge factors, the load imbalance in 2D is becoming more evident over the checkerboard partition. 
+Thus checker board with point-to-point communicates less and has better load balance over 2D partition
+with collective communication. These advantages  give scope for the checkerboard partition with point-to-point scale
+to higher number of nodes.
 
@@
 
-\caption{Communication and computation costs of different partitioning strategies on Kronecker graphs}
+\caption{Communication and computation costs of different partitioning strategies on Kronecker graphs on 4096 processors Rhea for $k=48$}
+\label{fig:kronecker}
 \end{figure}

### 2018-04-10 latex/icpp18/experiments.tex

@@
 %\subsection {Effect of communication scheme}
+\subsection {Effect of communication scheme}
+In \cref{fig:nnzsweep} we provide per-iteration computation and communication timings for different partitionings and communication schemes using $4096$ processors and a $64x64$ processor grid.
+\krfaun is the is the implementation of \mpifaun and is the only instance that uses collective communication.
+Comparing it with the \krcun that also uses uniform partitioning but with point-to-point communication, we notice that \mpifaun always incurs significantly higher communication cost.
+As the edge factor increases and matrix becomes denser, the communication cost of \mpifaun does not change as expected since it does not depend on the matrix sparsity.
+The communication cost of \krcun progressively increases with the edge factor, but even with an edge factor of $64$ \mpifaun takes about more time by a factor of two.
+Therefore, we conclude that even though there is a converging trend between the cost of collective and point-to-point communication, it is a rather slow convergence, and the point-to-point scheme should be the method of choice unless the matrix gets very dense.
 
@@
 \end{subfigure}
-
 \caption{Communication and computation costs of different partitioning strategies on Kronecker graphs on 4096 processors Rhea for $k=48$}
-\label{fig:kronecker}
+\label{fig:nnzsweep}
 \end{figure}

### 2018-04-10 latex/icpp18/experiments.tex
progress

--- 2018-04-10 latex/icpp18/experiments.tex
@@
 \newcommand{\krcun}{\textbf{CN}\xspace}
+\newcommand{\krcr}{\textbf{CR}\xspace}
+\newcommand{\krch1}{\textbf{CH1}\xspace}
+\newcommand{\krch2}{\textbf{CH2}\xspace}
 
@@
  strategies with the increase in number of non-zeros of the sparse matrices. 
+ We used PaToH~\cite{caay:99} for partitioning hypergraphs.
 \subsubsection{Datasets}\label{sec:datasets}
@@
 \subsection {Effect of partitioning strategies}
+In \cref{fig:nnzsweepcomm}, our first observation is that all partitionings except \krfaun yield comparable results in terms of communication, while there is some variation  in the computation times provided in \cref{fig:nnzsweepcomp}.
+
+\krcun gives similar k
 

### 2018-04-10 latex/icpp18/experiments.tex

--- 2018-04-10 latex/icpp18/experiments.tex
@@
 \label{sec:experiment}
+
 \newcommand{\NLS}{LUC }
+\newcommand{\krfaun}{\textbf{COL}\xspace}
+\newcommand{\krcun}{\textbf{CN}\xspace}
+\newcommand{\krcha}{\textbf{CH1}\xspace}
+\newcommand{\krchb}{\textbf{CH2}\xspace}
+\newcommand{\krcr}{\textbf{CR}\xspace}
+\newcommand{\krfh}{\textbf{FH}\xspace}
 
@@
 
-\subsubsection{Algorithms}
-
-In our experiments, we considered the following algorithms and partitionings:
-\begin{itemize}
-	\item \unp: \mpifaun Algorithm \cite{KBP16,KBP16MPIFAUN} with uniform natural  partition~(\unp) where each process holds an input matrix of size $m/P_r \times n/P_c$. 
-	\item \urp: The partitioning strategy in \unp\ could result in a significant computational load imbalance in with a skewed nonzero distribution of $\Am$.
-We alleviate this by randomly permuting the rows and columns of the matrix before executing \mpifaun, and call this scheme $\urp$.
-	\item \cpp: \distspnmf (\cref{alg:distnmf}) with checkerboard hypergraph partitioning explained in \cref{sec:cp1d}.
-	\item \crp: \distspnmf (\cref{alg:distnmf}) with randomized checkerboard partioning explained in \cref{sec:cp1d}.
-
-\end{itemize}
-
 %The Figure \ref{fig:stackexchangerr} shows the comparison of these algorithms on the stack exchange dataset explained in Section \ref{sec:datasets}.
@@
 %\subsection {Effect of communication scheme}
-
-\subsection {Effect of partitioning}
+\subsection {Effect of communication scheme}
 To understand the effect of partitioning and its impact on communication and computation, we performed
-experiments on synthetic Kronecker graph and the results are presented in Figure \ref{fig:kronecker}. 
+experiments on synthetic Kronecker graph and the results are presented in \cref{fig:nnzsweep}. 
 For $2^{20}$ vertices, we chose the edge factor of the kronecker graph as \{4,8,16,32,64\} and in
@@
 denser. We ran the baseline FAUN algorithm and the proposed algorithm with different partitioning 
-scheme as explained in Section \ref{sec:partitioning} on 4096 processors in Rhea for low rank $k=48$. 
+scheme as explained in \cref{sec:partitioning} on 4096 processors using a $64 \times 64$ processor grid on Rhea for low rank $k=48$.
+
+In \cref{fig:nnzsweep}, we provide per-iteration computation and communication timings for different partitionings and communication schemes using $4096$ processors and a $64x64$ processor grid.
+\krfaun is the is the implementation of \mpifaun and is the only instance that uses collective communication.
+Comparing it with the \krcun that also uses uniform partitioning but with point-to-point communication.
+\krcr decides the checkerboard topology randomly and uses point-to-point scheme as well.
+\krcha and \krchb correspond to 1D-like and 2D checkerboard hypergraph partitioning models with point-to-point communication.
+Finally, \krfh is the fine-graph hypergraph partitioning with point-to-point scheme, for which we only provide partition statistics for comparison and not the actual run time.
+
+We notice in \cref{fig:nnzsweep} that \krfaun always incurs significantly higher communication cost.
+As the edge factor increases and matrix becomes denser, the communication cost of \krfaun does not change as expected since it does not depend on the matrix sparsity.
+The communication cost of \krcun progressively increases with the edge factor, but even with an edge factor of $64$ \krfaun is about two times more costly.
+Therefore, we conclude that even though there is a converging trend between the cost of collective and point-to-point communication, it is a rather slow convergence, and the point-to-point scheme should be the method of choice unless matrix gets very dense.
+
+\subsection {Effect of partitioning strategies}
+In \cref{fig:nnzsweepcomm}, our first observation is that all partitionings except \krfaun yield comparable results in terms of communication, while there is some variation  in the computation times provided in \cref{fig:nnzsweepcomp}.
+\krcun gives similar k
+
 From the Figure \ref{fig:nnzsweepcomm},  it is evident that checkerboard partitioning with point-to-point
@@
 
+Therefore, we conclude that even though \krchb correctly models computation and communication costs of distributed sparse NMF, it does not yield better results in practice due to having too many constraints in the partitioning problem, which also renders it impractical due to costly precomputation for partitioning.
+\krcha yields a smaller hypergraph with reasonable partitioning cost and provides some benefits in terms of communication reduction.
+\krcr seems to produce good results overall as it provides good computation and communication balance.
+With these observations, we provide strong scalability experiments on real-world datasets in the next section with a detailed comparison of point-to-point and collective strategies as well has randomized and hypergraph checkerboard~(1d-like) partitioning.
+
 \subsection {Strong scaling}
+
+In this experiment, we considered the following algorithms and partitionings:
+\begin{itemize}
+	\item \unp: \mpifaun algorithm \cite{KBP16,KBP16MPIFAUN} with uniform partitioning~(\unp) where each process holds an input matrix of size $m/P_r \times n/P_c$. 
+	\item \urp: The partitioning strategy in \unp\ could result in a significant computational load imbalance in with a skewed nonzero distribution of $\Am$.
+We alleviate this by randomly permuting the rows and columns of the matrix before executing \mpifaun, and call this scheme $\urp$.
+	\item \cpp: \distspnmf (\cref{alg:distnmf}) with 1D-like checkerboard hypergraph partitioning explained in \cref{sec:cp1d}.
+	\item \crp: \distspnmf (\cref{alg:distnmf}) with randomized checkerboard partioning explained in \cref{sec:cp1d}.
+
+\end{itemize}
 

### 2018-04-10 latex/icpp18/experiments.tex
progress

@@
 \krcr seems to produce good results overall as it provides good computation and communication balance.
-With these observations, we provide strong scalability experiments on real-world datasets in the next section with a detailed comparison of point-to-point and collective strategies as well has randomized and hypergraph checkerboard~(1d-like) partitioning.
+We provide partition statistics with respect to these strategies in Table ?? which justify these observations.
 
 \subsection {Strong scaling}
+In this section, we provide strong scalability experiments on real-world datasets in the next section with a detailed comparison of point-to-point and collective strategies as well has randomized and hypergraph checkerboard~(1d-like) partitioning.
 

### 2018-04-10 latex/icpp18/experiments.tex
prog

@@
 
-Therefore, we conclude that even though \krchb correctly models computation and communication costs of distributed sparse NMF, it does not yield better results in practice due to having too many constraints in the partitioning problem, which also renders it impractical due to costly precomputation for partitioning.
+We conclude that even though \krchb correctly models computation and communication costs of distributed sparse NMF, it does not yield better results in practice due to having too many constraints in the partitioning problem, which also renders it impractical due to costly precomputation for partitioning.
 \krcha yields a smaller hypergraph with reasonable partitioning cost and provides some benefits in terms of communication reduction.
@@
 We provide partition statistics with respect to these strategies in Table ?? which justify these observations.
+
+\begin{table}
+\begin{center}
+\caption{Load balance and communication statistics for 256-way partitioning.\label{tab:loadcomm}}
+\vspace*{-1ex}
+\begin{footnotesize}
+  \begin{tabular}{ l | r | r | r | r | r }
+%    \multirow{2}{*}{Partitioning} & \multicolumn{2}{|c|}{Nnz } & \multicolumn{2}{|c|}{Comp. Load} & \multicolumn{2}{|c|}{Comm. Vol.} & \multicolumn{2}{|c}{Num. Msg} \\ \cline{2-9}
+    Partitioning & nz-ib  & row-ib      & col-ib     & com-max.      & com-avg \\ \hline
+%    \multicolumn{6}{c}{\emph{Delicious}} \\ \hline
+    fine-hp   & 547K      & 547K        & 1947K      & 1807K         & 199K      \\ \hline
+    fine-lb   & 564K      & 547K        & 1849K      & 1737K         & 309K      \\ \hline
+    fine-rd   & 550K      & 547K        & 2747K      & 2737K         & 1083K     \\ \hline
+    medium-gd & 598K      & 547K        & 2353K      & 2214K         & 624K      \\ \hline
+  \end{tabular}
+  \end{footnotesize}
+\end{center}
+\vspace*{-1ex}
+\end{table}
 

### 2018-04-10 latex/icpp18/experiments.tex
minor
partitioning done

@@
 In \cref{fig:nnzsweepcomm}, our first observation is that all partitionings except \krfaun yield comparable results in terms of communication, while there is some variation  in the computation times provided in \cref{fig:nnzsweepcomp}.
-\krcun gives similar k
-
-From the Figure \ref{fig:nnzsweepcomm},  it is evident that checkerboard partitioning with point-to-point
-communication is approximately 4x faster over the 2D partition with collective communication. In the case of
-computation, from Figure \ref{fig:nnzsweepcomp}, we can observe that in the case of denser matrices
-with higher edge factors, the load imbalance in 2D is becoming more evident over the checkerboard partition. 
-Thus checker board with point-to-point communicates less and has better load balance over 2D partition
-with collective communication. These advantages  give scope for the checkerboard partition with point-to-point scale
-to higher number of nodes.
+\krcun gives similar computation time to \krcr since matrix is randomly permuted.
+\krcha yields slightly higher computation time and no notable improvement in the communication time, as it involves multi-constraint partitioning with too many constraints, which is a difficult partitioning problem that partitioners such as PaToH might not solve very effectively.
+\krchb provides some advantage both in terms of computation and communication time, and this partitioning is feasible to compute as it involves hypergraph partitioning with only two constraints.
 
@@
 %    \multicolumn{6}{c}{\emph{Delicious}} \\ \hline
-    fine-hp   & 547K      & 547K        & 1947K      & 1807K         & 199K      \\ \hline
-    fine-lb   & 564K      & 547K        & 1849K      & 1737K         & 309K      \\ \hline
-    fine-rd   & 550K      & 547K        & 2747K      & 2737K         & 1083K     \\ \hline
-    medium-gd & 598K      & 547K        & 2353K      & 2214K         & 624K      \\ \hline
+    \krcun & 547K      & 547K        & 1947K      & 1807K         & 199K      \\ \hline
+    \krcha & 564K      & 547K        & 1849K      & 1737K         & 309K      \\ \hline
+    \krchb & 550K      & 547K        & 2747K      & 2737K         & 1083K     \\ \hline
+    \krcr  & 598K      & 547K        & 2353K      & 2214K         & 624K      \\ \hline
+    \krfh  & 598K      & 547K        & 2353K      & 2214K         & 624K      \\ \hline
   \end{tabular}

### 2018-04-10 latex/icpp18/experiments.tex

@@
 \krcr seems to produce good results overall as it provides good computation and communication balance.
-We provide partition statistics with respect to these strategies in Table ?? which justify these observations.
+We provide partition statistics with respect to these strategies in Table ref{tab:loadcomm} which justify these observations.
 
@@
 \begin{center}
-\caption{Load balance and communication statistics for 256-way partitioning.\label{tab:loadcomm}}
+\caption{Load balance and communication statistics for 4096-way partitioning with edge factor 64.\label{tab:loadcomm}}
 \vspace*{-1ex}
@@
 %    \multicolumn{6}{c}{\emph{Delicious}} \\ \hline
-    \krcun & 547K      & 547K        & 1947K      & 1807K         & 199K      \\ \hline
-    \krcha & 564K      & 547K        & 1849K      & 1737K         & 309K      \\ \hline
-    \krchb & 550K      & 547K        & 2747K      & 2737K         & 1083K     \\ \hline
-    \krcr  & 598K      & 547K        & 2353K      & 2214K         & 624K      \\ \hline
-    \krfh  & 598K      & 547K        & 2353K      & 2214K         & 624K      \\ \hline
+    \krcun & 1.30987      & 1        & 1.00392      & 15366         & 11876      \\ \hline
+    \krcha & 1.21477      & 1.27734        & 1.05098      & 20800         & 11038      \\ \hline
+    \krchb & 1.06479      & 1.30469        & 1.53725      & 23484         & 10853     \\ \hline
+    \krcr  & 1.38681      & 1.04688        & 1.05098      & 15475         & 11341      \\ \hline
+    \krfh  & 1.00115      & 1.05078        & 1.0549      & 47925         & 13802      \\ \hline
   \end{tabular}

### 2018-04-10 latex/icpp18/experiments.tex
imbalance table

--- 2018-04-10 latex/icpp18/experiments.tex
@@
 %    \multicolumn{6}{c}{\emph{Delicious}} \\ \hline
-    fine-hp   & 547K      & 547K        & 1947K      & 1807K         & 199K      \\ \hline
-    fine-lb   & 564K      & 547K        & 1849K      & 1737K         & 309K      \\ \hline
-    fine-rd   & 550K      & 547K        & 2747K      & 2737K         & 1083K     \\ \hline
-    medium-gd & 598K      & 547K        & 2353K      & 2214K         & 624K      \\ \hline
+    \krcun & 1.30987      & 1        & 1.00392      & 15366         & 11876      \\ \hline
+    \krcha & 1.21477      & 1.27734        & 1.05098      & 20800         & 11038      \\ \hline
+    \krchb & 1.06479      & 1.30469        & 1.53725      & 23484         & 10853     \\ \hline
+    \krcr  & 1.38681      & 1.04688        & 1.05098      & 15475         & 11341      \\ \hline
+    \krfh  & 1.00115      & 1.05078        & 1.0549      & 47925         & 13802      \\ \hline
   \end{tabular}

### 2018-04-10 latex/icpp18/experiments.tex
minor

--- 2018-04-10 latex/icpp18/experiments.tex
@@
 \krcr seems to produce good results overall as it provides good computation and communication balance.
-We provide partition statistics with respect to these strategies in Table ref{tab:loadcomm} which justify these observations.
+We provide partition statistics with respect to these strategies in \cref{tab:loadcomm} which justify these observations.
+We observe that \krfh incurs a significant communication imbalance with respect to other methods preventing scalability and increasing memory consumption of the bottleneck processor, and these results confirm those reported in \cite{KU15}.
 

### 2018-04-10 latex/icpp18/abstract.tex
one more submission icpp18

@@
 From the experiments, we could see that our proposed algorithm communicates atleast
-4x less than the collective and achieves upto 100x speed over the baseline FAUN on real world datasets.
+4x less than the collective and achieves upto 100x speed up over the baseline FAUN on real world datasets.
 Our algorithm was experimented in two different super computing platforms and we could scale up to 32000 processors on Bluegene/Q.

### 2018-04-10 latex/icpp18/conclusion.tex
conc

\section{conclusion}

In this paper, we compared various partitioning and communication strategies used in the literature in the context of non-negative matrix factorization.
We showed that an important difference in the parallel NMF algorithms is balancing matrix rows among processors, and this constraint renders state-of-the-art hypergraph partitioning methods less effective.
We employed variations of the MPI-FAUN implementation with point-to-point communication, and concluded that unless matrix at hand is quite dense, point-to-point communication significantly improves the scalability by reducing the communication volume.
With optimized implementations, we achieved scalability up to 32K nodes on a BG/Q supercomputer using partitioning schemes that are cheap to compute.
To the best of our knowledge, our work is the first high performance implementation of distributed NMF that takes the sparsity of the input matrix into consideration in communication to reduce the communication cost, and employs effective partitionings to further enhance parallel scalability.
Our immediate next steps for extending our work involve adding shared memory parallelism to obtain further speedup.

### 2018-05-17 distnmf/distanlsbpp.hpp
changes for gpu is still under progress

@@ -5,8 +5,13 @@
 #include "bppnnls.hpp"
 #include "aunmf.hpp"
 
+#ifdef BUILD_CUDA
+#define ONE_THREAD_MATRIX_SIZE 1000
+#include <omp.h>
+#endif
+
 template<class INPUTMATTYPE>
-class DistANLSBPP : public DistAUNMF<INPUTMATTYPE>{
+class DistANLSBPP : public DistAUNMF<INPUTMATTYPE> {
  private:
   ROWVEC localWnorm;
   ROWVEC Wnorm;
@@ -14,14 +19,67 @@ class DistANLSBPP : public DistAUNMF<INPUTMATTYPE>{
   void allocateMatrices() {
   }
 
+  void updateOtherGivenOneMultipleRHS(MAT &giventGiven, MAT &giventInput,
+                                      MAT *othermat) {
+    UINT numThreads = (giventInput.n_cols / ONE_THREAD_MATRIX_SIZE) + 1;
+    #pragma omp parallel for schedule(dynamic)
+    for (UINT i = 0; i < numThreads; i++) {
+      UINT spanStart = i * ONE_THREAD_MATRIX_SIZE;
+      UINT spanEnd = (i + 1) * ONE_THREAD_MATRIX_SIZE - 1;
+      if (spanEnd > giventInput.n_cols - 1) {
+        spanEnd = giventInput.n_cols - 1;
+      }
+      // if it is exactly divisible, the last iteration is unnecessary.
+      BPPNNLS<MAT, VEC > *subProblem;
+      if (spanStart <= spanEnd) {
+        if (spanStart == spanEnd) {
+          subProblem = new BPPNNLS<MAT, VEC >(giventGiven,
+                                              (VEC)giventInput.col(spanStart),
+                                              true);
+        } else {  // if (spanStart < spanEnd)
+          subProblem = new BPPNNLS<MAT, VEC >(giventGiven,
+                                              (MAT)giventInput.cols(spanStart, spanEnd),
+                                              true);
+        }
+#ifdef MPI_VERBOSE
+        #pragma omp parallel
+        {
+          DISTPRINTINFO ("Scheduling " << worh << " start=" << spanStart
+                         << ", end=" << spanEnd
+                         << ", tid=" << omp_get_thread_num());
+        }
+#endif
+        subProblem->solveNNLS();
+#ifdef MPI_VERBOSE
+        #pragma omp parallel
+        {
[…]

### 2018-05-17 distnmf/distnmf.cpp

@@ -10,6 +10,10 @@
 #include "naiveanlsbpp.hpp"
 #include "utils.hpp"
 #include "parsecommandline.hpp"
+#ifdef BUILD_CUDA
+#include <cuda.h>
+#include <cuda_runtime.h>
+#endif
 
 class DistNMFDriver {
   private:
@@ -31,6 +35,45 @@ class DistNMFDriver {
     int m_num_k_blocks;
     static const int kprimeoffset = 17;
     normtype m_input_normalization;
+
+#ifdef BUILD_CUDA
+    void printDevProp(cudaDeviceProp devProp) {
+        printf("Major revision number: %d\n", devProp.major);
+        printf("Minor revision number: %d\n", devProp.minor);
+        printf("Name: %s\n", devProp.name);
+        printf("Total global memory: %u\n", devProp.totalGlobalMem);
+        printf("Total shared memory per block: %u\n", devProp.sharedMemPerBlock);
+        printf("Total registers per block: %d\n", devProp.regsPerBlock);
+        printf("Warp size: %d\n", devProp.warpSize);
+        printf("Maximum memory pitch: %u\n", devProp.memPitch);
+        printf("Maximum threads per block: %d\n", devProp.maxThreadsPerBlock);
+        for (int i = 0; i < 3; ++i)
+            printf("Maximum dimension %d of block: %d\n", i, devProp.maxThreadsDim[i]);
+        for (int i = 0; i < 3; ++i)
+            printf("Maximum dimension %d of grid: %d\n", i, devProp.maxGridSize[i]);
+        printf("Clock rate: %d\n", devProp.clockRate);
+        printf("Total constant memory: %u\n", devProp.totalConstMem);
+        printf("Texture alignment: %u\n", devProp.textureAlignment);
+        printf("Concurrent copy and execution: %s\n", (devProp.deviceOverlap ? "Yes" : "No"));
+        printf("Number of multiprocessors: %d\n", devProp.multiProcessorCount);
+        printf("Kernel execution timeout: %s\n", (devProp.kernelExecTimeoutEnabled ? "Yes" : "No"));
+        return;
+    }
+    void gpuQuery() {
+        int devCount;
+        cudaGetDeviceCount(&devCount);
+        printf("CUDA Device Query...\n");
+        printf("There are %d CUDA devices.\n", devCount);
+        // Iterate through devices
+        for (int i = 0; i < devCount; ++i) {
+            // Get device properties
[…]

### 2018-05-22 latex/sc18/prelims.tex
fixing typos

@@
 Given matrices $\M{A}$ and $\M{B}$ that are $I_{A} \times R$ and $I_{B} \times R$, the KRP $\M{K} = \M{A} \Khat \M{B}$ is $I_{A}I_{B} \times R$. 
-It can be thought of as a row-wise Hadamard product, where $\M{K}(i+I_{A}(j{-}1),:) = \M{A}(i,:) \Hada \M{B}(j,:)$, or a column-wise Kronecker product, where $\M{K}(,c) =  \M{A}(:,c) \Kron \M{B}(:,c)$.
+It can be thought of as a row-wise Hadamard product, where $\M{K}(i+I_{A}(j{-}1),:) = \M{A}(i,:) \Hada \M{B}(j,:)$, or a column-wise Kronecker product, where $\M{K}(:,c) =  \M{A}(:,c) \Kron \M{B}(:,c)$.
 
@@
 \times I_{N}$ (and $I=\prod I_n$) into the columns of a matrix.
-The Matricized-Tensor Times Khatri-Rao Product or MTTKRP will be central to this work and takes the form $\Mn{M}{n} = \Mz{A}{n} \Mn{S}{n}$, where $\Mn{S}{n}= \Mn{H}{N} \Khat \cdots \Khat \Mn{H}{n+1} \Khat \Mn{H}{n-1} \Khat  \cdots \Khat \Mn{H}{1}$.
+The Matricized-Tensor Times Khatri-Rao Product or MTTKRP will be central to this work and takes the form $\Mn{M}{n} = \Mz{A}{n} \Mn{K}{n}$, where $\Mn{K}{n}= \Mn{H}{N} \Khat \cdots \Khat \Mn{H}{n+1} \Khat \Mn{H}{n-1} \Khat  \cdots \Khat \Mn{H}{1}$.
 
@@
 %\caption{$(\CPl,\epsilon) = \text{NNCP}(\TA,k)$}
-\caption{$\CP = \text{NNCP}(\TA,k)$}
+\caption{$\CP = \text{NNCP}(\TA,R)$}
 \label{alg:nncp}
 \begin{algorithmic}[1]
-\Require $\TA$ is $I_1\times \cdots \times I_N$ tensor, $k$ is approximation rank
+\Require $\TA$ is $I_1\times \cdots \times I_N$ tensor, $R$ is approximation rank
 \State \Comment{Initialize data}

### 2018-05-22 latex/sc18/experiments.tex
communication sweep for different config
one typo fix

@@
 \end{tikzpicture}
-\caption{Strong scaling of 3D synthetic tensor with dimension $1024\times 1024\times 1024$ on processor grids $2^k\times 2^k\times 2^k$ for $k\in\{0,1,2,3\}$.  The rank is fixed at 32.}
+\caption{Strong scaling of 3D synthetic tensor with dimension $1024\times 1024\times 1024$ on processor grids $2^k\times 2^k\times 2^k$ for $k\in\{0,1,2,3,4\}$.  The rank is fixed at 32.}
 \label{fig:strongsynthetic3D}
@@
 
+\begin{figure}
+\renewcommand{\datafile}{data/comm_sweep_scale_pgf_10.dat}
+\renewcommand{\numiterations}{10}
+\begin{tikzpicture}
+\begin{axis}[	
+	ybar stacked,
+	bar width=8pt,
+	width=\columnwidth,
+	height =.75\columnwidth,
+	%width=9cm, height=3.85cm,
+	ylabel={Time (s)}, 
+	xlabel={Rank $R$},
+	y label style={yshift=-.5cm},
+	ymin=0,
+	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
+	symbolic x coords={81x1x1x1-10, 27x3x1x1-10,9x9x1x1-10,9x3x3x1-10,3x3x3x3-10,},
+	%xtick=data,
+	xticklabels={,27x3x1x1,3x3x3x3,81x1x1x1,9x3x3x1,9x9x1x1},
+	legend style={at={(0.5,1.3)},anchor=north},
+	legend columns=-1,
+]
+	\setcolors
+	\addplot table[x=algo-k, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=algo-k, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=algo-k, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=algo-k, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=algo-k, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
+	\legend{PM,mTTV,NLS,Factor Comm,Gram Comm};
+\end{axis}
+%\labels
+\end{tikzpicture}
+\caption{Comm scheme for low rank $k$=10}
+\label{fig:commsweep10}
+\end{figure}
+
 %\begin{figure}

### 2018-05-22 latex/sc18/plots.tex
another minor change

@@
 	blue, fill=blue \\ 
-	green, fill=green \\ 
+	teal, fill=teal \\ 
 	red, pattern=crosshatch, pattern color=red \\
-	green, pattern=crosshatch, pattern color=green \\
+	teal, pattern=crosshatch, pattern color=teal \\
 	blue, pattern=crosshatch, pattern color=blue \\

### 2018-05-22 latex/sc18/experiments.tex
minor fixes

@@
 \end{tikzpicture}
-\caption{Strong scaling of 3D HSI real world data with dimension 1024 x1344 x 33 on processor grids of $1 \times k \times k$ for $k \in {1, 2, 4, 8, 16, 32}$. The rank is fixed at 32.}
+\caption{Strong scaling of 3D HSI real world data with dimension 1024 x1344 x 33 on processor grids of $k \times k\times 1$ for $k \in {1, 2, 4, 8, 16, 32}$. The rank is fixed at 32.}
 \label{fig:stronghsi3D}

### 2018-05-24 latex/sc18/experiments.tex
fixed label ordering for proc grid sweep exp

@@
 	ylabel={Time (s)}, 
-	xlabel={Rank $R$},
+	xlabel={Processor Grid},
 	y label style={yshift=-.5cm},
@@
 	%xtick=data,
-	xticklabels={,27x3x1x1,3x3x3x3,81x1x1x1,9x3x3x1,9x9x1x1},
+	xticklabels={,81x1x1x1,27x3x1x1,9x9x1x1,9x3x3x1,3x3x3x3},
 	legend style={at={(0.5,1.3)},anchor=north},
@@
 \end{tikzpicture}
-\caption{Comm scheme for low rank $k$=10}
+\caption{Time breakdown for $243\times243\times243\times243$ tensor and rank $R=10$ on 81 processors for varying processor grids.}
 \label{fig:commsweep10}

### 2018-05-25 latex/sc18/algorithm.tex
minor fixes
added table and some discussion in analysis section

@@
 
-\subsubsection{Algorithm Overview}
-
-The basic sequential algorithm is given in \Cref{alg:nncp}, and the parallel version is given in \Cref{alg:Par-NNCP-short}.
-We will refer to both the inner iteration, in which one factor matrix is updated (\cref{line:for} to \cref{line:endfor}), and the outer iteration, in which all factor matrices are updated (\cref{line:while} to \cref{line:endwhile}).
-In the parallel algorithm, the processors are organized into a logical multidimensional grid (tensor) with as many modes as the data tensor.
-The communication patterns used in the algorithm are all MPI collectives, including All-Reduce, Reduce-Scatter, and All-Gather.
-The processor communicators (across which the collectives are performed) include the set of all processors and the sets of processors within the same processor slice.
-Processors within a mode-$n$ slice all have the same $n$th coordinate.
-
-The method of enforcing the nonnegativity constraints of the linear least squares solve (or update) generally affects only local computation because each row of a factor matrix can be updated independently.
-In our algorithm, each processor solves the linear problem or computes the update for its subset of rows (see \cref{line:locNLS}). 
-The most expensive (and most complicated) part of the parallel algorithm is the computation of the MTTKRP, which corresponds to \cref{line:locMTTKRP,line:reduce-scatter,line:all-gather}.
-
-The details that are omitted from this presentation of the algorithm include the normalization of each factor matrix after it is computed and the computation of the residual error at the end of an outer iteration.
-The computations do involve both local computation and communication, but their costs are negligible.
-A more detailed pseudocode is given in \Cref{alg:Par-NNCP-long}.
-
-\subsubsection{Data Distribution}
-\label{sec:datadist}
-
-Given a logical processor grid of processors $P_1\times \cdots \times P_N$, we distribute the tensor $\TA$ in a block or Cartesian partition.
-Each processor owns a local tensor of dimensions $(I_1/P_1)\times \cdots \times (I_N/P_N)$, and only one copy of the tensor is stored.
-Locally, the tensor is stored linearly, with entries ordered in a natural mode-descending way that generalizes column-major layout of matrices.
-Given a processor $\V{p}=(\VE{p}{1},\dots,\VE{p}{N})$, we denote its local tensor by $\TA_{\V{p}}$.
-
-Each factor matrix is distributed across processors in a block row partition, so that each processor owns a subset of the rows.
-We use the notation $\Mn{H}{n}_{\V{p}}$, which has dimensions $I_n/P\times R$ to denote the local part of the $n$th factor matrix stored on processor $\V{p}$.
-However, we also make use a redundant distribution of the factor matrices across processors, because all processors in a mode-$n$ processor slice need access to the same entries of $\Mn{H}{n}$ to perform their computations.
-The notation $\Mn{H}{n}_{\VE{p}{n}}$ denotes the $I_n/P_n\times R$ submatrix of $\Mn{H}{n}$ that is redundantly stored on all processors whose $n$th coordinate is $\VE{p}{n}$ (there are $P/P_n$ such processors).
-
-Other matrices involved in the algorithm include $\Mn{M}{n}_{\V{p}}$, which is the result of the MTTKRP computation and has the same distribution scheme as $\Mn{H}{n}_{\V{p}}$, and $\Mn{G}{n}$, which is the $R\times R$ Gram matrix of the factor matrix $\Mn{H}{n}$ and is stored redundantly on all processors.
-
-\subsubsection{Inner Iteration}
-
-The inner iteration is displayed graphically in \Cref{fig:inner} for a 3-way example and an update of the $2$nd factor matrix.
-The main idea is that at the start of the $n$th inner iteration (\cref{fig:inner_a}), all of the data is in place for each processor to perform a local MTTKRP computation.
-This means that all processors in a slice redundantly own the same rows of the corresponding factor matrix (for all modes except $n$).
-After the local MTTKRP is computed (\cref{fig:inner_b}), each processor has computed a contribution to a subset of the rows of the global MTTKRP $\Mn{M}{n}$, but its contribution must be summed up with the contributions of all other processors in its mode-$n$ slice.
-This summation is performed with a Reduce-Scatter collective across the mode-$n$ processor slice that achieves a row-wise partition of the result (in \cref{fig:inner_c}, the light gray shading corresponds to the rows of $\Mn{M}{2}$ to which processor $(1,3,1)$ contributes and the dark gray shading corresponds to the rows it receives as output).
-The output distribution of the Reduce-Scatter is designed so that afterwards, the update of the factor matrix in that mode can be performed row-wise in parallel.
-Along with $\Mn{S}{n}$, which can be computed locally, each processor updates its own rows of the factor matrix given its rows of the MTTKRP result (\cref{fig:inner_d}).
-The remainder of the inner iteration is preparing and distributing the new factor matrix data for future inner iterations, which includes an All-Gather of the newly computed factor matrix $\Mn{H}{n}$ across mode-$n$ processor slices (\cref{fig:inner_e}) and recomputing $\Mn{G}{n}={\Mn{H}{n}}^\Tra\Mn{H}{n}$.
-
-\subsubsection{Analysis}
-
-We will analyze the cost of a single outer iteration.
-While the number of outer iterations is sensitive to the NLS method used, the outer iteration time is generally the same across NLS methods.
-
-\paragraph{\emph{Computation}}
-The local computation occurs at \cref{line:locMTTKRP,line:hadamard,line:locNLS,line:locSYRK}.
-The cost of \cref{line:hadamard} is $O(NR^2)$, the cost of \cref{line:locNLS} is $O(R^3I_n/P)$, which is a loose upper bound for BPP and other methods \cite{KBP16}, and the cost of \cref{line:locSYRK} is $O(R(I_n/P)^2)$.
-The sum of these three costs across all inner iterations is $O(R^2N^2+(R^3/P+R/P^2)\sum I_n)$, which is dominated by the cost of the MTTKRP.
-When using dimension trees to perform the MTTKRP (\cref{line:locMTTKRP}), we compute the cost amortized over all inner iterations.
-In this case, the cost is dominated by the two partial MTTKRP computations (from the root of the tree), which together are $O((R/P) \prod I_n)=O(IR/P)$ and dominate the costs of the multi-TTVs.
-We note that this cost involves the product of all the tensor dimensions, which is why it dominates, and we note that it scales linearly with $P$.
-
-\paragraph{\emph{Communication}}
-The communication within the inner iteration occurs at \cref{line:reduce-scatter,line:all-reduce,line:all-gather}.
-\Cref{line:all-reduce} involves $O(R^2)$ data and a collective across all processors.
-\Cref{line:reduce-scatter,line:all-gather} involve $O(I_nR/P_n)$ data across a subset of $P/P_n$ processors.
-Thus, the All-Reduce dominates the latency cost and the Reduce-Scatter/All-Gather dominate the bandwidth cost, for a total outer iteration communication cost of $O(R\sum I_n/P_n)$ words and $O(N\log P)$ messages.
-If the optimal processor grid can be chosen to minimize communication (assuming $P$ is sufficiently factorable), then the bandwidth cost can achieve a value of $O(RI^{1/N}/P^{1/N})$ by making the local tensors as cubical as possible.
-We note that this cost scales with $P^{1/N}$, which is far from linear scaling.
-However, it is proportional to the geometric mean of the tensor dimensions (on the order of one tensor dimension), which is much less than the computation cost dependence on the product of all dimensions.
-
-\paragraph{\emph{Memory}}
-The algorithm requires extra local memory to run.
-Aside from the memory required to store the local tensor of $O(I/P)$ words and factor matrices of cumulative size $O((R/P)\sum I_n)$, each processor must be able to store a redundant subset of the rows of the factor matrices it needs to perform MTTKRP computations.
-This corresponds to storing $P/P_n$ redundant copies of every factor matrix, which results in a local memory requirement of $O(R \sum I_n/P_n)$ for a general processor grid.
-The processor grid that minimizes communication also minimizes local memory, and the extra memory requirement can be as low as $O(RI^{1/N}/P^{1/N})$.
-
-The dimension tree algorithm also requires extra temporary memory space, but the space required tends to be much smaller than what is required to store the local tensor.
-If the tensor dimensions can be partitioned into two parts with approximately equal geometric means, the extra memory requirement for running a dimension tree is as small as $O(R\sqrt{I}/P)$, which is typically dominated by $O(I/P)$.
-
-\input{fig/Alg_header.tex}
-\begin{figure*}[t]
-\centering
-  \subfloat[Start $n$th iteration with redundant subset of rows of each input matrix. \label{fig:inner_a}]{\input{fig/AlgStart.tex}}  \quad
-  \subfloat[Compute local MTTKRP for contribution to output matrix $\Mn{M}{2}$. \label{fig:inner_b}]{\input{fig/AlgMTTKRP.tex}}  \quad
-  \subfloat[Reduce-Scatter to compute and distribute rows of $\Mn{M}{2}$. \label{fig:inner_c}]{\input{fig/AlgRS.tex}} \quad
-  \subfloat[Compute local NLS update to obtain $\Mn{H}{2}_{\V{p}}$ from $\Mn{M}{2}_{\V{p}}$ (and $\Mn{S}{2}$). \label{fig:inner_d}]{\input{fig/AlgNLS.tex}} \quad
-  \subfloat[All-Gather to collect rows of $\Mn{H}{2}$ needed for later inner iterations. \label{fig:inner_e}]{\input{fig/AlgAG.tex}}
-  \caption{Illustration of 2nd inner iteration of Par-NNCP algorithm for 3-way tensor on a $3\times3\times3$ processor grid, showing data distribution, communication, and computation across steps.
-Highlighted areas correspond to processor $(1,3,1)$ and its processor slice with which it communicates.
-The column normalization and computation of $\Mn{G}{2}$, which involve communication across all processors, is not shown here.}
-  \label{fig:inner} 
-\end{figure*}
-
 \begin{algorithm}
@@
 \end{algorithm}
+
+\subsubsection{Algorithm Overview}
+
+The basic sequential algorithm is given in \Cref{alg:nncp}, and the parallel version is given in \Cref{alg:Par-NNCP-short}.
+We will refer to both the inner iteration, in which one factor matrix is updated (\cref{line:for} to \cref{line:endfor}), and the outer iteration, in which all factor matrices are updated (\cref{line:while} to \cref{line:endwhile}).
+In the parallel algorithm, the processors are organized into a logical multidimensional grid (tensor) with as many modes as the data tensor.
+The communication patterns used in the algorithm are all MPI collectives, including All-Reduce, Reduce-Scatter, and All-Gather.
+The processor communicators (across which the collectives are performed) include the set of all processors and the sets of processors within the same processor slice.
+Processors within a mode-$n$ slice all have the same $n$th coordinate.
+
+The method of enforcing the nonnegativity constraints of the linear least squares solve (or update) generally affects only local computation because each row of a factor matrix can be updated independently.
+In our algorithm, each processor solves the linear problem or computes the update for its subset of rows (see \cref{line:locNLS}). 
+The most expensive (and most complicated) part of the parallel algorithm is the computation of the MTTKRP, which corresponds to \cref{line:locMTTKRP,line:reduce-scatter,line:all-gather}.
+
+The details that are omitted from this presentation of the algorithm include the normalization of each factor matrix after it is computed and the computation of the residual error at the end of an outer iteration.
+The computations do involve both local computation and communication, but their costs are negligible.
+A more detailed pseudocode is given in \Cref{alg:Par-NNCP-long}.
+
+\input{fig/Alg_header.tex}
+\begin{figure*}[t]
+\centering
+  \subfloat[Start $n$th iteration with redundant subset of rows of each input matrix. \label{fig:inner_a}]{\input{fig/AlgStart.tex}}  \quad
+  \subfloat[Compute local MTTKRP for contribution to output matrix $\Mn{M}{2}$. \label{fig:inner_b}]{\input{fig/AlgMTTKRP.tex}}  \quad
+  \subfloat[Reduce-Scatter to compute and distribute rows of $\Mn{M}{2}$. \label{fig:inner_c}]{\input{fig/AlgRS.tex}} \quad
+  \subfloat[Compute local NLS update to obtain $\Mn{H}{2}_{\V{p}}$ from $\Mn{M}{2}_{\V{p}}$ (and $\Mn{S}{2}$). \label{fig:inner_d}]{\input{fig/AlgNLS.tex}} \quad
+  \subfloat[All-Gather to collect rows of $\Mn{H}{2}$ needed for later inner iterations. \label{fig:inner_e}]{\input{fig/AlgAG.tex}}
+  \caption{Illustration of 2nd inner iteration of Par-NNCP algorithm for 3-way tensor on a $3\times3\times3$ processor grid, showing data distribution, communication, and computation across steps.
+Highlighted areas correspond to processor $(1,3,1)$ and its processor slice with which it communicates.
+The column normalization and computation of $\Mn{G}{2}$, which involve communication across all processors, is not shown here.}
+  \label{fig:inner} 
+\end{figure*}
+
+\subsubsection{Data Distribution}
+\label{sec:datadist}
+
+Given a logical processor grid of processors $P_1\times \cdots \times P_N$, we distribute the tensor $\TA$ in a block or Cartesian partition.
+Each processor owns a local tensor of dimensions $(I_1/P_1)\times \cdots \times (I_N/P_N)$, and only one copy of the tensor is stored.
+Locally, the tensor is stored linearly, with entries ordered in a natural mode-descending way that generalizes column-major layout of matrices.
+Given a processor $\V{p}=(\VE{p}{1},\dots,\VE{p}{N})$, we denote its local tensor by $\TA_{\V{p}}$.
+
+Each factor matrix is distributed across processors in a block row partition, so that each processor owns a subset of the rows.
+We use the notation $\Mn{H}{n}_{\V{p}}$, which has dimensions $I_n/P\times R$ to denote the local part of the $n$th factor matrix stored on processor $\V{p}$.
+However, we also make use a redundant distribution of the factor matrices across processors, because all processors in a mode-$n$ processor slice need access to the same entries of $\Mn{H}{n}$ to perform their computations.
+The notation $\Mn{H}{n}_{\VE{p}{n}}$ denotes the $I_n/P_n\times R$ submatrix of $\Mn{H}{n}$ that is redundantly stored on all processors whose $n$th coordinate is $\VE{p}{n}$ (there are $P/P_n$ such processors).
+
+Other matrices involved in the algorithm include $\Mn{M}{n}_{\V{p}}$, which is the result of the MTTKRP computation and has the same distribution scheme as $\Mn{H}{n}_{\V{p}}$, and $\Mn{G}{n}$, which is the $R\times R$ Gram matrix of the factor matrix $\Mn{H}{n}$ and is stored redundantly on all processors.
+
+\subsubsection{Inner Iteration}
+
+The inner iteration is displayed graphically in \Cref{fig:inner} for a 3-way example and an update of the $2$nd factor matrix.
+The main idea is that at the start of the $n$th inner iteration (\cref{fig:inner_a}), all of the data is in place for each processor to perform a local MTTKRP computation.
+This means that all processors in a slice redundantly own the same rows of the corresponding factor matrix (for all modes except $n$).
+After the local MTTKRP is computed (\cref{fig:inner_b}), each processor has computed a contribution to a subset of the rows of the global MTTKRP $\Mn{M}{n}$, but its contribution must be summed up with the contributions of all other processors in its mode-$n$ slice.
+This summation is performed with a Reduce-Scatter collective across the mode-$n$ processor slice that achieves a row-wise partition of the result (in \cref{fig:inner_c}, the light gray shading corresponds to the rows of $\Mn{M}{2}$ to which processor $(1,3,1)$ contributes and the dark gray shading corresponds to the rows it receives as output).
+The output distribution of the Reduce-Scatter is designed so that afterwards, the update of the factor matrix in that mode can be performed row-wise in parallel.
+Along with $\Mn{S}{n}$, which can be computed locally, each processor updates its own rows of the factor matrix given its rows of the MTTKRP result (\cref{fig:inner_d}).
+The remainder of the inner iteration is preparing and distributing the new factor matrix data for future inner iterations, which includes an All-Gather of the newly computed factor matrix $\Mn{H}{n}$ across mode-$n$ processor slices (\cref{fig:inner_e}) and recomputing $\Mn{G}{n}={\Mn{H}{n}}^\Tra\Mn{H}{n}$.
+
+\subsubsection{Analysis}
+
+\begin{table*}
+\centering
+\begin{tabular}{|c|cccc|}
+\hline
+ & \textbf{Computation} & \textbf{Memory - Local MTTKRP} & \textbf{Communication} & \textbf{Memory - Par.~Algorithm}  \\
+\hline
+Par-NNCP w/ opt.~proc.~grid & $\frac{IR}{P}$ & $\frac{RI^{1/2}}{P^{1/2}}$& $\frac{NRI^{1/N}}{P^{1/N}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$  \\
+Par-NNCP w/ gen.~proc.~grid & & & $R\sum_n \frac{I_n}{P_n}$ & $R\sum_n \frac{I_n}{P_n}$  \\
+NoDimTree & $\frac{NIR}{P}$ & $\displaystyle \max_n \frac{RI/I_n}{P/P_n}$  &  & \\
+\hline
+\end{tabular}
+\smallskip
+\caption{Leading-order per-iteration costs in terms of computation (flops), communication (words moved), and memory (words).
+We ignore constants but omit big-Oh notation for clarity.
+The first line corresponds to \cref{alg:Par-NNCP-short} with an optimal choice of processor grid and applying the dimension tree optimization locally.
+The second line corresponds to the changes in communication and memory for a general processor grid.
+The third line corresponds to the changes in computation and memory if the dimension tree optimization is not applied.}
+\label{tab:costs}
+\end{table*}
+
+We will analyze the cost of a single outer iteration.
+While the number of outer iterations is sensitive to the NLS method used, the outer iteration time is generally the same across NLS methods.
+We summarize the analysis in \cref{tab:costs}, showing the differences with and without an optimal processor grid and with and without using a dimension tree.
+
+\paragraph{\emph{Computation}}
+The local computation occurs at \cref{line:locMTTKRP,line:hadamard,line:locNLS,line:locSYRK}.
+The cost of \cref{line:hadamard} is $O(NR^2)$, the cost of \cref{line:locNLS} is $O(R^3I_n/P)$, which is a loose upper bound for BPP and other methods \cite{KBP16}, and the cost of \cref{line:locSYRK} is $O(R(I_n/P)^2)$.
+The sum of these three costs across all inner iterations is $O(R^2N^2+(R^3/P)\sum I_n+(R/P^2)\sum I_n^2)$, which is dominated by the cost of the MTTKRP.
+When using dimension trees to perform the MTTKRP (\cref{line:locMTTKRP}), we compute the cost amortized over all inner iterations.
+In this case, the cost is dominated by the two partial MTTKRP computations (from the root of the tree), which together are $O((R/P) \prod I_n)=O(IR/P)$ and dominate the costs of the multi-TTVs.
+We note that this cost involves the product of all the tensor dimensions, which is why it dominates, and we note that it scales linearly with $P$.
+
+\paragraph{\emph{Communication}}
+The communication within the inner iteration occurs at \cref{line:reduce-scatter,line:all-reduce,line:all-gather}.
+\Cref{line:all-reduce} involves $O(R^2)$ data and a collective across all processors.
+\Cref{line:reduce-scatter,line:all-gather} involve $O(I_nR/P_n)$ data across a subset of $P/P_n$ processors.
+Thus, the All-Reduce dominates the latency cost and the Reduce-Scatter/All-Gather dominate the bandwidth cost, for a total outer iteration communication cost of $O(R\sum I_n/P_n)$ words and $O(N\log P)$ messages.
+If the optimal processor grid can be chosen to minimize communication (assuming $P$ is sufficiently factorable), then the bandwidth cost can achieve a value of $O(NRI^{1/N}/P^{1/N})$ by making the local tensors as cubical as possible.
+We note that this cost scales with $P^{1/N}$, which is far from linear scaling.
+However, it is proportional to the geometric mean of the tensor dimensions (on the order of one tensor dimension), which is much less than the computation cost dependence on the product of all dimensions.
+
+\paragraph{\emph{Memory}}
+The algorithm requires extra local memory to run.
+Aside from the memory required to store the local tensor of $O(I/P)$ words and factor matrices of cumulative size $O((R/P)\sum I_n)$, each processor must be able to store a redundant subset of the rows of the factor matrices it needs to perform MTTKRP computations.
+This corresponds to storing $P/P_n$ redundant copies of every factor matrix, which results in a local memory requirement of $O(R \sum I_n/P_n)$ for a general processor grid.
+The processor grid that minimizes communication also minimizes local memory, and the extra memory requirement can be as low as $O(NRI^{1/N}/P^{1/N})$, which is typically dominated by $O(I/P)$.
+
+The dimension tree algorithm also requires extra temporary memory space, but the space required tends to be much smaller than what is required to store the local tensor.
+If the tensor dimensions can be partitioned into two parts with approximately equal geometric means, the extra memory requirement for running a dimension tree is as small as $O(R\sqrt{I/P})$, which is also typically dominated by $O(I/P)$.

### 2018-05-25 latex/sc18/algorithm.tex
added more details on dimension tree

@@
 An important optimization of the CP-ALS algorithm is to re-use temporary values across inner iterations \cite{PTC13a,KU16-TR,LCPSV17,Kaya17}.
-To illustrate the idea, consider a 3-way tensor $\T{X}$ approximated by $\dsquare{\M{U},\M{V},\M{W}}$ and the two MTTKRP computations $\underline{\Mz{X}{1}(\M{W}}\Khat\M{V})$ and $\underline{\Mz{X}{2}(\M{W}}\Khat\M{U})$ used to update factor matrices $\M{U}$ and $\M{V}$, respectively.
+To illustrate the idea, consider a 3-way tensor $\T{X}$ approximated by $\dsquare{\M{U},\M{V},\M{W}}$ and the two MTTKRP computations $\Mn{M}{1}=\underline{\Mz{X}{1}(\M{W}}\Khat\M{V})$ and $\Mn{M}{2}=\underline{\Mz{X}{2}(\M{W}}\Khat\M{U})$ used to update factor matrices $\M{U}$ and $\M{V}$, respectively.
 The underlined parts of the expressions correspond to the shared dependence of the outputs on the tensor $\T{X}$ and the third factor matrix $\M{W}$.
@@
 We refer to the computation that combines the temporary quantity with the other factor matrix to complete the MTTKRP computation as a multi-tensor-times-vector or \emph{multi-TTV}, as it consists of multiple operations that multiply a tensor times a set of vectors, each corresponding to a different mode.
-This terminology was used in \cite{HBJT17-TR}.
+This terminology was used in \cite{HBJT18}.
+
+To understand the steps of the partial MTTKRP and multi-TTV operations in more detail, we consider $\T{X}$ to be $I\times J\times K$ and $\M{U}$, $\M{V}$, and $\M{W}$ to have $R$ columns.
+Then 
+\begin{equation*}
+\MnE{M}{1}{ir} = \sum_{i,j} \TE{X}{ijk} \ME{V}{jr} \ME{W}{kr} 
+= \sum_{j} \ME{V}{jr} \sum_k \TE{X}{ijk} \ME{W}{kr} 
+= \sum_{j} \ME{V}{jr} \TE{T}{ijr}
+\end{equation*}
+where $\T{T}$ is an $I\times J\times R$ tensor that is the result of a partial MTTKRP between tensor $\T{X}$ and the single factor matrix $W$.
+Likewise,
+\begin{equation*}
+\MnE{M}{2}{jr} = \sum_{i,k} \TE{X}{ijk} \ME{U}{ir} \ME{W}{kr} 
+= \sum_{i} \ME{U}{ir} \sum_k \TE{X}{ijk} \ME{W}{kr} 
+= \sum_{i} \ME{U}{ir} \TE{T}{ijr}
+\end{equation*}
+and we see that the temporary tensor $\T{T}$ can be re-used.
+From these expressions, we can also see that computing $\T{T}$ (a partial MTTKRP) corresponds to a matrix-matrix multiplication, and computing each of $\Mn{M}{1}$ and $\Mn{M}{2}$ from $\T{T}$ (a multi-TTV) corresponds to $R$ independent matrix-vector multiplications.
+In this case, we compute $\Mn{M}{3}$ using a full MTTKRP.
 
@@
 The general approach can yield significant benefit, decreasing the computation by a factor of approximately $N/2$ for dense $N$-way tensors.
-The idea is introduced in \cite{PTC13a}, but we adopt the terminology and notation of \emph{dimension trees} used in \cite{KU16-TR,Kaya17}.
+The idea is introduced in \cite{PTC13a}, but we adopt the terminology and notation of \emph{dimension trees} used for sparse tensors in \cite{KU16-TR,Kaya17}.
 In this notation, the root node is labeled $\{1,\dots,N\}$ and corresponds to the original tensor, a leaf is labeled $\{n\}$ and corresponds to the $n$th MTTKRP result, and an internal node is labeled by a set of modes $\{i,\dots,j\}$ and corresponds to a temporary tensor whose values contribute to the MTTKRP results of modes $i,\dots,j$.
@@
 For dense tensors, the computational cost is dominated by the root's branches, which correspond to partial MTTKRP computations.
+We perform the splitting of modes at the root so that modes are chosen contiguously with the respect to the layout of the tensor entries in memory.
+In this way, each partial MTTKRP can be performed via BLAS's GEMM interface without reordering tensor entries in memory.
 All other edges in a tree correspond to multi-TTVs and are typically much cheaper.
+By organizing the memory layout of temporary quantities, the multi-TTV operations can be performed via a sequence of calls using BLAS's GEMV interface.
+By using the BLAS in our implementation, we are able to obtain high performance and on-node parallelism.
+
+No matter how the dimension tree is designed, the computational cost of each partial MTTKRP is $O(IR)$, where $I$ is the number of tensor entries and $R$ is the rank of the CP decomposition.
+This is the same operation count as a full MTTKRP.
+The computational cost of a multi-TTV is the number of entries in the temporary tensor, which is the product of a \emph{subset} of the original tensor dimensions multiplied by $R$.
+Thus, it is computationally cheaper than the partial MTTKRPs, but it is also typically bandwidth bound.
+
+The other subroutine necessary for implementing the dimension tree approach is the Khatri-Rao product of sets of factor matrices.
+We implement the operation as a row-wise Hadamard product of a set of factor matrix rows, and we use OpenMP parallelization to obtain on-node parallelism.
+The computational cost of this operation is also typically lower order, but the running time in practice suffers from being bandwidth bound.
 

### 2018-05-25 latex/sc18/conclusion.tex
added more details on dimension tree

@@
 This observation is supported by the theoretical analysis: although the communication does not scale as well with $P$, the total amount of data depends on a sum of tensor dimensions rather the product of the tensor dimensions, which determines the total amount of computation.
+For a relative comparison, consider memory-efficient parallel dense $n\times n$ matrix multiplication: the ratio of the $O(n^3/P)$ local computation to the $O(n^2/P^{1/2})$ communication is approximately the square root of the size of the local data, or $O((n^2/P)^{1/2})$.
+In the case of NNCP, the ratio of local computation to communication is $O((I/P)^{1-1/N}/N)$, which is approximately the size of the local data raised to the power $1-1/N$.
+This exponent is larger than $1/2$ and grows with $N$, and therefore it predicts the NNCP computation should be more computation bound than matrix multiplication.
+Note that this analysis does not consider the type of local computation; for small $R$, the local computation will likely be memory bandwidth bound, but the algorithm will spend more time on local computation than on interprocessor communication.
 

### 2018-05-25 latex/sc18/experiments.tex
added analysis of comm sweep experiment

@@
 
+\subsection{Varying Processor Grid}
+
+In order to illustrate the effect of processor grid choice on running time, we show in \cref{fig:commsweep10} a time breakdown over various processor grid choices for a 4D problem on 81 processors.
+Because the tensor is cubical and 81 has a restricted factorization into 4 numbers, there are 5 distinct processor grids.
+The overall takeaway is that the processor grid has very little effect on running time; in this experiment there is less than 10\% variation in overall time.
+While the optimal processor grid reduced the communication time by approximately $3\times$ compared to the other processor grids, the running time is dominated by local computation, so it had little effect on overall time.
+Furthermore, adjusting the processor grid affects the local tensor dimensions and the performance of the local computations, and the optimal processor grid led to slower local performance.
+For $R=10$, all of the local computation is memory bandwidth bound, and we believe the variations in running times to be effects of some temporary quantities fitting into smaller levels of cache.
+
+\begin{figure}
+\renewcommand{\datafile}{data/comm_sweep_scale_pgf_10.dat}
+\renewcommand{\numiterations}{10}
+\begin{tikzpicture}
+\begin{axis}[	
+	ybar stacked,
+	bar width=8pt,
+	width=\columnwidth,
+	height =.75\columnwidth,
+	%width=9cm, height=3.85cm,
+	ylabel={Time (s)}, 
+	xlabel={Processor Grid},
+	y label style={yshift=-.5cm},
+	ymin=0,
+	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
+	symbolic x coords={81x1x1x1-10, 27x3x1x1-10,9x9x1x1-10,9x3x3x1-10,3x3x3x3-10,},
+	%xtick=data,
+	xticklabels={,81x1x1x1,27x3x1x1,9x9x1x1,9x3x3x1,3x3x3x3},
+	legend style={at={(0.5,1.3)},anchor=north},
+	legend columns=-1,
+]
+	\setcolors
+	\addplot table[x=algo-k, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=algo-k, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=algo-k, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=algo-k, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=algo-k, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
+	\legend{PM,mTTV,NLS,Factor Comm,Gram Comm};
+\end{axis}
+%\labels
+\end{tikzpicture}
+\caption{Time breakdown for $243\times243\times243\times243$ tensor and rank $R=10$ on 81 processors for varying processor grids.}
+\label{fig:commsweep10}
+\end{figure}
+
 \subsection{Varying Approximation Rank}
@@
 \label{fig:ksweepneuro}
-\end{figure}
-
-\begin{figure}
-\renewcommand{\datafile}{data/comm_sweep_scale_pgf_10.dat}
-\renewcommand{\numiterations}{10}
-\begin{tikzpicture}
-\begin{axis}[	
-	ybar stacked,
-	bar width=8pt,
-	width=\columnwidth,
-	height =.75\columnwidth,
-	%width=9cm, height=3.85cm,
-	ylabel={Time (s)}, 
-	xlabel={Processor Grid},
-	y label style={yshift=-.5cm},
-	ymin=0,
-	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
-	symbolic x coords={81x1x1x1-10, 27x3x1x1-10,9x9x1x1-10,9x3x3x1-10,3x3x3x3-10,},
-	%xtick=data,
-	xticklabels={,81x1x1x1,27x3x1x1,9x9x1x1,9x3x3x1,3x3x3x3},
-	legend style={at={(0.5,1.3)},anchor=north},
-	legend columns=-1,
-]
-	\setcolors
-	\addplot table[x=algo-k, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo-k, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo-k, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo-k, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo-k, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
-	\legend{PM,mTTV,NLS,Factor Comm,Gram Comm};
-\end{axis}
-%\labels
-\end{tikzpicture}
-\caption{Time breakdown for $243\times243\times243\times243$ tensor and rank $R=10$ on 81 processors for varying processor grids.}
-\label{fig:commsweep10}
 \end{figure}

### 2018-05-29 latex/sc18/algorithm.tex
added figure and more details regarding dimtrees

@@
 We refer to the computation that combines the temporary quantity with the other factor matrix to complete the MTTKRP computation as a multi-tensor-times-vector or \emph{multi-TTV}, as it consists of multiple operations that multiply a tensor times a set of vectors, each corresponding to a different mode.
-This terminology was used in \cite{HBJT18}.
+%This terminology was used in \cite{HBJT18}.
 
@@
 = \sum_{j} \ME{V}{jr} \sum_k \TE{X}{ijk} \ME{W}{kr} 
-= \sum_{j} \ME{V}{jr} \TE{T}{ijr}
+= \sum_{j} \ME{V}{jr} \TE{T}{ijr},
 \end{equation*}
@@
 = \sum_{i} \ME{U}{ir} \sum_k \TE{X}{ijk} \ME{W}{kr} 
-= \sum_{i} \ME{U}{ir} \TE{T}{ijr}
+= \sum_{i} \ME{U}{ir} \TE{T}{ijr},
 \end{equation*}
@@
 
+\begin{figure}
+\input{fig/dim_tree.tex}
+\caption{Dimension tree example for $N=5$. 
+The data associated with the root node is the original tensor, the data associated with the leaf nodes are MTTKRP results, and the data associated with internal nodes are temporary tensors.  
+Edges labeled with PM correspond to partial MTTKRP computations, and edges labeled with mTTV correspond to multi-TTV computations.}
+\label{fig:DT}
+\end{figure}
+
 \Cref{fig:DT} illustrates a dimension tree for the case $N=5$.
@@
 
+\begin{figure}
+\subfloat[Partial MTTKRP to compute node $\{3,4,5\}$ from root node $\{1,2,3,4,5\}$, executed via one GEMM call. \label{fig:DT-PM}]{\includegraphics[width=\columnwidth]{fig/PM}} \\
+\subfloat[Multi-TTV to compute node $\{3\}$ from node $\{3,4,5\}$, executed via $R$ GEMV calls.
+Here $\Mz{T}{1} \lbrack r\rbrack$ refers to the $r$th contiguous block of $\Mz{T}{1}$.
+\label{fig:DT-mTTV}]{\includegraphics[width=\columnwidth]{fig/mTTV}}
+\caption{Data layout and dimensions for two example computations in dimension tree shown in \Cref{fig:DT}.
+In this notation, $\Mz{X}{3:5}$ is the matricization of input tensor $\T{X}$ with respect to modes 3 through 5, $\M{K}_{1:2} = \Mn{H}{2} \Khat \Mn{H}{1}$, $\T T$ is the temporary $I_3 \times I_4 \times I_5 \times R$ tensor corresponding to node $\{3,4,5\}$ in the dimension tree, $\M{K}_{4:5} = \Mn{H}{5} \Khat \Mn{H}{4}$, and $\Mn{M}{3}$ is the MTTKRP result for mode 3.}
+\label{fig:DTcartoon}
+\end{figure}
+
+\Cref{fig:DTcartoon} shows the data layout and dimensions of a partial MTTKRP and a multi-TTV taken from the example dimension tree in \Cref{fig:DT}.
+\Cref{fig:DT-PM} shows a partial MTTKRP between the input tensor $\T{X}$ and the Khatri-Rao product of the factor matrices in modes 1 and 2, which produces a temporary tensor $\T{T}$ corresponding to the $\{3,4,5\}$ node in the dimension tree.
+The key to efficiency in this computation is that the matricization of $\T{X}$ that assigns modes 1 through 2 to rows and modes 3 through 5 to columns is already column-major in memory.
+Thus, we can use the GEMM interface and compute the temporary tensor $\T{T}$ without reordering any tensor entries.
+\Cref{fig:DT-mTTV} depicts a multi-TTV that computes the results $\Mn{M}{3}$ from $\T{T}$ and the factor matrices in modes 4 and 5.
+Here, the tensor $\T{T}$ is matricized with respect to only its first mode (of dimension $I_3$), but this matricization is also column-major in memory.
+We choose the ordering of the modes of $\T{T}$ such that each of $R$ contiguous blocks is used to compute one column of the output matrix via a matrix-vector operation with a corresponding column of the Khati-Rao product of the other factor matrices.
+
 No matter how the dimension tree is designed, the computational cost of each partial MTTKRP is $O(IR)$, where $I$ is the number of tensor entries and $R$ is the rank of the CP decomposition.
@@
 The computational cost of a multi-TTV is the number of entries in the temporary tensor, which is the product of a \emph{subset} of the original tensor dimensions multiplied by $R$.
-Thus, it is computationally cheaper than the partial MTTKRPs, but it is also typically bandwidth bound.
+Thus, it is computationally cheaper than the partial MTTKRPs, but it is also memory bandwidth bound.
 
@@
 We implement the operation as a row-wise Hadamard product of a set of factor matrix rows, and we use OpenMP parallelization to obtain on-node parallelism.
-The computational cost of this operation is also typically lower order, but the running time in practice suffers from being bandwidth bound.
-
-\begin{figure}
-\input{fig/dim_tree.tex}
-\caption{Dimension tree example for $N=5$. 
-The data associated with the root node is the original tensor, the data associated with the leaf nodes are MTTKRP results, and the data associated with internal nodes are temporary tensors.  
-Edges labeled with PM correspond to partial MTTKRP computations, and edges labeled with mTTV correspond to multi-TTV computations.}
-\label{fig:DT}
-\end{figure}
+The computational cost of this operation is also typically lower order, but the running time in practice suffers from also being memory bandwidth bound.
 
@@
 The extra computation involved in computing the relative error is negligible.
-These identities have been used in previous work \cite{KB09,TensorBox,SK16,LKLHS2017}.
+These identities have been used previously \cite{KB09,TensorBox,SK16,LKLHS2017}.
 
@@
 Locally, the tensor is stored linearly, with entries ordered in a natural mode-descending way that generalizes column-major layout of matrices.
-Given a processor $\V{p}=(\VE{p}{1},\dots,\VE{p}{N})$, we denote its local tensor by $\TA_{\V{p}}$.
+Given a processor $\V{p}=(\VE{p}{1},\dots,\VE{p}{N})$, we denote its local tensor $\TA_{\V{p}}$.
 

### 2018-05-29 latex/sc18/conclusion.tex
added analysis of comm sweep experiment
added figure and more details regarding dimtrees

@@
 Besides the reduction in flops, the dimension tree approach enjoys better DGEMM performance and avoids memory-bound KRP computations.
+Furthermore, we see that tuning the processor grid had much less effect on overall performance.
+Not only do reductions in communication not matter as much as computation, but different local tensor sizes can also cause variations in local performance that outweigh the savings in communication.
 
-Given an efficient parallel algorithm for NNCP, it is now possible to compare the convergence properties of various NLS algorithms on large-scale tensor data that is too large to fit on single servers or would require too much computational time to analyze interactively.
-We hope to explore this in future work.
+%Given an efficient parallel algorithm for NNCP, it is now possible to compare the convergence properties of various NLS algorithms on large-scale tensor data that is too large to fit on single servers or would require too much computational time to analyze interactively.
+%We hope to explore this in future work.

### 2018-05-29 latex/sc18/algorithm.tex
minor changes

@@
 Par-NNCP w/ gen.~proc.~grid & & & $R\sum_n \frac{I_n}{P_n}$ & $R\sum_n \frac{I_n}{P_n}$  \\
-NoDimTree & $\frac{NIR}{P}$ & $\displaystyle \max_n \frac{RI/I_n}{P/P_n}$  &  & \\
+Par-NNCP w/o dim.~tree opt. & $\frac{NIR}{P}$ & $\displaystyle \max_n \frac{RI/I_n}{P/P_n}$  &  & \\
 \hline

### 2018-05-29 latex/sc18/introduction.tex
added figure and more details regarding dimtrees
minor changes

@@
 
+%The CP decomposition is a low-rank approximation of a multi-dimensional array, or tensor, which generalizes matrix approximations like the truncated singular value decomposition. 
+%As in Figure \ref{fig:cpdecomposition}, CP decomposition approximates the given input matrix as sum of $k$ rank-1 tensors.  
 The CP decomposition is a low-rank approximation of a multi-dimensional array, or tensor, which generalizes matrix approximations like the truncated singular value decomposition.
-As in Figure \ref{fig:cpdecomposition}, CP decomposition approximates the given input matrix as sum of $k$ rank-1 tensors.
+It approximates the input tensor by a sum of rank-one tensors, which are outer-products of vectors. 
 CP is often used for finding hidden patterns, or latent factors, within tensor data, particularly when the goal is to interpret the factors, and it is popular within the signal processing, machine learning, and scientific computing communities.
 
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=3in, height=1.5in]{fig/cpdecomposition.pdf}
-\caption{CP Decomposition}
-\label{fig:cpdecomposition}
-\end{center}
-\end{figure}
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics[width=3in, height=1.5in]{fig/cpdecomposition.pdf}
+%\caption{CP Decomposition}
+%\label{fig:cpdecomposition}
+%\end{center}
+%\end{figure}
 
@@
 
-As we detail in the related work, the general techniques for reducing computation and communication are not new.
-The recomputation avoidance was proposed in a sequential algorithm \cite{PTC13a}, and the parallelization scheme was proposed and analyzed for general tensors \cite{BKR17-TR} and the algorithm was implemented for 3D tensors \cite{LK+17b}.
-The main contribution of this paper is the combination and implementation of these ideas for general $N$D tensors, and we demonstrate superior performance to the existing implementation for 3D tensors \cite{LK+17b}.
+As we detail in the related work, the general techniques for reducing computation and communication have been used in similar contexts.
+The recomputation avoidance was proposed in a sequential algorithm \cite{PTC13a}, the parallelization scheme was proposed and analyzed for general tensors \cite{BKR17-TR}, and the algorithm was implemented for 3D tensors \cite{LK+17b}.
+%The main contribution of this paper is the combination and implementation of these ideas for general $N$-D tensors, and we demonstrate superior performance to the existing implementation for 3D tensors \cite{LK+17b}.
 
@@
 \begin{itemize}
-		\item we present a computation- and communication-efficient algorithm for NNCP with detailed cost analysis,
-		\item our algorithm is designed for general $N$-way tensors,
+		\item we present the first distributed-memory parallel implementation of NNCP algorithms for arbitrary-dimension dense tensors,
+		\item we optimize the use of dimension trees for dense tensors, avoiding recomputation across multiple MTTKRPs,
+		\item our parallel algorithm is communication optimal with a carefully chosen processor grid,
 		\item we demonstrate a performance improvement of up to $2.2\times$ over the existing state-of-the-art parallel software on 3D tensors,
-		\item we demonstrate efficient parallel scaling of up to $253\times$ on $512$ cores.
+		\item our implementation obtains efficient parallel scaling of up to $253\times$ on $512$ cores.
 \end{itemize}

### 2018-05-29 latex/sc18/appendix.tex
3D weak scaling in appendix

@@
 \end{algorithm}
+
+\begin{figure}
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/wk_3D_syn2_scale_pgf.dat}
+\renewcommand{\numiterations}{10}
+\weakthreeDtrue
+\breakdownplot
+%\labels
+\end{tikzpicture}
+\caption{Weak scaling of 3D synthetic tensors with (D) and without (N) the use of dimension trees.
+The tensor and processor grid dimensions are $128k\times 128k\times 128k$ and $k\times k\times k$ for $k\in\{1,2,3,4\}$, and the rank is fixed at 32.
+The reported times are per iteration.}
+\label{fig:weaksynthetic3D}
+\end{figure}

### 2018-05-29 latex/sc18/algorithm.tex
changes during meeting

@@
 
+\textcolor{red}{fill in blanks with duplicate values}
+
 \begin{table*}
@@
 Par-NNCP w/ opt.~proc.~grid & $\frac{IR}{P}$ & $\frac{RI^{1/2}}{P^{1/2}}$& $\frac{NRI^{1/N}}{P^{1/N}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$  \\
+%\hline
 Par-NNCP w/ gen.~proc.~grid & & & $R\sum_n \frac{I_n}{P_n}$ & $R\sum_n \frac{I_n}{P_n}$  \\
+%\hline
 Par-NNCP w/o dim.~tree opt. & $\frac{NIR}{P}$ & $\displaystyle \max_n \frac{RI/I_n}{P/P_n}$  &  & \\

### 2018-05-29 latex/sc18/experiments.tex
3D weak scaling in appendix
changes during meeting

@@
 
-Finally, we ran on 512 nodes, utilizing the 16 threads per node via parallel BLAS and OMP, for a $1024\times1024\times1024$ synthetic tensor and processor grid $8\times8\times8$.
-Here we observe a $253\times$ speedup which is the result of each node having less than 20 megabytes of data.
+\textcolor{red}{Finally, we ran on 512 nodes, utilizing the 16 threads per node via parallel BLAS and OMP, for a $1024\times1024\times1024$ synthetic tensor and processor grid $8\times8\times8$.
+Here we observe a $253\times$ speedup which is the result of each node having less than 20 megabytes of data.}
 

### 2018-05-29 latex/sc18/plots.tex

@@
 \newif\ifliavas
+% toggles weak 3D 
+\newif\ifweakthreeD
 
@@
 	ymin=0,
+	\ifweakthreeD
+		symbolic x coords={D1,N1,D8,N8,D27,N27},
+	\else
 		symbolic x coords={D1,N1,D16,N16,D81,N81,D256,N256},
+	\fi	
 	xtick=data,

### 2018-05-30 latex/sc18/algorithm.tex
full revised draft ready

--- 2018-05-29 latex/sc18/algorithm.tex
@@
 \hline
-Par-NNCP w/ opt.~proc.~grid & $\frac{IR}{P}$ & $\frac{RI^{1/2}}{P^{1/2}}$& $\frac{NRI^{1/N}}{P^{1/N}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$  \\
-Par-NNCP w/ gen.~proc.~grid & & & $R\sum_n \frac{I_n}{P_n}$ & $R\sum_n \frac{I_n}{P_n}$  \\
-NoDimTree & $\frac{NIR}{P}$ & $\displaystyle \max_n \frac{RI/I_n}{P/P_n}$  &  & \\
+Par-NNCP w/ opt.~proc.~grid & $\frac{IR}{P}$ & $\frac{RI^{1/2}}{P^{1/2}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$  \\
+%\hline
+Par-NNCP w/ gen.~proc.~grid & $\frac{IR}{P}$ & $\frac{RI^{1/2}}{P^{1/2}}$ & $\mathbf{R\sum_n \frac{I_n}{P_n}}$ & $\mathbf{R\sum_n \frac{I_n}{P_n}}$  \\
+%\hline
+Par-NNCP w/o dim.~tree opt. & $\mathbf{\frac{NIR}{P}}$ & $\displaystyle \mathbf{\max_n \frac{RI/I_n}{P/P_n}}$  &  $\frac{NRI^{1/N}}{P^{1/N}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$ \\
 \hline

### 2018-05-30 latex/sc18/prelims.tex
minor changes
full revised draft ready

@@
 In this case, every subproblem is a linear nonnegative least squares problem. Formally, the following problem 
-is solved iteratively for $n = 1 \cdots N$.
+is solved iteratively for $n = 1 \cdots N$:
 
 %\SplitN{\label{eqn:bcdnncp}}{
-$$\HH^{(n)} \leftarrow \argmin_{\HH \geq 0}  \| \Mn{S}{n} \HH^T - {\Mn{M}{n}}^T \|_F^2$$
+$$\HH^{(n)} \leftarrow \argmin_{\HH \geq 0}  \| \Mn{S}{n} \HH^T - {\Mn{M}{n}}^T \|_F^2.$$
 %}
+The BCD iteration is guaranteed to converge to a stationary point, but there is no guarantee that it converges to a global minimum.
+The number of outer iterations to convergence is problem dependent but typically ranges from the 10s to the 1000s.
 

### 2018-05-30 latex/sc18/experiments.tex

--- 2018-05-25 latex/sc18/experiments.tex
@@
 
-The performance on the cubical synthetic tensor is shown in \cref{fig:strongsynthetic3D}. 
+The performance on the cubical synthetic tensor is shown in \Cref{fig:strongsynthetic3D}. 
 We can observe from the figure that all the three algorithms scale nearly linearly as the problem remains compute bound. 
+Our algorithm (with the dimension tree optimization) achieves a speedup of $1771\times$ on 4096 cores over the same implementation running on 1 core.
 Recall from \label{sec:analysis} that the computation scales linearly with $1/P$ while the communication scales with $1/P^{1/N}=1/P^{1/3}$. 
@@
 
-Finally, we ran on 512 nodes, utilizing the 16 threads per node via parallel BLAS and OMP, for a $1024\times1024\times1024$ synthetic tensor and processor grid $8\times8\times8$.
-Here we observe a $253\times$ speedup which is the result of each node having less than 20 megabytes of data.
+%\textcolor{red}{Finally, we ran on 512 nodes, utilizing the 16 threads per node via parallel BLAS and OMP, for a $1024\times1024\times1024$ synthetic tensor and processor grid $8\times8\times8$.
+Here we observe a $253\times$ speedup which is the result of each node having less than 20 megabytes of data.}
 
@@
 
-In \cref{fig:strongsynthetic5D}, we benchmark performance for a 5D cubical tensor with each dimension set to 64.
+In \Cref{fig:strongsynthetic5D}, we benchmark performance for a 5D cubical tensor with each dimension set to 64.
 Because the tensor is 5D, we can no longer compare against NbAO-NTF.
@@
 %For eg., in the case of 3 x 3 x 3 x 3 processor grid with 81 nodes, we ran the proposed algorithm with and without dimension trees on a four mode tensor of size 384 x 384 x 384 x 384. 
-The results of the breakdown plot is shown in \cref{fig:weaksynthetic4D}. 
+The results of the breakdown plot is shown in \Cref{fig:weaksynthetic4D}. 
 In this case, the algorithm is compute bound with and without the use of the dimension tree, so the total time of the weak scaling remains fixed for both cases. 
@@
 \renewcommand{\numiterations}{10}
+\weakthreeDfalse
 \breakdownplot
@@
 
-In order to illustrate the effect of processor grid choice on running time, we show in \cref{fig:commsweep10} a time breakdown over various processor grid choices for a 4D problem on 81 processors.
+In order to illustrate the effect of processor grid choice on running time, we show in \Cref{fig:commsweep10} a time breakdown over various processor grid choices for a 4D problem on 81 processors.
 Because the tensor is cubical and 81 has a restricted factorization into 4 numbers, there are 5 distinct processor grids.
@@
 
-In \cref{fig:ksweepneuro} we compare performance for various ranks $R$ across all 3 algorithms, again used flat MPI. 
+In \Cref{fig:ksweepneuro} we compare performance for various ranks $R$ across all 3 algorithms, again used flat MPI. 
 Starting at $R=10$ we see the largest speed up of $2\times$ for our implementation with a dimension tree over NbAO-NTF. 
@@
 However, as the rank increases this speed up diminishes to $1.6\times$. 
-The loss of speed up is a result of the fact that, as we observed in \cref{fig:ksweep4DHSI}, the multi-TTV operations do not scale as well as the partial-MTTKRPs for increasing $R$.
+The loss of speed up is a result of the fact that, as we observed in \Cref{fig:ksweep4DHSI}, the multi-TTV operations do not scale as well as the partial-MTTKRPs for increasing $R$.
 Again, the performance of our implementation without the dimension tree optimization is comparable to NbAO-NTF.

### 2018-05-30 latex/sc18/abstract.tex
put changes in red

@@
 Our approach is also communication efficient, using a data distribution and parallel algorithm across a multidimensional processor grid that can be tuned to minimize communication.
-We benchmark our software on synthetic as well as hyperspectral image and neuroscience dynamic functional connectivity data, demonstrating that our algorithm scales well to 100s of nodes (up to 8192 cores) and is faster and more general than the currently available parallel software.
+We benchmark our software on synthetic as well as hyperspectral image and neuroscience dynamic functional connectivity data, demonstrating that our algorithm scales well to 100s of nodes (up to \red{4096} cores) and is faster and more general than the currently available parallel software.
 

### 2018-05-30 latex/sc18/algorithm.tex
put changes in red

@@
 
-To understand the steps of the partial MTTKRP and multi-TTV operations in more detail, we consider $\T{X}$ to be $I\times J\times K$ and $\M{U}$, $\M{V}$, and $\M{W}$ to have $R$ columns.
+\red{To understand the steps of the partial MTTKRP and multi-TTV operations in more detail, we consider $\T{X}$ to be $I\times J\times K$ and $\M{U}$, $\M{V}$, and $\M{W}$ to have $R$ columns.
 Then 
@@
 From these expressions, we can also see that computing $\T{T}$ (a partial MTTKRP) corresponds to a matrix-matrix multiplication, and computing each of $\Mn{M}{1}$ and $\Mn{M}{2}$ from $\T{T}$ (a multi-TTV) corresponds to $R$ independent matrix-vector multiplications.
-In this case, we compute $\Mn{M}{3}$ using a full MTTKRP.
+In this case, we compute $\Mn{M}{3}$ using a full MTTKRP.}
 
@@
 For dense tensors, the computational cost is dominated by the root's branches, which correspond to partial MTTKRP computations.
-We perform the splitting of modes at the root so that modes are chosen contiguously with the respect to the layout of the tensor entries in memory.
+\red{We perform the splitting of modes at the root so that modes are chosen contiguously with the respect to the layout of the tensor entries in memory.
 In this way, each partial MTTKRP can be performed via BLAS's GEMM interface without reordering tensor entries in memory.
@@
 By organizing the memory layout of temporary quantities, the multi-TTV operations can be performed via a sequence of calls using BLAS's GEMV interface.
-By using the BLAS in our implementation, we are able to obtain high performance and on-node parallelism.
+By using the BLAS in our implementation, we are able to obtain high performance and on-node parallelism.}
 
 \begin{figure}
-\subfloat[Partial MTTKRP to compute node $\{3,4,5\}$ from root node $\{1,2,3,4,5\}$, executed via one GEMM call. \label{fig:DT-PM}]{\includegraphics[width=\columnwidth]{fig/PM}} \\
-\subfloat[Multi-TTV to compute node $\{3\}$ from node $\{3,4,5\}$, executed via $R$ GEMV calls.
-Here $\Mz{T}{1} \lbrack r\rbrack$ refers to the $r$th contiguous block of $\Mz{T}{1}$.
-\label{fig:DT-mTTV}]{\includegraphics[width=\columnwidth]{fig/mTTV}}
-\caption{Data layout and dimensions for two example computations in dimension tree shown in \Cref{fig:DT}.
-In this notation, $\Mz{X}{3:5}$ is the matricization of input tensor $\T{X}$ with respect to modes 3 through 5, $\M{K}_{1:2} = \Mn{H}{2} \Khat \Mn{H}{1}$, $\T T$ is the temporary $I_3 \times I_4 \times I_5 \times R$ tensor corresponding to node $\{3,4,5\}$ in the dimension tree, $\M{K}_{4:5} = \Mn{H}{5} \Khat \Mn{H}{4}$, and $\Mn{M}{3}$ is the MTTKRP result for mode 3.}
+\subfloat[\red{Partial MTTKRP to compute node $\{3,4,5\}$ from root node $\{1,2,3,4,5\}$, executed via one GEMM call.} \label{fig:DT-PM}]{\includegraphics[width=\columnwidth]{fig/PM}} \\
+\subfloat[\red{Multi-TTV to compute node $\{3\}$ from node $\{3,4,5\}$, executed via $R$ GEMV calls.
+Here $\Mz{T}{1} \lbrack r\rbrack$ refers to the $r$th contiguous block of $\Mz{T}{1}$.} \label{fig:DT-mTTV}]{\includegraphics[width=\columnwidth]{fig/mTTV}}
+\caption{\red{Data layout and dimensions for two example computations in dimension tree shown in \Cref{fig:DT}.
+In this notation, $\Mz{X}{3:5}$ is the matricization of input tensor $\T{X}$ with respect to modes 3 through 5, $\M{K}_{1:2} = \Mn{H}{2} \Khat \Mn{H}{1}$, $\T T$ is the temporary $I_3 \times I_4 \times I_5 \times R$ tensor corresponding to node $\{3,4,5\}$ in the dimension tree, $\M{K}_{4:5} = \Mn{H}{5} \Khat \Mn{H}{4}$, and $\Mn{M}{3}$ is the MTTKRP result for mode 3.}}
 \label{fig:DTcartoon}
@@
 
-\Cref{fig:DTcartoon} shows the data layout and dimensions of a partial MTTKRP and a multi-TTV taken from the example dimension tree in \Cref{fig:DT}.
+\red{\Cref{fig:DTcartoon} shows the data layout and dimensions of a partial MTTKRP and a multi-TTV taken from the example dimension tree in \Cref{fig:DT}.
 \Cref{fig:DT-PM} shows a partial MTTKRP between the input tensor $\T{X}$ and the Khatri-Rao product of the factor matrices in modes 1 and 2, which produces a temporary tensor $\T{T}$ corresponding to the $\{3,4,5\}$ node in the dimension tree.
@@
 Here, the tensor $\T{T}$ is matricized with respect to only its first mode (of dimension $I_3$), but this matricization is also column-major in memory.
-We choose the ordering of the modes of $\T{T}$ such that each of $R$ contiguous blocks is used to compute one column of the output matrix via a matrix-vector operation with a corresponding column of the Khati-Rao product of the other factor matrices.
+We choose the ordering of the modes of $\T{T}$ such that each of $R$ contiguous blocks is used to compute one column of the output matrix via a matrix-vector operation with a corresponding column of the Khati-Rao product of the other factor matrices.}
 
-No matter how the dimension tree is designed, the computational cost of each partial MTTKRP is $O(IR)$, where $I$ is the number of tensor entries and $R$ is the rank of the CP decomposition.
+\red{No matter how the dimension tree is designed, the computational cost of each partial MTTKRP is $O(IR)$, where $I$ is the number of tensor entries and $R$ is the rank of the CP decomposition.
 This is the same operation count as a full MTTKRP.
 The computational cost of a multi-TTV is the number of entries in the temporary tensor, which is the product of a \emph{subset} of the original tensor dimensions multiplied by $R$.
-Thus, it is computationally cheaper than the partial MTTKRPs, but it is also memory bandwidth bound.
+Thus, it is computationally cheaper than the partial MTTKRPs, but it is also memory bandwidth bound.}
 
-The other subroutine necessary for implementing the dimension tree approach is the Khatri-Rao product of sets of factor matrices.
+\red{The other subroutine necessary for implementing the dimension tree approach is the Khatri-Rao product of sets of factor matrices.
 We implement the operation as a row-wise Hadamard product of a set of factor matrix rows, and we use OpenMP parallelization to obtain on-node parallelism.
-The computational cost of this operation is also typically lower order, but the running time in practice suffers from also being memory bandwidth bound.
+The computational cost of this operation is also typically lower order, but the running time in practice suffers from also being memory bandwidth bound.}
 
@@
 The inner iteration is displayed graphically in \Cref{fig:inner} for a 3-way example and an update of the $2$nd factor matrix.
-The main idea is that at the start of the $n$th inner iteration (\cref{fig:inner_a}), all of the data is in place for each processor to perform a local MTTKRP computation.
+The main idea is that at the start of the $n$th inner iteration (\Cref{fig:inner_a}), all of the data is in place for each processor to perform a local MTTKRP computation.
 This means that all processors in a slice redundantly own the same rows of the corresponding factor matrix (for all modes except $n$).
-After the local MTTKRP is computed (\cref{fig:inner_b}), each processor has computed a contribution to a subset of the rows of the global MTTKRP $\Mn{M}{n}$, but its contribution must be summed up with the contributions of all other processors in its mode-$n$ slice.
-This summation is performed with a Reduce-Scatter collective across the mode-$n$ processor slice that achieves a row-wise partition of the result (in \cref{fig:inner_c}, the light gray shading corresponds to the rows of $\Mn{M}{2}$ to which processor $(1,3,1)$ contributes and the dark gray shading corresponds to the rows it receives as output).
+After the local MTTKRP is computed (\Cref{fig:inner_b}), each processor has computed a contribution to a subset of the rows of the global MTTKRP $\Mn{M}{n}$, but its contribution must be summed up with the contributions of all other processors in its mode-$n$ slice.
+This summation is performed with a Reduce-Scatter collective across the mode-$n$ processor slice that achieves a row-wise partition of the result (in \Cref{fig:inner_c}, the light gray shading corresponds to the rows of $\Mn{M}{2}$ to which processor $(1,3,1)$ contributes and the dark gray shading corresponds to the rows it receives as output).
 The output distribution of the Reduce-Scatter is designed so that afterwards, the update of the factor matrix in that mode can be performed row-wise in parallel.
-Along with $\Mn{S}{n}$, which can be computed locally, each processor updates its own rows of the factor matrix given its rows of the MTTKRP result (\cref{fig:inner_d}).
-The remainder of the inner iteration is preparing and distributing the new factor matrix data for future inner iterations, which includes an All-Gather of the newly computed factor matrix $\Mn{H}{n}$ across mode-$n$ processor slices (\cref{fig:inner_e}) and recomputing $\Mn{G}{n}={\Mn{H}{n}}^\Tra\Mn{H}{n}$.
+Along with $\Mn{S}{n}$, which can be computed locally, each processor updates its own rows of the factor matrix given its rows of the MTTKRP result (\Cref{fig:inner_d}).
+The remainder of the inner iteration is preparing and distributing the new factor matrix data for future inner iterations, which includes an All-Gather of the newly computed factor matrix $\Mn{H}{n}$ across mode-$n$ processor slices (\Cref{fig:inner_e}) and recomputing $\Mn{G}{n}={\Mn{H}{n}}^\Tra\Mn{H}{n}$.
 
@@
 \smallskip
-\caption{Leading-order per-iteration costs in terms of computation (flops), communication (words moved), and memory (words).
+\caption{\red{Leading-order per-iteration costs in terms of computation (flops), communication (words moved), and memory (words).
 We ignore constants but omit big-Oh notation for clarity.
@@
 The second line corresponds to the changes in communication and memory for a general processor grid.
-The third line corresponds to the changes in computation and memory if the dimension tree optimization is not applied.}
+The third line corresponds to the changes in computation and memory if the dimension tree optimization is not applied.}}
 \label{tab:costs}
@@
 While the number of outer iterations is sensitive to the NLS method used, the outer iteration time is generally the same across NLS methods.
-We summarize the analysis in \cref{tab:costs}, showing the differences with and without an optimal processor grid and with and without using a dimension tree.
+\red{We summarize the analysis in \cref{tab:costs}, showing the differences with and without an optimal processor grid and with and without using a dimension tree.}
 
@@
 The cost of \cref{line:hadamard} is $O(NR^2)$, the cost of \cref{line:locNLS} is $O(R^3I_n/P)$, which is a loose upper bound for BPP and other methods \cite{KBP16}, and the cost of \cref{line:locSYRK} is $O(R(I_n/P)^2)$.
-The sum of these three costs across all inner iterations is $O(R^2N^2+(R^3/P)\sum I_n+(R/P^2)\sum I_n^2)$, which is dominated by the cost of the MTTKRP.
+The sum of these three costs across all inner iterations is \red{$O(R^2N^2+(R^3/P)\sum I_n+(R/P^2)\sum I_n^2)$}, which is dominated by the cost of the MTTKRP.
 When using dimension trees to perform the MTTKRP (\cref{line:locMTTKRP}), we compute the cost amortized over all inner iterations.

### 2018-05-30 latex/sc18/conclusion.tex
put changes in red

@@
 This observation is supported by the theoretical analysis: although the communication does not scale as well with $P$, the total amount of data depends on a sum of tensor dimensions rather the product of the tensor dimensions, which determines the total amount of computation.
-For a relative comparison, consider memory-efficient parallel dense $n\times n$ matrix multiplication: the ratio of the $O(n^3/P)$ local computation to the $O(n^2/P^{1/2})$ communication is approximately the square root of the size of the local data, or $O((n^2/P)^{1/2})$.
+\red{For a relative comparison, consider memory-efficient parallel dense $n\times n$ matrix multiplication: the ratio of the $O(n^3/P)$ local computation to the $O(n^2/P^{1/2})$ communication is approximately the square root of the size of the local data, or $O((n^2/P)^{1/2})$.
 In the case of NNCP, the ratio of local computation to communication is $O((I/P)^{1-1/N}/N)$, which is approximately the size of the local data raised to the power $1-1/N$.
 This exponent is larger than $1/2$ and grows with $N$, and therefore it predicts the NNCP computation should be more computation bound than matrix multiplication.
-Note that this analysis does not consider the type of local computation; for small $R$, the local computation will likely be memory bandwidth bound, but the algorithm will spend more time on local computation than on interprocessor communication.
+Note that this analysis does not consider the type of local computation; for small $R$, the local computation will likely be memory bandwidth bound, but the algorithm will spend more time on local computation than on interprocessor communication.}
 
@@
 Besides the reduction in flops, the dimension tree approach enjoys better DGEMM performance and avoids memory-bound KRP computations.
-Furthermore, we see that tuning the processor grid had much less effect on overall performance.
-Not only do reductions in communication not matter as much as computation, but different local tensor sizes can also cause variations in local performance that outweigh the savings in communication.
+\red{Furthermore, we see that tuning the processor grid had much less effect on overall performance.
+Not only do reductions in communication not matter as much as computation, but different local tensor sizes can also cause variations in local performance that outweigh the savings in communication. }
 

### 2018-05-30 latex/sc18/experiments.tex
put changes in red

@@
 We can observe from the figure that all the three algorithms scale nearly linearly as the problem remains compute bound. 
-Our algorithm (with the dimension tree optimization) achieves a speedup of $1771\times$ on 4096 cores over the same implementation running on 1 core.
+\red{Our algorithm (with the dimension tree optimization) achieves a speedup of $1771\times$ on 4096 cores over the same implementation running on 1 core.}
 Recall from \label{sec:analysis} that the computation scales linearly with $1/P$ while the communication scales with $1/P^{1/N}=1/P^{1/3}$. 
@@
 \end{tikzpicture}
-\caption{Strong scaling of 3D synthetic tensor with dimension $1024\times 1024\times 1024$ on processor grids $2^k\times 2^k\times 2^k$ for $k\in\{0,1,2,3,4\}$.  The rank is fixed at 32.}
+\caption{Strong scaling of 3D synthetic tensor with dimension $1024\times 1024\times 1024$ on processor grids $2^k\times 2^k\times 2^k$ for \red{$k\in\{0,1,2,3,4\}$}.  The rank is fixed at 32.}
 \label{fig:strongsynthetic3D}
@@
 
-\subsection{Varying Processor Grid}
+\subsection{\red{Varying Processor Grid}}
 
-In order to illustrate the effect of processor grid choice on running time, we show in \Cref{fig:commsweep10} a time breakdown over various processor grid choices for a 4D problem on 81 processors.
+\red{In order to illustrate the effect of processor grid choice on running time, we show in \Cref{fig:commsweep10} a time breakdown over various processor grid choices for a 4D problem on 81 processors.
 Because the tensor is cubical and 81 has a restricted factorization into 4 numbers, there are 5 distinct processor grids.
@@
 Furthermore, adjusting the processor grid affects the local tensor dimensions and the performance of the local computations, and the optimal processor grid led to slower local performance.
-For $R=10$, all of the local computation is memory bandwidth bound, and we believe the variations in running times to be effects of some temporary quantities fitting into smaller levels of cache.
+For $R=10$, all of the local computation is memory bandwidth bound, and we believe the variations in running times to be effects of some temporary quantities fitting into smaller levels of cache.}
 
@@
 \end{tikzpicture}
-\caption{Time breakdown for $243\times243\times243\times243$ tensor and rank $R=10$ on 81 processors for varying processor grids.}
+\caption{\red{Time breakdown for $243\times243\times243\times243$ tensor and rank $R=10$ on 81 processors for varying processor grids.}}
 \label{fig:commsweep10}

### 2018-05-30 latex/sc18/prelims.tex
put changes in red

@@
 Given matrices $\M{A}$ and $\M{B}$ that are $I_{A} \times R$ and $I_{B} \times R$, the KRP $\M{K} = \M{A} \Khat \M{B}$ is $I_{A}I_{B} \times R$. 
-It can be thought of as a row-wise Hadamard product, where $\M{K}(i+I_{A}(j{-}1),:) = \M{A}(i,:) \Hada \M{B}(j,:)$, or a column-wise Kronecker product, where $\M{K}(:,c) =  \M{A}(:,c) \Kron \M{B}(:,c)$.
+It can be thought of as a row-wise Hadamard product, where $\M{K}(i+I_{A}(j{-}1),:) = \M{A}(i,:) \Hada \M{B}(j,:)$, or a column-wise Kronecker product, where \red{$\M{K}(:,c) =  \M{A}(:,c) \Kron \M{B}(:,c)$}.
 
@@
 \times I_{N}$ (and $I=\prod I_n$) into the columns of a matrix.
-The Matricized-Tensor Times Khatri-Rao Product or MTTKRP will be central to this work and takes the form $\Mn{M}{n} = \Mz{A}{n} \Mn{K}{n}$, where $\Mn{K}{n}= \Mn{H}{N} \Khat \cdots \Khat \Mn{H}{n+1} \Khat \Mn{H}{n-1} \Khat  \cdots \Khat \Mn{H}{1}$.
+The Matricized-Tensor Times Khatri-Rao Product or MTTKRP will be central to this work and takes the form \red{$\Mn{M}{n} = \Mz{A}{n} \Mn{K}{n}$, where $\Mn{K}{n}= \Mn{H}{N} \Khat \cdots \Khat \Mn{H}{n+1} \Khat \Mn{H}{n-1} \Khat  \cdots \Khat \Mn{H}{1}$}.
 
@@
 %}
-The BCD iteration is guaranteed to converge to a stationary point, but there is no guarantee that it converges to a global minimum.
-The number of outer iterations to convergence is problem dependent but typically ranges from the 10s to the 1000s.
+\red{The BCD iteration is guaranteed to converge to a stationary point, but there is no guarantee that it converges to a global minimum.
+The number of outer iterations to convergence is problem dependent but typically ranges from 10s to 1000s.}
 

### 2018-05-30 latex/sc18/algorithm.tex
sc rebuttal submitted

@@
 \centering
-\begin{tabular}{|c|cccc|}
+\red{\begin{tabular}{|c|cccc|}
 \hline
@@
 \hline
-\end{tabular}
+\end{tabular}}
 \smallskip
@@
 While the number of outer iterations is sensitive to the NLS method used, the outer iteration time is generally the same across NLS methods.
-\red{We summarize the analysis in \cref{tab:costs}, showing the differences with and without an optimal processor grid and with and without using a dimension tree.}
+\red{We summarize the analysis in \Cref{tab:costs}, showing the differences with and without an optimal processor grid and with and without using a dimension tree.}
 

### 2018-06-06 latex/sc18/introduction.tex

@@
 The CP decomposition is a low-rank approximation of a multi-dimensional array, or tensor, which generalizes matrix approximations like the truncated singular value decomposition.
-It approximates the input tensor by a sum of rank-one tensors, which are outer-products of vectors. 
+\textcolor{red}{It approximates the input tensor by a sum of rank-one tensors, which are outer products of vectors. }
 CP is often used for finding hidden patterns, or latent factors, within tensor data, particularly when the goal is to interpret the factors, and it is popular within the signal processing, machine learning, and scientific computing communities.
@@
 
-We summarize our main contributions as follows:
+\red{We summarize our main contributions as follows:
 \begin{itemize}
@@
 		\item we demonstrate a performance improvement of up to $2.2\times$ over the existing state-of-the-art parallel software on 3D tensors,
-		\item our implementation obtains efficient parallel scaling of up to $253\times$ on $512$ cores.
-\end{itemize}
+		\item our implementation obtains efficient parallel scaling of up to $1771\times$ on 4096 cores.
+\end{itemize}}

### 2018-06-20 common/distutils.hpp
Adding relative directory paths for includes

@@ -4,9 +4,9 @@
 
 #include <mpi.h>
 #include <string>
-#include "distutils.h"
-#include "utils.h"
-#include "utils.hpp"
+#include "common/distutils.h"
+#include "common/utils.h"
+#include "common/utils.hpp"
 
 inline void mpitic() {
   // tictoc_stack.push(clock());

### 2018-06-20 common/ncpfactors.hpp

@@ -4,8 +4,8 @@
 #define COMMON_NCPFACTORS_HPP_
 
 #include <cassert>
-#include "tensor.hpp"
-#include "utils.h"
+#include "common/tensor.hpp"
+#include "common/utils.h"
 #ifdef MPI_DISTNTF
 #include <mpi.h>
 #endif
@@ -290,9 +290,9 @@ current_nrows *= rightkrp.n_rows;
     }
   }
   /*
-* this is for reinitializing random numbers across different
-* processors.
-*/
+   * this is for reinitializing random numbers across different
+   * processors.
+   */
   void randu(const int i_seed) {
     arma::arma_rng::set_seed(i_seed);
     for (int i = 0; i < this->m_modes; i++) {

### 2018-06-20 common/utils.hpp

@@ -9,7 +9,7 @@
 #include <vector>
 #include <stack>
 #include <typeinfo>
-#include "utils.h"
+#include "common/utils.h"
 #ifdef MKL_FOUND
 #include <mkl.h>
 #else

### 2018-06-20 dimtree/kobydt.hpp

@@ -5,8 +5,8 @@
 
 #include "dimtree/dimtrees.hpp"
 #include "dimtree/kdttensor.hpp"
-#include "ncpfactors.hpp"
-#include "tensor.hpp"
+#include "common/ncpfactors.hpp"
+#include "common/tensor.hpp"
 
 class KobyDimensionTree {
   ktensor *m_local_Y;

### 2018-06-20 distnmf/aunmf.hpp

@@ -7,8 +7,8 @@
 #include <armadillo>
 #include <string>
 #include <vector>
-#include "distnmf.hpp"
-#include "mpicomm.hpp"
+#include "distnmf/distnmf.hpp"
+#include "distnmf/mpicomm.hpp"
 
 /*
  * There are totally prxpc process.

### 2018-06-20 distnmf/naiveanlsbpp.hpp

@@ -3,8 +3,8 @@
 #ifndef DISTNMF_NAIVE_ANLS_BPP_HPP_
 #define DISTNMF_NAIVE_ANLS_BPP_HPP_
 #pragma once
-#include "bppnnls.hpp"
-#include "distnmf1D.hpp"
+#include "distnmf/distnmf1D.hpp"
+#include "nnls/bppnnls.hpp"
 
 template <class INPUTMATTYPE>
 class DistNaiveANLSBPP : public DistNMF1D<INPUTMATTYPE> {

### 2018-06-20 distntf/distauntf.hpp

@@ -6,12 +6,12 @@
 #include <armadillo>
 #include <string>
 #include <vector>
-#include "dimtree/kobydt.hpp"
-#include "distntf/distntfmpicomm.hpp"
-#include "distntftime.hpp"
-#include "luc.hpp"
-#include "common/ntf_utils.hpp"
 #include "common/distutils.hpp"
+#include "common/luc.hpp"
+#include "common/ntf_utils.hpp"
+#include "dimtree/kobydt.hpp"
+#include "distnmf/distntftime.hpp"
+#include "distntf/distntfmpicomm.hpp"
 
 /*
  * Tensor A of size is M1 x M2 x... x Mn is distributed among

### 2018-06-20 distntf/distntfio.hpp

@@ -7,11 +7,11 @@
 #include <armadillo>
 #include <string>
 #include <vector>
-#include "distntfmpicomm.hpp"
-#include "distutils.hpp"
-#include "ncpfactors.hpp"
-#include "npyio.hpp"
-#include "tensor.hpp"
+#include "common/distutils.hpp"
+#include "common/ncpfactors.hpp"
+#include "common/npyio.hpp"
+#include "common/tensor.hpp"
+#include "distntf/distntfmpicomm.hpp"
 
 /*
  * File name formats

### 2018-06-20 distntf/distntfmpicomm.hpp

@@ -7,7 +7,7 @@
 
 #include <mpi.h>
 #include <vector>
-#include "distntfutils.h"
+#include "distntf/distntfutils.h"
 namespace planc {
 
 class NTFMPICommunicator {
@@ -17,10 +17,10 @@ class NTFMPICommunicator {
   UVEC m_proc_grids;
   MPI_Comm m_cart_comm;
   // for mode communicators (*,...,p_n,...,*)
-  MPI_Comm* m_fiber_comm;
+  MPI_Comm *m_fiber_comm;
   // all communicators other than the given mode.
   // (p1,p2,...,p_n-1,*,p_n,...,p_M)
-  MPI_Comm* m_slice_comm;
+  MPI_Comm *m_slice_comm;
   UVEC m_fiber_ranks;
   UVEC m_slice_ranks;
 
@@ -57,7 +57,7 @@ class NTFMPICommunicator {
     }
   }
 
-  NTFMPICommunicator(int argc, char* argv[], const UVEC& i_dims)
+  NTFMPICommunicator(int argc, char *argv[], const UVEC &i_dims)
       : m_proc_grids(i_dims) {
     // Get the number of MPI processes
     MPI_Init(&argc, &argv);
@@ -123,13 +123,13 @@ class NTFMPICommunicator {
     delete[] m_slice_comm;
   }
 
-  const MPI_Comm& cart_comm() const { return m_cart_comm; }
-  void coordinates(int* o_c) const {
+  const MPI_Comm &cart_comm() const { return m_cart_comm; }
+  void coordinates(int *o_c) const {
     MPI_Cart_coords(m_cart_comm, m_global_rank, MPI_CART_DIMS, o_c);
   }
-  const MPI_Comm& fiber(const int i) const { return m_fiber_comm[i]; }
-  const MPI_Comm& slice(const int i) const { return m_slice_comm[i]; }
-  int rank(const int* i_coords) const {
+  const MPI_Comm &fiber(const int i) const { return m_fiber_comm[i]; }
+  const MPI_Comm &slice(const int i) const { return m_slice_comm[i]; }
+  int rank(const int *i_coords) const {
     int my_rank;
     MPI_Cart_rank(m_cart_comm, i_coords, &my_rank);
     return my_rank;

### 2018-06-20 nmf/bppnmf.hpp

@@ -4,8 +4,8 @@
 #define NMF_BPPNMF_HPP_
 
 #include <omp.h>
-#include "nmf.hpp"
-#include "bppnnls.hpp"
+#include "common/nmf.hpp"
+#include "nnls/bppnnls.hpp"
 
 // needed for precondition with hals
 #ifdef BUILD_SPARSE

### 2018-06-20 nmf/nmf.cpp

@@ -1,14 +1,14 @@
 /* Copyright 2016 Ramakrishnan Kannan */
 
-#include "nmf.hpp"
+#include "nmf/nmf.hpp"
 #include <stdio.h>
 #include <string>
-#include "aoadmm.hpp"
-#include "bppnmf.hpp"
-#include "hals.hpp"
-#include "mu.hpp"
-#include "parsecommandline.hpp"
-#include "utils.hpp"
+#include "nmf/aoadmm.hpp"
+#include "nmf/bppnmf.hpp"
+#include "nmf/hals.hpp"
+#include "nmf/mu.hpp"
+#include "common/parsecommandline.hpp"
+#include "common/utils.hpp"
 
 template <class NMFTYPE>
 void NMFDriver(int k, UWORD m, UWORD n, std::string AfileName,

### 2018-06-20 ntf/ntf.cpp

@@ -2,12 +2,12 @@
 
 #include <armadillo>
 #include <iostream>
-#include "auntf.hpp"
-#include "ncpfactors.hpp"
-#include "ntf_utils.hpp"
-#include "parsecommandline.hpp"
-#include "tensor.hpp"
-#include "utils.h"
+#include "common/ncpfactors.hpp"
+#include "common/ntf_utils.hpp"
+#include "common/parsecommandline.hpp"
+#include "common/tensor.hpp"
+#include "common/utils.h"
+#include "ntf/auntf.hpp"
 
 // ntf -d "2 3 4 5" -k 5 -t 20
 

### 2018-06-20 build/build.sh
header and script fixes. header files wrongly included

@@ -48,8 +48,8 @@ do
     pushd ../build_$SYSTEM\_dense_$cfg
     if [ "$SYSTEM" = "rhea" ]; then
         cmake $SRC_DIR/$cfg/
-    #we consider this as eos/titan
-    #we consider this as eos/titan
+    fi
+    #we consider this as eos/titan    
     if [ "$SYSTEM" = "eos" ]; then
         CC=CC CXX=CC cmake $SRC_DIR/$cfg/ -DCMAKE_IGNORE_MKL=1 -DCMAKE_BUILD_SPARSE=1
     fi
@@ -74,9 +74,7 @@ do
 
     if [ "$SYSTEM" = "titan" ]; then
         CC=CC CXX=CC cmake $SRC_DIR/$cfg/ -DCMAKE_IGNORE_MKL=1 -DCMAKE_BUILD_SPARSE=1
-    fi
-
-    
+    fi    
     make
     popd
 done

### 2018-06-20 common/utils.hpp

@@ -72,7 +72,7 @@ int random_sieve(const int nthprime) {
   mark = reinterpret_cast<int *>(calloc(nlimit, sizeof(int)));
 
   /* Calculate limit for k */
-  klimit = static_cast<int>(sqrt(static_cast<double>(nlimit) + 1)));
+  klimit = static_cast<int>(sqrt(static_cast<double>(nlimit) + 1));
 
   /* Mark the composites */
   /* Special case */

### 2018-06-20 distnmf/distio.hpp

@@ -6,7 +6,7 @@
 #include <unistd.h>
 #include <armadillo>
 #include <string>
-#include "distnmf/distutils.hpp"
+#include "common/distutils.hpp"
 #include "distnmf/mpicomm.hpp"
 
 /*

### 2018-06-20 distnmf/distnmf.hpp

@@ -4,9 +4,9 @@
 #define DISTNMF_DISTNMF_HPP_
 
 #include <string>
-#include "distnmftime.hpp"
+#include "common/nmf.hpp"
 #include "distnmf/mpicomm.hpp"
-#include "nmf/nmf.hpp"
+#include "distnmftime.hpp"
 #ifdef USE_PACOSS
 #include "pacoss.h"
 #endif

### 2018-06-20 distntf/distauntf.hpp

@@ -10,8 +10,8 @@
 #include "common/luc.hpp"
 #include "common/ntf_utils.hpp"
 #include "dimtree/kobydt.hpp"
-#include "distnmf/distntftime.hpp"
 #include "distntf/distntfmpicomm.hpp"
+#include "distntf/distntftime.hpp"
 
 /*
  * Tensor A of size is M1 x M2 x... x Mn is distributed among

### 2018-06-20 nmf/nmf.cpp

@@ -1,14 +1,14 @@
 /* Copyright 2016 Ramakrishnan Kannan */
 
-#include "nmf/nmf.hpp"
+#include "common/nmf.hpp"
 #include <stdio.h>
 #include <string>
+#include "common/parsecommandline.hpp"
+#include "common/utils.hpp"
 #include "nmf/aoadmm.hpp"
 #include "nmf/bppnmf.hpp"
 #include "nmf/hals.hpp"
 #include "nmf/mu.hpp"
-#include "common/parsecommandline.hpp"
-#include "common/utils.hpp"
 
 template <class NMFTYPE>
 void NMFDriver(int k, UWORD m, UWORD n, std::string AfileName,

### 2018-06-20 dimtree/ddt.hpp
refactoring koby dimension trees to dense dimension trees

@@ -1,14 +1,14 @@
 /* Copyright Koby Hayashi 2018 */
 
-#ifndef DIMTREE_KOBYDT_HPP_
-#define DIMTREE_KOBYDT_HPP_
+#ifndef DIMTREE_DDT_HPP_
+#define DIMTREE_DDT_HPP_
 
-#include "dimtree/dimtrees.hpp"
-#include "dimtree/kdttensor.hpp"
 #include "common/ncpfactors.hpp"
 #include "common/tensor.hpp"
+#include "dimtree/ddttensor.hpp"
+#include "dimtree/dimtrees.hpp"
 
-class KobyDimensionTree {
+class DenseDimensionTree {
   ktensor *m_local_Y;
   tensor *m_local_T;
   tensor *projection_Tensor;
@@ -20,10 +20,10 @@ class KobyDimensionTree {
   long int ldp;
   long int rdp;
 
-public:
-  KobyDimensionTree(const planc::Tensor &i_input_tensor,
-                    const planc::NCPFactors &i_ncp_factors,
-                    long int split_mode) {
+ public:
+  DenseDimensionTree(const planc::Tensor &i_input_tensor,
+                     const planc::NCPFactors &i_ncp_factors,
+                     long int split_mode) {
     m_local_T = reinterpret_cast<tensor *>(malloc(sizeof *m_local_T));
     m_local_Y = reinterpret_cast<ktensor *>(malloc(sizeof *m_local_Y));
     projection_Tensor =
@@ -98,7 +98,7 @@ public:
                 sizeof(double) * m_local_Y->dims[mode] * m_local_Y->rank);
   }
 
-  ~KobyDimensionTree() {
+  ~DenseDimensionTree() {
     for (long int i = 0; i < m_local_T->nmodes; i++) {
       free(m_local_Y->factors[i]);
     }
@@ -143,12 +143,12 @@ public:
       /**
           Updating the first factor matrix, do a partial MTTKRP
       */
-      D = ::direction::right; //   Contracting over the right modes
+      D = ::direction::right;  //   Contracting over the right modes
       LR_Ktensor_Reordering_existingY(
           m_local_Y, projection_Ktensor, s + 1,
-          opposite_direction(D)); // s+1 because s is included in the left
+          opposite_direction(D));  // s+1 because s is included in the left
 
-      if (projection_Ktensor->nmodes == 1) { // factor matrix output PM
+      if (projection_Ktensor->nmodes == 1) {  // factor matrix output PM
         /**
[…]

### 2018-06-20 dimtree/ddttensor.h

@@ -1,7 +1,7 @@
 /* Copyright Koby Hayashi 2018 */
 
-#ifndef DIMTREE_KDTTENSOR_H_
-#define DIMTREE_KDTTENSOR_H_
+#ifndef DIMTREE_DDTTENSOR_H_
+#define DIMTREE_DDTTENSOR_H_
 
 #include <cblas.h>
 #include <math.h>
@@ -233,4 +233,4 @@ void remove_mode_Ktensor(ktensor *Y, long int n);
 // mkl unavailable functions
 void vdMul(long int, double *, double *, double *);
 
-#endif  // DIMTREE_KDTTENSOR_H_
+#endif  // DIMTREE_DDTTENSOR_H_

### 2018-06-20 dimtree/ddttensor.hpp

@@ -1,9 +1,9 @@
 /* Copyright Koby Hayashi 2018 */
 
-#ifndef DIMTREE_KDTTENSOR_HPP_
-#define DIMTREE_KDTTENSOR_HPP_
+#ifndef DIMTREE_DDTTENSOR_HPP_
+#define DIMTREE_DDTTENSOR_HPP_
 
-#include "dimtree/kdttensor.h"
+#include "dimtree/ddttensor.h"
 
 /*
   Element wise multiplication between two vectors
@@ -1422,4 +1422,4 @@ direction opposite_direction(direction D) {
   }
 }
 
-#endif  // DIMTREE_KDTTENSOR_HPP_
+#endif  // DIMTREE_DDTTENSOR_HPP_

### 2018-06-20 dimtree/dimtrees.h

@@ -3,7 +3,7 @@
 #ifndef DIMTREE_DIMTREES_H_
 #define DIMTREE_DIMTREES_H_
 
-#include "dimtree/kdttensor.h"
+#include "dimtree/ddttensor.h"
 
 /**
   This .h file contains all the functions need to construct dimension

### 2018-06-20 dimtree/dimtrees.hpp

@@ -4,7 +4,7 @@
 #define DIMTREE_DIMTREES_HPP_
 
 #include "dimtree/dimtrees.h"
-#include "dimtree/kdttensor.hpp"
+#include "dimtree/ddttensor.hpp"
 
 /**
         Performs a tensor times a matrix based on some split of dimensions.

### 2018-06-20 distntf/distauntf.hpp

@@ -9,7 +9,7 @@
 #include "common/distutils.hpp"
 #include "common/luc.hpp"
 #include "common/ntf_utils.hpp"
-#include "dimtree/kobydt.hpp"
+#include "dimtree/ddt.hpp"
 #include "distntf/distntfmpicomm.hpp"
 #include "distntf/distntftime.hpp"
 
@@ -73,7 +73,7 @@ class DistAUNTF {
   double m_global_sqnorm_A;
   MAT hadamard_all_grams;
 
-  KobyDimensionTree *kdt;
+  DenseDimensionTree *kdt;
   // do the local syrk only for the current updated factor
   // and all reduce only for the current updated factor.
   // computes G^(current_mode)
@@ -409,8 +409,8 @@ class DistAUNTF {
       gather_ncp_factor(i);
     }
     if (this->m_enable_dim_tree) {
-      kdt = new KobyDimensionTree(m_input_tensor, m_gathered_ncp_factors,
-                                  m_input_tensor.modes() / 2);
+      kdt = new DenseDimensionTree(m_input_tensor, m_gathered_ncp_factors,
+                                   m_input_tensor.modes() / 2);
     }
 #ifdef DISTNTF_VERBOSE
     DISTPRINTINFO("local factor matrices::");

### 2018-06-20 ntf/auntf.hpp

@@ -12,7 +12,7 @@
 #include "common/ncpfactors.hpp"
 #include "common/ntf_utils.h"
 #include "common/tensor.hpp"
-#include "dimtree/kobydt.hpp"
+#include "dimtree/ddt.hpp"
 
 namespace planc {
 
@@ -44,7 +44,7 @@ class AUNTF {
   const algotype m_updalgo;
   LUC *m_luc;
   planc::Tensor *lowranktensor;
-  KobyDimensionTree *kdt;
+  DenseDimensionTree *kdt;
   bool m_enable_dim_tree;
 
  public:
@@ -86,8 +86,8 @@ class AUNTF {
   void dim_tree(bool i_dim_tree) {
     this->m_enable_dim_tree = i_dim_tree;
     if (i_dim_tree) {
-      this->kdt = new KobyDimensionTree(m_input_tensor, m_ncp_factors,
-                                        m_input_tensor.modes() / 2);
+      this->kdt = new DenseDimensionTree(m_input_tensor, m_ncp_factors,
+                                         m_input_tensor.modes() / 2);
     }
   }
   void num_it(const int i_n) { this->m_num_it = i_n; }

### 2018-06-20 latex/toms/algorithm.tex
seeding paper

\section{Parallel Algorithms}
\label{sec:parallel}

\subsection{Naive Parallel NMF Algorithm} 
\label{sec:naive}

In this section we present a naive parallelization of NMF algorithms, which has previously appeared in the context of a shared-memory parallel platform \cite{Fairbanks2015}. 
Each NLS problem with multiple right-hand sides can be parallelized based on the observation that each right-hand side is independent from the others. 
For example, we can solve several instances of Eq.~\eqref{eqn:single NLS} independently for different $\mathbf{b}$ where $\CC$ is fixed, which implies that we can optimize row blocks of $\WW$ and column blocks of $\HH$ in parallel. 
%\ramki{In this paper  we are using BPP to solve this Non-negative Least Squares (NLS) problem and call the algorithm as Naive Parallel BPP(\NaiveAlg). But the conversation in this section is not limited to \NaiveAlg and with out loss of generality is applicable to any algorithm that follows ANLS Algorithm \ref{alg:anlsnmf} such as Multiplicative Update (MU) and HALS as well.}

\begin{algorithm}[t!]
\caption{$[\WW,\HH] = \text{\NaiveAlg}(\AA,k)$}
\label{alg:naive}
\begin{algorithmic}[1]
\Require $\AA$ is an $m\times n$ matrix distributed both row-wise and column-wise across $p$ processors, $k$ is the approximation rank
\Require Local matrices: $\AA_{i}$ is $m/p\times n$, $\AA^{i}$ is $m\times n/p$, $\WW_i$ is $m/p\times k$, $\HH^i$ is $k\times n/p$
\State $p_i$ initializes $\HH^i$
\While{stopping criteria not satisfied}
	\Statex \quad\; \textbf{/* Compute $\WW$ given $\HH$ */} 
	\State collect $\HH$ on each processor using all-gather
		\label{line:naive:allgatherH}
	%\Statex \quad\; \textbf{/* Use Equation \eqref{eqn:muupdate}, \eqref{eqn:halsupdate} for \MU\ and \HALS\ respectively. ANLS/BPP will implement the SolveBPP function \ref{sec:BPP}.*/} 	
	\State $p_i$ computes $\WW_i \gets$ updateW$(\HH\HH^T,\AA_i\HH^T)$
		\label{line:naive:computeW}
	\Statex \quad\; \textbf{/* Compute $\HH$ given $\WW$ */} 
	\State collect $\WW$ on each processor using all-gather
		\label{line:naive:allgatherW}
	%\Statex \quad\; \textbf{/* Use Equation \eqref{eqn:muupdate},\eqref{eqn:halsupdate} for \MU\ and \HALS\ respectively. ANLS/BPP will implement the SolveBPP function \ref{sec:BPP}.*/} 	
	\State $p_i$ computes $(\HH^i)^T \gets$ updateH$(\WW^T\WW,(\WW^T\AA^i)^T)$
		\label{line:naive:computeH}
\EndWhile
\Ensure $\displaystyle \WW, \HH \approx \Argmin{\M{\tilde W} \geq 0, \M{\tilde H} \geq 0} \|\AA- \M{\tilde W} \M{\tilde H}\|$
\Ensure $\WW$ is an $m\times k$ matrix distributed row-wise across processors, $\HH$ is a $k\times n$ matrix distributed column-wise across processors
\end{algorithmic}
\end{algorithm}

\begin{figure}[t!]
\centering
\includegraphics[height=2.5in, width=\columnwidth]{fig/anlsbpp.pdf}
\caption{\NaiveAlg. Both rows and columns of $A$ are 1D distributed.  The algorithm works by (all-)gathering the entire fixed factor matrix to each processor and then performing the Local Update Computations to update the variable factor matrix.} 
\label{fig:naive}
\end{figure}

\renewcommand{\arraystretch}{1.25}
\newcommand{\smaller}{\scriptsize}
\begin{table*}[t!]
\begin{center}
\begin{tabular}{|c|c|c|c|c|}
\hline
\textbf{Algorithm} & \textbf{Flops} & \textbf{Words} & \textbf{Messages} & \textbf{Memory}  \\ \hline
\smaller \NaiveAlg & \smaller $4\frac{mnk}{p}+(m{+}n)k^2+F\lt(\frac mp,\frac np,k\rt)$ & \smaller $O((m+n)k)$ & \smaller $O(\log p)^*$ & \smaller $O\lt(\frac{mn}{p}+(m{+}n)k\rt)$ \\ \hline
\smaller \ParNMF\ ($m/p \geq n$) & \smaller $4\frac{mnk}{p}+\frac{(m+n)k^2}{p} + F\lt(\frac mp,\frac np,k\rt)$ & \smaller $O(nk)$ & \smaller $O(\log p)^*$ & \smaller $O\lt(\frac{mn}{p}+\frac{mk}{p}+nk\rt)$ \\ \hline
\smaller \ParNMF\ ($m/p < n$) & \smaller $4\frac{mnk}{p}+\frac{(m+n)k^2}{p}+F\lt(\frac mp,\frac np,k\rt)$ & \smaller $O\lt( \sqrt{\frac{mnk^2}{p}}\rt)$ & \smaller $O(\log p)^*$ & \smaller $O\lt(\frac{mn}{p}+\sqrt{\frac{mnk^2}{p}}\rt)$ \\ \hline
\smaller Lower Bound & $-$ & \smaller $\Omega\lt(\min\lt\{\sqrt{\frac{mnk^2}{p}},nk\rt\}\rt)$ & \smaller $\Omega(\log p)$ & \smaller $\frac{mn}{p}+\frac{(m+n)k}{p}$ \\ \hline
\end{tabular}
\normalsize
\end{center}
\caption{Leading order algorithmic costs for \NaiveAlg{} and \ParNMF{} (per iteration).  Note that the computation and memory costs assume the data matrix $\AA$ is dense, but the communication costs (words and messages) apply to both dense and sparse cases.
The function $F(\cdot)$ denotes the number of flops required for the particular NMF algorithm's Local Update Computation, aside from the matrix multiplications common across AU-NMF algorithms. 
Note that $F(m,n,k)$ is proportional to $m+n$ and not $mn$, so the term in the table scales linearly with $p$ (and not $p^2$) for all \LUC. \\
$^*$The stated latency cost assumes no communication is required in \LUC; \HALS\ requires $k\log p$ messages for normalization steps.}
\label{tab:costs}
\end{table*}%

Algorithm \ref{alg:naive} and Figure \ref{fig:naive} present a straightforward approach to parallelizing the independent subproblems.
Let us divide $\WW$ into row blocks $\WW_1, \ldots, \WW_p$ and $\HH$ into column blocks $\HH^1, \ldots, \HH^p$. 
We then double-partition the data matrix $\AA$ accordingly into row blocks $\AA_{1}, \ldots, \AA_p$ and column blocks $\AA^1, \ldots, \AA^p$ so that processor $i$ owns both $\AA_i$ and $\AA^i$ (see Figure \ref{fig:naive}).
With these partitions of the data and the variables, one can implement any AU-NMF algorithm in parallel, with only one communication step for each solve.

We summarize the algorithmic costs of Algorithm \ref{alg:naive} (derived in the following subsections) in Table \ref{tab:costs}.
This naive algorithm \cite{Fairbanks2015} has three main drawbacks: (1) it requires storing two copies of the data matrix (one in row distribution and one in column distribution) and both full factor matrices locally, (2) it does not parallelize the computation of $\HH\HH^T$ and $\WW^T\WW$ (each processor computes it redundantly), and (3) as we will see in Section \ref{sec:parNMF}, it communicates more data than necessary.

\subsubsection{Computation Cost}

The computation cost of Algorithm \ref{alg:naive} depends on the particular NMF algorithm used.
Thus, the computation at line \ref{line:naive:computeW} consists of computing $\AA_i\HH^T$, $\HH\HH^T$, and performing the algorithm-specific Local Update Computations for $m/p$ rows of $\WW$.
Likewise, the computation at line \ref{line:naive:computeH} consists of computing $\WW^T\AA^i$, $\WW^T\WW$, and performing the Local Update Computations for $n/p$ columns of $\HH$.
In the dense case, this amounts to $4mnk/p+(m+n)k^2+F(m/p,n/p,k)$ flops. 
Note that the first term has a constant 4 to account for both $\WW^T\AA$ and $\AA\HH^T$ and that the second term has a constant factor of 1 instead of 2 because the Gram computations ($\HH \HH^T$ and $\WW^T \WW$) exploit symmetry of the output matrix.
In the sparse case, processor $i$ performs $2(\nnz(\AA_i)+\nnz(\AA^i))k$ flops to compute $\AA^i\HH^T$ and $\WW^T\AA_i$ instead of $4mnk/p$.

\subsubsection{Communication Cost}

The size of $\WW$ is $mk$ words, and the size of $\HH$ is $nk$ words.
Thus, the communication cost of the all-gathers at lines \ref{line:naive:allgatherH} and \ref{line:naive:allgatherW}, based on the expression given in Section \ref{sec:collectives} is $\alpha\cdot 2\log p + \beta\cdot (m+n)k$.

\subsubsection{Memory Requirements} \label{sec:memory}

The local memory requirement includes storing each processor's part of matrices $\AA$, $\WW$, and $\HH$.
In the case of dense $\AA$, this is $2mn/p+(m+n)k/p$ words, as $\AA$ is stored twice; in the sparse case, processor $i$ requires $\nnz(\AA_i)+\nnz(\AA^i)$ words for the input matrix and $(m+n)k/p$ words for the output factor matrices.
Local memory is also required for storing temporary matrices $\WW$ and $\HH$ of size $(m+n)k$ words.

\subsection{\ParNMF}
\label{sec:parNMF}

We present our proposed algorithm, \ParNMF, as Algorithm \ref{alg:2D} and Figure \ref{fig:2D-Algorithm}. 
The main ideas of the algorithm are to (1) exploit the independence of Local Update Computations for rows of $\WW$ and columns of $\HH$ and (2) use communication-optimal matrix multiplication algorithms to set up the Local Update Computations.
The naive approach (Algorithm \ref{alg:naive}) shares the first property, by parallelizing over rows of $\WW$ and columns of $\HH$, but it uses parallel matrix multiplication algorithms that communicate more data than necessary.
The central intuition for communication-efficient parallel algorithms for computing $\HH\HH^T$, $\AA\HH^T$, $\WW^T\WW$, and $\WW^T\AA$ comes from a classification proposed by Demmel et al. \cite{DE+13}.
They consider three cases, depending on the relative sizes of the dimensions of the matrices and the number of processors; the four multiplies for NMF fall into either the ``one large dimension'' or ``two large dimensions" cases.
\ParNMF\ uses a careful data distribution in order to use a communication-optimal algorithm for each of the matrix multiplications, while at the same time exploiting the parallelism in the \LUC.

The algorithm uses a 2D distribution of the data matrix $\AA$ across a $p_r \times p_c$ grid of processors (with $p=p_rp_c$), as shown in Figure \ref{fig:2D-distribution}.
As we derive in the subsequent subsections, Algorithm \ref{alg:2D} performs an alternating method in parallel with a per-iteration bandwidth cost of $O\lt(\min\lt\{\sqrt{mnk^2/p},nk\rt\}\rt)$ words, latency cost of $O(\log p)$ messages, and load-balanced computation (up to the sparsity pattern of $\AA$ and convergence rates of local BPP computations). Figure \ref{fig:2D-Algorithm}, illustrates determining $\HH$ given $\WW$ with $pr=3$ and $pc=2$.

The main improvement of \ParNMF\ over \Naive\ involves the computation of $\AA\HH^T$ and $\WW^T\AA$.
By using a 2D distribution of the data matrix, no processor needs access to \emph{all} of one factor matrix, as in the case of \Naive, where each processor must access either all $m$ rows of $\WW$ or all $n$ columns of $\HH$.
Instead, with \ParNMF, each processor must access only $m/p_r$ of the rows of $\WW$ and $n/p_c$ of the columns of $\HH$, so the number of rows decreases as $p$ increases.
This implies the communication cost is reduced, as verified empirically in Figure \ref{fig:procsweep} (the extreme cases correspond to 1D distributions).

To minimize the communication cost and local memory requirements, in the typical case $p_r$ and $p_c$ are chosen so that $m/p_r\approx n/p_c\approx \sqrt{mn/p}$, in which case the bandwidth cost is $O\lt(\sqrt{mnk^2/p}\rt)$.
If the matrix is very tall and skinny, i.e., $m/p>n$, then we choose $p_r=p$ and $p_c=1$.
In this case, the distribution of the data matrix is 1D, and the bandwidth cost is $O(nk)$ words.

The matrix distributions for Algorithm \ref{alg:2D} are given in Figure \ref{fig:2D-distribution}; we use a 2D distribution of $\AA$ and 1D distributions of $\WW$ and $\HH$.
Recall from Table \ref{tab:notation} that $\M{M}_i$ and $\M{M}^i$  denote row and column blocks of $\M{M}$, respectively.
Thus, the notation $(\WW_i)_j$ denotes the $j$th row block within the $i$th row block of $\WW$.
Lines \ref{line:2DsyrkH}--\ref{line:2DcompW} compute $\WW$ for a fixed $\HH$, and lines \ref{line:2DsyrkW}--\ref{line:2DcompH} compute $\HH$ for a fixed $\WW$; note that the computations and communication patterns for the two alternating iterations are analogous.

In the rest of this section, we derive the per-iteration computation and communication costs, as well as the local memory requirements.
We also argue the communication-optimality of the algorithm in the dense case.
Table \ref{tab:costs} summarizes the results of this section and compares them to \NaiveAlg.

\begin{algorithm}[t!]
\caption{$[\WW,\HH] = \text{\ParNMF}(\AA,k)$}
\label{alg:2D}
\begin{algorithmic}[1]
\small
\Require $\AA$ is an $m\times n$ matrix distributed across a $p_r\times p_c$ grid of processors, $k$ is rank of approximation
\Require Local matrices: $\AA_{ij}$ is $m/p_r\times n/p_c$, $\WW_i$ is $m/p_r\times k$, $(\WW_i)_j$ is $m/p\times k$, $\HH_j$ is $k\times n/p_c$, and $(\HH_j)_i$ is $k\times n/p$
\State $p_{ij}$ initializes $(\HH_j)_i$
\While{stopping criteria not satisfied}
	\Statex \quad\; \textbf{/* Compute $\WW$ given $\HH$ */} 
	\State $p_{ij}$ computes $\M{U}_{ij}=(\HH_j)_i{(\HH_j)_i}^T$
		\label{line:2DsyrkH}
	\State compute $\HH\HH^T {=} \sum_{i,j} \M{U}_{ij}$ using all-reduce across all procs
		\label{line:2Dall-reduceH}
		\Comment{$\HH\HH^T$ is $k\times k$ and symmetric}
	\State $p_{ij}$ collects $\HH_j$ using all-gather across proc columns
		\label{line:2Dall-gatherH}
	\State $p_{ij}$ computes $\M{V}_{ij}=\AA_{ij}\HH_j^T$
		\label{line:2DNEW}
		\Comment{$\M{V}_{ij}$ is $m/p_r \times k$}
	\State compute $(\AA\HH^T)_i {=} \sum_j \M{V}_{ij}$ using reduce-scatter across proc row to achieve row-wise distribution of $(\AA\HH^T)_i$
		\Comment{$p_{ij}$ owns $m/p\times k$ submatrix $((\AA\HH^T)_i)_j$}
		\label{line:2Dreduce-scatterAHT}
		%\Statex \Comment{Use Equation \eqref{eqn:muupdate},\eqref{eqn:halsupdate} for \MU\ and \HALS\ respectively. ANLS/BPP will implement the SolveBPP function \ref{sec:BPP}.} 	
	\State $p_{ij}$ computes $(\WW_i)_j \gets$ UpdateW $(\HH\HH^T,((\AA\HH^T)_i)_j)$
		\label{line:2DcompW}
	\Statex \quad\; \textbf{/* Compute $\HH$ given $\WW$ */}
	\State $p_{ij}$ computes $\M{X}_{ij}={(\WW_i)_j}^T(\WW_i)_j$
		\label{line:2DsyrkW}
	\State compute $\WW^T\WW {=} \sum_{i,j} \M{X}_{ij}$ using all-reduce across all procs
		\label{line:2Dall-reduceW}
		\Comment{$\WW^T\WW$ is $k\times k$ and symmetric}
	\State $p_{ij}$ collects $\WW_i$ using all-gather across proc rows
		\label{line:2Dall-gatherW}
	\State $p_{ij}$ computes $\M{Y}_{ij}={\WW_i}^T\AA_{ij}$
		\label{line:2DNEH}
		\Comment{$\M{Y}_{ij}$ is $k\times n/p_c$}
	\State compute $(\WW^T\AA)^j = \sum_i \M{Y}_{ij}$ using reduce-scatter across proc columns to achieve column-wise distribution of $(\WW^T\AA)^j$
		\Comment{$p_{ij}$ owns $k\times n/p$ submatrix $((\WW^T\AA)^j)^i$}
		\label{line:2Dreduce-scatterWTA}
		%\Statex \Comment{Use Equation \eqref{eqn:muupdate},\eqref{eqn:halsupdate} for \MU\ and \HALS\ respectively. ANLS/BPP will implement the SolveBPP function \ref{sec:BPP}.}
	\State $p_{ij}$ computes $((\HH^j)^i)^T \gets$ UpdateH $(\WW^T\WW,(((\WW^T\AA)^j)^i)^T)$
		\label{line:2DcompH}
\EndWhile
\Ensure $\displaystyle \WW, \HH \approx \Argmin{\M{\tilde W} \geq 0, \M{\tilde H} \geq 0} \|\AA- \M{\tilde W} \M{\tilde H}\|$
\Ensure $\WW$ is an $m\times k$ matrix distributed row-wise across processors, $\HH$ is a $k\times n$ matrix distributed column-wise across processors
\normalsize
\end{algorithmic}
\end{algorithm}

\begin{figure}[t!]
\centering
\begin{subfigure}[b]{0.3\columnwidth}
\centering
\scalebox{.7}{\input{fig/1D-distribution}}
%\subcaption{1D Distribution with $p=p_r=4$ and $p_c=1$.}
\label{1D}
\end{subfigure} 
~
\begin{subfigure}[b]{0.6\columnwidth}
\centering
\scalebox{.75}{\input{fig/2D-distribution}}
%\subcaption{2D Distribution with $p_r=3$ and $p_c=2$.}
\label{2D}
\end{subfigure}
\caption{Data distributions for \ParNMF. 1D Distribution on left with $p=p_r=4$ and $p_c=1$. 2D Distribution on right with $p_r=3$ and $p_c=2$
Note that for the 2D distribution, $\AA_{ij}$ is $m/p_r \times m/p_c$, $\WW_i$ is $m/p_r \times k$, $(\WW_i)_j$ is $m/p\times k$, $\HH_j$ is $k\times n/p_c$, and $(\HH^j)^i$ is $k\times n/p$.}
\label{fig:2D-distribution}
\end{figure}
~
\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth, height=3in]{fig/mm2D.pdf}
\caption{Parallel matrix multiplications within \ParNMF\ for finding $\HH$ given $\WW$, with $p_r=3$ and $p_c=2$.  The computation of $\WW^T\WW$ appears on the far left; the rest of the figure depicts computation of $\WW^T\AA$.}
\label{fig:2D-Algorithm}
\end{figure*}

\subsubsection{Computation Cost}

Local matrix computations occur at lines \ref{line:2DsyrkH}, \ref{line:2DNEW}, \ref{line:2DsyrkW}, and \ref{line:2DNEH}.
In the case that $\AA$ is dense, each processor performs 
$$\frac np k^2+2\frac{m}{p_r}\frac{n}{p_c}k+\frac mp k^2+2\frac{m}{p_r}\frac{n}{p_c}k=4\frac{mnk}{p}+\frac{(m+n)k^2}{p}$$
flops. 
Recall that the second term on the right hand side has a constant factor of 1 instead of 2 because the local Gram computations (lines \ref{line:2DsyrkH} and \ref{line:2DsyrkW}) exploit symmetry.
In the case that $\AA$ is sparse, processor $(i,j)$ performs $(m+n)k^2/p$ flops in computing $\M{U}_{ij}$ and $\M{X}_{ij}$, and $4\nnz(\AA_{ij})k$ flops in computing $\M{V}_{ij}$ and $\M{Y}_{ij}$.
Local update computations occur at lines \ref{line:2DcompW} and \ref{line:2DcompH}.
In each case, the symmetric positive semi-definite matrix is $k\times k$ and the number of columns/rows of length $k$ to be computed are $m/p$ and $n/p$, respectively.
These costs together are given by $F(m/p,n/p,k)$.
There are computation costs associated with the all-reduce and reduce-scatter collectives (see Section \ref{sec:collectives}), both those contribute only to lower order terms: $O(k^2+mk/p_r+nk/p_c)$.

\subsubsection{Communication Cost}
\label{sec:alg:comm}

Communication occurs during six collective operations (lines \ref{line:2Dall-reduceH}, \ref{line:2Dall-gatherH}, \ref{line:2Dreduce-scatterAHT}, \ref{line:2Dall-reduceW}, \ref{line:2Dall-gatherW}, and \ref{line:2Dreduce-scatterWTA}).
We use the cost expressions presented in Section \ref{sec:collectives} for these collectives.
The communication cost of the all-reduces (lines \ref{line:2Dall-reduceH} and \ref{line:2Dall-reduceW}) is $\alpha \cdot 4\log p+\beta \cdot 2k^2$; 
the cost of the two all-gathers (lines \ref{line:2Dall-gatherH} and \ref{line:2Dall-gatherW}) is $\alpha \cdot \log p + \beta \cdot \lt( (p_r{-}1)nk/p + (p_c{-}1)mk/p\rt)$; and
the cost of the two reduce-scatters (lines \ref{line:2Dreduce-scatterAHT} and  \ref{line:2Dreduce-scatterWTA}) is $\alpha \cdot \log p + \beta \cdot \lt( (p_c{-}1)mk/p + (p_r{-}1)nk/p\rt)$.

We note that \LUC\ may introduce significant communication cost, depending on the NMF algorithm used.
The normalization of columns of $\WW$ within \HALS, for example, introduces an extra $k\log p$ latency cost.
We will ignore such costs in our general analysis.

In the case that $m/p<n$, we choose $p_r=\sqrt{mp/n} >1$ and $p_c=\sqrt{np/m}>1$, and these communication costs simplify to $\alpha \cdot O(\log p) + \beta \cdot O(mk/p_r+nk/p_c+k^2) = \alpha \cdot O(\log p) + \beta \cdot O(\sqrt{mnk^2/p}+k^2)$.
In the case that $m/p\geq n$, we choose $p_c=1$, and the costs simplify to $\alpha \cdot O(\log p) + \beta \cdot O(nk)$.

\ignore{
Thus, the total cost per iteration is
$$\gamma \cdot O\lt( \frac{mnk}{p} + C_\text{BPP}\lt(k,\frac{m+n}{p}\rt) \rt) + \beta \cdot O\lt(\frac{mk}{p_r}+\frac{nk}{p_c}+k^2\rt) + \alpha \cdot O(\log p).$$
In order to minimize the bandwidth cost of the algorithm, we choose $p_r=\sqrt{mp/n}$ and $p_c=\sqrt{np/m}$; this yields a per-iteration cost of 
$$\gamma \cdot O\lt( \frac{mnk}{p} + C_\text{BPP}\lt(k,\frac{m+n}{p}\rt) \rt) + \beta \cdot O\lt(\sqrt{\frac{mnk^2}{p}}+k^2\rt) + \alpha \cdot O(\log p).$$
We note that these choices of $p_r$ and $p_c$ are well defined assuming $m/p<n$ and $n/p<m$; if one of these assumptions is not satisfied we resort to Algorithm \ref{alg:1D}.
}

\subsubsection{Memory Requirements}
\label{sec:new_memory}

The local memory requirement includes storing each processor's part of matrices $\AA$, $\WW$, and $\HH$.
In the case of dense $\AA$, this is $mn/p+(m+n)k/p$ words; in the sparse case, processor $(i,j)$ requires $\nnz(\AA_{ij})$ words for the input matrix and $(m+n)k/p$ words for the output factor matrices.
Local memory is also required for storing temporary matrices $\WW_j$, $\HH_i$, $\M{V}_{ij}$, and $\M{Y}_{ij}$, of size $2mk/p_r+2nk/p_c$ words.

In the dense case, assuming $k<n/p_c$ and $k<m/p_r$, the local memory requirement is no more than a constant times the size of the original data.
For the optimal choices of $p_r$ and $p_c$, this assumption simplifies to $k<\max\lt\{\sqrt{mn/p},m/p\rt\}$. 
Note that the second argument of the max applies when the optimal distribution is 1D ($p_r=p$).

We note that if the temporary memory requirements become prohibitive, the computation of $((\AA \HH^T)_i)_j$ and $((\WW^T\AA)_j)_i$ via all-gathers and reduce-scatters can be blocked, decreasing the local memory requirements at the expense of greater latency costs.
When $\AA$ is sparse and $k$ is large enough, the memory footprint of the factor matrices can be larger than the input matrix.
In this case, the extra temporary memory requirements can become prohibitive; we observed this for a sparse data set with very large dimensions (see Section \ref{sec:webbase-2001}).
We leave the implementation of the blocked algorithm to future work.

\subsubsection{Communication Optimality}

In the case that $\AA$ is dense, Algorithm \ref{alg:2D} provably minimizes communication costs.
Theorem \ref{thm:LB} establishes the bandwidth cost lower bound for any algorithm that computes $\WW^T\AA$ or $\AA\HH^T$ each iteration.
A latency lower bound of $\Omega(\log p)$ exists in our communication model for any algorithm that aggregates global information \cite{CH+07}.
For NMF, this global aggregation is necessary in each iteration, for example, in order to compute residual error in the case that $\AA$ is distributed across all $p$ processors, because all processors have data that must be accumulated into the global error.
Based on the costs derived above, \ParNMF\ is communication optimal under the assumption $k<\sqrt{mn/p}$, matching these lower bounds to within constant factors.

\begin{theorem}[\cite{DE+13}]
\label{thm:LB}
Let $\AA \in \Rn{m \times n}$, $\WW \in \Rn{m \times k}$, and $\HH \in \Rn{k \times n}$ be dense matrices, with $k<n\leq m$.  If $k < \sqrt{mn/p}$, then any distributed-memory parallel algorithm on $p$ processors that load balances the matrix distributions and computes $\WW^T \AA$ and/or $\AA \HH^T$ must communicate at least $\Omega(\min\{\sqrt{mnk^2/p},nk\})$ words along its critical path.
\end{theorem}
\begin{proof}
The proof follows directly from \cite[Section II.B]{DE+13}.
Each matrix multiplication $\WW^T \AA$ and $\AA \HH^T$ has dimensions $k<n\leq m$, so the assumption $k<\sqrt{mn/p}$ ensures that neither multiplication has ``3 large dimensions.''
Thus, the communication lower bound is either $\Omega(\sqrt{mnk^2/p})$ in the case of $p>m/n$ (or ``2 large dimensions''), or $\Omega(nk)$, in the case of $p<m/n$ (or ``1 large dimension'').
If $p<m/n$, then $nk<\sqrt{mnk^2/p}$, so the lower bound can be written as $\Omega(\min\{\sqrt{mnk^2/p},nk\})$.
\end{proof}

We note that the communication costs of Algorithm \ref{alg:2D} are the same for dense and sparse data matrices (the data matrix itself is never communicated).
In the case that $\AA$ is sparse, this communication lower bound does not necessarily apply, as the required data movement depends on the sparsity pattern of $\AA$.
Thus, we cannot make claims of optimality in the sparse case (for general $\AA$).
The communication lower bounds for $\WW^T \AA$ and/or $\AA \HH^T$ (where $\AA$ is sparse) can be expressed in terms of hypergraphs that encode the sparsity structure of $\AA$ \cite{BDKS15}.
Indeed, hypergraph partitioners have been used to reduce communication and achieve load balance for a similar problem: computing a low-rank representation of a sparse tensor (without non-negativity constraints on the factors) \cite{KU15}.

### 2018-06-20 latex/toms/conclusion.tex
seeding paper

\section{Conclusion}\label{sec:conclusion}

In this paper, we propose a high-performance distributed-memory parallel framework for NMF algorithms that iteratively update the low rank factors in an alternating fashion.   
Our parallelization scheme is designed to avoid communication overheads and scales well to over 1500 cores. 
The framework is flexible, being (a) expressive enough to leverage many different NMF algorithms and (b) efficient for both sparse and dense matrices of sizes that span from a few hundreds to hundreds of millions.  
Our open-source software implementation is available for download.

%For the data sets on which we experimented, we showed that an efficient implementation of a naive parallel algorithm spends much of its time in interprocessor communication.
%For sparse problems, comparing p = 16 to p = 1536 (a factor increase of 96), we observe a speedups from \BPP\ of 59× (synthetic) and 22× (real world). For dense problems, comparing p = 96 to p = 1536 (a factor increase of 16), \BPP's speedup is 12× for both problems. The other NMF algorithms -- \MU\ and \HALS\, demonstrate similar scaling results in our framework.
For solving data mining problems at today's scale, parallel computation and distributed-memory systems are becoming prerequisites.
We argue in this paper that by using techniques from high-performance computing, the computations for NMF can be performed very efficiently.
Our framework allows for the HPC techniques (efficient matrix multiplication) to be separated from the data mining techniques (choice of NMF algorithm), and we compare data mining techniques at large scale, in terms of data sizes and number of processors.
One conclusion we draw from the empirical and theoretical observations is that the extra per-iteration cost of \BPP\ over alternatives like \MU\ and \HALS\ decreases as the number of processors $p$ increases, making \BPP\ more advantageous in terms of both quality and performance at larger scales.
By reporting time breakdowns that separate local computation from interprocessor communication, we also see that our parallelization scheme prevents communication from bottlenecking the overall computation; our comparison with a naive approach shows that communication can easily dominate the running time of each iteration. 
%In the case of \ParNMF, the problems remain computation bound on up to 600 processors, typically spending most of the time in local matrix multiplication or NLS solves.

%We focus in this work on BPP, %which is more expensive per-iteration than alternative methods like MU and HALS, 
%because it has been shown to reduce overall running time in the sequential case by requiring fewer iterations \cite{kim2011fast}.
%Because much of the time per iteration of \ParNMF\ is spent on local NLS, we believe further empirical exploration is necessary 
%to understand the proposed \ParNMF{}'s advantages for other AU-NMF algorithms such as MU and HALS. 
%%to confirm the advantages of BPP in the parallel case.
%We note that if we use the MU or HALS approach for determining low rank factors, the relative cost of interprocessor communication will grow, making the communication efficiency of our algorithm more important.

In future work, we would like to extend \ParNMF\  algorithm to dense and sparse tensors, computing the CANDECOMP/PARAFAC decomposition in parallel with non-negativity constraints on the factor matrices.
We plan on extending our software to include more NMF algorithms that fit the AU-NMF framework; these can be used for both matrices and tensors. 
We would also like to explore more intelligent distributions of sparse matrices: while our 2D distribution is based on evenly dividing rows and columns, it does not necessarily load balance the nonzeros of the matrix, which can lead to load imbalance in matrix multiplications.
We are interested in using graph and hypergraph partitioning techniques to load balance the memory and computation while at the same time reducing communication costs as much as possible.
%Finally, based our survey, experiments and knowledge, we would like to conclude that, so far we are the fastest distributed NMF algorithm in the literature. 
%Finally, we have not yet reached the limits of the scalability of \ParNMF; we would like to expand our benchmarks to larger numbers of nodes on the same size data sets to study performance behavior when communication costs completely dominate the running time.

### 2018-06-20 latex/toms/experiments.tex
seeding paper

\section{Experiments}
\label{sec:experiment}
\newcommand{\NLS}{LUC }

In this section, we describe our implementation of \ParNMF\ and evaluate its performance.
We identify a few synthetic and real world data sets to experiment with \ParNMF\ with dimensions that span from hundreds to millions. 
We compare the performance and exploring scaling behavior of different NMF algorithms -- \MU, \HALS, and ANLS/BPP (\BPP), implemented using the parallel \ParNMF\ framework.  
The code and the  datasets used for conducting the experiments can be downloaded from \url{https://github.com/ramkikannan/nmflibrary}.

\subsection{Experimental Setup}

\subsubsection{Data Sets}\label{sec:datasets}

We used sparse and dense matrices that are either synthetically generated or from real world applications. We explain the data sets in this section.

\begin{itemize}
\item Dense Synthetic Matrix: We generate a low rank matrix as the product of two uniform random matrices of size 207,360 $\times$ 100 and 100 $\times$ 138,240. 
The dimensions of this matrix are chosen to be evenly divisible for a particular set of processor grids.  
\item Sparse Synthetic Matrix: We generate a random sparse Erd\H{o}s-R\'{e}nyi matrix of the size 207,360 $\times$ 138,240 with density of 0.001.  That is, every entry is nonzero with probability 0.001.
\item Dense Real World Matrix ({\em Video}):  NMF is used on video data for background subtraction in order to detect moving objects. The low rank matrix $\hat{\AA} = \WW \HH$ represents background and the error matrix $\AA - \hat{\AA}$ represents moving objects.  %Detecting moving objects has many real-world applications such as traffic estimation \cite{Fujimoto2014} and security monitoring \cite{BSJJZ2015}.
In the case of detecting moving objects in streaming videos, the last several minutes of video is taken from the live video camera to construct the non-negative matrix. 
An algorithm to incrementally adjust the NMF based on the streaming video is presented in \cite{kim2013nonnegative}. 
To simulate this scenario, we collected a video in a busy intersection of the Georgia Tech campus at 20 frames per second. 
From this video, we took video for approximately 12 minutes and  then reshaped the matrix such that every RGB frame is a column of our matrix, so that the matrix is dense with size 1,013,400 $\times$ 13,824. 
\item Sparse Real World Matrix ({\em Webbase}): This data set is a directed sparse graph whose nodes correspond to webpages (URLs) and edges correspond to hyperlinks from one webpage to another.
The NMF output of this directed graph helps us understand clusters in graphs.
We consider two versions of the data set: {\em webbase-1M} and {\em webbase-2001}.
The dataset webbase-1M contains about 1 million nodes (1,000,005) and 3.1 million edges (3,105,536), and was first reported by Williams et al. \cite{Williams2009}.  
The version webbase-2001 has about 118 million nodes (118,142,155) and over 1 billion edges (1,019,903,190); it was first reported by Boldi and Vigna  \cite{Boldi2004}.  
Both data sets are available in the University of Florida Sparse Matrix Collection \cite{DH11} and the latter {\em webbase-2001} being the largest among the entire collection.
\item Text data ({\em Stack Exchange}): 
Stack Exchange is a network of question-and-answer websites on topics in varied fields, each site covering a specific topic, where questions, answers, and users are subject to a reputation award process. There are many Stack Exchange forums, such as {\em ask ubuntu, mathematics, latex}. 
We downloaded the latest anonymized dump of all 
user-contributed content on the Stack Exchange network from \cite{stkxchngdataset}. 
%Each site is formatted as a separate archive consisting of XML files zipped via 7-zip using bzip2 compression. 
%Each site archive includes Posts, Users, Votes, Comments, PostHistory and PostLinks. 
%For complete schema information, see the included readme.txt. 
We used only the questions from the most popular site called Stackoverflow and did not include the answers and comments. 
We removed the standard 571 English stop words (such as {\em are, am, be, above, below}) and then used snowball stemming available through the Natural Language Toolkit (NLTK) package \cite{LB2002}.
After this initial pre-processing, we deleted HTML tags (such as {\em lt, gt, em}) from the posts. 
The resulting bag-of-words matrix has a vocabulary of size 627,047 over 11,708,841 documents with 365,168,945 non-zero entries. 
In this data, the vocabulary is larger than the typical set of English words because it includes variables, constants, and other programming constructs of various programming languages from the user questions.
%We ran NMF on this dataset for finding 50 topics whose sample topic words are presented at the end of this section. 
\end{itemize}

The size of all the real world data sets were adjusted to the nearest size for uniformly distributing the matrix.

\subsubsection{Implementation Platform}

We conducted our experiments on ``Rhea'' at the Oak Ridge Leadership Computing Facility (OLCF).
Rhea is a commodity-type Linux cluster with a total of 512 nodes and a 4X FDR Infiniband interconnect.
Each node contains dual-socket 8-core Intel Sandy Bridge-EP processors and 128 GB of memory.
Each socket has a shared 20MB L3 cache, and each core has a private 256K L2 cache.

%\grey{Do we want to start using the acronym for our software and discuss it here? We will park this for camera ready}

Our objective of the implementation is using open source software as much as possible 
to promote reproducibility and reuse of our code.
The entire C++ code is developed using the matrix library Armadillo \cite{sanderson2010}. 
In Armadillo, the elements of the dense matrix are stored in column major order and the sparse matrices in Compressed Sparse Column (CSC) format.
For dense BLAS and LAPACK operations, we linked Armadillo with Intel MKL -- the default LAPACK/BLAS library in RHEA. It is also easy to link Armadillo with OpenBLAS \cite{xianyi2015}. 
We use Armadillo's own implementation of sparse matrix-dense matrix multiplication, the default GNU C++ Compiler (g++ (GCC) 4.8.2) and MPI library (Open MPI 1.8.4)  on RHEA.  We chose the commodity cluster with open source software so that the numbers presented here are representative of common use.

%\grey{Need to update this last phrase with compilation setup on Rhea}.

\subsubsection{Algorithms}

In our experiments, we considered the following algorithms: 
\begin{itemize}
	\item \MU: \ParNMF\ (Algorithm \ref{alg:2D}) with MU (Equation \eqref{eqn:muupdate})
	\item \HALS: \ParNMF\ (Algorithm \ref{alg:2D}) with HALS (Equation \eqref{eqn:halsupdate})
	\item \BPP: \ParNMF\ (Algorithm \ref{alg:2D}) with BPP (Section \ref{sec:BPP})
	\item \Naive: \NaiveAlg\ (Algorithm \ref{alg:naive}, Section \ref{sec:naive})
\end{itemize}

Our implementation of \Naive\ (Algorithm \ref{alg:naive}) uses BPP but can be easily to extended to \MU, \HALS, and other NMF algorithms. 
%A detailed comparison of \NaiveAlg\ with \ParNMF\ is made in our earlier work \cite{KBP16}. 
%We include some benchmark results from \Naive\ to reiterate the point that communication efficiency is key to obtaining reasonable performance, but we also omit other \Naive\ results in order to focus attention on comparisons among other algorithms.

For the algorithms based on \ParNMF, we use the processor grid that is closest to the theoretical optimum (see Section \ref{sec:alg:comm}) in order to minimize communication costs.
See Section \ref{sec:procgrid} for an empirical evaluation of varying processor grids for a particular algorithm and data set.

%We choose these three algorithms to confirm the following conclusions from the analysis of Section \ref{sec:parNMF}: the performance of a naive parallelization of \NaiveAlg\ (Algorithm \ref{alg:naive}) will be severely limited by communication overheads, and the correct choice of processor grid within Algorithm \ref{alg:2D} is necessary to optimize performance.
%To demonstrate the latter conclusion, we choose the two extreme choices of processor grids and test some data sets where a 1D processor grid is optimal (e.g., the Video matrix) and some where a squarish 2D grid is optimal (e.g., the Webbase matrix).

To ensure fair comparison among algorithms, the same random seed is used across different methods appropriately. 
That is, the initial random matrix $\HH$ is generated with the same random seed when testing with 
different algorithms (note that $\WW$ need not be initialized). 
In our experiments, we use number of iterations as the stopping criteria for all the algorithms.

%\grey{Need to update the following paragraph for our running times, maybe consider moving it elsewhere...}
While we would like to compare against other high-performance NMF algorithms in the literature, the only other distributed-memory implementations of which we're aware are implemented using Hadoop and are designed only for sparse matrices \cite{liao2014cloudnmf},
\cite{liu2010distributed}, \cite{gemulla2011large}, \cite{Yin2014} and \cite{Faloutsos2014}.
We stress that Hadoop is not designed for high performance computing of iterative numerical 
algorithms, requiring disk I/O between steps, so a run time comparison between a Hadoop 
implementation and a C++/MPI implementation is not a fair comparison of parallel algorithms.
A qualitative example of differences in run time is that a Hadoop implementation of the MU algorithm on 
a large sparse matrix of size $2^{17} \times 2^{16}$ with $2 \times {10^8}$ nonzeros (with k=8) 
takes on the order of 50 minutes per iteration \cite{liu2010distributed}, while our MU implementation 
takes 0.065 seconds per iteration for the synthetic data set (which is an order of magnitude larger in 
terms of rows, columns, and nonzeros) running on only 16 nodes. 
%Based on our survey, we are the fastest distributed NMF algorithm in the literature.

% input file with plotting macros
\input{plots.tex}

\subsection{Relative Error over Time} \label{sec:convergence}

There are various metrics to compare the quality of the 
NMF algorithms \cite{kim2013nonnegative}. The most common among these metrics are (a) relative error and (b) projected 
gradient. The former represents the closeness of the low rank approximation $\hat{\AA}\approx\WW\HH$, which is generally the optimization objective. 
The latter 
represent the quality of the produced low rank factors and the stationarity of the final solution. These 
metrics are also used as the stopping criterion for terminating the iteration of the NMF algorithm as in 
line \ref{algo:nmfloop} of Algorithm \ref{alg:aunmf}. Typically a combination of the number of iterations 
along with improvement of these metrics until a tolerance is met is used as stopping criterion. In this paper, we use 
relative error for the comparison as it is monotonically decreasing, as opposed to projected gradient of the 
low rank factors, which shows oscillations over iterations. The relative error can be formally defined as 
$\|\AA-\WW\HH\|_F/\|\AA\|_F$.

In Figure \ref{fig:convergence}, we measure the relative error at the end of every iteration (i.e., after the updates of both $\WW$ and $\HH$) for all three algorithms \MU, \HALS, and 
\BPP, and we plot the relative error over time (each mark represents an iteration).
We consider three real world datasets, \emph{video}, \emph{stack exchange} and \emph{webbase-1M}, and set $k=50$. 
We used only the number of iterations as stopping criterion and,  
just for this section, ran all the algorithms for 30 iterations. 
We note that the convergence behavior and computed factors can vary over different initializations; we used the same initial values across all three algorithms in these experiments.
Also, we observed that for these data sets, the convergence behavior was not sensitive to initialization (the final residual errors varied by less than 1\% in our experiments).
NMF solutions are guaranteed to be unique in certain cases, with mild assumptions on the input data \cite{BLRG14,HFS2016}, but we do not check those assumptions for these datasets.

To begin with, we explain the observations on the dense \emph{video} dataset presented in Figure \ref{fig:denserwerr}. 
The relative error of \MU\ is highest at 0.1812 after 30 iterations and \HALS's is the least with 0.1273; \BPP's relative error
is  0.1716 after 30 iterations. 
%For this data, although the per-iteration times vary across algorithms and \BPP\ has the longest per-iteration time, it converged more quickly than \HALS\ or \MU\ because it solves subproblems exactly.
%From the figure, we can observe that \BPP\  error didn't change after 29 iterations where as \HALS\ and \MU\ was still improving marginally at the 4th decimal even after 30 iterations.

We can observe that the relative error of \emph{stack exchange} from Figure \ref{fig:stackexchangeerr} is better 
than \emph{webbase-1M} from Figure \ref{fig:sparserwerr} over all three algorithms. 
In the case of the \emph{stack exchange} dataset, the relative errors after 30 iterations follow the pattern \MU\ $>$ \HALS\ $>$ \BPP, with values 0.8509, 0.8395, and 0.8377 respectively. 
%Unlike the \emph{video} dataset, both \MU\ and \HALS\ stopped improving after 23 iterations, 
%where as \BPP\ was still improving in the 4th decimal even though its error was better than the others. 
However, the difference in relative error for the \emph{webbase-1M} dataset is negligible, though the relative ordering of \MU\ $>$ \HALS\ $>$ \BPP\ is consistent, with values of 0.99927 for \MU\, 0.99920 for \HALS\ and 0.99919 for \BPP.

In general, for these datasets \BPP\ identifies better approximations and converges faster than \MU\ and \HALS\ despite the extra per-iteration time, which is consistent with the 
literature \cite{kim2013nonnegative,kim2011fast}. 
However, for the sparse datasets, the differences in relative error are small across the NMF algorithms.

%The Figure \ref{fig:stackexchangerr} shows the comparison of these algorithms on the stack exchange dataset explained in Section \ref{sec:datasets}.

%\captionsetup[subfigure]{labelfont=normalfont,textfont=normalfont,singlelinecheck=off,justification=raggedright}

\begin{figure}

% set which run to plot
\renewcommand{\run}{3}

\begin{subfigure}{0.3 \columnwidth}
\ylabeltrue
\begin{tikzpicture}
\renewcommand{\datafile}{data/denserwerr-time.dat}
\renewcommand{\run}{1}
\relerrplot
\renewcommand{\run}{3}
\end{tikzpicture}
\ylabelfalse
\subcaption{Video}
\label{fig:denserwerr}
\end{subfigure}
~
\begin{subfigure}{0.3 \columnwidth}
\begin{tikzpicture}
\renewcommand{\datafile}{data/stkx-5runs-err.dat}
\legendtrue
\relerrplot
\legendfalse
\end{tikzpicture}
\subcaption{Stack Exchange}
\label{fig:stackexchangeerr}
\end{subfigure}
~
\begin{subfigure}{0.3 \columnwidth}
\begin{tikzpicture}
\renewcommand{\datafile}{data/webbase1M-5runs-err.dat}
\relerrplot
\end{tikzpicture}
\subcaption{Webbase}
\label{fig:sparserwerr}
\end{subfigure}

\caption{Relative error comparison of \MU, \HALS, \BPP\ on real world datasets.}
\label{fig:convergence}
\end{figure}

\subsection{Time Per Iteration}

In this section we focus on per-iteration time of all the algorithms.
We report four types of experiments, varying the number of processors (Section \ref{sec:scaling}), the rank of the approximation (Section \ref{sec:ksweep}), the shape of the processor grid (Section \ref{sec:procgrid}), and scaling up the dataset size.
For each experiment we report a time breakdown in terms of the overall computation and communication steps (described in Section \ref{sec:perf-breakdown}) shared by all algorithms.

\subsubsection{Time Breakdown}
\label{sec:perf-breakdown}

To differentiate the computation and communication costs among the algorithms, we present the time breakdown among the various tasks within the algorithms for all performance experiments.
For Algorithm \ref{alg:2D}, there are three local computation tasks and three communication tasks to compute each of the factor matrices:
\begin{itemize}
	\item \textbf{MM}, computing a matrix multiplication with the local data matrix and one of the factor matrices;
	\item \textbf{\LUC}, local updates either using \BPP\ or applying the remaining work of the \MU\ or \HALS\ updates (i.e., the total time for both $UpdateW$ and $UpdateH$ functions); 
	\item \textbf{Gram}, computing the local contribution to the Gram matrix;
	\item \textbf{All-Gather}, to compute the global matrix multiplication;
	\item \textbf{Reduce-Scatter}, to compute the global matrix multiplication;
	\item \textbf{All-Reduce}, to compute the global Gram matrix.
\end{itemize}
In our results, we do not distinguish the costs of these tasks for $\WW$ and $\HH$ separately; we report their sum, though we note that we do not always expect balance between the two contributions for each task.
Algorithm \ref{alg:naive} performs all of these tasks except Reduce-Scatter and All-Reduce; all of its communication is in All-Gather.

\subsubsection{Scaling \texorpdfstring{$p$}{p}: Strong Scaling}
\label{sec:scaling}

Figure \ref{fig:scaling} presents a strong scaling experiment with four data sets: \emph{sparse synthetic}, \emph{dense synthetic}, \emph{webbase-1M}, and \emph{video}.
In this experiment, for each data set and algorithm, we use low rank $k=50$ and vary the number of processors (with fixed problem size).
We use $\{1,6,24,54,96\}$ nodes; since each node has 16 cores, this corresponds to $\{16,96,384,864,1536\}$ cores.
We report average per-iteration times.

We highlight three main observations from these experiments:
\begin{enumerate}
	\item \label{obs:1} \Naive\ is slower than all other algorithms for large $p$;
	\item \label{obs:2} \MU, \HALS, and \BPP\ (algorithms based on \ParNMF) scale up to over 1000 processors;
	\item \label{obs:3} the relative per-iteration cost of \LUC\ decreases as $p$ increases (for all algorithms), and therefore the extra per-iteration cost of \BPP\ (compared with \MU\ and \HALS) becomes negligible.
\end{enumerate}

\paragraph{Observation \ref{obs:1}} 
%We report \Naive\ performance only for the synthetic data sets (Figures \ref{fig:sparsesynstrongscale} and \ref{fig:densesynscaling}); the results for the real-world data sets are similar.
For the Sparse Synthetic data set, \Naive\ is $4.2\times$ slower than the fastest algorithm (\BPP) on 1536 processors; for the Dense Synthetic data set, \Naive\ is $1.6\times$ slower than the fastest algorithm (\MU) at that scale.
The slowdown increases to $7.7\times$ and $3.6\times$ for the sparse and dense real-world datasets, respectively.
Nearly all of this slowdown is due to the communication costs of \Naive.
Theoretical and practical evidence supporting the first observation is also reported in our previous paper \cite{KBP16}.
However, we also note that \Naive\ is the fastest algorithm for the smallest $p$ for each problem, which is largely due to reduced MM time.
Each algorithm performs exactly the same number of flops per MM; the efficiency of \Naive\ for small $p$ is due to cache effects.
For example, for the Dense Synthetic problem on 96 processors, the output matrix of \Naive's MM fits in L2 cache, but the output matrix of \ParNMF's MM does not; these effects disappear as $p$ increases.

\paragraph{Observation \ref{obs:2}}
Algorithms based on \ParNMF\ (\MU, \HALS, \BPP) scale well, up to over 1000 processors.
All algorithms' run times decrease as $p$ increases, with the exception of the Sparse Real World data set, in which case all algorithms slow down scaling from $p=864$ to $p=1536$ (we attribute this lack of scaling to load imbalance).
For sparse problems, comparing $p=16$ to $p=1536$ (a factor increase of 96), we observe speedups from \BPP\ of $59\times$ (synthetic) and $22\times$ (real world).
For dense problems, comparing $p=96$ to $p=1536$ (a factor increase of 16), \BPP's speedup is $12\times$ for both problems.
\MU\ and \HALS\ demonstrate similar scaling results.
For comparison, speedups for \Naive\ were $8\times$ and $3\times$ (sparse) and $6\times$ and $4\times$ (dense).

\paragraph{Observation \ref{obs:3}}
\MU, \HALS, and \BPP\ share all the same subroutines except those that are characterized as \LUC.
Considering only \LUC\ subroutines, \MU\ and \HALS\ require fewer operations than \BPP. However, \HALS\ has to make one additional communication for normalization of $\WW$. 
For small $p$, these cost differences are apparent in Figure \ref{fig:scaling}.
For example, for the sparse real world data set on 16 processors, \BPP's \LUC\ time is $16\times$ that of $\MU$, and the per iteration time differs by a factor of $4.5$.
However, as $p$ increases, the relative time spent in \LUC\ decreases, so the extra time taken by \BPP\ has less of an effect on the total per iteration time.
By contrast, for the dense real world data set on 1536 processors, \BPP\ spends a factor of 27 times more time in \LUC\ than \MU\ but only $11\%$ longer over the entire iteration.
For the synthetic data sets, \LUC\ takes $24\%$ (sparse) on 16 processors and $84\%$ (dense) on 96 processors, and that percentage drops to $11\%$ (sparse) and $15\%$ (dense) on 1536 processors.

These trends can also be seen theoretically (Table \ref{tab:costs}).
We expect local computations like MM, \LUC, and Gram to scale like $1/p$, assuming load balance is preserved.
If communication costs are dominated by the number of words being communicated (i.e., the communication is bandwidth bound), then we expect time spent in communication to scale like $1/\sqrt p$, and at least for dense problems, this scaling is the best possible.
Thus, communication costs will eventually dominate computation costs for all NMF problems, for sufficiently large $p$.
(Note that if communication is latency bound and proportional to the number of messages, then time spent communicating actually increases with $p$.)

The overall conclusion from this empirical and theoretical observation is that the extra per-iteration cost of \BPP\ over alternatives like \MU\ and \HALS\ decreases as the number of processors $p$ increases.
As shown in Section \ref{sec:convergence} the faster error reduction of \BPP\ typically reduces the overall time to solution compared with the alternatives even it requires more time for each iteration.
Our conclusion is that as we scale up $p$, this tradeoff is further relaxed so that \BPP\ becomes more and more advantageous for both quality and performance.

%\captionsetup[subfigure]{labelfont=normalfont,textfont=normalfont,singlelinecheck=off,justification=raggedright}

\begin{figure}

\naivetrue
\ksweepfalse
\legendtrue

\begin{subfigure}{\columnwidth}
%\centering
\begin{tikzpicture}
\renewcommand{\datafile}{data/sparsesynstrong_scale_pgf.dat}
\makeplot
\labels
\end{tikzpicture}
\subcaption{Sparse Synthetic}
\label{fig:sparsesynstrongscale}
\end{subfigure}
\legendfalse

\begin{subfigure}{\columnwidth}
%\centering
\begin{tikzpicture}
\renewcommand{\datafile}{data/densesynstrong_scale_pgf.dat}
\makeplot
\labels
\end{tikzpicture}
\subcaption{Dense Synthetic}
\label{fig:densesynscaling}
\end{subfigure}

%\naivefalse

\begin{subfigure}{\columnwidth}
%\centering
\renewcommand{\datafile}{data/sparserwstrong_scale_pgf.dat}
\begin{tikzpicture}
\makeplot
\labels
\end{tikzpicture}
\subcaption{Sparse Real World (webbase-1M)}
\label{fig:sparserwscaling}
\end{subfigure}

\begin{subfigure}{\columnwidth}
%\centering
\renewcommand{\datafile}{data/denserwstrong_scale_pgf.dat}
\begin{tikzpicture}
\makeplot
\labels
\end{tikzpicture}
\subcaption{Dense Real World (Video)}
\label{fig:denserwscaling}
\end{subfigure}

\caption{Per-iteration times with $k=50$, varying $p$ (strong scaling).}
\label{fig:scaling}
\end{figure}

\subsubsection{Scaling \texorpdfstring{$k$}{k}}
\label{sec:ksweep}

Figure \ref{fig:ksweep} presents an experiment scaling up the low rank value $k$ from 10 to 50 with each of the four data sets.
In this experiment, for each data set and algorithm, the problem size is fixed and the number of processors is fixed to $p=864$.
As in Section \ref{sec:scaling}, we report the average per-iteration times.
%We also omit \Naive\ data for the real world data sets to highlight the comparisons among \MU, \HALS, and \BPP.

We highlight two observations from these experiments:
\begin{enumerate}
	\item \label{obs:a} \Naive\ is plagued by communication time that increases linearly with $k$;
	\item \label{obs:b} \BPP's time increases more quickly with $k$ than those of \MU\ or \HALS;
\end{enumerate}

\paragraph{Observation \ref{obs:a}} 
We see from the synthetic data sets (Figures \ref{fig:sparsesynksweep} and \ref{fig:densesynksweep}) that the overall time of \Naive\ increases more rapidly with $k$ than any other algorithm and that the increase in time is due mainly to communication (All-Gather).
Table \ref{tab:costs} predicts that \Naive\ communication volume scales linearly with $k$, and we see that in practice the prediction is almost perfect with the synthetic problems.
This confirms that the communication is dominated by bandwidth costs and not latency costs (which are constant with respect to $k$).
We note that the communication cost of \ParNMF\ scales like $\sqrt k$, which is why we don't see as dramatic an increase in communication time for \MU, \HALS, or \BPP\ in Figure \ref{fig:ksweep}.

\paragraph{Observation \ref{obs:b}}
 Focusing attention on time spent in \NLS computations, we can compare how \MU, \HALS, and \BPP\ scale differently with $k$.
 We see a more rapid increase of \NLS time for \BPP\ than \MU\ or \HALS; this is expected because the \NLS computations unique to \BPP\ require between $O(k^3)$ and $O(k^4)$ operations (depending on the data) while the unique \NLS computations for \MU\ and \HALS\ are $O(k^2)$, with all other parameters fixed.
%It is generally observed that in both dense and sparse datasets, higher the low rank $k$ lesser the error \cite{kim2011fast}. For eg., the relative error of NMF algorithm with low rank $k=10$ will be more than the low rank $k=50$. 
Thus, the extra per-iteration cost of \BPP\ increases with $k$, so the advantage of \BPP\ of better error reduction must also increase with $k$ for it to remain superior at large values of $k$.
%\grey{can we confirm this in convergence section?}
We also note that although the number of operations within MM grows linearly with $k$, we do not observe much increase in time from $k=10$ to $k=50$; this is due to the improved efficiency of local MM for larger values of $k$.

\begin{figure}

\naivetrue
\ksweeptrue
\legendtrue

\begin{subfigure}{\columnwidth}
%\centering
\renewcommand{\datafile}{data/sparsesyn864_ksweep_scale_pgf.dat}
\begin{tikzpicture}
\makeplot
\labels
\end{tikzpicture}
\subcaption{Sparse Synthetic}
\label{fig:sparsesynksweep}
\end{subfigure}
\legendfalse

\begin{subfigure}{\columnwidth}
%\centering
\renewcommand{\datafile}{data/densesyn864_ksweep_scale_pgf.dat}
\begin{tikzpicture}
\makeplot
\labels
\end{tikzpicture}
\subcaption{Dense Synthetic}
\label{fig:densesynksweep}
\end{subfigure}

%\naivefalse

\begin{subfigure}{\columnwidth}
%\centering
\renewcommand{\datafile}{data/sparserwksweep_scale_pgf.dat}
\begin{tikzpicture}
\makeplot
\labels
\end{tikzpicture}
\subcaption{Sparse Real World (webbase-1M)}
\label{fig:sparserwksweep}
\end{subfigure}

\begin{subfigure}{\columnwidth}
\renewcommand{\datafile}{data/denserw_ksweep_scale_pgf.dat}
\begin{tikzpicture}
\makeplot
\labels
\end{tikzpicture}
\caption{Dense Real World (Video)}
\label{fig:denserwksweep}
\end{subfigure}

\caption{Per-iteration times with $p=864$, varying low rank $k$.}
\label{fig:ksweep}
\end{figure}

\subsubsection{Varying Processor Grid}
\label{sec:procgrid}

In this section we demonstrate the effect of the dimensions of the processor grid on per-iteration performance.
%As described in Section \ref{sec:alg:comm}, Algorithm \ref{alg:2D} is correct for any $p_r$ and $p_c$, but 
For a fixed total number of processors $p$, the communication cost of Algorithm \ref{alg:2D} varies with the choice of $p_r$ and $p_c$.
To minimize the amount of data communicated, the theoretical analysis suggests that the processor grid should be chosen to make the sizes of the local data matrix as square as possible.
This implies that if $m/p > n$, $p_r=p$ and $p_c=1$ is the optimal choice (a 1D processor grid); likewise if $n/p > m$ then a 1D processor grid with $p_r=1$ and $p_c=p$ is the optimal choice.
Otherwise, a 2D processor grid minimizes communication with $p_r \approx \sqrt{mp/n}$ and $p_c \approx \sqrt{np/m}$ (subject to integrality and $p_rp_c=p$).

Figure \ref{fig:procsweep} presents a benchmark of \BPP\ for the Sparse Synthetic data set for fixed values of $p$ and $k$.
We vary the processor grid dimensions from both 1D grids to the 2D grid that matches the theoretical optimum exactly.
Because the sizes of the Sparse Synthetic matrix are $172{,}800\times115{,}200$ and the number of processors is 1536, the theoretically optimal grid is $p_r = \sqrt{mp/n} = 48$ and $p_c = \sqrt{np/m} = 32$.
The experimental results confirm that this processor grid is optimal, and we see that the time spent communicating increases as the processor grid deviates from the optimum, with the 1D grids performing the worst.

\begin{figure}
\renewcommand{\datafile}{data/sparsesynpsweep_scale_pgf.dat}
\centering
\begin{tikzpicture}
\begin{axis}[
	ybar stacked,
	reverse legend,
	bar width=12pt,
	width=7cm, height=5cm,
	ylabel={Time (seconds)}, 
	y label style={yshift=-0.5cm},
	x label style={yshift=-0.5cm},
	ymin=0,
	symbolic x coords={1-1536,8-192,16-96,32-48,48-32,96-16,192-8,1536-1},
	xticklabels={$1{\times}1536$,$8{\times}192$,$16{\times}96$,$32{\times}48$,$48{\times}32$,$96{\times}16$,$192{\times}8$,$1536{\times}1$},
	xtick=data,
	xticklabel style={xshift=.1cm,rotate=45,anchor=east},
	xlabel={Processor Grid}, 
	%legend style={draw=none,row sep=-0.1cm},
	%legend style={at={(1,.5)},anchor=west}
	legend style={at={(0.5,1.2)},anchor=north},
	legend columns=-1,
	legend style={draw=none, cells={align=left}, nodes={scale=0.8}},
]
	\setcolors
	\addplot table[x=pg, y expr=(\thisrow{mm}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=pg, y expr=(\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=pg, y expr=(\thisrow{gram}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=pg, y expr=(\thisrow{allgather}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=pg, y expr=(\thisrow{reducescatter}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=pg, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
	\legend{MM, \NLS, Gram, All-Gather, Reduce-Scatter, All-Reduce};
\end{axis}
\end{tikzpicture}
\caption{Tuning processor grid for \BPP\ on Sparse Synthetic data set with $p=1536$ and $k=50$.}
\label{fig:procsweep}
\end{figure}

\subsubsection{Scaling up to Very Large Sparse Datasets}\label{sec:webbase-2001}

In this section, we test \ParNMF\ by scaling up the problem size.
While we've used \emph{webbase-1M} in previous experiments, we consider \emph{webbase-2001} in this section as it is the largest sparse data in University of Florida Sparse Matrix Collection \cite{DH11}.
The former dataset has about 1 million nodes and 3 million edges, whereas the latter dataset has over 100 million nodes and  1 billion edges (see Section \ref{sec:datasets} for more details).
Not only is the size of the input matrix increased by two orders of magnitude (because of the increase in the number of edges), but also the size of the output matrices is increased by two orders of magnitude (because of the increase in the number of nodes).

In fact, with a low rank of $k=50$, the size of the output matrices dominates that of the input matrix: $\WW$ and $\HH$ together require a total of 88 GB, while $\AA$ (stored in compressed column format) is only 16 GB.
At this scale, because each node (consisting of 16 cores) of Rhea has 128 GB of memory, multiple nodes are required to store the input and output matrices with room for other intermediate values.
As mentioned in Section \ref{sec:new_memory}, \ParNMF\ requires considerably more temporary memory than necessary when the output matrices require more memory than the input matrix.
While we were not limited by this property for the other sparse matrices, the \emph{webbase-2001} matrix dimensions are so large that we need the memories of tens of nodes to run the algorithm.
Thus, we report results only for the largest number of processors in our experiments: 1536 processors (96 nodes).
The extra temporary memory used by \ParNMF\ is a latency-minimizing optimization; the algorithm can be updated to avoid this extra memory cost using a blocked matrix multiplication algorithm.
The extra memory can be reduced to a negligible amount at the expense of more messages between processors and synchronizations across the parallel machine.
%We have not yet implemented this update.

%In the real-world, it is difficult to acquire very large dense datasets. Most of the very large dense 
%datasets will be either be image/video data or collected out of very large scientific experiments. It is 
%common in the DM/ML community to have very large sparse dataset. In this section, to benefit the Data 
%mining and machine learning community, we report the scalability of \ParNMF\ on very directed sparse 
%webgraph over 118 million nodes and nearly 1 billion edges.  In Section \ref{sec:datasets},  we 
%discussed the details of this {\em webbase-2001} dataset.

%It can be observed from our \ParNMF\ that we hold the dense matrices $\WW, \HH, (\WW_i)_j, 
%(\HH_j)_i, \WW^T\AA$ and $\AA \HH^T$ in memory for computation. As discussed in Section 
%{\ref{sec:memory}}, the estimated per process memory footprint of all the matrices together is $O(xx)$. In the 
%case of {\em Webbase-2001},  matrix of size 118,000,000 with low rank $k=$50 over 864 process, every 
%process  required xx GB and approximately 2GB for input sparse matrix. \ramki{Grey can we discuss to fill this up. I wanted this to be consistent with the table and the section.} We were running 16 process on every node, the estimated memory per-process along with temporary allocations,  the experiment couldn't run in 54 nodes with each node having 128GB. Hence, we are reporting the wall clock time for different tasks and the error plot for {\em Webbase-2001} over 96 nodes and 1536 processors in Figure \ref{fig:webbase2001}.

We present results for \emph{webbase-2001} in Figure \ref{fig:webbase2001}.
The average per-iteration timing results are consistent with the observations from other synthetic and real world sparse datasets as discussed in Section \ref{sec:scaling}, though the raw times are about 2 orders of magnitude larger, as expected. 
In the case of the error plot, as observed in other experiments, \BPP\ achieves smaller error (by 1\%) than other algorithms after converging; however \MU\ and \HALS\ initially outperform \BPP. 
We also see that \MU\ outperforms \HALS\ in the first 30 iterations. 
At the 30th iteration, the error for \HALS\ is still improving at the third decimal, whereas \MU's is improving at the fourth decimal. 
We suspect that over a greater number of iterations the error of \HALS\ could become smaller than that of \MU, which would be more consistent with other datasets.

\begin{figure}
\centering
\captionsetup[subfigure]{justification=centering}
\begin{subfigure}{0.40 \columnwidth}
\renewcommand{\datafile}{data/webbase118m_pgf.dat}
%\centering
\begin{tikzpicture}
\begin{axis}[
	ybar stacked,
	reverse legend,
	bar width=16pt,
	width=1.6in, height=1.8in,
	ylabel={Time (seconds)}, 
	y label style={yshift=-.5cm, xshift=-.5cm},
	x label style={yshift=-.5cm},
	ymin=0,
	symbolic x coords={1536-0,1536-1,1536-2},
	xticklabels={MU,HALS,ABPP},
	xtick=data,
	%xticklabel style={xshift=.15cm,rotate=45,anchor=east},
	%xlabel={NMF Algorithms}, 
	%legend style={draw=none,row sep=-0.1cm},
	%legend style={at={(1,.5)},anchor=west}
	legend style={at={(0.5,1.3)},anchor=north},
	legend columns=3,
	legend style={draw=none, cells={align=left}, nodes={scale=0.5}}
]
	\setcolors
	\addplot table[x=algo, y expr=(\thisrow{mm}/30)] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{nnls}/30)] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{gram}/30)] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{allgather}/30)] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{reducescatter}/30)] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{allreduce}/30)] {\datafile};
	\legend{MM, \NLS, Gram, All-Gather, Reduce-Scatter, All-Reduce};
\end{axis}
\end{tikzpicture}
\subcaption{Time}
\end{subfigure}
~
\begin{subfigure}{0.45 \columnwidth}
\begin{tikzpicture}
\renewcommand{\run}{1}
\legendtrue
\ylabeltrue
\widertrue
\renewcommand{\datafile}{data/webbase118m-err-time.dat}
\relerrplot
\widerfalse
\end{tikzpicture}
\subcaption{Error}
\end{subfigure}
\caption{NMF comparison on \emph{webbase-2001} for $k{=}50$ on 1536 processors.}
\label{fig:webbase2001}
\end{figure}

\subsection{Interpretation of Results}

We present results from two of the real world datasets in the Supplemental Material. 
The first example shows background separation of the \emph{video} data, and the second example shows topic modeling output on the \emph{stack exchange} text dataset. The details of these datasets are presented in Section \ref{sec:datasets}.

While the literature covers more detail about fine tuning NMF and different NMF variants for higher quality results on these two tasks, our main focus is to show how quickly we can produce baseline NMF solutions. 
In Figure 1 of the Supplemental Material, we can see the background is removed and the moving objects (e.g., cars) are visible. 
Similarly, Table 1 of Supplemental Material shows that the NMF solution discriminates among topics and and finds coherent keywords for each topic.

%\subsubsection{Moving Object Detection of Surveillance Video Data}
%
%As explained in the Section 6.1.1, we processed 12 minutes video that is captured from a 
%busy junction in Georgia Tech to separate the background and moving objects from this video. 
%In Figure \ref{fig:videoresults}, we present some sample frames to compare the input image with the separated background and moving objects.
%The background are the results of the low rank approximation 
%$\hat{\AA}=\WW\HH$ output yielded from our \ParNMF\ algorithm and the moving objects are given by $\AA-\hat{\AA}$. 
%We can clearly see the background remains static and the moving objects (e.g., cars) are visible. 
%
%\subsubsection{Topic Modeling of Stack Exchange Data}
%We ran 
%our \ParNMF\ algorithm on this dataset, which has nearly 12 million questions from the Stack Overflow site (under Stack Exchange) to produce 50 topics. 
%The matrix $\WW$ can be interpreted as {\em vocabulary-topic} distribution and the 
%$\HH$ as {\em topic-document} distribution.  We took the top 5 words for each of the 50 topics and 
%present them in Table \ref{tab:stackexchangetopics}. Typically a good topic generation satisfies 
%properties such as (a) finding discriminative rather than common words -- capturing words that can provide 
%some information; (b) finding different topics -- the similarity between different topics should be low; 
%(c) coherence - all the words that belong to one topic should be coherent.  There are some topic quality 
%metrics \cite{NLGB2010} that capture the usefulness of topic generation algorithm.  We can  see
% NMF generated generally high-quality and coherent topics. 
% Also, each of the topics are from different domains such as 
% databases, C/C++ programming, Java programming, and web technologies like PHP and HTML.

### 2018-06-20 latex/toms/qualityplots.tex
seeding paper

\pgfplotsset{
    discard if not/.style 2 args={
        x filter/.code={
            \edef\tempa{\thisrow{#1}}
            \edef\tempb{#2}
            \ifx\tempa\tempb
            \else
                \def\pgfmathresult{inf}
            \fi
        }
    }
}

\newcommand{\ksweeperrplot}{
\addplot [ultra thick, green, discard if not={algo}{0}] table [x=k, y=relerr] {\datafile};
\addplot [ultra thick, red, discard if not={algo}{1}] table [x=k, y=relerr] {\datafile};
\addplot [ultra thick, blue, discard if not={algo}{2}] table [x=k, y=relerr] {\datafile};
\legend{MU, HALS, BPP}
}
\newcommand{\ksweeperroptions}{
xlabel={Low Rank ($k$)},
ylabel={Rel. Error}
}

\newcommand{\itererrplot}{
\addplot [ultra thick, green, discard if not={algo}{0}] table [x=iter, y=relerr] {\datafile};
\addplot [ultra thick, red, discard if not={algo}{1}] table [x=iter, y=relerr] {\datafile};
\addplot [ultra thick, blue, discard if not={algo}{2}] table [x=iter, y=relerr] {\datafile};
\legend{MU, HALS, BPP}
}
\newcommand{\itererroptions}{
xlabel={Iteration},
ylabel={Rel. Error}
}

\begin{figure}[htbp]
\begin{center}
\begin{tikzpicture}
\renewcommand{\datafile}{data/densesyn_ksweep.dat}
\begin{axis}[\ksweeperroptions]
%\ksweeperrplot
\end{axis}
\end{tikzpicture}
\caption{Dense Synthetic - ksweep}
\label{fig:densesynksweeperr}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\begin{tikzpicture}
\renewcommand{\datafile}{data/sparsesyn_ksweep.dat}
\begin{axis}[\ksweeperroptions]
%\ksweeperrplot
\end{axis}
\end{tikzpicture}
\caption{Sparse Synthetic - ksweep}
\label{fig:sparsesynksweeperr}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\begin{tikzpicture}
\renewcommand{\datafile}{data/densesyn_iter.dat}
\begin{axis}[\itererroptions]
%\itererrplot
\end{axis}
\end{tikzpicture}
\caption{Dense Synthetic - Iteration vs Rel.Err}
\label{fig:densesynitererr}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\begin{tikzpicture}
\renewcommand{\datafile}{data/sparsesyn_iter.dat}
\begin{axis}[\itererroptions]
%\itererrplot
\end{axis}
\end{tikzpicture}
\caption{Sparse Synthetic - Iteration vs Rel.Err}
\label{fig:sparsesynitererr}
\end{center}
\end{figure}

### 2018-06-20 common/utils.hpp
Using steady clock timer from standard chrono library as clock instead of OpenMP clock

@@ -5,10 +5,11 @@
 #include <assert.h>
 #include <omp.h>
 #include <stdio.h>
+#include <chrono>
 #include <ctime>
-#include <vector>
 #include <stack>
 #include <typeinfo>
+#include <vector>
 #include "common/utils.h"
 #ifdef MKL_FOUND
 #include <mkl.h>
@@ -33,14 +34,16 @@ static ULONG powersof10[16] = {1,
                                100000000000000,
                                1000000000000000};
 
-static std::stack<clock_t> tictoc_stack;
+static std::stack<std::chrono::steady_clock::time_point> tictoc_stack;
 static std::stack<double> tictoc_stack_omp_clock;
 
-inline void tic() { tictoc_stack.push(clock()); }
+inline void tic() { tictoc_stack.push(std::chrono::steady_clock::now()); }
 
 inline double toc() {
-  double rc =
-      (static_cast<double>(clock() - tictoc_stack.top())) / CLOCKS_PER_SEC;
+  std::chrono::duration<double> time_span =
+      std::chrono::duration_cast<std::chrono::duration<double>>(
+          std::chrono::steady_clock::now() - tictoc_stack.top());
+  double rc = time_span.count();
   tictoc_stack.pop();
   return rc;
 }
@@ -246,7 +249,7 @@ void ARMAMKLSCSCMM(const SP_MAT &mklMat, char transa, const MAT &Bt,
   MKL_INT *pntrb = static_cast<MKL_INT *>(mklMat.col_ptrs);
   MKL_INT *pntre = pntrb + 1;
   mkl_dcscmm(&transa, &m, &n, &k, &alpha, matdescra, mklMat.values,
-             static_cast<MKL_INT *>mklMat.row_indices, pntrb, pntre,
+             static_cast<MKL_INT *> mklMat.row_indices, pntrb, pntre,
              static_cast<double *>(Bt.memptr()), &ldb, &beta, Ct, &ldc);
 }
 #endif

### 2018-06-20 latex/toms/abstract.tex

\begin{abstract}
Non-negative matrix factorization (NMF) is the problem of determining two non-negative low rank factors $\WW$ and $\HH$, for the given input matrix $\AA$, such that $\AA \approx \WW \HH$.  
NMF is a useful tool for many applications in different domains such as topic modeling in text mining, background separation in video analysis, and community detection in social networks.  
Despite its popularity in the data mining community, there is a lack of efficient parallel algorithms to solve the problem for big data sets.  
The main contribution of this work is a new, high-performance parallel computational framework for a broad class of NMF algorithms that iteratively solves alternating non-negative least squares (NLS) subproblems for $\WW$ and $\HH$. 
It maintains the data and factor matrices in memory (distributed across processors), uses MPI for interprocessor communication, and, in the dense case, provably minimizes communication costs (under mild assumptions).  
The framework is flexible and able to leverage a variety of NMF and NLS algorithms, including Multiplicative Update, Hierarchical Alternating Least Squares, and Block Principal Pivoting.
Our implementation allows us to benchmark and compare different algorithms on massive dense and sparse data matrices of size that spans from few hundreds of millions to billions.  
We demonstrate the scalability of our algorithm and compare it with baseline implementations, showing significant performance improvements. 
The code and the datasets used for conducting the experiments are available online.
\end{abstract}

### 2018-06-20 latex/toms/aunmf.tex

\section{Alternating-Updating NMF Algorithms}
\label{sec:aunmf}

We define Alternating-Updating NMF algorithms as those that (1) alternate between updating $\WW$ for a given $\HH$ and updating $\HH$ for a given $\WW$ and (2) use the Gram matrix associated with the fixed factor matrix and the product of the input data matrix $\AA$ with the fixed factor matrix.
We show the structure of the framework in Algorithm \ref{alg:aunmf}.

%Different algorithms determine $\WW$ and $\HH$ by partitioning these matrices in different ways and can be explained under Block Coordinate Descent (BCD) framework. 
%Generally, these matrices are determined as (a) 2-Blocks of entire matrix $\WW$ and $\HH$; (b) $2k$ blocks such as $[\ww^1, \ww^2, \cdots , \ww^m] \in \Rnplus{k}$ and $[\hh_1^T, \hh_2^T, \cdots , \hh_n^T] \in \Rnplus{k}$ \footnote{For convenience, $\M{x}^i$ is the $i$th column vector of matrix $\M{X}$ and $\M{x}_i$ is the $i$th row vector} and (c) $(m+n)k$ scalar blocks of $w_{ik} \in \WW$ and $h_{kj} \in \HH$.

\begin{algorithm}
\caption{$[\WW,\HH] = \text{AU-NMF}(A,k)$}
\label{alg:aunmf}
\begin{algorithmic}[1]
\Require $\AA$ is an $m\times n$ matrix, $k$ is the approximation rank
\State Initialize $\HH$ with a non-negative matrix in $\Rn{n\times k}_+$.
\While{stopping criteria not satisfied} \label{algo:nmfloop}
  \State Update $\WW$ using $\HH \HH^T$ and $\AA \HH^T$
  %\State Update $\WW$ as $\Argmin{\tilde \WW\geq 0}\NormBr{\AA-\tilde\WW\HH}_F$
  %\State Update all blocks defined over $\WW$
  \label{line:aunmf:W}
  \State Update $\HH$ using $\WW^T\WW$ and $\WW^T \AA$
  %\State Update $\HH$ as $\Argmin{\tilde\HH\geq 0}\NormBr{\AA-\WW\tilde\HH}_F$
  %\State Update all blocks defined over $\HH$
  \label{line:aunmf:H}
\EndWhile
\end{algorithmic}
\end{algorithm}

The specifics of lines \ref{line:aunmf:W} and  \ref{line:aunmf:H} depend on the NMF algorithm, and we refer to the computation associated with these lines as the Local Update Computations (\LUC), as they will not affect the parallelization schemes we define in Section \ref{sec:parNMF}.
Because these computations are performed locally, we use a function $F(m,n,k)$ to denote the number of flops required for each algorithm's \LUC\ (and we do not consider communication costs).
Note that $F(m,n,k)$ does not include the cost of computing $\HH\HH^T$, $\WW^T\WW$, $\WW^T\AA$, or $\AA\HH^T$.

We note that AU-NMF is very similar to a two-block, block coordinate descent (BCD) framework, but it has a key difference.
In the BCD framework where the two blocks are
the unknown factors $\WW$ and $\HH$, 
we \emph{solve} the following subproblems,
which have a unique solution for a full rank $\HH$ and $\WW$: 
\SplitN{\label{eqn:two block}} {
\WW &\leftarrow \Argmin{\tilde \WW\geq 0}\NormBr{\AA-\tilde\WW\HH}_F,\\
\HH &\leftarrow  \Argmin{\tilde\HH\geq 0}\NormBr{\AA-\WW\tilde\HH}_F.
}
Since each subproblem involves nonnegative least squares,
this two-block BCD method is also called
the Alternating Non-negative Least Squares (ANLS) method \cite{kim2013nonnegative}.
For example, Block Principal Pivoting (\BPP), discussed more in detail at Section \ref{sec:BPP}, is 
one algorithm that solves these NLS subproblems.
In the context of the AU-NMF algorithm,
 an ANLS method {\em maximally} reduces 
the overall NMF objective function value 
by finding the optimal solution for
 given $\HH$ and $\WW$ in lines \ref{line:aunmf:W} 
and \ref{line:aunmf:H} respectively.

There are other popular NMF algorithms 
that update the factor matrices alternatively
without maximally reducing the objective function value each time,
in the same sense as in ANLS. 
These updates do not necessarily solve each of the subproblems \eqref{eqn:two block} to optimality but simply improve the overall objective function \eqref{eqn:original NMF}.  
Such methods include Multiplicative Update (\MU) \cite{seung2001algorithms} and Hierarchical Alternating Least Squares (\HALS) \cite{cichocki2009nonnegative}, which was also proposed as Rank-one Residual Iteration (RRI) \cite{Ho2008}.
To show how these methods can fit into the AU-NMF framework, we discuss them in more detail in Sections \ref{sec:MU} and \ref{sec:HALS}.

The convergence properties of these different algorithms are discussed in detail by Kim, He and Park \cite{kim2013nonnegative}. 
We emphasize here that both \MU\ and \HALS\ require computing Gram matrices and matrix products of the input matrix and each factor matrix.
Therefore, if the update ordering follows the convention of updating all of $\WW$ followed by all of $\HH$, both methods fit into the AU-NMF framework. 
We note that both \MU\ and \HALS\ are defined for more general update orders, but for our purposes we constrain them to be AU-NMF algorithms.

While we focus on three NMF algorithms in this paper, we highlight that our framework is extensible to other NMF algorithms, including those based on Alternating Direction Method of Multipliers (ADMM) \cite{SF14}, Nesterov-based methods \cite{GTLY12}, or any other method that fits the framework of Algorithm \ref{alg:aunmf}.

%Here, we have explained the update equations of the NMF algorithms ANLS/BPP, HALS and MU. In all these three algorithms, the update equations uses $\AA\HH^T$ and $\HH\HH^T$ for line \ref{line:anls:W} and $\WW^T\AA$ and $\WW^T\WW$ for line \ref{line:anls:H}.

%As in the Figure \ref{fig:2blocks}, the ANLS/BPP \cite{kim2011fast} is an example of 2-Blocks BCD that determines $\WW$ and $\HH$ as matrix blocks by solving 
%the following equations using active set method iteratively until a stopping criteria is satisfied:

%\begin{figure*}
%\subfloat[][ANLS/BPP 2-Blocks BCD] {\includegraphics[height=1.5in, width=2.5in]{fig/bcd2blocks}\label{fig:2blocks}}
%\subfloat[][HALS 2k-Blocks BCD] {\includegraphics[height=1.5in, width=2.5in]{fig/bcd2kblocks}\label{fig:2kblocks}}
%\subfloat[][MU (m+n)k Scalar Blocks BCD] {\includegraphics[height=1.5in, width=2.5in]{fig/bcdscalarblocks}\label{fig:bcdscalarblocks}}
%\end{figure*}

\subsection{Multiplicative Update (\MU)}
\label{sec:MU}

In the case of \MU\ \cite{seung2001algorithms}, individual entries of $\WW$ and $\HH$ are updated with all other entries fixed.
In this case, the update rules are 
\SplitN{\label{eqn:muupdate}} {
w_{ij} &\leftarrow w_{ij} \frac{(\AA \HH^T)_{ij}}{(\WW \HH \HH^T)_{ij}}, \text{ and }\\
h_{ij} &\leftarrow  h_{ij} \frac{(\WW^T \AA)_{ij}}{(\WW^T \WW \HH)_{ij}}.
} 
Instead of performing these $(m+n)k$ in an arbitrary order, if all of $\WW$ is updated before $\HH$ (or vice-versa), this method also follows the AU-NMF framework.
After computing the Gram matrices $\HH\HH^T$ and $\WW^T \WW$ and the products $\AA\HH^T$ and $\WW^T\AA$, the extra cost of computing $\WW (\HH\HH^T)$ and $(\WW^T\WW)\HH$ is $F(m,n,k)=2(m+n)k^2$ flops to perform updates for all entries of $\WW$ and $\HH$, as the other elementwise operations affect only lower-order terms.
Thus, when \MU\ is used, lines \ref{line:aunmf:W} and \ref{line:aunmf:H} in Algorithm \ref{alg:aunmf} -- and functions UpdateW and UpdateH in Algorithms \ref{alg:naive} and \ref{alg:2D} -- implement the expressions in \eqref{eqn:muupdate}, given the previously computed matrices.

\subsection{Hierarchical Alternating Least Squares (\HALS)}
\label{sec:HALS}

In the case of \HALS\ \cite{cichocki2009nonnegative,CA2009}, updates are performed on individual columns of $\WW$ and rows of $\HH$ with all other entries in the factor matrices fixed.
This approach is a BCD method with $2k$ blocks, set to minimize the function
\begin{equation}
f(\ww^1,\cdots,\ww^k,\hh_1,\cdots,\hh_k)=\lt\|\AA-\sum_{i=1}^k \ww^i \hh_i \rt\|_F,\label{eq:hals_obj}
\end{equation}
where $\ww^i$ is the $i$th column of $\WW$ and $\hh_i$ is the $i$th row of $\HH$.
The update rules \cite[Algorithm 2]{CA2009} can be written in closed form:
\SplitN{\label{eqn:halsupdate}} {
\ww^i &\leftarrow \lt[ \ww^i + (\AA\HH^T)^i - \WW (\HH \HH^T)^i \rt]_+ \\
\ww^i &\leftarrow \frac{\ww^i}{\|\ww^i\|}, \text{ and } \\
\hh_i &\leftarrow \lt[ \hh_i + (\WW^T\AA)_i - (\WW^T \WW)_i\HH \rt]_+.
}

%\grey{We need to explain this normalization, citing a HALS paper or something similar.}
Note that the columns of $\WW$ and rows of $\HH$ are updated in order, so that the most up-to-date values are always used, and these $2k$ updates can be done in an arbitrary order.  However, if all the $\WW$ updates are done before $\HH$ (or vice-versa),  the method falls into the AU-NMF framework.
After computing the matrices $\HH\HH^T$, $\AA\HH^T$, $\WW^T\WW$, and $\WW^T\AA$, the extra computation is $F(m,n,k)=2(m+n)k^2$ flops for updating both $\WW$ and $\HH$.

Thus, when \HALS\ is used, lines \ref{line:aunmf:W} and \ref{line:aunmf:H} in Algorithm \ref{alg:aunmf} -- and functions UpdateW and UpdateH in Algorithms \ref{alg:naive} and \ref{alg:2D} -- implement the expressions in \eqref{eqn:halsupdate}, given the previously computed matrices.

%Note that the normalization of columns of $\WW$ is unique to $\; this does introduce extra 
%In practice, we also found without such normalization, the error was oscillating and it promised the monotonic error reduction to \HALS\ algorithm.

%According to \cite{kim2013nonnegative}, the above update function of MU does not find the optimal blocks $w_{ij}$ and discuss a different optimal update for scalar blocks function that is similar to HALS.

%In general, many of the NMF algorithms can be explained using the BCD framework. However, specifically it can be observed in all the above three different cases of NMF algorithms explained above, the blocks of the matrix $\WW$ is updated first followed by the updates to matrix $\HH$ and we always use the most recent blocks of $\WW$ and $\HH$ matrix. In this paper, we define iteration as one pass of updating all the blocks of low rank factors $\WW$ and $\HH$. That is., in the case of 2-Blocks, updating entire $\WW$, $\HH$; for $2k$-blocks finding all the vectors of $\WW$, $\HH$; and all the elements of $\WW$, $\HH$ for scalar blocks BCD.

\subsection{Alternating Nonnegative Least Squares with Block Principal Pivoting}
\label{sec:BPP}

%In this paper, we focus on and use the BPP method \cite{kim2011fast} to solve the NLS problem, as it is the fastest algorithm (in terms of number of iterations). 
%As argued in Section \ref{sec:aunmf}, we note that many NMF algorithms, including MU and HALS, can be used within our parallel frameworks (Algorithms \ref{alg:naive} and \ref{alg:2D}).

Block Principal Pivoting (BPP) is an active-set-like method for solving the NLS subproblems in Eq. \eqref{eqn:two block}.
The main subroutine of BPP is the single right-hand side NLS problem
\SplitN{\label{eqn:single NLS}}{
\min_{\xx\geq 0} \|\CC\xx-\mathbf{b}\|_2.
}

The Karush-Kuhn-Tucker (KKT) optimality conditions for  Eq.~\eqref{eqn:single NLS} are as follows
\SplitS{\label{eqn:KKT}}{
\yy &= \CC^T \CC \xx - \CC^T \mathbf{b} \\
\xx,\yy &\geq 0\\
%\xx &\geq 0\\
x_i y_i & = 0 \;\; \forall i .
}
The KKT conditions \eqref{eqn:KKT} states that at optimality, the support sets (i.e., the non-zero elements)
of $\xx$ and $\yy$ are complementary to each other. Therefore, Eq.~\eqref{eqn:KKT} is an instance of
the \emph{Linear Complementarity Problem} (LCP) which arises frequently in quadratic programming.
When $k\ll\min(m,n)$, active-set and active-set-like methods are very suitable because most
computations involve matrices of sizes $m\times k, n\times k$, and $k\times k$ which are
small and easy to handle.

If we knew which indices correspond to nonzero values in the optimal solution, then computing the solution is an unconstrained least squares problem on these indices.
In the optimal solution, call the set of indices $i$ such that $x_i=0$ the active set, and let the remaining indices be the passive set. The BPP algorithm works to find this final active set and passive set. 
%Since the above problem is convex, the correct partition of the optimal solution will satisfy the KKT condition (Eq.~\eqref{eqn:KKT}).  
It greedily swaps indices between the intermediate active and passive sets until finding a partition that satisfies the KKT condition. In the partition of the optimal solution, the values of the indices that belong to the active set will take zero. The values of the indices that belong to the passive set are determined by solving the unconstrained least squares problem restricted to the passive set. Kim, He and Park \cite{kim2011fast}, discuss the BPP algorithm in further detail. 
We use the notation
$$\XX \gets \text{SolveBPP}(\CC^T\CC,\CC^T\BB)$$
to define the (local) function for using BPP to solve Eq.~\eqref{eqn:single NLS} for every column of $\XX$.
We define $C_\text{BPP}(k,c)$ as the cost of \text{SolveBPP}, given the $k\times k$ matrix $\CC^T\CC$ and $k\times c$ matrix $\CC^T\BB$. 
\text{SolveBPP} mainly involves solving least squares problems over the intermediate passive sets. 
%and has a worst case of solving $k$ least squares problems. 
Our implementation uses the normal equations to solve the unconstrained least squares problems because the normal equations matrices have been pre-computed in order to check the KKT condition.
However, more numerically stable methods such as QR decomposition can also be used.

Thus, when \BPP\ is used, lines \ref{line:aunmf:W} and \ref{line:aunmf:H} in Algorithm \ref{alg:aunmf} -- and functions UpdateW and UpdateH in Algorithms \ref{alg:naive} and \ref{alg:2D} -- correspond to calls to SolveBPP.
The number of flops involved in SolveBPP is not a closed form expression; in this case $F(m,n,k)=C_\text{BPP}(k,m)+ C_\text{BPP}(k,n)$.

### 2018-06-20 latex/toms/bio.tex

\vspace{-0.5in}
\begin{IEEEbiography}[{\includegraphics[width=1in,height =2.25in,clip,keepaspectratio]{fig/ramki}}]{Ramakrishnan Kannan}
is a Computational Data Scientist in 
Oak Ridge National Laboratory.
% focusing on large scale data mining and machine learning algorithms on HPC systems and modern architectures with applications from scientific domain and many different internet services. 
He received his Ph.D in Computer Science from College of Computing, 
Georgia Institute of Technology advised by Prof. Haesun Park. 
He worked on Data Analytics group at IBM TJ Watson Research Center and 
was an IBM Master Inventor. He has M.Sc (Engg) from Indian Institute of Science.  
\end{IEEEbiography}

\vspace{-0.5in}
\begin{IEEEbiography}[{\includegraphics[width=1in,height =1.25in,clip,keepaspectratio]{fig/ballard}}]{Grey Ballard}
Grey Ballard is an Assistant Professor in the Computer Science Department at Wake Forest University. After receiving his PhD in computer science from the University of California Berkeley in 2013, he was a Truman Fellow at Sandia National Laboratories in Livermore, CA. 
He received his BS in math and computer science at Wake Forest in 2006 and his MA in math at Wake Forest in 2008.
%His research interests include numerical linear algebra, high performance computing, and computational science, particularly in developing algorithmic ideas that translate to improved implementations and more efficient software. 
His work has been recognized with the SIAM Linear Algebra Prize and three conference best paper awards, at SPAA, IPDPS, and ICDM.
%; the C.V. Ramamoorthy Distinguished Research Award at UC Berkeley; and the ACM Doctoral Dissertation Award - Honorable Mention.
\end{IEEEbiography}
\vspace{-0.5in}

\begin{IEEEbiography}[{\includegraphics[width=1in,height =1.25in,clip,keepaspectratio]{fig/hpark}}]{Haesun Park}
is a IEEE and SIAM Fellow and Professor in the 
School of Computational Science and Engineering, 
Georgia Institute of Technology. 
%Her research areas include numerical algorithms, data analysis, visual analytics, text mining, and parallel computing.
She has played major leadership  roles as the Executive Director
of the Center for Data Analytics, Georgia Tech, general chair for the SIAM 
Conference on Data Mining, and editorial board member of SIAM
and IEEE journals. She received a  Ph.D. in Computer Science 
from Cornell University in 1987 and B.S. in Mathematics from Seoul National 
University with the University President's Medal.
\end{IEEEbiography}

### 2018-06-20 latex/toms/differences.tex

\documentclass{article}

\usepackage{url}

\begin{document}
\title{Differences from our PPoPP Conference Paper \cite{KBP16}}
\author{Ramakrishnan Kannan, Grey Ballard, Haesun Park}
\maketitle

The main focus of our PPoPP'16 conference paper \cite{KBP16} is demonstrating and understanding scalability of an NMF algorithm using Alternating Non-negative Least Squares based on Block Principal Pivoting (ANLS/BPP) \cite{kim2011fast} on MPI-based supercomputing clusters. We compared our scalable implementation against a Naive ANLS/BPP parallelization scheme explained by Fairbanks, Kannan, Park and Bader \cite{Fairbanks2015}. We conducted our experiments on ``Edison'' -- a Cray XC30 supercomputer, at the National Energy Research Scientific Computing Center.

This submission is a major revision specifically targeted for the Data Mining and Machine Learning (DM/ML) community. We highlight the principal differences below.

\subsection*{MPI-FAUN Framework:}
The new contributions of this paper include (1) implementing a software framework called MPI-FAUN to compare ANLS/BPP with alternative NMF algorithms for large scale data sets, (2) benchmarking on a data analysis cluster and scaling up to over 1500 processors, and (3) providing an interpretation of results for real-world data sets. 
Our main contribution is a new, high-performance parallel computational framework for a broad class of NMF algorithms. The framework is efficient, scalable, flexible, and demonstrated to be effective for large-scale dense and sparse matrices. Based on our survey and knowledge, we are the fastest NMF implementation available in the literature.

\subsection*{Technical Details:}
We have completely rewritten the entire paper targeting the DM/ML community. Given the increased space of a journal submission, we provide new technical details including figures, diagrams, and experiments to appreciate the entire content and reproduce and extend our research.

\subsection*{Supported Algorithms:}
In our earlier work, we implemented only ANLS/BPP.  In this paper, we have implemented a framework to handle a wider range of NMF algorithms that include Multiplicative Update (MU) and Hierarchical Alternating Least Squares (HALS), with which we experiment here. The end users can run these distributed algorithms over command line.

\subsection*{Open Source Code:}
We have open-sourced our entire code through \url{https://github.com/ramkikannan/nmflibrary}. The users can not only try these algorithms off-the-shelf through command line interfaces but also extend their algorithms over our framework.

\subsection*{Commodity Cluster:}
Unlike our earlier paper that was run on Cray supercomputer, in this paper we specifically targeted running on experiments on a commodity MPI linux cluster, the ``Rhea'' system at Oak Ridge Leadership Computing Facility, which is designed for data analysis. It is common in DM/ML community from both academic and enterprise DM/ML to set up their own experimental cluster using their lab machines. Hence, we conducted experiment using commodity cluster with open source software such as armadillo, gcc, and OpenMPI. Unlike our earlier experiment with only 25 nodes, the results presented in this paper were conducted on nearly 100 nodes.

\subsection*{Experiment and Datasets:}
We have added a broader range of very large dense and sparse real world datasets. We have run experiments on the largest sparse data in the University of Florida Sparse Matrix Collection. Apart from substantiating our research by running our experiments on a larger cluster, we also motivate the readers by presenting the interpretation of the results.

%The entire document is prepared for DM/ML researchers from academic and enterprise community. We are eagerly waiting for the feedback, inputs and reviews from the reviewers.

\bibliographystyle{plain}
\bibliography{mybib}

\end{document}

### 2018-06-20 latex/toms/introduction.tex

\section{Introduction}

Non-negative Matrix Factorization (NMF) is the problem of finding two low rank factors $\WW\in \Rnplus{m\times k}$ and $\HH\in \Rnplus{k \times n}$ for a given input matrix  $\AA\in \Rnplus{m\times n}$, such that $\AA \approx \WW \HH$.
Here, $\Rnplus{m\times n}$ denotes the set of $m \times n$ matrices with non-negative real values.
Formally, the NMF problem \cite{seung2001algorithms} can be defined as \SplitN{\label{eqn:original NMF}}{
\min_{\WW \geq 0,\HH \geq 0} & \|\AA-\WW\HH\|_F,
}
where $\|\M{X}\|_F=(\sum_{ij} x_{ij}^2)^{1/2}$ is the Frobenius norm.

NMF is widely used in data mining and machine learning as a dimension reduction and factor analysis method. 
It is a natural fit for many real world problems as the non-negativity is inherent in many representations of real-world data and
the resulting low rank factors are expected to have a natural interpretation. The applications of NMF range from text mining \cite{pauca2004text},  computer vision \cite{hoyer2004non,Fujimoto2014,BSJJZ2015}, and bioinformatics \cite{kim2007sparse} to blind source separation  \cite{cichocki2009nonnegative}, unsupervised clustering \cite{kuang2012symmetric,kuang2013symnmf}  and many other areas.
In the typical case, $k \ll \min(m,n)$; for problems today, $m$ and $n$ can be on the order of millions or more, and $k$ is on the order of few tens to thousands.

There is a vast literature
on algorithms for NMF and their convergence properties \cite{kim2013nonnegative}.   
The commonly adopted NMF algorithms are -- (i) Multiplicative Update (\MU) \cite{seung2001algorithms} (ii) Hierarchical Alternating Least Squares (\HALS) \cite{cichocki2009nonnegative,Ho2008} (iii) NMF based on Alternating Nonnegative Least Squares and Block Principal Pivoting (\BPP) \cite{kim2011fast}, and (iv) Stochastic Gradient Descent (SGD) Updates \cite{gemulla2011large}. 
Most of the algorithms in NMF literature are based on alternately optimizing each of the low rank factors $\WW$ and $\HH$ while keeping the other fixed, in which case each subproblem is a constrained convex optimization problem. 
Subproblems can then be solved using standard optimization techniques such as projected gradient or interior point method; a detailed survey for solving such problems can be found in \cite{xiong2013survey,kim2013nonnegative}. 
In this paper, our implementation uses either \BPP, \MU, or \HALS. 
But our parallel framework is extensible to other algorithms (e.g., \cite{XY2013,HSL2016}) as-is or with a few modifications, as long as they fit an alternating-updating framework (defined in Section \ref{sec:aunmf}).

With the advent of large scale internet data and interest in Big Data, researchers have started studying scalability of many foundational machine learning algorithms. 
To illustrate the dimension of matrices commonly used in the machine learning community, we present a few examples. 
Nowadays the adjacency matrix of a billion-node social network is common. 
In the matrix representation of a video data, every frame contains three matrices for each RGB color, which is reshaped into a column.  
Thus in the case of a 4K video, every frame will take approximately 27 million rows (4096 row pixels x 2196 column pixels x 3 colors). 
Similarly, the popular representation of documents in text mining is a bag-of-words matrix, where the rows are the dictionary and the columns are the documents (e.g., webpages). 
Each entry $A_{ij}$ in the bag-of-words matrix is generally the frequency count of the word $i$  in the document $j$. 
Typically with the explosion of the new terms in social media, the number of words spans to millions. 
To handle such high-dimensional matrices, it is important to study low-rank approximation methods in a data-distributed and parallel computing environment.

In this work, we present an efficient algorithm and implementation using tools from the field of High-Performance Computing (HPC).
We maintain data in memory (distributed across processors), take advantage of optimized libraries like BLAS and LAPACK for local computational routines, and use the Message Passing Interface (MPI) standard to organize interprocessor communication.
%In particular, we use an MPI/C++ implementation, which offers several technical advantages such as (1) the ability to leverage state-of-the-art hardware, (2) efficient networks for communication between nodes, and (3) the availability of numerically stable and efficient BLAS and LAPACK routines. 
Furthermore, the current hardware trend is that available parallelism (and therefore aggregate computational rate) is increasing much more quickly than improvements in network bandwidth and latency, which implies that the relative cost of communication (compared to computation) is increasing.
To address this challenge, we analyze algorithms in terms of both their computation and communication costs.
In particular, we prove in Section \ref{sec:parNMF} that in the case of dense input and under a mild assumption, our proposed algorithm minimizes the amount of data communicated between processors to within a constant factor of the lower bound.

We call our implementation MPI-FAUN, an MPI-based Framework for Alternating-Updating Nonnegative matrix factorization algorithms.
A key attribute of our framework is that the efficiency does not require a loss of generality of NMF algorithms.
Our central observation is that most NMF algorithms, in particular those that alternate between updating each factor matrix, consist of two main tasks: (a) performing matrix multiplications and (b) solving Non-negative Least Squares (NLS) subproblems, either approximately or exactly.
More importantly, NMF algorithms tend to perform the same matrix multiplications, differing only in how they solve NLS subproblems, and the matrix multiplications often dominate the running time of the algorithms.
Our framework is designed to perform the matrix multiplications efficiently and organize the data so that the NLS subproblems can be solved independently in parallel, leveraging any of a number of possible methods.
We explore the overall efficiency of the framework and compare three different NMF methods in Section \ref{sec:experiment}, performing convergence, scalability, and parameter-tuning experiments on over 1500 processors.

\begin{table}[htp]
\begin{center}
\begin{tabular}{|c|c|c|c|c|}
\hline 
Dataset & Type & Matrix size & NMF Time \\ \hline
Video & Dense & 1 Million x 13,824 &  5.73 seconds \\
Stack Exchange & Sparse & 627,047 x 12 Million &  67 seconds \\
Webbase-2001 & Sparse & 118 Million x 118 Million & 25 minutes \\ \hline
\end{tabular}
\end{center}
\caption{\ParNMF\ on large real-world datasets. Reported time is for 30 iterations on 1536 processors with a low rank of 50.}
\label{tab:teaser}
\end{table}%

With our framework, we are able to explore several large-scale synthetic and real-world data sets, some dense and some sparse. In Table \ref{tab:teaser}, we present the NMF computation wall clock time on some very large real world datasets. We describe the results of the computation in Section \ref{sec:experiment}, showing the range of application of NMF and the ability of our framework to scale to large data sets.

%For example, we apply NMF to a surveillance video data set of size 44 GB in order to isolate the moving objects from the background view; once loaded into memory, the most efficient algorithm requires only 5.73 seconds on 1536 processors.
%We also apply our techniques to sparse data, analyzing Stack Exchange posts (text data) to infer discussion topics; this data set is 44 GB of XML file which takes 6.1 GB of sparse matrix representation, and our NMF implementation requires 67 seconds on 1536 processors to obtain meaningful results 
%Similarly, our algorithm can perform graph clustering using NMF on a very large sparse graph with 118 million nodes and 1 billion edges and the sparse representation of this matrix takes 18GB on disk 
%in nearly 15 minutes using 1536 processors, 
%\grey{do we want to report that the output dense matrices required 88GB?}.
%\grey{break up into multiple sentences}

A preliminary version of this work has already appeared as a conference paper \cite{KBP16}.
While the focus of the previous work was parallel performance of \BPP\ (Alternating Nonnegative Least Squares and Block Principal Pivoting), the goal of this paper is to explore more data analytic questions.
In particular, the new contributions of this paper include (1) implementing a software framework to compare \BPP\ with \MU\ (Multiplicative Update) and \HALS\ (Hierarchical Alternating Least Squares) for large scale data sets, (2) benchmarking on a data analysis cluster and scaling up to over 1500 processors, and (3) providing an interpretation of results for real-world data sets.
We provide a detailed comparison with other related work, including MapReduce implementations of NMF, in Section \ref{sec:related}.

Our main contribution is a new, high-performance parallel computational framework for a broad class of NMF algorithms. 
The framework is efficient, scalable, flexible, and demonstrated to be effective for large-scale dense and sparse matrices.  
Based on our survey and knowledge, we are the fastest NMF implementation available in the literature.  
The code and the datasets used for conducting the experiments can be downloaded from \url{https://github.com/ramkikannan/nmflibrary}.

### 2018-06-20 latex/toms/plots.tex

% macros for plotting
\newcommand{\datafile}{}
\newcommand{\run}{}
\newcommand{\numiterations}{30}
\newcommand{\minvalue}{1}

% toggles whether or not to plot naive results (if not, the rows of the data file also need to be commented out)
\newif\ifnaive
% toggles ksweep vs scaling plot
\newif\ifksweep
% toggles legend
\newif\iflegend
% toggles y axis label
\newif\ifylabel
% toggles wider plot area
\newif\ifwider

\newcommand{\setcolors}{
\pgfplotsset{cycle list={
	red, fill=red \\ 
	blue, fill=blue \\ 
	green, fill=green \\ 
	red, pattern=crosshatch, pattern color=red \\
	blue, pattern=crosshatch, pattern color=blue \\
	green, pattern=crosshatch, pattern color=green \\
}};
}

% set options for grouped bar plot
\newcommand{\plotoptions}{
	ybar stacked,
	reverse legend,
	bar width=6pt,
	width=9cm, height=3.85cm,
	ylabel={Time (seconds)}, 
	y label style={yshift=-.5cm},
	ymin=0,
	\ifnaive 
		\ifksweep
			symbolic x coords={10-0,10-1,10-2,10-3,,20-0,20-1,20-2,20-3,,30-0,30-1,30-2,30-3,,40-0,40-1,40-2,40-3,,50-0,50-1,50-2,50-3},
		\else
			symbolic x coords={16-0,16-1,16-2,16-3,,96-0,96-1,96-2,96-3,,384-0,384-1,384-2,384-3,,864-0,864-1,864-2,864-3,,1536-0,1536-1,1536-2,1536-3},
		\fi
		xticklabels={MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive},
	\else  
		\ifksweep
			symbolic x coords={10-0,10-1,10-2,,,20-0,20-1,20-2,,,30-0,30-1,30-2,,,40-0,40-1,40-2,,,50-0,50-1,50-2,},
		\else
			symbolic x coords={16-0,16-1,16-2,,,96-0,96-1,96-2,,,384-0,384-1,384-2,,,864-0,864-1,864-2,,,1536-0,1536-1,1536-2,}, 
		\fi
		xticklabels={MU,HALS,ABPP,MU,HALS,ABPP,MU,HALS,ABPP,MU,HALS,ABPP,MU,HALS,ABPP},
	\fi
	xtick=data,
	xticklabel style={xshift=.15cm,rotate=75,anchor=east},
	\ifksweep
		xlabel={Low Rank ($k$)}, 
	\else
		xlabel={Number of Processes ($p$)},
	\fi
	xlabel style={yshift=-0.75cm},
	%legend style={at={(0.4,-0.9)},anchor=north},
	legend style={at={(0.5,1.3)},anchor=north},
	legend columns=-1,
	legend style={draw=none, cells={align=left}, nodes={scale=0.8}}
}

% grouped bar plot command
\newcommand{\makeplot}{
\begin{axis}[\plotoptions]
	\setcolors
	\addplot table[x=algo, y expr=(\thisrow{mm}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{gram}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{allgather}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{reducescatter}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=algo, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
	\iflegend
		\legend{MM, \NLS, Gram, All-Gather, Reduce-Scatter, All-Reduce};
	\fi
\end{axis}
}

% labels for bar groups (manually positioned)
\newcommand{\labels}{
%\node [align=center,text width=3cm] at (1.25cm, -.95cm)   {\ifksweep 10 \else 16 \fi};
%\node [align=center,text width=3cm] at (3cm, -0.95cm)   {\ifksweep 20 \else 96 \fi};
%\node [align=center,text width=3cm] at (4.6cm, -0.95cm)   {\ifksweep 30 \else 384 \fi};
%\node [align=center,text width=3cm] at (6.3cm, -0.95cm) {\ifksweep 40 \else 864 \fi};
%\node [align=center,text width=3cm] at (8.15cm, -0.95cm) {\ifksweep 50 \else 1536 \fi};
\node [align=center,text width=3cm] at (1cm, -1.15cm)   {\ifksweep 10 \else 16 \fi};
\node [align=center,text width=3cm] at (2.4cm, -1.15cm)   {\ifksweep 20 \else 96 \fi};
\node [align=center,text width=3cm] at (3.7cm, -1.15cm)   {\ifksweep 30 \else 384 \fi};
\node [align=center,text width=3cm] at (5.1cm, -1.15cm) {\ifksweep 40 \else 864 \fi};
\node [align=center,text width=3cm] at (6.4cm, -1.15cm) {\ifksweep 50 \else 1536 \fi};
}

% allows for filtering rows from dat file
\pgfplotsset{
    discard if not/.style 2 args={
        x filter/.append code={
            \edef\tempa{\thisrow{#1}}
            \edef\tempb{#2}
            \ifx\tempa\tempb
            \else
                \def\pgfmathresult{inf}
            \fi
        }
    }
}

% set options for relative error over time plot
\newcommand{\relerrplotoptions}{
	xlabel=Time (s),
	\ifylabel
		ylabel=Relative Error,
		ylabel style={yshift=-.3cm},
	\fi
	\ifwider
		width=1.8in, height=2in,
	\else
		width=1.35in, height=2in,
	\fi
	y tick label style={
        /pgf/number format/.cd,
            precision=4,
	},
	legend style={draw=none, cells={align=left}, nodes={scale=0.7}}
}

% plot relative error over time, filtering rows out appropriately
\newcommand{\relerrplot}{
\begin{axis}[\relerrplotoptions]
	\addplot+ [discard if not={algo}{MU},discard if not={runs}{\run},green,thick,mark options={solid}] table [x={totaltime}, y={relerr}] {\datafile};
	\addplot+ [discard if not={algo}{HALS},discard if not={runs}{\run},red,thick,mark options={solid},mark=square*] table [x={totaltime}, y={relerr}] {\datafile};
	\addplot+ [discard if not={algo}{BPP},discard if not={runs}{\run},blue,thick,mark options={solid},mark=triangle*] table [x={totaltime}, y={relerr}] {\datafile};
	\iflegend
		\legend{MU,HALS,ABPP}
	\fi
\end{axis}
}

### 2018-06-20 latex/toms/prelim.tex

\section{Preliminaries}

\subsection{Notation}
\label{sec:notations}

Table \ref{tab:notation} summarizes the notation we use throughout this paper.
We use \emph{upper case} letters for matrices and \emph{lower case} letters for vectors.
%For example, $\AA$ is a matrix and $\aa$ is a column vector and $\aa^T$ is a row vector. 
We use both subscripts and superscripts for sub-blocks of matrices. 
For example, $\AA_i$ is the $i$th row block of matrix $\AA$, and $\AA^i$ is the $i$th column block.
Likewise, $\aa_i$ is the $i$th row of $\AA$, and $\aa^i$ is the $i$th column.
We use $m$ and $n$ to denote the numbers of rows and columns of $\AA$, respectively, and we assume without loss of generality $m\geq n$ throughout.

\begin{table}%[htdp]
\begin{center}
\begin{tabular}{|l|l|}
\hline
$\AA$ & Input matrix \\
$\WW$ & Left low rank factor \\
$\HH$ & Right low rank factor \\
$m$ & Number of rows of input matrix \\
$n$ & Number of columns of input matrix \\
$k$ & Low rank \\
$\M{M}_i$ & $i$th row block of matrix $\M{M}$ \\
$\M{M}^i$ & $i$th column block of matrix $\M{M}$  \\
$\M{M}_{ij}$ & $(i,j)$th subblock of $\M{M}$ \\
$p$ & Number of parallel processes \\
$p_r$ & Number of rows in processor grid \\
$p_c$ & Number of columns in processor grid \\
\hline
\end{tabular}
\end{center}
\caption{Notation}
\label{tab:notation}
\end{table}%

\subsection{Communication model}
\label{sec:comm-model}

To analyze our algorithms, we use the $\alpha$-$\beta$-$\gamma$ model of distributed-memory parallel computation.
In this model, interprocessor communication occurs in the form of messages sent between two processors across a bidirectional link (we assume a fully connected network).
We model the cost of a message of size $n$ words as $\alpha+n\beta$, where $\alpha$ is the per-message latency cost and $\beta$ is the per-word bandwidth cost.
Each processor can compute floating point operations (flops) on data that resides in its local memory; $\gamma$ is the per-flop computation cost.
With this communication model, we can predict the performance of an algorithm in terms of the number of flops it performs as well as the number of words and messages it communicates.
For simplicity, we will ignore the possibilities of overlapping computation with communication in our analysis.
For more details on the $\alpha$-$\beta$-$\gamma$ model, see \cite{TRG05,CH+07}.

\subsection{MPI collectives}
\label{sec:collectives}

Point-to-point messages can be organized into collective communication operations that involve more than two processors.
MPI provides an interface to the most commonly used collectives like broadcast, reduce, and gather, as the algorithms for these collectives can be optimized for particular network topologies and processor characteristics.
For a concise description of the most common collectives, see \cite[Figure 1]{CH+07}.
The algorithms we consider use the all-gather, reduce-scatter, and all-reduce collectives, so we review them here, along with their costs.
Our analysis assumes optimal collective algorithms are used (see \cite{TRG05,CH+07}), though our implementation relies on the underlying MPI implementation.

At the start of an all-gather collective, each of $p$ processors owns data of size $n/p$. 
After the all-gather, each processor owns a copy of the entire data of size $n$. 
The cost of an all-gather is $\alpha\cdot \log p + \beta \cdot \frac{p-1}{p}n$.
%
At the start of a reduce-scatter collective, each processor owns data of size $n$.
After the reduce-scatter, each processor owns a subset of the sum over all data, which is of size $n/p$.
This single collective is a more efficient way of implementing a reduce followed by a scatter.
(Note that the reduction can be computed with other associative operators besides addition.)
The cost of an reduce-scatter is $\alpha\cdot \log p + (\beta+\gamma) \cdot \frac{p-1}{p}n$.
%
At the start of an all-reduce collective, each processor owns data of size $n$.
After the all-reduce, each processor owns a copy of the sum over all data, which is also of size $n$.
The cost of an all-reduce is $2\alpha\cdot \log p + (2\beta+\gamma) \cdot \frac{p-1}{p}n$.
%
Note that the costs of each of the collectives are zero when $p=1$.

### 2018-06-20 latex/toms/stackxchangetopics.tex

\begin {tabular}{|ccccc||>{\columncolor [gray]{.8}}c>{\columncolor [gray]{.8}}c>{\columncolor [gray]{.8}}c>{\columncolor [gray]{.8}}c>{\columncolor [gray]{.8}}c|}%
\toprule 
\multicolumn{5}{|c||}{Top Keywords from Topics 1-25} & \multicolumn{5}{|>{\columncolor [gray]{0.8}}c|}{Top Keywords from Topics 26-50} \\ 
word1&word2&word3&word4&word5&word1&word2&word3&word4&word5\\\midrule %
refer&undefin&const&key&compil&echo&type=text&php&form&result\\%
text&field&box&word&static&test&perform&fail&unit&result\\%
imag&src&descript&alt=ent&size&tabl&key&queri&databas&insert\\%
button&click&event&form&add&user&email&usernam&login&log\\%
creat&bean&add&databas&except&data&json&store&read&databas\\%
string&static&final&catch&url&page&load&content&url&link\\%
width&height&color&left&display&privat&static&final&import&float\\%
app&applic&servic&thread&work&row&column&date&cell&valu\\%
ipsum&lorem&dolor&sit&amet&line&import&command&print&recent\\%
node&list&root&err&element&var&map&marker&match&url\\%
0x00&0xff&byte&0x01&0xc0&server&connect&client&messag&request\\%
file&directori&read&open&upload&number&byte&size&print&input\\%
function&call&event&work&variabl&object&properti&json&instanc&list\\%
int&char&const&static&doubl&array&element&valu&key&index\\%
public&overrid&virtual&static&extend&main&thread&program&frame&cout\\%
return&param&result&def&boolean&type&field&properti&argument&resolv\\%
info&thread&start&map&servic&select&item&queri&join&list\\%
error&syntax&found&symbol&fail&sourc&target&except&java&fail\\%
set&properti&virtual&default&updat&instal&version&packag&err&default\\%
case&break&switch&default&cout&code&work&problem&chang&write\\%
method&call&except&static&todo&void&overrid&protect&catch&extend\\%
href&nofollow&src&link&work&true&requir&boolean&option&valid\\%
end&def&dim&begin&properti&find&project&import&warn&referenc\\%
debug&request&filter&match&found&view&control&item&overrid&posit\\%
fals&boolean&fix&bool&autoincr&null&default&key&int(11&primari\\\bottomrule %
\end {tabular}%

### 2018-06-20 latex/toms/supplement.tex

\documentclass[10pt,journal,compsoc, onecolumn]{IEEEtran}

\pdfoutput=1 
\usepackage[varg]{txfonts}
\let\iint=\relax
\let\iiint=\relax
\let\iiiint=\relax
\let\idotsint=\relax

\let\subcaption\relax

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{color}
\usepackage{ramkidefns}
%for displaying topic words
\usepackage{pgfplotstable}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{colortbl}

\newcommand{\ParNMF}{MPI-FAUN}

\begin{document}

\begin{center}
\huge
Supplement Material for MPI-FAUN
\end{center}

In this section, we present results from two of the real world datasets. 
The first example shows an image processing example of background separation and moving object detection in surveillance video data, and the second example shows topic modeling output on the \emph{stack exchange} text dataset. The details of these datasets are presented in Section 6.1.1. 
While the literature covers more detail about fine tuning NMF and different NMF variants for higher quality results on these two tasks  \cite{ZT2011,B2014,AGHKT2014,KJJCH2015}, our main focus is to show how quickly we can produce a baseline NMF output and its real world interpretation.

\section{Moving Object Detection of Surveillance Video Data}

As explained in the Section 6.1.1, we processed 12 minutes video that is captured from a 
busy junction in Georgia Tech to separate the background and moving objects from this video. 
In Figure \ref{fig:videoresults}, we present some sample frames to compare the input image with the separated background and moving objects.
The background are the results of the low rank approximation 
$\hat{\AA}=\WW\HH$ output yielded from our \ParNMF\ algorithm and the moving objects are given by $\AA-\hat{\AA}$. 
We can clearly see the static background is captured in the second column and the moving objects (e.g., cars) are visible in the third column. 
The moving object matrix has been color-adjusted to make the cars more obvious, which has also magnified the noise.
%There are sufficient literature that utilizes the background $\hat{\AA}$ and produces a high  quality moving objects from the input image $\AA$, which is not the focus of this paper.

\section{Topic Modeling of Stack Exchange Data}

We downloaded the latest Stack Overflow  dump from its archive on 28-Jul-2016. The details of 
the preprocessing and the sparse matrix generation are explained in Section 6.1.1. We ran 
our \ParNMF\ algorithm on this dataset, which has nearly 12 million questions from the Stack Overflow site (under Stack Exchange) to produce 50 topics. 
The matrix $\WW$ can be interpreted as {\em vocabulary-topic} distribution and the 
$\HH$ as {\em topic-document} distribution.  We took the top 5 words for each of the 50 topics and 
present them in Table \ref{tab:stackexchangetopics}. Typically a good topic generation satisfies 
properties such as (a) finding discriminative rather than common words -- capturing words that can provide 
some information; (b) finding different topics -- the similarity between different topics should be low; 
(c) coherence - all the words that belong to one topic should be coherent.  There are some topic quality 
metrics \cite{NLGB2010} that capture the usefulness of topic generation algorithm.  We can  see
 NMF generated generally high-quality and coherent topics. 
 Also, each of the topics are from different domains such as 
 databases, C/C++ programming, Java programming, and web technologies like PHP and HTML.

%\pgfplotstableset{
%begin table=\begin{longtable},
%end table=\end{longtable},
%}

%\begin{table}
%\footnotesize
%\pgfplotstabletypeset[
%	col sep=comma,
%	every head row/.style={before row=\toprule,after row=\midrule},
%        every last row/.style={after row=\bottomrule},
%	columns={word1,word2,word3,word4,word5,word1,word2,word3,word4,word5},
%	display columns/0/.style={column type/.add={|}{}, select equal part entry of={0}{2}, string type},
%	display columns/1/.style={select equal part entry of={0}{2}, string type},
%	display columns/2/.style={select equal part entry of={0}{2}, string type},
%	display columns/3/.style={select equal part entry of={0}{2}, string type},
%	display columns/4/.style={column type/.add={}{||}, select equal part entry of={0}{2}, string type},
%	display columns/5/.style={column type/.add={>{\columncolor[gray]{.8}}}{},select equal part entry of={1}{2}, string type},
%	display columns/6/.style={column type/.add={>{\columncolor[gray]{.8}}}{},select equal part entry of={1}{2}, string type},
%	display columns/7/.style={column type/.add={>{\columncolor[gray]{.8}}}{},select equal part entry of={1}{2}, string type},
%	display columns/8/.style={column type/.add={>{\columncolor[gray]{.8}}}{},select equal part entry of={1}{2}, string type},
%	display columns/9/.style={column type/.add={>{\columncolor[gray]{.8}}}{},column type/.add={}{|}, select equal part entry of={1}{2}, string type},
%	%every even column/.style={column type/.add={>{\columncolor[gray]{.8}}}{}}
%	outfile=stackxchangetopics.tex
%]{data/topkwords.csv}
%\normalsize
%\caption{Top 5 words of 50 topics from {\em Stack Exchange} data set. \grey{Maybe make left 5 columns white and right 5 gray?  It's hard to see that there are two columns to connect the 5 related words.}}
%\label{tab:stackexchangetopics}
%\end{table}
%

\newcommand{\wdth}{1.65in}
\newcommand{\hght}{1.0725in}
\begin{figure*}[th!]
\centering
\begin{tabular}{ccc}
Input Frame($\AA$) & Background ($\WW\HH$) & Moving Object $\AA-\WW\HH$ \\
\includegraphics[height=\hght,width=\wdth]{fig/input1.png} &  
\includegraphics[height=\hght,width=\wdth]{fig/lr1.png}&
\includegraphics[height=\hght,width=\wdth]{fig/err1-edit.png} \\
\includegraphics[height=\hght,width=\wdth]{fig/input4000.png} &  
\includegraphics[height=\hght,width=\wdth]{fig/lr4000.png}&
\includegraphics[height=\hght,width=\wdth]{fig/err4000-edit.png} \\
\includegraphics[height=\hght,width=\wdth]{fig/input12000.png} &  
\includegraphics[height=\hght,width=\wdth]{fig/lr12000.png}&
\includegraphics[height=\hght,width=\wdth]{fig/err12000-edit.png} \\
\end{tabular}
\caption{Moving object detection for video data using NMF.  Each row of images corresponds to a particular frame in the video.  The left column is the original frame, the middle column is the reconstructed frame from the low-rank approximation (which captures the background), and the right column is the difference (which captures the moving objects) with color adjustment.}
\label{fig:videoresults}
\end{figure*}

\begin{table*}[th!]
\begin{center}
\footnotesize
\input{stackxchangetopics}
\normalsize
\end{center}
\caption{Top 5 words of 50 topics from {\em Stack Exchange} data set.}
%\grey{Maybe make left 5 columns white and right 5 gray?  It's hard to see that there are two columns to connect the 5 related words.}
\label{tab:stackexchangetopics}
\end{table*}

\bibliographystyle{IEEEtran}
\bibliography{paper}
\end{document}

### 2018-06-20 latex/toms/survey.tex

\section{Related Work}\label{sec:related}

In the data mining and machine learning literature there is an overlap between low rank approximations and matrix factorizations due to the nature of applications. 
Despite its name, non-negative matrix ``factorization'' is really a low rank approximation. 
Recently there is a growing interest in collaborative filtering based 
recommender systems. One of the popular techniques
for collaborative filtering is matrix factorization, often with nonnegativity constraints, 
and its implementation is widely available in many
off-the-shelf distributed machine learning libraries
such as GraphLab \cite{low2012}, MLLib \cite{meng2015mllib},
and many others \cite{satish2014,yun2014} as well.
However, we would like to clarify that collaborative
filtering using matrix factorization is a different problem than NMF: 
in the case of collaborative filtering, non-nonzeros in the matrix 
are considered to be missing entries, while in the case of NMF, non-nonzeros in the matrix correspond to true zero values.

There are several recent distributed NMF algorithms in the literature \cite{liao2014cloudnmf,Faloutsos2014,Yin2014,liu2010distributed}. 
Liu et al.\ propose running Multiplicative Update (MU) for KL divergence, squared loss, and ``exponential'' loss functions \cite{liu2010distributed}. 
Matrix multiplication, element-wise multiplication, and element-wise division are the building blocks of the MU algorithm. 
The authors discuss performing these matrix operations effectively in Hadoop for sparse matrices. 
Using similar approaches, Liao et al.\ implement an open source Hadoop-based MU algorithm and study its scalability on large-scale biological data sets \cite{liao2014cloudnmf}. 
Also, Yin, Gao, and Zhang present a scalable NMF that can perform frequent updates, which aim to use the most recently updated data \cite{Yin2014}. 
Similarly Faloutsos et al.\ propose a distributed, scalable method for decomposing matrices, tensors, and coupled data sets through stochastic gradient descent on a variety of objective functions \cite{Faloutsos2014}. 
The authors also provide an implementation that can enforce non-negative constraints on the factor matrices. 
All of these works use Hadoop to implement their algorithms.

We emphasize that our MPI-based approach has several advantages over Hadoop-based approaches:
\begin{itemize}
	\item efficiency -- our approach maintains data in memory, never communicating the data matrix, while Hadoop-based approaches must read/write data to/from disk and involves global shuffles of data matrix entries;
	\item generality -- our approach is well-designed for both dense and sparse data matrices, whereas Hadoop-based approaches generally require sparse inputs;
	\item privacy -- our approach allows processors to collaborate on computing an approximation without ever sharing their local input data (important for applications involving sensitive data, such as electronic health records), while Hadoop requires the user to relinquish control of data placement.
\end{itemize}

We note that Spark \cite{ZCFSS10} is a popular big-data processing infrastructure that is generally more efficient for iterative algorithms such as NMF than Hadoop, as it maintains data in memory and avoids file system I/O.
Even with a Spark implementation of previously proposed Hadoop-based NMF algorithm, we expect performance to suffer from expensive communication of input matrix entries, and Spark will not overcome the shortcomings of generality and privacy of the previous algorithms.
Although Spark has collaborative filtering libraries such as MLlib \cite{meng2015mllib}, which use matrix factorization and can impose non-negativity constraints, none of them implement pure NMF, and so we do not have a direct comparison against NMF running on Spark.
As mentioned above, the problem of collaborative filtering is different from NMF, and therefore different computations are performed at each iteration.

%Apart from the distributed implementation for NMF, we would also like to highlight few recent theoretical highlights with NMF under some mild assumptions. 
%Arora, Ge, Kannan and Moitra \cite{AGKM2012}, proposed a initial polynomial time running algorithm for topic modeling with NMF when there exists  anchor words, words that appear (with positive probability) in only one topic. 
%Later, Huang, Sidiropoulos and Swami \cite{HSS2016} improved up on these results under milder conditions using second-order moments. 
%The uniqueness of NMF has been of interest to the community and some recent results for both symmetric and non-symmetric matrices have been obtained using a geometrical point of view in \cite{HSS2013, HSS2014}.

Fairbanks et al. \cite{Fairbanks2015} present a parallel NMF algorithm designed for multicore machines.  
To demonstrate the importance of minimizing communication, we consider this approach to parallelizing an alternating-updating NMF algorithm in distributed memory (see Section \ref{sec:naive}).
While this naive algorithm exploits the natural parallelism available within the alternating iterations (the fact that rows of $\WW$ and columns of $\HH$ can be computed independently), it performs more communication than necessary to set up the independent problems.
We compare the performance of this algorithm with our proposed approach to demonstrate the importance of designing algorithms to minimize communication; that is, simply parallelizing the computation is not sufficient for satisfactory performance and parallel scalability.

Apart from distributed NMF algorithms using Hadoop and multicores, there are also implementations of the
MU algorithm in a distributed memory setting using X10 \cite{Grove2014} and on a GPU \cite{mejia2015nmf}.

### 2018-06-21 nmf/nmf.cpp
chrono works

@@ -50,10 +50,10 @@ void NMFDriver(int k, UWORD m, UWORD n, std::string AfileName,
   nmfAlgorithm.computeNMF();
   t2 = toc();
   INFO << "time taken:" << t2 << std::endl;
-  if (!WfileName.empty()) {
+  if (WfileName.compare("_w") != 0) {
     nmfAlgorithm.getLeftLowRankFactor().save(WfileName, arma::raw_ascii);
   }
-  if (!HfileName.empty()) {
+  if (HfileName.compare("_h") != 0) {
     nmfAlgorithm.getRightLowRankFactor().save(HfileName, arma::raw_ascii);
   }
 }

### 2018-06-21 test/testmm.cpp

@@ -1,6 +1,9 @@
 // to compile the code run as
-//g++ testmm.cpp -I$ARMADILLO_INCLUDE_DIR -L/lustre/atlas/proj-shared/csc209/ramki/rhea/libraries/openblas/lib/ -lopenblas --std=c++11
+// g++ testmm.cpp -I$ARMADILLO_INCLUDE_DIR
+// -L/lustre/atlas/proj-shared/csc209/ramki/rhea/libraries/openblas/lib/
+// -lopenblas --std=c++11
 
+#include <omp.h>
 #include <armadillo>
 #include <iostream>
 #include <stack>
@@ -8,42 +11,60 @@
 using namespace std;
 using namespace arma;
 
-static std::stack<clock_t> tictoc_stack;
-
-inline void tic() {
-    tictoc_stack.push(clock());
-}
+#ifdef USE_CHRONO
+static std::stack<std::chrono::steady_clock::time_point> tictoc_stack;
+inline void tic() { tictoc_stack.push(std::chrono::steady_clock::now()); }
 
 inline double toc() {
-    double rc = (static_cast<double>
-                 (clock() - tictoc_stack.top())) / CLOCKS_PER_SEC;
-    tictoc_stack.pop();
-    return rc;
+  std::chrono::duration<double> time_span =
+      std::chrono::duration_cast<std::chrono::duration<double>>(
+          std::chrono::steady_clock::now() - tictoc_stack.top());
+  double rc = time_span.count();
+  tictoc_stack.pop();
+  return rc;
 }
+#else
+static std::stack<double> tictoc_stack;
+inline void tic() { tictoc_stack.push(omp_get_wtime()); }
 
-#define PRINTMATINFO(A) cout << "::"#A"::" << (A).n_rows << "x" << (A).n_cols << endl;
+inline double toc() {
+  double rc = (static_cast<double>(omp_get_wtime() -
+                                   tictoc_stack.top()));  // / CLOCKS_PER_SEC;
+  tictoc_stack.pop();
+  return rc;
+}
+#endif
+
+#define PRINTMATINFO(A) \
+  cout << "::" #A "::" << (A).n_rows << "x" << (A).n_cols << endl;
 
 /*
-* multiplies as sparse mxk matrix with
-* dense kxn matrix
-*/
[…]

### 2018-06-22 latex/toms/algorithm.tex
latex didnt complete
fixed a couple latex bugs
sc papers in toms format

@@
-\section{Parallel Algorithms}
-\label{sec:parallel}
+% !TEX root = paper.tex
 
-\subsection{Naive Parallel NMF Algorithm} 
-\label{sec:naive}
+\section{Algorithm} 
+\label{sec:algorithm}
 
-In this section we present a naive parallelization of NMF algorithms, which has previously appeared in the context of a shared-memory parallel platform \cite{Fairbanks2015}. 
-Each NLS problem with multiple right-hand sides can be parallelized based on the observation that each right-hand side is independent from the others. 
-For example, we can solve several instances of Eq.~\eqref{eqn:single NLS} independently for different $\mathbf{b}$ where $\CC$ is fixed, which implies that we can optimize row blocks of $\WW$ and column blocks of $\HH$ in parallel.
-%\ramki{In this paper  we are using BPP to solve this Non-negative Least Squares (NLS) problem and call the algorithm as Naive Parallel BPP(\NaiveAlg).
-But the conversation in this section is not limited to \NaiveAlg and with out loss of generality is applicable to any algorithm that follows ANLS Algorithm \ref{alg:anlsnmf} such as Multiplicative Update (MU) and HALS as well.}
+\subsection{Dimension Trees}
+\label{sec:dimtrees}
 
-\begin{algorithm}[t!]
-\caption{$[\WW,\HH] = \text{\NaiveAlg}(\AA,k)$}
-\label{alg:naive}
+An important optimization of the CP-ALS algorithm is to re-use temporary values across inner iterations \cite{PTC13a,KU16-TR,LCPSV17,Kaya17}.
+To illustrate the idea, consider a 3-way tensor $\T{X}$ approximated by $\dsquare{\M{U},\M{V},\M{W}}$ and the two MTTKRP computations $\Mn{M}{1}=\underline{\Mz{X}{1}(\M{W}}\Khat\M{V})$ and $\Mn{M}{2}=\underline{\Mz{X}{2}(\M{W}}\Khat\M{U})$ used to update factor matrices $\M{U}$ and $\M{V}$, respectively.
+The underlined parts of the expressions correspond to the shared dependence of the outputs on the tensor $\T{X}$ and the third factor matrix $\M{W}$.
+Indeed, a temporary quantity, which we refer to as a \emph{partial MTTKRP}, can be computed and re-used across the two MTTKRP expressions.
+We refer to the computation that combines the temporary quantity with the other factor matrix to complete the MTTKRP computation as a multi-tensor-times-vector or \emph{multi-TTV}, as it consists of multiple operations that multiply a tensor times a set of vectors, each corresponding to a different mode.
+%This terminology was used in \cite{HBJT18}.
+
+To understand the steps of the partial MTTKRP and multi-TTV operations in more detail, we consider $\T{X}$ to be $I\times J\times K$ and $\M{U}$, $\M{V}$, and $\M{W}$ to have $R$ columns.
+Then 
+\begin{equation*}
+\MnE{M}{1}{ir} = \sum_{i,j} \TE{X}{ijk} \ME{V}{jr} \ME{W}{kr} 
+= \sum_{j} \ME{V}{jr} \sum_k \TE{X}{ijk} \ME{W}{kr} 
+= \sum_{j} \ME{V}{jr} \TE{T}{ijr},
+\end{equation*}
+where $\T{T}$ is an $I\times J\times R$ tensor that is the result of a partial MTTKRP between tensor $\T{X}$ and the single factor matrix $W$.
+Likewise,
+\begin{equation*}
+\MnE{M}{2}{jr} = \sum_{i,k} \TE{X}{ijk} \ME{U}{ir} \ME{W}{kr} 
+= \sum_{i} \ME{U}{ir} \sum_k \TE{X}{ijk} \ME{W}{kr} 
+= \sum_{i} \ME{U}{ir} \TE{T}{ijr},
+\end{equation*}
+and we see that the temporary tensor $\T{T}$ can be re-used.
+From these expressions, we can also see that computing $\T{T}$ (a partial MTTKRP) corresponds to a matrix-matrix multiplication, and computing each of $\Mn{M}{1}$ and $\Mn{M}{2}$ from $\T{T}$ (a multi-TTV) corresponds to $R$ independent matrix-vector multiplications.
+In this case, we compute $\Mn{M}{3}$ using a full MTTKRP.
+
+For a larger number of modes, a more general approach can organize the temporary quantities to be used over a maximal number of MTTKRPs.
+The general approach can yield significant benefit, decreasing the computation by a factor of approximately $N/2$ for dense $N$-way tensors.
+The idea is introduced in \cite{PTC13a}, but we adopt the terminology and notation of \emph{dimension trees} used for sparse tensors in \cite{KU16-TR,Kaya17}.
+In this notation, the root node is labeled $\{1,\dots,N\}$ and corresponds to the original tensor, a leaf is labeled $\{n\}$ and corresponds to the $n$th MTTKRP result, and an internal node is labeled by a set of modes $\{i,\dots,j\}$ and corresponds to a temporary tensor whose values contribute to the MTTKRP results of modes $i,\dots,j$.
+
+\begin{figure}
+\input{fig/dim_tree.tex}
+\caption{Dimension tree example for $N=5$. 
+The data associated with the root node is the original tensor, the data associated with the leaf nodes are MTTKRP results, and the data associated with internal nodes are temporary tensors.  
+Edges labeled with PM correspond to partial MTTKRP computations, and edges labeled with mTTV correspond to multi-TTV computations.}
+\label{fig:DT}
+\end{figure}
+
+\Cref{fig:DT} illustrates a dimension tree for the case $N=5$.
+Various shapes of binary trees are possible \cite{PTC13a,Kaya17}.
+For dense tensors, the computational cost is dominated by the root's branches, which correspond to partial MTTKRP computations.
+We perform the splitting of modes at the root so that modes are chosen contiguously with the respect to the layout of the tensor entries in memory.
+In this way, each partial MTTKRP can be performed via BLAS's GEMM interface without reordering tensor entries in memory.
+All other edges in a tree correspond to multi-TTVs and are typically much cheaper.
+By organizing the memory layout of temporary quantities, the multi-TTV operations can be performed via a sequence of calls using BLAS's GEMV interface.
+By using the BLAS in our implementation, we are able to obtain high performance and on-node parallelism.
+
+\begin{figure}
+\subfloat[Partial MTTKRP to compute node $\{3,4,5\}$ from root node $\{1,2,3,4,5\}$, executed via one GEMM call. \label{fig:DT-PM}]{\includegraphics[width=\columnwidth]{fig/PM}} \\
+\subfloat[Multi-TTV to compute node $\{3\}$ from node $\{3,4,5\}$, executed via $R$ GEMV calls.
+Here $\Mz{T}{1} \lbrack r\rbrack$ refers to the $r$th contiguous block of $\Mz{T}{1}$.
+\label{fig:DT-mTTV}]{\includegraphics[width=\columnwidth]{fig/mTTV}}
+\caption{Data layout and dimensions for two example computations in dimension tree shown in \Cref{fig:DT}.
+In this notation, $\Mz{X}{3:5}$ is the matricization of input tensor $\T{X}$ with respect to modes 3 through 5, $\M{K}_{1:2} = \Mn{H}{2} \Khat \Mn{H}{1}$, $\T T$ is the temporary $I_3 \times I_4 \times I_5 \times R$ tensor corresponding to node $\{3,4,5\}$ in the dimension tree, $\M{K}_{4:5} = \Mn{H}{5} \Khat \Mn{H}{4}$, and $\Mn{M}{3}$ is the MTTKRP result for mode 3.}
+\label{fig:DTcartoon}
+\end{figure}
+
+\Cref{fig:DTcartoon} shows the data layout and dimensions of a partial MTTKRP and a multi-TTV taken from the example dimension tree in \Cref{fig:DT}.
+\Cref{fig:DT-PM} shows a partial MTTKRP between the input tensor $\T{X}$ and the Khatri-Rao product of the factor matrices in modes 1 and 2, which produces a temporary tensor $\T{T}$ corresponding to the $\{3,4,5\}$ node in the dimension tree.
+The key to efficiency in this computation is that the matricization of $\T{X}$ that assigns modes 1 through 2 to rows and modes 3 through 5 to columns is already column-major in memory.
+Thus, we can use the GEMM interface and compute the temporary tensor $\T{T}$ without reordering any tensor entries.
+\Cref{fig:DT-mTTV} depicts a multi-TTV that computes the results $\Mn{M}{3}$ from $\T{T}$ and the factor matrices in modes 4 and 5.
+Here, the tensor $\T{T}$ is matricized with respect to only its first mode (of dimension $I_3$), but this matricization is also column-major in memory.
+We choose the ordering of the modes of $\T{T}$ such that each of $R$ contiguous blocks is used to compute one column of the output matrix via a matrix-vector operation with a corresponding column of the Khati-Rao product of the other factor matrices.
+
+No matter how the dimension tree is designed, the computational cost of each partial MTTKRP is $O(IR)$, where $I$ is the number of tensor entries and $R$ is the rank of the CP decomposition.
+This is the same operation count as a full MTTKRP.
+The computational cost of a multi-TTV is the number of entries in the temporary tensor, which is the product of a \emph{subset} of the original tensor dimensions multiplied by $R$.
+Thus, it is computationally cheaper than the partial MTTKRPs, but it is also memory bandwidth bound.
+
+The other subroutine necessary for implementing the dimension tree approach is the Khatri-Rao product of sets of factor matrices.
+We implement the operation as a row-wise Hadamard product of a set of factor matrix rows, and we use OpenMP parallelization to obtain on-node parallelism.
+The computational cost of this operation is also typically lower order, but the running time in practice suffers from also being memory bandwidth bound.
+
+\subsection{Relative Error Computation}
+\label{sec:error}
+
+Given a model $\T{M}=\CP$, we compute the relative error $\|\TA - \T{M}\|/\|\TA\|$ efficiently by using the identity $\|\TA-\T{M}\|^2 = \|\TA\|^2 - 2\langle \TA, \T{M} \rangle + \|\T{M}\|^2.$
+The quantity $\|\TA\|$ is fixed, and the other two terms can be computed cheaply given the temporary matrices computed during the course of the BCD algorithm.
+The second term can be computed using the identity $\langle \TA, \T{M} \rangle = \langle \Mn{M}{N}, \Mn{H}{N} \rangle$, where $\Mn{M}{N} = \Mz{A}{N} (\Mn{H}{N-1} \Khat \cdots \Khat \Mn{H}{1})$ is the MTTKRP result in the $N$th mode.
+The third term can be computed using the identity $\|\T{M}\|^2 = \V{1}^\Tra(\Mn{S}{N} \Hada \MnTra{H}{N} \Mn{H}{N})\V{1}$ where $\Mn{S}{N}=\MnTra{H}{1} \Mn{H}{1} \Hada \cdots \Hada \MnTra{H}{N-1} \Mn{H}{N-1}$.
+Both matrices $\Mn{M}{N}$ and $\Mn{S}{N}$ are computed during the course of the BCD algorithm for updating the factor matrix $\Mn{H}{N}$.
+The extra computation involved in computing the relative error is negligible.
+These identities have been used previously \cite{KB2009,TensorBox,SK16,LKLHS2017}.
+
+\subsection{Parallel Algorithm}
+
+\begin{algorithm}
+\caption{$\CP = \text{Par-NNCP}(\TA,R)$}
+\label{alg:Par-NNCP-short}
 \begin{algorithmic}[1]
-\Require $\AA$ is an $m\times n$ matrix distributed both row-wise and column-wise across $p$ processors, $k$ is the approximation rank
-\Require Local matrices: $\AA_{i}$ is $m/p\times n$, $\AA^{i}$ is $m\times n/p$, $\WW_i$ is $m/p\times k$, $\HH^i$ is $k\times n/p$
-\State $p_i$ initializes $\HH^i$
-\While{stopping criteria not satisfied}
-	\Statex \quad\; \textbf{/* Compute $\WW$ given $\HH$ */} 
-	\State collect $\HH$ on each processor using all-gather
-		\label{line:naive:allgatherH}
-	%\Statex \quad\; \textbf{/* Use Equation \eqref{eqn:muupdate}, \eqref{eqn:halsupdate} for \MU\ and \HALS\ respectively. ANLS/BPP will implement the SolveBPP function \ref{sec:BPP}.*/} 	
-	\State $p_i$ computes $\WW_i \gets$ updateW$(\HH\HH^T,\AA_i\HH^T)$
-		\label{line:naive:computeW}
-	\Statex \quad\; \textbf{/* Compute $\HH$ given $\WW$ */} 
-	\State collect $\WW$ on each processor using all-gather
-		\label{line:naive:allgatherW}
-	%\Statex \quad\; \textbf{/* Use Equation \eqref{eqn:muupdate},\eqref{eqn:halsupdate} for \MU\ and \HALS\ respectively. ANLS/BPP will implement the SolveBPP function \ref{sec:BPP}.*/} 	
-	\State $p_i$ computes $(\HH^i)^T \gets$ updateH$(\WW^T\WW,(\WW^T\AA^i)^T)$
-		\label{line:naive:computeH}
+\Require $\TA$ is an $I_1\times \cdots \times I_N$ tensor distributed across a $P_1\times \cdots \times P_N$ grid of $P$ processors, so that $\TA_{\V{p}}$ is $(I_1/P_1)\times \cdots \times (I_N/P_N)$ and is owned by processor $\V{p}=(p_1,\dots,p_N)$, $R$ is rank of approximation
+\For{$n=2$ to $N$}
+	\State Initialize $\Mn{H}{n}_{\V{p}}$ of dimensions $(I_n/P)\times R$ 
+	\State $\M[\overline]{G} = \text{Local-SYRK}(\Mn{H}{n}_{\V{p}})$
+	\State $\Mn{G}{n} = \text{All-Reduce}(\M[\overline]{G},\textsc{All-Procs})$
+	\State $\Mn{H}{n}_{p_n} = \text{All-Gather}(\Mn{H}{n}_{\V{p}},\textsc{Proc-Slice}(n,\VE{p}{n}))$
+\EndFor
+\State \Comment{Compute NNCP approximation}
+\While{not converged}
+	\label{line:while}
+	\State \Comment{Perform outer iteration of BCD}
+	\For{$n=1$ to $N$}
+		\label{line:for}
+		\State \Comment{Compute new factor matrix in $n$th mode}
+		\State $\M[\overline]{M} = \text{Local-MTTKRP}(\TA_{p_1\cdots p_N},\{\Mn{H}{i}_{p_i}\},n)$
+			\label{line:locMTTKRP}
+		\State $\Mn{M}{n}_{\V{p}} = \text{Reduce-Scatter}(\M[\overline]{M},\textsc{Proc-Slice}(n,\VE{p}{n}))$ 
+			\label{line:reduce-scatter}
+		\State $\Mn{S}{n} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$
+			\label{line:hadamard}
+		\State $\Mn{H}{n}_{\V{p}} = \text{Local-NLS-Update}(\Mn{S}{n},\Mn{M}{n}_{\V{p}})$
+			\label{line:locNLS}
+		\State \Comment{Organize data for later modes}
+		\State $\M[\overline]{G} = {\Mn{H}{n}_{\V{p}}}^\Tra\Mn{H}{n}_{\V{p}}$
+			\label{line:locSYRK}
+		\State $\Mn{G}{n} = \text{All-Reduce}(\M[\overline]{G},\textsc{All-Procs})$
+			\label{line:all-reduce}
+		\State $\Mn{H}{n}_{p_n} = \text{All-Gather}(\Mn{H}{n}_{\V{p}},\textsc{Proc-Slice}(n,\VE{p}{n}))$
+			\label{line:all-gather}
+	\EndFor 
+		\label{line:endfor}
 \EndWhile
-\Ensure $\displaystyle \WW, \HH \approx \Argmin{\M{\tilde W} \geq 0, \M{\tilde H} \geq 0} \|\AA- \M{\tilde W} \M{\tilde H}\|$
-\Ensure $\WW$ is an $m\times k$ matrix distributed row-wise across processors, $\HH$ is a $k\times n$ matrix distributed column-wise across processors
+	\label{line:endwhile}
+\Ensure $\TA \approx \CP$
+\Ensure Local matrices: $\Mn{H}{n}_{\V{p}}$ is $(I_n/P)\times R$ and owned by processor $\V{p}=(p_1,\dots,p_N)$, for $1\leq n \leq N$, $\V{\lambda}$ stored redundantly on every processor
 \end{algorithmic}
@@
 
-\begin{figure}[t!]
+\subsubsection{Algorithm Overview}
+
+The basic sequential algorithm is given in \Cref{alg:nncp}, and the parallel version is given in \Cref{alg:Par-NNCP-short}.
+We will refer to both the inner iteration, in which one factor matrix is updated (\cref{line:for} to \cref{line:endfor}), and the outer iteration, in which all factor matrices are updated (\cref{line:while} to \cref{line:endwhile}).
+In the parallel algorithm, the processors are organized into a logical multidimensional grid (tensor) with as many modes as the data tensor.
+The communication patterns used in the algorithm are all MPI collectives, including All-Reduce, Reduce-Scatter, and All-Gather.
+The processor communicators (across which the collectives are performed) include the set of all processors and the sets of processors within the same processor slice.
+Processors within a mode-$n$ slice all have the same $n$th coordinate.
+
+The method of enforcing the nonnegativity constraints of the linear least squares solve (or update) generally affects only local computation because each row of a factor matrix can be updated independently.
+In our algorithm, each processor solves the linear problem or computes the update for its subset of rows (see \cref{line:locNLS}). 
+The most expensive (and most complicated) part of the parallel algorithm is the computation of the MTTKRP, which corresponds to \cref{line:locMTTKRP,line:reduce-scatter,line:all-gather}.
+
+The details that are omitted from this presentation of the algorithm include the normalization of each factor matrix after it is computed and the computation of the residual error at the end of an outer iteration.
+The computations do involve both local computation and communication, but their costs are negligible.
+A more detailed pseudocode is given in \Cref{alg:Par-NNCP-long}.
+
+\input{fig/Alg_header.tex}
+\begin{figure*}[t]
 \centering
-\includegraphics[height=2.5in, width=\columnwidth]{fig/anlsbpp.pdf}
-\caption{\NaiveAlg.
-Both rows and columns of $A$ are 1D distributed.
-The algorithm works by (all-)gathering the entire fixed factor matrix to each processor and then performing the Local Update Computations to update the variable factor matrix.}
-\label{fig:naive}
-\end{figure}
-
-\renewcommand{\arraystretch}{1.25}
-\newcommand{\smaller}{\scriptsize}
-\begin{table*}[t!]
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|}
-\hline
-\textbf{Algorithm} & \textbf{Flops} & \textbf{Words} & \textbf{Messages} & \textbf{Memory}  \\ \hline
-\smaller \NaiveAlg & \smaller $4\frac{mnk}{p}+(m{+}n)k^2+F\lt(\frac mp,\frac np,k\rt)$ & \smaller $O((m+n)k)$ & \smaller $O(\log p)^*$ & \smaller $O\lt(\frac{mn}{p}+(m{+}n)k\rt)$ \\ \hline
-\smaller \ParNMF\ ($m/p \geq n$) & \smaller $4\frac{mnk}{p}+\frac{(m+n)k^2}{p} + F\lt(\frac mp,\frac np,k\rt)$ & \smaller $O(nk)$ & \smaller $O(\log p)^*$ & \smaller $O\lt(\frac{mn}{p}+\frac{mk}{p}+nk\rt)$ \\ \hline
-\smaller \ParNMF\ ($m/p < n$) & \smaller $4\frac{mnk}{p}+\frac{(m+n)k^2}{p}+F\lt(\frac mp,\frac np,k\rt)$ & \smaller $O\lt( \sqrt{\frac{mnk^2}{p}}\rt)$ & \smaller $O(\log p)^*$ & \smaller $O\lt(\frac{mn}{p}+\sqrt{\frac{mnk^2}{p}}\rt)$ \\ \hline
-\smaller Lower Bound & $-$ & \smaller $\Omega\lt(\min\lt\{\sqrt{\frac{mnk^2}{p}},nk\rt\}\rt)$ & \smaller $\Omega(\log p)$ & \smaller $\frac{mn}{p}+\frac{(m+n)k}{p}$ \\ \hline
-\end{tabular}
-\normalsize
-\end{center}
-\caption{Leading order algorithmic costs for \NaiveAlg{} and \ParNMF{} (per iteration).
-Note that the computation and memory costs assume the data matrix $\AA$ is dense, but the communication costs (words and messages) apply to both dense and sparse cases.
-The function $F(\cdot)$ denotes the number of flops required for the particular NMF algorithm's Local Update Computation, aside from the matrix multiplications common across AU-NMF algorithms. 
-Note that $F(m,n,k)$ is proportional to $m+n$ and not $mn$, so the term in the table scales linearly with $p$ (and not $p^2$) for all \LUC. \\
-$^*$The stated latency cost assumes no communication is required in \LUC; \HALS\ requires $k\log p$ messages for normalization steps.}
-\label{tab:costs}
-\end{table*}%
-
-Algorithm \ref{alg:naive} and Figure \ref{fig:naive} present a straightforward approach to parallelizing the independent subproblems.
-Let us divide $\WW$ into row blocks $\WW_1, \ldots, \WW_p$ and $\HH$ into column blocks $\HH^1, \ldots, \HH^p$. 
-We then double-partition the data matrix $\AA$ accordingly into row blocks $\AA_{1}, \ldots, \AA_p$ and column blocks $\AA^1, \ldots, \AA^p$ so that processor $i$ owns both $\AA_i$ and $\AA^i$ (see Figure \ref{fig:naive}).
-With these partitions of the data and the variables, one can implement any AU-NMF algorithm in parallel, with only one communication step for each solve.
-
-We summarize the algorithmic costs of Algorithm \ref{alg:naive} (derived in the following subsections) in Table \ref{tab:costs}.
-This naive algorithm \cite{Fairbanks2015} has three main drawbacks: (1) it requires storing two copies of the data matrix (one in row distribution and one in column distribution) and both full factor matrices locally, (2) it does not parallelize the computation of $\HH\HH^T$ and $\WW^T\WW$ (each processor computes it redundantly), and (3) as we will see in Section \ref{sec:parNMF}, it communicates more data than necessary.
-
-\subsubsection{Computation Cost}
-
-The computation cost of Algorithm \ref{alg:naive} depends on the particular NMF algorithm used.
-Thus, the computation at line \ref{line:naive:computeW} consists of computing $\AA_i\HH^T$, $\HH\HH^T$, and performing the algorithm-specific Local Update Computations for $m/p$ rows of $\WW$.
-Likewise, the computation at line \ref{line:naive:computeH} consists of computing $\WW^T\AA^i$, $\WW^T\WW$, and performing the Local Update Computations for $n/p$ columns of $\HH$.
-In the dense case, this amounts to $4mnk/p+(m+n)k^2+F(m/p,n/p,k)$ flops. 
-Note that the first term has a constant 4 to account for both $\WW^T\AA$ and $\AA\HH^T$ and that the second term has a constant factor of 1 instead of 2 because the Gram computations ($\HH \HH^T$ and $\WW^T \WW$) exploit symmetry of the output matrix.
-In the sparse case, processor $i$ performs $2(\nnz(\AA_i)+\nnz(\AA^i))k$ flops to compute $\AA^i\HH^T$ and $\WW^T\AA_i$ instead of $4mnk/p$.
-
-\subsubsection{Communication Cost}
-
-The size of $\WW$ is $mk$ words, and the size of $\HH$ is $nk$ words.
-Thus, the communication cost of the all-gathers at lines \ref{line:naive:allgatherH} and \ref{line:naive:allgatherW}, based on the expression given in Section \ref{sec:collectives} is $\alpha\cdot 2\log p + \beta\cdot (m+n)k$.
-
-\subsubsection{Memory Requirements} \label{sec:memory}
-
-The local memory requirement includes storing each processor's part of matrices $\AA$, $\WW$, and $\HH$.
-In the case of dense $\AA$, this is $2mn/p+(m+n)k/p$ words, as $\AA$ is stored twice; in the sparse case, processor $i$ requires $\nnz(\AA_i)+\nnz(\AA^i)$ words for the input matrix and $(m+n)k/p$ words for the output factor matrices.
-Local memory is also required for storing temporary matrices $\WW$ and $\HH$ of size $(m+n)k$ words.
-
-\subsection{\ParNMF}
-\label{sec:parNMF}
-
-We present our proposed algorithm, \ParNMF, as Algorithm \ref{alg:2D} and Figure \ref{fig:2D-Algorithm}. 
-The main ideas of the algorithm are to (1) exploit the independence of Local Update Computations for rows of $\WW$ and columns of $\HH$ and (2) use communication-optimal matrix multiplication algorithms to set up the Local Update Computations.
-The naive approach (Algorithm \ref{alg:naive}) shares the first property, by parallelizing over rows of $\WW$ and columns of $\HH$, but it uses parallel matrix multiplication algorithms that communicate more data than necessary.
-The central intuition for communication-efficient parallel algorithms for computing $\HH\HH^T$, $\AA\HH^T$, $\WW^T\WW$, and $\WW^T\AA$ comes from a classification proposed by Demmel et al.
-\cite{DE+13}.
-They consider three cases, depending on the relative sizes of the dimensions of the matrices and the number of processors; the four multiplies for NMF fall into either the ``one large dimension'' or ``two large dimensions" cases.
-\ParNMF\ uses a careful data distribution in order to use a communication-optimal algorithm for each of the matrix multiplications, while at the same time exploiting the parallelism in the \LUC.
-
-The algorithm uses a 2D distribution of the data matrix $\AA$ across a $p_r \times p_c$ grid of processors (with $p=p_rp_c$), as shown in Figure \ref{fig:2D-distribution}.
-As we derive in the subsequent subsections, Algorithm \ref{alg:2D} performs an alternating method in parallel with a per-iteration bandwidth cost of $O\lt(\min\lt\{\sqrt{mnk^2/p},nk\rt\}\rt)$ words, latency cost of $O(\log p)$ messages, and load-balanced computation (up to the sparsity pattern of $\AA$ and convergence rates of local BPP computations).
-Figure \ref{fig:2D-Algorithm}, illustrates determining $\HH$ given $\WW$ with $pr=3$ and $pc=2$.
-
-The main improvement of \ParNMF\ over \Naive\ involves the computation of $\AA\HH^T$ and $\WW^T\AA$.
-By using a 2D distribution of the data matrix, no processor needs access to \emph{all} of one factor matrix, as in the case of \Naive, where each processor must access either all $m$ rows of $\WW$ or all $n$ columns of $\HH$.
-Instead, with \ParNMF, each processor must access only $m/p_r$ of the rows of $\WW$ and $n/p_c$ of the columns of $\HH$, so the number of rows decreases as $p$ increases.
-This implies the communication cost is reduced, as verified empirically in Figure \ref{fig:procsweep} (the extreme cases correspond to 1D distributions).
-
-To minimize the communication cost and local memory requirements, in the typical case $p_r$ and $p_c$ are chosen so that $m/p_r\approx n/p_c\approx \sqrt{mn/p}$, in which case the bandwidth cost is $O\lt(\sqrt{mnk^2/p}\rt)$.
-If the matrix is very tall and skinny, i.e., $m/p>n$, then we choose $p_r=p$ and $p_c=1$.
-In this case, the distribution of the data matrix is 1D, and the bandwidth cost is $O(nk)$ words.
-
-The matrix distributions for Algorithm \ref{alg:2D} are given in Figure \ref{fig:2D-distribution}; we use a 2D distribution of $\AA$ and 1D distributions of $\WW$ and $\HH$.
-Recall from Table \ref{tab:notation} that $\M{M}_i$ and $\M{M}^i$  denote row and column blocks of $\M{M}$, respectively.
-Thus, the notation $(\WW_i)_j$ denotes the $j$th row block within the $i$th row block of $\WW$.
-Lines \ref{line:2DsyrkH}--\ref{line:2DcompW} compute $\WW$ for a fixed $\HH$, and lines \ref{line:2DsyrkW}--\ref{line:2DcompH} compute $\HH$ for a fixed $\WW$; note that the computations and communication patterns for the two alternating iterations are analogous.
-
-In the rest of this section, we derive the per-iteration computation and communication costs, as well as the local memory requirements.
-We also argue the communication-optimality of the algorithm in the dense case.
-Table \ref{tab:costs} summarizes the results of this section and compares them to \NaiveAlg.
-
-\begin{algorithm}[t!]
-\caption{$[\WW,\HH] = \text{\ParNMF}(\AA,k)$}
-\label{alg:2D}
-\begin{algorithmic}[1]
-\small
-\Require $\AA$ is an $m\times n$ matrix distributed across a $p_r\times p_c$ grid of processors, $k$ is rank of approximation
-\Require Local matrices: $\AA_{ij}$ is $m/p_r\times n/p_c$, $\WW_i$ is $m/p_r\times k$, $(\WW_i)_j$ is $m/p\times k$, $\HH_j$ is $k\times n/p_c$, and $(\HH_j)_i$ is $k\times n/p$
-\State $p_{ij}$ initializes $(\HH_j)_i$
-\While{stopping criteria not satisfied}
-	\Statex \quad\; \textbf{/* Compute $\WW$ given $\HH$ */} 
-	\State $p_{ij}$ computes $\M{U}_{ij}=(\HH_j)_i{(\HH_j)_i}^T$
-		\label{line:2DsyrkH}
-	\State compute $\HH\HH^T {=} \sum_{i,j} \M{U}_{ij}$ using all-reduce across all procs
-		\label{line:2Dall-reduceH}
-		\Comment{$\HH\HH^T$ is $k\times k$ and symmetric}
-	\State $p_{ij}$ collects $\HH_j$ using all-gather across proc columns
-		\label{line:2Dall-gatherH}
-	\State $p_{ij}$ computes $\M{V}_{ij}=\AA_{ij}\HH_j^T$
-		\label{line:2DNEW}
-		\Comment{$\M{V}_{ij}$ is $m/p_r \times k$}
-	\State compute $(\AA\HH^T)_i {=} \sum_j \M{V}_{ij}$ using reduce-scatter across proc row to achieve row-wise distribution of $(\AA\HH^T)_i$
-		\Comment{$p_{ij}$ owns $m/p\times k$ submatrix $((\AA\HH^T)_i)_j$}
-		\label{line:2Dreduce-scatterAHT}
-		%\Statex \Comment{Use Equation \eqref{eqn:muupdate},\eqref{eqn:halsupdate} for \MU\ and \HALS\ respectively. ANLS/BPP will implement the SolveBPP function \ref{sec:BPP}.} 	
-	\State $p_{ij}$ computes $(\WW_i)_j \gets$ UpdateW $(\HH\HH^T,((\AA\HH^T)_i)_j)$
-		\label{line:2DcompW}
-	\Statex \quad\; \textbf{/* Compute $\HH$ given $\WW$ */}
-	\State $p_{ij}$ computes $\M{X}_{ij}={(\WW_i)_j}^T(\WW_i)_j$
-		\label{line:2DsyrkW}
-	\State compute $\WW^T\WW {=} \sum_{i,j} \M{X}_{ij}$ using all-reduce across all procs
-		\label{line:2Dall-reduceW}
-		\Comment{$\WW^T\WW$ is $k\times k$ and symmetric}
-	\State $p_{ij}$ collects $\WW_i$ using all-gather across proc rows
-		\label{line:2Dall-gatherW}
-	\State $p_{ij}$ computes $\M{Y}_{ij}={\WW_i}^T\AA_{ij}$
-		\label{line:2DNEH}
-		\Comment{$\M{Y}_{ij}$ is $k\times n/p_c$}
-	\State compute $(\WW^T\AA)^j = \sum_i \M{Y}_{ij}$ using reduce-scatter across proc columns to achieve column-wise distribution of $(\WW^T\AA)^j$
-		\Comment{$p_{ij}$ owns $k\times n/p$ submatrix $((\WW^T\AA)^j)^i$}
-		\label{line:2Dreduce-scatterWTA}
-		%\Statex \Comment{Use Equation \eqref{eqn:muupdate},\eqref{eqn:halsupdate} for \MU\ and \HALS\ respectively. ANLS/BPP will implement the SolveBPP function \ref{sec:BPP}.}
-	\State $p_{ij}$ computes $((\HH^j)^i)^T \gets$ UpdateH $(\WW^T\WW,(((\WW^T\AA)^j)^i)^T)$
-		\label{line:2DcompH}
-\EndWhile
-\Ensure $\displaystyle \WW, \HH \approx \Argmin{\M{\tilde W} \geq 0, \M{\tilde H} \geq 0} \|\AA- \M{\tilde W} \M{\tilde H}\|$
-\Ensure $\WW$ is an $m\times k$ matrix distributed row-wise across processors, $\HH$ is a $k\times n$ matrix distributed column-wise across processors
-\normalsize
-\end{algorithmic}
-\end{algorithm}
-
-\begin{figure}[t!]
-\centering
-\begin{subfigure}[b]{0.3\columnwidth}
-\centering
-\scalebox{.7}{\input{fig/1D-distribution}}
-%\subcaption{1D Distribution with $p=p_r=4$ and $p_c=1$.}
-\label{1D}
-\end{subfigure} 
-~
-\begin{subfigure}[b]{0.6\columnwidth}
-\centering
-\scalebox{.75}{\input{fig/2D-distribution}}
-%\subcaption{2D Distribution with $p_r=3$ and $p_c=2$.}
-\label{2D}
-\end{subfigure}
-\caption{Data distributions for \ParNMF. 1D Distribution on left with $p=p_r=4$ and $p_c=1$. 2D Distribution on right with $p_r=3$ and $p_c=2$
-Note that for the 2D distribution, $\AA_{ij}$ is $m/p_r \times m/p_c$, $\WW_i$ is $m/p_r \times k$, $(\WW_i)_j$ is $m/p\times k$, $\HH_j$ is $k\times n/p_c$, and $(\HH^j)^i$ is $k\times n/p$.}
-\label{fig:2D-distribution}
-\end{figure}
-~
-\begin{figure*}[t!]
-\centering
-\includegraphics[width=0.8\textwidth, height=3in]{fig/mm2D.pdf}
-\caption{Parallel matrix multiplications within \ParNMF\ for finding $\HH$ given $\WW$, with $p_r=3$ and $p_c=2$.
-The computation of $\WW^T\WW$ appears on the far left; the rest of the figure depicts computation of $\WW^T\AA$.}
-\label{fig:2D-Algorithm}
+  \subfloat[Start $n$th iteration with redundant subset of rows of each input matrix. \label{fig:inner_a}]{\input{fig/AlgStart.tex}}  \quad
+  \subfloat[Compute local MTTKRP for contribution to output matrix $\Mn{M}{2}$. \label{fig:inner_b}]{\input{fig/AlgMTTKRP.tex}}  \quad
+  \subfloat[Reduce-Scatter to compute and distribute rows of $\Mn{M}{2}$. \label{fig:inner_c}]{\input{fig/AlgRS.tex}} \quad
+  \subfloat[Compute local NLS update to obtain $\Mn{H}{2}_{\V{p}}$ from $\Mn{M}{2}_{\V{p}}$ (and $\Mn{S}{2}$). \label{fig:inner_d}]{\input{fig/AlgNLS.tex}} \quad
+  \subfloat[All-Gather to collect rows of $\Mn{H}{2}$ needed for later inner iterations. \label{fig:inner_e}]{\input{fig/AlgAG.tex}}
+  \caption{Illustration of 2nd inner iteration of Par-NNCP algorithm for 3-way tensor on a $3\times3\times3$ processor grid, showing data distribution, communication, and computation across steps.
+Highlighted areas correspond to processor $(1,3,1)$ and its processor slice with which it communicates.
+The column normalization and computation of $\Mn{G}{2}$, which involve communication across all processors, is not shown here.}
+  \label{fig:inner} 
 \end{figure*}
 
-\subsubsection{Computation Cost}
+\subsubsection{Data Distribution}
+\label{sec:datadist}
 
-Local matrix computations occur at lines \ref{line:2DsyrkH}, \ref{line:2DNEW}, \ref{line:2DsyrkW}, and \ref{line:2DNEH}.
-In the case that $\AA$ is dense, each processor performs 
-$$\frac np k^2+2\frac{m}{p_r}\frac{n}{p_c}k+\frac mp k^2+2\frac{m}{p_r}\frac{n}{p_c}k=4\frac{mnk}{p}+\frac{(m+n)k^2}{p}$$
-flops. 
-Recall that the second term on the right hand side has a constant factor of 1 instead of 2 because the local Gram computations (lines \ref{line:2DsyrkH} and \ref{line:2DsyrkW}) exploit symmetry.
-In the case that $\AA$ is sparse, processor $(i,j)$ performs $(m+n)k^2/p$ flops in computing $\M{U}_{ij}$ and $\M{X}_{ij}$, and $4\nnz(\AA_{ij})k$ flops in computing $\M{V}_{ij}$ and $\M{Y}_{ij}$.
-Local update computations occur at lines \ref{line:2DcompW} and \ref{line:2DcompH}.
-In each case, the symmetric positive semi-definite matrix is $k\times k$ and the number of columns/rows of length $k$ to be computed are $m/p$ and $n/p$, respectively.
-These costs together are given by $F(m/p,n/p,k)$.
-There are computation costs associated with the all-reduce and reduce-scatter collectives (see Section \ref{sec:collectives}), both those contribute only to lower order terms: $O(k^2+mk/p_r+nk/p_c)$.
+Given a logical processor grid of processors $P_1\times \cdots \times P_N$, we distribute the tensor $\TA$ in a block or Cartesian partition.
+Each processor owns a local tensor of dimensions $(I_1/P_1)\times \cdots \times (I_N/P_N)$, and only one copy of the tensor is stored.
+Locally, the tensor is stored linearly, with entries ordered in a natural mode-descending way that generalizes column-major layout of matrices.
+Given a processor $\V{p}=(\VE{p}{1},\dots,\VE{p}{N})$, we denote its local tensor $\TA_{\V{p}}$.
 
-\subsubsection{Communication Cost}
-\label{sec:alg:comm}
+Each factor matrix is distributed across processors in a block row partition, so that each processor owns a subset of the rows.
+We use the notation $\Mn{H}{n}_{\V{p}}$, which has dimensions $I_n/P\times R$ to denote the local part of the $n$th factor matrix stored on processor $\V{p}$.
+However, we also make use a redundant distribution of the factor matrices across processors, because all processors in a mode-$n$ processor slice need access to the same entries of $\Mn{H}{n}$ to perform their computations.
+The notation $\Mn{H}{n}_{\VE{p}{n}}$ denotes the $I_n/P_n\times R$ submatrix of $\Mn{H}{n}$ that is redundantly stored on all processors whose $n$th coordinate is $\VE{p}{n}$ (there are $P/P_n$ such processors).
 
-Communication occurs during six collective operations (lines \ref{line:2Dall-reduceH}, \ref{line:2Dall-gatherH}, \ref{line:2Dreduce-scatterAHT}, \ref{line:2Dall-reduceW}, \ref{line:2Dall-gatherW}, and \ref{line:2Dreduce-scatterWTA}).
-We use the cost expressions presented in Section \ref{sec:collectives} for these collectives.
-The communication cost of the all-reduces (lines \ref{line:2Dall-reduceH} and \ref{line:2Dall-reduceW}) is $\alpha \cdot 4\log p+\beta \cdot 2k^2$; 
-the cost of the two all-gathers (lines \ref{line:2Dall-gatherH} and \ref{line:2Dall-gatherW}) is $\alpha \cdot \log p + \beta \cdot \lt( (p_r{-}1)nk/p + (p_c{-}1)mk/p\rt)$; and
-the cost of the two reduce-scatters (lines \ref{line:2Dreduce-scatterAHT} and  \ref{line:2Dreduce-scatterWTA}) is $\alpha \cdot \log p + \beta \cdot \lt( (p_c{-}1)mk/p + (p_r{-}1)nk/p\rt)$.
+Other matrices involved in the algorithm include $\Mn{M}{n}_{\V{p}}$, which is the result of the MTTKRP computation and has the same distribution scheme as $\Mn{H}{n}_{\V{p}}$, and $\Mn{G}{n}$, which is the $R\times R$ Gram matrix of the factor matrix $\Mn{H}{n}$ and is stored redundantly on all processors.
 
-We note that \LUC\ may introduce significant communication cost, depending on the NMF algorithm used.
-The normalization of columns of $\WW$ within \HALS, for example, introduces an extra $k\log p$ latency cost.
-We will ignore such costs in our general analysis.
+\subsubsection{Inner Iteration}
 
-In the case that $m/p<n$, we choose $p_r=\sqrt{mp/n} >1$ and $p_c=\sqrt{np/m}>1$, and these communication costs simplify to $\alpha \cdot O(\log p) + \beta \cdot O(mk/p_r+nk/p_c+k^2) = \alpha \cdot O(\log p) + \beta \cdot O(\sqrt{mnk^2/p}+k^2)$.
-In the case that $m/p\geq n$, we choose $p_c=1$, and the costs simplify to $\alpha \cdot O(\log p) + \beta \cdot O(nk)$.
+The inner iteration is displayed graphically in \Cref{fig:inner} for a 3-way example and an update of the $2$nd factor matrix.
+The main idea is that at the start of the $n$th inner iteration (\Cref{fig:inner_a}), all of the data is in place for each processor to perform a local MTTKRP computation.
+This means that all processors in a slice redundantly own the same rows of the corresponding factor matrix (for all modes except $n$).
+After the local MTTKRP is computed (\Cref{fig:inner_b}), each processor has computed a contribution to a subset of the rows of the global MTTKRP $\Mn{M}{n}$, but its contribution must be summed up with the contributions of all other processors in its mode-$n$ slice.
+This summation is performed with a Reduce-Scatter collective across the mode-$n$ processor slice that achieves a row-wise partition of the result (in \Cref{fig:inner_c}, the light gray shading corresponds to the rows of $\Mn{M}{2}$ to which processor $(1,3,1)$ contributes and the dark gray shading corresponds to the rows it receives as output).
+The output distribution of the Reduce-Scatter is designed so that afterwards, the update of the factor matrix in that mode can be performed row-wise in parallel.
+Along with $\Mn{S}{n}$, which can be computed locally, each processor updates its own rows of the factor matrix given its rows of the MTTKRP result (\Cref{fig:inner_d}).
+The remainder of the inner iteration is preparing and distributing the new factor matrix data for future inner iterations, which includes an All-Gather of the newly computed factor matrix $\Mn{H}{n}$ across mode-$n$ processor slices (\Cref{fig:inner_e}) and recomputing $\Mn{G}{n}={\Mn{H}{n}}^\Tra\Mn{H}{n}$.
 
-\ignore{
-Thus, the total cost per iteration is
-$$\gamma \cdot O\lt( \frac{mnk}{p} + C_\text{BPP}\lt(k,\frac{m+n}{p}\rt) \rt) + \beta \cdot O\lt(\frac{mk}{p_r}+\frac{nk}{p_c}+k^2\rt) + \alpha \cdot O(\log p).$$
-In order to minimize the bandwidth cost of the algorithm, we choose $p_r=\sqrt{mp/n}$ and $p_c=\sqrt{np/m}$; this yields a per-iteration cost of 
-$$\gamma \cdot O\lt( \frac{mnk}{p} + C_\text{BPP}\lt(k,\frac{m+n}{p}\rt) \rt) + \beta \cdot O\lt(\sqrt{\frac{mnk^2}{p}}+k^2\rt) + \alpha \cdot O(\log p).$$
-We note that these choices of $p_r$ and $p_c$ are well defined assuming $m/p<n$ and $n/p<m$; if one of these assumptions is not satisfied we resort to Algorithm \ref{alg:1D}.
-}
+\subsubsection{Analysis}
 
-\subsubsection{Memory Requirements}
-\label{sec:new_memory}
+\begin{table*}
+\centering
+\begin{tabular}{|c|cccc|}
+\hline
+ & \textbf{Computation} & \textbf{Memory - Local MTTKRP} & \textbf{Communication} & \textbf{Memory - Par.~Algorithm}  \\
+\hline
+Par-NNCP w/ opt.~proc.~grid & $\frac{IR}{P}$ & $\frac{RI^{1/2}}{P^{1/2}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$  \\
+%\hline
+Par-NNCP w/ gen.~proc.~grid & $\frac{IR}{P}$ & $\frac{RI^{1/2}}{P^{1/2}}$ & $\mathbf{R\sum_n \frac{I_n}{P_n}}$ & $\mathbf{R\sum_n \frac{I_n}{P_n}}$  \\
+%\hline
+Par-NNCP w/o dim.~tree opt. & $\mathbf{\frac{NIR}{P}}$ & $\displaystyle \mathbf{\max_n \frac{RI/I_n}{P/P_n}}$  &  $\frac{NRI^{1/N}}{P^{1/N}}$ & $\frac{NRI^{1/N}}{P^{1/N}}$ \\
+\hline
+\end{tabular}
+\smallskip
+\caption{Leading-order per-iteration costs in terms of computation (flops), communication (words moved), and memory (words).
+We ignore constants but omit big-Oh notation for clarity.
+The first line corresponds to \cref{alg:Par-NNCP-short} with an optimal choice of processor grid and applying the dimension tree optimization locally.
+The second line corresponds to the changes in communication and memory for a general processor grid.
+The third line corresponds to the changes in computation and memory if the dimension tree optimization is not applied.}
+\label{tab:costs}
+\end{table*}
 
-The local memory requirement includes storing each processor's part of matrices $\AA$, $\WW$, and $\HH$.
-In the case of dense $\AA$, this is $mn/p+(m+n)k/p$ words; in the sparse case, processor $(i,j)$ requires $\nnz(\AA_{ij})$ words for the input matrix and $(m+n)k/p$ words for the output factor matrices.
-Local memory is also required for storing temporary matrices $\WW_j$, $\HH_i$, $\M{V}_{ij}$, and $\M{Y}_{ij}$, of size $2mk/p_r+2nk/p_c$ words.
+We will analyze the cost of a single outer iteration.
+While the number of outer iterations is sensitive to the NLS method used, the outer iteration time is generally the same across NLS methods.
+We summarize the analysis in \Cref{tab:costs}, showing the differences with and without an optimal processor grid and with and without using a dimension tree.
 
-In the dense case, assuming $k<n/p_c$ and $k<m/p_r$, the local memory requirement is no more than a constant times the size of the original data.
-For the optimal choices of $p_r$ and $p_c$, this assumption simplifies to $k<\max\lt\{\sqrt{mn/p},m/p\rt\}$. 
-Note that the second argument of the max applies when the optimal distribution is 1D ($p_r=p$).
+\paragraph{\emph{Computation}}
+The local computation occurs at \cref{line:locMTTKRP,line:hadamard,line:locNLS,line:locSYRK}.
+The cost of \cref{line:hadamard} is $O(NR^2)$, the cost of \cref{line:locNLS} is $O(R^3I_n/P)$, which is a loose upper bound for BPP and other methods \cite{KBP16}, and the cost of \cref{line:locSYRK} is $O(R(I_n/P)^2)$.
+The sum of these three costs across all inner iterations is $O(R^2N^2+(R^3/P)\sum I_n+(R/P^2)\sum I_n^2)$, which is dominated by the cost of the MTTKRP.
+When using dimension trees to perform the MTTKRP (\cref{line:locMTTKRP}), we compute the cost amortized over all inner iterations.
+In this case, the cost is dominated by the two partial MTTKRP computations (from the root of the tree), which together are $O((R/P) \prod I_n)=O(IR/P)$ and dominate the costs of the multi-TTVs.
+We note that this cost involves the product of all the tensor dimensions, which is why it dominates, and we note that it scales linearly with $P$.
 
-We note that if the temporary memory requirements become prohibitive, the computation of $((\AA \HH^T)_i)_j$ and $((\WW^T\AA)_j)_i$ via all-gathers and reduce-scatters can be blocked, decreasing the local memory requirements at the expense of greater latency costs.
-When $\AA$ is sparse and $k$ is large enough, the memory footprint of the factor matrices can be larger than the input matrix.
-In this case, the extra temporary memory requirements can become prohibitive; we observed this for a sparse data set with very large dimensions (see Section \ref{sec:webbase-2001}).
-We leave the implementation of the blocked algorithm to future work.
+\paragraph{\emph{Communication}}
+The communication within the inner iteration occurs at \cref{line:reduce-scatter,line:all-reduce,line:all-gather}.
+\Cref{line:all-reduce} involves $O(R^2)$ data and a collective across all processors.
+\Cref{line:reduce-scatter,line:all-gather} involve $O(I_nR/P_n)$ data across a subset of $P/P_n$ processors.
+Thus, the All-Reduce dominates the latency cost and the Reduce-Scatter/All-Gather dominate the bandwidth cost, for a total outer iteration communication cost of $O(R\sum I_n/P_n)$ words and $O(N\log P)$ messages.
+If the optimal processor grid can be chosen to minimize communication (assuming $P$ is sufficiently factorable), then the bandwidth cost can achieve a value of $O(NRI^{1/N}/P^{1/N})$ by making the local tensors as cubical as possible.
+We note that this cost scales with $P^{1/N}$, which is far from linear scaling.
+However, it is proportional to the geometric mean of the tensor dimensions (on the order of one tensor dimension), which is much less than the computation cost dependence on the product of all dimensions.
 
-\subsubsection{Communication Optimality}
+\paragraph{\emph{Memory}}
+The algorithm requires extra local memory to run.
+Aside from the memory required to store the local tensor of $O(I/P)$ words and factor matrices of cumulative size $O((R/P)\sum I_n)$, each processor must be able to store a redundant subset of the rows of the factor matrices it needs to perform MTTKRP computations.
+This corresponds to storing $P/P_n$ redundant copies of every factor matrix, which results in a local memory requirement of $O(R \sum I_n/P_n)$ for a general processor grid.
+The processor grid that minimizes communication also minimizes local memory, and the extra memory requirement can be as low as $O(NRI^{1/N}/P^{1/N})$, which is typically dominated by $O(I/P)$.
 
-In the case that $\AA$ is dense, Algorithm \ref{alg:2D} provably minimizes communication costs.
-Theorem \ref{thm:LB} establishes the bandwidth cost lower bound for any algorithm that computes $\WW^T\AA$ or $\AA\HH^T$ each iteration.
-A latency lower bound of $\Omega(\log p)$ exists in our communication model for any algorithm that aggregates global information \cite{CH+07}.
-For NMF, this global aggregation is necessary in each iteration, for example, in order to compute residual error in the case that $\AA$ is distributed across all $p$ processors, because all processors have data that must be accumulated into the global error.
-Based on the costs derived above, \ParNMF\ is communication optimal under the assumption $k<\sqrt{mn/p}$, matching these lower bounds to within constant factors.
-
-\begin{theorem}[\cite{DE+13}]
-\label{thm:LB}
-Let $\AA \in \Rn{m \times n}$, $\WW \in \Rn{m \times k}$, and $\HH \in \Rn{k \times n}$ be dense matrices, with $k<n\leq m$.
-If $k < \sqrt{mn/p}$, then any distributed-memory parallel algorithm on $p$ processors that load balances the matrix distributions and computes $\WW^T \AA$ and/or $\AA \HH^T$ must communicate at least $\Omega(\min\{\sqrt{mnk^2/p},nk\})$ words along its critical path.
-\end{theorem}
-\begin{proof}
-The proof follows directly from \cite[Section II.B]{DE+13}.
-Each matrix multiplication $\WW^T \AA$ and $\AA \HH^T$ has dimensions $k<n\leq m$, so the assumption $k<\sqrt{mn/p}$ ensures that neither multiplication has ``3 large dimensions.''
-Thus, the communication lower bound is either $\Omega(\sqrt{mnk^2/p})$ in the case of $p>m/n$ (or ``2 large dimensions''), or $\Omega(nk)$, in the case of $p<m/n$ (or ``1 large dimension'').
-If $p<m/n$, then $nk<\sqrt{mnk^2/p}$, so the lower bound can be written as $\Omega(\min\{\sqrt{mnk^2/p},nk\})$.
-\end{proof}
-
-We note that the communication costs of Algorithm \ref{alg:2D} are the same for dense and sparse data matrices (the data matrix itself is never communicated).
-In the case that $\AA$ is sparse, this communication lower bound does not necessarily apply, as the required data movement depends on the sparsity pattern of $\AA$.
-Thus, we cannot make claims of optimality in the sparse case (for general $\AA$).
-The communication lower bounds for $\WW^T \AA$ and/or $\AA \HH^T$ (where $\AA$ is sparse) can be expressed in terms of hypergraphs that encode the sparsity structure of $\AA$ \cite{BDKS15}.
-Indeed, hypergraph partitioners have been used to reduce communication and achieve load balance for a similar problem: computing a low-rank representation of a sparse tensor (without non-negativity constraints on the factors) \cite{KU15}.
+The dimension tree algorithm also requires extra temporary memory space, but the space required tends to be much smaller than what is required to store the local tensor.
+If the tensor dimensions can be partitioned into two parts with approximately equal geometric means, the extra memory requirement for running a dimension tree is as small as $O(R\sqrt{I/P})$, which is also typically dominated by $O(I/P)$.

### 2018-06-22 latex/toms/conclusion.tex
sc papers in toms format

@@
-\section{Conclusion}\label{sec:conclusion}
+% !TEX root = paper.tex
 
-In this paper, we propose a high-performance distributed-memory parallel framework for NMF algorithms that iteratively update the low rank factors in an alternating fashion.   
-Our parallelization scheme is designed to avoid communication overheads and scales well to over 1500 cores. 
-The framework is flexible, being (a) expressive enough to leverage many different NMF algorithms and (b) efficient for both sparse and dense matrices of sizes that span from a few hundreds to hundreds of millions.
-Our open-source software implementation is available for download.
+\section{Conclusion} \label{sec:conclusion}
 
-%For the data sets on which we experimented, we showed that an efficient implementation of a naive parallel algorithm spends much of its time in interprocessor communication.
-%For sparse problems, comparing p = 16 to p = 1536 (a factor increase of 96), we observe a speedups from \BPP\ of 59× (synthetic) and 22× (real world).
-For dense problems, comparing p = 96 to p = 1536 (a factor increase of 16), \BPP's speedup is 12× for both problems.
-The other NMF algorithms -- \MU\ and \HALS\, demonstrate similar scaling results in our framework.
-For solving data mining problems at today's scale, parallel computation and distributed-memory systems are becoming prerequisites.
-We argue in this paper that by using techniques from high-performance computing, the computations for NMF can be performed very efficiently.
-Our framework allows for the HPC techniques (efficient matrix multiplication) to be separated from the data mining techniques (choice of NMF algorithm), and we compare data mining techniques at large scale, in terms of data sizes and number of processors.
-One conclusion we draw from the empirical and theoretical observations is that the extra per-iteration cost of \BPP\ over alternatives like \MU\ and \HALS\ decreases as the number of processors $p$ increases, making \BPP\ more advantageous in terms of both quality and performance at larger scales.
-By reporting time breakdowns that separate local computation from interprocessor communication, we also see that our parallelization scheme prevents communication from bottlenecking the overall computation; our comparison with a naive approach shows that communication can easily dominate the running time of each iteration.
-%In the case of \ParNMF, the problems remain computation bound on up to 600 processors, typically spending most of the time in local matrix multiplication or NLS solves.
+In this work, we present a new implementation for distributed-memory NNCP that will be made publicly available.
+The algorithm is general enough to handle any number of modes in the data tensor and can be adapted to use any NLS algorithm within the context of BCD (ALS).
+We use a dimension tree optimization to avoid unnecessary recomputation within the bottleneck local MTTKRP computation, and we use an efficient parallelization that minimizes communication cost.
+Our performance results show the ability to scale well to high processor counts, and we show favorable performance in comparison to state-of-the-art software for 3D tensors.
 
-%We focus in this work on BPP, %which is more expensive per-iteration than alternative methods like MU and HALS, 
-%because it has been shown to reduce overall running time in the sequential case by requiring fewer iterations \cite{kim2011fast}.
-%Because much of the time per iteration of \ParNMF\ is spent on local NLS, we believe further empirical exploration is necessary 
-%to understand the proposed \ParNMF{}'s advantages for other AU-NMF algorithms such as MU and HALS. 
-%%to confirm the advantages of BPP in the parallel case.
-%We note that if we use the MU or HALS approach for determining low rank factors, the relative cost of interprocessor communication will grow, making the communication efficiency of our algorithm more important.
+In particular, the performance results demonstrate that computing NNCP for dense tensors involves heavy computation relative to the sizes of the computed factor matrices.
+By avoiding the communication of tensor entries and communicating only the factor matrices, the parallel algorithm is nearly always compute bound.
+This observation is supported by the theoretical analysis: although the communication does not scale as well with $P$, the total amount of data depends on a sum of tensor dimensions rather the product of the tensor dimensions, which determines the total amount of computation.
+For a relative comparison, consider memory-efficient parallel dense $n\times n$ matrix multiplication: the ratio of the $O(n^3/P)$ local computation to the $O(n^2/P^{1/2})$ communication is approximately the square root of the size of the local data, or $O((n^2/P)^{1/2})$.
+In the case of NNCP, the ratio of local computation to communication is $O((I/P)^{1-1/N}/N)$, which is approximately the size of the local data raised to the power $1-1/N$.
+This exponent is larger than $1/2$ and grows with $N$, and therefore it predicts the NNCP computation should be more computation bound than matrix multiplication.
+Note that this analysis does not consider the type of local computation; for small $R$, the local computation will likely be memory bandwidth bound, but the algorithm will spend more time on local computation than on interprocessor communication.
 
-In future work, we would like to extend \ParNMF\  algorithm to dense and sparse tensors, computing the CANDECOMP/PARAFAC decomposition in parallel with non-negativity constraints on the factor matrices.
-We plan on extending our software to include more NMF algorithms that fit the AU-NMF framework; these can be used for both matrices and tensors. 
-We would also like to explore more intelligent distributions of sparse matrices: while our 2D distribution is based on evenly dividing rows and columns, it does not necessarily load balance the nonzeros of the matrix, which can lead to load imbalance in matrix multiplications.
-We are interested in using graph and hypergraph partitioning techniques to load balance the memory and computation while at the same time reducing communication costs as much as possible.
-%Finally, based our survey, experiments and knowledge, we would like to conclude that, so far we are the fastest distributed NMF algorithm in the literature. 
-%Finally, we have not yet reached the limits of the scalability of \ParNMF; we would like to expand our benchmarks to larger numbers of nodes on the same size data sets to study performance behavior when communication costs completely dominate the running time.
+We can also conclude from the performance results that the dimension tree optimization is the key to performance improvement over the state-of-the-art approaches.
+For 3D tensors, we observe a benefit larger than the theoretical 50\% reduction in computation, and for larger numbers of modes, the improvement is only magnified.
+Besides the reduction in flops, the dimension tree approach enjoys better DGEMM performance and avoids memory-bound KRP computations.
+Furthermore, we see that tuning the processor grid had much less effect on overall performance.
+Not only do reductions in communication not matter as much as computation, but different local tensor sizes can also cause variations in local performance that outweigh the savings in communication.
+
+%Given an efficient parallel algorithm for NNCP, it is now possible to compare the convergence properties of various NLS algorithms on large-scale tensor data that is too large to fit on single servers or would require too much computational time to analyze interactively.
+%We hope to explore this in future work.

### 2018-06-22 latex/toms/experiments.tex
sc papers in toms format

@@
-\section{Experiments}
-\label{sec:experiment}
-\newcommand{\NLS}{LUC }
+% !TEX root = paper.tex
 
-In this section, we describe our implementation of \ParNMF\ and evaluate its performance.
-We identify a few synthetic and real world data sets to experiment with \ParNMF\ with dimensions that span from hundreds to millions. 
-We compare the performance and exploring scaling behavior of different NMF algorithms -- \MU, \HALS, and ANLS/BPP (\BPP), implemented using the parallel \ParNMF\ framework.  
-The code and the  datasets used for conducting the experiments can be downloaded from \url{https://github.com/ramkikannan/nmflibrary}.
+\section{Performance Results} \label{sec:experiments}
 
-\subsection{Experimental Setup}
+\input{plots}
 
-\subsubsection{Data Sets}\label{sec:datasets}
+\subsection{Datasets}
 
-We used sparse and dense matrices that are either synthetically generated or from real world applications. We explain the data sets in this section.
+\subsubsection{Hyperspectral Images (HSI)}
+For comparison with previous work \cite{LK+17b}, we consider the same 3D hyperspectral imaging dataset called ``Souto\_wood\_pile'' \cite{FAN16}. 
+NNCP is often used on HSI data sets for classification and blind source separation of materials with differing spectral signatures.
+The hyperspectral datacube has dimensions $1024 \times 1344 \times 33$ and represents a set of 33 grayscale images of size 1344 $\times$ 1024 pixels sampled at wavelengths 400, 410, $\dots$, 720 nm, with each pixel value representing spectral radiance in $W m^{-2} sr^{-1} nm^{-1}$.
+We also consider the Nogueir\'{o} scene dataset, which is a sequence of 9 time-lapse HSI images of the same scene acquired at about 1-hour intervals.
+In each scene, hyperspectral images were acquired at about 1-hour intervals. 
+Each Nogueir\'{o} scene HSI image has the same properties as the Souto\_wood\_pile data set, so the corresponding tensor has dimensions $1024 \times 1344 \times 33 \times 9$.
 
-\begin{itemize}
-\item Dense Synthetic Matrix: We generate a low rank matrix as the product of two uniform random matrices of size 207,360 $\times$ 100 and 100 $\times$ 138,240. 
-The dimensions of this matrix are chosen to be evenly divisible for a particular set of processor grids.  
-\item Sparse Synthetic Matrix: We generate a random sparse Erd\H{o}s-R\'{e}nyi matrix of the size 207,360 $\times$ 138,240 with density of 0.001.
-That is, every entry is nonzero with probability 0.001.
-\item Dense Real World Matrix ({\em Video}):  NMF is used on video data for background subtraction in order to detect moving objects.
-The low rank matrix $\hat{\AA} = \WW \HH$ represents background and the error matrix $\AA - \hat{\AA}$ represents moving objects.
-%Detecting moving objects has many real-world applications such as traffic estimation \cite{Fujimoto2014} and security monitoring \cite{BSJJZ2015}.
-In the case of detecting moving objects in streaming videos, the last several minutes of video is taken from the live video camera to construct the non-negative matrix. 
-An algorithm to incrementally adjust the NMF based on the streaming video is presented in \cite{kim2013nonnegative}. 
-To simulate this scenario, we collected a video in a busy intersection of the Georgia Tech campus at 20 frames per second. 
-From this video, we took video for approximately 12 minutes and  then reshaped the matrix such that every RGB frame is a column of our matrix, so that the matrix is dense with size 1,013,400 $\times$ 13,824.
-\item Sparse Real World Matrix ({\em Webbase}): This data set is a directed sparse graph whose nodes correspond to webpages (URLs) and edges correspond to hyperlinks from one webpage to another.
-The NMF output of this directed graph helps us understand clusters in graphs.
-We consider two versions of the data set: {\em webbase-1M} and {\em webbase-2001}.
-The dataset webbase-1M contains about 1 million nodes (1,000,005) and 3.1 million edges (3,105,536), and was first reported by Williams et al. \cite{Williams2009}.  
-The version webbase-2001 has about 118 million nodes (118,142,155) and over 1 billion edges (1,019,903,190); it was first reported by Boldi and Vigna  \cite{Boldi2004}.  
-Both data sets are available in the University of Florida Sparse Matrix Collection \cite{DH11} and the latter {\em webbase-2001} being the largest among the entire collection.
-\item Text data ({\em Stack Exchange}): 
-Stack Exchange is a network of question-and-answer websites on topics in varied fields, each site covering a specific topic, where questions, answers, and users are subject to a reputation award process.
-There are many Stack Exchange forums, such as {\em ask ubuntu, mathematics, latex}.
-We downloaded the latest anonymized dump of all 
-user-contributed content on the Stack Exchange network from \cite{stkxchngdataset}. 
-%Each site is formatted as a separate archive consisting of XML files zipped via 7-zip using bzip2 compression. 
-%Each site archive includes Posts, Users, Votes, Comments, PostHistory and PostLinks. 
-%For complete schema information, see the included readme.txt. 
-We used only the questions from the most popular site called Stackoverflow and did not include the answers and comments. 
-We removed the standard 571 English stop words (such as {\em are, am, be, above, below}) and then used snowball stemming available through the Natural Language Toolkit (NLTK) package \cite{LB2002}.
-After this initial pre-processing, we deleted HTML tags (such as {\em lt, gt, em}) from the posts. 
-The resulting bag-of-words matrix has a vocabulary of size 627,047 over 11,708,841 documents with 365,168,945 non-zero entries. 
-In this data, the vocabulary is larger than the typical set of English words because it includes variables, constants, and other programming constructs of various programming languages from the user questions.
-%We ran NMF on this dataset for finding 50 topics whose sample topic words are presented at the end of this section. 
-\end{itemize}
+\subsubsection{Dynamic Functional Connectivity (dFC)}
+We also consider dynamic functional connectivity datasets that are generated from fMRI images of human brains.
+Given a 4D fMRI data set of voxel measurements across multiple timesteps, voxels containing brain data are partitioned into a set of regions of interest (specified using domain-specific knowledge), and a single time-series signal is aggregated for each region of interest.
+Then, an instantaneous correlation is computed for each time point and pair of regions, and this process is repeated for a number of subjects.
+Computing a CP decomposition of this data helps to discover patterns of brain connectivity among different regions and also differentiate among individuals.
+For our representative dFC data set, we consider 246 brain regions, which yields 30{,}012 unique pairs of regions, 1200 times steps, and 500 subjects, or a tensor of dimension $30{,}012\times 1200 \times 500$ \cite{VEU+12,THBGW17}.
 
-The size of all the real world data sets were adjusted to the nearest size for uniformly distributing the matrix.
+\subsubsection{Synthetic}
+Our synthetic data sets are constructed from a CP model with an exact low rank with no added noise.
+In this case we can confirm that the residual error of our algorithm with a random start converges to zero.
+For the purposes of benchmarking, we run a fixed number of iterations of the BCD algorithm rather than using a convergence check.
 
-\subsubsection{Implementation Platform}
+%For the synthetic datasets, our open source code supports (a) low-rank tensor (b) uniform random and (c) positive shifted normal distribution of $\mathfrak{N}(3,1)$ -- that is change the mean such that all the random numbers are positive.
+We considered three different synthetic matrices for different cases.
+For baseline comparison with Liavas \cite{LK+17b}, we considered a three mode
+%uniform of size 1024x1024x1024 on processor grids $2^k \times 2^k \times 2^k$ for $k \in {0,1,2,3}$. We used the uniform five 
+%mode synthetic tensor with dimension $64\times 64\times 64\times 64\times 64$ on processor grids 
+%$1\times1\times1\times1\times1$, $2\times1\times1\times1\times1$, $\dots$, $2\times2\times2\times2\times2$ for strong scaling 
+%experiments.  In the case of weak scaling of four mode synthetic tensors with (D) and without (N) the use of dimension trees.  The 
+%tensor and processor grid dimensions are $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$.
+In all the cases the dimensions were considered such that synthetic tensors can be accommodated even on single node with 64GB for scale up plots.
 
-We conducted our experiments on ``Rhea'' at the Oak Ridge Leadership Computing Facility (OLCF).
-Rhea is a commodity-type Linux cluster with a total of 512 nodes and a 4X FDR Infiniband interconnect.
-Each node contains dual-socket 8-core Intel Sandy Bridge-EP processors and 128 GB of memory.
-Each socket has a shared 20MB L3 cache, and each core has a private 256K L2 cache.
-
-%\grey{Do we want to start using the acronym for our software and discuss it here? We will park this for camera ready}
+\subsection{Machine Details}
+The entire experimentation was performed on Eos, a supercomputer at the Oak Ridge Leadership Computing Facility. 
+Eos is a 736-node Cray XC30 cluster of Intel Xeon E5-2670 processors with a total of 47.104TB of memory. 
+Its compute nodes are organized in blades where each blade contains 4 nodes, and every node has 2 sockets with 8 physical cores and 64GB memory. 
+The machine support Intel's hyper-threading (HT), but we restricted it because HT offers minimal improvement for BLAS and LAPACK operations. 
+In total, the Eos compute partition contains 11,776 traditional processor cores and our experiments used up to 4,096 cores (35\% of the machine).
 
@@
 to promote reproducibility and reuse of our code.
-The entire C++ code is developed using the matrix library Armadillo \cite{sanderson2010}. 
-In Armadillo, the elements of the dense matrix are stored in column major order and the sparse matrices in Compressed Sparse Column (CSC) format.
-For dense BLAS and LAPACK operations, we linked Armadillo with Intel MKL -- the default LAPACK/BLAS library in RHEA. It is also easy to link Armadillo with OpenBLAS \cite{xianyi2015}. 
-We use Armadillo's own implementation of sparse matrix-dense matrix multiplication, the default GNU C++ Compiler (g++ (GCC) 4.8.2) and MPI library (Open MPI 1.8.4)  on RHEA.
-We chose the commodity cluster with open source software so that the numbers presented here are representative of common use.
+%Unlike Liavas \cite{LK+17b} that uses Eigen matrix library \cite{eigenweb}, 
+We use Armadillo \cite{sanderson2010} for matrix representation
+and operations.  
+In Armadillo, the elements of the dense matrix are stored in column major order.
+For dense BLAS and LAPACK operations, we linked Armadillo with the default LAPACK/BLAS wrappers from Cray. 
+For compiler, we use GNU C++ Compiler (g++ (GCC) 6.3.0) and MPI library is from Cray.  We could also 
+compile and run the code in Rhea the commodity cluster from OLCF with entire open source libraries such as OpenBLAS and OpenMPI.
 
-%\grey{Need to update this last phrase with compilation setup on Rhea}.
+\subsection{Comparison Implementations}
+The implementation proposed by Liavas et al.
+\cite{LK+17b} is the only publicly available distributed-memory software (of which we are aware) for computing the CP decomposition of dense tensors, with or without constraints.
+We use the acronym NbAO-NTF for Nesterov-based Alternating Optimization Nonnegative Tensor Factorization to refer to their code.
 
-\subsubsection{Algorithms}
+It is based on the same parallel algorithm as our implementation, though it is limited to 3D tensors.
+The code uses MPI collectives for communication and Eigen \cite{eigenweb} as an interface to BLAS and LAPACK.
+We compiled the code linked to BLAS/LAPACK wrappers from Cray  (the same BLAS implementation used by our code) but we were unable to run multithreaded BLAS with their code.
+For fair comparison, we use a flat MPI configuration (one MPI process per core) on all comparisons between the two implementations.
 
-In our experiments, we considered the following algorithms: 
-\begin{itemize}
-	\item \MU: \ParNMF\ (Algorithm \ref{alg:2D}) with MU (Equation \eqref{eqn:muupdate})
-	\item \HALS: \ParNMF\ (Algorithm \ref{alg:2D}) with HALS (Equation \eqref{eqn:halsupdate})
-	\item \BPP: \ParNMF\ (Algorithm \ref{alg:2D}) with BPP (Section \ref{sec:BPP})
-	\item \Naive: \NaiveAlg\ (Algorithm \ref{alg:naive}, Section \ref{sec:naive})
-\end{itemize}
+We also point out a difference between the Nesterov-based algorithm and the BPP algorithm for solving the NLS subproblems.
+The Nesterov-based algorithm attempts an acceleration step using a linear combination of the current and proposed future step; however, it re-computes the residual error before deciding whether or not to accept or reject the acceleration step.
+This residual error cannot always be computed cheaply, using the technique described in \cref{sec:error}, and it contributes significantly (approximately 25\%) to the overall run time.
+Because the BPP algorithm does not require this extra computation, and studying convergence behavior of the different NLS algorithms is beyond the scope of this work, we remove the time spent in the acceleration step of NbAO-NTF in all our comparisons.
 
-Our implementation of \Naive\ (Algorithm \ref{alg:naive}) uses BPP but can be easily to extended to \MU, \HALS, and other NMF algorithms. 
-%A detailed comparison of \NaiveAlg\ with \ParNMF\ is made in our earlier work \cite{KBP16}. 
-%We include some benchmark results from \Naive\ to reiterate the point that communication efficiency is key to obtaining reasonable performance, but we also omit other \Naive\ results in order to focus attention on comparisons among other algorithms.
+Our proposed algorithm uses dimension trees, but we also benchmark our implementation without that optimization to highlight its importance.
+We use an existing implementation to perform the individual MTTKRPs \cite{HBJT18} with this approach.
 
-For the algorithms based on \ParNMF, we use the processor grid that is closest to the theoretical optimum (see Section \ref{sec:alg:comm}) in order to minimize communication costs.
-See Section \ref{sec:procgrid} for an empirical evaluation of varying processor grids for a particular algorithm and data set.
+\subsection{Strong Scaling}
 
-%We choose these three algorithms to confirm the following conclusions from the analysis of Section \ref{sec:parNMF}: the performance of a naive parallelization of \NaiveAlg\ (Algorithm \ref{alg:naive}) will be severely limited by communication overheads, and the correct choice of processor grid within Algorithm \ref{alg:2D} is necessary to optimize performance.
-%To demonstrate the latter conclusion, we choose the two extreme choices of processor grids and test some data sets where a 1D processor grid is optimal (e.g., the Video matrix) and some where a squarish 2D grid is optimal (e.g., the Webbase matrix).
+We perform two strong scaling experiments to compare performance with NbAO-NTF.
+The experiments use a cubical synthetic tensor and the HSI image used in \cite{LK+17b}, both of which are 3D.
 
-To ensure fair comparison among algorithms, the same random seed is used across different methods appropriately. 
-That is, the initial random matrix $\HH$ is generated with the same random seed when testing with 
-different algorithms (note that $\WW$ need not be initialized). 
-In our experiments, we use number of iterations as the stopping criteria for all the algorithms.
-
-%\grey{Need to update the following paragraph for our running times, maybe consider moving it elsewhere...}
-While we would like to compare against other high-performance NMF algorithms in the literature, the only other distributed-memory implementations of which we're aware are implemented using Hadoop and are designed only for sparse matrices \cite{liao2014cloudnmf},
-\cite{liu2010distributed}, \cite{gemulla2011large}, \cite{Yin2014} and \cite{Faloutsos2014}.
-We stress that Hadoop is not designed for high performance computing of iterative numerical 
-algorithms, requiring disk I/O between steps, so a run time comparison between a Hadoop 
-implementation and a C++/MPI implementation is not a fair comparison of parallel algorithms.
-A qualitative example of differences in run time is that a Hadoop implementation of the MU algorithm on 
-a large sparse matrix of size $2^{17} \times 2^{16}$ with $2 \times {10^8}$ nonzeros (with k=8) 
-takes on the order of 50 minutes per iteration \cite{liu2010distributed}, while our MU implementation 
-takes 0.065 seconds per iteration for the synthetic data set (which is an order of magnitude larger in 
-terms of rows, columns, and nonzeros) running on only 16 nodes. 
-%Based on our survey, we are the fastest distributed NMF algorithm in the literature.
-
-% input file with plotting macros
-\input{plots.tex}
-
-\subsection{Relative Error over Time} \label{sec:convergence}
-
-There are various metrics to compare the quality of the 
-NMF algorithms \cite{kim2013nonnegative}. The most common among these metrics are (a) relative error and (b) projected 
-gradient. The former represents the closeness of the low rank approximation $\hat{\AA}\approx\WW\HH$, which is generally the optimization objective. 
-The latter 
-represent the quality of the produced low rank factors and the stationarity of the final solution. These 
-metrics are also used as the stopping criterion for terminating the iteration of the NMF algorithm as in 
-line \ref{algo:nmfloop} of Algorithm \ref{alg:aunmf}. Typically a combination of the number of iterations 
-along with improvement of these metrics until a tolerance is met is used as stopping criterion. In this paper, we use 
-relative error for the comparison as it is monotonically decreasing, as opposed to projected gradient of the 
-low rank factors, which shows oscillations over iterations. The relative error can be formally defined as 
-$\|\AA-\WW\HH\|_F/\|\AA\|_F$.
-
-In Figure \ref{fig:convergence}, we measure the relative error at the end of every iteration (i.e., after the updates of both $\WW$ and $\HH$) for all three algorithms \MU, \HALS, and 
-\BPP, and we plot the relative error over time (each mark represents an iteration).
-We consider three real world datasets, \emph{video}, \emph{stack exchange} and \emph{webbase-1M}, and set $k=50$. 
-We used only the number of iterations as stopping criterion and,  
-just for this section, ran all the algorithms for 30 iterations. 
-We note that the convergence behavior and computed factors can vary over different initializations; we used the same initial values across all three algorithms in these experiments.
-Also, we observed that for these data sets, the convergence behavior was not sensitive to initialization (the final residual errors varied by less than 1\% in our experiments).
-NMF solutions are guaranteed to be unique in certain cases, with mild assumptions on the input data \cite{BLRG14,HFS2016}, but we do not check those assumptions for these datasets.
-
-To begin with, we explain the observations on the dense \emph{video} dataset presented in Figure \ref{fig:denserwerr}. 
-The relative error of \MU\ is highest at 0.1812 after 30 iterations and \HALS's is the least with 0.1273; \BPP's relative error
-is  0.1716 after 30 iterations. 
-%For this data, although the per-iteration times vary across algorithms and \BPP\ has the longest per-iteration time, it converged more quickly than \HALS\ or \MU\ because it solves subproblems exactly.
-%From the figure, we can observe that \BPP\  error didn't change after 29 iterations where as \HALS\ and \MU\ was still improving marginally at the 4th decimal even after 30 iterations.
-
-We can observe that the relative error of \emph{stack exchange} from Figure \ref{fig:stackexchangeerr} is better 
-than \emph{webbase-1M} from Figure \ref{fig:sparserwerr} over all three algorithms. 
-In the case of the \emph{stack exchange} dataset, the relative errors after 30 iterations follow the pattern \MU\ $>$ \HALS\ $>$ \BPP, with values 0.8509, 0.8395, and 0.8377 respectively. 
-%Unlike the \emph{video} dataset, both \MU\ and \HALS\ stopped improving after 23 iterations, 
-%where as \BPP\ was still improving in the 4th decimal even though its error was better than the others. 
-However, the difference in relative error for the \emph{webbase-1M} dataset is negligible, though the relative ordering of \MU\ $>$ \HALS\ $>$ \BPP\ is consistent, with values of 0.99927 for \MU\, 0.99920 for \HALS\ and 0.99919 for \BPP.
-
-In general, for these datasets \BPP\ identifies better approximations and converges faster than \MU\ and \HALS\ despite the extra per-iteration time, which is consistent with the 
-literature \cite{kim2013nonnegative,kim2011fast}. 
-However, for the sparse datasets, the differences in relative error are small across the NMF algorithms.
-
-%The Figure \ref{fig:stackexchangerr} shows the comparison of these algorithms on the stack exchange dataset explained in Section \ref{sec:datasets}.
-
-%\captionsetup[subfigure]{labelfont=normalfont,textfont=normalfont,singlelinecheck=off,justification=raggedright}
+The performance on the cubical synthetic tensor is shown in \Cref{fig:strongsynthetic3D}. 
+We can observe from the figure that all the three algorithms scale nearly linearly as the problem remains compute bound. 
+Our algorithm (with the dimension tree optimization) achieves a speedup of $1771\times$ on 4096 cores over the same implementation running on 1 core.
+Recall from \label{sec:analysis} that the computation scales linearly with $1/P$ while the communication scales with $1/P^{1/N}=1/P^{1/3}$. 
+As is evident from the figure, the communication cost does not degrade performance even for thousands of cores. 
+Our proposed algorithm with dimension trees is 35\% faster than NbAO-NTF at 512 cores (with similar relative difference for other core counts).
+This performance improvement is due in large part to the 50\% reduction in arithmetic operations provided by the dimension tree optimization.
+There is little difference in performance between our implementation without dimension trees and NbAO-NTF.
 
 \begin{figure}
-
-% set which run to plot
-\renewcommand{\run}{3}
-
-\begin{subfigure}{0.3 \columnwidth}
-\ylabeltrue
 \begin{tikzpicture}
-\renewcommand{\datafile}{data/denserwerr-time.dat}
-\renewcommand{\run}{1}
-\relerrplot
-\renewcommand{\run}{3}
+\renewcommand{\datafile}{data/str_3D_syn.dat}
+\renewcommand{\numiterations}{1}
+\liavastrue
+\strongscalingplot
 \end{tikzpicture}
-\ylabelfalse
-\subcaption{Video}
-\label{fig:denserwerr}
-\end{subfigure}
-~
-\begin{subfigure}{0.3 \columnwidth}
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/stkx-5runs-err.dat}
-\legendtrue
-\relerrplot
-\legendfalse
-\end{tikzpicture}
-\subcaption{Stack Exchange}
-\label{fig:stackexchangeerr}
-\end{subfigure}
-~
-\begin{subfigure}{0.3 \columnwidth}
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/webbase1M-5runs-err.dat}
-\relerrplot
-\end{tikzpicture}
-\subcaption{Webbase}
-\label{fig:sparserwerr}
-\end{subfigure}
-
-\caption{Relative error comparison of \MU, \HALS, \BPP\ on real world datasets.}
-\label{fig:convergence}
+\caption{Strong scaling of 3D synthetic tensor with dimension $1024\times 1024\times 1024$ on processor grids $2^k\times 2^k\times 2^k$ for $k\in\{0,1,2,3,4\}$.  The rank is fixed at 32.}
+\label{fig:strongsynthetic3D}
 \end{figure}
 
-\subsection{Time Per Iteration}
+\Cref{fig:stronghsi3D} shows the strong scaling on the HSI data. 
+In this case, our proposed algorithm with dimension trees is over $2\times$ faster than NbAO-NTF, but part of this speedup is due to differences in the NLS update algorithms.
+For the low core count, the dimension tree provides a 60\% speedup compared to the MTTKRP time in NbAO-NTF.
+At the high core counts for this experiment, the local MTTKRP is no longer the dominating cost.
+%The proposed algorithm with classical MTTKRP was performing better on higher number of processors over the baseline. 
+%Similarly, at this stage, the SVD used for Nestrov Non-negative Least Squares on Liavas dominates the cost over MTTKRP that makes it even slower than the our proposed algorithm without dimension trees.
 
-In this section we focus on per-iteration time of all the algorithms.
-We report four types of experiments, varying the number of processors (Section \ref{sec:scaling}), the rank of the approximation (Section \ref{sec:ksweep}), the shape of the processor grid (Section \ref{sec:procgrid}), and scaling up the dataset size.
-For each experiment we report a time breakdown in terms of the overall computation and communication steps (described in Section \ref{sec:perf-breakdown}) shared by all algorithms.
-
-\subsubsection{Time Breakdown}
-\label{sec:perf-breakdown}
-
-To differentiate the computation and communication costs among the algorithms, we present the time breakdown among the various tasks within the algorithms for all performance experiments.
-For Algorithm \ref{alg:2D}, there are three local computation tasks and three communication tasks to compute each of the factor matrices:
-\begin{itemize}
-	\item \textbf{MM}, computing a matrix multiplication with the local data matrix and one of the factor matrices;
-	\item \textbf{\LUC}, local updates either using \BPP\ or applying the remaining work of the \MU\ or \HALS\ updates (i.e., the total time for both $UpdateW$ and $UpdateH$ functions); 
-	\item \textbf{Gram}, computing the local contribution to the Gram matrix;
-	\item \textbf{All-Gather}, to compute the global matrix multiplication;
-	\item \textbf{Reduce-Scatter}, to compute the global matrix multiplication;
-	\item \textbf{All-Reduce}, to compute the global Gram matrix.
-\end{itemize}
-In our results, we do not distinguish the costs of these tasks for $\WW$ and $\HH$ separately; we report their sum, though we note that we do not always expect balance between the two contributions for each task.
-Algorithm \ref{alg:naive} performs all of these tasks except Reduce-Scatter and All-Reduce; all of its communication is in All-Gather.
-
-\subsubsection{Scaling \texorpdfstring{$p$}{p}: Strong Scaling}
-\label{sec:scaling}
-
-Figure \ref{fig:scaling} presents a strong scaling experiment with four data sets: \emph{sparse synthetic}, \emph{dense synthetic}, \emph{webbase-1M}, and \emph{video}.
-In this experiment, for each data set and algorithm, we use low rank $k=50$ and vary the number of processors (with fixed problem size).
-We use $\{1,6,24,54,96\}$ nodes; since each node has 16 cores, this corresponds to $\{16,96,384,864,1536\}$ cores.
-We report average per-iteration times.
-
-We highlight three main observations from these experiments:
-\begin{enumerate}
-	\item \label{obs:1} \Naive\ is slower than all other algorithms for large $p$;
-	\item \label{obs:2} \MU, \HALS, and \BPP\ (algorithms based on \ParNMF) scale up to over 1000 processors;
-	\item \label{obs:3} the relative per-iteration cost of \LUC\ decreases as $p$ increases (for all algorithms), and therefore the extra per-iteration cost of \BPP\ (compared with \MU\ and \HALS) becomes negligible.
-\end{enumerate}
-
-\paragraph{Observation \ref{obs:1}} 
-%We report \Naive\ performance only for the synthetic data sets (Figures \ref{fig:sparsesynstrongscale} and \ref{fig:densesynscaling}); the results for the real-world data sets are similar.
-For the Sparse Synthetic data set, \Naive\ is $4.2\times$ slower than the fastest algorithm (\BPP) on 1536 processors; for the Dense Synthetic data set, \Naive\ is $1.6\times$ slower than the fastest algorithm (\MU) at that scale.
-The slowdown increases to $7.7\times$ and $3.6\times$ for the sparse and dense real-world datasets, respectively.
-Nearly all of this slowdown is due to the communication costs of \Naive.
-Theoretical and practical evidence supporting the first observation is also reported in our previous paper \cite{KBP16}.
-However, we also note that \Naive\ is the fastest algorithm for the smallest $p$ for each problem, which is largely due to reduced MM time.
-Each algorithm performs exactly the same number of flops per MM; the efficiency of \Naive\ for small $p$ is due to cache effects.
-For example, for the Dense Synthetic problem on 96 processors, the output matrix of \Naive's MM fits in L2 cache, but the output matrix of \ParNMF's MM does not; these effects disappear as $p$ increases.
-
-\paragraph{Observation \ref{obs:2}}
-Algorithms based on \ParNMF\ (\MU, \HALS, \BPP) scale well, up to over 1000 processors.
-All algorithms' run times decrease as $p$ increases, with the exception of the Sparse Real World data set, in which case all algorithms slow down scaling from $p=864$ to $p=1536$ (we attribute this lack of scaling to load imbalance).
-For sparse problems, comparing $p=16$ to $p=1536$ (a factor increase of 96), we observe speedups from \BPP\ of $59\times$ (synthetic) and $22\times$ (real world).
-For dense problems, comparing $p=96$ to $p=1536$ (a factor increase of 16), \BPP's speedup is $12\times$ for both problems.
-\MU\ and \HALS\ demonstrate similar scaling results.
-For comparison, speedups for \Naive\ were $8\times$ and $3\times$ (sparse) and $6\times$ and $4\times$ (dense).
-
-\paragraph{Observation \ref{obs:3}}
-\MU, \HALS, and \BPP\ share all the same subroutines except those that are characterized as \LUC.
-Considering only \LUC\ subroutines, \MU\ and \HALS\ require fewer operations than \BPP. However, \HALS\ has to make one additional communication for normalization of $\WW$. 
-For small $p$, these cost differences are apparent in Figure \ref{fig:scaling}.
-For example, for the sparse real world data set on 16 processors, \BPP's \LUC\ time is $16\times$ that of $\MU$, and the per iteration time differs by a factor of $4.5$.
-However, as $p$ increases, the relative time spent in \LUC\ decreases, so the extra time taken by \BPP\ has less of an effect on the total per iteration time.
-By contrast, for the dense real world data set on 1536 processors, \BPP\ spends a factor of 27 times more time in \LUC\ than \MU\ but only $11\%$ longer over the entire iteration.
-For the synthetic data sets, \LUC\ takes $24\%$ (sparse) on 16 processors and $84\%$ (dense) on 96 processors, and that percentage drops to $11\%$ (sparse) and $15\%$ (dense) on 1536 processors.
-
-These trends can also be seen theoretically (Table \ref{tab:costs}).
-We expect local computations like MM, \LUC, and Gram to scale like $1/p$, assuming load balance is preserved.
-If communication costs are dominated by the number of words being communicated (i.e., the communication is bandwidth bound), then we expect time spent in communication to scale like $1/\sqrt p$, and at least for dense problems, this scaling is the best possible.
-Thus, communication costs will eventually dominate computation costs for all NMF problems, for sufficiently large $p$.
-(Note that if communication is latency bound and proportional to the number of messages, then time spent communicating actually increases with $p$.)
-
-The overall conclusion from this empirical and theoretical observation is that the extra per-iteration cost of \BPP\ over alternatives like \MU\ and \HALS\ decreases as the number of processors $p$ increases.
-As shown in Section \ref{sec:convergence} the faster error reduction of \BPP\ typically reduces the overall time to solution compared with the alternatives even it requires more time for each iteration.
-Our conclusion is that as we scale up $p$, this tradeoff is further relaxed so that \BPP\ becomes more and more advantageous for both quality and performance.
-
-%\captionsetup[subfigure]{labelfont=normalfont,textfont=normalfont,singlelinecheck=off,justification=raggedright}
+%\textcolor{red}{Finally, we ran on 512 nodes, utilizing the 16 threads per node via parallel BLAS and OMP, for a $1024\times1024\times1024$ synthetic tensor and processor grid $8\times8\times8$.
+Here we observe a $253\times$ speedup which is the result of each node having less than 20 megabytes of data.}
 
 \begin{figure}
-
-\naivetrue
-\ksweepfalse
-\legendtrue
-
-\begin{subfigure}{\columnwidth}
-%\centering
 \begin{tikzpicture}
-\renewcommand{\datafile}{data/sparsesynstrong_scale_pgf.dat}
-\makeplot
-\labels
+\renewcommand{\datafile}{data/str_3D_HSI.dat}
+\renewcommand{\numiterations}{10}
+\liavastrue
+\strongscalingplot
 \end{tikzpicture}
-\subcaption{Sparse Synthetic}
-\label{fig:sparsesynstrongscale}
-\end{subfigure}
-\legendfalse
-
-\begin{subfigure}{\columnwidth}
-%\centering
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/densesynstrong_scale_pgf.dat}
-\makeplot
-\labels
-\end{tikzpicture}
-\subcaption{Dense Synthetic}
-\label{fig:densesynscaling}
-\end{subfigure}
-
-%\naivefalse
-
-\begin{subfigure}{\columnwidth}
-%\centering
-\renewcommand{\datafile}{data/sparserwstrong_scale_pgf.dat}
-\begin{tikzpicture}
-\makeplot
-\labels
-\end{tikzpicture}
-\subcaption{Sparse Real World (webbase-1M)}
-\label{fig:sparserwscaling}
-\end{subfigure}
-
-\begin{subfigure}{\columnwidth}
-%\centering
-\renewcommand{\datafile}{data/denserwstrong_scale_pgf.dat}
-\begin{tikzpicture}
-\makeplot
-\labels
-\end{tikzpicture}
-\subcaption{Dense Real World (Video)}
-\label{fig:denserwscaling}
-\end{subfigure}
-
-\caption{Per-iteration times with $k=50$, varying $p$ (strong scaling).}
-\label{fig:scaling}
+\caption{Strong scaling of 3D HSI real world data with dimension 1024 x1344 x 33 on processor grids of $k \times k\times 1$ for $k \in {1, 2, 4, 8, 16, 32}$. The rank is fixed at 32.}
+\label{fig:stronghsi3D}
 \end{figure}
 
-\subsubsection{Scaling \texorpdfstring{$k$}{k}}
-\label{sec:ksweep}
-
-Figure \ref{fig:ksweep} presents an experiment scaling up the low rank value $k$ from 10 to 50 with each of the four data sets.
-In this experiment, for each data set and algorithm, the problem size is fixed and the number of processors is fixed to $p=864$.
-As in Section \ref{sec:scaling}, we report the average per-iteration times.
-%We also omit \Naive\ data for the real world data sets to highlight the comparisons among \MU, \HALS, and \BPP.
-
-We highlight two observations from these experiments:
-\begin{enumerate}
-	\item \label{obs:a} \Naive\ is plagued by communication time that increases linearly with $k$;
-	\item \label{obs:b} \BPP's time increases more quickly with $k$ than those of \MU\ or \HALS;
-\end{enumerate}
-
-\paragraph{Observation \ref{obs:a}} 
-We see from the synthetic data sets (Figures \ref{fig:sparsesynksweep} and \ref{fig:densesynksweep}) that the overall time of \Naive\ increases more rapidly with $k$ than any other algorithm and that the increase in time is due mainly to communication (All-Gather).
-Table \ref{tab:costs} predicts that \Naive\ communication volume scales linearly with $k$, and we see that in practice the prediction is almost perfect with the synthetic problems.
-This confirms that the communication is dominated by bandwidth costs and not latency costs (which are constant with respect to $k$).
-We note that the communication cost of \ParNMF\ scales like $\sqrt k$, which is why we don't see as dramatic an increase in communication time for \MU, \HALS, or \BPP\ in Figure \ref{fig:ksweep}.
-
-\paragraph{Observation \ref{obs:b}}
- Focusing attention on time spent in \NLS computations, we can compare how \MU, \HALS, and \BPP\ scale differently with $k$.
- We see a more rapid increase of \NLS time for \BPP\ than \MU\ or \HALS; this is expected because the \NLS computations unique to \BPP\ require between $O(k^3)$ and $O(k^4)$ operations (depending on the data) while the unique \NLS computations for \MU\ and \HALS\ are $O(k^2)$, with all other parameters fixed.
-%It is generally observed that in both dense and sparse datasets, higher the low rank $k$ lesser the error \cite{kim2011fast}.
-For eg., the relative error of NMF algorithm with low rank $k=10$ will be more than the low rank $k=50$.
-Thus, the extra per-iteration cost of \BPP\ increases with $k$, so the advantage of \BPP\ of better error reduction must also increase with $k$ for it to remain superior at large values of $k$.
-%\grey{can we confirm this in convergence section?}
-We also note that although the number of operations within MM grows linearly with $k$, we do not observe much increase in time from $k=10$ to $k=50$; this is due to the improved efficiency of local MM for larger values of $k$.
+In \Cref{fig:strongsynthetic5D}, we benchmark performance for a 5D cubical tensor with each dimension set to 64.
+Because the tensor is 5D, we can no longer compare against NbAO-NTF.
+We see a $13-16\times$ speed up using a dimension tree over not using one for this problem.
+As predicted, the dimension tree optimization saves relatively more arithmetic for higher-order tensors.
+However, the reduction in leading order flop cost is only $2.5\times$ for $N=5$; the rest of the speedup comes from more efficient DGEMM performance and avoiding memory-bound KRP computation.  
+That is, although the flop count of KRP computation is lower order, it still contributes to the run time because it is inefficient.
+Also, for tensors with balanced dimensions, the dimension tree approach yields more favorable shapes for DGEMM.
 
 \begin{figure}
-
-\naivetrue
-\ksweeptrue
-\legendtrue
-
-\begin{subfigure}{\columnwidth}
-%\centering
-\renewcommand{\datafile}{data/sparsesyn864_ksweep_scale_pgf.dat}
 \begin{tikzpicture}
-\makeplot
-\labels
+\renewcommand{\datafile}{data/str_5D_syn.dat}
+\renewcommand{\numiterations}{10}
+\liavasfalse
+\strongscalingplot
 \end{tikzpicture}
-\subcaption{Sparse Synthetic}
-\label{fig:sparsesynksweep}
-\end{subfigure}
-\legendfalse
-
-\begin{subfigure}{\columnwidth}
-%\centering
-\renewcommand{\datafile}{data/densesyn864_ksweep_scale_pgf.dat}
-\begin{tikzpicture}
-\makeplot
-\labels
-\end{tikzpicture}
-\subcaption{Dense Synthetic}
-\label{fig:densesynksweep}
-\end{subfigure}
-
-%\naivefalse
-
-\begin{subfigure}{\columnwidth}
-%\centering
-\renewcommand{\datafile}{data/sparserwksweep_scale_pgf.dat}
-\begin{tikzpicture}
-\makeplot
-\labels
-\end{tikzpicture}
-\subcaption{Sparse Real World (webbase-1M)}
-\label{fig:sparserwksweep}
-\end{subfigure}
-
-\begin{subfigure}{\columnwidth}
-\renewcommand{\datafile}{data/denserw_ksweep_scale_pgf.dat}
-\begin{tikzpicture}
-\makeplot
-\labels
-\end{tikzpicture}
-\caption{Dense Real World (Video)}
-\label{fig:denserwksweep}
-\end{subfigure}
-
-\caption{Per-iteration times with $p=864$, varying low rank $k$.}
-\label{fig:ksweep}
+\caption{Strong scaling of 5D synthetic tensor with dimension $64\times 64\times 64\times 64\times 64$ on processor grids $1\times1\times1\times1\times1$, $2\times1\times1\times1\times1$, $\dots$, $2\times2\times2\times2\times2$.
+The rank is fixed at 32.}
+\label{fig:strongsynthetic5D}
 \end{figure}
 
-\subsubsection{Varying Processor Grid}
-\label{sec:procgrid}
+\subsection{Weak Scaling Time Breakdown}
 
-In this section we demonstrate the effect of the dimensions of the processor grid on per-iteration performance.
-%As described in Section \ref{sec:alg:comm}, Algorithm \ref{alg:2D} is correct for any $p_r$ and $p_c$, but 
-For a fixed total number of processors $p$, the communication cost of Algorithm \ref{alg:2D} varies with the choice of $p_r$ and $p_c$.
-To minimize the amount of data communicated, the theoretical analysis suggests that the processor grid should be chosen to make the sizes of the local data matrix as square as possible.
-This implies that if $m/p > n$, $p_r=p$ and $p_c=1$ is the optimal choice (a 1D processor grid); likewise if $n/p > m$ then a 1D processor grid with $p_r=1$ and $p_c=p$ is the optimal choice.
-Otherwise, a 2D processor grid minimizes communication with $p_r \approx \sqrt{mp/n}$ and $p_c \approx \sqrt{np/m}$ (subject to integrality and $p_rp_c=p$).
-
-Figure \ref{fig:procsweep} presents a benchmark of \BPP\ for the Sparse Synthetic data set for fixed values of $p$ and $k$.
-We vary the processor grid dimensions from both 1D grids to the 2D grid that matches the theoretical optimum exactly.
-Because the sizes of the Sparse Synthetic matrix are $172{,}800\times115{,}200$ and the number of processors is 1536, the theoretically optimal grid is $p_r = \sqrt{mp/n} = 48$ and $p_c = \sqrt{np/m} = 32$.
-The experimental results confirm that this processor grid is optimal, and we see that the time spent communicating increases as the processor grid deviates from the optimum, with the 1D grids performing the worst.
+We also perform a weak scaling experiment to understand the time it takes to solve bigger problems with more processors.
+In this experiment, we use a synthetic 4D tensor and keep the amount of tensor data assigned to each processor constant, with tensor and processor grid of dimensions $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$ and the rank fixed at 32.
+%For eg., in the case of 3 x 3 x 3 x 3 processor grid with 81 nodes, we ran the proposed algorithm with and without dimension trees on a four mode tensor of size 384 x 384 x 384 x 384. 
+The results of the breakdown plot is shown in \Cref{fig:weaksynthetic4D}. 
+In this case, the algorithm is compute bound with and without the use of the dimension tree, so the total time of the weak scaling remains fixed for both cases. 
+Using a dimension tree, the time is completely dominated by the MTTKRP computation.  
+Without using a dimension tree, we observe that the KRP is expensive and yields a $2.5\times$ slower total run time even in the 4D case.
 
 \begin{figure}
-\renewcommand{\datafile}{data/sparsesynpsweep_scale_pgf.dat}
-\centering
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/wk_4D_syn.dat}
+\renewcommand{\numiterations}{10}
+\weakthreeDfalse
+\breakdownplot
+%\labels
+\end{tikzpicture}
+\caption{Weak scaling of 4D synthetic tensors with (D) and without (N) the use of dimension trees.
+The tensor and processor grid dimensions are $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$, and the rank is fixed at 32.
+The reported times are per iteration.}
+\label{fig:weaksynthetic4D}
+\end{figure}
+
+\subsection{Varying Processor Grid}
+
+In order to illustrate the effect of processor grid choice on running time, we show in \Cref{fig:commsweep10} a time breakdown over various processor grid choices for a 4D problem on 81 processors.
+Because the tensor is cubical and 81 has a restricted factorization into 4 numbers, there are 5 distinct processor grids.
+The overall takeaway is that the processor grid has very little effect on running time; in this experiment there is less than 10\% variation in overall time.
+While the optimal processor grid reduced the communication time by approximately $3\times$ compared to the other processor grids, the running time is dominated by local computation, so it had little effect on overall time.
+Furthermore, adjusting the processor grid affects the local tensor dimensions and the performance of the local computations, and the optimal processor grid led to slower local performance.
+For $R=10$, all of the local computation is memory bandwidth bound, and we believe the variations in running times to be effects of some temporary quantities fitting into smaller levels of cache.
+
+\begin{figure}
+\renewcommand{\datafile}{data/comm_sweep_scale_pgf_10.dat}
+\renewcommand{\numiterations}{10}
 \begin{tikzpicture}
@@
 	ybar stacked,
-	reverse legend,
-	bar width=12pt,
-	width=7cm, height=5cm,
-	ylabel={Time (seconds)}, 
-	y label style={yshift=-0.5cm},
-	x label style={yshift=-0.5cm},
+	bar width=8pt,
+	width=\columnwidth,
+	height =.75\columnwidth,
+	%width=9cm, height=3.85cm,
+	ylabel={Time (s)}, 
+	xlabel={Processor Grid},
+	y label style={yshift=-.5cm},
 	ymin=0,
-	symbolic x coords={1-1536,8-192,16-96,32-48,48-32,96-16,192-8,1536-1},
-	xticklabels={$1{\times}1536$,$8{\times}192$,$16{\times}96$,$32{\times}48$,$48{\times}32$,$96{\times}16$,$192{\times}8$,$1536{\times}1$},
-	xtick=data,
-	xticklabel style={xshift=.1cm,rotate=45,anchor=east},
-	xlabel={Processor Grid}, 
-	%legend style={draw=none,row sep=-0.1cm},
-	%legend style={at={(1,.5)},anchor=west}
-	legend style={at={(0.5,1.2)},anchor=north},
+	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
+	symbolic x coords={81x1x1x1-10, 27x3x1x1-10,9x9x1x1-10,9x3x3x1-10,3x3x3x3-10,},
+	%xtick=data,
+	xticklabels={,81x1x1x1,27x3x1x1,9x9x1x1,9x3x3x1,3x3x3x3},
+	legend style={at={(0.5,1.3)},anchor=north},
 	legend columns=-1,
-	legend style={draw=none, cells={align=left}, nodes={scale=0.8}},
 ]
 	\setcolors
-	\addplot table[x=pg, y expr=(\thisrow{mm}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=pg, y expr=(\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=pg, y expr=(\thisrow{gram}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=pg, y expr=(\thisrow{allgather}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=pg, y expr=(\thisrow{reducescatter}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=pg, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
-	\legend{MM, \NLS, Gram, All-Gather, Reduce-Scatter, All-Reduce};
+	\addplot table[x=algo-k, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=algo-k, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=algo-k, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=algo-k, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=algo-k, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
+	\legend{PM,mTTV,NLS,Factor Comm,Gram Comm};
 \end{axis}
+%\labels
 \end{tikzpicture}
-\caption{Tuning processor grid for \BPP\ on Sparse Synthetic data set with $p=1536$ and $k=50$.}
-\label{fig:procsweep}
+\caption{Time breakdown for $243\times243\times243\times243$ tensor and rank $R=10$ on 81 processors for varying processor grids.}
+\label{fig:commsweep10}
 \end{figure}
 
-\subsubsection{Scaling up to Very Large Sparse Datasets}\label{sec:webbase-2001}
+\subsection{Varying Approximation Rank}
 
-In this section, we test \ParNMF\ by scaling up the problem size.
-While we've used \emph{webbase-1M} in previous experiments, we consider \emph{webbase-2001} in this section as it is the largest sparse data in University of Florida Sparse Matrix Collection \cite{DH11}.
-The former dataset has about 1 million nodes and 3 million edges, whereas the latter dataset has over 100 million nodes and  1 billion edges (see Section \ref{sec:datasets} for more details).
-Not only is the size of the input matrix increased by two orders of magnitude (because of the increase in the number of edges), but also the size of the output matrices is increased by two orders of magnitude (because of the increase in the number of nodes).
+One of the challenges of the CP (and NNCP) decomposition in practice is the choice of decomposition rank.
+The most common technique is to compute multiple CP decompositions for various ranks.
+As the rank $R$ increases, the approximation error  $\|\TA - \T{M}\|$ decreases with the better approximation power of more parameters. 
+However, the benefit of increasing $R$ eventually diminishes if the data can be well approximated with a CP model.
+%that an increase in $R$ in lower values, say from 5 to 10, will have significant improvement in relative error over increase in higher 
+%values, like 100 to 105. 
+%Hence, it is common practice in the community to sweep $k$, to obtain better approximation error within the 
+%manageable computation. 
+Towards this end, we experiment with various values of $R$ to observe the relative increase in running time for two real-world data sets.
 
-In fact, with a low rank of $k=50$, the size of the output matrices dominates that of the input matrix: $\WW$ and $\HH$ together require a total of 88 GB, while $\AA$ (stored in compressed column format) is only 16 GB.
-At this scale, because each node (consisting of 16 cores) of Rhea has 128 GB of memory, multiple nodes are required to store the input and output matrices with room for other intermediate values.
-As mentioned in Section \ref{sec:new_memory}, \ParNMF\ requires considerably more temporary memory than necessary when the output matrices require more memory than the input matrix.
-While we were not limited by this property for the other sparse matrices, the \emph{webbase-2001} matrix dimensions are so large that we need the memories of tens of nodes to run the algorithm.
-Thus, we report results only for the largest number of processors in our experiments: 1536 processors (96 nodes).
-The extra temporary memory used by \ParNMF\ is a latency-minimizing optimization; the algorithm can be updated to avoid this extra memory cost using a blocked matrix multiplication algorithm.
-The extra memory can be reduced to a negligible amount at the expense of more messages between processors and synchronizations across the parallel machine.
-%We have not yet implemented this update.
+\Cref{fig:ksweep4DHSI} shows the time breakdown of our implementation using a dimension tree on the 4D HSI dataset for $R=\{10,\dots,50\}$. 
+We observe an overall time increase with increased $R$, but each part of the computation scales slightly differently.
+The multi-TTV computation ({\em mTTV}) scales linearly with the increasing $R$,  whereas the partial MTTKRP ({\em PM}) is scaling super-linearly. 
+This is because mTTV is cast as matrix-vector products (DGEMV) and PM is cast as matrix-matrix products (DGEMM).
+As $R$ increases from 10 to 50, DGEMM performance improves but DGEMV performance is constant.  
+The local NLS time is increasing with $O(R^3)$ as expected and the All-Reduce required for the Gram matrices scales with $O(R^2)$, becoming a significant cost for larger $R$.
 
-%In the real-world, it is difficult to acquire very large dense datasets. Most of the very large dense 
-%datasets will be either be image/video data or collected out of very large scientific experiments. It is 
-%common in the DM/ML community to have very large sparse dataset. In this section, to benefit the Data 
-%mining and machine learning community, we report the scalability of \ParNMF\ on very directed sparse 
-%webgraph over 118 million nodes and nearly 1 billion edges.  In Section \ref{sec:datasets},  we 
-%discussed the details of this {\em webbase-2001} dataset.
-
-%It can be observed from our \ParNMF\ that we hold the dense matrices $\WW, \HH, (\WW_i)_j, 
-%(\HH_j)_i, \WW^T\AA$ and $\AA \HH^T$ in memory for computation. As discussed in Section 
-%{\ref{sec:memory}}, the estimated per process memory footprint of all the matrices together is $O(xx)$. In the 
-%case of {\em Webbase-2001},  matrix of size 118,000,000 with low rank $k=$50 over 864 process, every 
-%process  required xx GB and approximately 2GB for input sparse matrix.
-\ramki{Grey can we discuss to fill this up.
-I wanted this to be consistent with the table and the section.} We were running 16 process on every node, the estimated memory per-process along with temporary allocations,  the experiment couldn't run in 54 nodes with each node having 128GB.
-Hence, we are reporting the wall clock time for different tasks and the error plot for {\em Webbase-2001} over 96 nodes and 1536 processors in Figure \ref{fig:webbase2001}.
-
-We present results for \emph{webbase-2001} in Figure \ref{fig:webbase2001}.
-The average per-iteration timing results are consistent with the observations from other synthetic and real world sparse datasets as discussed in Section \ref{sec:scaling}, though the raw times are about 2 orders of magnitude larger, as expected.
-In the case of the error plot, as observed in other experiments, \BPP\ achieves smaller error (by 1\%) than other algorithms after converging; however \MU\ and \HALS\ initially outperform \BPP. 
-We also see that \MU\ outperforms \HALS\ in the first 30 iterations. 
-At the 30th iteration, the error for \HALS\ is still improving at the third decimal, whereas \MU's is improving at the fourth decimal. 
-We suspect that over a greater number of iterations the error of \HALS\ could become smaller than that of \MU, which would be more consistent with other datasets.
+In \Cref{fig:ksweepneuro} we compare performance for various ranks $R$ across all 3 algorithms, again used flat MPI. 
+Starting at $R=10$ we see the largest speed up of $2\times$ for our implementation with a dimension tree over NbAO-NTF. 
+This is due to a combination of the dimension tree performing fewer flops in the MTTKRPs and KRPs. 
+However, as the rank increases this speed up diminishes to $1.6\times$. 
+The loss of speed up is a result of the fact that, as we observed in \Cref{fig:ksweep4DHSI}, the multi-TTV operations do not scale as well as the partial-MTTKRPs for increasing $R$.
+Again, the performance of our implementation without the dimension tree optimization is comparable to NbAO-NTF.
 
 \begin{figure}
-\centering
-\captionsetup[subfigure]{justification=centering}
-\begin{subfigure}{0.40 \columnwidth}
-\renewcommand{\datafile}{data/webbase118m_pgf.dat}
-%\centering
+\renewcommand{\datafile}{data/ksw_4D_HSI.dat}
+\renewcommand{\numiterations}{10}
 \begin{tikzpicture}
@@
 	ybar stacked,
-	reverse legend,
-	bar width=16pt,
-	width=1.6in, height=1.8in,
-	ylabel={Time (seconds)}, 
-	y label style={yshift=-.5cm, xshift=-.5cm},
-	x label style={yshift=-.5cm},
+	bar width=8pt,
+	width=\columnwidth,
+	height =.75\columnwidth,
+	%width=9cm, height=3.85cm,
+	ylabel={Time (s)}, 
+	xlabel={Rank $R$},
+	y label style={yshift=-.5cm},
 	ymin=0,
-	symbolic x coords={1536-0,1536-1,1536-2},
-	xticklabels={MU,HALS,ABPP},
-	xtick=data,
-	%xticklabel style={xshift=.15cm,rotate=45,anchor=east},
-	%xlabel={NMF Algorithms}, 
-	%legend style={draw=none,row sep=-0.1cm},
-	%legend style={at={(1,.5)},anchor=west}
+	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
+	symbolic x coords={D10,D20,D30,D40,D50},
+	%xtick=data,
+	xticklabels={,10,20,30,40,50},
 	legend style={at={(0.5,1.3)},anchor=north},
-	legend columns=3,
-	legend style={draw=none, cells={align=left}, nodes={scale=0.5}}
+	legend columns=-1,
 ]
 	\setcolors
-	\addplot table[x=algo, y expr=(\thisrow{mm}/30)] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{nnls}/30)] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{gram}/30)] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{allgather}/30)] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{reducescatter}/30)] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{allreduce}/30)] {\datafile};
-	\legend{MM, \NLS, Gram, All-Gather, Reduce-Scatter, All-Reduce};
+	\addplot table[x=alg-K, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-K, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-K, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-K, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-K, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
+	\legend{PM,mTTV,NLS,Factor Comm,Gram Comm};
+\end{axis}
+%\labels
+\end{tikzpicture}
+\caption{Per-iteration time breakdown of our implementation (using dimension trees) over various ranks for a time-lapse HSI dataset with dimensions $1344\times 1024\times 33 \times 9$ on 64 processors arranged in a $8\times8\times1\times1$ grid.}
+\label{fig:ksweep4DHSI}
+\end{figure}
+
+\begin{figure}
+\begin{tikzpicture}
+\begin{axis}[legend style={draw=none, anchor=north,  cells={align=left}, nodes={scale=0.7}}, legend pos=north west, xlabel=Low Rank $R$, ylabel=Time(s), ylabel near ticks]
+	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
+	\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
+	\addplot+ [discard if not={alg}{L},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
+	\legend{DimTree,NoDimTree,NbAO-NTF}
 \end{axis}
 \end{tikzpicture}
-\subcaption{Time}
-\end{subfigure}
-~
-\begin{subfigure}{0.45 \columnwidth}
-\begin{tikzpicture}
-\renewcommand{\run}{1}
-\legendtrue
-\ylabeltrue
-\widertrue
-\renewcommand{\datafile}{data/webbase118m-err-time.dat}
-\relerrplot
-\widerfalse
-\end{tikzpicture}
-\subcaption{Error}
-\end{subfigure}
-\caption{NMF comparison on \emph{webbase-2001} for $k{=}50$ on 1536 processors.}
-\label{fig:webbase2001}
+\caption{Overall running time for dFC dataset with dimensions $30{,}012\times 1200 \times 500$ on 1440 cores arranged in a $2\times 6\times 120$ processor grid and varying choices of rank $R$.}
+\label{fig:ksweepneuro}
 \end{figure}
 
-\subsection{Interpretation of Results}
+%\begin{figure}
 
-We present results from two of the real world datasets in the Supplemental Material. 
-The first example shows background separation of the \emph{video} data, and the second example shows topic modeling output on the \emph{stack exchange} text dataset.
-The details of these datasets are presented in Section \ref{sec:datasets}.
+% set which run to plot
+%\renewcommand{\run}{3}
+%
+%\begin{subfigure}{0.3 \columnwidth}
+%\ylabeltrue
+%\begin{tikzpicture}
+%\renewcommand{\datafile}{data/denserwerr-time.dat}
+%\renewcommand{\run}{1}
+%\relerrplot
+%\renewcommand{\run}{3}
+%\end{tikzpicture}
+%\ylabelfalse
+%\subcaption{Video}
+%\label{fig:denserwerr}
+%\end{subfigure}
+%~
+%\begin{subfigure}{0.3 \columnwidth}
+%\begin{tikzpicture}
+%\renewcommand{\datafile}{data/stkx-5runs-err.dat}
+%\legendtrue
+%\relerrplot
+%\legendfalse
+%\end{tikzpicture}
+%\subcaption{Stack Exchange}
+%\label{fig:stackexchangeerr}
+%\end{subfigure}
+%~
+%\begin{subfigure}{0.3 \columnwidth}
+%\begin{tikzpicture}
+%\renewcommand{\datafile}{data/webbase1M-5runs-err.dat}
+%\relerrplot
+%\end{tikzpicture}
+%\subcaption{Webbase}
+%\label{fig:sparserwerr}
+%\end{subfigure}
+%
+%\caption{Relative error comparison of \MU, \HALS, \BPP\ on real world datasets.}
+%\label{fig:convergence}
+%\end{figure}
 
-While the literature covers more detail about fine tuning NMF and different NMF variants for higher quality results on these two tasks, our main focus is to show how quickly we can produce baseline NMF solutions.
-In Figure 1 of the Supplemental Material, we can see the background is removed and the moving objects (e.g., cars) are visible. 
-Similarly, Table 1 of Supplemental Material shows that the NMF solution discriminates among topics and and finds coherent keywords for each topic.
-
-%\subsubsection{Moving Object Detection of Surveillance Video Data}
+% set which run to plot
+%\renewcommand{\run}{3}
 %
-%As explained in the Section 6.1.1, we processed 12 minutes video that is captured from a 
-%busy junction in Georgia Tech to separate the background and moving objects from this video. 
-%In Figure \ref{fig:videoresults}, we present some sample frames to compare the input image with the separated background and moving objects.
-%The background are the results of the low rank approximation 
-%$\hat{\AA}=\WW\HH$ output yielded from our \ParNMF\ algorithm and the moving objects are given by $\AA-\hat{\AA}$. 
-%We can clearly see the background remains static and the moving objects (e.g., cars) are visible. 
+%\begin{subfigure}{0.3 \columnwidth}
+%\ylabeltrue
+%\begin{tikzpicture}
+%\renewcommand{\datafile}{data/denserwerr-time.dat}
+%\renewcommand{\run}{1}
+%\strongscalingplot
+%\renewcommand{\run}{3}
+%\end{tikzpicture}
+%\ylabelfalse
+%\subcaption{Video}
+%\label{fig:denserwerr}
+%\end{subfigure}
+%~
+%\begin{subfigure}{0.3 \columnwidth}
+%\begin{tikzpicture}
+%\renewcommand{\datafile}{data/stkx-5runs-err.dat}
+%\legendtrue
+%\strongscalingplot
+%\legendfalse
+%\end{tikzpicture}
+%\subcaption{Stack Exchange}
+%\label{fig:stackexchangeerr}
+%\end{subfigure}
+%~
+%\begin{subfigure}{0.3 \columnwidth}
+%\begin{tikzpicture}
+%\renewcommand{\datafile}{data/webbase1M-5runs-err.dat}
+%\strongscalingplot
+%\end{tikzpicture}
+%\subcaption{Webbase}
+%\label{fig:sparserwerr}
+%\end{subfigure}
 %
-%\subsubsection{Topic Modeling of Stack Exchange Data}
-%We ran 
-%our \ParNMF\ algorithm on this dataset, which has nearly 12 million questions from the Stack Overflow site (under Stack Exchange) to produce 50 topics. 
-%The matrix $\WW$ can be interpreted as {\em vocabulary-topic} distribution and the 
-%$\HH$ as {\em topic-document} distribution.  We took the top 5 words for each of the 50 topics and 
-%present them in Table \ref{tab:stackexchangetopics}. Typically a good topic generation satisfies 
-%properties such as (a) finding discriminative rather than common words -- capturing words that can provide 
-%some information; (b) finding different topics -- the similarity between different topics should be low; 
-%(c) coherence - all the words that belong to one topic should be coherent.  There are some topic quality 
-%metrics \cite{NLGB2010} that capture the usefulness of topic generation algorithm.  We can  see
-% NMF generated generally high-quality and coherent topics. 
-% Also, each of the topics are from different domains such as 
-% databases, C/C++ programming, Java programming, and web technologies like PHP and HTML.
+%\caption{Relative error comparison of \MU, \HALS, \BPP\ on real world datasets.}
+%\label{fig:convergence}
+%\end{figure}

### 2018-06-22 latex/toms/survey.tex
sc papers in toms format

@@
-\section{Related Work}\label{sec:related}
+% !TEX root = paper.tex
 
-In the data mining and machine learning literature there is an overlap between low rank approximations and matrix factorizations due to the nature of applications. 
-Despite its name, non-negative matrix ``factorization'' is really a low rank approximation. 
-Recently there is a growing interest in collaborative filtering based 
-recommender systems. One of the popular techniques
-for collaborative filtering is matrix factorization, often with nonnegativity constraints, 
-and its implementation is widely available in many
-off-the-shelf distributed machine learning libraries
-such as GraphLab \cite{low2012}, MLLib \cite{meng2015mllib},
-and many others \cite{satish2014,yun2014} as well.
-However, we would like to clarify that collaborative
-filtering using matrix factorization is a different problem than NMF: 
-in the case of collaborative filtering, non-nonzeros in the matrix 
-are considered to be missing entries, while in the case of NMF, non-nonzeros in the matrix correspond to true zero values.
+\section{Related Work} 
+\label{sec:survey}
 
-There are several recent distributed NMF algorithms in the literature \cite{liao2014cloudnmf,Faloutsos2014,Yin2014,liu2010distributed}. 
-Liu et al.\ propose running Multiplicative Update (MU) for KL divergence, squared loss, and ``exponential'' loss functions \cite{liu2010distributed}. 
-Matrix multiplication, element-wise multiplication, and element-wise division are the building blocks of the MU algorithm. 
-The authors discuss performing these matrix operations effectively in Hadoop for sparse matrices. 
-Using similar approaches, Liao et al.\ implement an open source Hadoop-based MU algorithm and study its scalability on large-scale biological data sets \cite{liao2014cloudnmf}. 
-Also, Yin, Gao, and Zhang present a scalable NMF that can perform frequent updates, which aim to use the most recently updated data \cite{Yin2014}. 
-Similarly Faloutsos et al.\ propose a distributed, scalable method for decomposing matrices, tensors, and coupled data sets through stochastic gradient descent on a variety of objective functions \cite{Faloutsos2014}.
-The authors also provide an implementation that can enforce non-negative constraints on the factor matrices. 
-All of these works use Hadoop to implement their algorithms.
+The formulation of NNCP with least squares error and algorithms for computing it go back to \cite{Paatero97,WW01}, developed in part as a generalization of nonnegative matrix factorization algorithms \cite{LS99} to tensors.
+Sidiropoulos et al.
+\cite{SLFHPF2017} provide a more detailed and complete survey that includes basic tensor factorization models with and without constraints, broad coverage of algorithms, and recent driving applications.
+The tensor operations discussed and the notation used in this paper follow Kolda and Bader's survey \cite{KB2009}.
 
-We emphasize that our MPI-based approach has several advantages over Hadoop-based approaches:
-\begin{itemize}
-	\item efficiency -- our approach maintains data in memory, never communicating the data matrix, while Hadoop-based approaches must read/write data to/from disk and involves global shuffles of data matrix entries;
-	\item generality -- our approach is well-designed for both dense and sparse data matrices, whereas Hadoop-based approaches generally require sparse inputs;
-	\item privacy -- our approach allows processors to collaborate on computing an approximation without ever sharing their local input data (important for applications involving sensitive data, such as electronic health records), while Hadoop requires the user to relinquish control of data placement.
-\end{itemize}
+Recently, there has been growing interest in scaling tensor operations to bigger data and more processors in both the data mining/machine learning and the high performance computing communities. 
+For sparse tensors, there have been parallelization efforts to compute CP decompositions both on shared-memory platforms \cite{SRSK2015,LCPSV17} as well as distributed-memory platforms \cite{KU16,SK16,KU18}, and these approaches can be generalized to constrained problems \cite{SBK2017}.
+The focus of this work is on dense tensors, but many of the ideas for sparse tensors are applicable to the dense case, including parallel data distributions, communication pattern, and techniques to avoid recomputation across modes.
 
-We note that Spark \cite{ZCFSS10} is a popular big-data processing infrastructure that is generally more efficient for iterative algorithms such as NMF than Hadoop, as it maintains data in memory and avoids file system I/O.
-Even with a Spark implementation of previously proposed Hadoop-based NMF algorithm, we expect performance to suffer from expensive communication of input matrix entries, and Spark will not overcome the shortcomings of generality and privacy of the previous algorithms.
-Although Spark has collaborative filtering libraries such as MLlib \cite{meng2015mllib}, which use matrix factorization and can impose non-negativity constraints, none of them implement pure NMF, and so we do not have a direct comparison against NMF running on Spark.
-As mentioned above, the problem of collaborative filtering is different from NMF, and therefore different computations are performed at each iteration.
+In particular, Liavas et al. \cite{LK+17b} extend a parallel algorithm designed for sparse tensors \cite{SK16} to the 3D dense case.
+They use the ``medium-grained'' dense tensor distribution and row-wise factor matrix distribution, which is exactly the same as our distribution strategy (see \cref{sec:datadist}), and they use a Nesterov-based algorithm to enforce the nonnegativity constraints.
+Their code is publicly available, and we compare our performance with theirs in \cref{sec:experiments}.
+A similar data distribution and parallel algorithm for computing a single dense MTTKRP computation is proposed by Ballard, Knight, and Rouse \cite{BKR17-TR}. 
+They prove that the algorithm is communication optimal, but they do not provide an implementation.
+Another approach to parallelizing NNCP decomposition of dense tensors is presented by Phan and Cichocki \cite{PC11}, but they use a dynamic tensor factorization, which performs different, more independent computations across processors.
 
-%Apart from the distributed implementation for NMF, we would also like to highlight few recent theoretical highlights with NMF under some mild assumptions. 
-%Arora, Ge, Kannan and Moitra \cite{AGKM2012}, proposed a initial polynomial time running algorithm for topic modeling with NMF when there exists  anchor words, words that appear (with positive probability) in only one topic.
-%Later, Huang, Sidiropoulos and Swami \cite{HSS2016} improved up on these results under milder conditions using second-order moments. 
-%The uniqueness of NMF has been of interest to the community and some recent results for both symmetric and non-symmetric matrices have been obtained using a geometrical point of view in \cite{HSS2013, HSS2014}.
+The idea of using dimension trees (discussed in \cref{sec:dimtrees}) to avoid recomputation within MTTKRPs across modes is introduced in \cite{PTC13a} for computing the CP decomposition of dense tensors.
+It has also been used for sparse CP \cite{LCPSV17,KU18} and other tensor computations \cite{KU16}.
 
-Fairbanks et al. \cite{Fairbanks2015} present a parallel NMF algorithm designed for multicore machines.  
-To demonstrate the importance of minimizing communication, we consider this approach to parallelizing an alternating-updating NMF algorithm in distributed memory (see Section \ref{sec:naive}).
-While this naive algorithm exploits the natural parallelism available within the alternating iterations (the fact that rows of $\WW$ and columns of $\HH$ can be computed independently), it performs more communication than necessary to set up the independent problems.
-We compare the performance of this algorithm with our proposed approach to demonstrate the importance of designing algorithms to minimize communication; that is, simply parallelizing the computation is not sufficient for satisfactory performance and parallel scalability.
+%Given these recent work on the sparse tensors, we survey the literature around dense tensor -- the interest of this paper.
+Unlike sparse tensor, there are very few activities around dense tensors that are important for analyzing tensors from images and scientific world such as 2D Ronchigram from pytography data.
+With the advent of the deep learning and the interest of the community in processing higher order tensors from hyper spectral image and medical imaging, we have used these real world in conducting experiments of this paper.
+%
+%Austin, Ballard, and Kolda \cite{ABK16} distributed-memory parallel implementation for the Tucker decomposition to compress 8TB of scientific data.
+Tucker decomposition is unconstrained and a completely different decomposition over constrained Non-negative Tensor Factorization explained in this paper.
+Recently,  Ballard, Knight, Rouse \cite{BKR17-TR} have proposed Communication Lower Bounds for Matricized Tensor Times Khatri-Rao Product under particular distribution of dense tensors.
+The proposed algorithm \ref{alg:2D} is motivated out of this work.
+%
+%With this detailed survey, we would like to highlight the contribution of this paper.
 
-Apart from distributed NMF algorithms using Hadoop and multicores, there are also implementations of the
-MU algorithm in a distributed memory setting using X10 \cite{Grove2014} and on a GPU \cite{mejia2015nmf}.
+%\subsection{Contributions:}

### 2018-08-22 latex/toms/experiments.tex
pgfplots not working

@@
 
-%\begin{figure}
+\begin{figure}
 
-% set which run to plot
-%\renewcommand{\run}{3}
-%
-%\begin{subfigure}{0.3 \columnwidth}
-%\ylabeltrue
-%\begin{tikzpicture}
-%\renewcommand{\datafile}{data/denserwerr-time.dat}
-%\renewcommand{\run}{1}
-%\relerrplot
-%\renewcommand{\run}{3}
-%\end{tikzpicture}
-%\ylabelfalse
-%\subcaption{Video}
-%\label{fig:denserwerr}
-%\end{subfigure}
-%~
-%\begin{subfigure}{0.3 \columnwidth}
-%\begin{tikzpicture}
-%\renewcommand{\datafile}{data/stkx-5runs-err.dat}
-%\legendtrue
-%\relerrplot
-%\legendfalse
-%\end{tikzpicture}
-%\subcaption{Stack Exchange}
-%\label{fig:stackexchangeerr}
-%\end{subfigure}
-%~
-%\begin{subfigure}{0.3 \columnwidth}
-%\begin{tikzpicture}
-%\renewcommand{\datafile}{data/webbase1M-5runs-err.dat}
-%\relerrplot
-%\end{tikzpicture}
-%\subcaption{Webbase}
-%\label{fig:sparserwerr}
-%\end{subfigure}
-%
-%\caption{Relative error comparison of \MU, \HALS, \BPP\ on real world datasets.}
-%\label{fig:convergence}
-%\end{figure}
+%set which run to plot
+\renewcommand{\run}{32}
+\renewcommand{\xmax}{4.9}
+\begin{subfigure}{0.3 \columnwidth}
+\ylabeltrue
+\widertrue
+\legendfalse
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/accuracylr81pksweep.csv}
+\renewcommand{\run}{32}
+\relerrplot
+\end{tikzpicture}
+\ylabelfalse
+\subcaption{Low rank $k=32$}
+\label{fig:accuracylr81k32}
+\end{subfigure}
+~
+\begin{subfigure}{0.3 \columnwidth}
+\widertrue
+\legendfalse
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/accuracylr81pksweep.csv}
+\renewcommand{\run}{64}
+\relerrplot
+\end{tikzpicture}
+\subcaption{Low rank $k=64$}
+\label{fig:accuracylr81k64}
+\end{subfigure}
+~
+\begin{subfigure}{0.3 \columnwidth}
+\ylabeltrue
+\widertrue
+\legendtrue
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/accuracylr81pksweep.csv}
+\renewcommand{\run}{96}
+\relerrplot
+\end{tikzpicture}
+\ylabelfalse
+\subcaption{Low rank $k=96$}
+\label{fig:accuracylr81k96}
+\end{subfigure}
 
-% set which run to plot
-%\renewcommand{\run}{3}
-%
-%\begin{subfigure}{0.3 \columnwidth}
-%\ylabeltrue
-%\begin{tikzpicture}
-%\renewcommand{\datafile}{data/denserwerr-time.dat}
-%\renewcommand{\run}{1}
-%\strongscalingplot
-%\renewcommand{\run}{3}
-%\end{tikzpicture}
-%\ylabelfalse
-%\subcaption{Video}
-%\label{fig:denserwerr}
-%\end{subfigure}
-%~
-%\begin{subfigure}{0.3 \columnwidth}
-%\begin{tikzpicture}
-%\renewcommand{\datafile}{data/stkx-5runs-err.dat}
-%\legendtrue
-%\strongscalingplot
-%\legendfalse
-%\end{tikzpicture}
-%\subcaption{Stack Exchange}
-%\label{fig:stackexchangeerr}
-%\end{subfigure}
-%~
-%\begin{subfigure}{0.3 \columnwidth}
-%\begin{tikzpicture}
-%\renewcommand{\datafile}{data/webbase1M-5runs-err.dat}
-%\strongscalingplot
-%\end{tikzpicture}
-%\subcaption{Webbase}
-%\label{fig:sparserwerr}
-%\end{subfigure}
-%
-%\caption{Relative error comparison of \MU, \HALS, \BPP\ on real world datasets.}
-%\label{fig:convergence}
-%\end{figure}
+\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Low Rank Tensor of size 129x129x129x129 on 3x3x3x3 Processor Grid.}
+\label{fig:convergencelowrank}
+\end{figure}
+
+\begin{figure}
+
+%set which run to plot
+\renewcommand{\run}{32}
+\renewcommand{\xmax}{9.9}
+\begin{subfigure}{0.3 \columnwidth}
+\ylabeltrue
+\widertrue
+\legendfalse
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/accuracyuni81pksweep.csv}
+\renewcommand{\run}{32}
+\relerrplot
+\end{tikzpicture}
+\ylabelfalse
+\subcaption{Low rank $k=32$}
+\label{fig:accuracylr81k32}
+\end{subfigure}
+~
+\begin{subfigure}{0.3 \columnwidth}
+\widertrue
+\legendfalse
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/accuracyuni81pksweep.csv}
+\renewcommand{\run}{64}
+\relerrplot
+\end{tikzpicture}
+\subcaption{Low rank $k=64$}
+\label{fig:accuracylr81k64}
+\end{subfigure}
+~
+\begin{subfigure}{0.3 \columnwidth}
+\ylabeltrue
+\widertrue
+\legendtrue
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/accuracyuni81pksweep.csv}
+\renewcommand{\run}{96}
+\relerrplot
+\end{tikzpicture}
+\ylabelfalse
+\subcaption{Low rank $k=96$}
+\label{fig:accuracylr81k96}
+\end{subfigure}
+
+\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform Tensor of size 129x129x129x129 on 3x3x3x3 Processor Grid.}
+\label{fig:convergenceuniform}
+\end{figure}
+
+\begin{figure}
+
+%set which run to plot
+\renewcommand{\numiterations}{29}
+\renewcommand{\nogpu}{1}
+\renewcommand{\xmax}{128}
+
+\begin{subfigure}{0.45 \columnwidth}
+\ylabeltrue
+\widertrue
+\legendfalse
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/accuracylr81pksweep.csv}
+\renewcommand{\nogpu}{0}
+\cpuvsgpuplot
+\end{tikzpicture}
+\ylabelfalse
+\subcaption{CPU}
+\label{fig:cpulr81}
+\end{subfigure}
+~
+\begin{subfigure}{0.45 \columnwidth}
+\widertrue
+\legendtrue
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/accuracylr81pksweep.csv}
+\renewcommand{\nogpu}{1}
+\cpuvsgpuplot
+\end{tikzpicture}
+\subcaption{GPU}
+\label{fig:gpulr81}
+\end{subfigure}
+~
+\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Low Rank Tensor of size 129x129x129x129 on 3x3x3x3 Processor Grid on CPU.}
+\label{fig:cpuvsgpulowrank}
+\end{figure}
+
+\begin{figure}
+
+%set which run to plot
+\renewcommand{\numiterations}{29}
+\renewcommand{\nogpu}{0}
+\renewcommand{\xmax}{150}
+
+\begin{subfigure}{0.45 \columnwidth}
+\ylabeltrue
+\widertrue
+\legendfalse
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/accuracyuni81pksweep.csv}
+\cpuvsgpuplot
+\end{tikzpicture}
+\ylabelfalse
+\subcaption{CPU}
+\label{fig:cpuuni81}
+\end{subfigure}
+~
+\begin{subfigure}{0.45 \columnwidth}
+\widertrue
+\legendtrue
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/accuracyuni81pksweep.csv}
+\renewcommand{\nogpu}{1}
+\cpuvsgpuplot
+\end{tikzpicture}
+\subcaption{GPU}
+\label{fig:gpuuni81}
+\end{subfigure}
+~
+\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform Tensor of size 129x129x129x129 on 3x3x3x3 Processor Grid on CPU.}
+\label{fig:cpuvsgpuuniform}
+\end{figure}
+
+\renewcommand{\numiterations}{29}
+\begin{figure}
+\begin{subfigure}{0.45 \columnwidth}
+\ylabeltrue
+\widertrue
+\legendfalse
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/lucuni81pksweep.csv}
+\lucplot
+\end{tikzpicture}
+\ylabelfalse
+\subcaption{Uniform}
+\label{fig:lucuni81}
+\end{subfigure}
+~
+\begin{subfigure}{0.45 \columnwidth}
+\widertrue
+\legendtrue
+\begin{tikzpicture}
+\renewcommand{\datafile}{data/luclr81pksweep.csv}
+\renewcommand{\nogpu}{1}
+\lucplot
+\end{tikzpicture}
+\subcaption{Low Rank}
+\label{fig:luclr81}
+\end{subfigure}
+~
+\caption{LUC comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform and Low Rank Tensor of size 129x129x129x129 on 3x3x3x3 Processor Grid on CPU.}
+\label{fig:luccomp}
+\end{figure}

### 2018-08-22 latex/toms/plots.tex
pgfplots not working

@@
+% !TEX root = experiments.tex
+
 % macros for plotting
 \newcommand{\datafile}{}
-\newcommand{\run}{}
+\newcommand{\alg}{}
 \newcommand{\numiterations}{30}
 \newcommand{\minvalue}{1}
+\newcommand{\run}{16}
+\newcommand{\nogpu}{0}
+\newcommand{\xmax}{10}
 
@@
 \newif\ifwider
+% toggles including liavas results
+\newif\ifliavas
+% toggles weak 3D 
+\newif\ifweakthreeD
 
@@
 	blue, fill=blue \\ 
-	green, fill=green \\ 
+	teal, fill=teal \\ 
 	red, pattern=crosshatch, pattern color=red \\
+	teal, pattern=crosshatch, pattern color=teal \\
 	blue, pattern=crosshatch, pattern color=blue \\
-	green, pattern=crosshatch, pattern color=green \\
 }};
@@
 % set options for grouped bar plot
-\newcommand{\plotoptions}{
+\newcommand{\breakdownplotoptions}{
 	ybar stacked,
 	reverse legend,
-	bar width=6pt,
-	width=9cm, height=3.85cm,
-	ylabel={Time (seconds)}, 
+	bar width=8pt,
+	%width=9cm, height=3.85cm,
+	%width=\textwidth,
+	ylabel={Time (s)}, 
 	y label style={yshift=-.5cm},
 	ymin=0,
-	\ifnaive 
-		\ifksweep
-			symbolic x coords={10-0,10-1,10-2,10-3,,20-0,20-1,20-2,20-3,,30-0,30-1,30-2,30-3,,40-0,40-1,40-2,40-3,,50-0,50-1,50-2,50-3},
+	\ifweakthreeD
+		symbolic x coords={D1,N1,D8,N8,D27,N27,D64,N64},
 	\else
-			symbolic x coords={16-0,16-1,16-2,16-3,,96-0,96-1,96-2,96-3,,384-0,384-1,384-2,384-3,,864-0,864-1,864-2,864-3,,1536-0,1536-1,1536-2,1536-3},
-		\fi
-		xticklabels={MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive,MU,HALS,ABPP,Naive},
-	\else  
-		\ifksweep
-			symbolic x coords={10-0,10-1,10-2,,,20-0,20-1,20-2,,,30-0,30-1,30-2,,,40-0,40-1,40-2,,,50-0,50-1,50-2,},
-		\else
-			symbolic x coords={16-0,16-1,16-2,,,96-0,96-1,96-2,,,384-0,384-1,384-2,,,864-0,864-1,864-2,,,1536-0,1536-1,1536-2,}, 
-		\fi
-		xticklabels={MU,HALS,ABPP,MU,HALS,ABPP,MU,HALS,ABPP,MU,HALS,ABPP,MU,HALS,ABPP},
+		symbolic x coords={D1,N1,D16,N16,D81,N81,D256,N256},
 	\fi	
 	xtick=data,
-	xticklabel style={xshift=.15cm,rotate=75,anchor=east},
-	\ifksweep
-		xlabel={Low Rank ($k$)}, 
-	\else
-		xlabel={Number of Processes ($p$)},
-	\fi
-	xlabel style={yshift=-0.75cm},
-	%legend style={at={(0.4,-0.9)},anchor=north},
 	legend style={at={(0.5,1.3)},anchor=north},
 	legend columns=-1,
-	legend style={draw=none, cells={align=left}, nodes={scale=0.8}}
+	reverse legend
 }
 
-% grouped bar plot command
-\newcommand{\makeplot}{
-\begin{axis}[\plotoptions]
+% stacked bar plot command
+\newcommand{\breakdownplot}{
+\begin{axis}[\breakdownplotoptions]
 	\setcolors
-	\addplot table[x=algo, y expr=(\thisrow{mm}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{gram}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{allgather}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{reducescatter}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
-	\iflegend
-		\legend{MM, \NLS, Gram, All-Gather, Reduce-Scatter, All-Reduce};
-	\fi
+	\addplot table[x=alg-p, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-p, y expr=(\thisrow{krp}/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-p, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-p, y expr=((\thisrow{nnls}+\thisrow{gram}+\thisrow{allreduce})/(\minvalue*\numiterations))] {\datafile};
+	\addplot table[x=alg-p, y expr=((\thisrow{err_compute}+\thisrow{err_communication})/(\minvalue*\numiterations))] {\datafile};
+	\legend{MTTKRP,KRP,Factor Comm,NLS,Error};
 \end{axis}
@@
 
+% set options for strongscaling plot
+\newcommand{\strongscalingplotoptions}{
+		log basis y={2},
+		log basis x={2},
+		\ifliavas
+			xlabel=Cores,
+		\else
+			xlabel=Nodes,
+		\fi
+		ylabel=Time (s),
+		y tick label style={
+	        /pgf/number format/.cd,
+            	precision=4,
+		},
+	legend style={draw=none, cells={align=left}, nodes={scale=0.7}}
+}
+
+% plot time over p, filtering rows out appropriately
+\newcommand{\strongscalingplot}{
+\begin{loglogaxis}[\strongscalingplotoptions]
+	%\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={p}, y={totaltime}] {\datafile};
+	\ifliavas
+		\addplot+ [discard if not={alg}{DF},thick,mark options={solid},mark=square*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
+		\addplot+ [discard if not={alg}{NF},thick,mark options={solid},mark=square*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
+	\else
+		\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=triangle*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
+		\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
+	\fi
+	\ifliavas
+		\addplot+ [discard if not={alg}{L},thick,mark options={solid}] table [x={p}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {\datafile};
+		\legend{DimTree,NoDimTree,NbAO-NTF \cite{LK+17b}}
+	\else
+		\legend{DimTree,NoDimTree,FlatDimTree,FlatNoDimTree};
+	\fi
+\end{loglogaxis}
+}
+
 % set options for relative error over time plot
@@
 	xlabel=Time (s),
+	xmin=0,
+	xmax=\xmax,
+	domain=0:\xmax,
 	\ifylabel
@@
 \begin{axis}[\relerrplotoptions]
-	\addplot+ [discard if not={algo}{MU},discard if not={runs}{\run},green,thick,mark options={solid}] table [x={totaltime}, y={relerr}] {\datafile};
-	\addplot+ [discard if not={algo}{HALS},discard if not={runs}{\run},red,thick,mark options={solid},mark=square*] table [x={totaltime}, y={relerr}] {\datafile};
-	\addplot+ [discard if not={algo}{BPP},discard if not={runs}{\run},blue,thick,mark options={solid},mark=triangle*] table [x={totaltime}, y={relerr}] {\datafile};
+	\addplot+ [unbounded coords=jump,discard if not={algo}{0},discard if not={k}{\run},discard if not={gpu}{1},green,thick,mark options={solid}] table [x={totaltime}, y={relerr}] {\datafile};
+	\addplot+ [unbounded coords=jump,discard if not={algo}{1},discard if not={k}{\run},discard if not={gpu}{1},red,thick,mark options={solid},mark=square*] table [x={totaltime}, y={relerr}] {\datafile};
+	\addplot+ [unbounded coords=jump,discard if not={algo}{2},discard if not={k}{\run},discard if not={gpu}{1},blue,thick,mark options={solid},mark=triangle*] table [x={totaltime}, y={relerr}] {\datafile};
+		\addplot+ [unbounded coords=jump,discard if not={algo}{4},discard if not={k}{\run},discard if not={gpu}{1},magenta,thick,mark options={solid},mark=diamond*] table [x={totaltime}, y={relerr}] {\datafile};
+			\addplot+ [unbounded coords=jump,discard if not={algo}{5},discard if not={k}{\run},discard if not={gpu}{1},black,thick,mark options={solid},mark=pentagon*] table [x={totaltime}, y={relerr}] {\datafile};
 	\iflegend
-		\legend{MU,HALS,ABPP}
+		\legend{MU,HALS,ABPP,ADMM,Nestrov}
 	\fi
@@
 }
+
+% set options for relative error over time plot
+\newcommand{\cpuvsgpuoptions}{
+	xlabel=Low Rank,
+	xmin=0,
+	xmax=150, 
+	domain=0:150,
+	\ifylabel
+		ylabel=Time(s),
+		ylabel style={yshift=-.3cm},
+	\fi
+	\ifwider
+		width=2in, height=2in,
+	\else
+		width=1.35in, height=2in,
+	\fi
+	y tick label style={
+        /pgf/number format/.cd,
+            precision=4,
+	},
+	legend style={draw=none, cells={align=left}, nodes={scale=0.7}}
+}
+
+% plot relative error over time, filtering rows out appropriately
+\newcommand{\cpuvsgpuplot}{
+\begin{axis}[\cpuvsgpuoptions]
+	\addplot+ [discard if not={algo}{0},discard if not={gpu}{\nogpu},discard if not={iter}{\numiterations},green,thick,mark options={solid}] table [x={k}, y={totaltime}] {\datafile};
+	\addplot+ [discard if not={algo}{1},discard if not={gpu}{\nogpu},discard if not={iter}{\numiterations},red,thick,mark options={solid},mark=square*] table [x={k}, y={totaltime}] {\datafile};
+	\addplot+ [discard if not={algo}{2},discard if not={gpu}{\nogpu},discard if not={iter}{\numiterations},blue,thick,mark options={solid},mark=triangle*] table [x={k}, y={totaltime}] {\datafile};
+		\addplot+ [discard if not={algo}{4},discard if not={gpu}{\nogpu},discard if not={iter}{\numiterations},magenta,thick,mark options={solid},mark=diamond*] table [x={k}, y={totaltime}] {\datafile};
+			\addplot+ [discard if not={algo}{5},discard if not={gpu}{\nogpu},discard if not={iter}{\numiterations},black,thick,mark options={solid},mark=pentagon*] table [x={k}, y={totaltime}] {\datafile};
+	\iflegend
+		\legend{MU,HALS,ABPP,ADMM,Nestrov}
+	\fi
+\end{axis}
+}
+
+\newcommand{\lucplot}{
+\begin{axis}[\cpuvsgpuoptions]
+	\addplot+ [unbounded coords=jump,discard if not={algo}{0},green,thick,mark options={solid}] table [x={k}, y expr=((\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
+	\addplot+ [unbounded coords=jump,discard if not={algo}{1},red,thick,mark options={solid},mark=square*] table [x={k}, y expr=((\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
+	\addplot+ [unbounded coords=jump,discard if not={algo}{2},blue,thick,mark options={solid},mark=triangle*] table [x={k}, y expr=((\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
+		\addplot+ [unbounded coords=jump,discard if not={algo}{4},magenta,thick,mark options={solid},mark=diamond*] table [x={k}, y expr=((\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
+			\addplot+ [unbounded coords=jump,discard if not={algo}{5},black,thick,mark options={solid},mark=pentagon*] table [x={k}, y expr=((\thisrow{nnls}/(\minvalue*\numiterations))] {\datafile};
+	\iflegend
+		\legend{MU,HALS,ABPP,ADMM,Nestrov}
+	\fi
+\end{axis}
+}

### 2018-08-30 latex/toms/plots.tex
fixed tex compilation errors (beware, first compilation takes a very long time, but subsequent ones aren't bad)

@@
             	precision=4,
+		% change "directory" back to tikz
+                /tikz/.cd,
 		},
@@
             precision=4,
+            % change "directory" back to tikz
+                /tikz/.cd,
 	},

### 2018-09-06 latex/toms/experiments.tex
fixed tex compilation errors (beware, first compilation takes a very long time, but subsequent ones aren't bad)
change 129 to 162 for synthetic

@@
 \renewcommand{\xmax}{4.9}
-\begin{subfigure}{0.3 \columnwidth}
+\subfloat[Low rank $k=32$ \label{fig:accuracylr81k32}]{
 \ylabeltrue
@@
 \ylabelfalse
-\subcaption{Low rank $k=32$}
-\label{fig:accuracylr81k32}
-\end{subfigure}
-~
-\begin{subfigure}{0.3 \columnwidth}
+}
+\subfloat[Low rank $k=64$ \label{fig:accuracylr81k64}]{
 \widertrue
@@
 \end{tikzpicture}
-\subcaption{Low rank $k=64$}
-\label{fig:accuracylr81k64}
-\end{subfigure}
-~
-\begin{subfigure}{0.3 \columnwidth}
+}
+\subfloat[Low rank $k=96$ \label{fig:accuracylr81k96}]{
 \ylabeltrue
@@
 \ylabelfalse
-\subcaption{Low rank $k=96$}
-\label{fig:accuracylr81k96}
-\end{subfigure}
-
+}
 \caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Low Rank Tensor of size 129x129x129x129 on 3x3x3x3 Processor Grid.}
@@
 \begin{figure}
-
 %set which run to plot
@@
 \renewcommand{\xmax}{9.9}
-\begin{subfigure}{0.3 \columnwidth}
+\subfloat[Low rank $k=32$ \label{fig:accuracylr81k32}]{
 \ylabeltrue
@@
 \ylabelfalse
-\subcaption{Low rank $k=96$}
-\label{fig:accuracylr81k96}
-\end{subfigure}
-
-\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform Tensor of size 129x129x129x129 on 3x3x3x3 Processor Grid.}
+}
+\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid.}
 \label{fig:convergenceuniform}
@@
 
-\begin{subfigure}{0.45 \columnwidth}
+\subfloat[CPU \label{fig:cpulr81}]{
 \ylabeltrue
@@
 \ylabelfalse
-\subcaption{CPU}
-\label{fig:cpulr81}
-\end{subfigure}
-~
-\begin{subfigure}{0.45 \columnwidth}
+}
+\subfloat[GPU \label{fig:gpulr81}]{
 \widertrue
@@
 \end{tikzpicture}
-\subcaption{GPU}
-\label{fig:gpulr81}
-\end{subfigure}
-~
-\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Low Rank Tensor of size 129x129x129x129 on 3x3x3x3 Processor Grid on CPU.}
+}
+\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Low Rank Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid on CPU.}
 \label{fig:cpuvsgpulowrank}
@@
 
-\begin{subfigure}{0.45 \columnwidth}
+\subfloat[CPU \label{fig:cpuuni81}]{
 \ylabeltrue
@@
 \ylabelfalse
-\subcaption{CPU}
-\label{fig:cpuuni81}
-\end{subfigure}
-~
-\begin{subfigure}{0.45 \columnwidth}
+}
+\subfloat[GPU \label{fig:gpuuni81}]{
 \widertrue
@@
 \end{tikzpicture}
-\subcaption{GPU}
-\label{fig:gpuuni81}
-\end{subfigure}
-~
-\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform Tensor of size 129x129x129x129 on 3x3x3x3 Processor Grid on CPU.}
+}
+\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid on CPU.}
 \label{fig:cpuvsgpuuniform}
@@
 \begin{figure}
-\begin{subfigure}{0.45 \columnwidth}
+\subfloat[Uniform \label{fig:lucuni81}]{
 \ylabeltrue
@@
 \ylabelfalse
-\subcaption{Uniform}
-\label{fig:lucuni81}
-\end{subfigure}
-~
-\begin{subfigure}{0.45 \columnwidth}
+}
+\subfloat[Low Rank \label{fig:luclr81}]{
 \widertrue
@@
 \end{tikzpicture}
-\subcaption{Low Rank}
-\label{fig:luclr81}
-\end{subfigure}
-~
-\caption{LUC comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform and Low Rank Tensor of size 129x129x129x129 on 3x3x3x3 Processor Grid on CPU.}
+}
+\caption{LUC comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform and Low Rank Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid on CPU.}
 \label{fig:luccomp}

### 2018-09-07 latex/toms/experiments.tex
made axes uniform with toggles for each set of plots

@@
 \renewcommand{\xmax}{4.9}
+\renewcommand{\ymin}{0}
+\renewcommand{\ymax}{.25}
 \subfloat[Low rank $k=32$ \label{fig:accuracylr81k32}]{
+\tikzsetnextfilename{accuracylr81k32}
 \ylabeltrue
@@
 \subfloat[Low rank $k=64$ \label{fig:accuracylr81k64}]{
+\tikzsetnextfilename{accuracylr81k64}
 \widertrue
@@
 \subfloat[Low rank $k=96$ \label{fig:accuracylr81k96}]{
+\tikzsetnextfilename{accuracylr81k96}
 \ylabeltrue
@@
 }
-\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Low Rank Tensor of size 129x129x129x129 on 3x3x3x3 Processor Grid.}
+\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size 129x129x129x129 on 3x3x3x3 Processor Grid.}
 \label{fig:convergencelowrank}
@@
 \renewcommand{\xmax}{9.9}
+\renewcommand{\ymin}{.4}
+\renewcommand{\ymax}{.52}
 \subfloat[Low rank $k=32$ \label{fig:accuracylr81k32}]{
+\tikzsetnextfilename{accuracylr81k32}
 \ylabeltrue
@@
 }
-\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid.}
+\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Synthetic Uniform Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid.}
 \label{fig:convergenceuniform}
@@
 \renewcommand{\xmax}{128}
+\renewcommand{\ymax}{12}
 
 \subfloat[CPU \label{fig:cpulr81}]{
+\tikzsetnextfilename{cpulr81}
 \ylabeltrue
@@
 \subfloat[GPU \label{fig:gpulr81}]{
+\tikzsetnextfilename{gpulr81}
 \widertrue
@@
 }
-\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Low Rank Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid on CPU.}
+\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid on CPU.}
 \label{fig:cpuvsgpulowrank}
@@
 \renewcommand{\nogpu}{0}
-\renewcommand{\xmax}{150}
+\renewcommand{\xmax}{125}
+\renewcommand{\ymax}{12}
 
 \subfloat[CPU \label{fig:cpuuni81}]{
+\tikzsetnextfilename{cpuuni81}
 \ylabeltrue
@@
 \subfloat[GPU \label{fig:gpuuni81}]{
+\tikzsetnextfilename{gpuuni81}
 \widertrue
@@
 }
-\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid on CPU.}
+\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Synthetic Uniform Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid on CPU.}
 \label{fig:cpuvsgpuuniform}
@@
 \renewcommand{\numiterations}{29}
+\renewcommand{\xmax}{125}
+\renewcommand{\ymax}{.05}
 \begin{figure}
 \subfloat[Uniform \label{fig:lucuni81}]{
+\tikzsetnextfilename{lucuni81}
 \ylabeltrue
@@
 \subfloat[Low Rank \label{fig:luclr81}]{
+\tikzsetnextfilename{luclr81}
 \widertrue
@@
 }
-\caption{LUC comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform and Low Rank Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid on CPU.}
+\caption{LUC comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Synthetic Uniform and Low Rank Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid on CPU.}
 \label{fig:luccomp}

### 2018-09-07 latex/toms/plots.tex
made axes uniform with toggles for each set of plots

@@
 \newcommand{\xmax}{10}
+\newcommand{\ymin}{0}
+\newcommand{\ymax}{10}
 
@@
 	xmax=\xmax,
-	domain=0:\xmax,
+	%domain=0:\xmax,
+	ymin=\ymin,
+	ymax=\ymax,
 	\ifylabel
@@
 	\fi
-	y tick label style={
-        /pgf/number format/.cd,
-            precision=4,
-            % change "directory" back to tikz
-                /tikz/.cd,
-	},
+%	y tick label style={
+%        /pgf/number format/.cd,
+%            precision=4,
+%            % change "directory" back to tikz
+%                /tikz/.cd,
+%	},
 	legend style={draw=none, cells={align=left}, nodes={scale=0.7}}
@@
 	\iflegend
-		\legend{MU,HALS,ABPP,ADMM,Nestrov}
+		\legend{MU,HALS,ABPP,ADMM,Nesterov}
 	\fi
@@
 	xmin=0,
-	xmax=150, 
-	domain=0:150,
+	xmax=\xmax, 
+	ymin=0,
+	ymax=\ymax,
 	\ifylabel
@@
 	\fi
-	y tick label style={
-        /pgf/number format/.cd,
-            precision=4,
-            % change "directory" back to tikz
-                /tikz/.cd,
-	},
-	legend style={draw=none, cells={align=left}, nodes={scale=0.7}}
+	legend style={draw=none, cells={align=left}, nodes={scale=0.7}},
+	legend pos={north west}
 }

### 2018-09-07 latex/toms/experiments.tex
uniform plot was using lowrank details

@@
 }
-\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size 129x129x129x129 on 3x3x3x3 Processor Grid.}
+\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid.}
 \label{fig:convergencelowrank}
@@
 \renewcommand{\ymax}{.52}
-\subfloat[Low rank $k=32$ \label{fig:accuracylr81k32}]{
-\tikzsetnextfilename{accuracylr81k32}
+\subfloat[Low rank $k=32$ \label{fig:accuracyuni81k32}]{
+\tikzsetnextfilename{accuracyuni81k32}
 \ylabeltrue
@@
 }
-\subfloat[Low rank $k=64$ \label{fig:accuracylr81k64}]{
-\tikzsetnextfilename{accuracylr81k64}
+\subfloat[Low rank $k=64$ \label{fig:accuracyuni81k64}]{
+\tikzsetnextfilename{accuracyuni81k64}
 \widertrue
@@
 }
-\subfloat[Low rank $k=96$ \label{fig:accuracylr81k96}]{
-\tikzsetnextfilename{accuracylr81k96}
+\subfloat[Low rank $k=96$ \label{fig:accuracyuni81k96}]{
+\tikzsetnextfilename{accuracyuni81k96}
 \ylabeltrue

### 2018-12-07 latex/slides/Sandia-LDRD-pres/talk.tex
added SIAM-ALA talk

--- 2018-04-02 latex/slides/talk.tex
@@
 
-\input{../sc18/fig/Alg_header.tex}
+\input{../../sc18/fig/Alg_header.tex}
 
@@
 \scalebox{1.5}{
-\only<1>{\input{../sc18/fig/AlgStart.tex}}
-\only<2>{\input{../sc18/fig/AlgMTTKRP.tex}}
-\only<3>{\input{../sc18/fig/AlgRS.tex}}
-\only<4>{\input{../sc18/fig/AlgNLS.tex}}
-\only<5>{\input{../sc18/fig/AlgAG.tex}}
+\only<1>{\input{../../sc18/fig/AlgStart.tex}}
+\only<2>{\input{../../sc18/fig/AlgMTTKRP.tex}}
+\only<3>{\input{../../sc18/fig/AlgRS.tex}}
+\only<4>{\input{../../sc18/fig/AlgNLS.tex}}
+\only<5>{\input{../../sc18/fig/AlgAG.tex}}
 }
@@
 \centering
-  \subfloat[Start]{\input{../sc18/fig/AlgStart.tex}} \quad
-  \subfloat[Local MTTKRP]{\input{../sc18/fig/AlgMTTKRP.tex}} \quad
-  \subfloat[Reduce-Scatter]{\input{../sc18/fig/AlgRS.tex}} \quad
-  \subfloat[Local NLS]{\input{../sc18/fig/AlgNLS.tex}} \quad
-  \subfloat[All-Gather]{\input{../sc18/fig/AlgAG.tex}}
+  \subfloat[Start]{\input{../../sc18/fig/AlgStart.tex}} \quad
+  \subfloat[Local MTTKRP]{\input{../../sc18/fig/AlgMTTKRP.tex}} \quad
+  \subfloat[Reduce-Scatter]{\input{../../sc18/fig/AlgRS.tex}} \quad
+  \subfloat[Local NLS]{\input{../../sc18/fig/AlgNLS.tex}} \quad
+  \subfloat[All-Gather]{\input{../../sc18/fig/AlgAG.tex}}
 \end{figure}
@@
 
-\input{../sc18/fig/dim_tree.tex}
+\input{../../sc18/fig/dim_tree.tex}
 \end{center}
@@
 
-\input{../sc18/plots}
+\input{../../sc18/plots}
 
@@
 \begin{tikzpicture}[scale=.9]
-\renewcommand{\datafile}{../sc18/data/str_3D_syn.dat}
+\renewcommand{\datafile}{../../sc18/data/str_3D_syn.dat}
 \renewcommand{\numiterations}{42}
@@
 \begin{tikzpicture}[scale=.9]
-\renewcommand{\datafile}{../sc18/data/str_5D_syn.dat}
+\renewcommand{\datafile}{../../sc18/data/str_5D_syn.dat}
 \renewcommand{\numiterations}{10}
@@
 \begin{tikzpicture}
-\renewcommand{\datafile}{../sc18/data/wk_4D_syn.dat}
+\renewcommand{\datafile}{../../sc18/data/wk_4D_syn.dat}
 \renewcommand{\numiterations}{10}
@@
 \begin{figure}
-\renewcommand{\datafile}{../sc18/data/ksw_4D_HSI.dat}
+\renewcommand{\datafile}{../../sc18/data/ksw_4D_HSI.dat}
 \renewcommand{\numiterations}{10}

### 2018-12-11 latex/toms/experiments.tex
real world completed
committing experimental plots from python

--- 2018-09-06 latex/toms/experiments.tex
@@
 
-\subsection{Strong Scaling}
-
-We perform two strong scaling experiments to compare performance with NbAO-NTF.
-The experiments use a cubical synthetic tensor and the HSI image used in \cite{LK+17b}, both of which are 3D.
-
-The performance on the cubical synthetic tensor is shown in \Cref{fig:strongsynthetic3D}. 
-We can observe from the figure that all the three algorithms scale nearly linearly as the problem remains compute bound. 
-Our algorithm (with the dimension tree optimization) achieves a speedup of $1771\times$ on 4096 cores over the same implementation running on 1 core.
-Recall from \label{sec:analysis} that the computation scales linearly with $1/P$ while the communication scales with $1/P^{1/N}=1/P^{1/3}$. 
-As is evident from the figure, the communication cost does not degrade performance even for thousands of cores. 
-Our proposed algorithm with dimension trees is 35\% faster than NbAO-NTF at 512 cores (with similar relative difference for other core counts).
-This performance improvement is due in large part to the 50\% reduction in arithmetic operations provided by the dimension tree optimization.
-There is little difference in performance between our implementation without dimension trees and NbAO-NTF.
-
-\begin{figure}
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/str_3D_syn.dat}
-\renewcommand{\numiterations}{1}
-\liavastrue
-\strongscalingplot
-\end{tikzpicture}
-\caption{Strong scaling of 3D synthetic tensor with dimension $1024\times 1024\times 1024$ on processor grids $2^k\times 2^k\times 2^k$ for $k\in\{0,1,2,3,4\}$.  The rank is fixed at 32.}
-\label{fig:strongsynthetic3D}
-\end{figure}
-
-\Cref{fig:stronghsi3D} shows the strong scaling on the HSI data. 
-In this case, our proposed algorithm with dimension trees is over $2\times$ faster than NbAO-NTF, but part of this speedup is due to differences in the NLS update algorithms.
-For the low core count, the dimension tree provides a 60\% speedup compared to the MTTKRP time in NbAO-NTF.
-At the high core counts for this experiment, the local MTTKRP is no longer the dominating cost.
-%The proposed algorithm with classical MTTKRP was performing better on higher number of processors over the baseline. 
-%Similarly, at this stage, the SVD used for Nestrov Non-negative Least Squares on Liavas dominates the cost over MTTKRP that makes it even slower than the our proposed algorithm without dimension trees.
-
-%\textcolor{red}{Finally, we ran on 512 nodes, utilizing the 16 threads per node via parallel BLAS and OMP, for a $1024\times1024\times1024$ synthetic tensor and processor grid $8\times8\times8$.
-Here we observe a $253\times$ speedup which is the result of each node having less than 20 megabytes of data.}
-
-\begin{figure}
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/str_3D_HSI.dat}
-\renewcommand{\numiterations}{10}
-\liavastrue
-\strongscalingplot
-\end{tikzpicture}
-\caption{Strong scaling of 3D HSI real world data with dimension 1024 x1344 x 33 on processor grids of $k \times k\times 1$ for $k \in {1, 2, 4, 8, 16, 32}$. The rank is fixed at 32.}
-\label{fig:stronghsi3D}
-\end{figure}
-
-In \Cref{fig:strongsynthetic5D}, we benchmark performance for a 5D cubical tensor with each dimension set to 64.
-Because the tensor is 5D, we can no longer compare against NbAO-NTF.
-We see a $13-16\times$ speed up using a dimension tree over not using one for this problem.
-As predicted, the dimension tree optimization saves relatively more arithmetic for higher-order tensors.
-However, the reduction in leading order flop cost is only $2.5\times$ for $N=5$; the rest of the speedup comes from more efficient DGEMM performance and avoiding memory-bound KRP computation.  
-That is, although the flop count of KRP computation is lower order, it still contributes to the run time because it is inefficient.
-Also, for tensors with balanced dimensions, the dimension tree approach yields more favorable shapes for DGEMM.
-
-\begin{figure}
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/str_5D_syn.dat}
-\renewcommand{\numiterations}{10}
-\liavasfalse
-\strongscalingplot
-\end{tikzpicture}
-\caption{Strong scaling of 5D synthetic tensor with dimension $64\times 64\times 64\times 64\times 64$ on processor grids $1\times1\times1\times1\times1$, $2\times1\times1\times1\times1$, $\dots$, $2\times2\times2\times2\times2$.
-The rank is fixed at 32.}
-\label{fig:strongsynthetic5D}
-\end{figure}
-
-\subsection{Weak Scaling Time Breakdown}
-
-We also perform a weak scaling experiment to understand the time it takes to solve bigger problems with more processors.
-In this experiment, we use a synthetic 4D tensor and keep the amount of tensor data assigned to each processor constant, with tensor and processor grid of dimensions $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$ and the rank fixed at 32.
-%For eg., in the case of 3 x 3 x 3 x 3 processor grid with 81 nodes, we ran the proposed algorithm with and without dimension trees on a four mode tensor of size 384 x 384 x 384 x 384. 
-The results of the breakdown plot is shown in \Cref{fig:weaksynthetic4D}. 
-In this case, the algorithm is compute bound with and without the use of the dimension tree, so the total time of the weak scaling remains fixed for both cases. 
-Using a dimension tree, the time is completely dominated by the MTTKRP computation.  
-Without using a dimension tree, we observe that the KRP is expensive and yields a $2.5\times$ slower total run time even in the 4D case.
-
-\begin{figure}
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/wk_4D_syn.dat}
-\renewcommand{\numiterations}{10}
-\weakthreeDfalse
-\breakdownplot
-%\labels
-\end{tikzpicture}
-\caption{Weak scaling of 4D synthetic tensors with (D) and without (N) the use of dimension trees.
-The tensor and processor grid dimensions are $128k\times 128k\times 128k\times 128k$ and $k\times k\times k\times k$ for $k\in\{1,2,3,4\}$, and the rank is fixed at 32.
-The reported times are per iteration.}
-\label{fig:weaksynthetic4D}
-\end{figure}
-
-\subsection{Varying Processor Grid}
-
-In order to illustrate the effect of processor grid choice on running time, we show in \Cref{fig:commsweep10} a time breakdown over various processor grid choices for a 4D problem on 81 processors.
-Because the tensor is cubical and 81 has a restricted factorization into 4 numbers, there are 5 distinct processor grids.
-The overall takeaway is that the processor grid has very little effect on running time; in this experiment there is less than 10\% variation in overall time.
-While the optimal processor grid reduced the communication time by approximately $3\times$ compared to the other processor grids, the running time is dominated by local computation, so it had little effect on overall time.
-Furthermore, adjusting the processor grid affects the local tensor dimensions and the performance of the local computations, and the optimal processor grid led to slower local performance.
-For $R=10$, all of the local computation is memory bandwidth bound, and we believe the variations in running times to be effects of some temporary quantities fitting into smaller levels of cache.
-
-\begin{figure}
-\renewcommand{\datafile}{data/comm_sweep_scale_pgf_10.dat}
-\renewcommand{\numiterations}{10}
-\begin{tikzpicture}
-\begin{axis}[	
-	ybar stacked,
-	bar width=8pt,
-	width=\columnwidth,
-	height =.75\columnwidth,
-	%width=9cm, height=3.85cm,
-	ylabel={Time (s)}, 
-	xlabel={Processor Grid},
-	y label style={yshift=-.5cm},
-	ymin=0,
-	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
-	symbolic x coords={81x1x1x1-10, 27x3x1x1-10,9x9x1x1-10,9x3x3x1-10,3x3x3x3-10,},
-	%xtick=data,
-	xticklabels={,81x1x1x1,27x3x1x1,9x9x1x1,9x3x3x1,3x3x3x3},
-	legend style={at={(0.5,1.3)},anchor=north},
-	legend columns=-1,
-]
-	\setcolors
-	\addplot table[x=algo-k, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo-k, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo-k, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo-k, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=algo-k, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
-	\legend{PM,mTTV,NLS,Factor Comm,Gram Comm};
-\end{axis}
-%\labels
-\end{tikzpicture}
-\caption{Time breakdown for $243\times243\times243\times243$ tensor and rank $R=10$ on 81 processors for varying processor grids.}
-\label{fig:commsweep10}
-\end{figure}
-
 \subsection{Varying Approximation Rank}
@@
 
-\Cref{fig:ksweep4DHSI} shows the time breakdown of our implementation using a dimension tree on the 4D HSI dataset for $R=\{10,\dots,50\}$. 
-We observe an overall time increase with increased $R$, but each part of the computation scales slightly differently.
-The multi-TTV computation ({\em mTTV}) scales linearly with the increasing $R$,  whereas the partial MTTKRP ({\em PM}) is scaling super-linearly. 
-This is because mTTV is cast as matrix-vector products (DGEMV) and PM is cast as matrix-matrix products (DGEMM).
-As $R$ increases from 10 to 50, DGEMM performance improves but DGEMV performance is constant.  
-The local NLS time is increasing with $O(R^3)$ as expected and the All-Reduce required for the Gram matrices scales with $O(R^2)$, becoming a significant cost for larger $R$.
-
-In \Cref{fig:ksweepneuro} we compare performance for various ranks $R$ across all 3 algorithms, again used flat MPI. 
-Starting at $R=10$ we see the largest speed up of $2\times$ for our implementation with a dimension tree over NbAO-NTF. 
-This is due to a combination of the dimension tree performing fewer flops in the MTTKRPs and KRPs. 
-However, as the rank increases this speed up diminishes to $1.6\times$. 
-The loss of speed up is a result of the fact that, as we observed in \Cref{fig:ksweep4DHSI}, the multi-TTV operations do not scale as well as the partial-MTTKRPs for increasing $R$.
-Again, the performance of our implementation without the dimension tree optimization is comparable to NbAO-NTF.
+%%%%%%%%%%%%%%%%%
+%Convergence Plot for Synthetic low rank
+%%%%%%%%%%%%%%%%%
 
 \begin{figure}
-\renewcommand{\datafile}{data/ksw_4D_HSI.dat}
-\renewcommand{\numiterations}{10}
-\begin{tikzpicture}
-\begin{axis}[	
-	ybar stacked,
-	bar width=8pt,
-	width=\columnwidth,
-	height =.75\columnwidth,
-	%width=9cm, height=3.85cm,
-	ylabel={Time (s)}, 
-	xlabel={Rank $R$},
-	y label style={yshift=-.5cm},
-	ymin=0,
-	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
-	symbolic x coords={D10,D20,D30,D40,D50},
-	%xtick=data,
-	xticklabels={,10,20,30,40,50},
-	legend style={at={(0.5,1.3)},anchor=north},
-	legend columns=-1,
-]
-	\setcolors
-	\addplot table[x=alg-K, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=alg-K, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=alg-K, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=alg-K, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
-	\addplot table[x=alg-K, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
-	\legend{PM,mTTV,NLS,Factor Comm,Gram Comm};
-\end{axis}
-%\labels
-\end{tikzpicture}
-\caption{Per-iteration time breakdown of our implementation (using dimension trees) over various ranks for a time-lapse HSI dataset with dimensions $1344\times 1024\times 33 \times 9$ on 64 processors arranged in a $8\times8\times1\times1$ grid.}
-\label{fig:ksweep4DHSI}
+\subfloat[Low rank $k=64$ \label{fig:accuracylr81k64}]{
+\includegraphics[width=0.45\textwidth, height=2in]{data/plots/accuracylr64.pdf}
+}
+\subfloat[Low rank $k=96$ \label{fig:accuracylr81k96}]{
+\includegraphics[width=0.45\textwidth, height=2in]{data/plots/accuracylr96.pdf}
+}
+\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size 162x162x162x162 on 81 Titan Nodes as 3x3x3x3 Processor Grid.}
+\label{fig:convergencelowrank}
 \end{figure}
 
+%%%%%%%%%%%%%%%%%
+%Convergence Plot for Realworld
+%%%%%%%%%%%%%%%%%
+
 \begin{figure}
-\begin{tikzpicture}
-\begin{axis}[legend style={draw=none, anchor=north,  cells={align=left}, nodes={scale=0.7}}, legend pos=north west, xlabel=Low Rank $R$, ylabel=Time(s), ylabel near ticks]
-	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
-	\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
-	\addplot+ [discard if not={alg}{L},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
-	\legend{DimTree,NoDimTree,NbAO-NTF}
-\end{axis}
-\end{tikzpicture}
-\caption{Overall running time for dFC dataset with dimensions $30{,}012\times 1200 \times 500$ on 1440 cores arranged in a $2\times 6\times 120$ processor grid and varying choices of rank $R$.}
-\label{fig:ksweepneuro}
+	\subfloat[Low rank $k=64$ \label{fig:accuracyrwk64}]{
+		\includegraphics[width=0.45\textwidth, height=2in]{data/plots/accuracyrw64.pdf}
+	}
+	\subfloat[Low rank $k=96$ \label{fig:accuracyrwk96}]{
+		\includegraphics[width=0.45\textwidth, height=2in]{data/plots/accuracyrw96.pdf}
+	}
+	\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Realworld Low Rank Tensor of size $200 \times 500 \times 64 \times 192$ on 64 Titan Nodes as $2\times4\times2\times4$ Processor Grid.}
+	\label{fig:convergencerealworld}
 \end{figure}
+
+%%%%%%%%%%%%%%%%%
+%CPU VS GPU on Synthetic Low Rank and Realworld
+%%%%%%%%%%%%%%%%%
 
@@
 
-%set which run to plot
-\renewcommand{\run}{32}
-\renewcommand{\xmax}{4.9}
-\subfloat[Low rank $k=32$ \label{fig:accuracylr81k32}]{
-\ylabeltrue
-\widertrue
-\legendfalse
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/accuracylr81pksweep.csv}
-\renewcommand{\run}{32}
-\relerrplot
-\end{tikzpicture}
-\ylabelfalse
-}
-\subfloat[Low rank $k=64$ \label{fig:accuracylr81k64}]{
-\widertrue
-\legendfalse
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/accuracylr81pksweep.csv}
-\renewcommand{\run}{64}
-\relerrplot
-\end{tikzpicture}
-}
-\subfloat[Low rank $k=96$ \label{fig:accuracylr81k96}]{
-\ylabeltrue
-\widertrue
-\legendtrue
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/accuracylr81pksweep.csv}
-\renewcommand{\run}{96}
-\relerrplot
-\end{tikzpicture}
-\ylabelfalse
-}
-\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Low Rank Tensor of size 129x129x129x129 on 3x3x3x3 Processor Grid.}
-\label{fig:convergencelowrank}
-\end{figure}
-
-\begin{figure}
-%set which run to plot
-\renewcommand{\run}{32}
-\renewcommand{\xmax}{9.9}
-\subfloat[Low rank $k=32$ \label{fig:accuracylr81k32}]{
-\ylabeltrue
-\widertrue
-\legendfalse
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/accuracyuni81pksweep.csv}
-\renewcommand{\run}{32}
-\relerrplot
-\end{tikzpicture}
-\ylabelfalse
-}
-\subfloat[Low rank $k=64$ \label{fig:accuracylr81k64}]{
-\widertrue
-\legendfalse
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/accuracyuni81pksweep.csv}
-\renewcommand{\run}{64}
-\relerrplot
-\end{tikzpicture}
-}
-\subfloat[Low rank $k=96$ \label{fig:accuracylr81k96}]{
-\ylabeltrue
-\widertrue
-\legendtrue
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/accuracyuni81pksweep.csv}
-\renewcommand{\run}{96}
-\relerrplot
-\end{tikzpicture}
-\ylabelfalse
-}
-\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid.}
-\label{fig:convergenceuniform}
-\end{figure}
-
-\begin{figure}
-
-%set which run to plot
-\renewcommand{\numiterations}{29}
-\renewcommand{\nogpu}{1}
-\renewcommand{\xmax}{128}
-
 \subfloat[CPU \label{fig:cpulr81}]{
-\ylabeltrue
-\widertrue
-\legendfalse
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/accuracylr81pksweep.csv}
-\renewcommand{\nogpu}{0}
-\cpuvsgpuplot
-\end{tikzpicture}
-\ylabelfalse
+\includegraphics[width=0.45\textwidth, height=2in]{data/plots/cpuvsgpulr810.pdf}
 }
 \subfloat[GPU \label{fig:gpulr81}]{
-\widertrue
-\legendtrue
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/accuracylr81pksweep.csv}
-\renewcommand{\nogpu}{1}
-\cpuvsgpuplot
-\end{tikzpicture}
+\includegraphics[width=0.45\textwidth, height=2in]{data/plots/cpuvsgpulr811.pdf}
 }
-\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Low Rank Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid on CPU.}
+\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size 162x162x162x162 on 81 Titan Nodes as 3x3x3x3 Processor Grid on CPU and GPU.}
 \label{fig:cpuvsgpulowrank}
@@
 
+%%%%%%%%%%%%%%%%%
+%LUC Running Time Comparison
+%%%%%%%%%%%%%%%%%
+
 \begin{figure}
-
-%set which run to plot
-\renewcommand{\numiterations}{29}
-\renewcommand{\nogpu}{0}
-\renewcommand{\xmax}{150}
-
-\subfloat[CPU \label{fig:cpuuni81}]{
-\ylabeltrue
-\widertrue
-\legendfalse
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/accuracyuni81pksweep.csv}
-\cpuvsgpuplot
-\end{tikzpicture}
-\ylabelfalse
+\subfloat[Synthetic Low Rank \label{fig:luclr81}]{
+\includegraphics[width=0.45\textwidth, height=2in]{data/plots/luclr81.pdf}
 }
-\subfloat[GPU \label{fig:gpuuni81}]{
-\widertrue
-\legendtrue
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/accuracyuni81pksweep.csv}
-\renewcommand{\nogpu}{1}
-\cpuvsgpuplot
-\end{tikzpicture}
+\subfloat[Tiny Imagenet \label{fig:lucrw}]{
+\includegraphics[width=0.45\textwidth, height=2in]{data/plots/lucrw64.pdf}
 }
-\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid on CPU.}
-\label{fig:cpuvsgpuuniform}
+\caption{LUC comparison of \MU, \HALS, \BPP, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size 162x162x162x162 on 81 Titan nodes as 3x3x3x3 Processor Grid on CPU and Realworld TinyImagenet Dataset on 64 Titan nodes as 2x4x2x4 processor grid}
+\label{fig:luccomp}
 \end{figure}
 
-\renewcommand{\numiterations}{29}
+\subsection{Strong Scaling Time Breakdown}
+
+%%%%%%%%%%%%%%%%%
+%Strong Scaling Synthetic Low Rank
+%%%%%%%%%%%%%%%%%
 \begin{figure}
-\subfloat[Uniform \label{fig:lucuni81}]{
-\ylabeltrue
-\widertrue
-\legendfalse
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/lucuni81pksweep.csv}
-\lucplot
-\end{tikzpicture}
-\ylabelfalse
+\subfloat[Synthetic 3D Low Rank \label{fig:strsca3d}]{
+\includegraphics[width=\textwidth, height=3in]{data/plots/strsca3d.pdf}
+}\\
+\subfloat[Synthetic 4D Low Rank \label{fig:strsca4d}]{
+\includegraphics[width=\textwidth, height=3in]{data/plots/strsca4d.pdf}
 }
-\subfloat[Low Rank \label{fig:luclr81}]{
-\widertrue
-\legendtrue
-\begin{tikzpicture}
-\renewcommand{\datafile}{data/luclr81pksweep.csv}
-\renewcommand{\nogpu}{1}
-\lucplot
-\end{tikzpicture}
+\caption {Strong Scaling on Synthetic 3D and 4D low rank tensors.
+For 3D, the input tensor is of size 1024x1024x1024 and 256x256x256x256 for 4D on 8,16, 32, 64 and 128 nodes in Titan.
+The algorithms alg=0,1,2,4,5 are \MU, \HALS, \BPP, \ADMM, \Nestrov\ respectively.}
+\label{fig:synstrongscaling}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%
+%Strong Scaling Realworld Low Rank
+%%%%%%%%%%%%%%%%%
+\begin{figure}
+\includegraphics[width=\textwidth, height=3in]{data/plots/ssrw_tinyimagenet_no16.pdf}
+\caption {Strong Scaling on Realworld Tiny Imagenet Dataset on 32, 64, 128, 256, 512 and 1024 nodes in Titan.
+The algorithms alg=0,1,2,4,5,6 are \MU, \HALS, \BPP, \ADMM, \Nestrov, \CPALS\ respectively.}
+\label{fig:synstrongscaling}
+\end{figure}
+
+\subsection{Weak Scaling Time Breakdown}
+
+%%%%%%%%%%%%%%%%%
+%Weak Scaling Synthetic Low Rank
+%%%%%%%%%%%%%%%%%
+\begin{figure}
+\subfloat[Synthetic 3D Low Rank \label{fig:wksca3d}]{
+\includegraphics[width=\textwidth, height=3in]{data/plots/wksca3d.pdf}
+}\\
+\subfloat[Synthetic 4D Low Rank \label{fig:wksca4d}]{
+\includegraphics[width=\textwidth, height=3in]{data/plots/wksca4d.pdf}
 }
-\caption{LUC comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov on 4D Synthetic Uniform and Low Rank Tensor of size 162x162x162x162 on 3x3x3x3 Processor Grid on CPU.}
-\label{fig:luccomp}
+\caption {Weak Scaling on Synthetic 3D and 4D low rank tensors.
+For 3D, the input tensor is of size 128x128x128, 256x256x256, 378x378x378 and 512x512x512 on 1, 8, 27 and 64 Titan Nodes.
+The 4D tensors were of size 128, 256, 405, 512 for each mode on 1, 16, 81 and 256 Titan nodes.
+The algorithms alg=0,1,2,4,5 are \MU, \HALS, \BPP, \ADMM, \Nestrov\ respectively.}
+\label{fig:synweakscaling}
 \end{figure}
+
+\subsection{Varying Processor Grid}
+
+%%%%%%%%%%%%%%%%%
+%Configuration Sweep
+%%%%%%%%%%%%%%%%%
+\begin{figure}
+\includegraphics[width=\textwidth, height=2.5in]{data/plots/confsweep4d.pdf}
+\caption{Processor Grid sweep of a 162x162x162x162 synthetic low rank tensor on 81 nodes in Titan as 81x1x1x1, 27x3x1x1, 9x3x3x3, 3x3x3x3 grids represented as p=1,2,3,4 respectively.
+The algorithms alg=0,1,2,4,5 are \MU, \HALS, \BPP, \ADMM, \Nestrov\ respectively.}
+\label{fig:confsweep4D}
+\end{figure}

### 2018-12-11 latex/toms/data/timebreakdown.py

@@ -34,10 +34,10 @@ for bar, hatch in zip(bars, hatches): # hatch_colors, fill_colors):
     bar.set_hatch(hatch)
     #bar.set_color(fill_color)
     #bar.set_hatch_color(mcolors.rgb_to_hsv(hatch_color))
-plt.legend(loc=9, ncol=3)
+plt.legend(bbox_to_anchor=(1.04,0.5), loc='center left', ncol=1)
 plt.ylabel('Running Time (in Secs)')
-plt.gcf().set_size_inches(9, 8)
-plt.savefig(output_file,format='pdf')
+#plt.gcf().set_size_inches(9, 8)
+plt.savefig(output_file,format='pdf',bbox_inches='tight')
 
 print 'check output file ' + output_file
 

### 2018-12-14 latex/toms/algorithm.tex
Initial commit for TOMS paper algorithm section

@@
 If the tensor dimensions can be partitioned into two parts with approximately equal geometric means, the extra memory requirement for running a dimension tree is as small as $O(R\sqrt{I/P})$, which is also typically dominated by $O(I/P)$.
+
+\subsection{NLS Algorithms}
+This subsection deals with the nonnegative least squares solves needed to update the factor at each inner iteration of the algorithm (\cref{line:locNLS} in \Cref{alg:Par-NNCP-short}).
+Our framework is capable of supporting any alternating-updating solver~\cite{KBP16}.
+The NLS solvers which fit this framework and are implemented in our software package are Multiplicative Update~\cite{LS99}, Hierarchical Alternating Least Squares~\cite{CP2009}, Block Principal Pivoting~\cite{KP2011}, Alternating Direction Method of Multipliers~\cite{HSL2016} and Nesterov-type algorithm~\cite{LKLHS2017}.
+We briefly describe the different solvers below.
+
+\subsubsection{Multiplicative Update (MU)}
+\subsubsection{Hierarchical Alternating Least Squares (HALS)}
+\subsubsection{Block Principal Pivoting (BPP)}
+\subsubsection{Alternating Direction Method of Multipliers (ADMM)}
+\subsubsection{Nesterov-type algorithm}

### 2018-12-14 latex/toms/abstract.tex
added draft abstract

@@
+% !TEX root = paper.tex
+
 \begin{abstract}
-Non-negative matrix factorization (NMF) is the problem of determining two non-negative low rank factors $\WW$ and $\HH$, for the given input matrix $\AA$, such that $\AA \approx \WW \HH$.  
-NMF is a useful tool for many applications in different domains such as topic modeling in text mining, background separation in video analysis, and community detection in social networks.  
-Despite its popularity in the data mining community, there is a lack of efficient parallel algorithms to solve the problem for big data sets.  
-The main contribution of this work is a new, high-performance parallel computational framework for a broad class of NMF algorithms that iteratively solves alternating non-negative least squares (NLS) subproblems for $\WW$ and $\HH$.
-It maintains the data and factor matrices in memory (distributed across processors), uses MPI for interprocessor communication, and, in the dense case, provably minimizes communication costs (under mild assumptions).
-The framework is flexible and able to leverage a variety of NMF and NLS algorithms, including Multiplicative Update, Hierarchical Alternating Least Squares, and Block Principal Pivoting.
-Our implementation allows us to benchmark and compare different algorithms on massive dense and sparse data matrices of size that spans from few hundreds of millions to billions.  
-We demonstrate the scalability of our algorithm and compare it with baseline implementations, showing significant performance improvements. 
-The code and the datasets used for conducting the experiments are available online.
+We consider the problem of low-rank approximation of massive dense non-negative matrix and tensor data, for example to discover latent patterns in video and imaging applications.
+As the size of data sets grows, single workstations are hitting bottlenecks in both computation time and available memory.
+We propose a distributed-memory parallel computing solution to handle massive data sets, loading the input data across the memories of multiple nodes and performing efficient and scalable parallel algorithms to compute the low-rank approximation.
+We present a software package called PLANC, which implements our solution and allows for extension in terms of data (dense or sparse), algorithm (e.g., from multiplicative updating techniques to alternating direction method of multipliers), and architecture (we exploit GPUs to accelerate the computation in this work).
+We describe our parallel distributions and algorithms, which are careful to avoid unnecessary communication and computation, show how to extend the software to include new algorithms and/or constraints, and report efficiency and scalability results for both synthetic and real-world data sets coming from a variety of fields.
 \end{abstract}

### 2018-12-14 latex/toms/algorithm.tex

@@
 \subsection{NLS Algorithms}
-This subsection deals with the nonnegative least squares solves needed to update the factor at each inner iteration of the algorithm (\cref{line:locNLS} in \Cref{alg:Par-NNCP-short}).
-Our framework is capable of supporting any alternating-updating solver~\cite{KBP16}.
+This subsection deals with the nonnegative least squares (NLS) solves needed to update the factor at each inner iteration of the algorithm (\cref{line:locNLS} in \Cref{alg:Par-NNCP-short}).
+The general problem to be solved in each inner iteration is a constrained least squares problem of the form,
+
+\begin{align}
+\M{X} \leftarrow \underset{\M{X} \ge 0}{\argmin} \left\lVert \M{A}\M{X} -\M{B}\right\rVert^2_F
+\label{eqn:nls}
+\end{align}
+
+In the case of updating the local factor matrix $\Mn{H}{n}_{\V{p}}$ we need to solve~\ref{eqn:nls} with $\M{X} = \Mn{H}{n}_{\V{p}}$, $\M{A} = \Mn{S}{n}$ and $\M{B} = \Mn{M}{n}_{\V{p}}$.
+Our framework is capable of supporting any alternating-updating NLS solver~\cite{KBP16}.
 The NLS solvers which fit this framework and are implemented in our software package are Multiplicative Update~\cite{LS99}, Hierarchical Alternating Least Squares~\cite{CP2009}, Block Principal Pivoting~\cite{KP2011}, Alternating Direction Method of Multipliers~\cite{HSL2016} and Nesterov-type algorithm~\cite{LKLHS2017}.
 We briefly describe the different solvers below.
+Note that the descriptions are for the general form of the NLS solver as shown in~\ref{eqn:nls}.
 
 \subsubsection{Multiplicative Update (MU)}
+The MU solve is an elementwise operation. The update rule for each element of $\M{X}$ is as follows,
+\begin{align}
+\M{X}(i,j) \leftarrow \M{X}(i,j)\frac{\M{B}(i,j)}{\left(\M{A}\M{X}\right)(i,j)}
+\end{align}
+While this rule does not solve~\ref{eqn:nls} to optimality it ensures a reduction in the objective value from the initial value of $\M{X}$.
+
 \subsubsection{Hierarchical Alternating Least Squares (HALS)}
+HALS updates are performed on individual rows of $\M{X}$. The update rule for row $i$ can be written in closed form as,
+\begin{align}
+\M{X}(i,:) \leftarrow \left[ \M{X}(i,:) + \frac{\M{B}(i,:) - \left(\M{A}(i,:)\M{X}\right)}{\M{A}(i,i)} \right]_+
+\end{align}
+The rows of $\M{X}$ are updated in order so that the latest values are used in every update step.
+
 \subsubsection{Block Principal Pivoting (BPP)}

### 2018-12-18 latex/toms/experiments.tex
mouse plots completed

@@
 \begin{figure}
-\includegraphics[width=\textwidth, height=3in]{data/plots/ssrw_tinyimagenet_no16.pdf}
+\includegraphics[width=\textwidth, height=3in]{data/plots/ssrw_tinyimagenet_breakdown.pdf}
 \caption {Strong Scaling on Realworld Tiny Imagenet Dataset on 32, 64, 128, 256, 512 and 1024 nodes in Titan.
@@
 \end{figure}
+
+%%%%%%%%%%%%%%%%%
+%Mouse Data
+%%%%%%%%%%%%%%%%%
+\begin{figure}
+	\subfloat[Low rank $k=64$ \label{fig:accuracyrwk64}]{
+		\includegraphics[width=0.45\textwidth, height=2in]{data/plots/accuracymouse64.pdf}
+	}
+	\subfloat[Low rank $k=96$ \label{fig:accuracyrwk96}]{
+		\includegraphics[width=0.45\textwidth, height=2in]{data/plots/accuracymouse96.pdf}
+	}
+	\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Realworld Low Rank Tensor of size $1040\times1392\times69\times25$ on 64 Titan Nodes as $8\times8\times1\times1$ Processor Grid.}
+	\label{fig:convergencemouse}
+\end{figure}
+
+\begin{figure}
+   \includegraphics[width=0.45\textwidth, height=2in]{data/plots/lucmouse64.pdf}
+	\caption{LUC comparison of \MU, \HALS, \BPP, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size $1040\times1392\times69\times25$ on 64 Titan nodes as $8\times8\times1\times1$ Processor Grid}
+	\label{fig:luccompmouse}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=\textwidth, height=3in]{data/plots/ssrw_mouse.pdf}
+	\caption {Strong Scaling on Realworld Mouse MRI Dataset on 32, 64, 128, 256, 512 and 1024 nodes in Titan. The algorithms alg=0,1,2,4,5,6 are \MU, \HALS, \BPP, \ADMM, \Nestrov, \CPALS\ respectively.}
+	\label{fig:synstrongscaling}
+\end{figure}

### 2018-12-18 common/distutils.hpp

@@ -43,7 +43,9 @@ inline double mpitoc() {
   tictoc_stack.pop();
   return rc;
 }
-
+/**
+ * Captures the memory usage of every mpi process
+ */
 inline void memusage(const int myrank, std::string event) {
   // Based on the answer from stackoverflow
   // http://stackoverflow.com/questions/669438/how-to-get-memory-usage-at-run-time-in-c
@@ -65,17 +67,26 @@ inline void memusage(const int myrank, std::string event) {
   }
 }
 
-// n is the size of the tensor on that dimension
-// p is the number of splits of n
-// r is the rank of the mpi process in that mode. fiber rank
+/**
+ * The dimension a particular rank holds out of 
+ * the global dimension n across p processes.
+ * @param[in] n is the global size of the tensor on that dimension
+ * @param[in] p is the number of splits of n. 
+ *            Typically number of processes on a particular mode
+ * @param[in] r is the rank of the mpi process in that mode. fiber rank
+ */
 inline int itersplit(int n, int p, int r) {
   int split = (r < n % p) ? n / p + 1 : n / p;
   return split;
 }
 
-// n is the size of the tensor on that dimension
-// p is the number of splits of n.
-// r is the rank of the mpi process in that mode. fiber rank
+/**
+ * Returns the start idx of the current rank r
+ * for a global dimension n across p processes. 
+ * @param[in] n is the size of the tensor on that dimension
+ * @param[in] p is the number of splits of n.
+ * @param[in] r is the rank of the mpi process in that mode. fiber rank
+ */
 inline int startidx(int n, int p, int r) {
   int rem = n % p;
   int idx =

### 2018-12-18 common/luc.hpp

@@ -1,219 +1,0 @@
-/* Copyright Ramakrishnan Kannan 2017 */
-#ifndef COMMON_LUC_HPP_
-#define COMMON_LUC_HPP_
-
-#ifdef MPI_DISTNTF
-#include <mpi.h>
-#endif
-#include "common/ncpfactors.hpp"
-#include "nnls/bppnnls.hpp"
-#include "nnls/nnls.hpp"
-
-namespace planc {
-class LUC {
- private:
-  // create variable required for your local function.
-  NCPFactors *factors;
-  int i_k;
-  const algotype m_updalgo;
-
-  // ADMM auxiliary variables
-  NCPFactors *m_local_ncp_aux;
-  NCPFactors *m_local_ncp_aux_t;
-  NCPFactors *m_temp_local_ncp_aux_t;
-  MAT L;
-  MAT Lt;
-  MAT tempgram;
-
-  int admm_iter;
-  double tolerance;
-
-  MAT updateMU(const MAT &gram, const MAT &rhs, const int mode) {
-    MAT H(factors->factor(mode));
-    MAT temp = H * gram + EPSILON;
-    H = (H % rhs.t()) / temp;
-    return H.t();
-  }
-
-  MAT updateHALS(const MAT &gram, const MAT &rhs, const int mode) {
-    MAT H(factors->factor(mode));
-
-    // iterate over all columns of H
-    for (int i = 0; i < this->i_k; i++) {
-      VEC updHi = H.col(i) + ((rhs.row(i)).t() - H * gram.col(i));
-      fixNumericalError<VEC>(&updHi);
-      double normHi = arma::norm(updHi, 2);
-      normHi *= normHi;
-      double globalnormHi = normHi;
-#ifdef MPI_DISTNTF
-      MPI_Allreduce(&normHi, &globalnormHi, 1, MPI_DOUBLE, MPI_SUM,
-                    MPI_COMM_WORLD);
-#endif  // shared memory can just check normHi
-      if (globalnormHi > 0) {
-        H.col(i) = updHi;
-      }
-    }
-    return H.t();
-  }
-
-  MAT updateBPP(const MAT &gram, const MAT &rhs) {
-    BPPNNLS<MAT, VEC> subProblem(gram, rhs, true);
-    subProblem.solveNNLS();
-    return subProblem.getSolutionMatrix();
-  }
-
-  MAT updateAOADMM(const MAT &gram, const MAT &rhs, const int mode) {
[…]

### 2018-12-18 common/ncpfactors.hpp

@@ -10,25 +10,34 @@
 #include <mpi.h>
 #endif
 
-// ncp_factors contains the factors of the ncp
-// every ith factor is of size n_i * k
-// number of factors is called as mode of the tensor
-// all idxs are zero idx.
+/**
+ *  ncp_factors contains the factors of the ncp
+ * every ith factor is of size n_i * k
+ * number of factors is called as mode of the tensor
+ * all idxs are zero idx.
+ */
 
 namespace planc {
 
 class NCPFactors {
-  MAT *ncp_factors;
-  int m_modes;
-  int m_k;
-  UVEC m_dimensions;
-  // in the distributed mode all the processes has same lambda
+  MAT *ncp_factors;   /// Array of factors .One factor for every mode.
+  int m_modes;        /// Number of modes in tensor
+  int m_k;            /// Low rank
+  UVEC m_dimensions;  /// Vector of dimensions for every mode
+  /// in the distributed mode all the processes has same lambda
   VEC m_lambda;
-  // normalize the factors of a matrix
+  /// normalize the factors of a matrix
   bool freed_ncp_factors;
 
  public:
-  // constructors
+  /**
+   * constructor that takes the dimensions of every mode, low rank k
+   * All the factors will be initialized with random uniform number
+   * @param[in] vector of dimension for every mode.
+   * @param[in] low rank
+   * @param[in] trans. takes true or false. Transposes every factor.
+   */
+
   NCPFactors(const UVEC &i_dimensions, const int &i_k, bool trans) {
     this->m_dimensions = i_dimensions;
     this->m_modes = i_dimensions.n_rows;
@@ -84,23 +93,36 @@ ncp_factors[i].clear();
   }
 
   // getters
+  /// returns low rank
   int rank() const { return m_k; }
+  /// dimensions of every mode
   UVEC dimensions() const { return m_dimensions; }
+  /// factor matrix of a mode i_n
   MAT &factor(const int i_n) const { return ncp_factors[i_n]; }
+  /// returns number of modes
   int modes() const { return m_modes; }
+  /// returns the lambda vector
   VEC lambda() const { return m_lambda; }
 
   // setters
+  /**
+   * Set the mode i_n with the given factor matrix
+   * @param[in] i_n mode for which given factor matrix will be updated
+   * @param[in] i_factor factor matrix
+   */
[…]

### 2018-12-18 common/nmf.hpp

@@ -17,36 +17,32 @@
 // #define COLLECTSTATS 1
 // #endif
 
-namespace planc{
+namespace planc {
 
 // T must be a either an instance of MAT or sp_MAT
 template <class T>
 class NMF {
  protected:
-  // input MATrix of size mxn
-  T A;
-
-  // low rank factors with size mxk and nxk respectively.
-  MAT W, H;
+  T A;  /// input matrix of size mxn
+  MAT W, H;  /// left and low rank factors of size mxk and nxk respectively
   MAT Winit, Hinit;
-  UINT m, n, k;
+  UINT m, n, k;  /// rows, columns and lowrank  
 
   /*
    * Collected statistics are
    * iteration Htime Wtime totaltime normH normW densityH densityW relError
    */
   MAT stats;
-  double objective_err;
+  double objective_err;  /// objective error at any particular iteration
   double normA, normW, normH;
   double densityW, densityH;
   bool cleared;
-  int m_num_iterations;
-  std::string input_file_name;
+  int m_num_iterations;  /// number of iterations
+  std::string input_file_name; 
   MAT errMtx;       // used for error computation.
   T A_err_sub_mtx;  // used for error computation.
-
-  // The regularization is a vector of two values. The first value specifies
-  // L2 regularization values and the second is L1 regularization.
+  /// The regularization is a vector of two values. The first value specifies
+  /// L2 regularization values and the second is L1 regularization.
   FVEC m_regW;
   FVEC m_regH;
 
@@ -65,12 +61,13 @@ class NMF {
     this->stats(iteration, 8) = this->objective_err;
   }
 
-  /*
+  /**
    * For both L1 and L2 regularizations we only adjust the
    * HtH or WtW. The regularization is a vector of two values.
    * The first value specifies L2 regularization values
    * and the second is L1 regularization.
-   * Mostly we expect
+   * param[in] regularization as a vector
+   * param[out] Gram matrix
    */
   void applyReg(const FVEC &reg, MAT *AtA) {
     // Frobenius norm regularization
@@ -87,6 +84,10 @@ class NMF {
       (*AtA) = (*AtA) + 2 * lambda_l1 * onematrix;
     }
   }
+
+  /**
+   *  L2 normalize column vectors of W
+   */
 
   void normalize_by_W() {
     MAT W_square = arma::pow(this->W, 2);
@@ -110,6 +111,11 @@ class NMF {
   }
 
  public:
+  /**
[…]

### 2018-12-18 common/npyio.hpp

@@ -12,13 +12,13 @@
 namespace planc {
 class NumPyArray {
  private:
-  size_t m_word_size;
+  int64_t m_word_size;
   bool m_fortran_order;
-  size_t m_modes;
+  int64_t m_modes;
   UVEC m_dims;
   void parse_npy_header(FILE* fp) {
     char buffer[256];
-    size_t res = fread(buffer, sizeof(char), 11, fp);
+    int64_t res = fread(buffer, sizeof(char), 11, fp);
     if (res != 11) {
       ERR << "Something wrong. Could not read header " << std::endl;
       exit(-1);
@@ -31,13 +31,13 @@ class NumPyArray {
 
     // fortran order is column major order
     // C order is row major order
-    size_t loc1 = header.find("fortran_order");
+    int64_t loc1 = header.find("fortran_order");
     loc1 += 16;
     this->m_fortran_order = (header.substr(loc1, 4) == "True" ? true : false);
 
     // obtain dimensions
     loc1 = header.find("(");
-    size_t loc2 = header.find(")");
+    int64_t loc2 = header.find(")");
     if (loc1 < 0 || loc2 < 0) {
       ERR << "could not find ()" << std::endl;
       exit(-1);
@@ -54,7 +54,7 @@ class NumPyArray {
     std::stringstream ss(str_shape);
     std::string s;
     ss.str(str_shape);
-    size_t i = 0;
+    int64_t i = 0;
     while (getline(ss, s, ',')) {
       this->m_dims[i++] = ::atoi(s.c_str());
     }
@@ -87,7 +87,7 @@ class NumPyArray {
     }
     parse_npy_header(fp);
     this->m_input_tensor = new Tensor(this->m_dims);
-    size_t nread = fread(m_input_tensor->m_data, sizeof(m_input_tensor->m_data),
+    int64_t nread = fread(m_input_tensor->m_data, sizeof(m_input_tensor->m_data),
                          m_input_tensor->numel(), fp);
     if (nread != m_input_tensor->numel()) {
       WARN << "something wrong ::read::" << nread

### 2018-12-18 common/ntf_utils.hpp

@@ -6,9 +6,12 @@
 #include "common/ncpfactors.hpp"
 #include "common/tensor.hpp"
 
-// A is of size m x k
-// B is of size n x k
-// Returns C of size mn x k
+/**
+ * Returns the khatri-rao product between two matrices.
+ * @param[in] A is of size m x k matrix
+ * @param[in]  B is of size n x k
+ * @param[out] Returns C of size mn x k
+ */
 void khatrirao(const MAT &i_A, const MAT &i_B, MAT *o_C) {
   assert(i_A.n_cols == i_B.n_cols);
   VEC acol = arma::zeros<VEC>(i_A.n_rows);
@@ -22,12 +25,26 @@ void khatrirao(const MAT &i_A, const MAT &i_B, MAT *o_C) {
     }
   }
 }
+/**
+ * Returns the kronecker product between two vectors
+ * @param[in] acol column vector of size m
+ * @param[in] bcol is column vector of size n
+ * @param[out] output vector of size mn
+ */
 inline void kronecker(const VEC &i_acol, const VEC &i_bcol, VEC *o_c) {
   for (int j = 0; j < i_acol.n_rows; j++) {
     (*o_c).rows(arma::span(j * i_bcol.n_rows, (j + 1) * i_bcol.n_rows - 1)) =
         i_acol(j) * i_bcol;
   }
 }
+/**
+ * Return the mttkrp of mode i_n of tensor X. That is., determine
+ * the KRP leaving out i_n and multiply with mode i_n factor
+ * @param[in] mode i_n
+ * @param[in] Tensor X
+ * @param[in] NCPFactors
+ * @param[out] MTTKRP of factor mode i_n 
+ */
 
 void mttkrp(const int i_n, const planc::Tensor &X, planc::NCPFactors &i_F,
             MAT *o_mttkrp) {

### 2018-12-18 common/parsecommandline.hpp

@@ -83,6 +83,12 @@ class ParseCommandLine {
   }
 
  public:
+  /**
+   * Constructor that takes the number of arguments and the 
+   * command line parameters.
+   * @param[in] argc - number of arguments
+   * @param[in] **argv - command line parameters.
+   */ 
   ParseCommandLine(int argc, char **argv) : m_argc(argc), m_argv(argv) {
     this->m_num_modes = 0;
     this->m_pr = 1;
@@ -95,9 +101,9 @@ class ParseCommandLine {
     this->m_lucalgo = ANLSBPP;
     this->m_compute_error = 0;
     this->m_input_normalization = NONE;
-    this->m_dim_tree = 0;
+    this->m_dim_tree = 1;
   }
-
+  /// parses the command line parameters
   void parseplancopts() {
     int opt, long_index;
     while ((opt = getopt_long(this->m_argc, this->m_argv,
@@ -169,6 +175,8 @@ class ParseCommandLine {
     }
   }
 
+  /// print the configuration received through the command line paramters
+
   void printConfig() {
     std::cout << "a::" << this->m_lucalgo << "::i::" << this->m_Afile_name
               << "::k::" << this->m_k << "::m::" << this->m_globalm
@@ -216,26 +224,76 @@ class ParseCommandLine {
          << "-i Ainput -o nmfoutput -t 10 -p \"3 2\" --sparsity=0.3"
          << "-r \"0.0001 0 0 0.0001\" " << std::endl;
   }
-
+  /// returns the low rank. Passed as parameter --lowrank or -k
   UWORD lowrankk() { return m_k; }
+  /// return global rows. Passed as parameter -d
   UWORD globalm() { return m_globalm; }
+  //// returns the global columns. Passed as parameter -d
   UWORD globaln() { return m_globaln; }
+  /**
+   * L2 regularization as the first parameter and L1 as second 
+   * for left lowrank factor W. Passed as parameter --regularizer 
+   * with pair of values in double quotes for W and H "l2W l1W l2H l1H"
+   */  
   FVEC regW() { return m_regW; }
+  /**
+   * L2 regularization as the first parameter and L1 as second 
+   * for right lowrank factor H. Passed as parameter --regularizer 
+   * with pair of values in double quotes for W and H "l2W l1W l2H l1H"
+   */  
   FVEC regH() { return m_regH; }
+  /// Returns the NMF algorithm to run. Passed as parameter --algo or -a
[…]

### 2018-12-18 common/tensor.hpp

@@ -13,7 +13,7 @@
 #include "common/utils.h"
 
 namespace planc {
-/*
+/**
  * Data is stored such that the unfolding \f$Y_0\f$ is column
  * major.  This means the flattening \f$Y_{N-1}\f$ is row-major,
  * and any other flattening \f$Y_n\f$ can be represented as a set
@@ -76,6 +76,9 @@ class Tensor {
     this->m_data = NULL;
     freed_on_destruction = false;
   }
+  /**
+   * Constructor that takes only dimensions of every mode as a vector
+   */
   Tensor(const UVEC &i_dimensions)
       : m_dimensions(i_dimensions),
         m_modes(i_dimensions.n_rows),
@@ -95,8 +98,10 @@ class Tensor {
     freed_on_destruction = false;
     randu();
   }
-  // Need when copying from matrix to Tensor.
-  // otherwise copy constructor will be called.
+  /**
+   * Need when copying from matrix to Tensor. otherwise copy constructor
+   * will be called. The data will be passed in row major order
+   */
   Tensor(const UVEC &i_dimensions, double *i_data)
       : m_dimensions(i_dimensions),
         m_modes(i_dimensions.n_rows),
@@ -112,7 +117,10 @@ class Tensor {
       freed_on_destruction = true;
     }
   }
-  // copy constructor
+
+  /**
+   *  copy constructor
+   */
   Tensor(const Tensor &src) {
     clear();
     this->m_numel = src.numel();
@@ -137,6 +145,10 @@ class Tensor {
     }
     return *this;
   }
+  /**
+   * Clears the data and also destroys the storage
+   */
+
   void clear() {
     if (this->m_numel > 0 && this->m_data != NULL) {
       delete[] this->m_data;  // destroy storage in this
@@ -146,42 +158,63 @@ class Tensor {
     }
   }
 
+  /// Return the number of modes. It is a scalar value.
   int modes() const { return m_modes; }
+  /// Returns a vector of dimensions on every mode
   UVEC dimensions() const { return m_dimensions; }
   UVEC global_idx() const { return m_global_idx; }
+
+  /**
+   * Returns the dimension of the input mode.
+   * @param[in] a mode of the tensor.
+   * @return dimension of ith mode
+   */
+
   int dimension(int i) const { return m_dimensions[i]; }
-  int numel() const { return m_numel; }
+  /// Returns total number of elements
+  UWORD numel() const { return m_numel; }
 
[…]

### 2018-12-18 common/utils.hpp

@@ -37,8 +37,14 @@ static ULONG powersof10[16] = {1,
 static std::stack<std::chrono::steady_clock::time_point> tictoc_stack;
 static std::stack<double> tictoc_stack_omp_clock;
 
+
+/// start the timer. easy to call as tic(); some code; double t=toc();
 inline void tic() { tictoc_stack.push(std::chrono::steady_clock::now()); }
 
+/***
+ * Returns the time taken between the most recent tic() to itself.
+ * @return time in seconds.  
+*/
 inline double toc() {
   std::chrono::duration<double> time_span =
       std::chrono::duration_cast<std::chrono::duration<double>>(

### 2018-12-18 dimtree/ddt.hpp

@@ -40,7 +40,7 @@ class DenseDimensionTree {
     m_local_T->dims = (long int *)malloc(sizeof(long int) * m_local_T->nmodes);
     m_local_Y->nmodes = i_ncp_factors.modes();
     m_local_Y->rank = i_ncp_factors.rank();
-    ROWVEC temp_vec = arma::prod(i_ncp_factors.lambda());
+    VEC temp_vec = i_ncp_factors.lambda();
     m_local_Y->lambdas = temp_vec.memptr();
     m_local_Y->dims_product = arma::prod(i_ncp_factors.dimensions());
 

### 2018-12-18 distnmf/aunmf.hpp

@@ -10,15 +10,15 @@
 #include "distnmf/distnmf.hpp"
 #include "distnmf/mpicomm.hpp"
 
-/*
+/**
  * There are totally prxpc process.
  * Each process will hold the following
- * An A of size (globalm/pr) x (globaln/pc)
- * Here each process m=globalm/pr and n=globaln/pc
- * H of size (globaln/p)xk
- * W of size (globalm/p)xk
- * A is mxn matrix
- * H is nxk matrix
+ * An A of size \f$\frac{globalm}{p_r} \times \frac{globaln}{p_c}\f$
+ * Here each process \f$m=\frac{globalm}{p_r} and n=\frac{globaln}{p_c}\f$
+ * H of size \f$\frac{globaln}{p} \times k\f$
+ * W of size \f${globalm}{p} \times k\f$
+ * A is \f$m \times n\f$ matrix
+ * H is \f$n \times k\f$ matrix
  */
 namespace planc {
 
@@ -27,32 +27,32 @@ class DistAUNMF : public DistNMF<INPUTMATTYPE> {
   // needed in derived algorithms to
   // call BPP routines
  protected:
-  MAT HtH;    // H is of size (globaln/p)*k;
-  MAT WtW;    // W is of size (globaln/p)*k;
-  MAT AHtij;  // AHtij is of size k*(globalm/p)
-  MAT WtAij;  // WtAij is of size k*(globaln/p)
-  MAT Wt;     // Wt is of size k*(globalm/p)
-  MAT Ht;     // Ht is of size k*(globaln/p)
+  MAT HtH;    /// H is of size (globaln/p)*k;
+  MAT WtW;    /// W is of size (globaln/p)*k;
+  MAT AHtij;  /// AHtij is of size k*(globalm/p)
+  MAT WtAij;  /// WtAij is of size k*(globaln/p)
+  MAT Wt;     /// Wt is of size k*(globalm/p)
+  MAT Ht;     /// Ht is of size k*(globaln/p)
 
   virtual void updateW() = 0;
   virtual void updateH() = 0;
 
  private:
   // Things needed while solving for W
-  MAT localHtH;         // H is of size (globaln/p)*k;
-  MAT Hjt, Hj;          // Hj is of size n*k;
-  MAT AijHj, AijHjt;    // AijHj is of size m*k;
-  INPUTMATTYPE A_ij_t;  // n*m
[…]

### 2018-12-18 distnmf/distals.hpp

@@ -4,18 +4,23 @@
 #define DISTNMF_DISTALS_HPP_
 #include "distnmf/aunmf.hpp"
 
+/**
+ * Unconstrained least squares. Should match the SVD
+ * objective error.  
+ */
+
 namespace planc {
 
 template <class INPUTMATTYPE>
 class DistALS : public DistAUNMF<INPUTMATTYPE> {
  protected:
-  // update W given HtH and AHt
+  /**
+   * update W given HtH and AHt
+   * AHtij is of size \f$ k \times \frac{globalm}/{p} \f$.
+   * this->W is of size \f$ \frac{globalm}{p} \times k \f$
+   * this->HtH is of size kxk
+  */
   void updateW() {
-    // AHtij is of size k*(globalm/p).
-    // this->W is of size (globalm/p)xk
-    // this->HtH is of size kxk
-    // w_ij = w_ij .* (AH)_ij/(WHtH)_ij
-    // Here ij is the element of W matrix.
     this->Wt = arma::solve(arma::trimatl(this->HtH), this->AHtij);
     this->W = this->Wt.t();
     DISTPRINTINFO("ALS::updateW::HtH::" << PRINTMATINFO(this->HtH)
@@ -26,13 +31,13 @@ class DistALS : public DistAUNMF<INPUTMATTYPE> {
                                         << norm(this->AHtij, "fro")
                                         << "::W::" << norm(this->W, "fro"));
   }
-
+  /**
+   * updateH given WtAij and WtW
+   *  WtAij is of size \f$k \times \frac{globaln}{p} \f$
+   * this->H is of size \f$ \frac{globaln}{p} \times k\f$
+   * this->WtW is of size kxk
+   */
   void updateH() {
-    // WtAij is of size k*(globaln/p)
-    // this->H is of size (globaln/p)xk
-    // this->WtW is of size kxk
-    // h_ij = h_ij .* WtAij.t()/(HWtW)_ij
-    // Here ij is the element of H matrix.
     this->Ht = arma::solve(arma::trimatl(this->WtW), this->WtAij);
     this->H = this->Ht.t();
     DISTPRINTINFO("ALS::updateH::WtW::" << PRINTMATINFO(this->WtW)

### 2018-12-18 distnmf/distanlsbpp.hpp

@@ -6,6 +6,11 @@
 #include "distnmf/aunmf.hpp"
 #include "nnls/bppnnls.hpp"
 
+/**
+ * Provides the updateW and updateH for the
+ * distributed ANLS/BPP algorithm.
+ */
+
 #ifdef BUILD_CUDA
 #define ONE_THREAD_MATRIX_SIZE 1000
 #include <omp.h>
@@ -23,6 +28,9 @@ class DistANLSBPP : public DistAUNMF<INPUTMATTYPE> {
 
   void allocateMatrices() {}
 
+  /**
+   * Multi threaded ANLS/BPP using openMP
+   */
   void updateOtherGivenOneMultipleRHS(const MAT& giventGiven,
                                       const MAT& giventInput, MAT* othermat) {
     UINT numThreads = (giventInput.n_cols / ONE_THREAD_MATRIX_SIZE) + 1;
@@ -75,34 +83,24 @@ class DistANLSBPP : public DistAUNMF<INPUTMATTYPE> {
   }
 
  protected:
-  // updateW given HtH and AHt
+  /**
+   * update W given HtH and AHt
+   * AHtij is of size \f$ k \times \frac{globalm}/{p}\f$.
+   * this->W is of size \f$\frac{globalm}{p} \times k \f$
+   * this->HtH is of size kxk
+  */
   void updateW() {
     updateOtherGivenOneMultipleRHS(this->HtH, this->AHtij, &this->W);
-    // BPPNNLS<MAT, VEC> subProblem(this->HtH, this->AHtij, true);
-    // subProblem.solveNNLS();
-    // this->Wt = subProblem.getSolutionMatrix();
-    // fixNumericalError<MAT>(&(this->Wt));
     this->Wt = this->W.t();
-
-    // localWnorm = sum(this->W % this->W);
-    // tic();
-    // MPI_Allreduce(localWnorm.memptr(), Wnorm.memptr(), this->k, MPI_FLOAT,
-    //               MPI_SUM, MPI_COMM_WORLD);
-    // double temp = toc();
-    // this->time_stats.allgather_duration(temp);
-    // for (int i = 0; i < this->k; i++) {
-    //     this->W.col(i) = this->W.col(i) / sqrt(Wnorm(i));
-    // }
-    // this->Wt = this->W.t();
   }
-
-  // updateH given WtW and WtA
+  /**
+   * updateH given WtAij and WtW
[…]

### 2018-12-18 distnmf/distaoadmm.hpp

@@ -76,7 +76,7 @@ class DistAOADMM : public DistAUNMF<INPUTMATTYPE> {
   }
 
  protected:
-  // updateW given HtH and AHt
+  /// updateW given HtH and AHt
   void updateW() {
     // Calculate modified Gram Matrix
     // tempHtH = arma::conv_to<MAT >::from(this->HtH);
@@ -158,7 +158,7 @@ class DistAOADMM : public DistAUNMF<INPUTMATTYPE> {
         stop_iter = true;
     }
   }
-  // updateH given WtW and WtA
+  /// updateH given WtW and WtA
   void updateH() {
     // Calculate the Gram Matrix
     tempWtW = arma::conv_to<MAT>::from(this->WtW);

### 2018-12-18 distnmf/disthals.hpp

@@ -4,24 +4,26 @@
 #define DISTNMF_DISTHALS_HPP_
 
 #include "distnmf/aunmf.hpp"
+/**
+ * emulating Jingu's code
+ * https://github.com/kimjingu/nonnegfac-matlab/blob/master/nmf.m
+ * function hals_iterSolver
+ */
 
 namespace planc {
 
 template <class INPUTMATTYPE>
 class DistHALS : public DistAUNMF<INPUTMATTYPE> {
  protected:
-  // emulating Jingu's code
-  // https://github.com/kimjingu/nonnegfac-matlab/blob/master/nmf.m
-  // function hals_iterSolver
-
+  /**
+   * AHtij is of size \f$ k \times \frac{globalm}/{p}\f$.
+   * this->W is of size \f$\frac{globalm}{p} \times k\f$
+   * this->HtH is of size kxk
+   * Eq 14(a) page 7 of JGO paper
+   * \f$W(:,i)=[W(:,i) + (AH(:,i)-WH^TH(:,i))/H^TH(i,i)]_+\f$
+   * column normalize W_i
+   */
   void updateW() {
-    // AHtij is of size k*(globalm/p).
-    // this->W is of size (globalm/p)xk
-    // this->HtH is of size kxk
-    // Eq 14(a) page 7 of JGO paper
-    // W(:,i)=[W(:,i) + (AH(:,i)-W*HtH(:,i))/HtH(i,i)]_+
-    // column normalize W_i
-    // Here ij is the element of W matrix.
     for (int i = 0; i < this->k; i++) {
       // W(:,i) = max(W(:,i) * HHt_reg(i,i) + AHt(:,i) - W *
       // HHt_reg(:,i),epsilon);
@@ -54,13 +56,15 @@ class DistHALS : public DistAUNMF<INPUTMATTYPE> {
     this->Wt = this->W.t();
   }
 
+  /**
+   * WtAij is of size \f$k \times \frac{globaln}{p} \f$
+   * this->H is of size \f$ \frac{globaln}{p} \times k\f$
+   * this->WtW is of size kxk
+   * Eq 14(b) page 7 of JGO paper
+   * \f$ H(:,i) = H(:,i) + WtAij(:,i) - HW^TW(:,i)\f$
+   * Here ij is the element of H matrix.
+   */
   void updateH() {
-    // WtAij is of size k*(globaln/p)
-    // this->H is of size (globaln/p)xk
[…]

### 2018-12-18 distnmf/distio.hpp

@@ -9,7 +9,7 @@
 #include "common/distutils.hpp"
 #include "distnmf/mpicomm.hpp"
 
-/*
+/**
  * File name formats
  * A is the filename
  * 1D distribution Arows_totalpartitions_rank or Acols_totalpartitions_rank
@@ -26,9 +26,9 @@ template <class MATTYPE>
 class DistIO {
  private:
   const MPICommunicator& m_mpicomm;
-  MATTYPE m_Arows;  // ONED_ROW and ONED_DOUBLE
-  MATTYPE m_Acols;  // ONED_COL and ONED_DOUBLE
-  MATTYPE m_A;      // TWOD
+  MATTYPE m_Arows;  /// ONED_ROW and ONED_DOUBLE
+  MATTYPE m_Acols;  /// ONED_COL and ONED_DOUBLE
+  MATTYPE m_A;      /// TWOD
   // don't start getting prime number from 2;
   static const int kPrimeOffset = 10;
   // Hope no one hits on this number.
@@ -42,7 +42,7 @@ class DistIO {
 #endif
 
   const iodistributions m_distio;
-  /*
+  /**
    * A random matrix is always needed for sparse case
    * to get the pattern. That is., the indices where
    * numbers are being filled. We will use this pattern
@@ -138,7 +138,7 @@ class DistIO {
   }
 #endif
 
-  /*
+  /**
    * Uses the pattern from the input matrix X but
    * the value is computed as low rank.
    */
@@ -311,11 +311,21 @@ class DistIO {
  public:
   DistIO<MATTYPE>(const MPICommunicator& mpic, const iodistributions& iod)
       : m_mpicomm(mpic), m_distio(iod) {}
-  /*
+  /**
    * We need m,n,pr,pc only for rand matrices. If otherwise we are
    * expecting the file will hold all the details.
    * If we are loading by file name we dont need distio flag.
-   *
+   * If the filename is rand_lowrank/rand_uniform, appropriate
+   * random functions will be called. Otherwise, it will be loaded from file.
+   * @param[in] file_name. For random matrices rand_lowrank/rand_uniform
+   * @param[in] m - globalm. Needed only for random matrices. 
+   *                otherwise, we will know from file.
+   * @param[in] n - globaln. Needed only for random matrices
+   * @param[in] k - low rank. Used for generating synthetic lowrank matrices.
+   * @param[in] sparsity - sparsity factor between 0-1 for sparse matrices.
[…]

### 2018-12-18 distnmf/distmu.hpp

@@ -4,6 +4,12 @@
 #define DISTNMF_DISTMU_HPP_
 #include "distnmf/aunmf.hpp"
 
+/**
+ * Distributed MU factorization.
+ * Offers implementation for the pure virtual function
+ * updateW and updateH based on MU.
+ */
+
 namespace planc {
 
 template <class INPUTMATTYPE>
@@ -14,13 +20,15 @@ class DistMU : public DistAUNMF<INPUTMATTYPE> {
   ROWVEC Wnorm;
 
  protected:
-  // update W given HtH and AHt
+  /**
+   * update W given HtH and AHt
+   * AHtij is of size \f$ k \times \frac{globalm}/{p}\f$.
+   * this->W is of size \f$\frac{globalm}{p} \times k\f$
+   * this->HtH is of size kxk
+   * \f$w_{ij} = w_{ij} .* \frac{(AH)_{ij}}{(WH^TH)_{ij}}\f$
+   * Here ij is the element of W matrix.
+   */
   void updateW() {
-    // AHtij is of size k*(globalm/p).
-    // this->W is of size (globalm/p)xk
-    // this->HtH is of size kxk
-    // w_ij = w_ij .* (AH)_ij/(WHtH)_ij
-    // Here ij is the element of W matrix.
     WHtH = (this->W * this->HtH) + EPSILON;
 #ifdef MPI_VERBOSE
     DISTPRINTINFO("::WHtH::" << endl << this->WHtH);
@@ -50,13 +58,15 @@ class DistMU : public DistAUNMF<INPUTMATTYPE> {
        }*/
     this->Wt = this->W.t();
   }
-
+  /**
+   * updateH given WtAij and WtW
+   * WtAij is of size \f$k \times \frac{globaln}{p} \f$
+   * this->H is of size \f$ \frac{globaln}{p} \times k\f$
+   * this->WtW is of size kxk
+   * \f$h_{ij} = \frac{h_{ij} .* WtAij.t()}{(HW^TW)_{ij}}\f$
+   * Here ij is the element of H matrix.
+   */  
   void updateH() {
-    // WtAij is of size k*(globaln/p)
-    // this->H is of size (globaln/p)xk
-    // this->WtW is of size kxk
-    // h_ij = h_ij .* WtAij.t()/(HWtW)_ij
-    // Here ij is the element of H matrix.
     HWtW = this->H * this->WtW + EPSILON;
     this->H = (this->H % this->WtAij.t()) / HWtW;
 #ifdef MPI_VERBOSE

### 2018-12-18 distnmf/distnmf.hpp

@@ -11,18 +11,7 @@
 #include "pacoss.h"
 #endif
 
-/*
- * There are totally prxpc process.
- * Each process will hold the following
- *   An A of size (m/pr) x (n/pc)
- * H of size (n/p)xk
- * W of size (m/p)xk
- * A is mxn matrix
- * H is nxk matrix
- */
-
 namespace planc {
-
 template <typename INPUTMATTYPE>
 class DistNMF : public NMF<INPUTMATTYPE> {
  protected:
@@ -43,6 +32,14 @@ class DistNMF : public NMF<INPUTMATTYPE> {
   ROWVEC Wnorm;
 
  public:
+  /**
+   * There are totally prxpc process.
+   * Each process will hold the following
+   * @param[in] A of size \f$\frac{globalm}{p_r} \times \frac{globaln}{p_c}\f$
+   * @param[in] right low rank factor H of size \f$\frac{globaln}{p} \times k\f$
+   * @param[in] left low rank factor W of size \f$\frac{globalm}{p} \times k\f$
+   * @param[in] MPI Communicator for row and column communicators
+   */  
   DistNMF(const INPUTMATTYPE &input, const MAT &leftlowrankfactor,
           const MAT &rightlowrankfactor, const MPICommunicator &communicator)
       : NMF<INPUTMATTYPE>(input, leftlowrankfactor, rightlowrankfactor),
@@ -83,12 +80,19 @@ class DistNMF : public NMF<INPUTMATTYPE> {
     this->m_colcomm = colcomm;
   }
 #endif
+  /// returns globalm
   const int globalm() const { return m_globalm; }
+  /// returns globaln
   const int globaln() const { return m_globaln; }
+  /// returns global squared norm of A
   const double globalsqnorma() const { return m_globalsqnormA; }
+  /// return the current error
   void compute_error(const uint &ce) { this->m_compute_error = ce; }
+  /// returns the flag to compute error or not.
   const bool is_compute_error() const { return (this->m_compute_error); }
+  /// returns the NMF algorithm
   void algorithm(algotype dat) { this->m_algorithm = dat; }
+  /// Reports the time
   void reportTime(const double temp, const std::string &reportstring) {
     double mintemp, maxtemp, sumtemp;
     MPI_Allreduce(&temp, &maxtemp, 1, MPI_DOUBLE, MPI_MAX, MPI_COMM_WORLD);
@@ -102,6 +106,7 @@ class DistNMF : public NMF<INPUTMATTYPE> {
[…]

### 2018-12-18 distnmf/distnmftime.hpp

@@ -3,6 +3,10 @@
 #ifndef DISTNMF_DISTNMFTIME_HPP_
 #define DISTNMF_DISTNMFTIME_HPP_
 
+/**
+ * Class and function for collecting time statistics 
+ */
+
 namespace planc {
 
 class DistNMFTime {

### 2018-12-18 distnmf/mpicomm.hpp

@@ -11,14 +11,21 @@
 #include "pacoss/pacoss.h"
 #endif
 
+/**
+ * Class and function for 2D MPI communicator
+ * with row and column communicators
+ */
+
 namespace planc {
 
 class MPICommunicator {
  private:
-  int m_rank;
-  int m_numProcs;
-  int m_row_rank, m_row_size;
-  int m_col_rank, m_col_size;
+  int m_rank;  /// Local Rank
+  int m_numProcs;  /// Total number of mpi process
+  int m_row_rank;
+  int m_row_size; 
+  int m_col_rank
+  int m_col_size;
   int m_pr, m_pc;
 
   // for 2D communicators
@@ -107,11 +114,17 @@ class MPICommunicator {
     printConfig();
 #endif
   }
+  /// returns the global rank
   const int rank() const { return m_rank; }
+  /// returns the total number of mpi processes
   const int size() const { return m_numProcs; }
+  /// returns its rank in the row processor grid
   const int row_rank() const { return m_row_rank; }
+  /// returns the rank in the column processor grid
   const int col_rank() const { return m_col_rank; }
+  /// Total number of row processors
   const int pr() const { return m_pr; }
+  /// Total number of column processor
   const int pc() const { return m_pc; }
   const MPI_Comm *commSubs() const { return m_commSubs; }
 };

### 2018-12-18 distntf/distauntf.hpp

@@ -11,7 +11,7 @@
 #include "distntf/distntfmpicomm.hpp"
 #include "distntf/distntftime.hpp"
 
-/*
+/**
  * Tensor A of size is M1 x M2 x... x Mn is distributed among
  * P1 x P2 x ... x Pn grid of P processors. That means, every
  * processor has (M1/P1) x (M2/P2) x ... x (Mn/Pn) tensor as
@@ -82,9 +82,13 @@ class DistAUNTF {
   MAT hadamard_all_grams;
 
   DenseDimensionTree *kdt;
-  // do the local syrk only for the current updated factor
-  // and all reduce only for the current updated factor.
-  // computes G^(current_mode)
+
+  /**
+   * do the local syrk only for the current updated factor
+   * and all reduce only for the current updated factor.
+   * computes G^(current_mode)
+   * @param[in] current_mode
+   */
   void update_global_gram(const int current_mode) {
     // computing U
     MPITIC;  // gram
@@ -110,6 +114,13 @@ class DistAUNTF {
     this->time_stats.allreduce_duration(temp);
   }
 
+  /**
+   * Apply L1 and L2 regularization on the gram of hadamards
+   * @param[in] L1 - L1 Regularization parameters as a float
+   * @param[in] L2 - L2 Regularization parameters as a float
+   * @param[out] AtA - L1/L2 regularization applied gram matrix
+   */
+
   void applyReg(float lambda_l2, float lambda_l1, MAT *AtA) {
     // Frobenius norm regularization
     if (lambda_l2 > 0) {
@@ -124,8 +135,11 @@ class DistAUNTF {
     }
   }
 
-  /*
-   * This iterates over all grams and find hadamard of the grams
+  /**
+   * This iterates over all grams leaving out current mode
+   * and finds hadamard of the grams
+   * 
+   * @param[in] current_mode. 
    */
   void gram_hadamard(int current_mode) {
     global_gram.ones();
@@ -140,8 +154,13 @@ class DistAUNTF {
     this->time_stats.compute_duration(temp);
     this->time_stats.gram_duration(temp);
   }
+  /**
+   * factor matrices all gather only on the current updated factor
+   * It performs the gather over the slice communicator
+   * @param[in] current_mode
+   */
 
-  // factor matrices all gather only on the current update factor
+  // 
   void gather_ncp_factor(const int current_mode) {
     m_gathered_ncp_factors_t.factor(current_mode).zeros();
[…]

### 2018-12-18 distntf/distntf.cpp

@@ -203,6 +203,7 @@ class DistNTF {
   }
 
  public:
+  /// Driver function for the distntf
   DistNTF(int argc, char *argv[]) {
     this->m_argc = argc;
     this->m_argv = argv;

### 2018-12-18 distntf/distntfanlsbpp.hpp

@@ -12,6 +12,13 @@ namespace planc {
 
 class DistNTFANLSBPP : public DistAUNTF {
  protected:
+  /**
+   * This is openmp multithreaded ANLS/BPP update function.
+   * Given the MTTKRP and the hadamard of all the grams, we 
+   * determine the factor matrix to be updated. 
+   * @param[in] Mode of the factor to be updated
+   * @returns The new updated factor
+   */
   MAT update(const int mode) {
     MAT othermat(this->m_local_ncp_factors_t.factor(mode));
     if (m_nls_sizes[mode] > 0) {

### 2018-12-18 distntf/distntfaoadmm.hpp

@@ -20,6 +20,13 @@ class DistNTFAOADMM : public DistAUNTF {
   double tolerance;
 
  protected:
+   /**
+   * This is openmp multithreaded ANLS/BPP update function.
+   * Given the MTTKRP and the hadamard of all the grams, we 
+   * determine the factor matrix to be updated. 
+   * @param[in] Mode of the factor to be updated
+   * @returns The new updated factor
+   */
   MAT update(const int mode) {
     // return variable
     MAT updated_fac(this->m_local_ncp_factors.factor(mode));

### 2018-12-18 distntf/distntfcpals.hpp

@@ -9,6 +9,13 @@ namespace planc {
 
 class DistNTFCPALS : public DistAUNTF {
  protected:
+  /**
+   * This is unconstrained CP ALS update function.
+   * Given the MTTKRP and the hadamard of all the grams, we 
+   * determine the factor matrix to be updated. 
+   * @param[in] Mode of the factor to be updated
+   * @returns The new updated factor
+   */
   MAT update(const int mode) {
     MAT Ht = this->m_local_ncp_factors.factor(mode).t();
     if (m_nls_sizes[mode] > 0) {

### 2018-12-18 distntf/distntfhals.hpp

@@ -9,6 +9,13 @@ namespace planc {
 
 class DistNTFHALS : public DistAUNTF {
  protected:
+   /**
+   * This is HALS update function.
+   * Given the MTTKRP and the hadamard of all the grams, we 
+   * determine the factor matrix to be updated. 
+   * @param[in] Mode of the factor to be updated
+   * @returns The new updated factor
+   */
   MAT update(const int mode) {
     MAT H(this->m_local_ncp_factors.factor(mode));
     // iterate over all columns of H

### 2018-12-18 distntf/distntfio.hpp

@@ -237,13 +237,15 @@ class DistNTFIO {
     rc.clear();
     return this->m_global_dims;
   }
-  /*
-    Reading from real input file.
-    Expecting a .tensor text file and .bin file.
-    .info text file will be written by root processor alone
-    .bin file will be an mpi io file
-    Read distributed tensor
-*/
+  /**
+   * Writes distributed tensor.
+   * Expecting a .tensor text file and .bin file.
+   * .info text file will be written by root processor alone
+   * .bin file will be an mpi io file
+   * Read distributed tensor
+   * @param[in] filename
+   * @param[in] local_tensor
+   */
   void write_dist_tensor(const std::string filename,
                          const Tensor &local_tensor) {
     // all processes reading the file_name.info file.
@@ -371,7 +373,9 @@ class DistNTFIO {
         MAT current_factor = factort.t();
         current_factor.save(sw.str(), arma::raw_ascii);
       }
+      sw.clear();
     }
+    sw.clear();
     sw << output_file_name << "_lambda"
        << "_" << MPI_SIZE;
     if (ISROOT) {

### 2018-12-18 distntf/distntfmpicomm.hpp

@@ -16,10 +16,9 @@ class NTFMPICommunicator {
   int m_num_procs;
   UVEC m_proc_grids;
   MPI_Comm m_cart_comm;
-  // for mode communicators (*,...,p_n,...,*)
+  /// for mode communicators (*,...,p_n,...,*)
   MPI_Comm *m_fiber_comm;
-  // all communicators other than the given mode.
-  // (p1,p2,...,p_n-1,*,p_n,...,p_M)
+  /// all communicators other than the given mode (p1,p2,..,p_n-1,*,p_n,..,p_M)
   MPI_Comm *m_slice_comm;
   UVEC m_fiber_ranks;
   UVEC m_slice_ranks;
@@ -64,6 +63,9 @@ class NTFMPICommunicator {
     }
   }
 
+  /**
+   * Constructor for setting up the nD grid communicators
+   */
   NTFMPICommunicator(int argc, char *argv[], const UVEC &i_dims)
       : m_proc_grids(i_dims) {
     // Get the number of MPI processes
@@ -141,32 +143,38 @@ class NTFMPICommunicator {
   const MPI_Comm &cart_comm() const { return m_cart_comm; }
   void coordinates(int *o_c) const {
     MPI_Cart_coords(m_cart_comm, m_global_rank, MPI_CART_DIMS, o_c);
-  }
-
+  }  
   std::vector<int> coordinates() const { return this->m_coords; }
-
+  /// Returns the fiber communicator. 
   const MPI_Comm &fiber(const int i) const { return m_fiber_comm[i]; }
+  /// Returns the slice communicator
   const MPI_Comm &slice(const int i) const { return m_slice_comm[i]; }
+  /// Returns the rank of current MPI process given the cartesian coordinates
   int rank(const int *i_coords) const {
     int my_rank;
     MPI_Cart_rank(m_cart_comm, i_coords, &my_rank);
     return my_rank;
   }
+  /// Returns the number of MPI processors
   int size() const { return m_num_procs; }
   int size(const int i_d) const {
     int n_procs;
     MPI_Comm_size(m_slice_comm[i_d], &n_procs);
     return n_procs;
   }
+  /// Returns the fiber rank on a particular fiber grid
   int fiber_rank(int i) const { return m_fiber_ranks[i]; }
+  /// Returns the slice rank on a particular slice grid.
   int slice_rank(int i) const { return m_slice_ranks[i]; }
+  /// Returns the process grid for which the communicators are setup
   UVEC proc_grids() const { return this->m_proc_grids; }
[…]

### 2018-12-18 distntf/distntfmu.hpp

@@ -9,6 +9,13 @@ namespace planc {
 
 class DistNTFMU : public DistAUNTF {
  protected:
+ /**
+   * This is MU update function.
+   * Given the MTTKRP and the hadamard of all the grams, we 
+   * determine the factor matrix to be updated. 
+   * @param[in] Mode of the factor to be updated
+   * @returns The new updated factor
+   */
   MAT update(const int mode) {
     MAT H(this->m_local_ncp_factors.factor(mode));
     if (m_nls_sizes[mode] > 0) {

### 2018-12-18 distntf/distntfnes.hpp

@@ -133,6 +133,8 @@ class DistNTFNES : public DistAUNTF {
 
   MAT update(const int mode) {
     double L, mu, lambda, q, alpha, alpha_prev, beta;
+    int outer_iter = this->current_it();
+    int iter = 0;
     if (mode == 0) {
       m_prev_t.set_lambda(m_prox_t.lambda());
     }
@@ -162,6 +164,7 @@ class DistNTFNES : public DistAUNTF {
                              (modified_gram * m_acc_t.factor(mode)));
       if (stop_iter(mode)) break;
 
+      iter++;
       Htprev = Ht;
       Ht = m_acc_t.factor(mode) - ((1 / (L + lambda)) * m_grad_t.factor(mode));
       fixNumericalError<MAT>(&Ht, EPSILON_1EMINUS16);
@@ -176,6 +179,9 @@ class DistNTFNES : public DistAUNTF {
       Ht.zeros();
       m_grad_t.set(mode, Ht);
     }
+    PRINTROOT("Nesterov Update::mode::" << mode
+              << "::outer_iter::" << outer_iter
+              << "::NLS inner_iter::" << iter);
     m_prox_t.set(mode, Ht);
     m_prox_t.distributed_normalize_rows(mode);
 

### 2018-12-18 utilities/py2mpiio.cpp

@@ -4,8 +4,9 @@
 
 /*
 * For converting numpy array to distributed io
-* Compile as g++ py2mpiio.cpp  -I/ccs/home/ramki/nmflibrary/ 
-* -I$ARMADILLO_INCLUDE_DIR  -I/ccs/home/ramki/rhea/libraries/openblas/include/
+* Compile as g++ -D_FILE_OFFSET_BITS=64 py2mpiio.cpp 
+* -I/ccs/home/ramki/nmflibrary/  -I$ARMADILLO_INCLUDE_DIR  
+* -I/ccs/home/ramki/rhea/libraries/openblas/include/
 */
 
 int main(int argc, char *argv[]) {

### 2019-01-11 distnmf/conv-mat-to-ss.py
dist ntf was producing wrong output file names

@@ -1,48 +1,0 @@
-#!/usr/bin/python2.7
-# Read a matrix in coordinate format and create another matrix file with sparse struct headers
-
-import sys
-
-matname = sys.argv[1]
-outmatname = sys.argv[2]
-
-
-print matname
-
-f = open(matname, "r")
-nz = 0;
-nrows = 0;
-ncols = 0;
-for line in f:
-  line = line.split()
-  nz = nz + 1
-  line[0] = int(line[0])
-  line[1] = int(line[1])
-  if line[0] + 1 > nrows:
-    nrows = line[0] + 1
-  if line[1] + 1 > ncols:
-    ncols = line[1] + 1
-  if nz % 100000 == 0:
-    print nz
-f.close()
-
-f = open(matname, "r")
-fo = open(outmatname, "w")
-fo.write('%d %d\n' % (2, nz))
-fo.write('%d %d\n' % (nrows, ncols))
-nz = 0;
-for line in f:
-  line = line.split()
-  line[0] = int(line[0]) + 1
-  line[1] = int(line[1]) + 1
-  if len(line) == 2:
-    val = 1.0
-  else:
-    val = float(line[2]);
-  nz = nz + 1
-  if nz % 100000 == 0:
-    print nz
-  fo.write('%d %d %lf\n' % (line[0], line[1], val))
-
-fo.close()
-f.close()

### 2019-01-11 distntf/distntfio.hpp

@@ -374,10 +374,9 @@ class DistNTFIO {
         current_factor.save(sw.str(), arma::raw_ascii);
       }
       sw.clear();
+      sw.str("");
     }
-    sw.clear();
-    sw << output_file_name << "_lambda"
-       << "_" << MPI_SIZE;
+    sw << output_file_name << "_lambda" << "_" << MPI_SIZE;
     if (ISROOT) {
       ntfsolver->lambda().save(sw.str(), arma::raw_ascii);
     }

### 2019-01-11 common/ncpfactors.hpp
was not working with lowrank 1 and 2. set_labmmbda was iterating over modes instead of k. split criteria for dim tree fails if left is first mode and the rest of right

@@ -115,9 +115,7 @@ ncp_factors[i].clear();
     this->ncp_factors[i_n] = i_factor;
   }
   /// sets the lambda vector
-  void set_lambda(VEC new_lambda) {
-    for (int i = 0; i < this->m_modes; i++) m_lambda(i) = new_lambda(i);
-  }
+  void set_lambda(const VEC &new_lambda) { m_lambda = new_lambda; }
   // compute gram of all local factors
   /**
    * Return the hadamard of the factor grams
@@ -295,7 +293,7 @@ current_nrows *= rightkrp.n_rows;
     out = rc;
   }
   /**
-   *  prints just the information about the factors. 
+   *  prints just the information about the factors.
    * it will NOT print the factor
    */
   void printinfo() {
@@ -322,8 +320,8 @@ current_nrows *= rightkrp.n_rows;
     std::cout << this->ncp_factors[i_n];
   }
   /**
-   * Transposes the entire factor matrix. 
-   * @param[out] factor_t that contains the transpose of every factor matrix. 
+   * Transposes the entire factor matrix.
+   * @param[out] factor_t that contains the transpose of every factor matrix.
    */
   void trans(NCPFactors &factor_t) {
     for (int i = 0; i < this->m_modes; i++) {
@@ -370,8 +368,8 @@ current_nrows *= rightkrp.n_rows;
   /**
    * initializes the local tensor with the given seed.
    * this is for reinitializing random numbers across different
-   * processors. 
-   * @param[in] i_seed 
+   * processors.
+   * @param[in] i_seed
    */
   void randu(const int i_seed) {
     arma::arma_rng::set_seed(i_seed);
@@ -382,8 +380,8 @@ current_nrows *= rightkrp.n_rows;
         ncp_factors[i].zeros();
       }
     }
-  }  
-  /// this is for reinitializing zeros across different processors.  
+  }
+  /// this is for reinitializing zeros across different processors.
   void zeros() {
     for (int i = 0; i < this->m_modes; i++) {
       ncp_factors[i].zeros();
@@ -393,8 +391,8 @@ current_nrows *= rightkrp.n_rows;
   // Distribution normalization of factor matrices
   // To be used for MPI code only
   /**
-   * Distributed column normalize of all the modes 
-   * across different processors. 
+   * Distributed column normalize of all the modes
+   * across different processors.
    */
   void distributed_normalize() {
     double local_colnorm;
@@ -413,7 +411,7 @@ current_nrows *= rightkrp.n_rows;
   }
[…]

### 2019-01-11 distntf/distauntf.hpp

@@ -138,8 +138,8 @@ class DistAUNTF {
   /**
    * This iterates over all grams leaving out current mode
    * and finds hadamard of the grams
-   * 
-   * @param[in] current_mode. 
+   *
+   * @param[in] current_mode.
    */
   void gram_hadamard(int current_mode) {
     global_gram.ones();
@@ -160,7 +160,7 @@ class DistAUNTF {
    * @param[in] current_mode
    */
 
-  // 
+  //
   void gather_ncp_factor(const int current_mode) {
     m_gathered_ncp_factors_t.factor(current_mode).zeros();
     // Had this comment for debugging memory corruption in all_gather
@@ -402,7 +402,7 @@ class DistAUNTF {
     for (int mode = 0; mode < this->m_modes; mode++) {
       if (mode != current_mode) this->m_stale_mttkrp[mode] = true;
     }
-  } 
+  }
 
   virtual void accelerate() {}
 
@@ -495,9 +495,10 @@ class DistAUNTF {
   VEC lambda() { return m_local_ncp_factors.lambda(); }
   /// Returns the current outer iteration of the computeNTF
   int current_it() const { return this->m_current_it; }
-  /// Returns the current error 
+  /// Returns the current error
   double current_error() const { return this->m_rel_error; }
-  /// MTTKRP can be computed with or without dimension trees. Dimtree is default.
+  /// MTTKRP can be computed with or without dimension trees. Dimtree is
+  /// default.
   void dim_tree(bool i_dim_tree) {
     this->m_enable_dim_tree = i_dim_tree;
     if (this->m_enable_dim_tree) {
@@ -545,9 +546,9 @@ class DistAUNTF {
   /**
    * Returns the factor matrix by collected it across all the processors.
    * Don't call this during the computeNTF loop. It is right now
-   * called after all the iterations to dump the factor matrix. 
+   * called after all the iterations to dump the factor matrix.
    * @param[in] mode of the factor matrix to be collected across processors
-   * @param[in] factor_matrix. To be allocated and freed by the caller. 
+   * @param[in] factor_matrix. To be allocated and freed by the caller.
    */
   void factor(int mode, double *factor_matrix) {
     gather_ncp_factor(mode);
@@ -588,18 +589,25 @@ class DistAUNTF {
       while (temp_cum_prod(split_mode) < split_criteria) {
         split_mode++;
       }
-      // check to see if split mode is left or right.
-      size_t current_left = temp_cum_prod(split_mode);
[…]

### 2019-01-14 latex/toms/algorithm.tex
Dim tree algorithm rough draft.
Initial BPP description commit

@@
 The computational cost of this operation is also typically lower order, but the running time in practice suffers from also being memory bandwidth bound.
+
+\begin{algorithm}
+\caption{Simple Dim Tree}
+\label{alg:simpleDT}
+\begin{algorithmic}
+\Require $\T{X}$ is an $N$-way tensor, $n \in [N]$, $s$ split dimension, $n$ mode to compute MTTKRP for, projection tensor $\hat{\T{X}}$, ktensor $\hat{\T{Y}}$ corresponding to $\hat{\T{X}}$.
+
+\If{n == 0 or n == N-1}
+	\If{n = 0}
+		\State $\M{K} \leftarrow KRP(s+1:N)$
+		\State $\hat{\T{X}} \leftarrow \T{X}(0:s,s+1:N) \cdot \M{K}$
+		\For{i=1:R}
+			\State $\M{M}(:,i) \leftarrow \hat{\T{X}}_i(0,1:s) \cdot KRP(1:s)$
+		\EndFor
+	\Else % N-1 = n
+		\State $\M{K} \leftarrow KRP(0:s)$
+		\State $\hat{\T{X}} \leftarrow \Tr{X}(0:s,s+1:N)^\Tra \cdot \M{K}$
+		\For{i=1:R}
+			\State $\M{M}(:,i) \leftarrow \hat{\T{X}}_i(s+1,s+2:N) \cdot KRP(s+2:N)$
+		\EndFor
+	\EndIf
+\
+\Else
+	\If{n $\leq$ s}
+		\State Y = s
+	\Else
+		\State Y = N
+	\EndIf
+	\For{i=1:R}
+		\State  $\M{K} \leftarrow KRP(n-1)$
+		\State $\hat{\T{X}}_i \leftarrow \M{K}_i \cdot \hat{\T{X}}_i(n-1,n:Y)$ % Update to X
+	\EndFor
+	\For{i=1:R}
+		\State  $\M{K} \leftarrow KRP(n+1:Y)$
+		\State $\M{M}(:,i) \leftarrow \hat{\T{X}}_i(n,n+1:Y) \cdot \M{K}_i$
+	\EndFor
+\EndIf
+
+\end{algorithmic}
+\end{algorithm}
 
@@
 
-In the case of updating the local factor matrix $\Mn{H}{n}_{\V{p}}$ we need to solve~\ref{eqn:nls} with $\M{X} = \Mn{H}{n}_{\V{p}}$, $\M{A} = \Mn{S}{n}$ and $\M{B} = \Mn{M}{n}_{\V{p}}$.
+In the case of updating the local factor matrix $\Mn{H}{n}_{\V{p}}$ we need to solve~\Cref{eqn:nls} with $\M{X} = \Mn{H}{n}_{\V{p}}$, $\M{A} = \Mn{S}{n}$ and $\M{B} = \Mn{M}{n}_{\V{p}}$.
 Our framework is capable of supporting any alternating-updating NLS solver~\cite{KBP16}.
@@
 We briefly describe the different solvers below.
-Note that the descriptions are for the general form of the NLS solver as shown in~\ref{eqn:nls}.
+Note that the descriptions are for the general form of the NLS solver as shown in~\Cref{eqn:nls}.
 
@@
 \end{align}
-While this rule does not solve~\ref{eqn:nls} to optimality it ensures a reduction in the objective value from the initial value of $\M{X}$.
+While this rule does not solve~\Cref{eqn:nls} to optimality it ensures a reduction in the objective value from the initial value of $\M{X}$.
 
@@
 \subsubsection{Block Principal Pivoting (BPP)}
+BPP is an active-set like method for solving NLS problems. The main subroutine of BPP is the single right hand side version of~\Cref{eqn:nls},
+\begin{align}
+\V{x} \leftarrow \underset{\V{x} \ge 0}{\argmin} \left\lVert \M{A}\V{x} -\V{b}\right\rVert^2_F
+\label{eqn:snls}
+\end{align}
+The Karash-Kuhn-Tucker (KKT) optimality conditions for~\Cref{eqn:snls} are shown below,
+\begin{subequations}
+\begin{align}
+\V{y} &= \M{A}^\Tra\M{A}\V{x} - \M{A}^\Tra\V{b} \\
+\V{y} &\geq \V{0} \\
+\V{x} &\geq \V{0} \\
+x_i y_i &= 0 \quad \forall i
+\end{align}
+\label{eqn:bppkkt}
+\end{subequations}
+The complementary slackness criteria from the KKT conditions forces the support sets, i.e.
+the non-zero elements, of $\V{x}$ and $\V{y}$ to be disjoint.
+This makes~\Cref{eqn:bppkkt} an instance of the \emph{Linear Complementarity Problem} which arises frequently in quadratic programming.
+When $k \ll \min(m.n)$, active-set methods are preferred since they deal with small matrices of size $m \times k$, $n \times k$ and $k \times k$.
+In the optimal solution, the active-set is the set of indices where $x_i = 0$ and the remaining indices are referred to as the passive set.
+Once the active-set is found, we can find the optimal solution to~\Cref{eqn:snls} by using solving an unconstrained least squares problem on the passive set of indices.
+The BPP algorithm attempts to find the active-set by greedily swapping indices between the intermediate active and passive sets until it finds a solution that satisfies the KKT conditions.
+Kim and Park discuss the method in greater detail in~\cite{KP2011}.
+The unconstrained least squares is solved using the Normal Equations method in the software though other methods can also be used.
+
 \subsubsection{Alternating Direction Method of Multipliers (ADMM)}
+
 \subsubsection{Nesterov-type algorithm}

### 2019-01-15 latex/toms/algorithm.tex
dim tree alg
Included descriptions of ADMM and Nesterov methods

@@
 \begin{align}
-\M{X} \leftarrow \underset{\M{X} \ge 0}{\argmin} \left\lVert \M{A}\M{X} -\M{B}\right\rVert^2_F
+\M{X} \leftarrow \underset{\M{X} \ge \M{0}}{\argmin} \left\lVert \M{A}\M{X} -\M{B}\right\rVert^2_F
 \label{eqn:nls}
@@
 \subsubsection{Alternating Direction Method of Multipliers (ADMM)}
+In the ADMM solver the optimization problem~\Cref{eqn:nls} is reformulated by introducing an auxiliary variable $\M{\hat{X}}$ of size $k \times n$,
+\begin{align}
+\begin{split}
+&\underset{\M{X},\M{\hat{X}}}{\min} \quad \frac 1 2 \left\lVert \M{A}\M{\hat{X}} - \M{B}\right\rVert_F^2 + r(\M{X}) \\
+&\textnormal{subject to } \M{X} = \M{\hat{X}}^\Tra 
+\end{split}
+\label{eqn:admmnls}
+\end{align}
+Here $r(\cdot)$ is the indicator function of $\mathbb{R}_+$. The updates for the ADMM algorithm becomes,
+\begin{align}
+\begin{split}
+\M{\hat{X}} &\leftarrow \left(\M{A}^\Tra\M{A} + \rho\M{I} \right)^{-1}\left(\M{A}^\Tra\M{B} + \rho\left(\M{X} + \M{U}\right)^\Tra\right) \\
+\M{X} &\leftarrow \underset{\M{X}}{\argmin}\ r(\M{X}) + \frac{\rho}{2} \left\lVert \M{X} - \M{\hat{X}}^\Tra + \M{U}\right\rVert_F^2 \\
+\M{U} &\leftarrow \M{U} + \M{X} - \M{\hat{X}}^\Tra
+\end{split}
+\end{align}
+where $\M{U}$ is the scaled version of the dual variables corresponding to the equality constraints $\M{X} = \M{\hat{X}}^\Tra$ and $\rho$ is a step size specified by the user.
+The advantage of using ADMM is the clever splitting of the non-negativity constraints into updates of two blocks of variables $\M{X}$ and $\M{\hat{X}}$.
+This allows for an unconstrained least squares solve for $\M{\hat{X}}$ and element-wise projections onto $\mathbb{R}_+$ for $\M{X}$.
+One important fact to notice is that the same matrix $\M{A}^\Tra\M{B}$ and matrix inverse $\Br{\M{A}^\Tra\M{A} + \rho \M{I}}^{-1}$ is used for all the updates.
+We can therefore cache $\M{A}^\Tra\M{B}$ and the Cholesky decomposition of  $\Br{\M{A}^\Tra\M{A} + \rho \M{I}}$ to save some computations.
+A comprehensive guide to the ADMM method, convergence properties and selection of optimal $\rho$ can be found in~\cite{boyd2011distributed}.
 
 \subsubsection{Nesterov-type algorithm}
+The Nesterov-type algorithm in this package was introduced by Liavas et al~\cite{LKLHS2017}.
+Their method solves a modified version of NLS problem~\Cref{eqn:nls} with the introduction of a proximal term.
+The proximal term is useful to handle ill-conditioned instances and guarantee strong convexity.
+The objective function tackled is,
+\begin{align}
+\underset{\M{X} \ge \M{0}}{\min} f_p(\M{X}) := \frac 1 2 \left\lVert \M{B}^\Tra - \M{X}^\Tra\M{A}^\Tra \right\rVert_F^2 + \frac{\lambda}{2}\left\lVert\M{X} -\M{X_*}\right\rVert_F^2
+\label{eqn:nesnls}
+\end{align}
+The gradient of $f_p$ is given by the expression,
+\begin{align}
+\nabla f_p(\M{X}) = -(\M{B}^\Tra - \M{X}^\Tra\M{A}^\Tra)\M{A} + \lambda(\M{X}-\M{X_*})
+\end{align}
+The first order updates to $\M{X}$ are iterative and sketched below. They are repeated until a termination criteria is triggered. Different criteria are discussed in~\cite{LKLHS2017},
+\begin{align}
+\begin{split}
+\M{Y}_{k+1} &\leftarrow \left[ \M{X}_k - \alpha \nabla f_p(\M{X}_k)\right]_+ \\
+\M{X}_{k+1} &\leftarrow \M{Y}_{k+1} + \beta \Br{\M{Y}_{k+1} - \M{Y}_k}
+\end{split}
+\label{eqn:nesupd}
+\end{align}
+Here $[\cdot]_+$ is the projection operator in to $\mathbb{R}_+$.
+The selection of hyperparameters $\lambda, \alpha$ and $\beta$ depends on the spectrum of $\M{A}\M{A}^\Tra$ and is necessary for developing an Nesterov-like method for solving~\Cref{eqn:nesnls}.
+Details of the selection procedure and different cases can be found in the original paper~\cite{LKLHS2017}.
+In addition to the acceleration performed during each NLS solve~\Cref{eqn:nesupd} we can also perform an acceleration step for every outer iteration in~\Cref{alg:Par-NNCP-short}~\Cref{line:while}.
+In this step all factor matrices are updated using the previous iterate values till the objective stops decreasing.

### 2019-01-15 latex/toms/algorithm.tex
Dimension Tree section edits

--- 2019-01-14 latex/toms/algorithm.tex
@@
 
+\Cref{alg:simpleDT} gives pseudo code for the dimension tree implemented in PLANC.
+This code follows a dimension tree format which computes MTTKRPS in ascending order, that is 0 to $N$.
+There is a need for extra storage space to save the results of partial MTTKRPs and multi-TTVs.
+This is denoted as a tensor $\hat{\T{X}}$.
+Note that this tensor is always $R \ \times$ a subset of contiguous tensor dimensions.
+This means that $\hat{\T{X}}$ can always be thought of as a column major matrix with $R$ columns.
+The multi-TTV operations require that the columns of $\hat{\T{X}}$ be logically reshaped into smaller matrices to perform $R$ matrix vector products.
+$\hat{\T{X}}^i$ denotes accessing the data which makes up the ith column of $\hat{\T{X}}$.
+The amount of additional memory needed for $\hat{\T{X}}$  is the size of the larger partial MTTKRP and is $O(I)$ if $R$ is less than the smallest tensor dimension.
+
 \begin{algorithm}
@@
 \begin{algorithmic}
-\Require $\T{X}$ is an $N$-way tensor, $n \in [N]$, $s$ split dimension, $n$ mode to compute MTTKRP for, projection tensor $\hat{\T{X}}$, ktensor $\hat{\T{Y}}$ corresponding to $\hat{\T{X}}$.
-
+\Require $\T{X}$ is an $N$-way tensor, $n \in [N]$, $s$ split dimension, $n$ loops over $0:N-1$, and projection tensor $\hat{\T{X}}$ for storing intermediate tensor products.
 \If{n == 0 or n == N-1}
 	\If{n = 0}
-		\State $\M{K} \leftarrow KRP(s+1:N)$
+		\State $\M{K} \leftarrow (\M{H}^{(s+1)} \bigodot ... \bigodot \M{H}^{(N)}) $
 		\State $\hat{\T{X}} \leftarrow \T{X}(0:s,s+1:N) \cdot \M{K}$
 		\For{i=1:R}
-			\State $\M{M}(:,i) \leftarrow \hat{\T{X}}_i(0,1:s) \cdot KRP(1:s)$
+			\State $\M{M}(:,i) \leftarrow \hat{\T{X}}^i(0,1:s) \cdot (\M{H}^{(1)} \bigodot ... \bigodot \M{H}^{(s)})$
 		\EndFor
 	\Else % N-1 = n
-		\State $\M{K} \leftarrow KRP(0:s)$
+		\State $\M{K} \leftarrow (\M{H}^{(0)} \bigodot ... \bigodot \M{H}^{(s)})$
 		\State $\hat{\T{X}} \leftarrow \Tr{X}(0:s,s+1:N)^\Tra \cdot \M{K}$
 		\For{i=1:R}
-			\State $\M{M}(:,i) \leftarrow \hat{\T{X}}_i(s+1,s+2:N) \cdot KRP(s+2:N)$
+			\State $\M{M}(:,i) \leftarrow \hat{\T{X}}^i(s+1,s+2:N) \cdot (\M{H}^{(s+2)} \bigodot ... \bigodot \M{H}^{(N)})$
 		\EndFor
@@
 	\For{i=1:R}
-		\State  $\M{K} \leftarrow KRP(n-1)$
-		\State $\hat{\T{X}}_i \leftarrow \M{K}_i \cdot \hat{\T{X}}_i(n-1,n:Y)$ % Update to X
+		\State  $\M{K} \leftarrow \M{H}^{(n-1)}$
+		\State $\hat{\T{X}}^i \leftarrow \M{K}_i \cdot \hat{\T{X}}^i(n-1,n:Y)$ % Update to X
 	\EndFor
 	\For{i=1:R}
-		\State  $\M{K} \leftarrow KRP(n+1:Y)$
-		\State $\M{M}(:,i) \leftarrow \hat{\T{X}}_i(n,n+1:Y) \cdot \M{K}_i$
+		\State  $\M{K} \leftarrow (\M{H}^{(n+1)} \bigodot ... \bigodot \M{H}^{(Y)})$
+		\State $\M{M}(:,i) \leftarrow \hat{\T{X}}^i(n,n+1:Y) \cdot \M{K}_i$
 	\EndFor
 \EndIf
-
 \end{algorithmic}

### 2019-01-16 distntf/distauntf.hpp
accouting time for nesterov and admm, global size mismatched for io tensor

@@ -558,6 +558,10 @@ class DistAUNTF {
     std::vector<int> recvcnts(fiber_size, 0);
     std::vector<int> displs(fiber_size, 0);
 
+    DISTPRINTINFO("Collecting mode::" << mode << "::sendcnt::" << sendcnt
+                                      << "::fiber_size::" << fiber_size
+                                      << "::global_size::" << global_size);
+
     // int dimsize = m_factor_local_dims[current_mode];
     for (int i = 0; i < fiber_size; i++) {
       recvcnts[i] = itersplit(global_size, fiber_size, i) * m_low_rank_k;
@@ -589,10 +593,10 @@ class DistAUNTF {
       while (temp_cum_prod(split_mode) < split_criteria) {
         split_mode++;
       }
-      PRINTROOT("KDT Split Mode::"
-                << split_mode << "::split criteria::" << split_criteria
-                << "::cum prod::" << std::endl
-                << temp_cum_prod << std::endl);
+      PRINTROOT("KDT Split Mode::" << split_mode
+                                   << "::split criteria::" << split_criteria
+                                   << "::cum prod::" << std::endl
+                                   << temp_cum_prod << std::endl);
 
       // check to see if split mode is left or right.
       if (split_mode > 0) {

### 2019-01-16 distntf/distntf.cpp

@@ -78,17 +78,8 @@ class DistNTF {
       // }
       // return;
     }
-    // planc::Tensor A(dio.A()->dimensions(), dio.A()->m_data);
     planc::Tensor A;
     A = dio.A();
-    if (m_Afile_name.compare(0, rand_prefix.size(), rand_prefix) != 0) {
-      UVEC local_dims = A.dimensions();
-      /*MPI_Allreduce(&localm, &(this->m_globalm), 1, MPI_INT,
-       *            MPI_SUM, mpicomm.commSubs()[0]);
-       * MPI_Allreduce(&localn, &(this->m_globaln), 1, MPI_INT,
-       *            MPI_SUM, mpicomm.commSubs()[1]);*/
-      this->m_global_dims = local_dims % this->m_proc_grids;
-    }
     INFO << mpicomm.rank()
          << "::Completed generating tensor A=" << A.dimensions()
          << "::start indices::" << A.global_idx()

### 2019-01-16 distntf/distntfaoadmm.hpp

@@ -18,12 +18,17 @@ class DistNTFAOADMM : public DistAUNTF {
   MAT tempgram;
   int admm_iter;
   double tolerance;
+  double chol_time;
+  double stop_iter_time;
+  double proj_time;
+  double solve_time;
+  double norm_time;
 
  protected:
-   /**
+  /**
    * This is ADMM based update function.
-   * Given the MTTKRP and the hadamard of all the grams, we 
-   * determine the factor matrix to be updated. 
+   * Given the MTTKRP and the hadamard of all the grams, we
+   * determine the factor matrix to be updated.
    * @param[in] Mode of the factor to be updated
    * @returns The new updated factor
    */
@@ -35,6 +40,7 @@ class DistNTFAOADMM : public DistAUNTF {
     // Set up ADMM iteration
     double alpha = 0.0;
 
+    MPITIC;
     if (m_nls_sizes[mode] > 0) {
       alpha = arma::trace(this->global_gram) / this->m_local_ncp_factors.rank();
       alpha = (alpha > 0) ? alpha : 0.01;
@@ -43,6 +49,7 @@ class DistNTFAOADMM : public DistAUNTF {
       L = arma::chol(tempgram, "lower");
       Lt = L.t();
     }
+    chol_time += MPITOC;
     bool stop_iter = false;
 
     // Start ADMM loop from here
@@ -50,7 +57,7 @@ class DistNTFAOADMM : public DistAUNTF {
       if (m_nls_sizes[mode] > 0) {
         prev_fac = updated_fac;
         m_local_ncp_aux_t.set(mode, m_local_ncp_aux.factor(mode).t());
-
+        MPITIC;
         m_temp_local_ncp_aux_t.set(
             mode, arma::solve(arma::trimatl(L),
                               this->ncp_local_mttkrp_t[mode] +
@@ -59,14 +66,15 @@ class DistNTFAOADMM : public DistAUNTF {
         m_local_ncp_aux_t.set(mode,
                               arma::solve(arma::trimatu(Lt),
                                           m_temp_local_ncp_aux_t.factor(mode)));
-
+        solve_time += MPITOC;
         // Update factor matrix
         updated_fac = m_local_ncp_aux_t.factor(mode).t();
+        MPITIC;
         fixNumericalError<MAT>(&(updated_fac), EPSILON_1EMINUS16);
         updated_fac = updated_fac - m_local_ncp_aux.factor(mode);
         updated_fac.for_each(
[…]

### 2019-01-16 distntf/distntfnes.hpp

@@ -19,11 +19,15 @@ class DistNTFNES : public DistAUNTF {
   double delta1;  // Termination value for absmax
   double delta2;  // Termination value for min
   // Acceleration Variables
-  int acc_start;   // Starting iteration for acceleration
-  int acc_exp;     // Step size exponent
-  int acc_fails;   // Acceleration Fails
-  int fail_limit;  // Acceleration Fail limit
+  int acc_start;                   // Starting iteration for acceleration
+  int acc_exp;                     // Step size exponent
+  int acc_fails;                   // Acceleration Fails
+  int fail_limit;                  // Acceleration Fail limit
   const int NUM_INNER_ITERS = 20;  // Capping inner iter
+  double eig_time;
+  double stop_iter_time;
+  double proj_time;
+  double norm_time;
 
  protected:
   inline double get_lambda(double L, double mu) {
@@ -54,6 +58,7 @@ class DistNTFNES : public DistAUNTF {
   }
 
   bool stop_iter(const int mode) {
+    MPITIC;
     bool stop = false;
     double local_absmax, local_min, global_absmax, global_min;
     local_absmax = 0.0;
@@ -72,7 +77,7 @@ class DistNTFNES : public DistAUNTF {
                   MPI_COMM_WORLD);
 
     if (global_absmax <= delta1 && global_min >= -delta2) stop = true;
-
+    stop_iter_time += MPITOC;
     return stop;
   }
 
@@ -144,8 +149,9 @@ class DistNTFNES : public DistAUNTF {
     MAT Htprev = Ht;
     m_acc_t.set(mode, Ht);
     modified_gram = this->global_gram;
-
+    MPITIC;
     VEC eigval = arma::eig_sym(modified_gram);
+    eig_time += MPITOC;
     L = eigval.max();
     mu = eigval.min();
     lambda = get_lambda(L, mu);
@@ -168,8 +174,10 @@ class DistNTFNES : public DistAUNTF {
       iter++;
       Htprev = Ht;
       Ht = m_acc_t.factor(mode) - ((1 / (L + lambda)) * m_grad_t.factor(mode));
+      MPITIC;
       fixNumericalError<MAT>(&Ht, EPSILON_1EMINUS16);
       Ht.for_each([](MAT::elem_type &val) { val = val > 0.0 ? val : 0.0; });
+      proj_time += MPITOC;
       alpha_prev = alpha;
       alpha = get_alpha(alpha_prev, q);
       beta =
[…]

### 2019-01-16 latex/slides/SIAM-ALA18/fig/modeled.tex

%!TEX root = ../talk.tex

\begin{tikzpicture}[scale=.8]
\renewcommand{\datafile}{data/modeled.dat}
\begin{axis}[
	title=Modeled Strong-Scaling Comparison,
	xlabel=Processors,
	ylabel=Words Communicated,
	width=\columnwidth,
	xmode=log,
	log basis x={2},
	xmin=1,xmax=2^31,
	xtick={1,32,1024,32768,1048576,33554432,1073741824},
	ymode=log,
	legend pos=south west
	]
	\addplot [black,thick,dashed,mark options={solid},mark=x] table [x={P}, y={MM}] {\datafile};
	\addlegendentry{Matrix Multiplication}
	\addplot [red,thick,dashed,mark options={solid},mark=asterisk] table [x={P}, y={Alg3}] {\datafile};
	\addlegendentry{Stationary Tensor}
	\addplot [blue,thick,dashed,mark options={solid},mark=o] table [x={P}, y={Alg4}] {\datafile};
	\addlegendentry{General Tensor}
\end{axis}
\end{tikzpicture}

### 2019-01-16 latex/slides/SIAM-ALA18/talk.tex

\documentclass[xcolor=dvipsnames]{beamer}

\usepackage{setspace}
\usepackage{beamerthemesplit}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage{times}
\usepackage[T1]{fontenc}
\usepackage{color}
\usepackage{colortbl}
\usepackage[noend]{algorithm,algpseudocode}
\usepackage{multimedia,xmpmulti}
\usepackage{multirow,multicol}
\usepackage[absolute,overlay]{textpos}
\usepackage{multimedia,xmpmulti}
\usepackage{tabularx,multicol}
\usepackage{url}
\usepackage{ulem}

\input{tensor_header}

%% Tikz - For making pretty pictures
\usepackage{tikz}
\usetikzlibrary{3d}
\usetikzlibrary{patterns}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}

\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{etoolbox}
%\pgfplotsset{compat=1.12}
\usetikzlibrary{decorations.pathreplacing}
\makeatletter
\newcommand{\gettikzxy}[3]{%
  \tikz@scan@one@point\pgfutil@firstofone#1\relax
  \edef#2{\the\pgf@x}%
  \edef#3{\the\pgf@y}%
}
\makeatother
\newcommand{\datafile}{}

\algrenewcommand{\algorithmiccomment}[1]{\textit{$\%$ #1}}
\newcommand{\lt}{\left}
\newcommand{\rt}{\right}
\newcommand{\ra}{\rightarrow}
\newcommand{\dis}{\displaystyle}
\newcommand{\bc}[3]{\left\langle#1, #2, #3\right\rangle}
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\ignore}[1]{}
\newtheorem{thm}{Theorem}
\newtheorem{lem}[thm]{Lemma}
\definecolor{bgblue}{rgb}{0.8,0.85,1}
\definecolor{bggreen}{rgb}{0.8,1,0.8}
\definecolor{bgred}{rgb}{1,0.85,0.8}
\definecolor{Blue}{rgb}{.08,0,.7}
\setbeamercolor{greentextbox}{fg=black,bg=bggreen}

\definecolor{wfugold}{rgb}{0.6196078,0.494117647,0.21960784}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\multiplycolor}{red}
\newcommand{\zero}{}
\newcommand{\cred}[1]{\textcolor{red}{#1}}
\newcommand{\cblue}[1]{\textcolor{blue}{#1}}
\newcommand{\cgold}[1]{\textcolor{wfugold}{#1}}

\newcommand{\UVW}{\llbracket \M U, \M V, \M W \rrbracket}
\newcommand{\plusequals}{\mathrel{+}=}
\algdef{SE}[FORDOTS]{ForDots}{EndForDots}{$\ddots$}{}

\graphicspath{{fig/}}

\mode<presentation>

\usetheme{Warsaw}
\usecolortheme[named=wfugold]{structure}
%Madrid
%\usecolortheme{dolphin}
%\useinnertheme{rounded}
%\usefonttheme{serif}
\setbeamertemplate{navigation symbols}{} % gets rid of navigation bars
\setbeamertemplate{footline}
{
  \hbox{
  \begin{beamercolorbox}[wd=.33\paperwidth,ht=2.25ex,dp=1ex,left]{author in head/foot}%
    \usebeamerfont{author in head/foot}
    Ballard
  \end{beamercolorbox}%
  \begin{beamercolorbox}[wd=.34\paperwidth,ht=2.25ex,dp=1ex,center]{title in head/foot}%
    \usebeamerfont{title in head/foot}

  \end{beamercolorbox}%
  \begin{beamercolorbox}[wd=.33\paperwidth,ht=2.25ex,dp=1ex,right]{date in head/foot}%
    \usebeamerfont{date in head/foot}
    \insertframenumber{} \hspace*{2ex} 
  \end{beamercolorbox}}%
}

\begin{document}

\title{Communication-Optimal Algorithms for \\ CP Decompositions of Dense Tensors}
\author[Ballard]{\textbf{Grey Ballard}, Koby Hayashi, Ramakrishnan Kannan, Nicholas Knight, Kathryn Rouse}
\institute{\includegraphics[scale=.15]{wfulogo}}
\date{May 8, 2018}
\titlegraphic{SIAM Conference on Applied Linear Algebra \\ \small MS04: Constrained Low-Rank Matrix
and Tensor Approximations}
\begin{frame}[label=title,plain]
\maketitle
\end{frame}
\addtocounter{framenumber}{-1}

\begin{frame}[label=summary]
\frametitle{Summary}

\begin{itemize}
	\item We establish communication lower bounds for matricized-tensor times Khatri-Rao product (MTTKRP)
	\begin{itemize}
		\item key kernel for computing CP decomposition
	\end{itemize}
	\vfill
	\item We present optimal parallel dense MTTKRP algorithm
	\begin{itemize}
		\item attains the lower bound to within constant factors
	\end{itemize}
	\vfill
	\item We implement and benchmark optimal CP-ALS algorithm
	\begin{itemize}
		\item remains computation bound and scales well
		\item dimension tree optimization avoids redundant computation
	\end{itemize}
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{CP Decomposition: sum of outer products}

Matrix: \quad $\displaystyle \M{M} \approx \sum_{r=1}^R \MC{U}{r} (\sigma_r \MC{V}{r}^T)$
\begin{center}
\scalebox{.14}{\includegraphics{matrix-LR-SOP}}
\end{center}
\vfill
Tensor: \quad $\displaystyle \T{X}\approx \sum_{r=1}^R \MC{U}{r}\circ\MC{V}{r}\circ\MC{W}{r}$
\begin{center}
\scalebox{.27}{\includegraphics{tensor-LR-SOP}}
\end{center}
\vfill
\begin{center}
This is known as the CANDECOMP or PARAFAC or \\ canonical polyadic or \textcolor{wfugold}{CP decomposition}
\end{center}

\end{frame}

\begin{frame}
\frametitle{CP Optimization Problem}

\begin{center}
\scalebox{.15}{\includegraphics{tensor-LR-SOP}}
\end{center}
\vfill
\begin{flushleft} 
For fixed rank $R$, we want to solve
\end{flushleft}
%\vspace{-1.5cm}
$$\min_{\M{U},\M{V},\M{W}} \left\|\T{X}-\sum_{r=1}^R \MC{U}{r}\circ\MC{V}{r}\circ\MC{W}{r}\right\| $$
\vfill
\begin{flushright}
which is a nonlinear, nonconvex optimization problem
\end{flushright}
\vfill
{\small 
\begin{itemize}
	\item in the matrix case, the SVD gives us the optimal solution
	\item in the tensor case, need iterative optimization scheme
\end{itemize}
}

\end{frame}

\begin{frame}
\frametitle{Alternating Least Squares (ALS)}

Fixing all but one factor matrix, we have a linear LS problem:
$$\min_{\M{V}} \left\|\T{X}-\sum_{r=1}^R \MC[\hat]{U}{r}\circ\MC{V}{r}\circ\MC[\hat]{W}{r}\right\| $$
or equivalently
$$\min_{\M{V}} \left\|\Mz{X}{2}-\M{V}(\M[\hat]{W} \Khat \M[\hat]{U})^\Tra\right\|_F $$

\begin{center}

\vfill

$\Khat$ is the \emph{Khatri-Rao} product, a column-wise Kronecker product or row-wise Hadamard (element-wise) product

\vfill

ALS works by alternating over factor matrices, updating one at a time by solving the corresponding linear LS problem

\end{center}

\end{frame}

\begin{frame}
\frametitle{CP-ALS}

\textbf{Repeat}
\begin{enumerate}
	\item Solve $\M{U}(\M{V}^\Tra\M{V} \Hada \M{W}^\Tra\M{W}) = \textcolor{wfugold}{\Mz{X}{1}(\M{W} \Khat \M{V})}$ for $\M{U}$ 
	%\item Normalize columns of $\M{U}$
	\item Solve $\M{V}(\M{U}^\Tra\M{U} \Hada \M{W}^\Tra\M{W}) = \textcolor{wfugold}{\Mz{X}{2}(\M{W} \Khat \M{U})}$ for $\M{V}$ 
	%\item Normalize columns of $\M{V}$
	\item Solve $\M{W}(\M{U}^\Tra\M{U} \Hada \M{V}^\Tra\M{V}) = \textcolor{wfugold}{\Mz{X}{3}(\M{V} \Khat \M{U})}$ for $\M{W}$ 
	%\item Normalize columns of $\M{W}$ and store norms in $\V{\lambda}$
\end{enumerate}

\vfill

\begin{center}
{\small
Linear least squares problems solved via normal equations \\ using identity $(\M{A}\Khat\M{B})^\Tra(\M{A}\Khat\M{B}) = \M{A}^\Tra\M{A} \Hada \M{B}^\Tra\M{B}$, \\ where $\Hada$ is Hadamard product
\vfill
All optimization schemes that compute the gradient \\ must also compute \textcolor{wfugold}{MTTKRP} in all modes: e.g.,
$$\frac{\partial f}{\partial \M{V}} = \M{V}(\M{U}^\Tra\M{U} \Hada \M{W}^\Tra\M{W}) - \textcolor{wfugold}{\Mz{X}{2}(\M{W} \Khat \M{U})}$$
}
\end{center}

\end{frame}

\begin{frame}
\frametitle{MTTKRP bottleneck}

\begin{itemize}
	\item How do we compute MTTKRP efficiently?
	\vfill
	\item How do we parallelize MTTKRP efficiently?
	\begin{itemize}
		\item how do we load balance computation?
		\item how do we minimize communication?
	\end{itemize}
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{MTTKRP via Matrix Multiplication}

$$\text{MTTKRP:  } \quad \textcolor{wfugold}{\M{M} = \Mz{X}{2}(\M{W} \Khat \M{U})}$$

\vfill

Standard approach to MTTKRP for dense tensors
\begin{enumerate}
	\item ``form'' matricized tensor (a matrix)
	\item compute Khatri-Rao product (a matrix)
	\item call matrix-matrix multiplication subroutine
\end{enumerate}

\vfill

\begin{center}
Can we communicate less by exploiting tensor structure? \\ (avoiding forming explicit Khatri-Rao product)
\end{center}

\end{frame}

\begin{frame}
\frametitle{MTTKRP for \only<1>{3}\only<2>{$N$}-way Tensors}

\textbf{Matrix equation}:  
\only<1>{$$\M{M} = \Mz{X}{2}(\M{W} \Khat \M{U})$$}
\only<2>{$$\Mn{M}{n} = \Mz{X}{n}(\Mn{U}{N} \Khat \cdots \Khat \Mn{U}{n+1} \Khat \Mn{U}{n-1} \Khat \cdots \Khat \Mn{U}{1})$$}

\vfill

\textbf{Element equation}:  
\only<1>{$$\ME{M}{jr} = \sum_{i=1}^I \sum_{k=1}^K \TE{X}{ijk} \ME{U}{ir} \ME{W}{kr}$$}
\only<2>{$$\MnE{M}{n}{i_nr} = \sum \TE{X}{i_1\dots i_N} \prod_{m\neq n} \MnE{U}{m}{i_mr} $$}

\vfill

\textbf{Example pseudocode}:

\begin{center}
\begin{algorithmic}
\only<1>{
\For{$i=1$ to $I$}
	\For{$j=1$ to $J$}
		\For{$k=1$ to $K$}
			\For{$r=1$ to $R$}
				\State $\M M(j,r) \plusequals \T{X}(i,j,k) \cdot \M{U}(i,r) \cdot \M{W}(k,r)$
			\EndFor
		\EndFor
	\EndFor
\EndFor
}\only<2>{
\For{$i_1=1$ to $I_1$}
	\ForDots
		\For{$i_N=1$ to $I_N$}
			\For{$r=1$ to $R$}
				\State \footnotesize $\Mn{M}{n}(i_n,r) \plusequals \T{X}(i_1,\dots,i_N) \cdot \Mn{U}{1}(i_1,r)  \cdots  \Mn{U}{N}(i_N,r)$
			\EndFor
		\EndFor
	\EndForDots
\EndFor
}
\end{algorithmic}
\end{center}

\end{frame}

\begin{frame}
\frametitle{Communication Lower Bounds for MTTKRP}

MTTKRP is a set of nested loops that accesses arrays
\begin{itemize}
	\small
	\item Nick's PhD thesis was ``Communication-Optimal Loop Nests"
	\item References: thesis \cite{Knight15} and paper \cite{CDKSY13}
\end{itemize}

\vfill

From Nick's thesis...
\begin{itemize}
	\item tabulate how the arrays are accessed
	\item use H\"{o}lder-Brascamp-Lieb-type inequality in LB proof
	\item solve linear program to get tightest lower bound
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{MTTKRP Lower Bound Proofs}

\begin{itemize}
	\item Lower bound argument follows \cite{CDKSY13} almost directly
	%\begin{itemize}
	%	\item solve linear program involving $\Delta$ for tightest bound
	%\end{itemize}
	\vfill
	\pause
	\item Gotcha: the number of nested loops is not constant
	\begin{itemize}
		%\item LB becomes too low by a factor of $O(N)$ (number of loops)
		\item Fixed using a technique similar to one used for tightening the constant in matrix multiplication lower bound \cite{SvdG17}
	\end{itemize}
	\vfill
	\pause
	\item Gotcha: memory-independent bounds most relevant
	\begin{itemize}
		\item inspiration from matrix multiplication \cite{BDHLS12-SS,DE+13}
	\end{itemize}
	\vfill
	\pause
	\item Key assumption: algorithm is not allowed to pre-compute and re-use temporary values 
	\begin{itemize}
		\item e.g., forming explicit Khatri-Rao product
		\item e.g., computing and re-using ``partial'' MTTKRP
	\end{itemize}
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Parallel Communication Lower Bound}

\begin{theorem}
Any parallel MTTKRP algorithm involving a tensor with $I_k=I^{1/N}$ for all $k$ and that evenly distributes one copy of the input and output performs at least
\begin{equation*}
\Omega\left( \left( \frac{NIR}{P} \right)^{\frac{N}{2N-1}} + NR\left(\frac{I}{P}\right)^{1/N} \right)
\end{equation*}
sends and receives.  (Either term can dominate.)
\end{theorem}

\vfill

\small
\begin{itemize}
	\footnotesize
	\item $N$ is the number of modes
	\item $I$ is the number of tensor entries
	\item $I_k$ is the dimension of the $k$th mode
	\item $R$ is the rank of the CP model
	\item $P$ is the number of processors
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Communication-Optimal Parallel Algorithm (3D)}

\begin{columns}

\begin{column}{.6\textwidth}

\newcommand{\procdim}{3}
\newcommand{\proc}{\draw[black,shift={(-.5,-.5)}] (0,0) grid (\procdim,\procdim);}
\newcommand{\highlight}{gray!75}
\newcommand{\commhighlight}{gray!25}

\begin{center}
\begin{tikzpicture}[x={(-0.5cm,-0.4cm)}, y={(1cm,0cm)}, z={(0cm,1cm)},every node/.append style={transform shape}]

% highlight all-gather/reduce-scatter patterns
% 1st slide is proc data (leave highlighting for all slides)
% 2nd slide is all-gather in 1st mode
\only<2>{
% front face
\begin{scope}[canvas is yz plane at x=.5]
	% highlight front face of proc comm
	\draw[fill=\commhighlight,shift={(-.5,-.5)}] (0,0) rectangle (3,1);
	% highlight block of 1st factor matrix
	\draw[fill=\commhighlight,shift={(-1,.5)},xscale=.5] (0,0) rectangle (-1,-1);
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=-90]
	% highlight right face of proc comm
	\draw[fill=\commhighlight,shift={(-.5,-.5)}] (0,0) rectangle (3,1);
\end{scope}
% top face
\begin{scope}[canvas is yx plane at z=.5,yscale=-1,rotate=0]
	% highlight top face of proc comm
	\draw[fill=\commhighlight,shift={(-.5,-.5)}] (0,0) rectangle (3,3);
\end{scope}
}
% 3rd slide is all-gather in 3rd mode
\only<3>{
% front face
\begin{scope}[canvas is yz plane at x=.5]
	% highlight front face of proc comm
	\draw[fill=\commhighlight,shift={(-.5,.5)}] (0,0) rectangle (3,-3);
	% highlight block of 1st factor matrix now owned by processor
	\draw[fill=\highlight,shift={(-1,.5)},xscale=.5] (0,0) rectangle (-1,-1);
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=-90]
	% highlight right face of proc comm
	\draw[fill=\commhighlight,shift={(-.5,.5)}] (0,0) rectangle (1,-3);
	% highlight block of 3rd factor matrix
	\draw[fill=\commhighlight,yscale=.5,shift={(-.5,-6)}] (0,0) rectangle (1,-1);
\end{scope}
% top face
\begin{scope}[canvas is yx plane at z=.5,yscale=-1,rotate=0]
	% highlight top face of proc comm
	\draw[fill=\commhighlight,shift={(-.5,-.5)}] (0,0) rectangle (3,1);
\end{scope}
}
% 4th and 5th slides are local computation
\only<4-5>{
% front face
\begin{scope}[canvas is yz plane at x=.5]
	% highlight block of 1st factor matrix now owned by processor
	\draw[fill=\highlight,shift={(-1,.5)},xscale=.5] (0,0) rectangle (-1,-1);
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=-90]
	% highlight block of 3rd factor matrix now owned by processor
	\draw[fill=\highlight,yscale=.5,shift={(-.5,-6)}] (0,0) rectangle (1,-1);
\end{scope}
}
\only<5>{
% front face
\begin{scope}[canvas is yz plane at x=.5]
	% highlight block of 2nd factor matrix contribution computed by processor
	\draw[fill=\highlight,shift={(1.5,-3)},yscale=.5] (0,0) rectangle (1,-1);
\end{scope}
}
% 6th slide is reduce-scatter in 2nd mode
\only<6>{
% front face
\begin{scope}[canvas is yz plane at x=.5]
	% highlight front face of proc comm
	\draw[fill=\commhighlight,shift={(1.5,.5)}] (0,0) rectangle (1,-3);
	% highlight block of 2nd factor matrix involved in reduce scatter
	\draw[fill=\commhighlight,shift={(1.5,-3)},yscale=.5] (0,0) rectangle (1,-1);
	% highlight block of 2nd factor matrix of computed output
	\draw[fill=\highlight,shift={(1.5,-3)},yscale=.5] (0,0) rectangle (1/9,-1);
	% highlight block of 1st factor matrix now owned by processor
	\draw[fill=\highlight,shift={(-1,.5)},xscale=.5] (0,0) rectangle (-1,-1);
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=-90]
	% highlight right face of proc comm
	\draw[fill=\commhighlight,shift={(-.5,.5)}] (0,0) rectangle (3,-3);
	% highlight block of 3rd factor matrix now owned by processor
	\draw[fill=\highlight,yscale=.5,shift={(-.5,-6)}] (0,0) rectangle (1,-1);
\end{scope}
% top face
\begin{scope}[canvas is yx plane at z=.5,yscale=-1,rotate=0]
	% highlight top face of proc comm
	\draw[fill=\commhighlight,shift={(1.5,-.5)}] (0,0) rectangle (1,3);
\end{scope}
}

% highlight one processor's data
% front face
\begin{scope}[canvas is yz plane at x=.5,shift={(1.5,-1.5)}]
	% highlight front face of tensor block
	\draw[fill=\highlight,shift={(0,1)}] (0,0) rectangle (1,1);
	% highlight block of 1st factor matrix
	\draw[fill=\highlight,shift={(-2.5,2-1/3)},xscale=.5] (0,0) rectangle (-1,-1/9);
	% highlight block of 2nd factor matrix
	\draw[shift={(0,-1.5)},yscale=.5] (0,0) rectangle (1/9,-1);
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=-90,shift={(-.5,-3)}]
	% highlight front face of tensor block
	\draw[fill=\highlight,shift={(0,2.5)}] (0,0) rectangle (1,1);
	% highlight block of 3rd factor matrix
	\draw[fill=\highlight,yscale=.5] (0,0) rectangle (1/9,-1);
\end{scope}
% top face
\begin{scope}[canvas is yx plane at z=.5,yscale=-1,rotate=0]
	% highlight front face of tensor block
	\draw[fill=\highlight,shift={(1.5,-.5)}] (0,0) rectangle (1,1);
\end{scope}

% draw tensor
% front face
\begin{scope}[canvas is yz plane at x=.5,rotate=-90]
	\proc
\end{scope}
% top face
\begin{scope}[canvas is yx plane at z=.5,yscale=-1,rotate=0]
	\proc
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=180]
	\proc
\end{scope}

% draw factor matrices
% front face
\begin{scope}[canvas is yz plane at x=.5,shift={(1.5,-.5)}]
	% draw 1st factor matrix
	\draw[shift={(-2.5,0)},xscale=.5] (0,-2) grid (-1,1);
	\node[draw=none] at (-3.5,-.5) {$\Mn{U}{1}$};
	% draw 2nd factor matrix
	\draw[shift={(0,-2.5)},yscale=.5] (-2,0) grid (1,-1);
	\node[draw=none] at (-.5,-3.5) {$\Mn{M}{2}$};
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=-90,shift={(1.5,-.5)}]
	% draw 2nd factor matrix
	\draw[shift={(0,-2.5)},yscale=.5] (-2,0) grid (1,-1);
	\node[draw=none] at (-.5,-3.5) {$\Mn{U}{3}$};
\end{scope}

\end{tikzpicture}
\end{center}

\end{column}
\begin{column}{.45\textwidth}

\footnotesize
Each processor
\begin{enumerate}
	\item Starts with one subtensor and subset of rows of each input factor matrix
	\uncover<2->{\item All-Gathers all the rows needed from $\Mn{U}{1}$}
	\uncover<3->{\item All-Gathers all the rows needed from $\Mn{U}{3}$}
	\uncover<4->{\item Computes its contribution to rows of $\Mn{M}{2}$ (local MTTKRP)}
	\uncover<6->{\item Reduce-Scatters to compute and distribute $\Mn{M}{2}$ evenly}
\end{enumerate}

\end{column}

\end{columns}

\end{frame}

\begin{frame}
\frametitle{Theoretical Comparisons}

\begin{center}
%\renewcommand{\arraystretch}{2}
\begin{tabular}{|c|ccc|}
\hline
& \textbf{Lower Bound} & \textbf{New Algorithm} & \textbf{Standard (MM$^*$)}  \\
\hline
\textbf{Words} & \multirow{2}{*}{$\Omega\left( NR\left(\frac IP \right)^{1/N} \right)$} & \multirow{2}{*}{$O\left(NR\left(\frac IP \right)^{1/N} \right)$} & \multirow{2}{*}{$O\left( I^{1/N}R \right)$} \\
\small (``small'' $P$) & & & \\
\hline
\uncover<2>{\textbf{Words}} & \uncover<2>{\multirow{2}{*}{$\Omega\left(\left(\frac{NIR}{P}\right)^{\frac{N}{2N-1}} \right)$} & \multirow{2}{*}{$O\left( \left(\frac{NIR}{P}\right)^{\frac{N}{2N-1}}  \right)$} & \multirow{2}{*}{$O\left( \left(\frac{IR}{P}\right)^{2/3} \right)$}} \\
\uncover<2>{\small (``large'' $P$)} & & & \\
\hline
\end{tabular}
\end{center}

\begin{itemize}
	\small
	\item For relatively small $P$ (or small $R$) and even dimensions, parallel ``stationary'' algorithm attains lower bound
	\begin{itemize}
		\small
		\item same algorithm for sparse \cite{SK16} and dense 3D \cite{LK+17b} 
	\end{itemize}
	\pause
	\item For larger $P$ (or $R$), then we need more general algorithm to attain lower bound
	\begin{itemize}
	%	\item also parallelize over columns of output matrix
		\item involves communicating the tensor
	\end{itemize}
%	\item Simple version of this parallel algorithm in \cite{LK+17}
\end{itemize}

\vfill

\begin{center}
\small 
$^*$communication-optimal matrix multiplication from \cite{DE+13}
\end{center}

\end{frame}

\begin{frame}
\frametitle{Modeled Communication Costs}

\begin{center}
\input{fig/modeled}
\end{center}

\end{frame}

\begin{frame}
\frametitle{What about for a full CP-ALS iteration?}

A full iteration of CP-ALS includes computing all $N$ MTTKRPs

\vfill

\textbf{Lower Bound}\\
Lower bound for single MTTKRP applies to computing all $N$

\vfill

\textbf{Algorithm}\\
We can compute all $N$ with same communication as just 1
\begin{itemize}
	\small
	\item lots of data overlap across MTTKRPs
	\item more computation required, but not that much more
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Avoiding re-communication across MTTKRPs}

\begin{algorithmic}
\While{not converged}
	\For{$n=1$ to $N$}
		\State \Comment{Compute new factor matrix in $n$th mode}
		\State \textcolor{wfugold}{$\M{M} = \text{Local-MTTKRP}(\T{X}_{p_1\cdots p_N},\{\Mn{U}{i}_{p_i}\},n)$}
		\State \textcolor{wfugold}{$\Mn{M}{n}_{\V{p}} = \text{Reduce-Scatter}(\M{M},\textsc{Proc-Slice}(n,\VE{p}{n}))$} 
		\State \textcolor{gray}{$\Mn{S}{n} = \Mn{G}{1} \Hada \cdots \Hada \Mn{G}{n-1} \Hada \Mn{G}{n+1} \Hada \cdots \Hada \Mn{G}{N}$}
		\State \textcolor{gray}{$\Mn{U}{n}_{\V{p}} = \text{Local-Update}(\Mn{S}{n},\Mn{M}{n}_{\V{p}})$}
		\State \Comment{Organize data for later modes}
		\State \textcolor{gray}{$\M{H} = {\Mn{U}{n}_{\V{p}}}^\Tra\Mn{U}{n}_{\V{p}}$}
		\State \textcolor{gray}{$\Mn{G}{n} = \text{All-Reduce}(\M{H},\textsc{All-Procs})$}
		\State \textcolor{wfugold}{$\Mn{U}{n}_{p_n} = \text{All-Gather}(\Mn{U}{n}_{\V{p}},\textsc{Proc-Slice}(n,\VE{p}{n}))$}
	\EndFor 
\EndWhile
\end{algorithmic}

\vfill

\begin{center}
\footnotesize
Compute factor matrix, communicate it \textbf{once} for use in all other $N{-}1$ modes
\end{center}

\end{frame}

\begin{frame}
\frametitle{Avoiding recomputation across MTTKRPs}

\footnotesize
We re-use communication \emph{and} computation across MTTKRPs
\normalsize
$$\Mn{M}{1}=\underline{\Mz{X}{1}\left(\Mn{U}{3}\rt.}\Khat\lt.\Mn{U}{2}\rt) \hspace{.5cm} \text{  and  } \hspace{.5cm} \Mn{M}{2}=\underline{\Mz{X}{2}\lt(\Mn{U}{3}\rt.}\Khat\lt.\Mn{U}{1}\rt)$$

\vfill
\pause

\footnotesize
We organize intermediate values in ``dimension tree'' \cite{PTC13a,LCPSV17,KU18}
\normalsize

\only<2>{
\begin{center}
\begin{tikzpicture}

\node (123) at (4,2) {$\{1,2,3\}$};
\node (12) at (2,1) {$\{1,2\}$};
\node (3) at (6,1) {$\Mn{M}{3}$};
\node (1) at (1,0) {$\Mn{M}{1}$};
\node (2) at (3,0) {$\Mn{M}{2}$};

\scriptsize
\path[draw] (123) edge [left,align=right] node {PM} (12);
\path[draw] (123) edge [right,align=left] node {PM} (3);
\path[draw] (12) edge [left,align=right] node {mTTV} (1);
\path[draw] (12) edge [right,align=left] node {mTTV} (2);
\normalsize

\end{tikzpicture}
\end{center}
}\only<3>{
\begin{center}
\begin{tikzpicture}

\node (12345) at (4,2) {$\{1,2,3,4,5\}$};
\node (12) at (2,1) {$\{1,2\}$};
\node (345) at (6,1) {$\{3,4,5\}$};
\node (1) at (1,0) {$\Mn{M}{1}$};
\node (2) at (3,0) {$\Mn{M}{2}$};
\node (3) at (5,0) {$\Mn{M}{3}$};
\node (45) at (7,0) {$\{4,5\}$};
\node (4) at (6,-1) {$\Mn{M}{4}$};
\node (5) at (8,-1) {$\Mn{M}{5}$};

\scriptsize
\path[draw] (12345) edge [left,align=right] node {PM} (12);
\path[draw] (12345) edge [right,align=left] node {PM} (345);
\path[draw] (12) edge [left,align=right] node {mTTV} (1);
\path[draw] (12) edge [right,align=left] node {mTTV} (2);
\path[draw] (345) edge [left,align=right] node {mTTV} (3);
\path[draw] (345) edge [right,align=left] node {mTTV} (45);
\path[draw] (45) edge [left,align=right] node {mTTV} (4);
\path[draw] (45) edge [right,align=left] node {mTTV} (5);
\normalsize

\end{tikzpicture}
\end{center}
}

\only<2-3>{
\begin{textblock}{15}(2.5,14.5)
\footnotesize
PM = Partial MTTKRP \hspace{1cm} mTTV = multi-Tensor-Times-Vector
\end{textblock}
}

\end{frame}

\begin{frame}
\frametitle{Implementation}

\begin{itemize}
	\item Uses CP-ALS for non-negative CP problems
	\begin{itemize}
		\item minimize least squares loss function
		\item use block principal pivoting \cite{KP11} to solve subproblems
	\end{itemize}
	\vfill
	\item Avoids redundant communication across MTTKRPs
	\vfill
	\item Avoids redundant computation across MTTKRPs \\ using dimension trees 
\end{itemize}

\end{frame}

% input file with plotting macros
\input{fig/plots}

\begin{frame}
\frametitle{Strong Scaling Results (3D)}

\begin{figure}
\begin{tikzpicture}
\renewcommand{\datafile}{data/str_3D_syn.dat}
\renewcommand{\numiterations}{1}
\liavastrue
\strongscalingplot
\end{tikzpicture}
\caption{$1024{\times} 1024{\times} 1024$ tensor on $2^k{\times} 2^k{\times} 2^k$ proc grids ($R=32$)}
\end{figure}

\end{frame}

\begin{frame}
\frametitle{Strong Scaling Results (5D)}

\begin{figure}
\begin{tikzpicture}
\renewcommand{\datafile}{data/str_5D_syn.dat}
\renewcommand{\numiterations}{10}
\liavasfalse
\strongscalingplot
\end{tikzpicture}
\caption{$64\times 64\times 64\times 64\times 64$ tensor ($R=32$)}
\end{figure}

\end{frame}

\begin{frame}
\frametitle{Varying Rank Results (3D)}

\begin{figure}
\begin{tikzpicture}
\begin{axis}[legend style={draw=none, anchor=north,  cells={align=left}, nodes={scale=0.7}}, legend pos=north west, xlabel=Low Rank $R$, ylabel=Time(s), ylabel near ticks]
	\addplot+ [discard if not={alg}{D},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
	\addplot+ [discard if not={alg}{N},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
	\addplot+ [discard if not={alg}{L},thick,mark options={solid},mark=square*] table [x={k}, y expr=(\thisrow{total}/(\minvalue*\numiterations))] {data/kswp_neuro.dat};
	\legend{DimTree,NoDimTree,NbAO-NTF \cite{LK+17b}}
\end{axis}
\end{tikzpicture}
\caption{$30{,}012\times 1200 \times 500$ tensor on $120\times 6\times 2$ proc grid}
\label{fig:ksweepneuro}
\end{figure}

\end{frame}

\begin{frame}
\frametitle{Varying Rank Results (4D)}

\begin{figure}
\renewcommand{\datafile}{data/ksw_4D_HSI.dat}
\renewcommand{\numiterations}{10}
\scalebox{.75}{
\begin{tikzpicture}
\begin{axis}[	
	ybar stacked,
	bar width=8pt,
	width=\columnwidth,
	height =.75\columnwidth,
	%width=9cm, height=3.85cm,
	ylabel={Time (s)}, 
	xlabel={Rank $R$},
	y label style={yshift=-.15cm},
	ymin=0,
	%symbolic x coords={D10,N10,,D20,N20,,D30,N30,,D40,N40,,D50,N50},
	symbolic x coords={D10,D20,D30,D40,D50},
	%xtick=data,
	xticklabels={,10,,20,,30,,40,,50},
	legend style={at={(0.5,1.3)},anchor=north},
	legend columns=-1,
]
	\setcolors
	\addplot table[x=alg-K, y expr=(\thisrow{mttkrp}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=alg-K, y expr=(\thisrow{mttv}/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=alg-K, y expr=((\thisrow{nnls}+\thisrow{gram})/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=alg-K, y expr=((\thisrow{allgather}+\thisrow{reducescatter})/(\minvalue*\numiterations))] {\datafile};
	\addplot table[x=alg-K, y expr=(\thisrow{allreduce}/(\minvalue*\numiterations))] {\datafile};
	\legend{Partial MTTKRP,multi-TTV,Loc Update,Factor Comm,Gram Comm};
\end{axis}
%\labels
\end{tikzpicture}
}
\caption{$1344\times 1024\times 33 \times 9$ tensor on $8\times8\times1\times1$ proc grid}
\end{figure}

\end{frame}

\againframe{summary}

%%%%%%%%%%%%%%%%%%%%
% reference slides

\begin{frame}[allowframebreaks]
\frametitle{References}
\tiny
\bibliographystyle{alpha}
\bibliography{local}
\normalsize
\end{frame}

\begin{frame}
\frametitle{Fibers}

\begin{center}

\scalebox{.8}{\includegraphics{fibers}}

\vfill

A tensor can be decomposed into the \emph{fibers} of each mode \\
(fibers are vectors -- fix all indices but one)

\end{center}

\end{frame}

\begin{frame}
\frametitle{Matricized Tensors}

\begin{center}

\scalebox{.25}{\includegraphics{unfoldings}}

\vfill

A tensor can be reshaped into a matrix, \\ 
called a \emph{matricized tensor} or \emph{unfolding}, for a given mode, \\
where each column is a fiber

\end{center}

\end{frame}

\begin{frame}
\frametitle{Sequential Communication Lower Bound}

\begin{theorem}
For sufficiently large $I$, any sequential MTTKRP algorithm performs at least
\begin{equation*}
\Omega\left(\frac{NIR}{M^{1-1/N}}\right)
\end{equation*}
loads and stores to/from slow memory.
\end{theorem}

\vfill

\small
\begin{itemize}
	\item $N$ is the number of modes
	\item $I$ is the number of tensor entries
	\item $R$ is the rank of the CP model
	\item $M$ is the size of the fast memory
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Communication-Optimal Sequential Algorithm (3D)}

\begin{columns}

\begin{column}{.55\textwidth}

\newcommand{\matdim}{4}
\newcommand{\mat}{\draw[black,shift={(-.5,-.5)}] (0,0) grid (\matdim,\matdim);}
\newcommand{\highlight}{gray!50}

\begin{center}
\begin{tikzpicture}[x={(-0.5cm,-0.4cm)}, y={(1cm,0cm)}, z={(0cm,1cm)},every node/.append style={transform shape},scale=.8]

% highlight tensor block and factor matrix subcolumns (do first to put in background)
% front face
\begin{scope}[canvas is yz plane at x=.5,shift={(1.5,-1.5)}]
	% highlight front face of tensor block
	\draw[dashed,fill=\highlight,shift={(0,0)}] (0,0) rectangle (-1,-1);
	\node[draw=none] at (-1.25,-.5) {\small $b$};
	\node[draw=none] at (-.5,-1.25) {\small $b$};
	% highlight block of 1st factor matrix
	\draw[fill=\highlight,shift={(-2.5,0)},xscale=.5] (0,0) rectangle (-1,-1);
	% highlight block of 2nd factor matrix
	\draw[fill=\highlight,shift={(0,-2.5)},yscale=.5] (0,0) rectangle (-1,-1);
	\only<2>{
	% highlight subcolumn of 1st factor matrix
	\draw[densely dotted,thick,shift={(-2.8,0)}] (0,0) -- (0,-1);
	% highlight subcolumn of 2nd factor matrix
	\draw[densely dotted,thick,shift={(0,-2.8)}] (0,0) -- (-1,0);
	}
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\matdim-.5),rotate=-90,shift={(.5,-4)}]
	% highlight block of 3rd factor matrix
	\draw[fill=\highlight,yscale=.5] (0,0) rectangle (-1,-1);
	\only<2>{
	% highlight subcolumn of 3rd factor matrix
	\draw[densely dotted,thick,shift={(0,-.3)}] (0,0) -- (-1,0);
	}
\end{scope}
% right face, but in middle of tensor
\begin{scope}[canvas is zx plane at y=1.5,rotate=-90,shift={(-.5,-2.5)}]
	% highlight right face of tensor block
	\draw[dashed,fill=\highlight] (0,0) rectangle (1,1);
	\node[draw=none] at (.75,-.2) {\small $b$};
\end{scope}
% top face, but in middle of tensor
\begin{scope}[canvas is yx plane at z=-1.5,yscale=-1,rotate=0,shift={(.5,-.5)}]
	\draw (0,0) node {\LARGE $\cdot$};
	% highlight top face of tensor block
	\draw[dashed,fill=\highlight] (0,0) rectangle (1,1);
\end{scope}

% draw tensor
% front face
\begin{scope}[canvas is yz plane at x=.5,rotate=-90]
	\mat
\end{scope}
% top face
\begin{scope}[canvas is yx plane at z=.5,yscale=-1,rotate=0]
	\mat
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\matdim-.5),rotate=180]
	\mat
\end{scope}

% draw factor matrices
% front face
\begin{scope}[canvas is yz plane at x=.5,shift={(1.5,-1.5)}]
	% draw 1st factor matrix
	\draw[shift={(-2.5,0)},xscale=.5] (0,-2) grid (-1,2);
	\node[draw=none] at (-3.5,0) {$\Mn{U}{1}$};
	% draw 2nd factor matrix
	\draw[shift={(0,-2.5)},yscale=.5] (-2,0) grid (2,-1);
	\node[draw=none] at (0,-3.5) {$\Mn{M}{2}$};
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\matdim-.5),rotate=-90,shift={(1.5,-1.5)}]
	% draw 3nd factor matrix
	\draw[shift={(0,-2.5)},yscale=.5] (-2,0) grid (2,-1);
	\node[draw=none] at (0,-3.5) {$\Mn{U}{3}$};
\end{scope}

\end{tikzpicture}
\end{center}

\end{column}
\begin{column}{.45\textwidth}

\footnotesize
\begin{enumerate}
	\item Loop over $b\times \cdots \times b$ blocks of the tensor
	\uncover<2>{\item With block in memory, loop over subcolumns of input factor matrices, updating corresponding subcolumn of output matrix}
\end{enumerate}

\vfill

\footnotesize
\begin{itemize}
	\uncover<2>{\item choose $b \approx M^{1/N}$}
\end{itemize}

\end{column}

\end{columns}

\end{frame}

\begin{frame}
\frametitle{Theoretical Comparisons}

\begin{center}
\renewcommand{\arraystretch}{2}
\begin{tabular}{|c|ccc|}
\hline
& \textbf{Lower Bound} & \textbf{New Algorithm} & \textbf{Standard (MM)} \\
\hline
\textbf{Flops} & - & $NIR$ & $2IR$ \\
\textbf{Words} & $\Omega\left( \frac{NIR}{M^{1-1/N}} \right)$ & $O\left( I+\frac{NIR}{M^{1-1/N}} \right)$ & $O\left( I+\frac{IR}{M^{1/2}} \right)$ \\
\textbf{Temp Mem} & - & - & $\frac{IR}{I_n}$ \\
\hline
\end{tabular}
\end{center}

\vfill
\pause

\begin{itemize}
	\item New algorithm performs $N/2$ more flops than standard
	\item For relatively small $R$, $I$ term dominates communication
	\begin{itemize}
		\item we expect this to be the typical case in practice
	\end{itemize}
	\item For relatively large $R$, new algorithm communicates less
	\begin{itemize}
		\item better exponent on $M$
	\end{itemize}
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{MTTKRP Loop Nest}

\begin{algorithmic}
\For{$i_1=1$ to $I_1$}
	\ForDots
		\For{$i_N=1$ to $I_N$}
			\For{$r=1$ to $R$}
				\State \footnotesize $\Mn{M}{n}(i_n,r) \plusequals \T{X}(i_1,\dots,i_N) * \Mn{U}{1}(i_1,r) * \cdots * \Mn{U}{N}(i_N,r)$
			\EndFor
		\EndFor
	\EndForDots
\EndFor
\end{algorithmic}

\vfill

\begin{center}
$\Large \Delta= \quad\quad$
\begin{tabular}{c|cccccc}
& $i_1$ & $\cdots$ & $i_n$ & $\cdots$ & $i_N$ & $r$ \\
\hline 
$\Mn{U}{1}$ & 1 & & & & & 1 \\
$\vdots$ & & $\ddots$ & & & & $\vdots$ \\
$\Mn{M}{n}$ & & & 1 & & & 1 \\ 
$\vdots$ & & & & $\ddots$ & & $\vdots$ \\
$\Mn{U}{N}$ & & & & & 1 & 1 \\
$\T{X}$ & 1 & $\cdots$ & 1 & $\cdots$ & 1 & 
\end{tabular}
\end{center}

\end{frame}

\end{document}

### 2019-01-16 latex/toms/ack.tex

\section{Acknowledgements}
\small

This material is based upon work supported by the National Science Foundation Graduate Research Fellowship under Grant No. OAC-1642385.
This manuscript has been co-authored by UT-Battelle, LLC under Contract No. DE-AC05-00OR22725 with the U.S. Department of Energy.  This project was partially funded by the Laboratory Director's Research and Development fund. This research used resources of the Oak Ridge Leadership Computing Facility at the Oak Ridge National Laboratory, which is supported by the Office of Science of the U.S. Department of Energy.

%This research used resources of the National Energy Research Scientific Computing Center, a DOE Office of Science User Facility supported by the Office of Science of the U.S. Department of Energy under Contract No. DE-AC02-05CH11231.

Also, partial funding for this work was provided by AFOSR Grant FA9550-13-1-0100, National Science Foundation (NSF) grants IIS-1348152, ACI-1338745, and ACI-1642385, Defense Advanced Research Projects Agency (DARPA) XDATA program grant FA8750-12-2-0309.
%We also thank NSF for the travel grant to present this work in the conference through the grant CCF-1552229. 
%The United States Government retains and the publisher, by accepting the article for publication, acknowledges that the United States Government retains a non-exclusive, paid-up, irrevocable, world-wide license to publish or reproduce the published form of this manuscript, or allow others to do so, for United States Government purposes. The Department of Energy will provide public access to these results of federally sponsored research in accordance with the DOE Public Access Plan (\url{http://energy.gov/downloads/doepublic-access-plan}). 
 Any opinions, findings and conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the views of the USDOE, NERSC, AFOSR, NSF or DARPA.

 \normalsize

### 2019-01-16 latex/toms/experiments.tex

@@
 }
-\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size 162x162x162x162 on 81 Titan Nodes as 3x3x3x3 Processor Grid.}
+\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size 384x384x384x384 on 81 Titan Nodes as 3x3x3x3 Processor Grid.}
 \label{fig:convergencelowrank}
@@
 	}
-	\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Realworld Low Rank Tensor of size $200 \times 500 \times 64 \times 192$ on 64 Titan Nodes as $2\times4\times2\times4$ Processor Grid.}
+	\caption{Relative error comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Realworld Low Rank Tensor of size $200 \times 500 \times 64 \times 192$ on 64 Titan Nodes as $4\times4\times1\times4$ Processor Grid.}
 	\label{fig:convergencerealworld}
@@
 }
-\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size 162x162x162x162 on 81 Titan Nodes as 3x3x3x3 Processor Grid on CPU and GPU.}
+\caption{Timing comparison of \MU, \HALS, \BPP\, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size 384x384x384x384 on 81 Titan Nodes as 3x3x3x3 Processor Grid on CPU and GPU.}
 \label{fig:cpuvsgpulowrank}

### 2019-01-16 latex/toms/introduction.tex

@@
+% !TEX root = paper.tex
+
 \section{Introduction}
 
-Non-negative Matrix Factorization (NMF) is the problem of finding two low rank factors $\WW\in \Rnplus{m\times k}$ and $\HH\in \Rnplus{k \times n}$ for a given input matrix  $\AA\in \Rnplus{m\times n}$, such that $\AA \approx \WW \HH$.
-Here, $\Rnplus{m\times n}$ denotes the set of $m \times n$ matrices with non-negative real values.
-Formally, the NMF problem \cite{seung2001algorithms} can be defined as \SplitN{\label{eqn:original NMF}}{
-\min_{\WW \geq 0,\HH \geq 0} & \|\AA-\WW\HH\|_F,
+%The CP decomposition is a low-rank approximation of a multi-dimensional array, or tensor, which generalizes matrix approximations like the truncated singular value decomposition. 
+%As in Figure \ref{fig:cpdecomposition}, CP decomposition approximates the given input matrix as sum of $k$ rank-1 tensors.  
+The CP decomposition is a low-rank approximation of a multi-dimensional array, or tensor, which generalizes matrix approximations like the truncated singular value decomposition.
+It approximates the input tensor by a sum of rank-one tensors, which are outer products of vectors.
+CP is often used for finding hidden patterns, or latent factors, within tensor data, particularly when the goal is to interpret the factors, and it is popular within the signal processing, machine learning, and scientific computing communities.
+
+%\begin{figure}[htbp]
+%\begin{center}
+%\includegraphics[width=3in, height=1.5in]{fig/cpdecomposition.pdf}
+%\caption{CP Decomposition}
+%\label{fig:cpdecomposition}
+%\end{center}
+%\end{figure}
+
+To aid in interpretability, domain-specific constraints are often imposed on the computed factors.
+We focus in this paper on dense tensors (when nearly all of the tensor entries are nonzero) and on constraining solutions to have nonnegative entries, which is useful when the tensor data itself is nonnegative.
+Formally, NNCP can be defined as
+\SplitN{\label{eqn:nncp}}{
+ \min_{\HH^{(i)} \geq 0} & \| \TA - \sum_{r=1}^R \Mn{H}{1}(:,r) \circ \cdots \circ \Mn{H}{N}(:,r) \|^2
 }
-where $\|\M{X}\|_F=(\sum_{ij} x_{ij}^2)^{1/2}$ is the Frobenius norm.
+where $\Mn{H}{1}(:,i) \circ \cdots \circ \Mn{H}{N}(:,i)$ is the outer product of the $i^{th}$ vector from all the $N$ factors that yields a rank one tensor $\T{M}$ and $\sum_{r=1}^R \Mn{H}{1}(:,r) \circ \cdots \circ \Mn{H}{N}(:,r)$ results in a sum of $R$ rank one tensors that will be of the same dimension as the input tensor $\TA$.
+For example, in imaging and microscopy applications, tensor values often correspond to intensities, and NNCP can be used to cluster and analyze the data in a lower-dimensional space \cite{JC+16}.
+In this work, we consider two such applications: a series of time-lapse hyperspectral images \cite{FAN16} and a dynamic functional correlation data set generated from functional magnetic resonance images of human brains \cite{VEU+12}.
 
-NMF is widely used in data mining and machine learning as a dimension reduction and factor analysis method. 
-It is a natural fit for many real world problems as the non-negativity is inherent in many representations of real-world data and
-the resulting low rank factors are expected to have a natural interpretation.
-The applications of NMF range from text mining \cite{pauca2004text},  computer vision \cite{hoyer2004non,Fujimoto2014,BSJJZ2015}, and bioinformatics \cite{kim2007sparse} to blind source separation  \cite{cichocki2009nonnegative}, unsupervised clustering \cite{kuang2012symmetric,kuang2013symnmf}  and many other areas.
-In the typical case, $k \ll \min(m,n)$; for problems today, $m$ and $n$ can be on the order of millions or more, and $k$ is on the order of few tens to thousands.
+One approach to handling multidimensional data is to ``matricize'' it, combining sets of modes to reshape the data into a matrix, so that standard matrix  methods like principal component analysis or nonnegative matrix factorization can be applied.
+While this approach can be effective in certain cases, reshaping the data destroys multidimensional relationships among entries that the matrix methods cannot recover.
+By maintaining the tensor structure of the data, the low-rank representations preserve these relationships, often producing better and more interpretable results.
 
-There is a vast literature
-on algorithms for NMF and their convergence properties \cite{kim2013nonnegative}.   
-The commonly adopted NMF algorithms are -- (i) Multiplicative Update (\MU) \cite{seung2001algorithms} (ii) Hierarchical Alternating Least Squares (\HALS) \cite{cichocki2009nonnegative,Ho2008} (iii) NMF based on Alternating Nonnegative Least Squares and Block Principal Pivoting (\BPP) \cite{kim2011fast}, and (iv) Stochastic Gradient Descent (SGD) Updates \cite{gemulla2011large}.
-Most of the algorithms in NMF literature are based on alternately optimizing each of the low rank factors $\WW$ and $\HH$ while keeping the other fixed, in which case each subproblem is a constrained convex optimization problem.
-Subproblems can then be solved using standard optimization techniques such as projected gradient or interior point method; a detailed survey for solving such problems can be found in \cite{xiong2013survey,kim2013nonnegative}.
-In this paper, our implementation uses either \BPP, \MU, or \HALS. 
-But our parallel framework is extensible to other algorithms (e.g., \cite{XY2013,HSL2016}) as-is or with a few modifications, as long as they fit an alternating-updating framework (defined in Section \ref{sec:aunmf}).
+However, tensor methods are more complicated both mathematically and computationally.
+The kernel computations within standard algorithms for computing NNCP can be formulated as matrix computations, but the complicated layout of tensors in memory prevents the straightforward use of BLAS and LAPACK libraries.
+In particular, the matrix formulation of subcomputations involve different views of the tensor data, so no single layout yields a column- or row-major matrix layout for all subcomputations.
+Likewise, the parallelization approach for tensor methods is not a straightforward application of parallel matrix computation algorithms.
 
-With the advent of large scale internet data and interest in Big Data, researchers have started studying scalability of many foundational machine learning algorithms. 
-To illustrate the dimension of matrices commonly used in the machine learning community, we present a few examples. 
-Nowadays the adjacency matrix of a billion-node social network is common. 
-In the matrix representation of a video data, every frame contains three matrices for each RGB color, which is reshaped into a column.  
-Thus in the case of a 4K video, every frame will take approximately 27 million rows (4096 row pixels x 2196 column pixels x 3 colors). 
-Similarly, the popular representation of documents in text mining is a bag-of-words matrix, where the rows are the dictionary and the columns are the documents (e.g., webpages). 
-Each entry $A_{ij}$ in the bag-of-words matrix is generally the frequency count of the word $i$  in the document $j$. 
-Typically with the explosion of the new terms in social media, the number of words spans to millions. 
-To handle such high-dimensional matrices, it is important to study low-rank approximation methods in a data-distributed and parallel computing environment.
+In developing an efficient parallel algorithm for computing a NNCP of a dense tensor, the key is to parallelize the bottleneck computation known as Matricized-Tensor Times Khatri-Rao Product (MTTKRP), which is performed repeatedly for each mode of the tensor.
+The parallelization must load balance the computation, minimize communication across processors, and distribute the results so that the rest of the computation can be performed independently.
+In our algorithm, not only do we load balance the computation, but we also compute and store temporary values that can be used across MTTKRPs of different modes using a technique known as dimension trees, significantly reducing the computational cost compared to standard approaches.
+Our parallelization strategy also avoids communicating tensor entries and minimizes the communication of factor matrix entries, helping the algorithm to remain computation bound and scalable to high core counts.
 
-In this work, we present an efficient algorithm and implementation using tools from the field of High-Performance Computing (HPC).
-We maintain data in memory (distributed across processors), take advantage of optimized libraries like BLAS and LAPACK for local computational routines, and use the Message Passing Interface (MPI) standard to organize interprocessor communication.
-%In particular, we use an MPI/C++ implementation, which offers several technical advantages such as (1) the ability to leverage state-of-the-art hardware, (2) efficient networks for communication between nodes, and (3) the availability of numerically stable and efficient BLAS and LAPACK routines.
-Furthermore, the current hardware trend is that available parallelism (and therefore aggregate computational rate) is increasing much more quickly than improvements in network bandwidth and latency, which implies that the relative cost of communication (compared to computation) is increasing.
-To address this challenge, we analyze algorithms in terms of both their computation and communication costs.
-In particular, we prove in Section \ref{sec:parNMF} that in the case of dense input and under a mild assumption, our proposed algorithm minimizes the amount of data communicated between processors to within a constant factor of the lower bound.
+As we detail in the related work, the general techniques for reducing computation and communication have been used in similar contexts.
+The recomputation avoidance was proposed in a sequential algorithm \cite{PTC13a}, the parallelization scheme was proposed and analyzed for general tensors \cite{BKR17-TR}, and the algorithm was implemented for 3D tensors \cite{LK+17b}.
+%The main contribution of this paper is the combination and implementation of these ideas for general $N$-D tensors, and we demonstrate superior performance to the existing implementation for 3D tensors \cite{LK+17b}.
 
-We call our implementation MPI-FAUN, an MPI-based Framework for Alternating-Updating Nonnegative matrix factorization algorithms.
-A key attribute of our framework is that the efficiency does not require a loss of generality of NMF algorithms.
-Our central observation is that most NMF algorithms, in particular those that alternate between updating each factor matrix, consist of two main tasks: (a) performing matrix multiplications and (b) solving Non-negative Least Squares (NLS) subproblems, either approximately or exactly.
-More importantly, NMF algorithms tend to perform the same matrix multiplications, differing only in how they solve NLS subproblems, and the matrix multiplications often dominate the running time of the algorithms.
-Our framework is designed to perform the matrix multiplications efficiently and organize the data so that the NLS subproblems can be solved independently in parallel, leveraging any of a number of possible methods.
-We explore the overall efficiency of the framework and compare three different NMF methods in Section \ref{sec:experiment}, performing convergence, scalability, and parameter-tuning experiments on over 1500 processors.
-
-\begin{table}[htp]
-\begin{center}
-\begin{tabular}{|c|c|c|c|c|}
-\hline 
-Dataset & Type & Matrix size & NMF Time \\ \hline
-Video & Dense & 1 Million x 13,824 &  5.73 seconds \\
-Stack Exchange & Sparse & 627,047 x 12 Million &  67 seconds \\
-Webbase-2001 & Sparse & 118 Million x 118 Million & 25 minutes \\ \hline
-\end{tabular}
-\end{center}
-\caption{\ParNMF\ on large real-world datasets. Reported time is for 30 iterations on 1536 processors with a low rank of 50.}
-\label{tab:teaser}
-\end{table}%
-
-With our framework, we are able to explore several large-scale synthetic and real-world data sets, some dense and some sparse.
-In Table \ref{tab:teaser}, we present the NMF computation wall clock time on some very large real world datasets.
-We describe the results of the computation in Section \ref{sec:experiment}, showing the range of application of NMF and the ability of our framework to scale to large data sets.
-
-%For example, we apply NMF to a surveillance video data set of size 44 GB in order to isolate the moving objects from the background view; once loaded into memory, the most efficient algorithm requires only 5.73 seconds on 1536 processors.
-%We also apply our techniques to sparse data, analyzing Stack Exchange posts (text data) to infer discussion topics; this data set is 44 GB of XML file which takes 6.1 GB of sparse matrix representation, and our NMF implementation requires 67 seconds on 1536 processors to obtain meaningful results
-%Similarly, our algorithm can perform graph clustering using NMF on a very large sparse graph with 118 million nodes and 1 billion edges and the sparse representation of this matrix takes 18GB on disk
-%in nearly 15 minutes using 1536 processors, 
-%\grey{do we want to report that the output dense matrices required 88GB?}.
-%\grey{break up into multiple sentences}
-
-A preliminary version of this work has already appeared as a conference paper \cite{KBP16}.
-While the focus of the previous work was parallel performance of \BPP\ (Alternating Nonnegative Least Squares and Block Principal Pivoting), the goal of this paper is to explore more data analytic questions.
-In particular, the new contributions of this paper include (1) implementing a software framework to compare \BPP\ with \MU\ (Multiplicative Update) and \HALS\ (Hierarchical Alternating Least Squares) for large scale data sets, (2) benchmarking on a data analysis cluster and scaling up to over 1500 processors, and (3) providing an interpretation of results for real-world data sets.
-We provide a detailed comparison with other related work, including MapReduce implementations of NMF, in Section \ref{sec:related}.
-
-Our main contribution is a new, high-performance parallel computational framework for a broad class of NMF algorithms. 
-The framework is efficient, scalable, flexible, and demonstrated to be effective for large-scale dense and sparse matrices.  
-Based on our survey and knowledge, we are the fastest NMF implementation available in the literature.  
-The code and the datasets used for conducting the experiments can be downloaded from \url{https://github.com/ramkikannan/nmflibrary}.
+We summarize our main contributions as follows:
+\begin{itemize}
+		\item we present the first distributed-memory parallel implementation of NNCP algorithms for arbitrary-dimension dense tensors,
+		\item we optimize the use of dimension trees for dense tensors, avoiding recomputation across multiple MTTKRPs,
+		\item our parallel algorithm is communication optimal with a carefully chosen processor grid,
+		\item we demonstrate a performance improvement of up to $2.2\times$ over the existing state-of-the-art parallel software on 3D tensors,
+		\item our implementation obtains efficient parallel scaling of up to $1771\times$ on 4096 cores.
+\end{itemize}

### 2019-01-25 latex/toms/experiments.tex
changed the size of the synthetic low rank experiments

@@
 }
-\caption{LUC comparison of \MU, \HALS, \BPP, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size 162x162x162x162 on 81 Titan nodes as 3x3x3x3 Processor Grid on CPU and Realworld TinyImagenet Dataset on 64 Titan nodes as 2x4x2x4 processor grid}
+\caption{LUC comparison of \MU, \HALS, \BPP, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size $384 \times 384 \times 384 \times 384$ on 81 Titan nodes as 3x3x3x3 Processor Grid on CPU and Realworld TinyImagenet Dataset on 64 Titan nodes as 2x4x2x4 processor grid}
 \label{fig:luccomp}
@@
 \includegraphics[width=\textwidth, height=2.5in]{data/plots/confsweep4d.pdf}
-\caption{Processor Grid sweep of a 162x162x162x162 synthetic low rank tensor on 81 nodes in Titan as 81x1x1x1, 27x3x1x1, 9x3x3x3, 3x3x3x3 grids represented as p=1,2,3,4 respectively.
+\caption{Processor Grid sweep of a $384 \times 384 \times 384 \times 384$ synthetic low rank tensor on 81 nodes in Titan as 81x1x1x1, 27x3x1x1, 9x3x3x3, 3x3x3x3 grids represented as p=1,2,3,4 respectively.
 The algorithms alg=0,1,2,4,5 are \MU, \HALS, \BPP, \ADMM, \Nestrov\ respectively.}

### 2019-02-08 latex/toms/experiments.tex
minor edits/organization of experiments section

@@
 \subsection{Datasets}
+
+\GB{This section is out of date...}
 
@@
 \subsection{Machine Details}
+
+\GB{This section is out of date...}
+
 The entire experimentation was performed on Eos, a supercomputer at the Oak Ridge Leadership Computing Facility. 
@@
 
-\subsection{Comparison Implementations}
-The implementation proposed by Liavas et al.
-\cite{LK+17b} is the only publicly available distributed-memory software (of which we are aware) for computing the CP decomposition of dense tensors, with or without constraints.
-We use the acronym NbAO-NTF for Nesterov-based Alternating Optimization Nonnegative Tensor Factorization to refer to their code.
+\subsection{Microbenchmarks with Synthetic Data}
 
-It is based on the same parallel algorithm as our implementation, though it is limited to 3D tensors.
-The code uses MPI collectives for communication and Eigen \cite{eigenweb} as an interface to BLAS and LAPACK.
-We compiled the code linked to BLAS/LAPACK wrappers from Cray  (the same BLAS implementation used by our code) but we were unable to run multithreaded BLAS with their code.
-For fair comparison, we use a flat MPI configuration (one MPI process per core) on all comparisons between the two implementations.
+\subsubsection{Comparison across Algorithms}
 
-We also point out a difference between the Nesterov-based algorithm and the BPP algorithm for solving the NLS subproblems.
-The Nesterov-based algorithm attempts an acceleration step using a linear combination of the current and proposed future step; however, it re-computes the residual error before deciding whether or not to accept or reject the acceleration step.
-This residual error cannot always be computed cheaply, using the technique described in \cref{sec:error}, and it contributes significantly (approximately 25\%) to the overall run time.
-Because the BPP algorithm does not require this extra computation, and studying convergence behavior of the different NLS algorithms is beyond the scope of this work, we remove the time spent in the acceleration step of NbAO-NTF in all our comparisons.
+\begin{itemize}
+	\item MU and CP-ALS clearly cheapest, should scale like $k^2$
+	\item Nesterov most expensive, due to inner iterations
+	\item HALS, AO-ADMM, Nesterov all involve communication during ``LUC"
+\end{itemize}
 
-Our proposed algorithm uses dimension trees, but we also benchmark our implementation without that optimization to highlight its importance.
-We use an existing implementation to perform the individual MTTKRPs \cite{HBJT18} with this approach.
+%%%%%%%%%%%%%%%%%
+%LUC Running Time Comparison
+%%%%%%%%%%%%%%%%%
+
+\begin{figure}
+\subfloat[Synthetic Low Rank \label{fig:luclr81}]{
+\includegraphics[width=0.45\textwidth, height=2in]{data/plots/luclr81.pdf}
+}
+\subfloat[Tiny Imagenet \label{fig:lucrw}]{
+\includegraphics[width=0.45\textwidth, height=2in]{data/plots/lucrw64.pdf}
+}
+\caption{LUC comparison of \MU, \HALS, \BPP, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size $384 \times 384 \times 384 \times 384$ on 81 Titan nodes as 3x3x3x3 Processor Grid on CPU and Realworld TinyImagenet Dataset on 64 Titan nodes as 2x4x2x4 processor grid}
+\label{fig:luccomp}
+\end{figure}
+
+\subsubsection{Comparison across Processor Grids}
+
+%%%%%%%%%%%%%%%%%
+%Configuration Sweep
+%%%%%%%%%%%%%%%%%
+\begin{figure}
+\includegraphics[width=\textwidth, height=2.5in]{data/plots/confsweep4d.pdf}
+\caption{Processor Grid sweep of a $384 \times 384 \times 384 \times 384$ synthetic low rank tensor on 81 nodes in Titan as 81x1x1x1, 27x3x1x1, 9x3x3x3, 3x3x3x3 grids represented as p=1,2,3,4 respectively.
+The algorithms alg=0,1,2,4,5 are \MU, \HALS, \BPP, \ADMM, \Nestrov\ respectively.}
+\label{fig:confsweep4D}
+\end{figure}
+
+\subsection{Scaling Studies with Synthetic Data}
+
+\subsubsection{Weak Scaling}
+
+%%%%%%%%%%%%%%%%%
+%Weak Scaling Synthetic Low Rank
+%%%%%%%%%%%%%%%%%
+\begin{figure}
+\subfloat[Synthetic 3D Low Rank \label{fig:wksca3d}]{
+\includegraphics[width=\textwidth, height=3in]{data/plots/wksca3d.pdf}
+}\\
+\subfloat[Synthetic 4D Low Rank \label{fig:wksca4d}]{
+\includegraphics[width=\textwidth, height=3in]{data/plots/wksca4d.pdf}
+}
+\caption {Weak Scaling on Synthetic 3D and 4D low rank tensors.
+For 3D, the input tensor is of size 128x128x128, 256x256x256, 378x378x378 and 512x512x512 on 1, 8, 27 and 64 Titan Nodes.
+The 4D tensors were of size 128, 256, 405, 512 for each mode on 1, 16, 81 and 256 Titan nodes.
+The algorithms alg=0,1,2,4,5 are \MU, \HALS, \BPP, \ADMM, \Nestrov\ respectively.}
+\label{fig:synweakscaling}
+\end{figure}
+
+\subsubsection{Strong Scaling}
+
+%%%%%%%%%%%%%%%%%
+%Strong Scaling Synthetic Low Rank
+%%%%%%%%%%%%%%%%%
+\begin{figure}
+\subfloat[Synthetic 3D Low Rank \label{fig:strsca3d}]{
+\includegraphics[width=\textwidth, height=3in]{data/plots/strsca3d.pdf}
+}\\
+\subfloat[Synthetic 4D Low Rank \label{fig:strsca4d}]{
+\includegraphics[width=\textwidth, height=3in]{data/plots/strsca4d.pdf}
+}
+\caption {Strong Scaling on Synthetic 3D and 4D low rank tensors.
+For 3D, the input tensor is of size 1024x1024x1024 and 256x256x256x256 for 4D on 8,16, 32, 64 and 128 nodes in Titan.
+The algorithms alg=0,1,2,4,5 are \MU, \HALS, \BPP, \ADMM, \Nestrov\ respectively.}
+\label{fig:synstrongscaling}
+\end{figure}
+
+\GB{Everything after this point is still disorganized...}
 
@@
 
-%%%%%%%%%%%%%%%%%
-%LUC Running Time Comparison
-%%%%%%%%%%%%%%%%%
-
-\begin{figure}
-\subfloat[Synthetic Low Rank \label{fig:luclr81}]{
-\includegraphics[width=0.45\textwidth, height=2in]{data/plots/luclr81.pdf}
-}
-\subfloat[Tiny Imagenet \label{fig:lucrw}]{
-\includegraphics[width=0.45\textwidth, height=2in]{data/plots/lucrw64.pdf}
-}
-\caption{LUC comparison of \MU, \HALS, \BPP, \ADMM, \Nestrov\ on 4D Synthetic Low Rank Tensor of size $384 \times 384 \times 384 \times 384$ on 81 Titan nodes as 3x3x3x3 Processor Grid on CPU and Realworld TinyImagenet Dataset on 64 Titan nodes as 2x4x2x4 processor grid}
-\label{fig:luccomp}
-\end{figure}
-
 \subsection{Strong Scaling Time Breakdown}
-
-%%%%%%%%%%%%%%%%%
-%Strong Scaling Synthetic Low Rank
-%%%%%%%%%%%%%%%%%
-\begin{figure}
-\subfloat[Synthetic 3D Low Rank \label{fig:strsca3d}]{
-\includegraphics[width=\textwidth, height=3in]{data/plots/strsca3d.pdf}
-}\\
-\subfloat[Synthetic 4D Low Rank \label{fig:strsca4d}]{
-\includegraphics[width=\textwidth, height=3in]{data/plots/strsca4d.pdf}
-}
-\caption {Strong Scaling on Synthetic 3D and 4D low rank tensors.
-For 3D, the input tensor is of size 1024x1024x1024 and 256x256x256x256 for 4D on 8,16, 32, 64 and 128 nodes in Titan.
-The algorithms alg=0,1,2,4,5 are \MU, \HALS, \BPP, \ADMM, \Nestrov\ respectively.}
-\label{fig:synstrongscaling}
-\end{figure}
 
@@
 
-%%%%%%%%%%%%%%%%%
-%Weak Scaling Synthetic Low Rank
-%%%%%%%%%%%%%%%%%
-\begin{figure}
-\subfloat[Synthetic 3D Low Rank \label{fig:wksca3d}]{
-\includegraphics[width=\textwidth, height=3in]{data/plots/wksca3d.pdf}
-}\\
-\subfloat[Synthetic 4D Low Rank \label{fig:wksca4d}]{
-\includegraphics[width=\textwidth, height=3in]{data/plots/wksca4d.pdf}
-}
-\caption {Weak Scaling on Synthetic 3D and 4D low rank tensors.
-For 3D, the input tensor is of size 128x128x128, 256x256x256, 378x378x378 and 512x512x512 on 1, 8, 27 and 64 Titan Nodes.
-The 4D tensors were of size 128, 256, 405, 512 for each mode on 1, 16, 81 and 256 Titan nodes.
-The algorithms alg=0,1,2,4,5 are \MU, \HALS, \BPP, \ADMM, \Nestrov\ respectively.}
-\label{fig:synweakscaling}
-\end{figure}
-
 \subsection{Varying Processor Grid}
-
-%%%%%%%%%%%%%%%%%
-%Configuration Sweep
-%%%%%%%%%%%%%%%%%
-\begin{figure}
-\includegraphics[width=\textwidth, height=2.5in]{data/plots/confsweep4d.pdf}
-\caption{Processor Grid sweep of a $384 \times 384 \times 384 \times 384$ synthetic low rank tensor on 81 nodes in Titan as 81x1x1x1, 27x3x1x1, 9x3x3x3, 3x3x3x3 grids represented as p=1,2,3,4 respectively.
-The algorithms alg=0,1,2,4,5 are \MU, \HALS, \BPP, \ADMM, \Nestrov\ respectively.}
-\label{fig:confsweep4D}
-\end{figure}
 

### 2019-02-14 latex/slides/CSAIL-workshop19/talk.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Beamer Presentation
% LaTeX Template
% Version 1.0 (10/11/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND THEMES
%----------------------------------------------------------------------------------------

\documentclass{beamer}

\mode<presentation> {

% The Beamer class comes with a number of default slide themes
% which change the colors and layouts of slides. Below this is a list
% of all the themes, uncomment each in turn to see what they look like.

%\usetheme{default}
%\usetheme{AnnArbor}
%\usetheme{Antibes}
%\usetheme{Bergen}
%\usetheme{Berkeley}
%\usetheme{Berlin}
%\usetheme{Boadilla}
%\usetheme{CambridgeUS}
%\usetheme{Copenhagen}
%\usetheme{Darmstadt}
%\usetheme{Dresden}
%\usetheme{Frankfurt}
%\usetheme{Goettingen}
%\usetheme{Hannover}
%\usetheme{Ilmenau}
%\usetheme{JuanLesPins}
%\usetheme{Luebeck}
\usetheme{Madrid}
%\usetheme{Malmoe}
%\usetheme{Marburg}
%\usetheme{Montpellier}
%\usetheme{PaloAlto}
%\usetheme{Pittsburgh}
%\usetheme{Rochester}
%\usetheme{Singapore}
%\usetheme{Szeged}
%\usetheme{Warsaw}

% As well as themes, the Beamer class has a number of color themes
% for any slide theme. Uncomment each of these in turn to see how it
% changes the colors of your current slide theme.

%\usecolortheme{albatross}
%\usecolortheme{beaver}
%\usecolortheme{beetle}
%\usecolortheme{crane}
%\usecolortheme{dolphin}
%\usecolortheme{dove}
%\usecolortheme{fly}
%\usecolortheme{lily}
%\usecolortheme{orchid}
%\usecolortheme{rose}
%\usecolortheme{seagull}
%\usecolortheme{seahorse}
%\usecolortheme{whale}
%\usecolortheme{wolverine}

%\setbeamertemplate{footline} % To remove the footer line in all slides uncomment this line
%\setbeamertemplate{footline}[page number] % To replace the footer line in all slides with a simple slide count uncomment this line

\setbeamertemplate{navigation symbols}{} % To remove the navigation symbols from the bottom of all slides uncomment this line
}

\usepackage{graphicx} % Allows including images
\usepackage{booktabs} % Allows the use of \toprule, \midrule and \bottomrule in tables
\usepackage{mathtools}
\usepackage{hyperref}

% My definitions
\newcommand{\ones}{\mathbbm{1}}
\newcommand{\e}{\vec{e}}
\newcommand{\tr}{\text{tr}}
\newcommand{\bs}{\boldsymbol}
\mathchardef\mhyphen="2D
\newcommand{\C}{\mathbb{C}}
\newcommand{\R}{\mathbb{R}}

% auto sized delimiters

\newcommand{\br}[1]{\left[#1\right]}
\newcommand{\pr}[1]{\left(#1\right)}
\newcommand{\ceil}[1]{\left\lceil#1\right\rceil}
\newcommand{\floor}[1]{\left\lfloor#1\right\rfloor}
\newcommand{\abs}[1]{\left|#1\right|}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\dom}[1]{\mathrm{dom}\pr{#1}}

%default delimiter for Pr and E
\DeclarePairedDelimiter{\defaultDelim}{[}{]}
\DeclarePairedDelimiter{\probDelim}{\{}{\}}
\DeclarePairedDelimiter{\roundDelim}{(}{)}

\DeclareMathOperator{\capPr}{\mathbb{P}}
\renewcommand{\Pr}[2][]{\capPr_{#1}\probDelim*{#2}}
\DeclareMathOperator{\capE}{\mathbb{E}}
\newcommand{\E}[2][]{\underset{#1}{\capE}\defaultDelim*{#2}}
\DeclareMathOperator{\capVar}{Var}
\newcommand{\Var}[2][]{\capVar_{#1}\defaultDelim*{#2}}
\DeclareMathOperator{\capVCD}{VCDim}
\newcommand{\VCD}[2][]{\capVCD_{#1}\roundDelim*{#2}}
%\DeclareMathOperator*{}{} puts subscript below

%% Tikz - For making pretty pictures
\usepackage{tikz}
\usetikzlibrary{3d}
\usetikzlibrary{patterns}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}

\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{etoolbox}
%\pgfplotsset{compat=1.12}
\usetikzlibrary{decorations.pathreplacing}
\makeatletter
\newcommand{\gettikzxy}[3]{%
  \tikz@scan@one@point\pgfutil@firstofone#1\relax
  \edef#2{\the\pgf@x}%
  \edef#3{\the\pgf@y}%
}
\makeatother

% Appendix Numbering
\usepackage{appendixnumberbeamer}

\graphicspath{{fig/}}
\input{tensor_header}

%----------------------------------------------------------------------------------------
%	TITLE PAGE
%----------------------------------------------------------------------------------------

\title[PLANC]{PLANC: Parallel Low Rank Approximations with Non-negativity Constraints} % The short title appears at the bottom of every slide, the full title is only on the title page

\author[Srinivas Eswar]{Ramakrishnan Kannan, Michael Matheson, Grey Ballard, \textbf{Srinivas Eswar}, Koby Hayashi, Haesun Park} % Your name
\institute[GT] % Your institution as it will appear on the bottom of every slide, may be shorthand to save space
{
Ph.D student, School of CSE \\
Georgia Institute of Technology \\ % Your institution for the title page
\medskip
%\textit{seswar3@gatech.edu} % Your email address
Advisors: Rich Vuduc and Haesun Park
}
\date{January 26, 2019} % Date, can be changed to a custom date
\titlegraphic{Workshop on Compiler Techniques for Sparse Tensor Algebra \\ \vspace{0.3in}
\scriptsize Acknowledgement: This work was partly sponsored by NSF, Sandia and ORNL}

\begin{document}

\begin{frame}
\titlepage % Print the title page as the first slide
\end{frame}

%\begin{frame}
%\frametitle{Overview} % Table of contents slide, comment this block out to remove it
%\tableofcontents % Throughout your presentation, if you choose to use \section{} and \subsection{} commands, these will automatically be printed on this slide as an overview of your presentation
%\end{frame}

%----------------------------------------------------------------------------------------
%	PRESENTATION SLIDES
%----------------------------------------------------------------------------------------

%------------------------------------------------
%\section{First Section} % Sections can be created in order to organize your presentation into discrete blocks, all sections and subsections are automatically printed in the table of contents as an overview of the talk
%------------------------------------------------

%\subsection{Subsection Example} % A subsection can be created just before a set of slides with a common theme to further break down your presentation into chunks
\begin{frame}[label=summary]
\frametitle{Summary}
\begin{itemize}
\item PLANC is an open source, scalable and flexible software package to compute Non-negative Tensor Factorisation.
\pause
\item Implements state of the art communication avoiding algorithm for matricised-tensor times Khatri-Rao product (MTTKRP).
\pause
\item Popular optimisation methods like Block Principal Pivoting, Alternating Direction Method of Multipliers, first order Nesterov methods etc. for Non-negative Least Squares are included. 
\pause
\item NTF is an important contributor towards explainable AI with a wide range of applications like spectral unmixing, scientific visualization, healthcare analytics, topic modelling etc.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{CP Decomposition}
Matrix: \quad $\displaystyle \M{M} \approx \sum_{r=1}^R \MC{U}{r} (\sigma_r \MC{V}{r}^T)$
\begin{center}
\scalebox{.07}{\includegraphics{matrix-LR-SOP}}
\end{center}
\vfill
Tensor: \quad $\displaystyle \T{X}\approx \sum_{r=1}^R (\lambda_r \MC{U}{r})\circ\MC{V}{r}\circ\MC{W}{r}$
\begin{center}
\scalebox{.15}{\includegraphics{tensor-LR-SOP}}
\end{center}
\vfill
\begin{itemize}
\item This is known as the CANDECOMP or PARAFAC or canonical polyadic or \textbf{CP decomposition}. It approximate tensors as sum of outer products or rank-1 tensors.
\item NNCP imposes \textbf{non-negativity} constraints on the factor matrices to aid interpretability.
\end{itemize}
\end{frame}

%\begin{frame}
%\frametitle{NNCP Decomposition}
%\begin{itemize} 
%\item NNCP imposes non-negativity constraints on each of the factor matrices.
%\begin{align*}
%\T{X}\approx \sum_{r=1}^R &(\lambda_r \MC{U}{r})\circ\MC{V}{r}\circ\MC{W}{r} \\
%\textnormal{subject to} \\
%\MC{u}{r} &\geq \V{0} \\
%\MC{v}{r} &\geq \V{0} \\
%\MC{w}{r} &\geq \V{0} 
%\end{align*}
%\item NNCP results in a ``parts"-based decomposition with each column $\MC{u}{r}, \MC{v}{r}, \MC{w}{r}$ can only be scaled and added together.
%\item This aids in the interpretability of the factorisation especially if the input data is non-negative.
%\end{itemize}
%\end{frame}

\begin{frame}
\frametitle{Computational Bottlenecks}
\begin{itemize}
\item The MTTKRP is the major bottleneck for NNCP.
$$\Mn{M}{1} = \Mz{X}{1}(\M{W} \Khat \M{V})$$
$$\ME{M}{ir} = \sum_{j=1}^J \sum_{k=1}^K \TE{X}{ijk} \ME{V}{jr} \ME{W}{kr}$$
\item Standard approach is to explicitly matricise the tensor and form the Khatri-Rao product before calling DGEMM.
\item Can we do better? Avoid matricisation of the tensor and full Khatri-Rao products ...
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Communication Lower Bounds}
\begin{itemize}
\item Following the nested arrays lower bounds~\cite{BKR2017}.
\begin{theorem}
Any parallel MTTKRP algorithm involving a tensor with $I_k=I^{1/N}$ for all $k$ and that evenly distributes one copy of the input and output performs at least
\begin{equation*}
\Omega\left( \left( \frac{NIR}{P} \right)^{\frac{N}{2N-1}} + NR\left(\frac{I}{P}\right)^{1/N} \right)
\end{equation*}
sends and receives.  (Either term can dominate.)
\end{theorem}
\item Key Assumptions: algorithm is not allowed to pre-compute and re-use temporary values.
\item $\Omega\pr{NR\pr{\frac I P}^{1/N}}$ is the most frequently occurring case for relatively small $P$ or $R$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Shared Memory Optimisation - Dimension Trees}
\begin{itemize}
\item Reuse computations across MTTKRPs.
$$\Mn{M}{1}=\underline{\Mz{X}{1}(\Mn{U}{3}}\Khat\Mn{U}{2}) \quad \Mn{M}{2}=\underline{\Mz{X}{2}(\Mn{U}{3}}\Khat\Mn{U}{1})$$
\item Utilise a ``dimension tree" to store and reuse partial products~\cite{PTC13a,LKLHS2017,HBJT18}. 
\begin{center}
\begin{tikzpicture}

\node (123) at (4,2) {$\{1,2,3\}$};
\node (12) at (2,1) {$\{1,2\}$};
\node (3) at (6,1) {$\Mn{M}{3}$};
\node (1) at (1,0) {$\Mn{M}{1}$};
\node (2) at (3,0) {$\Mn{M}{2}$};

\scriptsize
\path[draw] (123) edge [left,align=right] node {PM} (12);
\path[draw] (123) edge [right,align=left] node {PM} (3);
\path[draw] (12) edge [left,align=right] node {mTTV} (1);
\path[draw] (12) edge [right,align=left] node {mTTV} (2);
\normalsize

\end{tikzpicture}
\end{center}
\item PM = partial MTTKRP 
\item mTTV = multi-Tensor-Times-Vector
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Distributed Memory Optimisation - Communication Avoiding}

\begin{columns}

\begin{column}{.6\textwidth}

\newcommand{\procdim}{3}
\newcommand{\proc}{\draw[black,shift={(-.5,-.5)}] (0,0) grid (\procdim,\procdim);}
\newcommand{\highlight}{gray!75}
\newcommand{\commhighlight}{gray!25}

\begin{center}
\begin{tikzpicture}[x={(-0.5cm,-0.4cm)}, y={(1cm,0cm)}, z={(0cm,1cm)},every node/.append style={transform shape}]

% highlight all-gather/reduce-scatter patterns
% 1st slide is proc data (leave highlighting for all slides)
% 2nd slide is all-gather in 1st mode
\only<2>{
% front face
\begin{scope}[canvas is yz plane at x=.5]
	% highlight front face of proc comm
	\draw[fill=\commhighlight,shift={(-.5,-.5)}] (0,0) rectangle (3,1);
	% highlight block of 1st factor matrix
	\draw[fill=\commhighlight,shift={(-1,.5)},xscale=.5] (0,0) rectangle (-1,-1);
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=-90]
	% highlight right face of proc comm
	\draw[fill=\commhighlight,shift={(-.5,-.5)}] (0,0) rectangle (3,1);
\end{scope}
% top face
\begin{scope}[canvas is yx plane at z=.5,yscale=-1,rotate=0]
	% highlight top face of proc comm
	\draw[fill=\commhighlight,shift={(-.5,-.5)}] (0,0) rectangle (3,3);
\end{scope}
}
% 3rd slide is all-gather in 3rd mode
\only<3>{
% front face
\begin{scope}[canvas is yz plane at x=.5]
	% highlight front face of proc comm
	\draw[fill=\commhighlight,shift={(-.5,.5)}] (0,0) rectangle (3,-3);
	% highlight block of 1st factor matrix now owned by processor
	\draw[fill=\highlight,shift={(-1,.5)},xscale=.5] (0,0) rectangle (-1,-1);
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=-90]
	% highlight right face of proc comm
	\draw[fill=\commhighlight,shift={(-.5,.5)}] (0,0) rectangle (1,-3);
	% highlight block of 3rd factor matrix
	\draw[fill=\commhighlight,yscale=.5,shift={(-.5,-6)}] (0,0) rectangle (1,-1);
\end{scope}
% top face
\begin{scope}[canvas is yx plane at z=.5,yscale=-1,rotate=0]
	% highlight top face of proc comm
	\draw[fill=\commhighlight,shift={(-.5,-.5)}] (0,0) rectangle (3,1);
\end{scope}
}
% 4th and 5th slides are local computation
\only<4-5>{
% front face
\begin{scope}[canvas is yz plane at x=.5]
	% highlight block of 1st factor matrix now owned by processor
	\draw[fill=\highlight,shift={(-1,.5)},xscale=.5] (0,0) rectangle (-1,-1);
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=-90]
	% highlight block of 3rd factor matrix now owned by processor
	\draw[fill=\highlight,yscale=.5,shift={(-.5,-6)}] (0,0) rectangle (1,-1);
\end{scope}
}
\only<5>{
% front face
\begin{scope}[canvas is yz plane at x=.5]
	% highlight block of 2nd factor matrix contribution computed by processor
	\draw[fill=\highlight,shift={(1.5,-3)},yscale=.5] (0,0) rectangle (1,-1);
\end{scope}
}
% 6th slide is reduce-scatter in 2nd mode
\only<6>{
% front face
\begin{scope}[canvas is yz plane at x=.5]
	% highlight front face of proc comm
	\draw[fill=\commhighlight,shift={(1.5,.5)}] (0,0) rectangle (1,-3);
	% highlight block of 2nd factor matrix involved in reduce scatter
	\draw[fill=\commhighlight,shift={(1.5,-3)},yscale=.5] (0,0) rectangle (1,-1);
	% highlight block of 2nd factor matrix of computed output
	\draw[fill=\highlight,shift={(1.5,-3)},yscale=.5] (0,0) rectangle (1/9,-1);
	% highlight block of 1st factor matrix now owned by processor
	\draw[fill=\highlight,shift={(-1,.5)},xscale=.5] (0,0) rectangle (-1,-1);
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=-90]
	% highlight right face of proc comm
	\draw[fill=\commhighlight,shift={(-.5,.5)}] (0,0) rectangle (3,-3);
	% highlight block of 3rd factor matrix now owned by processor
	\draw[fill=\highlight,yscale=.5,shift={(-.5,-6)}] (0,0) rectangle (1,-1);
\end{scope}
% top face
\begin{scope}[canvas is yx plane at z=.5,yscale=-1,rotate=0]
	% highlight top face of proc comm
	\draw[fill=\commhighlight,shift={(1.5,-.5)}] (0,0) rectangle (1,3);
\end{scope}
}
\only<7>{
% front face
\begin{scope}[canvas is yz plane at x=.5]
	% highlight block of 2nd factor matrix of computed output
	\draw[fill=\highlight,shift={(1.5,-3)},yscale=.5] (0,0) rectangle (1/9,-1);
\end{scope}
}

% highlight one processor's data
% front face
\begin{scope}[canvas is yz plane at x=.5,shift={(1.5,-1.5)}]
	% highlight front face of tensor block
	\draw[fill=\highlight,shift={(0,1)}] (0,0) rectangle (1,1);
	% highlight block of 1st factor matrix
	\draw[fill=\highlight,shift={(-2.5,2-1/3)},xscale=.5] (0,0) rectangle (-1,-1/9);
	% highlight block of 2nd factor matrix
	\draw[shift={(0,-1.5)},yscale=.5] (0,0) rectangle (1/9,-1);
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=-90,shift={(-.5,-3)}]
	% highlight front face of tensor block
	\draw[fill=\highlight,shift={(0,2.5)}] (0,0) rectangle (1,1);
	% highlight block of 3rd factor matrix
	\draw[fill=\highlight,yscale=.5] (0,0) rectangle (1/9,-1);
\end{scope}
% top face
\begin{scope}[canvas is yx plane at z=.5,yscale=-1,rotate=0]
	% highlight front face of tensor block
	\draw[fill=\highlight,shift={(1.5,-.5)}] (0,0) rectangle (1,1);
\end{scope}

% draw tensor
% front face
\begin{scope}[canvas is yz plane at x=.5,rotate=-90]
	\proc
\end{scope}
% top face
\begin{scope}[canvas is yx plane at z=.5,yscale=-1,rotate=0]
	\proc
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=180]
	\proc
\end{scope}

% draw factor matrices
% front face
\begin{scope}[canvas is yz plane at x=.5,shift={(1.5,-.5)}]
	% draw 1st factor matrix
	\draw[shift={(-2.5,0)},xscale=.5] (0,-2) grid (-1,1);
	\node[draw=none] at (-3.5,-.5) {$\Mn{U}{1}$};
	% draw 2nd factor matrix
	\draw[shift={(0,-2.5)},yscale=.5] (-2,0) grid (1,-1);
	\node[draw=none] at (-.5,-3.5) {$\Mn{M}{2}$};
\end{scope}
% right face
\begin{scope}[canvas is zx plane at y=(\procdim-.5),rotate=-90,shift={(1.5,-.5)}]
	% draw 3rd factor matrix
	\draw[shift={(0,-2.5)},yscale=.5] (-2,0) grid (1,-1);
	\node[draw=none] at (-.5,-3.5) {$\Mn{U}{3}$};
\end{scope}

\end{tikzpicture}
\end{center}

\end{column}
\begin{column}{.45\textwidth}

\footnotesize
Each processor
\begin{enumerate}
	\item Starts with one subtensor and subset of rows of each input factor matrix
	\uncover<2->{\item All-Gathers all the rows needed from $\Mn{U}{1}$}
	\uncover<3->{\item All-Gathers all the rows needed from $\Mn{U}{3}$}
	\uncover<4->{\item Computes its contribution to rows of $\Mn{M}{2}$ (local MTTKRP)}
	\uncover<6->{\item Reduce-Scatters to compute and distribute $\Mn{M}{2}$ evenly}
	\uncover<7->{\item Solve local NLS problem using $\Mn{M}{2}$ and $\Mn{U}{2}$}
\end{enumerate}

\end{column}

\end{columns}

\end{frame}

\begin{frame}
\frametitle{Performance Plots - Strong Scaling}
\begin{figure}
\includegraphics[width=0.7\textwidth]{strsca4d}
\caption{Strong Scaling on synthetic tensor of dimensions $256 \times 256 \times 256 \times 256$ on 8, 16, 32, 64 and 128 nodes of Titan}
\end{figure}
\begin{center}
Can achieve \textbf{nearly linear} scaling since NNCP is compute bound.
\end{center}
\end{frame}

\begin{frame}
\frametitle{Performance Plots - CPU vs GPU}
\begin{columns}[onlytextwidth]
\begin{column}{.5\textwidth}
\begin{figure}
  \includegraphics[width=\textwidth]{cpuvsgpulr810}
  \caption{CPU}
\end{figure}
\end{column}
\hfill
\begin{column}{.5\textwidth}
\begin{figure}
  \includegraphics[width=\textwidth]{cpuvsgpulr811}
  \caption{GPU}
\end{figure}
\end{column}
\end{columns}
\begin{center}
\footnotesize{4D synthetic tensor of dimensions $384 \times 384 \times 384 \times 384$ on 81 Titan nodes as a $3 \times 3 \times 3 \times 3$ grid with varying low rank.} 
\end{center}
\begin{center}
Offloading DGEMM calls to GPU can provide \textbf{7X speedup}.
\end{center}
\end{frame}

\begin{frame}
\frametitle{Compiler Challenges and Extensions to the Sparse Setting}
\begin{enumerate}
\item Dimension Tree ordering.
	\begin{itemize}
	\item Combinatorial explosion for sparse case (contrasted with the single split choice for the dense case).
	\item Sparse case involves growth in intermediate values.
	\end{itemize}
\item Communication Pattern establishment and load balancing.
	\begin{itemize}
	\item Automatic communicator setup given a processor grid and tensor operation.
	\item Automatic data distribution using the communication-avoiding loop optimisation~\cite{knight2015communication, DR16arxiv}.
	\end{itemize}
\item Block Parallelism in Least Squares Solvers.
	\begin{itemize}
	\item Active Set orderings can be grouped in an embarrassingly parallel call.
	\item Sparse case with masking matrix has a similar RHS pattern.
	\end{itemize}
\item Binary Bloat
	\begin{itemize}
	\item Separate binaries for GPU/CPU and Sparse/Dense.
	\end{itemize}
\end{enumerate}
\end{frame}

%\againframe{summary}
\begin{frame}[label=summary]
\frametitle{Summary}
\begin{itemize}
\item PLANC is an open source, scalable and flexible software package to compute Non-negative Tensor Factorisation.
\item Implements state of the art communication avoiding algorithm for matricised-tensor times Khatri-Rao product (MTTKRP).
\item Popular optimisation methods like Block Principal Pivoting, Alternating Direction Method of Multipliers, first order Nesterov methods etc. for Non-negative Least Squares are included. 
\item NTF is an important contributor towards explainable AI with a wide range of applications like spectral unmixing, scientific visualization, healthcare analytics, topic modelling etc.
\item Coming soon as a miniapp on OLCF machines. \href{https://github.com/ramkikannan/planc/}{https://github.com/ramkikannan/planc}.
\end{itemize}
\end{frame}

\begin{frame}[allowframebreaks]
\frametitle{References}
\tiny
\bibliographystyle{alpha}
\bibliography{slides}
\normalsize
\end{frame}

\appendix

\begin{frame}
\frametitle{NNCP Algorithm}
\begin{itemize}
\item For given rank $R$ we formulate NNCP as the following optimisation problem,
\begin{align*}
\underset{\M{U},\M{V},\M{W} \geq \M{0}}{\min}\norm{\T{X}-\sum_{r=1}^R \lambda_r\MC{U}{r}\circ\MC{V}{r}\circ\MC{W}{r}}
\end{align*}
\item Non-linear and non-convex problem.
\item Solve via Alternating Non-negative Least Squares in an iterative manner using Block Coordinate Descent.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Alternating Non-negative Least Squares (ANLS)}
\begin{itemize}
\item Fixing all factor matrices but one results in a linear NLS problem,
\begin{align*}
&\underset{\M{U} \geq \M{0}}{\min}\norm{\T{X}-\sum_{r=1}^R \MC{U}{r}\circ\MC[\hat]{V}{r}\circ{\MC[\hat]{W}{r}}} \\
&\textnormal{or equivalently,}\\
&\underset{\M{U} \geq \M{0}}{\min} \norm{\Mz{X}{1} -\M{U}(\M[\hat]{W} \Khat \M[\hat]{V})^\Tra}_F
\end{align*}
\item $\Khat$ is the Khatri-Rao product (column-wise Kronecker Product) of the factor matrices.
\item Utilising the identity $(\M{A}\Khat\M{B})^\Tra(\M{A}\Khat\M{B}) = \M{A}^\Tra\M{A} \Hada \M{B}^\Tra\M{B}$, we can cast the above problem as NLS solves via normal equations.
$$\underset{\M{U} \geq \M{0}}{\min} \norm{\Mz{X}{1}(\M[\hat]{W} \Khat \M[\hat]{V}) -\M{U}\pr{\M[\hat]{W}^\Tra\M[\hat]{W} \Hada \M[\hat]{V}^\Tra\M[\hat]{V}}}_F$$
where $\Hada$ is the Hadamard Product.
\end{itemize}
\end{frame}

\end{document}
